### 2012-11-28 synmut.tex
Initial commit

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ARTICLE ABOUT FATE OF SYNONYMOUS MUTATIONS IN HIV
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[12pt,a4paper,notitlepage,onecolumn]{article}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{(Rise and fall of synonymous mutations)}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym}
\usepackage{color}
\usepackage{subfig}
\usepackage{graphicx}
\usepackage[font=small, format=hang, labelfont={sf,bf}, figurename=Fig.]{caption}
\usepackage{natbib}
\usepackage{pslatex}
\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\DeclareMathOperator\de{d\!}
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{u}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{\Title}
\author{\Author}
\date{\today}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\maketitle

\begin{abstract}
\noindent
Intrapatient HIV evolution is goverened by selection on the protein level in the
arms race with the immune system (killer T-cells and antibodies). Synonymous
mutations do not have an immunity-related phenotype and are often assumed to be
neutral. In this paper, we show that synonymous changes in epitope-rich regions
are often deleterious but still reach frequencies of order one.  We analyze time
series of viral sequences from the V1-C5 part of {\it env} within individual
hosts and observe that synonymous derived alleles rarely fix in the
viral population. Simulations suggest that such synonymous mutations
have a (Malthisuan) selection coefficient of the order of $-0.001$, and that
they are brought up to high frequency by linkage to neighbouring beneficial
nonsynonymous alleles (genetic draft). As far as the biological causes are
concerned, we detect a negative correlation between fixation of an allele and
its involvement in evolutionarily conserved RNA stem-loop structures.
This phenonenon is not observed in other parts of the HIV genome, in which
selective sweeps are less dense and the genetic architecture less constrained.
\end{abstract}

\section{Introduction}
\section{Results}

\begin{figure}
\begin{center}
\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_nonsynp8}
\caption{Allele frequency trajectories of typical patient, C3-V5, nonsynonymous
(solid) and synonymous mutations (dashed lines). Most synonymous mutations are
not fixed.}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.49\linewidth]{Bunnik2008_fixmid_syn_ShankanonShanka}}
\subfloat{\includegraphics[width=0.49\linewidth]{Shankarappa_fixmid_syn_V_regions}}
\caption{Fixation probability of derived synonymous alleles is strongly
suppressed in C3-V5 versus other parts of the {\it env} gene (left panel).
Especially hard is fixation of new alleles in conserved regions flanking the V
loops (right panel). The black dashed line is the prediction from neutral
theory, for comparison purposes.}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.49\linewidth]{fixation_probability_shortgenome_N_1e4_epitopes_example_longer}}
\subfloat{\includegraphics[width=0.49\linewidth]{fixation_probability_N_1e4_3ingredients_3rddel_envnonenv_stopall}}
\caption{Simulations show that the suppression of fixation probability can be
generated by linkage to sweeping nonsynonymous alleles nearby. Two possible
scenarios are competition between escape mutants (left panel) and time-dependent
selection due to immune sytem recognition (right panel).}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.49\linewidth]{mixed_Shankarappa_Bunnik2008_Liu_fixation_reactivity_Vandflanking_fromSHAPE}}
\subfloat{\includegraphics[width=0.49\linewidth]{mixed_Shankarappa_Bunnik2008_Liu_fixation_reactivity_nonVandflanking}}
\caption{Watts et al. have measured the reactivity of HIV nucleotides to {\it in
vitro} chemical attack and shown that some nucleotides are more likely to be
involved in RNA secondary folds. C1-C5 regions, in particular, show conserved
stem-loop structures~\citep{watts_architecture_2009}. We show that among all
derived alleles in those regions reaching frequencies of order one, there is a negative
correlation between fixation and involvement in a base pairing in a RNA stem
(left panel). The rest of the genome does not show any correlation (right
panel). There might be too few silent polymorphisms in the first place, or the
signal might be masked by a lot of non-functional RNA structures.}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.49\linewidth]{Henn_density_polymorphisms}}
\subfloat{\includegraphics[width=0.49\linewidth]{Henn_density_polymorphisms_syn_over_chances}}
\caption{The total density of polymorphisms (mostly nonsynonymous ones) is
highest in the V regions (left panel). The density of synonymous mutations only,
however, is not enriched there (right panel). This could be due to a more
deleterious effect of synonymous mutations.}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.49\linewidth]{fixation_loss_shortgenome_area_ada_frac_del_eff_coi_0_01_nescepi_6_heat.pdf}}
\subfloat{\includegraphics[width=0.49\linewidth]{fixation_loss_shortgenome_area_ada_frac_del_eff_coi_0_01_nescepi_6_nonsyn_heat.pdf}}

\caption{Simulations on the escape competition scenario show that the density of
 selective sweeps and the size of the deleterious effects of synonymous
 mutations are the main driving forces of the phenomenon. A convex fixation
 probability is recovered, as seen in the data, along the diagonal (left panel):
 more dense sweeps can support more deleterious linked mutations. The density of
 sweeps is limited, however, by the nonsynonymous fixation probability, which is
 quite close to neutrality (right panel). Moreover, strong competition between
 escape mutants is required, so that several escape mutants are ``found'' by HIV
within a few months of antibody production.}

\end{center}
\end{figure}

\section{Discussion}
\section{Methods}
\section*{Acknowledgements}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{natbib}
\bibliography{bib}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2012-11-29 synmut.tex
Introduction draft

@@
 \newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
-\newcommand{\mut}{u}
+\newcommand{\mut}{\mu}
 \newcommand{\mfit}{\langle F\rangle}
@@
 \newcommand{\OO}{\mathcal{O}}
+\newcommand{\env}{\textit{env}}
+\newcommand{\rev}{\textit{rev}}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \section{Introduction}
+
+HIV evolves rapidly within a single host during the course of the infection. The
+driving forces shaping this process are the high mutation rate and the strong
+selection imposed by the host immune system via a wealth of mechanisms, notably
+killer T cells (CTLs) and neutralizing
+antibodies~\citep{pantaleo_immunopathogenesis_1996}.
+
+In a nutshell, when the host develops a CTL or antibody respose against a
+particular viral epitope, rare HIV variants carrying mutated versions of the
+epitope, called {\it escape mutants}, acquire a fitness advantage and spread
+rapidly in the viral population, within a few months (see
+\figurename~\ref{fig:aft}, solid lines). During chronic infection, the
+(Malthusian) effect size of this beneficial mutations is of the order of
+$0.01$~\citep{neher_recombination_2010}. The viral \env{} gene shows the fastest
+rates of adaptation, because is both rich of CTL epitopes and targeted by
+antibodies; its sequence diverges at rates of the order of $1\%$ per
+year~\citep{shankarappa_consistent_1999}.
+
+Many nucleotide polymorphisms are escape mutations, and in particular are
+nonsynonymous, i.e. they appear in protein coding regions and change the amino
+acid sequence. Nonetheless, nucleotide changes unrelated to immune escape are
+seen, in \env{} and elsewhere, and some of them become abundant alike, often
+rapidly. In particular, it is not uncommon for synonymous mutations to reach
+frequencies of order one within months from their first appearance (see
+\figurename~\ref{fig:aft}, dashed lines). The biological function of these
+mutations in the economy of HIV is not well understood. By definition, the
+immunological phenotype, which is decided at the protein level, is unaffected,
+but other biological and ecological aspects of the viral lifestyle might be
+involved. In practice, a couple of RNA-level phenotypes are
+known. For example, within \env{} a certain RNA sequence, called \rev{}
+response element (RRE), is used by HIV to enhance nuclear export of some of its
+transcripts~\citep{fernandes_hiv-1_2012}. Another case is the interaction
+between viral reverse transcriptase, viral ssRNA, and the host
+tRNA$^\text{Lys3}$: the latter is required for priming viral replication and
+bound by a specifical pseudoknotted RNA structure in the viral 5' untranslated
+region~\citep{barat_interaction_1991, paillart_vitro_2002}.
+
+Crucially for evolutionary studies, the minor phenotypes caused by synonymous
+mutations might have an effect on viral fitness. For instance, recent studies
+have shown that genetically engineered HIV strains with skewed codon usage bias
+(CUB) patterns towards more or less abundant tRNAs replicate better or worse,
+respectively~\citep{ngumbela_quantitative_2008, li_codon-usage-based_2012}.
+In this study, we try to characterize the fitness effects of synonymous
+polymorphisms that, at some point during the infection, become abundant in the
+viral population.
+
+One simple way to assess the neutrality of synonymous mutations is to look at
+their level of conservation. Deleterious mutations at functional sites are
+expected to be absent or rare across the viral population; vice versa, mutant
+alleles that reach high frequencies are expected to be neutral. Confirmatorily,
+population genetics shows that the equilibrium frequency of a deleterious allele
+with fitness $-s$ is $\mut / |s|$, where $\mut$ is the mutation rate per site
+per generation; neutral alleles have no equilibrium frequency and can slowly fix
+via genetic drift~\citep{ewens_mathematical_2004}. This approach, albeit
+intuitive, works only under the assumption of independent sites. If the focal
+synonymous mutant is linked to another, nonneutral allele, its frequency is the
+result of the combined fitness effects of both sites. Since recombination in HIV
+is known to be rather rare~\citep{neher_recombination_2010,
+batorsky_estimate_2011}, the genetic context of the synonymous change at hand
+must be taken into account.
+
 \section{Results}
-
+We start from time series of viral nucleotide sequences from single patients,
+which span several years of chronic
+infection~\citep{shankarappa_consistent_1999, bunnik_autologous_2008,
+liu_selection_2006}. Plotting the allele frequencies against time for all
+polymorphic sites, it is evident that, although nonsynonymous changes are
+widespread, synonymous ones are also present in various occasions at high
+frequency (see \figurename~\ref{fig:aft}).
 \begin{figure}
@@
 (solid) and synonymous mutations (dashed lines). Most synonymous mutations are
-not fixed.}
+not fixed. Colors are set according to the position of the site along the C3-V5
+region (red to blue). Data from Ref.~\cite{shankarappa_consistent_1999}.}
+\label{fig:aft}
 \end{center}
@@
 loops (right panel). The black dashed line is the prediction from neutral
-theory, for comparison purposes.}
+theory, for comparison purposes. Data from
+Refs.~\cite{shankarappa_consistent_1999, bunnik_autologous_2008}.}
 \end{center}
@@
 \begin{center}
-\subfloat{\includegraphics[width=0.49\linewidth]{fixation_probability_shortgenome_N_1e4_epitopes_example_longer}}
-\subfloat{\includegraphics[width=0.49\linewidth]{fixation_probability_N_1e4_3ingredients_3rddel_envnonenv_stopall}}
-\caption{Simulations show that the suppression of fixation probability can be
-generated by linkage to sweeping nonsynonymous alleles nearby. Two possible
-scenarios are competition between escape mutants (left panel) and time-dependent
-selection due to immune sytem recognition (right panel).}
+\includegraphics[width=\linewidth]{Shankarappa_fix_loss_dt_times}
+\caption{Fixation or extinction times for synonymous alleles starting from
+intermediate frequencies. The colored bands are the final fixation probabilities
+expected from neutral theory; the observed alleles are fixed less frequently
+than expected. The timescale of fixation/extinction is approximately 500 days,
+corresponding to a selective effect of $\sim -0.001$.}
+\label{fig:fixtimes}
 \end{center}
@@
 panel). There might be too few silent polymorphisms in the first place, or the
-signal might be masked by a lot of non-functional RNA structures.}
+signal might be masked by a lot of non-functional RNA structures. Data from
+Refs.~\cite{shankarappa_consistent_1999, bunnik_autologous_2008,
+liu_selection_2006}.}
 \end{center}
@@
 
+%\begin{figure}
+%\begin{center}
+%\subfloat{\includegraphics[width=0.49\linewidth]{Henn_density_polymorphisms}}
+%\subfloat{\includegraphics[width=0.49\linewidth]{Henn_density_polymorphisms_syn_over_chances}}
+%\caption{The total density of polymorphisms (mostly nonsynonymous ones) is
+%highest in the V regions (left panel). The density of synonymous mutations only,
+%however, is not enriched there (right panel). This could be due to a more
+%deleterious effect of synonymous mutations.}
+%\end{center}
+%\end{figure}
+
 \begin{figure}
 \begin{center}
-\subfloat{\includegraphics[width=0.49\linewidth]{Henn_density_polymorphisms}}
-\subfloat{\includegraphics[width=0.49\linewidth]{Henn_density_polymorphisms_syn_over_chances}}
-\caption{The total density of polymorphisms (mostly nonsynonymous ones) is
-highest in the V regions (left panel). The density of synonymous mutations only,
-however, is not enriched there (right panel). This could be due to a more
-deleterious effect of synonymous mutations.}
+\subfloat{\includegraphics[width=0.49\linewidth]{fixation_probability_shortgenome_N_1e4_epitopes_example_longer}}
+\subfloat{\includegraphics[width=0.49\linewidth]{fixation_probability_N_1e4_3ingredients_3rddel_envnonenv_stopall}}
+\caption{Simulations show that the suppression of fixation probability can be
+generated by linkage to sweeping nonsynonymous alleles nearby. Two possible
+scenarios are competition between escape mutants (left panel) and time-dependent
+selection due to immune sytem recognition (right panel).}
 \end{center}

### 2012-12-02 synmut.tex
draft0

--- 2012-11-28 synmut.tex
@@
 selective sweeps are less dense and the genetic architecture less constrained.
+%In absence of antiretroviral treatment, HIV is very successful at producing
+%mutants that are able to stay undetected by the host immune system for months,
+%boosting the infection~\citep{richman_rapid_2003, bunnik_autologous_2008,
+%moore_limited_2009}. The high mutation rate at the core of this process,
+%however, also generates genetic noise in the background of beneficial alleles.
+%Because of the limited ourcrossing of HIV, hitchhikers usually stay linked to
+%the focal allele for a time comparable to its sweeping time, i.e. a few
+%months~\citep{neher_recombination_2010, batorsky_estimate_2011}. The later fate
+%of these accessory mutations depends more and more, as they decouple genetically
+%from the escape allele via recombination, on their own fitness effects. In this
+%study we show that, in a genetic region particularly dense of selective sweeps,
+%synonymous hitchhikers tend to revert on a time scale of the order of 500 days,
+%because of their deleterious fitness effects. In addition, we provide evidence
+%that the biological origin for this deleterious effects resides, at least
+%partially, in the disruption of macroevolutionarily conserved RNA secondary
+%structures, termed ``insulating stems''.
 \end{abstract}
@@
 \section{Introduction}
+
+HIV evolves rapidly within a single host during the course of the infection. The
+driving forces shaping this process are the high mutation rate and the strong
+selection imposed by the host immune system via killer T cells (CTLs) and
+neutralizing antibodies (AB)~\citep{pantaleo_immunopathogenesis_1996}.
+
+When the host develops a CTL or AB respose against a particular viral epitope,
+rare HIV variants carrying mutated versions of the epitope, called {\it escape
+mutants}, acquire a fitness advantage and spread rapidly in the viral
+population, within a few months (see \figurename~\ref{fig:aft}, solid lines).
+During chronic infection, the (Malthusian) effect size of this beneficial
+mutations is of the order of $s_e \sim 0.01$~\citep{neher_recombination_2010}.
+The viral \env{} gene shows the fastest rates of adaptation, because is both
+rich of CTL epitopes and targeted by antibodies; its sequence diverges at rates
+of the order of $1\%$ per year~\citep{shankarappa_consistent_1999}.
+
+Many nucleotide polymorphisms that reach high frequency in the viral population
+are escape mutations: all of them are {\it nonsynonymous} (i.e. they appear in
+protein coding regions and change the amino acid sequence). Nonetheless,
+nucleotide changes unrelated to immune escape are not rare, in \env{} and
+elsewhere, and some of them become abundant alike, often rapidly. In particular,
+it is not uncommon for {\it synonymous} mutations to reach frequencies of order
+one within months from their first appearance (see \figurename~\ref{fig:aft},
+dashed lines). The biological function of these mutations within the economy of
+HIV is not well understood. By definition, the immunological phenotype, which is
+decided at the protein level, is unaffected, but other biological and ecological
+aspects of the viral lifestyle might be involved. In practice, a couple of
+RNA-level phenotypes are known. For example, within \env{} a certain RNA
+sequence, called \rev{} response element (RRE), is used by HIV to enhance
+nuclear export of some of its transcripts~\citep{fernandes_hiv-1_2012}. Another
+well studied case is the interaction between viral reverse transcriptase, viral
+ssRNA, and the host tRNA$^\text{Lys3}$: the latter is required for priming
+reverse transcription (RT) and bound by a specifical pseudoknotted RNA structure
+in the viral 5' untranslated region~\citep{barat_interaction_1991,
+paillart_vitro_2002}.
+
+Crucially for evolutionary purposes, the minor phenotypes caused by synonymous
+mutations might have an effect on viral fitness. For instance, recent studies
+have shown that genetically engineered HIV strains with skewed codon usage bias
+(CUB) patterns towards more or less abundant tRNAs replicate better or worse,
+respectively~\citep{ngumbela_quantitative_2008, li_codon-usage-based_2012}. In
+this study, we try to characterize the fitness effects of synonymous
+polymorphisms that, at some point during chronic infection, become abundant in
+the viral population.
+
+One simple way to assess the neutrality of synonymous mutations is to look at
+their level of conservation. Deleterious mutations at functional sites are
+expected to be absent or rare across the viral population; vice versa, mutant
+alleles that reach high frequencies are expected to be neutral. If genetic sites
+are independent, the equilibrium frequency of a deleterious allele with fitness
+$-s$ is $\mut / |s|$, where $\mut$ is the mutation rate per site per generation;
+neutral alleles have no equilibrium frequency and can slowly fix via genetic
+drift~\citep{ewens_mathematical_2004}. If the focal synonymous mutant is linked
+to another nonneutral allele, however, its frequency is the result of the
+combined fitness effects of both sites, and simple conservation-level analyses
+fail. Since recombination in HIV is known to happen
+rarely~\citep{neher_recombination_2010, batorsky_estimate_2011}, the genetic
+context of the synonymous change at hand must be taken into account. Our
+results underline the importance of the latter scenario for intrapatient HIV
+evolution.
+
 \section{Results}
-
+We analyze from time series of viral nucleotide sequences from single patients,
+which span several years of chronic
+infection~\citep{shankarappa_consistent_1999, bunnik_autologous_2008,
+liu_selection_2006}. Plotting the allele frequencies against time for all
+polymorphic sites, it is evident that, although nonsynonymous changes are
+widespread, synonymous ones are also present in various occasions at high
+frequency (see \figurename~\ref{fig:aft}).
 \begin{figure}
@@
 (solid) and synonymous mutations (dashed lines). Most synonymous mutations are
-not fixed.}
+not fixed. Colors are set according to the position of the site along the C3-V5
+region (red to blue). Data from Ref.~\cite{shankarappa_consistent_1999}.}
+\label{fig:aft}
 \end{center}
 \end{figure}
+Moreover, many mutations appear to change rapidly in frequency as a flock,
+especially at the lower end of the frequency spectrum. These observations lead
+to the idea that linkage might be widespread at the time scale of the typical
+selective sweep, i.e. a few months~\citep{neher_recombination_2010}. It seems
+therefore likely that synonymous mutations are not attaining high frequencies by
+genetic drift but are rather exponentially amplified by selection on linked
+nonsynonymous sites, a process known as {\it genetic
+draft}~\citep{neher_genetic_2011}.
 
+To test this hypothesis, we focus on the long-time fate of such synonymous
+derived alleles {\it after} they are already abundant. According to the neutral
+theory of independent sites, (a) an allele starting at frequency $\nu$ will
+reach fixation, in the long-time limit, with a probability $P_f(\nu) = \nu$, and
+get extinct otherwise; (b) the time required to reach either boundary, fixation
+or extinction, is essentialy the inverse of the population size (in
+generations), hence much longer than five years for HIV~\citep{boltz_ultrasensitive_2012}.
 \begin{figure}
@@
 \subfloat{\includegraphics[width=0.49\linewidth]{Shankarappa_fixmid_syn_V_regions}}
-\caption{Fixation probability of derived synonymous alleles is strongly
-suppressed in C3-V5 versus other parts of the {\it env} gene (left panel).
-Especially hard is fixation of new alleles in conserved regions flanking the V
-loops (right panel). The black dashed line is the prediction from neutral
-theory, for comparison purposes.}
+\caption{Left panel: fixation probability of derived synonymous alleles is strongly
+suppressed in C3-V5 versus other parts of the {\it env} gene, and of
+nonsynonymous ones.
+Right panel: especially hard is fixation of new alleles in conserved regions flanking the V
+loops. The black dashed line is the prediction from neutral
+theory, for comparison purposes. Data from
+Refs.~\cite{shankarappa_consistent_1999, bunnik_autologous_2008}.}
+\label{fig:fixp}
 \end{center}
@@
 
+The left panel of \figurename~\ref{fig:fixp} shows the fixation probability of
+derived alleles seen within various frequency windows. With regard to synonymous
+alleles, the average over the whole HIV genome (organce line) matches closely
+the expected result from neutral theory (dashed line). A different picture
+emerges when the most variable part of the genome, V1-C5 region of \env{},
+is zoomed into (red line). Two independent datasets, from
+refs.~\citep{shankarappa_consistent_1999, bunnik_autologous_2008} yield the same
+result, i.e. that synonymous mutations hardly fix at all.
+
+A related observation can be made on the nonsynonymous alleles (blue line).
+Although we see no difference between different parts of the HIV genome, the
+neutral-like shape of $P_f$ is at odds with the na\"ive expectation that most
+nonsynonymous alleles reaching high frequency are to-be-fixed escape mutations.
+
+Moreover, as illustrated in \figurename~\ref{fig:fixtimes}, the time for
+synonymous alleles to reach the first boundary starting from intermediate
+frequencies is of the order of 500 days which, assuming a conservative
+generation time of one day~\citep{perelson_hiv-1_1996}, correspond to the same
+number of generations. This time would only be possible in a neutral scenario
+if the population size of HIV were extremely small, of the order of $10^3$, much
+smaller then both theory and experiments
+indicate~\citep{boltz_ultrasensitive_2012}.
+% WE SHOULD REALLY CITE OURSELVES IN THE (HOPEFUL) FUTURE HERE :-)
+\begin{figure}
+\begin{center}
+\includegraphics[width=\linewidth]{Shankarappa_fix_loss_dt_times}
+\caption{Fixation or extinction times for synonymous alleles starting from
+intermediate frequencies. The colored bands are the final fixation
+probabilities expected from neutral theory; the observed alleles are fixed
+less frequently than expected. The timescale of fixation/extinction is
+approximately 500 days, corresponding to a selective effect of $\sim -0.001$.}
+\label{fig:fixtimes}
+\end{center}
+\end{figure}
+
+Taken together, these findings suggest that synonymous (and some nonsynonymous)
+alleles are not propagating independently by genetic drift, but rather
+hitchhiking on neighbouring beneficial alleles. This conclusion is also
+consistent with the currently accepted estimate for the outcrossing rate of HIV,
+of the order of $1\%$ per generation. Although synonymous alleles can be drafted
+to high frequency by linkage and selection, they decouple on a time scale of the
+order of 100 days, which is also the average time it takes a (nonsynonymous)
+selected allele to spread through the population during chronic
+infection~\citep{neher_recombination_2010}.
+
+\figurename~\ref{fig:fixp} (left panel) also suggests that at least some
+synonymous mutations in the V1-C5 region should be deleterious. We analyze this
+hypothesis in two ways. First of all, we look for possible biological causes for
+such a behaviour. Furthermore, we perform computer simulations of viral
+evolution under a wide range of parameters in order to reproduce the results for
+the fixation probability and study the regions of the parameter space compatible
+with the observations.
+
+One possible {\it a priori} explanation for deleterious fitness effects of
+synonymous mutations is the presence of secondary structures in the viral RNA.
+If any RNA secondary structures are functionally relevant for HIV replication,
+mutations in nucleotides involved in those base pairs are expected to revert
+more easily than by pure chance, to keep those structures in place. The
+propensity of nucleotides to form base pairs in HIV has been measured
+recently\citep{watts_architecture_2009}, and the HIV genome has resulted to be
+highly structured~\citep{forsdyke_reciprocal_1995, watts_architecture_2009}.
+Watts {\it et al.} have measured the propensity of any nucleoside to be attacked by
+external chemicals (SHAPE reactivity) and shown that a higher reactivity
+corresponds to a smaller likelihood of being involved in an RNA helix. As far as
+the V1-C5 region is concerned, it has been shown that the short C1-5 regions
+have a high tendency to form {\it insulating
+stems}~\citep{watts_architecture_2009, sanjuan_interplay_2011}. Since the V1-5
+loops have to be highly variable to ensure immune escape, their destabilizing
+contribution to the viral RNA structure is though to be buffered by the
+surrounding, base-pair rich stems. To test whether or not insulating stems might
+be the reason for the reduced fixation probability, we partition the data
+depending on their position in the V1-C5 region. Consistently with the
+expectations, $P_f$ is more heavily readuced in the stems than in
+the loops (see \figurename~\ref{fig:fixp}, right panel).
+
+Despite this indirect result supporting the involvement of RNA secondary
+structure in the deleterious effects of synonymous mutations in V1-C5, we set
+out to seek more direct evidence. We partition all synonymous alleles observed
+at intermediate frequencies above 10-15\% depending on their final destiny
+(fixation or extinction). Subsequently, we align our sequences to the reference
+NL4-3 strain used in ref.~\citep{watts_architecture_2009} and assign them SHAPE
+reactivities. As shown in \figurename~\ref{fig:SHAPE} (left panel) in a
+cumulative histogram, the reactivity of fixed alleles are systematically larger
+than of alleles that are doomed to extinction. In other words, alleles that are
+likely to be breaking RNA helices are also more likely to revert and finally be
+lost from the population.
+\begin{figure}
+\begin{center}
+\subfloat{\includegraphics[width=0.49\linewidth]{mixed_Shankarappa_Bunnik2008_Liu_fixation_reactivity_Vandflanking_fromSHAPE}}
+\subfloat{\includegraphics[width=0.49\linewidth]{mixed_Shankarappa_Bunnik2008_Liu_fixation_reactivity_nonVandflanking}}
+\caption{Watts et al. have measured the reactivity of HIV nucleotides to {\it
+in vitro} chemical attack and shown that some nucleotides are more likely to
+be involved in RNA secondary folds. C1-C5 regions, in particular, show
+conserved stem-loop structures~\citep{watts_architecture_2009}. We show that
+among all derived alleles in those regions reaching frequencies of order one,
+there is a negative correlation between fixation and involvement in a base
+pairing in a RNA stem (left panel). The rest of the genome does not show any
+correlation (right panel). There might be too few silent polymorphisms in the
+first place, or the signal might be masked by non-functional RNA
+structures. Data from Refs.~\cite{shankarappa_consistent_1999,
+bunnik_autologous_2008, liu_selection_2006}.}
+\label{fig:SHAPE}
+\end{center}
+\end{figure}
+
+Since the insulating stems are not the only RNA secondary structures found by
+the authors of ref.~\citep{watts_architecture_2009}, we repeated the same
+partitioning based on fate on the rest of the HIV genome. As shown in the right
+panel of \figurename~\ref{fig:SHAPE}, there is no evidence that a similar
+mechanism is acting outside of the V1-C5 region. This is not fully surprising,
+because it is well known that single stranded RNA chains tend to form a lot of
+spontaneous structures, most of which have no biological function. Furthermore,
+the averaging over many different genomic regions might worsen the
+signal-to-noise ratio, which is already quite low, and result in
+indistinguishable histograms.
+
+In addition to RNA secondary structure, we have considered other possible
+explanations for a fitness effect of synonymous mutations, in particular codon
+usage bias (CUB). HIV is known to prefer A-rich codons over highly expressed
+human housekeeping genes~\citep{jenkins_extent_2003}. Moreover, codon-optimized
+and -pessimized viruses have recently been generated and shown to replicate
+better or worse than wild type strains,
+respectively~\citep{li_codon-usage-based_2012, ngumbela_quantitative_2008,
+coleman_virus_2008}. We do not found, however, evidence for any contribution of
+CUB to the ultimate fate of synonymous alleles. Several lines of thought support
+this result. First of all, although codon-optimized HIV seems to perform better
+{\it in vitro}, the distance in CUB between HIV and human genes is not shrinking
+at the macroevolutionary level. Second, within a single patient, we do not
+observe any bias towards more human-like CUB in the synonymous mutations that
+reach fixation rather than extinction. Third, it is a common phenomenon for
+retroviruses to use variously different codons from their hosts, and CUB effects
+on fitness are thought to be so small that divergent nucleotide composition has
+been suggested as a possible mechanism for viral
+speciation~\citep{bronson_nucleotide_1994}. Fourth, CUB in the V1-C5 region is
+not very different from other parts of the HIV genome, whereas the reduced
+fixation probability is only observed there. In conclusion, although we cannot
+exclude an effect of CUB on fitness as a general rule, we expect it to be a
+minor effect in our context.
+
+We also perform computer simulations of evolving viral population under
+selection and rare recombination. For this purpose, we use the recently
+published package FFPopSim, which includes a module dedicated to intrapatient
+HIV evolution~\citep{zanini_ffpopsim:_2012}. We analyze many combination of
+parameters such as population size, recombination rate, selection coefficient
+and density of escape mutations, epitope length, deleterious effect of
+synonymous mutations. The purpose of the simulations is to probe the population
+genetics mechanisms required to reproduce the fixation probabilities of
+\figurename~\ref{fig:fixp}, simultaneously for synonymous and nonsynonymous
+alleles.
+
+The main result of the simulations is that genetic draft can indeed generate a
+depression in fixation probabilities, but only under relatively tight
+circumstances. In terms of model confidence, this is a good result: the
+probability of obtaining the experimental curves by pure chance is negligible.
+\figurename~\ref{fig:simfixp} shows two instances of successful simulations.
 \begin{figure}
@@
 selection due to immune sytem recognition (right panel).}
+\label{fig:simfixp}
 \end{center}
@@
 
-\begin{figure}
-\begin{center}
-\subfloat{\includegraphics[width=0.49\linewidth]{mixed_Shankarappa_Bunnik2008_Liu_fixation_reactivity_Vandflanking_fromSHAPE}}
-\subfloat{\includegraphics[width=0.49\linewidth]{mixed_Shankarappa_Bunnik2008_Liu_fixation_reactivity_nonVandflanking}}
-\caption{Watts et al. have measured the reactivity of HIV nucleotides to {\it in
-vitro} chemical attack and shown that some nucleotides are more likely to be
-involved in RNA secondary folds. C1-C5 regions, in particular, show conserved
-stem-loop structures~\citep{watts_architecture_2009}. We show that among all
-derived alleles in those regions reaching frequencies of order one, there is a negative
-correlation between fixation and involvement in a base pairing in a RNA stem
-(left panel). The rest of the genome does not show any correlation (right
-panel). There might be too few silent polymorphisms in the first place, or the
-signal might be masked by a lot of non-functional RNA structures.}
-\end{center}
-\end{figure}
+First of all, in order for $P_f$ to be reduced to levels seen in the data, most
+synonymous mutants habe to be slightly deleterious. The reason is that if
+synonymous alleles are an even mixture of deleterious and neutral ones, the
+latter are much more likely to be carried to high frequencies by genetic draft,
+because they present no burden to escape mutations. The synonymous $P_f$ is then
+dominated by these neutral alleles and the depression induced by the few
+deleterious ones becomes undetectable.
 
-\begin{figure}
-\begin{center}
-\subfloat{\includegraphics[width=0.49\linewidth]{Henn_density_polymorphisms}}
-\subfloat{\includegraphics[width=0.49\linewidth]{Henn_density_polymorphisms_syn_over_chances}}
-\caption{The total density of polymorphisms (mostly nonsynonymous ones) is
-highest in the V regions (left panel). The density of synonymous mutations only,
-however, is not enriched there (right panel). This could be due to a more
-deleterious effect of synonymous mutations.}
-\end{center}
-\end{figure}
-
+Furthermore, the two crucial parameters that control the fixation probability
+are the following: (a) the deleterious effects of hitchhikers compared to
+the beneficial effects of escape mutants, and (b) the density of escape
+mutations. Intuitively, a higher density of escape mutations (i.e., epitopes)
+enables a larger degree of genetic draft, because escape mutations start to
+combine and their effects add up. In \figurename~\ref{fig:simheat} (left panel),
+we show that this is indeed the case in simulations. For any combination of
+parameters, we get a curve like those shown in \figurename~\ref{fig:simfixp} and
+we measure the area between the curve and the diagonal (neutral prediction). An
+area around 0.20 to 0.35 correspond to a concave curve; an area of 0.45 means no
+genetic draft at all; an area of 0 means a neutral-like fixation probability.
+The data-like parameter region that produces concave $P_f$ curves is roughly
+marked by the ellipse and confirms the aforementioned intuitive expectation.
 \begin{figure}
@@
 \subfloat{\includegraphics[width=0.49\linewidth]{fixation_loss_shortgenome_area_ada_frac_del_eff_coi_0_01_nescepi_6_nonsyn_heat.pdf}}
-
 \caption{Simulations on the escape competition scenario show that the density of
@@
 within a few months of antibody production.}
-
+\label{fig:simheat}
 \end{center}
@@
 
+Third, the effects of more dense escape mutations also affects the fixation
+probability of {\it nonsynonymous} alleles. In real data, among all
+nonsynonymous changes, there is no way to distinguish {\it a priori} between
+escape mutations and hitchhikers. The fixation probability of nonsynonymous
+derived alleles is an average over both, weighted by their abundances at the
+starting frequency. In case of denser epitopes, this average is enriched in
+escape mutations, most of which are destined to fix as soon as they reach their
+establishment frequency $s/N \ll 0.1$~\citep{desai_beneficial_2007}. The
+measured $P_f$ for nonsynonymous alleles in HIV is, however, in hardly any
+excess above the neutral line (see \figurename~\ref{fig:fixp}, blue line). In
+order to reconcile simulations with data on this point, several requirements
+have to be met. First of all, the density of epitopes has to be fairly small,
+lest they dominate the average and cause a strong rise in $P_f$. The parameter
+region compatible with the data is indicated in the right panel of
+\figurename~\ref{fig:simheat} with a circle in the top-left corner,
+corresponding to deleterious effects of synonymous changes of roughly $-0.001$.
+
+Moreover, even at the smallest epitope densities for which genetic draft is
+still effective, it is still much more likely for an escape mutation to both
+reach a high frequency and subsequently fix, compared to a nonsynonymous
+hitchhiker. In HIV, however, as shown in \figurename~\ref{fig:fixtimes}, it is
+quite common for a nonsynonymous mutation (solid lines) to reach majority quicky
+and yet eventually get extinct. This conundrum is suggestive of an active
+mechanism, in real HIV infections, for decreasing the fitness of escape mutants
+over time. We test two possible such mechamisms that are biologically plausible
+and find some evidence in the clinical literature: time-dependent selection and
+within-epitope competition. The former concept refers to the possibility that
+the host immune system slowly recognizes the escape mutants and, by deploying
+its countermeasures (antibodies and killer T cells), reduces its fitness during
+a selective sweep, dooming it to extinction despite the initial quick rise in
+frequency. An example for the fixation probabilities generated by this kind of
+models is shown in \figurename~\ref{fig:simfixp} (right panel). This scenario
+finds some backing in the clinical literature: an antibody response to escape
+mutants is indeed present; the delay of this response compared to appearance of
+the escape mutation, a few months, matches the average sweep time of an escape
+mutant. It is therefore not unlikely that the immune system catch the mutant
+during the sweep and neutralize it before fixation~\citep{richman_rapid_2003,
+bunnik_autologous_2008}. The alternative scenario, however, namely
+within-epitope competition, is also valid. The basic idea is that, since the
+mutation rate of HIV is high, several different escape mutations can be seeded
+at the same time and start to spread. Their fitness benefits are not additive,
+because each of them is essentially sufficient to warrant invisibility to the
+host immune system (e.g. to suppress antibody binding). As a consequence, the
+fastest allele to establish is most likely to reach fixation; later escape
+mutants can still reach frequencies of order one because they initially only
+replace wild type viruses; they are eventually swept away by the more abundant,
+early established mutant. This kind of epistatic interactions within epitopes is
+proven able to reduce fixation probabilities in simulations
+(\figurename~\ref{fig:simfixp}, left panel). Furthermore, the emergence of
+multiple sweeping nonsynonymous mutations in real HIV infections has been shown
+previously~\citep{moore_limited_2009}.
+
 \section{Discussion}
+In the HIV research community, synonymous changes have been known for a long
+time to have potential fitness effects, for instance via the RRE element or the
+priming of RT via tRNA$^\text{Lys3}$~\citep{fernandes_hiv-1_2012,
+paillart_vitro_2002}. Despite the large {\it corpus} of biochemical studies on
+these phenomena, the evolutionary significance of nonneutral synonymous changes
+reimains to be assessed. In this study, we show that, in the epitope-rich V1-C5
+\env{} region, synonymous changes with deleterious effects are actively lost
+during the course of the infection. Note that, although most of these alleles
+are indeed rare at the population level, i.e. across HIV-infected subjects, that
+fact alone is not sufficient to infer fitness effects, because of the
+confounding effect of phylogeny. Conversely, the positive correlation between
+intrapatient extinction and interpatient rarity implies, on the light of our
+findings, that these alleles should be constantly depleted from the pool of
+circulating HIV strains by natural selection. A mechanistic insight into this
+process is not trivial, however, because the time scale of the intrapatient
+depletion is long, $s^{-1} \sim 10^3~\text{days}$, whereas transmission is
+thought to happen mostly during early infection~\citep{brenner_high_2007}.
+
+With respect to the correlation with RNA seconary structure, functional
+significance of the insulating stems has been proposed
+previously~\citep{watts_architecture_2009, sanjuan_interplay_2011}, but its
+direct microevolutionary consequences are less clear. Our analysis is akin to
+that in ref.~\citep{sanjuan_interplay_2011} in terms of overall goals, but
+provides direct evidence that insulating stems are relevant for viral fitness
+{\it in vivo}. In absence of clear-cut biochemical experiments, our results
+confirm the hypothesis that insulating stems might be needed to organize the
+viral RNA during packaging in such a way as to avoid unwanted structures caused
+by mutations in the hypervariable loops. In addition, although the fitness
+effects involved are quite small (ten times smaller than the typical benefit of
+an escape mutation), the consistency across unrelated patients makes these stems
+an attractive potential target for therapeutic drugs. Unfortunately, our
+approach yields a rather weak, if locally highly significant, signal. Whether or
+not similar RNA secondary structures are present in other regions of the genome
+remains largely an unanswered question, although we do not find evidence in this
+sense.
+
+As far as population genetics models are concerned, our study uncovers the
+subtle balance of evolutionary forces governing intrapatient HIV evolution. The
+fixation and extinction times and probabilities represent a rich and simple
+summary statistics to test sequencing data and computer simulation upon, as
+noted independently in ref.~\citep{strelkowa_clonal_2012} in the context of
+influenza. Furthermore, our results emphasize the inadequacy of independent-site
+models of HIV evolution, especially in the light of transient effects on
+sweeping sites, such as time-dependent selection and within-epitope negative
+epistasis. Although a final word about which of these mechanisms is more
+widespread is yet to be spoken, both intuition and biological evidence from the
+literature support a mixed scenario~\citep{richman_rapid_2003,
+moore_limited_2009}. Note also that, unlike influenza, HIV does recombine if
+rarely, hence clonal interference as studied in
+ref.~\citep{strelkowa_clonal_2012} is only a short-term effect. In conclusion,
+we regard two consequences of this state of affairs as particularly relevant for
+clinical purposes. On the one hand, the intervention of the host immune system
+appears, if late, effective in fighting escape mutants, so that an active
+stimulation of the host immune systems towards a more prompt response might be a
+viable treatment route. On the other hand, if HIV is indeed able to generate
+several escape mutants at the same time, as both data and calculations seem to
+indicate, an early response against some of them might not suffice to control
+the viral load.
+
 \section{Methods}
+\comment{to be written\dots}
 \section*{Acknowledgements}
+\comment{to be written\dots}
 

### 2012-12-03 synmut.tex
typo + caption package

@@
 \usepackage{graphicx}
-\usepackage[font=small, format=hang, labelfont={sf,bf}, figurename=Fig.]{caption}
+%\usepackage[font=small, format=hang, labelfont={sf,bf}, figurename=Fig.]{caption}
 \usepackage{natbib}
@@
 escape mutations, most of which are destined to fix as soon as they reach their
-establishment frequency $s/N \ll 0.1$~\citep{desai_beneficial_2007}. The
+establishment frequency $(sN)^{-1} \ll 0.1$~\citep{desai_beneficial_2007}. The
 measured $P_f$ for nonsynonymous alleles in HIV is, however, in hardly any

### 2012-12-04 synmut.tex
Coalescent tree + C2-V5

@@
 \includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_nonsynp8}
-\caption{Allele frequency trajectories of typical patient, C3-V5, nonsynonymous
+\caption{Allele frequency trajectories of typical patient, C2-V5, nonsynonymous
 (solid) and synonymous mutations (dashed lines). Most synonymous mutations are
-not fixed. Colors are set according to the position of the site along the C3-V5
+not fixed. Colors are set according to the position of the site along the C2-V5
 region (red to blue). Data from Ref.~\cite{shankarappa_consistent_1999}.}
@@
 \caption{Left panel: fixation probability of derived synonymous alleles is strongly
-suppressed in C3-V5 versus other parts of the {\it env} gene, and of
+suppressed in C2-V5 versus other parts of the {\it env} gene, and of
 nonsynonymous ones.

### 2012-12-06 synmut.tex
major rearrangements, rewriting -> first draft from RAN, needs fleshing out, new figures etc

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\documentclass[12pt,a4paper,notitlepage,onecolumn]{article}
+\documentclass[rmp, twocolumn]{revtex4}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
-\newcommand{\Title}{(Rise and fall of synonymous mutations)}
+\newcommand{\Title}{Deleterious synonymous mutations hitch-hike to high frequency in HIV \env~evolution}
 \newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \usepackage[english]{babel}
@@
 \newcommand{\rev}{\textit{rev}}
+\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{document}
 \title{\Title}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{document}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\maketitle
 
@@
 
+\maketitle
+
 \section{Introduction}
 
-HIV evolves rapidly within a single host during the course of the infection. The
-driving forces shaping this process are the high mutation rate and the strong
-selection imposed by the host immune system via killer T cells (CTLs) and
-neutralizing antibodies (AB)~\citep{pantaleo_immunopathogenesis_1996}.
-
-When the host develops a CTL or AB respose against a particular viral epitope,
-rare HIV variants carrying mutated versions of the epitope, called {\it escape
-mutants}, acquire a fitness advantage and spread rapidly in the viral
-population, within a few months (see \figurename~\ref{fig:aft}, solid lines).
+HIV evolves rapidly within a single host during the course of the infection.
+This evolution is driven by strong selection imposed by the host immune system via killer T cells (CTLs) and neutralizing antibodies (AB)~\citep{pantaleo_immunopathogenesis_1996} and facilitated by the high mutation rate of HIV \citep{mansky}.
+When the host develops a CTL or AB response against a particular viral epitope, mutations that reduce or prevent recognition of the epitope frequently emerge.
+Escape mutations in epitopes targeted by CTLs typically evolve early in infection and spread rapidly through the population \citep{McMichael}.
+Later in infection, the most rapidly evolving part of the HIV genome are the so called variable loops of the envelope proteins that need to avoid recognition by neutralizing anti bodies.
+Mutation in \env~ spread through the population within a few months (see \figurename~\ref{fig:aft}, solid lines).
 During chronic infection, the (Malthusian) effect size of this beneficial
 mutations is of the order of $s_e \sim 0.01$~\citep{neher_recombination_2010}.
-The viral \env{} gene shows the fastest rates of adaptation, because is both
-rich of CTL epitopes and targeted by antibodies; its sequence diverges at rates
-of the order of $1\%$ per year~\citep{shankarappa_consistent_1999}.
 
-Many nucleotide polymorphisms that reach high frequency in the viral population
-are escape mutations: all of them are {\it nonsynonymous} (i.e. they appear in
-protein coding regions and change the amino acid sequence). Nonetheless,
-nucleotide changes unrelated to immune escape are not rare, in \env{} and
-elsewhere, and some of them become abundant alike, often rapidly. In particular,
-it is not uncommon for {\it synonymous} mutations to reach frequencies of order
-one within months from their first appearance (see \figurename~\ref{fig:aft},
-dashed lines). The biological function of these mutations within the economy of
-HIV is not well understood. By definition, the immunological phenotype, which is
-decided at the protein level, is unaffected, but other biological and ecological
-aspects of the viral lifestyle might be involved. In practice, a couple of
-RNA-level phenotypes are known. For example, within \env{} a certain RNA
-sequence, called \rev{} response element (RRE), is used by HIV to enhance
+These escape mutations are selected for their effect on the amino acid sequence of the viral proteins.
+The viral genome, however, needs to meet additional constraints such as efficient processing and translation, nuclear export, and packaging into the viral capsid which operate on the RNA level.
+Purifying selection beyond the protein sequence is therefore expected, while it seems reasonable that the bulk of positive selection through the immune system should be restricted to amino acid sequences.
+A couple important RNA elements  are well characterized.
+For example, within \env{} a certain RNA sequence, called \rev{} response element (RRE), is used by HIV to enhance
 nuclear export of some of its transcripts~\citep{fernandes_hiv-1_2012}. Another
@@
 in the viral 5' untranslated region~\citep{barat_interaction_1991,
-paillart_vitro_2002}.
-
-Crucially for evolutionary purposes, the minor phenotypes caused by synonymous
-mutations might have an effect on viral fitness. For instance, recent studies
+paillart_vitro_2002}. Recent studies
 have shown that genetically engineered HIV strains with skewed codon usage bias
 (CUB) patterns towards more or less abundant tRNAs replicate better or worse,
-respectively~\citep{ngumbela_quantitative_2008, li_codon-usage-based_2012}. In
-this study, we try to characterize the fitness effects of synonymous
-polymorphisms that, at some point during chronic infection, become abundant in
-the viral population.
+respectively~\citep{ngumbela_quantitative_2008, li_codon-usage-based_2012}.
+
+INFLUENZA PSEUDO VACCINE.
+
+SYNONYMOUS CONSERVATION. DO WE HAVE A PLOT OF GENOME WIDE CONSERVATION, MAYBE FOR SUPPLEMENT?
+
+Despite evidence for functional importance of specific RNA sequences, synonymous mutations are commonly used as approximate neutral markers in studies of viral evolution.
+Neutral markers allow to make inference about the stochastic forces driving evolution \citep{smth}.
+Here, we characterize the dynamics of synonymous mutations in \env{} and show that a substantial fraction of these mutations are deleterious.
+The central quantity we investigate is the probability of fixation of a mutation, conditional on its population frequency.
+Even though the synonymous mutations are deleterious and cannot be used as neutral markers, we show the degree to which they hitch-hike with nearby non-synonymous mutations is very informative.
+Extending the analysis of fixation probabilities to the non-synonymous mutations, we show that time dependent selection or strong competition of escape mutations inside the same epitope are necessary to explain the observed patterns of fixation and loss.
 
@@
 neutral alleles have no equilibrium frequency and can slowly fix via genetic
-drift~\citep{ewens_mathematical_2004}. If the focal synonymous mutant is linked
+drift~\citep{ewens_mathematical_2004}.
+t If the focal synonymous mutant is linked
 to another nonneutral allele, however, its frequency is the result of the
@@
 \section{Results}
-We analyze from time series of viral nucleotide sequences from single patients,
-which span several years of chronic
-infection~\citep{shankarappa_consistent_1999, bunnik_autologous_2008,
-liu_selection_2006}. Plotting the allele frequencies against time for all
-polymorphic sites, it is evident that, although nonsynonymous changes are
-widespread, synonymous ones are also present in various occasions at high
-frequency (see \figurename~\ref{fig:aft}).
+A neutral mutation segregating at frequency $\nu$ has a probability $\nu$ to spread through the population and fix, while it is lost with probability $1-\nu$.
+This is a simple consequence of the fact that exactly one of present $N$ individuals will be the common ancestor of the entire population at a particular locus and this ancestor has a probability $\nu$ of carrying this mutations, see illustration in \FIG{illustration}.
+Deleterious or beneficial mutations, in contrast, should fix less or more often, respectively.
+Time series sequence data therefore suggest a simple way to investigate average properties of different classes of mutations.
+
+\paragraph{Synonymous mutations in \env, C2-V5 are mostly deleterious}
+
+\FIG{aft} shows time series data of the frequencies of all mutations observed \env, C2-V5, in patient ??\citep{shankarappa_consistent_1999,liu_selection_2006}.
+Despite many synonymous mutations reaching high frequency, very few fix.
+This observation in further quantified in panels ?
+and ?, that stratify the data of ??
+patients (see methods) according to the frequency at which different mutations are observed.
+Considering all mutations in a frequency interval $\nu_0$ at some time $t_i$, we calculate the fraction that is found at frequency 1, frequency 0, or at intermediate frequency at later time points $t_f$.
+Plotting these fixed, lost, and polymorphic fraction against the time interval $t_f-t_i$, we see that most synonymous mutations segregate for roughly one year and are lost  much more frequently than expected.
+The ultimate probability of loss or fixation is shown as a function of the initial frequency $\nu_0$ in panel ??.
+In contrast to synonymous mutations, the non-synonymous seem to follow more a less the neutral expectation -- a point to which we will come back below.
+
 \begin{figure}
@@
 \includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_nonsynp8}
-\caption{Allele frequency trajectories of typical patient, C2-V5, nonsynonymous
-(solid) and synonymous mutations (dashed lines). Most synonymous mutations are
-not fixed. Colors are set according to the position of the site along the C2-V5
-region (red to blue). Data from Ref.~\cite{shankarappa_consistent_1999}.}
+\includegraphics[width=\linewidth]{Shankarappa_fix_loss_dt_times}
+\caption{Synonymous mutations rarely fix in \env, C2-V5: Panel A shows mutation 
+frequency trajectories observed in patient ?? \cite{shankarappa_consistent_1999};  Nonsynonymous
+and synonymous mutations are shown as solid and dashed lines, respectively. 
+Colors indicate the position of the site along the C2-V5 region (red to blue) ADD COLORBAR. MAYBE MAKE FIGURE WITH SYNONYMOUS AND NONSYN SEPARATELY.
+While non-synonymous mutations frequently fix, very few synonymous mutations do even though they are frequently observed at intermediate frequencies.
+Panel B quantifies time course of loss and fixation of synonymous mutations observed in a frequency interval $\nu_0$.
+The ultimate fraction of synonymous mutations that fix as a function of intermediate frequency $\nu_0$.
+}
 \label{fig:aft}
@@
 \end{figure}
-Moreover, many mutations appear to change rapidly in frequency as a flock,
-especially at the lower end of the frequency spectrum. These observations lead
-to the idea that linkage might be widespread at the time scale of the typical
-selective sweep, i.e. a few months~\citep{neher_recombination_2010}. It seems
-therefore likely that synonymous mutations are not attaining high frequencies by
-genetic drift but are rather exponentially amplified by selection on linked
-nonsynonymous sites, a process known as {\it genetic
-draft}~\citep{neher_genetic_2011}.
 
-To test this hypothesis, we focus on the long-time fate of such synonymous
-derived alleles {\it after} they are already abundant. According to the neutral
-theory of independent sites, (a) an allele starting at frequency $\nu$ will
-reach fixation, in the long-time limit, with a probability $P_f(\nu) = \nu$, and
-get extinct otherwise; (b) the time required to reach either boundary, fixation
-or extinction, is essentialy the inverse of the population size (in
-generations), hence much longer than five years for HIV~\citep{boltz_ultrasensitive_2012}.
+\citet{bunnik_autologous_2008} present a longitudinal data sets on the entire \env~gene of ??
+patients at ??
+time points with ??-??
+sequences each.
+Repeating the above analysis separately on the C2-V5 region studied above and the remainder of \env~ reveal strikingly different behavior inside and outside the hyper variable region.
+Within C2-V5, this data fully confirms the observations made in the data set by \citet{shankarappa_consistent_1999}.
+In the remainder of \env, however, observed synonymous mutations behave as if they were neutral; see \FIG{fixp}.
+
+ARE OBSERVED SYNONYMOUS MUTATIONS OUTSIDE C2-V5 NEUTRAL?
+DOES LOSS/FIX CORRELATE WITH CONSERVATION.
+CAN WE LOOK AT THE AVERAGE LEVEL OF CONSERVATION STRATIFIED BY MAX FREQ?
+MAYBE WE COULD HAVE ONE -- COMPLETELY CIRCULAR -- FIGURE SHOWING LOSS/FIX VS CONSERVATION.
+
+These observations suggest that many of the synonymous mutations in the part of \env~that includes the hyper variable regions are deleterious, while outside this regions only roughly neutral mutations are polymorphic.
+
 \begin{figure}
@@
 \subfloat{\includegraphics[width=0.49\linewidth]{Bunnik2008_fixmid_syn_ShankanonShanka}}
-\subfloat{\includegraphics[width=0.49\linewidth]{Shankarappa_fixmid_syn_V_regions}}
 \caption{Left panel: fixation probability of derived synonymous alleles is strongly
@@
 
-The left panel of \figurename~\ref{fig:fixp} shows the fixation probability of
-derived alleles seen within various frequency windows. With regard to synonymous
-alleles, the average over the whole HIV genome (organce line) matches closely
-the expected result from neutral theory (dashed line). A different picture
-emerges when the most variable part of the genome, V1-C5 region of \env{},
-is zoomed into (red line). Two independent datasets, from
-refs.~\citep{shankarappa_consistent_1999, bunnik_autologous_2008} yield the same
-result, i.e. that synonymous mutations hardly fix at all.
+\paragraph{Synonymous mutations in C2-V5 tend to disrupt conserved RNA stems}
+One possible {\it a priori} explanation for lack of fixation of synonymous mutations in C2-V5 are  secondary structures in the viral RNA.
+If any RNA secondary structures are relevant for HIV replication,
+mutations in nucleotides involved in those base pairs are expected  to be deleterious and to revert preferentially. 
+Many functionally important secondary structure elements have been characterized, including  the  \rev{} response element (RRE) to enhance nuclear export of some of its transcripts~\citep{fernandes_hiv-1_2012}.
+Another well studied case is the interaction between viral reverse transcriptase, viral ssRNA, and the host tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription (RT) and bound by a specifical pseudoknotted RNA structure in the viral 5' untranslated region~\citep{barat_interaction_1991, paillart_vitro_2002}.
+It has been suggested early on that parts of the viral genome that has the potential to form stems as better conserved that the remainder \citep{forsdyke_reciprocal_1995}.
 
-A related observation can be made on the nonsynonymous alleles (blue line).
-Although we see no difference between different parts of the HIV genome, the
-neutral-like shape of $P_f$ is at odds with the na\"ive expectation that most
-nonsynonymous alleles reaching high frequency are to-be-fixed escape mutations.
-
-Moreover, as illustrated in \figurename~\ref{fig:fixtimes}, the time for
-synonymous alleles to reach the first boundary starting from intermediate
-frequencies is of the order of 500 days which, assuming a conservative
-generation time of one day~\citep{perelson_hiv-1_1996}, correspond to the same
-number of generations. This time would only be possible in a neutral scenario
-if the population size of HIV were extremely small, of the order of $10^3$, much
-smaller then both theory and experiments
-indicate~\citep{boltz_ultrasensitive_2012}.
-% WE SHOULD REALLY CITE OURSELVES IN THE (HOPEFUL) FUTURE HERE :-)
-\begin{figure}
-\begin{center}
-\includegraphics[width=\linewidth]{Shankarappa_fix_loss_dt_times}
-\caption{Fixation or extinction times for synonymous alleles starting from
-intermediate frequencies. The colored bands are the final fixation
-probabilities expected from neutral theory; the observed alleles are fixed
-less frequently than expected. The timescale of fixation/extinction is
-approximately 500 days, corresponding to a selective effect of $\sim -0.001$.}
-\label{fig:fixtimes}
-\end{center}
-\end{figure}
-
-Taken together, these findings suggest that synonymous (and some nonsynonymous)
-alleles are not propagating independently by genetic drift, but rather
-hitchhiking on neighbouring beneficial alleles. This conclusion is also
-consistent with the currently accepted estimate for the outcrossing rate of HIV,
-of the order of $1\%$ per generation. Although synonymous alleles can be drafted
-to high frequency by linkage and selection, they decouple on a time scale of the
-order of 100 days, which is also the average time it takes a (nonsynonymous)
-selected allele to spread through the population during chronic
-infection~\citep{neher_recombination_2010}.
-
-\figurename~\ref{fig:fixp} (left panel) also suggests that at least some
-synonymous mutations in the V1-C5 region should be deleterious. We analyze this
-hypothesis in two ways. First of all, we look for possible biological causes for
-such a behaviour. Furthermore, we perform computer simulations of viral
-evolution under a wide range of parameters in order to reproduce the results for
-the fixation probability and study the regions of the parameter space compatible
-with the observations.
-
-One possible {\it a priori} explanation for deleterious fitness effects of
-synonymous mutations is the presence of secondary structures in the viral RNA.
-If any RNA secondary structures are functionally relevant for HIV replication,
-mutations in nucleotides involved in those base pairs are expected to revert
-more easily than by pure chance, to keep those structures in place. The
-propensity of nucleotides to form base pairs in HIV has been measured
-recently\citep{watts_architecture_2009}, and the HIV genome has resulted to be
-highly structured~\citep{forsdyke_reciprocal_1995, watts_architecture_2009}.
-Watts {\it et al.} have measured the propensity of any nucleoside to be attacked by
-external chemicals (SHAPE reactivity) and shown that a higher reactivity
-corresponds to a smaller likelihood of being involved in an RNA helix. As far as
-the V1-C5 region is concerned, it has been shown that the short C1-5 regions
-have a high tendency to form {\it insulating
-stems}~\citep{watts_architecture_2009, sanjuan_interplay_2011}. Since the V1-5
-loops have to be highly variable to ensure immune escape, their destabilizing
-contribution to the viral RNA structure is though to be buffered by the
-surrounding, base-pair rich stems. To test whether or not insulating stems might
-be the reason for the reduced fixation probability, we partition the data
-depending on their position in the V1-C5 region. Consistently with the
-expectations, $P_f$ is more heavily readuced in the stems than in
-the loops (see \figurename~\ref{fig:fixp}, right panel).
-
-Despite this indirect result supporting the involvement of RNA secondary
-structure in the deleterious effects of synonymous mutations in V1-C5, we set
-out to seek more direct evidence. We partition all synonymous alleles observed
+Recently, the propensity of nucleotides of the HIV genome to form base pairs has been measured using the SHAPE assay (a biochemical reaction preferentially altering unpaired bases) \citep{watts_architecture_2009}.
+The SHAPE assay has shown that the variable regions V1 to V5 tend to be unpaired, while the conserved regions between those variable regions form stems.
+We partition all synonymous alleles observed
 at intermediate frequencies above 10-15\% depending on their final destiny
@@
 lost from the population.
-\begin{figure}
-\begin{center}
-\subfloat{\includegraphics[width=0.49\linewidth]{mixed_Shankarappa_Bunnik2008_Liu_fixation_reactivity_Vandflanking_fromSHAPE}}
-\subfloat{\includegraphics[width=0.49\linewidth]{mixed_Shankarappa_Bunnik2008_Liu_fixation_reactivity_nonVandflanking}}
-\caption{Watts et al. have measured the reactivity of HIV nucleotides to {\it
-in vitro} chemical attack and shown that some nucleotides are more likely to
-be involved in RNA secondary folds. C1-C5 regions, in particular, show
-conserved stem-loop structures~\citep{watts_architecture_2009}. We show that
-among all derived alleles in those regions reaching frequencies of order one,
-there is a negative correlation between fixation and involvement in a base
-pairing in a RNA stem (left panel). The rest of the genome does not show any
-correlation (right panel). There might be too few silent polymorphisms in the
-first place, or the signal might be masked by non-functional RNA
-structures. Data from Refs.~\cite{shankarappa_consistent_1999,
-bunnik_autologous_2008, liu_selection_2006}.}
-\label{fig:SHAPE}
-\end{center}
-\end{figure}
-
-Since the insulating stems are not the only RNA secondary structures found by
-the authors of ref.~\citep{watts_architecture_2009}, we repeated the same
-partitioning based on fate on the rest of the HIV genome. As shown in the right
-panel of \figurename~\ref{fig:SHAPE}, there is no evidence that a similar
-mechanism is acting outside of the V1-C5 region. This is not fully surprising,
-because it is well known that single stranded RNA chains tend to form a lot of
-spontaneous structures, most of which have no biological function. Furthermore,
-the averaging over many different genomic regions might worsen the
-signal-to-noise ratio, which is already quite low, and result in
-indistinguishable histograms.
+We then split the synonymous mutations in the C2-V5 region further into conserved and variable regions and found that the biggest depression in fixation probability is observed in the conserved stems, while the variable loops show little deviations from the neutral signature; see \FIG{SHAPE}B.
 
@@
 
-We also perform computer simulations of evolving viral population under
-selection and rare recombination. For this purpose, we use the recently
-published package FFPopSim, which includes a module dedicated to intrapatient
-HIV evolution~\citep{zanini_ffpopsim:_2012}. We analyze many combination of
-parameters such as population size, recombination rate, selection coefficient
-and density of escape mutations, epitope length, deleterious effect of
-synonymous mutations. The purpose of the simulations is to probe the population
-genetics mechanisms required to reproduce the fixation probabilities of
-\figurename~\ref{fig:fixp}, simultaneously for synonymous and nonsynonymous
-alleles.
-
-The main result of the simulations is that genetic draft can indeed generate a
-depression in fixation probabilities, but only under relatively tight
-circumstances. In terms of model confidence, this is a good result: the
-probability of obtaining the experimental curves by pure chance is negligible.
-\figurename~\ref{fig:simfixp} shows two instances of successful simulations.
 \begin{figure}
 \begin{center}
-\subfloat{\includegraphics[width=0.49\linewidth]{fixation_probability_shortgenome_N_1e4_epitopes_example_longer}}
-\subfloat{\includegraphics[width=0.49\linewidth]{fixation_probability_N_1e4_3ingredients_3rddel_envnonenv_stopall}}
-\caption{Simulations show that the suppression of fixation probability can be
-generated by linkage to sweeping nonsynonymous alleles nearby. Two possible
-scenarios are competition between escape mutants (left panel) and time-dependent
-selection due to immune sytem recognition (right panel).}
-\label{fig:simfixp}
+\subfloat{\includegraphics[width=0.49\linewidth]{mixed_Shankarappa_Bunnik2008_Liu_fixation_reactivity_Vandflanking_fromSHAPE}}
+\subfloat{\includegraphics[width=0.49\linewidth]{Shankarappa_fixmid_syn_V_regions.pdf}}
+\caption{Watts et al. have measured the reactivity of HIV nucleotides to {\it
+in vitro} chemical attack and shown that some nucleotides are more likely to
+be involved in RNA secondary folds. C1-C5 regions, in particular, show
+conserved stem-loop structures~\citep{watts_architecture_2009}. We show that
+among all derived alleles in those regions reaching frequencies of order one,
+there is a negative correlation between fixation and involvement in a base
+pairing in a RNA stem (left panel). The rest of the genome does not show any
+correlation (right panel). There might be too few silent polymorphisms in the
+first place, or the signal might be masked by non-functional RNA
+structures. Data from Refs.~\cite{shankarappa_consistent_1999,
+bunnik_autologous_2008, liu_selection_2006}.}
+\label{fig:SHAPE}
 \end{center}
@@
 
-First of all, in order for $P_f$ to be reduced to levels seen in the data, most
-synonymous mutants habe to be slightly deleterious. The reason is that if
-synonymous alleles are an even mixture of deleterious and neutral ones, the
-latter are much more likely to be carried to high frequencies by genetic draft,
-because they present no burden to escape mutations. The synonymous $P_f$ is then
-dominated by these neutral alleles and the depression induced by the few
-deleterious ones becomes undetectable.
+\paragraph{Deleterious mutations are brought to high frequency by hitch-hiking}
+While the observation that some fraction of synonymous mutations is deleterious is not unexpected, it seems odd that we observe them at high population frequency -- at least in some regions of the genome.
+The region of \env~in which we observe deleterious mutations at high frequency is special in that it undergoes frequent adaptive changes to evade recognition by neutralizing antibodies \cite{Williamson}.
+Due to the limited amount of recombination in HIV \cite{neher_recombination_2010,batorsky_2011}, deleterious mutations that are linked to adaptive variants can reach high frequency \citep{maynard_smith}.
 
+The potential for hitch-hiking is already apparent from the allele frequency trajectories in \FIG{aft}, where many mutations appear to change rapidly in frequency as a flock.
+Deleterious synonymous mutations can be amplified exponentially by selection on linked nonsynonymous sites, a process known as {\it genetic draft}~\citep{neher_genetic_2011}.
+In order to be advected to high frequency by a linked adaptive mutation, the deleterious effect of the mutation has to be substantially smaller than the adaptive effect.
+The latter was estimated to be on the order of $s_a = 0.01$ per day.
+The approximate magnitude of the deleterious effects can be estimated from \FIG{fixtimes}, that shows the distribution of times for synonymous alleles to reach the fix or get lost starting from intermediate frequencies.
+The typical time to loss is of the order of 500 days.
+If this loss is driven by the deleterious effect of the mutation, this corresponds to deleterious effects of roughly $0.002$ per day.
+
+To get a better idea of the range of parameters that are compatible with the observations and our interpretation, we  perform computer simulations of evolving viral populations under selection and rare recombination.
+For this purpose, we use the recently published package FFPopSim, which includes a module dedicated to intra-patient
+HIV evolution~\citep{zanini_ffpopsim:_2012}. We analyze many combinations of
+parameters such as population size, recombination rate, selection coefficient
+and density of escape mutations, deleterious effect of synonymous mutation.
+
+The main result of the simulations is that genetic draft can indeed bring weakly deleterious mutations to high frequencies and result in a dependence of the fixation probability on initial frequency that is compatible with observations.
+We quantify the reduction in fixation probability by the area under the diagonal....
+Since neutral mutations are much more likely to rise to high frequency than deleterious ones, the majority of the synonymous mutations needs to be slightly deleterious observe a significant reduction of $p_{fix}$.
 Furthermore, the two crucial parameters that control the fixation probability
@@
 combine and their effects add up. In \figurename~\ref{fig:simheat} (left panel),
-we show that this is indeed the case in simulations. For any combination of
-parameters, we get a curve like those shown in \figurename~\ref{fig:simfixp} and
-we measure the area between the curve and the diagonal (neutral prediction). An
-area around 0.20 to 0.35 correspond to a concave curve; an area of 0.45 means no
-genetic draft at all; an area of 0 means a neutral-like fixation probability.
-The data-like parameter region that produces concave $P_f$ curves is roughly
-marked by the ellipse and confirms the aforementioned intuitive expectation.
+we show that this is indeed the case in simulations.
+
+SHOW THAT THE DEPRESSION WORKS.
+
 \begin{figure}
@@
 
-Third, the effects of more dense escape mutations also affects the fixation
-probability of {\it nonsynonymous} alleles. In real data, among all
-nonsynonymous changes, there is no way to distinguish {\it a priori} between
-escape mutations and hitchhikers. The fixation probability of nonsynonymous
-derived alleles is an average over both, weighted by their abundances at the
-starting frequency. In case of denser epitopes, this average is enriched in
-escape mutations, most of which are destined to fix as soon as they reach their
-establishment frequency $(sN)^{-1} \ll 0.1$~\citep{desai_beneficial_2007}. The
-measured $P_f$ for nonsynonymous alleles in HIV is, however, in hardly any
-excess above the neutral line (see \figurename~\ref{fig:fixp}, blue line). In
-order to reconcile simulations with data on this point, several requirements
-have to be met. First of all, the density of epitopes has to be fairly small,
-lest they dominate the average and cause a strong rise in $P_f$. The parameter
-region compatible with the data is indicated in the right panel of
-\figurename~\ref{fig:simheat} with a circle in the top-left corner,
-corresponding to deleterious effects of synonymous changes of roughly $-0.001$.
-
-Moreover, even at the smallest epitope densities for which genetic draft is
-still effective, it is still much more likely for an escape mutation to both
-reach a high frequency and subsequently fix, compared to a nonsynonymous
-hitchhiker. In HIV, however, as shown in \figurename~\ref{fig:fixtimes}, it is
-quite common for a nonsynonymous mutation (solid lines) to reach majority quicky
-and yet eventually get extinct. This conundrum is suggestive of an active
-mechanism, in real HIV infections, for decreasing the fitness of escape mutants
-over time. We test two possible such mechamisms that are biologically plausible
-and find some evidence in the clinical literature: time-dependent selection and
-within-epitope competition. The former concept refers to the possibility that
-the host immune system slowly recognizes the escape mutants and, by deploying
-its countermeasures (antibodies and killer T cells), reduces its fitness during
-a selective sweep, dooming it to extinction despite the initial quick rise in
+However, if hitch-hiking is driven by non-synonymous mutations that are unconditionally beneficial, we should find that non-synonymous mutations almost always fix once they reach high frequencies -- in contrast with \FIG{fixp} that shows that non-synonymous mutations fix as if they were neutral.
+We know, however, that non-synonymous variation in the variable regions is driven by positive selection.
+Inspecting the trajectories of non-synonymous mutations suggest the rapid rise and fall of many alleles.
+We test two possible such mechanisms that are biologically plausible
+and could explain the transient rise of non-synonymous mutations: time-dependent selection and
+within-epitope competition. If the immune system starts recognizes the escape mutant before its fixation, the mutant might cease to be beneficial and disappears despite its quick initial rise in
 frequency. An example for the fixation probabilities generated by this kind of
-models is shown in \figurename~\ref{fig:simfixp} (right panel). This scenario
-finds some backing in the clinical literature: an antibody response to escape
-mutants is indeed present; the delay of this response compared to appearance of
-the escape mutation, a few months, matches the average sweep time of an escape
-mutant. It is therefore not unlikely that the immune system catch the mutant
-during the sweep and neutralize it before fixation~\citep{richman_rapid_2003,
-bunnik_autologous_2008}. The alternative scenario, however, namely
-within-epitope competition, is also valid. The basic idea is that, since the
-mutation rate of HIV is high, several different escape mutations can be seeded
-at the same time and start to spread. Their fitness benefits are not additive,
-because each of them is essentially sufficient to warrant invisibility to the
-host immune system (e.g. to suppress antibody binding). As a consequence, the
-fastest allele to establish is most likely to reach fixation; later escape
-mutants can still reach frequencies of order one because they initially only
-replace wild type viruses; they are eventually swept away by the more abundant,
-early established mutant. This kind of epistatic interactions within epitopes is
-proven able to reduce fixation probabilities in simulations
-(\figurename~\ref{fig:simfixp}, left panel). Furthermore, the emergence of
-multiple sweeping nonsynonymous mutations in real HIV infections has been shown
+models is shown in \figurename~\ref{fig:simfixp} (right panel). In support of this idea, \citet{richman_rapid_2003,
+bunnik_autologous_2008} report antibody responses to escape mutants.
+These respones are delayed by a few months, roughly matching the average sweep time of an escape mutant.
+Alternatively, several different escape mutations in the same epitope can arise almost simultaneously and start to spread.
+Their fitness benefits are not additive, because each of them is essentially sufficient to escape.
+As a consequence, several mutations rise to high frequency, while the escape with the smallest cost is most likely to eventually fix.
+In simulations, this kind of epistatic interactions within epitopes is reduces fixation probabilities in simulations (\figurename~\ref{fig:simfixp}, left panel).
+The emergence of multiple sweeping nonsynonymous mutations in real HIV infections has been shown
 previously~\citep{moore_limited_2009}.
@@
 \section{Discussion}
-In the HIV research community, synonymous changes have been known for a long
-time to have potential fitness effects, for instance via the RRE element or the
-priming of RT via tRNA$^\text{Lys3}$~\citep{fernandes_hiv-1_2012,
-paillart_vitro_2002}. Despite the large {\it corpus} of biochemical studies on
-these phenomena, the evolutionary significance of nonneutral synonymous changes
-reimains to be assessed. In this study, we show that, in the epitope-rich V1-C5
-\env{} region, synonymous changes with deleterious effects are actively lost
-during the course of the infection. Note that, although most of these alleles
-are indeed rare at the population level, i.e. across HIV-infected subjects, that
-fact alone is not sufficient to infer fitness effects, because of the
-confounding effect of phylogeny. Conversely, the positive correlation between
-intrapatient extinction and interpatient rarity implies, on the light of our
-findings, that these alleles should be constantly depleted from the pool of
-circulating HIV strains by natural selection. A mechanistic insight into this
-process is not trivial, however, because the time scale of the intrapatient
-depletion is long, $s^{-1} \sim 10^3~\text{days}$, whereas transmission is
-thought to happen mostly during early infection~\citep{brenner_high_2007}.
+Despite several known functional roles for RNA secondary structure in the HIV genome, synonymous mutations are often used as approximately neutral markers in evolutionary studies of viruses.
+We have shown that the majority of synonymous mutations in the conserved regions C2-C5 of the \env~gene are deleterious.
+Comparison with recent biochemical studies of binding propensity of bases in RNA genome suggest that these mutations are deleterious in part because they disrupt stems in RNA secondary structure.
+Furthermore, we provide evidence that these mutations are brought to high frequency through linkage to adaptive mutations.
+The latter  mutations are only transiently adaptive, either through a coevolution with the immune system or redundant escape within an epitope.
 
-With respect to the correlation with RNA seconary structure, functional
-significance of the insulating stems has been proposed
-previously~\citep{watts_architecture_2009, sanjuan_interplay_2011}, but its
-direct microevolutionary consequences are less clear. Our analysis is akin to
-that in ref.~\citep{sanjuan_interplay_2011} in terms of overall goals, but
-provides direct evidence that insulating stems are relevant for viral fitness
-{\it in vivo}. In absence of clear-cut biochemical experiments, our results
-confirm the hypothesis that insulating stems might be needed to organize the
-viral RNA during packaging in such a way as to avoid unwanted structures caused
-by mutations in the hypervariable loops. In addition, although the fitness
-effects involved are quite small (ten times smaller than the typical benefit of
-an escape mutation), the consistency across unrelated patients makes these stems
-an attractive potential target for therapeutic drugs. Unfortunately, our
-approach yields a rather weak, if locally highly significant, signal. Whether or
-not similar RNA secondary structures are present in other regions of the genome
-remains largely an unanswered question, although we do not find evidence in this
-sense.
+Our observations and conclusion rely heavily on longitudinal data in which the dynamics of mutations can be explicitly observed.
+The fact that deleterious mutations can be brought to high frequencies through hitch-hiking underscores the vigorousness of the coevolution with the immune system.
+The fact that multiple escape mutations in the same epitope -- as is indeed observed in studies of antibody escape \citep{sdfsd} -- are necessary to explain the patterns of fixation of non-synonymous mutations points towards a large populations size that rapidly discovers adaptive mutations.
+A similar point has been made recently by \citet{boltz_ultrasensitive_2012} in the context of preexisting drug resistance mutations.
+
+The observed hitch-hiking highlights the importance of linkage due to infrequent recombination for the evolution of HIV \citep{neher_recombination_2010,batorsky_estimate_2011,joseffson_smth_2011}.
+The recombination rate has been estimated to be on the order of $\rho = 10^{-5}$ per base and day.
+It takes roughly $t_{sw} = s^{-1} \log \nu_0$ generations for an adaptive mutation with growth rate $s$ to rise from an initially low frequency $\nu_0\sim \mu$ to frequency one.
+This implies that a region of length $l = (\rho t_{sw})^{-1} = s/(\rho \log \nu_0)$ remains linked to the adaptive mutation.
+With $s=0.01$, $l\approx 100$ bases which is consistent with strong linkage between the variable loops and the stems in between.
+Furthermore, we do not expect much hitch-hiking to extend far beyond the variable regions consistent with the lack of signal out side of C5-V5.
+In case of much stronger selection -- such as observed during early CTL escape or drug resistance evolution -- the linked  region is of course a lot larger.
+
+The functional significance of the insulating RNA structure stems between the hyper variable loops has been proposed previously~\citep{watts_architecture_2009, sanjuan_interplay_2011}.
+Our analysis is akin to that in ref.~\citep{sanjuan_interplay_2011} in terms of overall goals, but provides direct evidence that insulating stems are relevant for viral fitness {\it in vivo}.
+Our analysis is limited by the availability of longitudinal data which requires a focus on the the variable regions of \env.
+Conserved RNA structures likely exist (and several are known) in different parts of the HIV genome.
+In absence of repeated adaptive substitutions in the vicinity that cause hitch-hiking, the deleterious synonymous mutations will remain at low frequencies and can only be observed by deep sequencing methods.
 

### 2012-12-08 synmut.tex
some changes applied

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
@@
 \newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
-
 \usepackage[english]{babel}
@@
 \usepackage{graphicx}
-%\usepackage[font=small, format=hang, labelfont={sf,bf}, figurename=Fig.]{caption}
 \usepackage{natbib}
@@
 selective sweeps are less dense and the genetic architecture less constrained.
-%In absence of antiretroviral treatment, HIV is very successful at producing
-%mutants that are able to stay undetected by the host immune system for months,
-%boosting the infection~\citep{richman_rapid_2003, bunnik_autologous_2008,
-%moore_limited_2009}. The high mutation rate at the core of this process,
-%however, also generates genetic noise in the background of beneficial alleles.
-%Because of the limited ourcrossing of HIV, hitchhikers usually stay linked to
-%the focal allele for a time comparable to its sweeping time, i.e. a few
-%months~\citep{neher_recombination_2010, batorsky_estimate_2011}. The later fate
-%of these accessory mutations depends more and more, as they decouple genetically
-%from the escape allele via recombination, on their own fitness effects. In this
-%study we show that, in a genetic region particularly dense of selective sweeps,
-%synonymous hitchhikers tend to revert on a time scale of the order of 500 days,
-%because of their deleterious fitness effects. In addition, we provide evidence
-%that the biological origin for this deleterious effects resides, at least
-%partially, in the disruption of macroevolutionarily conserved RNA secondary
-%structures, termed ``insulating stems''.
 \end{abstract}
@@
 HIV evolves rapidly within a single host during the course of the infection.
-This evolution is driven by strong selection imposed by the host immune system via killer T cells (CTLs) and neutralizing antibodies (AB)~\citep{pantaleo_immunopathogenesis_1996} and facilitated by the high mutation rate of HIV \citep{mansky}.
-When the host develops a CTL or AB response against a particular viral epitope, mutations that reduce or prevent recognition of the epitope frequently emerge.
-Escape mutations in epitopes targeted by CTLs typically evolve early in infection and spread rapidly through the population \citep{McMichael}.
-Later in infection, the most rapidly evolving part of the HIV genome are the so called variable loops of the envelope proteins that need to avoid recognition by neutralizing anti bodies.
-Mutation in \env~ spread through the population within a few months (see \figurename~\ref{fig:aft}, solid lines).
-During chronic infection, the (Malthusian) effect size of this beneficial
-mutations is of the order of $s_e \sim 0.01$~\citep{neher_recombination_2010}.
+This evolution is driven by strong selection imposed by the host immune system
+via killer T cells (CTLs) and neutralizing antibodies
+(AB)~\citep{pantaleo_immunopathogenesis_1996} and facilitated by the high
+mutation rate of HIV~\citep{mansky_lower_1995}. When the host develops a CTL or
+AB response against a particular viral epitope, mutations that reduce or prevent
+recognition of the epitope frequently emerge. Escape mutations in epitopes
+targeted by CTLs typically evolve early in infection and spread rapidly through
+the population~\citep{mcmichael_immune_2009}. Later in infection, the most
+rapidly evolving part of the HIV genome are the so called variable loops of the
+envelope protein gp120, which need to avoid recognition by neutralizing ABs.
+Mutations in \env~, the gene encoding for gp120, spread through the population
+within a few months (see \figurename~\ref{fig:aft}, solid lines). During chronic
+infection, the (Malthusian) effect size of this beneficial mutations is of the
+order of $s_a \sim 0.01$~\citep{neher_recombination_2010}.
 
-These escape mutations are selected for their effect on the amino acid sequence of the viral proteins.
-The viral genome, however, needs to meet additional constraints such as efficient processing and translation, nuclear export, and packaging into the viral capsid which operate on the RNA level.
-Purifying selection beyond the protein sequence is therefore expected, while it seems reasonable that the bulk of positive selection through the immune system should be restricted to amino acid sequences.
-A couple important RNA elements  are well characterized.
-For example, within \env{} a certain RNA sequence, called \rev{} response element (RRE), is used by HIV to enhance
+These escape mutations are selected for their effect on the amino acid sequence
+of the viral proteins. The viral genome, however, needs to meet additional
+constraints such as efficient processing and translation, nuclear export, and
+packaging into the viral capsid: all these processes operate at the RNA level. A
+few important RNA elements are well characterized. For example, a certain RNA
+sequence, called \rev{} response element (RRE), is used by HIV to enhance
 nuclear export of some of its transcripts~\citep{fernandes_hiv-1_2012}. Another
@@
 in the viral 5' untranslated region~\citep{barat_interaction_1991,
-paillart_vitro_2002}. Recent studies
-have shown that genetically engineered HIV strains with skewed codon usage bias
-(CUB) patterns towards more or less abundant tRNAs replicate better or worse,
+paillart_vitro_2002}. Furthermore, recent studies have shown that genetically
+engineered HIV strains with skewed codon usage bias (CUB) patterns towards more
+or less abundant tRNAs replicate better or worse,
 respectively~\citep{ngumbela_quantitative_2008, li_codon-usage-based_2012}.
+Purifying selection beyond the protein sequence is therefore expected, while it
+seems reasonable that the bulk of positive selection through the immune system
+should be restricted to amino acid sequences.
 
@@
 
-Despite evidence for functional importance of specific RNA sequences, synonymous mutations are commonly used as approximate neutral markers in studies of viral evolution.
-Neutral markers allow to make inference about the stochastic forces driving evolution \citep{smth}.
-Here, we characterize the dynamics of synonymous mutations in \env{} and show that a substantial fraction of these mutations are deleterious.
-The central quantity we investigate is the probability of fixation of a mutation, conditional on its population frequency.
-Even though the synonymous mutations are deleterious and cannot be used as neutral markers, we show the degree to which they hitch-hike with nearby non-synonymous mutations is very informative.
-Extending the analysis of fixation probabilities to the non-synonymous mutations, we show that time dependent selection or strong competition of escape mutations inside the same epitope are necessary to explain the observed patterns of fixation and loss.
+Despite evidence for functional importance of specific RNA sequences, synonymous
+mutations are commonly used as approximate neutral markers in studies of viral
+evolution. Neutral markers allow to make inference about the stochastic forces
+driving evolution~\citep{smth}. Here, we characterize the dynamics of synonymous
+mutations in \env{} and show that a substantial fraction of these mutations are
+deleterious. The central quantity we investigate is the probability of fixation
+of a mutation, conditional on its population frequency. Even though the
+synonymous mutations are deleterious and cannot be used as neutral markers, we
+show that the degree to which they hitchhike with nearby non-synonymous
+mutations is very informative; their ability to hitchhike for extended times is
+itself rooted in the small recombination rate of
+HIV~\citep{neher_recombination_2010, batorsky_estimate_2011}. Extending the
+analysis of fixation probabilities to the non-synonymous mutations, we show that
+time dependent selection or strong competition of escape mutations inside the
+same epitope are necessary to explain the observed patterns of fixation and
+loss.
 
-One simple way to assess the neutrality of synonymous mutations is to look at
-their level of conservation. Deleterious mutations at functional sites are
-expected to be absent or rare across the viral population; vice versa, mutant
-alleles that reach high frequencies are expected to be neutral. If genetic sites
-are independent, the equilibrium frequency of a deleterious allele with fitness
-$-s$ is $\mut / |s|$, where $\mut$ is the mutation rate per site per generation;
-neutral alleles have no equilibrium frequency and can slowly fix via genetic
-drift~\citep{ewens_mathematical_2004}.
-t If the focal synonymous mutant is linked
-to another nonneutral allele, however, its frequency is the result of the
-combined fitness effects of both sites, and simple conservation-level analyses
-fail. Since recombination in HIV is known to happen
-rarely~\citep{neher_recombination_2010, batorsky_estimate_2011}, the genetic
-context of the synonymous change at hand must be taken into account. Our
-results underline the importance of the latter scenario for intrapatient HIV
-evolution.
+%One simple way to assess the neutrality of synonymous mutations is to look at
+%their level of conservation. Deleterious mutations at functional sites are
+%expected to be absent or rare across the viral population; vice versa, mutant
+%alleles that reach high frequencies are expected to be neutral. If genetic sites
+%are independent, the equilibrium frequency of a deleterious allele with fitness
+%$-s$ is $\mut / |s|$, where $\mut$ is the mutation rate per site per generation;
+%neutral alleles have no equilibrium frequency and can slowly fix via genetic
+%drift~\citep{ewens_mathematical_2004}. This intuitive picture does not hold in
+%presence of genetic linkage and, in particular, for HIV evolution, because
+%recombination is rare~\citep{neher_recombination_2010, batorsky_estimate_2011}.
+%A more likely scenario, at least in escape mutation-rich regions, is the
+%following: if the focal synonymous mutant is linked to a beneficial allele
+%nearby, the latter essentially carries the synonymous allele toward high
+%frequencies for a time on the order of the inverse of the recombination rate.
+%The two slowly decouple afterwards, and the fitness effect of the synonymous
+%allele starts to be visible on its own. On the one hand, genetic linkage and
+%hitchhiking confound the interprentation of conservation levels; on the other,
+%as we show below, purifying selection can be still observed with the help of
+%longitudinal data.
 
 \section{Results}
-A neutral mutation segregating at frequency $\nu$ has a probability $\nu$ to spread through the population and fix, while it is lost with probability $1-\nu$.
-This is a simple consequence of the fact that exactly one of present $N$ individuals will be the common ancestor of the entire population at a particular locus and this ancestor has a probability $\nu$ of carrying this mutations, see illustration in \FIG{illustration}.
-Deleterious or beneficial mutations, in contrast, should fix less or more often, respectively.
-Time series sequence data therefore suggest a simple way to investigate average properties of different classes of mutations.
 
-\paragraph{Synonymous mutations in \env, C2-V5 are mostly deleterious}
+A neutral mutation segregating at frequency $\nu$ has a probability $\nu$ to
+spread through the population and fix, while it is lost with probability
+$1-\nu$. This is a simple consequence of the fact that exactly one of present
+$N$ individuals will be the common ancestor of the entire population at a
+particular locus and this ancestor has a probability $\nu$ of carrying this
+mutations, see illustration in \FIG{fixp}. Deleterious or beneficial
+mutations, in contrast, should fix less or more often, respectively. Time series
+sequence data therefore suggest a simple way to investigate average properties
+of different classes of mutations.
 
-\FIG{aft} shows time series data of the frequencies of all mutations observed \env, C2-V5, in patient ??\citep{shankarappa_consistent_1999,liu_selection_2006}.
-Despite many synonymous mutations reaching high frequency, very few fix.
-This observation in further quantified in panels ?
-and ?, that stratify the data of ??
-patients (see methods) according to the frequency at which different mutations are observed.
-Considering all mutations in a frequency interval $\nu_0$ at some time $t_i$, we calculate the fraction that is found at frequency 1, frequency 0, or at intermediate frequency at later time points $t_f$.
-Plotting these fixed, lost, and polymorphic fraction against the time interval $t_f-t_i$, we see that most synonymous mutations segregate for roughly one year and are lost  much more frequently than expected.
-The ultimate probability of loss or fixation is shown as a function of the initial frequency $\nu_0$ in panel ??.
-In contrast to synonymous mutations, the non-synonymous seem to follow more a less the neutral expectation -- a point to which we will come back below.
+\paragraph{Synonymous polymorphisms in \env, C2-V5 are mostly deleterious}
+
+\FIG{aft} shows time series data of the frequencies of all mutations observed
+\env, C2-V5, in patient 8\citep{shankarappa_consistent_1999,liu_selection_2006}.
+Despite many synonymous mutations reaching high frequency, very few fix. This
+observation in further quantified in panels \ref{fig:fixp1} and
+\ref{fig:fixp2}, that stratify the data of
+7-10 patients (see methods) according to the frequency at which different
+mutations are observed. Considering all mutations in a frequency interval
+$\nu_0$ at some time $t_i$, we calculate the fraction that is found at frequency
+1, frequency 0, or at intermediate frequency at later time points $t_f$.
+Plotting these fixed, lost, and polymorphic fraction against the time interval
+$t_f-t_i$, we see that most synonymous mutations segregate for roughly one year
+and are lost  much more frequently than expected. The ultimate probability of
+loss or fixation is shown as a function of the initial frequency $\nu_0$ in
+panel ??. In contrast to synonymous mutations, the non-synonymous seem to follow
+more a less the neutral expectation -- a point to which we will come back below.
 
@@
 \includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_nonsynp8}
-\includegraphics[width=\linewidth]{Shankarappa_fix_loss_dt_times}
-\caption{Synonymous mutations rarely fix in \env, C2-V5: Panel A shows mutation 
-frequency trajectories observed in patient ?? \cite{shankarappa_consistent_1999};  Nonsynonymous
-and synonymous mutations are shown as solid and dashed lines, respectively. 
-Colors indicate the position of the site along the C2-V5 region (red to blue) ADD COLORBAR. MAYBE MAKE FIGURE WITH SYNONYMOUS AND NONSYN SEPARATELY.
-While non-synonymous mutations frequently fix, very few synonymous mutations do even though they are frequently observed at intermediate frequencies.
-Panel B quantifies time course of loss and fixation of synonymous mutations observed in a frequency interval $\nu_0$.
-The ultimate fraction of synonymous mutations that fix as a function of intermediate frequency $\nu_0$.
-}
+\caption{Synonymous mutations rarely fix in \env, C2-V5: mutation frequency
+ trajectories observed in patient 8~\cite{shankarappa_consistent_1999};
+ Nonsynonymous and synonymous mutations are shown as solid and dashed lines,
+ respectively. Colors indicate the position of the site along the C2-V5 region
+ (red to blue) MAYBE MAKE FIGURE WITH SYNONYMOUS AND NONSYN
+ SEPARATELY. While non-synonymous mutations frequently fix, very few synonymous
+mutations do even though they are frequently observed at intermediate
+frequencies.}
 \label{fig:aft}
@@
 
-\citet{bunnik_autologous_2008} present a longitudinal data sets on the entire \env~gene of ??
-patients at ??
-time points with ??-??
-sequences each.
-Repeating the above analysis separately on the C2-V5 region studied above and the remainder of \env~ reveal strikingly different behavior inside and outside the hyper variable region.
-Within C2-V5, this data fully confirms the observations made in the data set by \citet{shankarappa_consistent_1999}.
-In the remainder of \env, however, observed synonymous mutations behave as if they were neutral; see \FIG{fixp}.
+\citet{bunnik_autologous_2008} present a longitudinal data sets on the entire
+\env~gene of 3 patients at $\sim 5$ time points with approximately 5-20
+sequences each. Repeating the above analysis separately on the C2-V5 region
+studied above and the remainder of \env~ reveal strikingly different behavior
+inside and outside the hypervariable region. Within C2-V5, this data fully
+confirms the observations made in the data set by
+\citet{shankarappa_consistent_1999}. In the remainder of \env, however, observed
+synonymous mutations behave as if they were neutral; see \FIG{fixp}.
 
@@
 
-These observations suggest that many of the synonymous mutations in the part of \env~that includes the hyper variable regions are deleterious, while outside this regions only roughly neutral mutations are polymorphic.
+These observations suggest that many of the synonymous mutations in the part of
+\env~that includes the hypervariable regions are deleterious, while outside
+this regions mostly roughly neutral mutations are polymorphic.
 
@@
 \begin{center}
-\subfloat{\includegraphics[width=0.49\linewidth]{Bunnik2008_fixmid_syn_ShankanonShanka}}
-\caption{Left panel: fixation probability of derived synonymous alleles is strongly
-suppressed in C2-V5 versus other parts of the {\it env} gene, and of
-nonsynonymous ones.
-Right panel: especially hard is fixation of new alleles in conserved regions flanking the V
-loops. The black dashed line is the prediction from neutral
-theory, for comparison purposes. Data from
+\subfloat{\includegraphics[width=0.9\linewidth]{Shankarappa_fix_loss_dt_times}
+\label{fig:fixp1}}\\
+\subfloat{\includegraphics[width=0.9\linewidth]{Bunnik2008_fixmid_syn_ShankanonShanka}
+\label{fig:fixp2}}
+\caption{Left panel: time course of loss and fixation of synonymous mutations
+ observed in a frequency interval $\nu_0$. The ultimate fraction of synonymous
+ mutations that fix as a function of intermediate frequency $\nu_0$ is the
+ fixation probability.  Right panel: fixation probability of derived synonymous
+alleles is strongly suppressed in C2-V5 versus other parts of the {\it env}
+gene, and of nonsynonymous ones. Data from
 Refs.~\cite{shankarappa_consistent_1999, bunnik_autologous_2008}.}
@@
 \paragraph{Synonymous mutations in C2-V5 tend to disrupt conserved RNA stems}
-One possible {\it a priori} explanation for lack of fixation of synonymous mutations in C2-V5 are  secondary structures in the viral RNA.
-If any RNA secondary structures are relevant for HIV replication,
-mutations in nucleotides involved in those base pairs are expected  to be deleterious and to revert preferentially. 
-Many functionally important secondary structure elements have been characterized, including  the  \rev{} response element (RRE) to enhance nuclear export of some of its transcripts~\citep{fernandes_hiv-1_2012}.
-Another well studied case is the interaction between viral reverse transcriptase, viral ssRNA, and the host tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription (RT) and bound by a specifical pseudoknotted RNA structure in the viral 5' untranslated region~\citep{barat_interaction_1991, paillart_vitro_2002}.
-It has been suggested early on that parts of the viral genome that has the potential to form stems as better conserved that the remainder \citep{forsdyke_reciprocal_1995}.
+One possible {\it a priori} explanation for lack of fixation of synonymous
+mutations in C2-V5 are  secondary structures in the viral RNA. If any RNA
+secondary structures are relevant for HIV replication, mutations in nucleotides
+involved in those base pairs are expected  to be deleterious and to revert
+preferentially.  Many functionally important secondary structure elements have
+been characterized, including  the  \rev{} response element (RRE) to enhance
+nuclear export of some of its transcripts~\citep{fernandes_hiv-1_2012}. Another
+well studied case is the interaction between viral reverse transcriptase, viral
+ssRNA, and the host tRNA$^\text{Lys3}$: the latter is required for priming
+reverse transcription (RT) and bound by a specifical pseudoknotted RNA structure
+in the viral 5' untranslated region~\citep{barat_interaction_1991,
+paillart_vitro_2002}. It has been suggested early on that parts of the viral
+genome that has the potential to form stems is better conserved that the
+remainder~\citep{forsdyke_reciprocal_1995}.
 
-Recently, the propensity of nucleotides of the HIV genome to form base pairs has been measured using the SHAPE assay (a biochemical reaction preferentially altering unpaired bases) \citep{watts_architecture_2009}.
-The SHAPE assay has shown that the variable regions V1 to V5 tend to be unpaired, while the conserved regions between those variable regions form stems.
-We partition all synonymous alleles observed
-at intermediate frequencies above 10-15\% depending on their final destiny
-(fixation or extinction). Subsequently, we align our sequences to the reference
-NL4-3 strain used in ref.~\citep{watts_architecture_2009} and assign them SHAPE
-reactivities. As shown in \figurename~\ref{fig:SHAPE} (left panel) in a
-cumulative histogram, the reactivity of fixed alleles are systematically larger
-than of alleles that are doomed to extinction. In other words, alleles that are
-likely to be breaking RNA helices are also more likely to revert and finally be
-lost from the population.
-We then split the synonymous mutations in the C2-V5 region further into conserved and variable regions and found that the biggest depression in fixation probability is observed in the conserved stems, while the variable loops show little deviations from the neutral signature; see \FIG{SHAPE}B.
+Recently, the propensity of nucleotides of the HIV genome to form base pairs has
+been measured using the SHAPE assay (a biochemical reaction preferentially
+altering unpaired bases)~\citep{watts_architecture_2009}. The SHAPE assay has
+shown that the variable regions V1 to V5 tend to be unpaired, while the
+conserved regions between those variable regions form stems. We partition all
+synonymous alleles observed at intermediate frequencies above 10-15\% depending
+on their final destiny (fixation or extinction). Subsequently, we align our
+sequences to the reference NL4-3 strain used in
+ref.~\citep{watts_architecture_2009} and assign them SHAPE reactivities. As
+shown in \FIG{SHAPE} (left panel) in a cumulative histogram, the reactivity of
+fixed alleles are systematically larger than of alleles that are doomed to
+extinction. In other words, alleles that are likely to be breaking RNA helices
+are also more likely to revert and finally be lost from the population. We then
+split the synonymous mutations in the C2-V5 region further into conserved and
+variable regions and found that the biggest depression in fixation probability
+is observed in the conserved stems, while the variable loops show little
+deviations from the neutral signature; see \FIG{SHAPE}B.
 
@@
 respectively~\citep{li_codon-usage-based_2012, ngumbela_quantitative_2008,
-coleman_virus_2008}. We do not found, however, evidence for any contribution of
+coleman_virus_2008}. We do not find, however, evidence for any contribution of
 CUB to the ultimate fate of synonymous alleles. Several lines of thought support
@@
 \begin{center}
-\subfloat{\includegraphics[width=0.49\linewidth]{mixed_Shankarappa_Bunnik2008_Liu_fixation_reactivity_Vandflanking_fromSHAPE}}
-\subfloat{\includegraphics[width=0.49\linewidth]{Shankarappa_fixmid_syn_V_regions.pdf}}
+\subfloat{\includegraphics[width=0.9\linewidth]{mixed_Shankarappa_Bunnik2008_Liu_fixation_reactivity_Vandflanking_fromSHAPE}}\\
+\subfloat{\includegraphics[width=0.9\linewidth]{Shankarappa_fixmid_syn_V_regions.pdf}}
 \caption{Watts et al. have measured the reactivity of HIV nucleotides to {\it
@@
 \paragraph{Deleterious mutations are brought to high frequency by hitch-hiking}
-While the observation that some fraction of synonymous mutations is deleterious is not unexpected, it seems odd that we observe them at high population frequency -- at least in some regions of the genome.
-The region of \env~in which we observe deleterious mutations at high frequency is special in that it undergoes frequent adaptive changes to evade recognition by neutralizing antibodies \cite{Williamson}.
-Due to the limited amount of recombination in HIV \cite{neher_recombination_2010,batorsky_2011}, deleterious mutations that are linked to adaptive variants can reach high frequency \citep{maynard_smith}.
+While the observation that some fraction of synonymous mutations is deleterious
+is not unexpected, it seems odd that we observe them at high population
+frequency -- at least in some regions of the genome. The region of \env~in which
+we observe deleterious mutations at high frequency is special in that it
+undergoes frequent adaptive changes to evade recognition by neutralizing
+antibodies~\cite{williamson_adaptation_2003}. Due to the limited amount of
+recombination in HIV \cite{neher_recombination_2010,batorsky_estimate_2011},
+deleterious mutations that are linked to adaptive variants can reach high
+frequency~\citep{smith_hitch-hiking_1974}.
 
-The potential for hitch-hiking is already apparent from the allele frequency trajectories in \FIG{aft}, where many mutations appear to change rapidly in frequency as a flock.
-Deleterious synonymous mutations can be amplified exponentially by selection on linked nonsynonymous sites, a process known as {\it genetic draft}~\citep{neher_genetic_2011}.
-In order to be advected to high frequency by a linked adaptive mutation, the deleterious effect of the mutation has to be substantially smaller than the adaptive effect.
-The latter was estimated to be on the order of $s_a = 0.01$ per day.
-The approximate magnitude of the deleterious effects can be estimated from \FIG{fixtimes}, that shows the distribution of times for synonymous alleles to reach the fix or get lost starting from intermediate frequencies.
-The typical time to loss is of the order of 500 days.
-If this loss is driven by the deleterious effect of the mutation, this corresponds to deleterious effects of roughly $0.002$ per day.
+The potential for hitch-hiking is already apparent from the allele frequency
+trajectories in \FIG{aft}, where many mutations appear to change rapidly in
+frequency as a flock. Deleterious synonymous mutations can be amplified
+exponentially by selection on linked nonsynonymous sites, a process known as
+{\it genetic draft}~\citep{gillespie_genetic_2000, neher_genetic_2011}. In order
+to be advected to high frequency by a linked adaptive mutation, the deleterious
+effect of the mutation has to be substantially smaller than the adaptive effect.
+The latter was estimated to be on the order of $s_a = 0.01$ per day. The
+approximate magnitude of the deleterious effects can be estimated from
+\FIG{fixp} (left panel), that shows the distribution of times for synonymous
+alleles to reach the fix or get lost starting from intermediate frequencies. The
+typical time to loss is of the order of 500 days. If this loss is driven by the
+deleterious effect of the mutation, this corresponds to deleterious effects of
+roughly $s_d \sim - 0.002$ per day.
 
-To get a better idea of the range of parameters that are compatible with the observations and our interpretation, we  perform computer simulations of evolving viral populations under selection and rare recombination.
-For this purpose, we use the recently published package FFPopSim, which includes a module dedicated to intra-patient
-HIV evolution~\citep{zanini_ffpopsim:_2012}. We analyze many combinations of
-parameters such as population size, recombination rate, selection coefficient
-and density of escape mutations, deleterious effect of synonymous mutation.
+To get a better idea of the range of parameters that are compatible with the
+observations and our interpretation, we  perform computer simulations of
+evolving viral populations under selection and rare recombination. For this
+purpose, we use the recently published package FFPopSim, which includes a module
+dedicated to intra-patient HIV evolution~\citep{zanini_ffpopsim:_2012}. We
+analyze many combinations of parameters such as population size, recombination
+rate, selection coefficient and density of escape mutations, deleterious effect
+of synonymous mutation.
 
-The main result of the simulations is that genetic draft can indeed bring weakly deleterious mutations to high frequencies and result in a dependence of the fixation probability on initial frequency that is compatible with observations.
-We quantify the reduction in fixation probability by the area under the diagonal....
-Since neutral mutations are much more likely to rise to high frequency than deleterious ones, the majority of the synonymous mutations needs to be slightly deleterious observe a significant reduction of $p_{fix}$.
-Furthermore, the two crucial parameters that control the fixation probability
+The main result of the simulations is that genetic draft can indeed bring weakly
+deleterious mutations to high frequencies and result in a dependence of the
+fixation probability on initial frequency that is compatible with observations.
+We quantify the reduction in fixation probability by the area under the
+diagonal~\comment{EXPLAIN!} Since neutral mutations are much more likely to rise
+to high frequency than deleterious ones, the majority of the synonymous
+mutations needs to be slightly deleterious observe a significant reduction of
+$P_\text{fix}$. Furthermore, the two crucial parameters that control the fixation probability
 are the following: (a) the deleterious effects of hitchhikers compared to
@@
 
-SHOW THAT THE DEPRESSION WORKS.
-
 \begin{figure}
 \begin{center}
-\subfloat{\includegraphics[width=0.49\linewidth]{fixation_loss_shortgenome_area_ada_frac_del_eff_coi_0_01_nescepi_6_heat.pdf}}
-\subfloat{\includegraphics[width=0.49\linewidth]{fixation_loss_shortgenome_area_ada_frac_del_eff_coi_0_01_nescepi_6_nonsyn_heat.pdf}}
-\caption{Simulations on the escape competition scenario show that the density of
- selective sweeps and the size of the deleterious effects of synonymous
- mutations are the main driving forces of the phenomenon. A convex fixation
- probability is recovered, as seen in the data, along the diagonal (left panel):
- more dense sweeps can support more deleterious linked mutations. The density of
- sweeps is limited, however, by the nonsynonymous fixation probability, which is
- quite close to neutrality (right panel). Moreover, strong competition between
- escape mutants is required, so that several escape mutants are ``found'' by HIV
-within a few months of antibody production.}
+\subfloat{\includegraphics[width=0.9\linewidth]{fixation_loss_shortgenome_distance_ada_frac_del_eff_coi_various.pdf}}\\
+\subfloat{\includegraphics[width=0.9\linewidth]{fixation_loss_shortgenome_area_ada_frac_del_eff_coi_0_01_nescepi_6_heat.pdf}}\\
+\subfloat{\includegraphics[width=0.9\linewidth]{fixation_loss_shortgenome_area_ada_frac_del_eff_coi_0_01_nescepi_6_nonsyn_heat.pdf}}
+\caption{The depression in $P_\text{fix}$ depends on the deleterious effect size
+ of the synonymous alleles (panel A). Simulations on the escape competition
+ scenario show that the density of selective sweeps and the size of the
+ deleterious effects of synonymous mutations are the main driving forces of the
+ phenomenon. A convex fixation probability is recovered, as seen in the data,
+ along the diagonal (panel B): more dense sweeps can support more deleterious
+ linked mutations. The density of sweeps is limited, however, by the
+ nonsynonymous fixation probability, which is quite close to neutrality (panel
+ C). Moreover, strong competition between escape mutants is required, so that
+ several escape mutants are ``found'' by HIV within a few months of antibody
+production.}
 \label{fig:simheat}
@@
 
-However, if hitch-hiking is driven by non-synonymous mutations that are unconditionally beneficial, we should find that non-synonymous mutations almost always fix once they reach high frequencies -- in contrast with \FIG{fixp} that shows that non-synonymous mutations fix as if they were neutral.
-We know, however, that non-synonymous variation in the variable regions is driven by positive selection.
-Inspecting the trajectories of non-synonymous mutations suggest the rapid rise and fall of many alleles.
-We test two possible such mechanisms that are biologically plausible
-and could explain the transient rise of non-synonymous mutations: time-dependent selection and
-within-epitope competition. If the immune system starts recognizes the escape mutant before its fixation, the mutant might cease to be beneficial and disappears despite its quick initial rise in
-frequency. An example for the fixation probabilities generated by this kind of
-models is shown in \figurename~\ref{fig:simfixp} (right panel). In support of this idea, \citet{richman_rapid_2003,
-bunnik_autologous_2008} report antibody responses to escape mutants.
-These respones are delayed by a few months, roughly matching the average sweep time of an escape mutant.
-Alternatively, several different escape mutations in the same epitope can arise almost simultaneously and start to spread.
-Their fitness benefits are not additive, because each of them is essentially sufficient to escape.
-As a consequence, several mutations rise to high frequency, while the escape with the smallest cost is most likely to eventually fix.
-In simulations, this kind of epistatic interactions within epitopes is reduces fixation probabilities in simulations (\figurename~\ref{fig:simfixp}, left panel).
-The emergence of multiple sweeping nonsynonymous mutations in real HIV infections has been shown
-previously~\citep{moore_limited_2009}.
+However, if hitch-hiking is driven by non-synonymous mutations that are
+unconditionally beneficial, we should find that non-synonymous mutations almost
+always fix once they reach high frequencies -- in contrast with \FIG{fixp} that
+shows that non-synonymous mutations fix as if they were neutral. We know,
+however, that non-synonymous variation in the variable regions is driven by
+positive selection. Inspecting the trajectories of non-synonymous mutations
+suggest the rapid rise and fall of many alleles.  We test two possible such
+mechanisms that are biologically plausible and could explain the transient rise
+of non-synonymous mutations: time-dependent selection and within-epitope
+competition. If the immune starts recognizes the escape mutant before its
+fixation, the mutant might cease to be beneficial and disappears despite its
+quick initial rise in frequency.  In support of this idea,
+\citet{richman_rapid_2003, bunnik_autologous_2008} report antibody responses to
+escape mutants. These respones are delayed by a few months, roughly matching the
+average sweep time of an escape mutant. Alternatively, several different escape
+mutations in the same epitope can arise almost simultaneously and start to
+spread. Their fitness benefits are not additive, because each of them is
+essentially sufficient to escape. As a consequence, several mutations rise to
+high frequency, while the escape with the smallest cost is most likely to
+eventually fix. In simulations, this kind of epistatic interactions within
+epitopes reduces fixation probabilities in simulations.  The emergence of
+multiple sweeping nonsynonymous mutations in real HIV infections has been shown
+previously~\citep{moore_limited_2009, bar_early_2012}.
 
 \section{Discussion}
-Despite several known functional roles for RNA secondary structure in the HIV genome, synonymous mutations are often used as approximately neutral markers in evolutionary studies of viruses.
-We have shown that the majority of synonymous mutations in the conserved regions C2-C5 of the \env~gene are deleterious.
-Comparison with recent biochemical studies of binding propensity of bases in RNA genome suggest that these mutations are deleterious in part because they disrupt stems in RNA secondary structure.
-Furthermore, we provide evidence that these mutations are brought to high frequency through linkage to adaptive mutations.
-The latter  mutations are only transiently adaptive, either through a coevolution with the immune system or redundant escape within an epitope.
+Despite several known functional roles for RNA secondary structure in the HIV
+genome, synonymous mutations are often used as approximately neutral markers in
+evolutionary studies of viruses. We have shown that the majority of synonymous
+mutations in the conserved regions C2-C5 of the \env~gene are deleterious.
+Comparison with recent biochemical studies of binding propensity of bases in RNA
+genome suggest that these mutations are deleterious in part because they disrupt
+stems in RNA secondary structure. Furthermore, we provide evidence that these
+mutations are brought to high frequency through linkage to adaptive mutations.
+The latter  mutations are only transiently adaptive, either through a
+coevolution with the immune system or redundant escape within an epitope.
 
-Our observations and conclusion rely heavily on longitudinal data in which the dynamics of mutations can be explicitly observed.
-The fact that deleterious mutations can be brought to high frequencies through hitch-hiking underscores the vigorousness of the coevolution with the immune system.
-The fact that multiple escape mutations in the same epitope -- as is indeed observed in studies of antibody escape \citep{sdfsd} -- are necessary to explain the patterns of fixation of non-synonymous mutations points towards a large populations size that rapidly discovers adaptive mutations.
-A similar point has been made recently by \citet{boltz_ultrasensitive_2012} in the context of preexisting drug resistance mutations.
+Our observations and conclusion rely heavily on longitudinal data in which the
+dynamics of mutations can be explicitly observed. The fact that deleterious
+mutations can be brought to high frequencies through hitch-hiking underscores
+the vigorousness of the coevolution with the immune system. The fact that
+multiple escape mutations in the same epitope -- as is indeed observed in
+studies of antibody escape~\citep{moore_limited_2009, bar_early_2012} -- are
+necessary to explain the patterns of fixation of non-synonymous mutations points
+towards a large populations size that rapidly discovers adaptive mutations. A
+similar point has been made recently by Boltz {\it et al.} in the context of
+preexisting drug resistance mutations~\citep{boltz_ultrasensitive_2012}.
 
-The observed hitch-hiking highlights the importance of linkage due to infrequent recombination for the evolution of HIV \citep{neher_recombination_2010,batorsky_estimate_2011,joseffson_smth_2011}.
-The recombination rate has been estimated to be on the order of $\rho = 10^{-5}$ per base and day.
-It takes roughly $t_{sw} = s^{-1} \log \nu_0$ generations for an adaptive mutation with growth rate $s$ to rise from an initially low frequency $\nu_0\sim \mu$ to frequency one.
-This implies that a region of length $l = (\rho t_{sw})^{-1} = s/(\rho \log \nu_0)$ remains linked to the adaptive mutation.
-With $s=0.01$, $l\approx 100$ bases which is consistent with strong linkage between the variable loops and the stems in between.
-Furthermore, we do not expect much hitch-hiking to extend far beyond the variable regions consistent with the lack of signal out side of C5-V5.
-In case of much stronger selection -- such as observed during early CTL escape or drug resistance evolution -- the linked  region is of course a lot larger.
+The observed hitch-hiking highlights the importance of linkage due to infrequent
+recombination for the evolution of HIV
+\citep{neher_recombination_2010,batorsky_estimate_2011,
+josefsson_majority_2011}. The recombination rate has been estimated to be on the
+order of $\rho = 10^{-5}$ per base and day. It takes roughly $t_{sw} = s^{-1}
+\log \nu_0$ generations for an adaptive mutation with growth rate $s$ to rise
+from an initially low frequency $\nu_0\sim \mu$ to frequency one. This implies
+that a region of length $l = (\rho t_{sw})^{-1} = s/ \rho \log \nu_0$ remains
+linked to the adaptive mutation. With $s=0.01$, $l\approx 100$ bases which is
+consistent with strong linkage between the variable loops and the stems in
+between. Furthermore, we do not expect much hitch-hiking to extend far beyond
+the variable regions consistent with the lack of signal out side of C5-V5. In
+case of much stronger selection -- such as observed during early CTL escape or
+drug resistance evolution -- the linked  region is of course a lot larger.
 
-The functional significance of the insulating RNA structure stems between the hyper variable loops has been proposed previously~\citep{watts_architecture_2009, sanjuan_interplay_2011}.
-Our analysis is akin to that in ref.~\citep{sanjuan_interplay_2011} in terms of overall goals, but provides direct evidence that insulating stems are relevant for viral fitness {\it in vivo}.
-Our analysis is limited by the availability of longitudinal data which requires a focus on the the variable regions of \env.
-Conserved RNA structures likely exist (and several are known) in different parts of the HIV genome.
-In absence of repeated adaptive substitutions in the vicinity that cause hitch-hiking, the deleterious synonymous mutations will remain at low frequencies and can only be observed by deep sequencing methods.
+The functional significance of the insulating RNA structure stems between the
+hyper variable loops has been proposed
+previously~\citep{watts_architecture_2009, sanjuan_interplay_2011}.
+\citet{sanjuan_interplay_2011} have shown that insulating stems are relevant for
+viral fitness {\it in vivo}. Our analysis is limited by the availability of
+longitudinal data which requires a focus on the the variable regions of \env.
+Conserved RNA structures likely exist (and several are known) in different parts
+of the HIV genome. In absence of repeated adaptive substitutions in the vicinity
+that cause hitch-hiking, the deleterious synonymous mutations will remain at low
+frequencies and can only be observed by deep sequencing methods.
 
@@
 literature support a mixed scenario~\citep{richman_rapid_2003,
-moore_limited_2009}. Note also that, unlike influenza, HIV does recombine if
-rarely, hence clonal interference as studied in
+moore_limited_2009, bar_early_2012}. Note also that, unlike influenza, HIV does
+recombine if rarely, hence clonal interference as studied in
 ref.~\citep{strelkowa_clonal_2012} is only a short-term effect. In conclusion,
 we regard two consequences of this state of affairs as particularly relevant for
-clinical purposes. On the one hand, the intervention of the host immune system
-appears, if late, effective in fighting escape mutants, so that an active
-stimulation of the host immune systems towards a more prompt response might be a
-viable treatment route. On the other hand, if HIV is indeed able to generate
-several escape mutants at the same time, as both data and calculations seem to
-indicate, an early response against some of them might not suffice to control
-the viral load.
+clinical purposes.
 

### 2012-12-15 synmut.tex
review, more supplementary figures

@@
 \newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
-\newcommand{\Title}{Deleterious synonymous mutations hitch-hike to high frequency in HIV \env~evolution}
+\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV \env~evolution}
 \newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
@@
 \usepackage[utf8x]{inputenc}
-\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym}
+\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym,textcomp}
 \usepackage{color}
-\usepackage{subfig}
 \usepackage{graphicx}
+\usepackage[caption=false]{subfig}
 \usepackage{natbib}
@@
 \newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
+\newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}
 
@@
 \section{Introduction}
-
 HIV evolves rapidly within a single host during the course of the infection.
@@
 mutation rate of HIV~\citep{mansky_lower_1995}. When the host develops a CTL or
-AB response against a particular viral epitope, mutations that reduce or prevent
+AB response against a particular HIV epitope, mutations in the viral genome that reduce or prevent
 recognition of the epitope frequently emerge. Escape mutations in epitopes
-targeted by CTLs typically evolve early in infection and spread rapidly through
-the population~\citep{mcmichael_immune_2009}. Later in infection, the most
+targeted by CTLs typically evolve during early infection and spread rapidly through
+the population~\citep{mcmichael_immune_2009}. During chronic infection, the most
 rapidly evolving part of the HIV genome are the so called variable loops of the
@@
 Mutations in \env~, the gene encoding for gp120, spread through the population
-within a few months (see \figurename~\ref{fig:aft}, solid lines). During chronic
-infection, the (Malthusian) effect size of this beneficial mutations is of the
+within a few months (see \figurename~\ref{fig:aft}, solid lines).
+The (Malthusian) effect size of these beneficial mutations is of the
 order of $s_a \sim 0.01$~\citep{neher_recombination_2010}.
 
-These escape mutations are selected for their effect on the amino acid sequence
-of the viral proteins. The viral genome, however, needs to meet additional
-constraints such as efficient processing and translation, nuclear export, and
-packaging into the viral capsid: all these processes operate at the RNA level. A
-few important RNA elements are well characterized. For example, a certain RNA
+These escape mutations are strongly selected for their effect on the amino acid
+sequence of the viral proteins. Conversely, synonymous mutations are commonly
+used as approximate neutral markers in studies of viral evolution. Neutral
+markers are very useful in practice, because they can be used to make inferences
+about the stochastic forces driving evolution~\citep{yang_statistical_2000}.
+The viral genome, however, needs to satisfy further constraints in addition to
+immune escape, such as efficient processing and translation, nuclear export, and
+packaging into the viral capsid: all these processes operate at the RNA level
+and are sensitive to synonymous changes. A
+few functionally important RNA elements are well characterized. For example, a certain RNA
 sequence, called \rev{} response element (RRE), is used by HIV to enhance
@@
 
-SYNONYMOUS CONSERVATION. DO WE HAVE A PLOT OF GENOME WIDE CONSERVATION, MAYBE FOR SUPPLEMENT?
+%SYNONYMOUS CONSERVATION. DO WE HAVE A PLOT OF GENOME WIDE CONSERVATION, MAYBE
+%FOR SUPPLEMENT? YES
 
-Despite evidence for functional importance of specific RNA sequences, synonymous
-mutations are commonly used as approximate neutral markers in studies of viral
-evolution. Neutral markers allow to make inference about the stochastic forces
-driving evolution~\citep{smth}. Here, we characterize the dynamics of synonymous
-mutations in \env{} and show that a substantial fraction of these mutations are
-deleterious. The central quantity we investigate is the probability of fixation
-of a mutation, conditional on its population frequency. Even though the
-synonymous mutations are deleterious and cannot be used as neutral markers, we
-show that the degree to which they hitchhike with nearby non-synonymous
-mutations is very informative; their ability to hitchhike for extended times is
-itself rooted in the small recombination rate of
+Here, we characterize the dynamics of synonymous mutations in \env{} and show
+that a substantial fraction of these mutations are deleterious. We further show
+that, although such synonymous mutations cannot be used as neutral markers, the
+degree to which they hitchhike with nearby nonsynonymous mutations is very
+informative. Their ability to hitchhike for extended times, which is a core
+requirement for our analysis, is rooted in the small recombination rate of
 HIV~\citep{neher_recombination_2010, batorsky_estimate_2011}. Extending the
-analysis of fixation probabilities to the non-synonymous mutations, we show that
+analysis of fixation probabilities to the nonsynonymous mutations, we show that
 time dependent selection or strong competition of escape mutations inside the
@@
 
-%One simple way to assess the neutrality of synonymous mutations is to look at
-%their level of conservation. Deleterious mutations at functional sites are
-%expected to be absent or rare across the viral population; vice versa, mutant
-%alleles that reach high frequencies are expected to be neutral. If genetic sites
-%are independent, the equilibrium frequency of a deleterious allele with fitness
-%$-s$ is $\mut / |s|$, where $\mut$ is the mutation rate per site per generation;
-%neutral alleles have no equilibrium frequency and can slowly fix via genetic
-%drift~\citep{ewens_mathematical_2004}. This intuitive picture does not hold in
-%presence of genetic linkage and, in particular, for HIV evolution, because
-%recombination is rare~\citep{neher_recombination_2010, batorsky_estimate_2011}.
-%A more likely scenario, at least in escape mutation-rich regions, is the
-%following: if the focal synonymous mutant is linked to a beneficial allele
-%nearby, the latter essentially carries the synonymous allele toward high
-%frequencies for a time on the order of the inverse of the recombination rate.
-%The two slowly decouple afterwards, and the fitness effect of the synonymous
-%allele starts to be visible on its own. On the one hand, genetic linkage and
-%hitchhiking confound the interprentation of conservation levels; on the other,
-%as we show below, purifying selection can be still observed with the help of
-%longitudinal data.
-
 \section{Results}
-
+The central quantity we investigate is the probability of fixation
+of a mutation, conditional on its population frequency.
 A neutral mutation segregating at frequency $\nu$ has a probability $\nu$ to
-spread through the population and fix, while it is lost with probability
-$1-\nu$. This is a simple consequence of the fact that exactly one of present
-$N$ individuals will be the common ancestor of the entire population at a
-particular locus and this ancestor has a probability $\nu$ of carrying this
-mutations, see illustration in \FIG{fixp}. Deleterious or beneficial
-mutations, in contrast, should fix less or more often, respectively. Time series
-sequence data therefore suggest a simple way to investigate average properties
+spread through the population and fix; in the rest of the cases, i.e. with probability
+$1-\nu$, it goes extinct. This is a simple consequence of the fact that (a) exactly
+one of the $N$ individuals in the current population will be the common ancestor
+of the entire future population at a
+particular locus and (b) this ancestor has a probability $\nu$ of carrying the
+mutation, see illustration in \FIG{fixp}. Deleterious or beneficial
+mutations fix less or more often than neutral ones, respectively. Time series
+sequence data enable a direct observation of both the current frequency $\nu$ of
+any particular mutation and its future fate (fixation or extinction).
+They therefore represent a simple way to investigate average properties
 of different classes of mutations.
 
-\paragraph{Synonymous polymorphisms in \env, C2-V5 are mostly deleterious}
+\subsection{Synonymous polymorphisms in \env, C2-V5 are mostly deleterious}
 
 \FIG{aft} shows time series data of the frequencies of all mutations observed
-\env, C2-V5, in patient 8\citep{shankarappa_consistent_1999,liu_selection_2006}.
-Despite many synonymous mutations reaching high frequency, very few fix. This
-observation in further quantified in panels \ref{fig:fixp1} and
-\ref{fig:fixp2}, that stratify the data of
-7-10 patients (see methods) according to the frequency at which different
-mutations are observed. Considering all mutations in a frequency interval
-$\nu_0$ at some time $t_i$, we calculate the fraction that is found at frequency
-1, frequency 0, or at intermediate frequency at later time points $t_f$.
-Plotting these fixed, lost, and polymorphic fraction against the time interval
-$t_f-t_i$, we see that most synonymous mutations segregate for roughly one year
-and are lost  much more frequently than expected. The ultimate probability of
-loss or fixation is shown as a function of the initial frequency $\nu_0$ in
-panel ??. In contrast to synonymous mutations, the non-synonymous seem to follow
-more a less the neutral expectation -- a point to which we will come back below.
-
+\env~, C2-V5, in patient p8~\citep{shankarappa_consistent_1999,
+liu_selection_2006}. Despite many synonymous mutations reaching high frequency,
+very few fix. This observation is quantified in panels \ref{fig:fixp1} and
+\ref{fig:fixp2}, which stratify the data of 7-10 patients according to the
+frequency at which different mutations are observed (see methods). Considering
+all mutations in a frequency interval around $\nu_0$ at some time $t_i$, we
+calculate the fraction that is found at frequency 1, at frequency 0, or at
+intermediate frequency at later times $t_f$. Plotting these fixed, lost, and
+polymorphic fraction against the time interval $t_f-t_i$, we see that most
+synonymous mutations segregate for roughly one year and are lost much more
+frequently than expected. The long-time probability of fixation versus
+extinction is shown as a function of the initial frequency $\nu_0$ in
+panel~\ref{fig:fixp2}. In contrast to synonymous mutations, the nonsynonymous
+seem to follow more a less the neutral expectation -- a point to which we will
+come back below. 
 \begin{figure}
@@
  (red to blue) MAYBE MAKE FIGURE WITH SYNONYMOUS AND NONSYN
- SEPARATELY. While non-synonymous mutations frequently fix, very few synonymous
+ SEPARATELY. While nonsynonymous mutations frequently fix, very few synonymous
 mutations do even though they are frequently observed at intermediate
@@
 
-\citet{bunnik_autologous_2008} present a longitudinal data sets on the entire
+\citet{bunnik_autologous_2008} present a longitudinal dataset on the entire
 \env~gene of 3 patients at $\sim 5$ time points with approximately 5-20
-sequences each. Repeating the above analysis separately on the C2-V5 region
+sequences each (see methods). Repeating the above analysis separately on the C2-V5 region
 studied above and the remainder of \env~ reveal strikingly different behavior
@@
 
-ARE OBSERVED SYNONYMOUS MUTATIONS OUTSIDE C2-V5 NEUTRAL?
-DOES LOSS/FIX CORRELATE WITH CONSERVATION.
-CAN WE LOOK AT THE AVERAGE LEVEL OF CONSERVATION STRATIFIED BY MAX FREQ?
-MAYBE WE COULD HAVE ONE -- COMPLETELY CIRCULAR -- FIGURE SHOWING LOSS/FIX VS CONSERVATION.
+%ARE OBSERVED SYNONYMOUS MUTATIONS OUTSIDE C2-V5 NEUTRAL? (?? SOME!)
+%DOES LOSS/FIX CORRELATE WITH CONSERVATION? YES.
+%MAYBE WE COULD HAVE ONE -- COMPLETELY CIRCULAR -- FIGURE SHOWING LOSS/FIX VS CONSERVATION: YES.
+%CAN WE LOOK AT THE AVERAGE LEVEL OF CONSERVATION STRATIFIED BY MAX FREQ? TRICKY: the maximal freq is achieved by hitchhiking...
 
-These observations suggest that many of the synonymous mutations in the part of
-\env~that includes the hypervariable regions are deleterious, while outside
-this regions mostly roughly neutral mutations are polymorphic.
+These observations suggest that many of the synonymous polymorphisms in the part
+of \env~that includes the hypervariable regions are deleterious, while outside
+this regions polymorphisms are mostly roughly neutral.
 
@@
 
-\paragraph{Synonymous mutations in C2-V5 tend to disrupt conserved RNA stems}
+\subsection{Synonymous mutations in C2-V5 tend to disrupt conserved RNA stems}
 One possible {\it a priori} explanation for lack of fixation of synonymous
@@
 preferentially. Many functionally important secondary structure elements have
-been characterized, including  the  \rev{} response element (RRE) to enhance
-nuclear export of some of its transcripts~\citep{fernandes_hiv-1_2012}. Another
-well studied case is the interaction between viral reverse transcriptase, viral
-ssRNA, and the host tRNA$^\text{Lys3}$: the latter is required for priming
-reverse transcription (RT) and bound by a specifical pseudoknotted RNA structure
-in the viral 5' untranslated region~\citep{barat_interaction_1991,
+been characterized, including  the RRE~\citep{fernandes_hiv-1_2012} the
+5' UTR pseudoknot interacting with thehost tRNA$^\text{Lys3}$~\citep{barat_interaction_1991,
 paillart_vitro_2002}. It has been suggested early on that parts of the viral
@@
 ref.~\citep{watts_architecture_2009} and assign them SHAPE reactivities. As
-shown in \FIG{SHAPE} (left panel) in a cumulative histogram, the reactivity of
-fixed alleles are systematically larger than of alleles that are doomed to
-extinction. In other words, alleles that are likely to be breaking RNA helices
-are also more likely to revert and finally be lost from the population. We then
-split the synonymous mutations in the C2-V5 region further into conserved and
+shown in \FIG{SHAPEA} in a cumulative histogram, the reactivity of fixed alleles
+are systematically larger than of alleles that are doomed to extinction
+(Kolmogorov-Smirnov test, $P\approx 2~\text{\textperthousand}$). In other
+words, alleles that are likely to be breaking RNA helices are also more likely
+to revert and finally be lost from the population.  As a complementary analysis,
+we split the synonymous mutations in the C2-V5 region further into conserved and
 variable regions and found that the biggest depression in fixation probability
 is observed in the conserved stems, while the variable loops show little
-deviations from the neutral signature; see \FIG{SHAPE}B.
+deviation from the neutral signature, see \FIG{SHAPEB}.
 
@@
 minor effect in our context.
-
 \begin{figure}
 \begin{center}
-\subfloat{\includegraphics[width=0.9\linewidth]{mixed_Shankarappa_Bunnik2008_Liu_fixation_reactivity_Vandflanking_fromSHAPE}}\\
-\subfloat{\includegraphics[width=0.9\linewidth]{Shankarappa_fixmid_syn_V_regions.pdf}}
+\subfloat{\includegraphics[width=0.9\linewidth]{mixed_Shankarappa_Bunnik2008_Liu_fixation_reactivity_Vandflanking_fromSHAPE}
+\label{fig:SHAPEA}}\\
+\subfloat{\includegraphics[width=0.9\linewidth]{Shankarappa_fixmid_syn_V_regions.pdf}\label{fig:SHAPEB}}
 \caption{Watts et al. have measured the reactivity of HIV nucleotides to {\it
@@
 
-\paragraph{Deleterious mutations are brought to high frequency by hitch-hiking}
+\subsection{Deleterious mutations are brought to high frequency by hitch-hiking}
 While the observation that some fraction of synonymous mutations is deleterious
 is not unexpected, it seems odd that we observe them at high population
-frequency -- at least in some regions of the genome. The region of \env~in which
+frequency -- at least in some regions of the genome. The region of \env~ in which
 we observe deleterious mutations at high frequency is special in that it
@@
 antibodies~\cite{williamson_adaptation_2003}. Due to the limited amount of
-recombination in HIV \cite{neher_recombination_2010,batorsky_estimate_2011},
+recombination in HIV~\cite{neher_recombination_2010,batorsky_estimate_2011},
 deleterious mutations that are linked to adaptive variants can reach high
@@
 
-The potential for hitch-hiking is already apparent from the allele frequency
+The potential for hitchhiking is already apparent from the allele frequency
 trajectories in \FIG{aft}, where many mutations appear to change rapidly in
@@
 effect of the mutation has to be substantially smaller than the adaptive effect.
-The latter was estimated to be on the order of $s_a = 0.01$ per day. The
-approximate magnitude of the deleterious effects can be estimated from
-\FIG{fixp} (left panel), that shows the distribution of times for synonymous
+The latter was estimated to be on the order of $s_a = 0.01$ per day~\citep{neher_recombination_2010}.
+The approximate magnitude of the deleterious effects can be estimated from
+\FIG{fixp1}, that shows the distribution of times for synonymous
 alleles to reach the fix or get lost starting from intermediate frequencies. The
@@
 purpose, we use the recently published package FFPopSim, which includes a module
-dedicated to intra-patient HIV evolution~\citep{zanini_ffpopsim:_2012}. We
+dedicated to intrapatient HIV evolution~\citep{zanini_ffpopsim:_2012}. We
 analyze many combinations of parameters such as population size, recombination
@@
 fixation probability on initial frequency that is compatible with observations.
-We quantify the reduction in fixation probability by the area under the
-diagonal~\comment{EXPLAIN!} Since neutral mutations are much more likely to rise
-to high frequency than deleterious ones, the majority of the synonymous
-mutations needs to be slightly deleterious observe a significant reduction of
-$P_\text{fix}$. Furthermore, the two crucial parameters that control the fixation probability
-are the following: (a) the deleterious effects of hitchhikers compared to
-the beneficial effects of escape mutants, and (b) the density of escape
-mutations. Intuitively, a higher density of escape mutations (i.e., epitopes)
-enables a larger degree of genetic draft, because escape mutations start to
-combine and their effects add up. In \figurename~\ref{fig:simheat} (left panel),
+First of all, since neutral mutations
+are much more likely to rise to high frequency than deleterious ones, the
+majority of the synonymous mutations needs to be slightly deleterious observe a
+significant reduction of $P_\text{fix}$.
+In order to further quantify the reduction in fixation probability, we look at the
+difference between the neutral curve ($P(\nu) = \nu$) and the measured fixation
+probability and calculate its area (see inset of \FIG{simfixpvar}). The minimal and maximal values for this
+area are zero (neutral-like curve) and 0.5 (no fixation at all),
+respectively. The convex curve seen in the HIV data corresponds to an area of
+approximately 0.17. Various simulation curves are shown in \FIG{simfixpvar}, and
+the area of the data curve is shown in the inset.
+In \FIGS{simheat1}{simheat2}, we explore the parameter space: the combinations that yield areas close to the
+experimental result are roughly indicated by ellipses. The two crucial parameters
+that control the fixation probability are the following: (a) the deleterious
+effects of hitchhikers compared to the beneficial effects of escape mutants, and
+(b) the density of escape mutations. Intuitively, a higher density of escape
+mutations (i.e., epitopes) enables a larger degree of genetic draft, because
+escape mutations start to combine and their effects add up. In \FIG{simheat1},
 we show that this is indeed the case in simulations.
@@
 \begin{center}
-\subfloat{\includegraphics[width=0.9\linewidth]{fixation_loss_shortgenome_distance_ada_frac_del_eff_coi_various.pdf}}\\
-\subfloat{\includegraphics[width=0.9\linewidth]{fixation_loss_shortgenome_area_ada_frac_del_eff_coi_0_01_nescepi_6_heat.pdf}}\\
-\subfloat{\includegraphics[width=0.9\linewidth]{fixation_loss_shortgenome_area_ada_frac_del_eff_coi_0_01_nescepi_6_nonsyn_heat.pdf}}
+\subfloat{\includegraphics[width=0.9\linewidth]{fixation_loss_shortgenome_distance_ada_frac_del_eff_coi_various.pdf}
+\label{fig:simfixpvar}}\\
+\subfloat{\includegraphics[width=0.9\linewidth]{fixation_loss_shortgenome_area_ada_frac_del_eff_coi_0_01_nescepi_6_heat.pdf}
+\label{fig:simheat1}}\\
+\subfloat{\includegraphics[width=0.9\linewidth]{fixation_loss_shortgenome_area_ada_frac_del_eff_coi_0_01_nescepi_6_nonsyn_heat.pdf}
+\label{fig:simheat2}}
 \caption{The depression in $P_\text{fix}$ depends on the deleterious effect size
@@
 
-However, if hitch-hiking is driven by non-synonymous mutations that are
-unconditionally beneficial, we should find that non-synonymous mutations almost
+However, if hitch-hiking is driven by nonsynonymous mutations that are
+unconditionally beneficial, we should find that nonsynonymous mutations almost
 always fix once they reach high frequencies -- in contrast with \FIG{fixp} that
-shows that non-synonymous mutations fix as if they were neutral. We know,
-however, that non-synonymous variation in the variable regions is driven by
-positive selection. Inspecting the trajectories of non-synonymous mutations
+shows that nonsynonymous mutations fix as if they were neutral. We know,
+however, that nonsynonymous variation in the variable regions is driven by
+positive selection. Inspecting the trajectories of nonsynonymous mutations
 suggest the rapid rise and fall of many alleles.  We test two possible such
 mechanisms that are biologically plausible and could explain the transient rise
-of non-synonymous mutations: time-dependent selection and within-epitope
-competition. If the immune starts recognizes the escape mutant before its
-fixation, the mutant might cease to be beneficial and disappears despite its
+of nonsynonymous mutations: time-dependent selection and within-epitope
+competition. If the immune system recognizes the escape mutant before its
+fixation, the mutant might cease to be beneficial and disappear despite its
 quick initial rise in frequency.  In support of this idea,
@@
 spread. Their fitness benefits are not additive, because each of them is
-essentially sufficient to escape. As a consequence, several mutations rise to
-high frequency, while the escape with the smallest cost is most likely to
+essentially sufficient to escape. As a consequence, several escape mutations rise to
+high frequency, while the escape with the smallest cost in terms of replication,
+packaging, etc. is most likely to
 eventually fix. In simulations, this kind of epistatic interactions within
-epitopes reduces fixation probabilities in simulations.  The emergence of
+epitopes reduces the fixation probability. The emergence of
 multiple sweeping nonsynonymous mutations in real HIV infections has been shown
 previously~\citep{moore_limited_2009, bar_early_2012}.
+See the supplementary material for examples of successful simulations in both scenarios.
 
@@
 Comparison with recent biochemical studies of binding propensity of bases in RNA
-genome suggest that these mutations are deleterious in part because they disrupt
-stems in RNA secondary structure. Furthermore, we provide evidence that these
+genome suggest that these mutations are deleterious, at least in part, because they disrupt
+stems in RNA secondary structures. Furthermore, we provide evidence that these
 mutations are brought to high frequency through linkage to adaptive mutations.
@@
 dynamics of mutations can be explicitly observed. The fact that deleterious
-mutations can be brought to high frequencies through hitch-hiking underscores
-the vigorousness of the coevolution with the immune system. The fact that
+mutations can be brought to high frequencies through hitchhiking underscores
+the intensity of the coevolution with the immune system. The fact that
 multiple escape mutations in the same epitope -- as is indeed observed in
 studies of antibody escape~\citep{moore_limited_2009, bar_early_2012} -- are
-necessary to explain the patterns of fixation of non-synonymous mutations points
+necessary to explain the patterns of fixation of nonsynonymous mutations points
 towards a large populations size that rapidly discovers adaptive mutations. A
@@
 
-The observed hitch-hiking highlights the importance of linkage due to infrequent
+The observed hitchhiking highlights the importance of linkage due to infrequent
 recombination for the evolution of HIV
@@
 josefsson_majority_2011}. The recombination rate has been estimated to be on the
-order of $\rho = 10^{-5}$ per base and day. It takes roughly $t_{sw} = s^{-1}
-\log \nu_0$ generations for an adaptive mutation with growth rate $s$ to rise
+order of $\rho = 10^{-5}$ per base and day. It takes roughly $t_{sw} = s_a^{-1}
+\log \nu_0$ generations for an adaptive mutation with growth rate $s_a$ to rise
 from an initially low frequency $\nu_0\sim \mu$ to frequency one. This implies
-that a region of length $l = (\rho t_{sw})^{-1} = s/ \rho \log \nu_0$ remains
-linked to the adaptive mutation. With $s=0.01$, $l\approx 100$ bases which is
+that a region of length $l = (\rho t_{sw})^{-1} = s_a / \rho \log \nu_0$ remains
+linked to the adaptive mutation. With $s_a=0.01$, $l\approx 100$ bases which is
 consistent with strong linkage between the variable loops and the stems in
-between. Furthermore, we do not expect much hitch-hiking to extend far beyond
+between. Furthermore, we do not expect hitchhiking to extend far beyond
 the variable regions consistent with the lack of signal out side of C5-V5. In
 case of much stronger selection -- such as observed during early CTL escape or
-drug resistance evolution -- the linked  region is of course a lot larger.
+drug resistance evolution -- the linked  region is of course much larger.
 
@@
 viral fitness {\it in vivo}. Our analysis is limited by the availability of
-longitudinal data which requires a focus on the the variable regions of \env.
-Conserved RNA structures likely exist (and several are known) in different parts
-of the HIV genome. In absence of repeated adaptive substitutions in the vicinity
-that cause hitch-hiking, the deleterious synonymous mutations will remain at low
+longitudinal data which requires a focus on the the variable regions of \env~.
+Conserved RNA structures most likely exist in different parts
+of the HIV genome (several are known). In absence of repeated adaptive substitutions in the vicinity
+that cause hitchhiking, the deleterious synonymous mutations remain at low
 frequencies and can only be observed by deep sequencing methods.
@@
 fixation and extinction times and probabilities represent a rich and simple
-summary statistics to test sequencing data and computer simulation upon, as
-noted independently in ref.~\citep{strelkowa_clonal_2012} in the context of
-influenza. Furthermore, our results emphasize the inadequacy of independent-site
+summary statistics to test sequencing data and computer simulation upon. A
+similar method has been recently used in a longitudinal study of
+influenza~\citep{strelkowa_clonal_2012}. The propagators suggested in that
+paper, however, represent ratios between (certain kinds of) nonsynonymous
+mutations and synonymous ones, hence they are inadequate to investigate
+synonymous changes themselves. Those authors also conclude that several
+beneficial mutations segregate simultaneously in influenza, a scenario
+remarkably similar to our within-epitope competition picture. These results
+jointly suggest that viral evolution proceeds by multiple concurrent sweeps
+rather then by successive fixation~\citep{desai_beneficial_2007, neher_rate_2010}.
+
+Finally, our results emphasize the inadequacy of independent site
 models of HIV evolution, especially in the light of transient effects on
 sweeping sites, such as time-dependent selection and within-epitope negative
-epistasis. Although a final word about which of these mechanisms is more
+epistasis. Although a final word about which mechanism is more
 widespread is yet to be spoken, both intuition and biological evidence from the
@@
 recombine if rarely, hence clonal interference as studied in
-ref.~\citep{strelkowa_clonal_2012} is only a short-term effect. In conclusion,
-we regard two consequences of this state of affairs as particularly relevant for
-clinical purposes.
+ref.~\citep{strelkowa_clonal_2012} is only a short-term effect.
 

### 2012-12-20 synmut.tex
Methods

@@
 \section{Methods}
-\comment{to be written\dots}
+\subsection{Sequence data collection}
+Longitudinal intrapatient viral RNA sequences were collected for published
+studies~\citep{shankarappa_consistent_1999,
+liu_selection_2006, bunnik_autologous_2008} and downloaded from the Los Alamos
+National Laboratory (LANL) HIV sequence database~\citep{LANL2012}. The sequences from
+some patients showed signs of HIV compartimentalization into subpopulations and
+were discarded; a grand total of 11
+patients with approximately 6 time points each and 10 sequences per time point
+were analyzed. The time interval or resolution between two ocnsecutive sequences
+was approximately 6 to 18 months.
+
+\subsection{Sequence analisys}
+The good sequences were aligned within each patient
+via the translated amino acid sequence, using
+Muscle~\citep{edgar_muscle:_2004}, and to the NL4-3 reference sequence probed
+by \citet{watts_architecture_2009}. Within each patient, a consensus RNA
+sequence at the first time point was used to classify alleles as ancestral or
+derived at all sites. Problematic sites that included large frequencies of gaps
+were excluded from the analysis, because variable regions are known to be
+subject to frequent indels, while our analysis is limited to nucleotide
+substitutions. Time series of allele frequencies were extracted from the
+sequences.
+
+The synonymity of a mutation was assigned if the rest of the codon was
+in the ancestral state and using the standard genetic code. Cases where more
+than one mutation within the codon was observed were discarded. Slightly
+different criteria for synonymous/nonsynonymous discrimination yielded similar
+results.
+
+\subsection{Fixation probability and secondary structure}
+For the estimate of times to fixation/extinction, polymorphisms were
+binned by frequency and the time to reaching the first boundary (fixation or
+extinction) was stored. For the fixation probability, the long-time limit of the
+resulting curves was used, excluding polymorphisms that arose late in the
+clinical history (and would have had no time to reach either boundary).
+
+For the correlation analysis with RNA secondary structure, the SHAPE scores were
+downloaded from the journal website~\citep{watts_architecture_2009}. By virtue
+of the alignment of the longitudinal sequences with the reference used by
+\citet{watts_architecture_2009}, SHAPE reactivities were assigned to most sites.
+Problematic assignments in indel-rich regions were excluded from the analysis.
+In order to restrict the analysis to synonymous polymorphisms, a lower frequency
+threshold of 0.15 was used (other thresholds yielded the same results). Since
+very few polymorphisms hitchhike beyond, say, a frequency of 0.5, this pool is
+enriched for to-be-lost mutations; hence the "lost" curve in \FIG{SHAPEA}
+contains much more points than the "fixed" one.
+
+The V loops and flanking regions were identified manually starting from the
+annotated reference HXB2 sequence from the LANL HIV database~\citep{LANL2012}. A
+similar approach was used to label the C2-V5 region sequenced in
+ref.~\citep{shankarappa_consistent_1999}.
+
+\subsection{Computer simulations}
+Simulations were performed using the recently published software
+FFPopSim~\citep{zanini_ffpopsim:_2012}. Both full-length HIV genomes and
+\env{}-only simulations were performed and yielded comparable results. For each
+set of parameters, approximately 100 simulation runs were averaged over. In each
+run, a random fitness landscape with specified statistical properties (e.g.
+density of beneficial sites, average deleterious effect of synonymous changes) was generated.
+Although the curves shown in \FIG{simfixpvar} are not very smooth, small
+parameter changes resulted in overall consistent trends across many repetitions.
+
+For the discussion of simulation parameters, the areas below or above the neutral
+diagonal were estimated from the binned fixation probabilities using the linear
+interpolation between the bin centers. This measure is sufficiently precise for
+our purposes, because the HIV data are quite scarse themselves.
+
 \section*{Acknowledgements}

### 2012-12-21 synmut.tex
small changes here and there

@@
 \noindent
-Intrapatient HIV evolution is goverened by selection on the protein level in the
-arms race with the immune system (killer T-cells and antibodies). Synonymous
-mutations do not have an immunity-related phenotype and are often assumed to be
-neutral. In this paper, we show that synonymous changes in epitope-rich regions
-are often deleterious but still reach frequencies of order one.  We analyze time
-series of viral sequences from the V1-C5 part of {\it env} within individual
-hosts and observe that synonymous derived alleles rarely fix in the
-viral population. Simulations suggest that such synonymous mutations
-have a (Malthisuan) selection coefficient of the order of $-0.001$, and that
-they are brought up to high frequency by linkage to neighbouring beneficial
-nonsynonymous alleles (genetic draft). As far as the biological causes are
-concerned, we detect a negative correlation between fixation of an allele and
-its involvement in evolutionarily conserved RNA stem-loop structures.
-This phenonenon is not observed in other parts of the HIV genome, in which
-selective sweeps are less dense and the genetic architecture less constrained.
+
+Intrapatient HIV evolution is goverened by selection on the protein level in
+the arms race with the immune system (killer T-cells and antibodies).
+Synonymous mutations do not have an immunity-related phenotype and are often
+assumed to be neutral. In this paper, we show that synonymous changes in
+epitope-rich regions are often deleterious but reach frequencies of order one
+via genetic hitchhiking.  We analyze time series of viral sequences from the
+V1-C5 part of {\it env}, the envelope gene, within individual hosts and observe
+that synonymous derived alleles fix in the viral population much less
+frequently than expected from neutral models. Instead, synonymous changes tend
+to revert on a time scale of two years, suggesting a selection coefficient of the
+order of $-1~$\textperthousand. Extensive computer simulations support these
+findings quantitatively. We explore possible biological causes and detect a
+negative correlation between fixation of an allele and its involvement in
+secondary RNA stem-loop structures, indicating that such structures are
+functionally relevant for HIV replication {\it in vivo}. This phenonenon is not
+observed in other parts of the HIV genome, in which selective sweeps are less
+dense and the genetic architecture less constrained.
+
 \end{abstract}
-
 \maketitle
@@
 \section{Introduction}
+
 HIV evolves rapidly within a single host during the course of the infection.
@@
 via killer T cells (CTLs) and neutralizing antibodies
-(AB)~\citep{pantaleo_immunopathogenesis_1996} and facilitated by the high
+(ABs)~\citep{pantaleo_immunopathogenesis_1996} and facilitated by the high
 mutation rate of HIV~\citep{mansky_lower_1995}. When the host develops a CTL or
-AB response against a particular HIV epitope, mutations in the viral genome that reduce or prevent
-recognition of the epitope frequently emerge. Escape mutations in epitopes
-targeted by CTLs typically evolve during early infection and spread rapidly through
-the population~\citep{mcmichael_immune_2009}. During chronic infection, the most
-rapidly evolving part of the HIV genome are the so called variable loops of the
-envelope protein gp120, which need to avoid recognition by neutralizing ABs.
-Mutations in \env~, the gene encoding for gp120, spread through the population
-within a few months (see \figurename~\ref{fig:aft}, solid lines).
-The (Malthusian) effect size of these beneficial mutations is of the
+AB response against a particular HIV epitope, mutations in the viral genome that
+reduce or prevent recognition of the epitope frequently emerge. Escape mutations
+in epitopes targeted by CTLs typically evolve during early infection and spread
+rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
+infection, the most rapidly evolving part of the HIV genome are the so called
+variable loops of the envelope protein gp120, which need to avoid recognition by
+neutralizing ABs.  Mutations in \env~, the gene encoding for gp120, spread
+through the population within a few months (see \figurename~\ref{fig:aft}, solid
+lines).  The (Malthusian) effect size of these beneficial mutations is of the
 order of $s_a \sim 0.01$~\citep{neher_recombination_2010}.
@@
 packaging into the viral capsid: all these processes operate at the RNA level
-and are sensitive to synonymous changes. A
-few functionally important RNA elements are well characterized. For example, a certain RNA
-sequence, called \rev{} response element (RRE), is used by HIV to enhance
-nuclear export of some of its transcripts~\citep{fernandes_hiv-1_2012}. Another
-well studied case is the interaction between viral reverse transcriptase, viral
-ssRNA, and the host tRNA$^\text{Lys3}$: the latter is required for priming
-reverse transcription (RT) and bound by a specifical pseudoknotted RNA structure
-in the viral 5' untranslated region~\citep{barat_interaction_1991,
-paillart_vitro_2002}. Furthermore, recent studies have shown that genetically
-engineered HIV strains with skewed codon usage bias (CUB) patterns towards more
-or less abundant tRNAs replicate better or worse,
-respectively~\citep{ngumbela_quantitative_2008, li_codon-usage-based_2012}.
-Purifying selection beyond the protein sequence is therefore expected, while it
-seems reasonable that the bulk of positive selection through the immune system
-should be restricted to amino acid sequences.
-
-INFLUENZA PSEUDO VACCINE.
+and are sensitive to synonymous changes. A few functionally important RNA
+elements are well characterized. For example, a certain RNA sequence in the HIV
+genome, called \rev{} response element (RRE), enhances nuclear export of viral
+transcripts~\citep{fernandes_hiv-1_2012}. Another well studied case is the
+interaction between viral reverse transcriptase, viral ssRNA, and the host
+tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription
+(RT) and bound by a specifical pseudoknotted RNA structure in the viral 5'
+untranslated region~\citep{barat_interaction_1991, paillart_vitro_2002}.
+Nucleotide-level fitness effects have been observed beyond RNA structure as
+well. Recent studies have shown that genetically engineered HIV strains with
+skewed codon usage bias (CUB) patterns towards more or less abundant tRNAs
+replicate better or worse, respectively~\citep{ngumbela_quantitative_2008,
+li_codon-usage-based_2012}. A similar conclusion has been reached about
+influenza, and codon-pessimized influenza strains have been shown to be good
+live attenuated vaccines in mice~\citep{mueller_live_2010}. Purifying selection
+beyond the protein sequence is therefore expected, while it seems reasonable
+that the bulk of positive selection through the immune system be restricted to
+amino acid sequences.
 
@@
 
-Here, we characterize the dynamics of synonymous mutations in \env{} and show
-that a substantial fraction of these mutations are deleterious. We further show
-that, although such synonymous mutations cannot be used as neutral markers, the
-degree to which they hitchhike with nearby nonsynonymous mutations is very
-informative. Their ability to hitchhike for extended times, which is a core
-requirement for our analysis, is rooted in the small recombination rate of
+In this paper, we characterize the dynamics of synonymous mutations in \env{}
+and show that a substantial fraction of these mutations is deleterious. We
+further show that, although such synonymous mutations cannot be used as neutral
+markers, the degree to which they hitchhike with nearby nonsynonymous mutations
+is very informative. Their ability to hitchhike for extended times, which is a
+core requirement for our analysis, is rooted in the small recombination rate of
 HIV~\citep{neher_recombination_2010, batorsky_estimate_2011}. Extending the
@@
 \section{Results}
-The central quantity we investigate is the probability of fixation
-of a mutation, conditional on its population frequency.
-A neutral mutation segregating at frequency $\nu$ has a probability $\nu$ to
-spread through the population and fix; in the rest of the cases, i.e. with probability
-$1-\nu$, it goes extinct. This is a simple consequence of the fact that (a) exactly
-one of the $N$ individuals in the current population will be the common ancestor
-of the entire future population at a
-particular locus and (b) this ancestor has a probability $\nu$ of carrying the
-mutation, see illustration in \FIG{fixp}. Deleterious or beneficial
-mutations fix less or more often than neutral ones, respectively. Time series
-sequence data enable a direct observation of both the current frequency $\nu$ of
-any particular mutation and its future fate (fixation or extinction).
-They therefore represent a simple way to investigate average properties
-of different classes of mutations.
+
+The central quantity we investigate is the probability of fixation of a
+mutation, conditional on its population frequency.  A neutral mutation
+segregating at frequency $\nu$ has a probability $P_\text{fix}(\nu) = \nu$ to
+spread through the population and fix; in the rest of the cases, i.e. with
+probability $1-\nu$, it goes extinct. This is a simple consequence of the fact
+that (a) exactly one of the $N$ individuals in the current population will be
+the common ancestor of the entire future population at a particular locus and
+(b) this ancestor has a probability $\nu$ of carrying the mutation, see
+illustration in \FIG{fixp}.  Deleterious or beneficial mutations fix less or
+more often than neutral ones, respectively. Time series sequence data enable a
+direct observation of both the current frequency $\nu$ of any particular
+mutation and its future fate (fixation or extinction). They therefore represent
+a simple way to investigate average properties of different classes of
+mutations.
 
@@
 \FIG{aft} shows time series data of the frequencies of all mutations observed
-\env~, C2-V5, in patient p8~\citep{shankarappa_consistent_1999,
-liu_selection_2006}. Despite many synonymous mutations reaching high frequency,
-very few fix. This observation is quantified in panels \ref{fig:fixp1} and
-\ref{fig:fixp2}, which stratify the data of 7-10 patients according to the
-frequency at which different mutations are observed (see methods). Considering
-all mutations in a frequency interval around $\nu_0$ at some time $t_i$, we
-calculate the fraction that is found at frequency 1, at frequency 0, or at
-intermediate frequency at later times $t_f$. Plotting these fixed, lost, and
-polymorphic fraction against the time interval $t_f-t_i$, we see that most
-synonymous mutations segregate for roughly one year and are lost much more
-frequently than expected. The long-time probability of fixation versus
-extinction is shown as a function of the initial frequency $\nu_0$ in
-panel~\ref{fig:fixp2}. In contrast to synonymous mutations, the nonsynonymous
-seem to follow more a less the neutral expectation -- a point to which we will
-come back below. 
+\env~, C2-V5, in patient p8~\citep{shankarappa_consistent_1999}. Despite many
+synonymous mutations reaching high frequency (dashed lines), very few fix. This
+observation is quantified in panels \FIG{fixp1} and \ref{fig:fixp2}, which
+stratify the data of 7 (resp. 10) patients according to the frequency at which
+different mutations are observed (see methods). Considering all mutations in a
+frequency interval around $\nu_0$ at some time $t_i$, we calculate the fraction
+that is found at frequency 1, at frequency 0, or at intermediate frequency at
+later times $t_f$. Plotting these fixed, lost, and polymorphic fraction against
+the time interval $t_f-t_i$, we see that most synonymous mutations segregate for
+roughly two years and are lost much more frequently than expected. The long-time
+probability of fixation versus extinction is shown as a function of the initial
+frequency $\nu_0$ in panel~\ref{fig:fixp2} (red line). In contrast to synonymous
+mutations, the nonsynonymous seem to follow more a less the neutral expectation
+(blue line) -- a point to which we will come back below.
+
 \begin{figure}
@@
 \env~gene of 3 patients at $\sim 5$ time points with approximately 5-20
-sequences each (see methods). Repeating the above analysis separately on the C2-V5 region
-studied above and the remainder of \env~ reveal strikingly different behavior
-inside and outside the hypervariable region. Within C2-V5, this data fully
-confirms the observations made in the data set by
-\citet{shankarappa_consistent_1999}. In the remainder of \env, however, observed
-synonymous mutations behave as if they were neutral; see \FIG{fixp}.
+sequences each (see methods). Repeating the above analysis separately on the
+C2-V5 region studied above and the remainder of \env~ reveal striking
+differences (see \FIG{fixp}). Within C2-V5, this data fully confirms the
+observations made in the data set by \citet{shankarappa_consistent_1999} (red
+line). In the remainder of \env, however, observed synonymous mutations behave
+as if they were neutral (orange line).
 
@@
 %DOES LOSS/FIX CORRELATE WITH CONSERVATION? YES.
-%MAYBE WE COULD HAVE ONE -- COMPLETELY CIRCULAR -- FIGURE SHOWING LOSS/FIX VS CONSERVATION: YES.
+%MAYBE WE COULD HAVE ONE -- COMPLETELY CIRCULAR -- FIGURE SHOWING LOSS/FIX VS CONSERVATION: SUPPLEMENTARY?
 %CAN WE LOOK AT THE AVERAGE LEVEL OF CONSERVATION STRATIFIED BY MAX FREQ? TRICKY: the maximal freq is achieved by hitchhiking...
@@
 \subsection{Synonymous mutations in C2-V5 tend to disrupt conserved RNA stems}
+
 One possible {\it a priori} explanation for lack of fixation of synonymous
@@
 preferentially. Many functionally important secondary structure elements have
-been characterized, including  the RRE~\citep{fernandes_hiv-1_2012} the
-5' UTR pseudoknot interacting with thehost tRNA$^\text{Lys3}$~\citep{barat_interaction_1991,
-paillart_vitro_2002}. It has been suggested early on that parts of the viral
-genome that has the potential to form stems is better conserved that the
+been characterized, including  the RRE~\citep{fernandes_hiv-1_2012} and the 5'
+UTR pseudoknot interacting with the host
+tRNA$^\text{Lys3}$~\citep{barat_interaction_1991, paillart_vitro_2002}. It has
+been suggested early on that parts of the viral genome that has the potential to
+form stems is better conserved than the
 remainder~\citep{forsdyke_reciprocal_1995}.
@@
 ref.~\citep{watts_architecture_2009} and assign them SHAPE reactivities. As
-shown in \FIG{SHAPEA} in a cumulative histogram, the reactivity of fixed alleles
-are systematically larger than of alleles that are doomed to extinction
-(Kolmogorov-Smirnov test, $P\approx 2~\text{\textperthousand}$). In other
-words, alleles that are likely to be breaking RNA helices are also more likely
-to revert and finally be lost from the population.  As a complementary analysis,
-we split the synonymous mutations in the C2-V5 region further into conserved and
-variable regions and found that the biggest depression in fixation probability
-is observed in the conserved stems, while the variable loops show little
-deviation from the neutral signature, see \FIG{SHAPEB}.
+shown in \FIG{SHAPEA} in a cumulative histogram, the reactivities of fixed
+alleles (red line) are systematically larger than of alleles that are doomed to
+extinction (blue line) (Kolmogorov-Smirnov test, $P\approx
+2~\text{\textperthousand}$). In other words, alleles that are likely to be
+breaking RNA helices are also more likely to revert and finally be lost from the
+population. As a control, the average over non-observed but potentially
+available polymorphisms lies between the two curves (green line), as expected
+(because only some of them will be helix breakers). Furthermore, as a
+complementary analysis, we split the synonymous mutations in the extended V1-V5
+region further into conserved and variable regions and found that the biggest
+depression in fixation probability is observed in the conserved stems, while the
+variable loops show little deviation from the neutral signature, see
+\FIG{SHAPEB}.
 
@@
 \subsection{Deleterious mutations are brought to high frequency by hitch-hiking}
+
 While the observation that some fraction of synonymous mutations is deleterious
 is not unexpected, it seems odd that we observe them at high population
-frequency -- at least in some regions of the genome. The region of \env~ in which
-we observe deleterious mutations at high frequency is special in that it
-undergoes frequent adaptive changes to evade recognition by neutralizing
+frequency -- at least in some regions of the genome. The region of \env~ in
+which we observe deleterious mutations at high frequency, however, is special in
+that it undergoes frequent adaptive changes to evade recognition by neutralizing
 antibodies~\cite{williamson_adaptation_2003}. Due to the limited amount of
@@
 analyze many combinations of parameters such as population size, recombination
-rate, selection coefficient and density of escape mutations, deleterious effect
+rate, selection coefficient and density of escape mutations, and deleterious effect
 of synonymous mutation.
@@
 fixation probability on initial frequency that is compatible with observations.
-First of all, since neutral mutations
-are much more likely to rise to high frequency than deleterious ones, the
-majority of the synonymous mutations needs to be slightly deleterious observe a
-significant reduction of $P_\text{fix}$.
-In order to further quantify the reduction in fixation probability, we look at the
-difference between the neutral curve ($P(\nu) = \nu$) and the measured fixation
-probability and calculate its area (see inset of \FIG{simfixpvar}). The minimal and maximal values for this
-area are zero (neutral-like curve) and 0.5 (no fixation at all),
-respectively. The convex curve seen in the HIV data corresponds to an area of
-approximately 0.17. Various simulation curves are shown in \FIG{simfixpvar}, and
-the area of the data curve is shown in the inset.
-In \FIGS{simheat1}{simheat2}, we explore the parameter space: the combinations that yield areas close to the
-experimental result are roughly indicated by ellipses. The two crucial parameters
-that control the fixation probability are the following: (a) the deleterious
-effects of hitchhikers compared to the beneficial effects of escape mutants, and
-(b) the density of escape mutations. Intuitively, a higher density of escape
-mutations (i.e., epitopes) enables a larger degree of genetic draft, because
-escape mutations start to combine and their effects add up. In \FIG{simheat1},
-we show that this is indeed the case in simulations.
+First of all, since neutral mutations are much more likely to rise to high
+frequency than deleterious ones, the majority of the synonymous mutations needs
+to be slightly deleterious observe a significant reduction of $P_\text{fix}$.
+In order to further quantify the reduction in fixation probability, we look at
+the difference between the neutral curve ($P_\text{fix}(\nu) = \nu$) and the
+measured fixation probability and calculate its area (see inset of
+\FIG{simfixpvar}). The minimal and maximal values for this area are zero
+(neutral-like curve) and 0.5 (no fixation at all), respectively. The HIV data
+correspond to an area under the diagonal of approximately 0.2 for synonymous
+changes, and a very small area over the diagonal for nonsynonymous changes.
+Various simulation curves are shown in \FIG{simfixpvar}. Then, in
+\FIGS{simheat1}{simheat2}, we explore the parameter space: the combinations that
+yield areas close to the experimental result are roughly indicated by ellipses.
+The two crucial parameters that control the fixation probability are the
+following: (a) the deleterious effects of hitchhikers compared to the beneficial
+effects of escape mutants, and (b) the density of escape mutations. Intuitively,
+a higher density of escape mutations (i.e., epitopes) enables a larger degree of
+genetic draft, because escape mutations from different epitopes start to combine
+and their effects add up. In \FIG{simheat1}, we show that this is indeed the
+case in simulations.
 
@@
 \label{fig:simheat2}}
-\caption{The depression in $P_\text{fix}$ depends on the deleterious effect size
+\caption{\comment{USE A GOOD FIG WITH THE CORRECT $s_a$!} The depression in $P_\text{fix}$ depends on the deleterious effect size
  of the synonymous alleles (panel A). Simulations on the escape competition
@@
 
-However, if hitch-hiking is driven by nonsynonymous mutations that are
+However, if hitchhiking is driven by nonsynonymous mutations that are
 unconditionally beneficial, we should find that nonsynonymous mutations almost
@@
 positive selection. Inspecting the trajectories of nonsynonymous mutations
-suggest the rapid rise and fall of many alleles.  We test two possible such
+suggest the rapid rise and fall of many alleles.  We test two possible
 mechanisms that are biologically plausible and could explain the transient rise
@@
 between. Furthermore, we do not expect hitchhiking to extend far beyond
-the variable regions consistent with the lack of signal out side of C5-V5. In
+the variable regions consistent with the lack of signal out side of C1-V5. In
 case of much stronger selection -- such as observed during early CTL escape or
@@
 viral fitness {\it in vivo}. Our analysis is limited by the availability of
-longitudinal data which requires a focus on the the variable regions of \env~.
+longitudinal data which requires a focus on the the variable regions of \env.
 Conserved RNA structures most likely exist in different parts

### 2012-12-23 synmut.tex
minor edits

--- 2012-12-15 synmut.tex
@@
 \newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
+\newcommand{\env}{\textit{env}}
+
 \newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV \env~evolution}
@@
 \newcommand{\OO}{\mathcal{O}}
-\newcommand{\env}{\textit{env}}
 \newcommand{\rev}{\textit{rev}}
@@
 \newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}
+\newcommand{\shankaregion{C2-C5}
 
@@
 \noindent
-Intrapatient HIV evolution is goverened by selection on the protein level in the
+Intrapatient HIV evolution is governed by selection on the protein level in the
 arms race with the immune system (killer T-cells and antibodies). Synonymous
@@
 neutral. In this paper, we show that synonymous changes in epitope-rich regions
-are often deleterious but still reach frequencies of order one.  We analyze time
-series of viral sequences from the V1-C5 part of {\it env} within individual
+are often deleterious but still reach high frequencies.  We analyze time
+series of viral sequences from the \shankaregion~part of {\it env} within individual
 hosts and observe that synonymous derived alleles rarely fix in the
 viral population. Simulations suggest that such synonymous mutations
-have a (Malthisuan) selection coefficient of the order of $-0.001$, and that
-they are brought up to high frequency by linkage to neighbouring beneficial
-nonsynonymous alleles (genetic draft). As far as the biological causes are
-concerned, we detect a negative correlation between fixation of an allele and
-its involvement in evolutionarily conserved RNA stem-loop structures.
-This phenonenon is not observed in other parts of the HIV genome, in which
-selective sweeps are less dense and the genetic architecture less constrained.
+have a (Malthusian) selection coefficient of the order of $-0.001$, and that
+they are brought up to high frequency by hitchhiking with neighboring beneficial
+nonsynonymous alleles. We detect a negative correlation between fixation of an allele and
+its involvement in RNA stem-loop structures.
+Deleterious synonymous mutations are observed in other parts of the HIV genome, in which
+selective sweeps are less dense and hitch-hiking not as strong.
 \end{abstract}
@@
 envelope protein gp120, which need to avoid recognition by neutralizing ABs.
-Mutations in \env~, the gene encoding for gp120, spread through the population
+Mutations in \env, the gene encoding for gp120, spread through the population
 within a few months (see \figurename~\ref{fig:aft}, solid lines).
@@
 and are sensitive to synonymous changes. A
-few functionally important RNA elements are well characterized. For example, a certain RNA
+few functionally important RNA elements are well characterized and show substantially 
+less synonymous variation across the HIV population (see Fig.~S1). For example, a certain RNA
 sequence, called \rev{} response element (RRE), is used by HIV to enhance
@@
 ssRNA, and the host tRNA$^\text{Lys3}$: the latter is required for priming
-reverse transcription (RT) and bound by a specifical pseudoknotted RNA structure
+reverse transcription (RT) and bound by a pseudoknotted RNA structure
 in the viral 5' untranslated region~\citep{barat_interaction_1991,
@@
 respectively~\citep{ngumbela_quantitative_2008, li_codon-usage-based_2012}.
+Similarly, influenza can be dramatically attenuated by reducing its translation efficiency \citep{pseudo_vaccine}.
 Purifying selection beyond the protein sequence is therefore expected, while it
@@
 should be restricted to amino acid sequences.
-
-INFLUENZA PSEUDO VACCINE.
-
-%SYNONYMOUS CONSERVATION. DO WE HAVE A PLOT OF GENOME WIDE CONSERVATION, MAYBE
-%FOR SUPPLEMENT? YES
 
@@
 degree to which they hitchhike with nearby nonsynonymous mutations is very
-informative. Their ability to hitchhike for extended times, which is a core
-requirement for our analysis, is rooted in the small recombination rate of
+informative. Their ability to hitchhike for extended times is rooted in the small recombination rate of
 HIV~\citep{neher_recombination_2010, batorsky_estimate_2011}. Extending the
@@
 synonymous mutations segregate for roughly one year and are lost much more
-frequently than expected. The long-time probability of fixation versus
+frequently than expected if they were neutral. The long-time probability of fixation versus
 extinction is shown as a function of the initial frequency $\nu_0$ in
@@
 of \env~that includes the hypervariable regions are deleterious, while outside
-this regions polymorphisms are mostly roughly neutral.
+this regions polymorphisms are mostly roughly neutral. Note that this does not imply that all 
+synonymous mutations are neutral -- only those mutations observed at high frequencies tend 
+to be neutral.
 
@@
 involved in those base pairs are expected to be deleterious and to revert
-preferentially. Many functionally important secondary structure elements have
-been characterized, including  the RRE~\citep{fernandes_hiv-1_2012} the
-5' UTR pseudoknot interacting with thehost tRNA$^\text{Lys3}$~\citep{barat_interaction_1991,
-paillart_vitro_2002}. It has been suggested early on that parts of the viral
-genome that has the potential to form stems is better conserved that the
-remainder~\citep{forsdyke_reciprocal_1995}.
+preferentially. Indeed, synonymous variation across the HIV population is variable, 
+suggesting purifying selection beyond the protein coding level.
 

### 2012-12-23 synmut.tex

--- 2012-12-21 synmut.tex
@@
 \noindent
-
-Intrapatient HIV evolution is goverened by selection on the protein level in
-the arms race with the immune system (killer T-cells and antibodies).
-Synonymous mutations do not have an immunity-related phenotype and are often
-assumed to be neutral. In this paper, we show that synonymous changes in
-epitope-rich regions are often deleterious but reach frequencies of order one
-via genetic hitchhiking.  We analyze time series of viral sequences from the
-V1-C5 part of {\it env}, the envelope gene, within individual hosts and observe
-that synonymous derived alleles fix in the viral population much less
-frequently than expected from neutral models. Instead, synonymous changes tend
-to revert on a time scale of two years, suggesting a selection coefficient of the
-order of $-1~$\textperthousand. Extensive computer simulations support these
-findings quantitatively. We explore possible biological causes and detect a
-negative correlation between fixation of an allele and its involvement in
-secondary RNA stem-loop structures, indicating that such structures are
-functionally relevant for HIV replication {\it in vivo}. This phenonenon is not
-observed in other parts of the HIV genome, in which selective sweeps are less
-dense and the genetic architecture less constrained.
+Intrapatient HIV evolution is governed by selection on the protein level in the
+arms race with the immune system (killer T-cells and antibodies). Synonymous
+mutations do not have an immunity-related phenotype and are often assumed to be
+neutral. In this paper, we show that synonymous changes in epitope-rich regions
+are often deleterious but still reach high frequencies.  We analyze time
+series of viral sequences from the \shankaregion~part of {\it env} within individual
+hosts and observe that synonymous derived alleles rarely fix in the
+viral population. Simulations suggest that such synonymous mutations
+have a (Malthusian) selection coefficient of the order of $-0.001$, and that
+they are brought up to high frequency by hitchhiking with neighboring beneficial
+nonsynonymous alleles. We detect a negative correlation between fixation of an allele and
+its involvement in RNA stem-loop structures.
+Deleterious synonymous mutations are observed in other parts of the HIV genome, in which
+selective sweeps are less dense and hitch-hiking not as strong.
+\end{abstract}
 

### 2013-01-02 synmut.tex
don't know exactly
figures updates
merge few details

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
-\newcommand{\env}{\textit{env}}
-
-\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV \env~evolution}
-\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
 \usepackage[english]{babel}
@@
 \usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
-\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\DeclareMathOperator\de{d\!}
 \newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
@@
 \newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}
-\newcommand{\shankaregion{C2-C5}
-
+\newcommand{\env}{\textit{env}}
+\newcommand{\shankaregion}{C2-C5}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
+\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV \env~evolution}
+\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
+\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 have a (Malthusian) selection coefficient of the order of $-0.001$, and that
-they are brought up to high frequency by hitchhiking with neighboring beneficial
+they are brought up to high frequency by hitchhiking on neighboring beneficial
 nonsynonymous alleles. We detect a negative correlation between fixation of an allele and
 its involvement in RNA stem-loop structures.
-Deleterious synonymous mutations are observed in other parts of the HIV genome, in which
+Deleterious synonymous mutations are not observed as abundantly in other parts of the HIV genome, in which
 selective sweeps are less dense and hitch-hiking not as strong.
-\end{abstract}
 
@@
 \begin{center}
-\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_nonsynp8}
+\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_nonsynp8.pdf}
 \caption{Synonymous mutations rarely fix in \env, C2-V5: mutation frequency

### 2013-01-04 synmut.tex
edits throughout, added a few references

@@
 via killer T cells (CTLs) and neutralizing antibodies
-(ABs)~\citep{pantaleo_immunopathogenesis_1996} and facilitated by the high
+(ABs)~\citep{rambaut_causes_2004} and facilitated by the high
 mutation rate of HIV~\citep{mansky_lower_1995}. When the host develops a CTL or
@@
 variable loops of the envelope protein gp120, which need to avoid recognition by
-neutralizing ABs.  Mutations in \env~, the gene encoding for gp120, spread
+neutralizing ABs.  Mutations in \env, the gene encoding for gp120, spread
 through the population within a few months (see \figurename~\ref{fig:aft}, solid
-lines).  The (Malthusian) effect size of these beneficial mutations is of the
-order of $s_a \sim 0.01$~\citep{neher_recombination_2010}.
+lines).
 
@@
 sequence of the viral proteins. Conversely, synonymous mutations are commonly
-used as approximate neutral markers in studies of viral evolution. Neutral
-markers are very useful in practice, because they can be used to make inferences
-about the stochastic forces driving evolution~\citep{yang_statistical_2000}.
-The viral genome, however, needs to satisfy further constraints in addition to
-immune escape, such as efficient processing and translation, nuclear export, and
-packaging into the viral capsid: all these processes operate at the RNA level
-and are sensitive to synonymous changes. A few functionally important RNA
+used as approximately neutral markers in studies of viral evolution. Neutral
+markers are very useful in practice, because there evolution can be compared to
+that of putatively functional sites to detect purifying or directional selection
+\citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In most species,
+however, it is  known that synonymous codons don't evolve completely neutrally
+but that some codons are favored over others \citep{plotkin_synonymous_2011}.
+In addition to maintaining protein function and avoiding immune
+recognition, the HIV genome has to ensure efficient processing and translation,
+nuclear export, and packaging into the viral capsid: all these processes operate at the RNA
+level and are sensitive to synonymous changes. A few functionally important RNA
 elements are well characterized. For example, a certain RNA sequence in the HIV
@@
 replicate better or worse, respectively~\citep{ngumbela_quantitative_2008,
-li_codon-usage-based_2012}. A similar conclusion has been reached about
-influenza, and codon-pessimized influenza strains have been shown to be good
-live attenuated vaccines in mice~\citep{mueller_live_2010}. Purifying selection
-beyond the protein sequence is therefore expected, while it seems reasonable
-that the bulk of positive selection through the immune system be restricted to
-amino acid sequences.
+li_codon-usage-based_2012}. A similar conclusion has been reached in influenza,
+and codon-pessimized strains have been shown to be good live attenuated vaccines
+in mice~\citep{mueller_live_2010}. Purifying selection beyond the protein
+sequence is therefore expected, while it seems reasonable that the bulk of
+positive selection through the immune system be restricted to amino acid
+sequences.
 
@@
 markers, the degree to which they hitchhike with nearby nonsynonymous mutations
-is very informative. Their ability to hitchhike for extended times, which is a
-core requirement for our analysis, is rooted in the small recombination rate of
+is very informative both about their on deleterious effect, as well as the
+strength and nature of selection for nearby positively selected sites.
+Their ability to hitchhike for extended times, which is a core requirement for
+our analysis, is rooted in the small recombination rate of
 HIV~\citep{neher_recombination_2010, batorsky_estimate_2011}. Extending the
@@
 probability $1-\nu$, it goes extinct. This is a simple consequence of the fact
-that (a) exactly one of the $N$ individuals in the current population will be
+that (i) exactly one of the $N$ individuals in the current population will be
 the common ancestor of the entire future population at a particular locus and
-(b) this ancestor has a probability $\nu$ of carrying the mutation, see
+(ii) this ancestor has a probability $\nu$ of carrying the mutation, see
 illustration in \FIG{fixp}.  Deleterious or beneficial mutations fix less or
@@
 direct observation of both the current frequency $\nu$ of any particular
-mutation and its future fate (fixation or extinction). They therefore represent
-a simple way to investigate average properties of different classes of
-mutations.
+mutation and its future fate (fixation or extinction).
 
-\subsection{Synonymous polymorphisms in \env, C2-V5 are mostly deleterious}
+\subsection{Synonymous polymorphisms in \env, C2-V5, are mostly deleterious}
 
 \FIG{aft} shows time series data of the frequencies of all mutations observed
-\env~, C2-V5, in patient p8~\citep{shankarappa_consistent_1999}. Despite many
+\env, C2-V5, in patient p8~\citep{shankarappa_consistent_1999}. Despite many
 synonymous mutations reaching high frequency (dashed lines), very few fix. This
 observation is quantified in panels \FIG{fixp1} and \ref{fig:fixp2}, which
-stratify the data of 7 (resp. 10) patients according to the frequency at which
-different mutations are observed (see methods). Considering all mutations in a
+stratify the data of 7 (resp. 10 ???) patients according to the frequency at
+which different mutations are observed (see methods). Considering all mutations in a
 frequency interval around $\nu_0$ at some time $t_i$, we calculate the fraction
@@
 the time interval $t_f-t_i$, we see that most synonymous mutations segregate for
-roughly two years and are lost much more frequently than expected. The long-time
+roughly one year and are lost much more frequently than expected. The long-time
 probability of fixation versus extinction is shown as a function of the initial
@@
 mutations, the nonsynonymous seem to follow more a less the neutral expectation
-(blue line) -- a point to which we will come back below.
+(blue line) -- a point to which we will come back below. Note that very few
+synonymous polymorphisms remain polymorphic for more than years, confirming that
+the viral populations investigated here are well mixed and didn't split into
+several isolated subpopulations. ? out of the ?? patients studied in
+\citet{shankarappa_consistent_1999} did display such a split and were excluded
+from the analysis; see methods and supplement.
 
@@
 this regions polymorphisms are mostly roughly neutral. Note that this does not imply that all 
-synonymous mutations are neutral -- only those mutations observed at high frequencies tend 
-to be neutral.
+synonymous mutations outside C2-V5 are neutral -- but those mutations observed
+at high frequencies tend to be neutral. Within C2-V5, however, deleterious
+mutations reach high frequency.
 
@@
 involved in those base pairs are expected to be deleterious and to revert
-preferentially. Many functionally important secondary structure elements have
-been characterized, including  the RRE~\citep{fernandes_hiv-1_2012} and the 5'
-UTR pseudoknot interacting with the host
-tRNA$^\text{Lys3}$~\citep{barat_interaction_1991, paillart_vitro_2002}. It has
-been suggested early on that parts of the viral genome that has the potential to
-form stems is better conserved than the
-remainder~\citep{forsdyke_reciprocal_1995}.
+preferentially. In addition to the specific functional secondary structure
+elements discussed above (e.g., the RRE), \citet{forsdyke_reciprocal_1995}
+observed that parts of the viral genome that has potential to
+form stems (as predicted by RNA folding algorithms) is more conserved than the
+remainder.
 
@@
 available polymorphisms lies between the two curves (green line), as expected
-(because only some of them will be helix breakers). Furthermore, as a
-complementary analysis, we split the synonymous mutations in the extended V1-V5
-region further into conserved and variable regions and found that the biggest
-depression in fixation probability is observed in the conserved stems, while the
-variable loops show little deviation from the neutral signature, see
-\FIG{SHAPEB}.
+(because only some of these mutations will interfere with stem loop formation).
+To test the hypothesis that mutations in C2-V5 are lost since break stems
+in the conserved between the variable loops, we split the synonymous mutations
+in the extended V1-V5 region further into conserved and variable regions and
+found that the biggest depression in fixation probability is observed in the
+conserved stems, while the variable loops show little deviation from the
+neutral signature, see \FIG{SHAPEB}. This is consistent with important stem
+structures in conservered regions between loops.
 
@@
 usage bias (CUB). HIV is known to prefer A-rich codons over highly expressed
-human housekeeping genes~\citep{jenkins_extent_2003}. Moreover, codon-optimized
+human codons~\citep{jenkins_extent_2003}. Moreover, codon-optimized
 and -pessimized viruses have recently been generated and shown to replicate
 better or worse than wild type strains,
-respectively~\citep{li_codon-usage-based_2012, ngumbela_quantitative_2008,
+respectively~\citep{li_codon-usage-based_2012,ngumbela_quantitative_2008,
 coleman_virus_2008}. We do not find, however, evidence for any contribution of
@@
 {\it in vitro}, the distance in CUB between HIV and human genes is not shrinking
-at the macroevolutionary level. Second, within a single patient, we do not
+at the macroevolutionary level REF?. Second, within a single patient, we do not
 observe any bias towards more human-like CUB in the synonymous mutations that
@@
 minor effect in our context.
+
 \begin{figure}
@@
 \FIG{fixp1}, that shows the distribution of times for synonymous
-alleles to reach the fix or get lost starting from intermediate frequencies. The
+alleles to fix or get lost starting from intermediate frequencies. The
 typical time to loss is of the order of 500 days. If this loss is driven by the
@@
 evolving viral populations under selection and rare recombination. For this
-purpose, we use the recently published package FFPopSim, which includes a module
+purpose, we use FFPopSim, which includes a module
 dedicated to intrapatient HIV evolution~\citep{zanini_ffpopsim:_2012}. We
@@
 
-The main result of the simulations is that genetic draft can indeed bring weakly
+We find that genetic draft can indeed bring weakly
 deleterious mutations to high frequencies and result in a dependence of the
 fixation probability on initial frequency that is compatible with observations.
-First of all, since neutral mutations are much more likely to rise to high
+The ranges of parameters that are compatible put tight bounds several
+parameters of HIV evolution.
+Since neutral mutations are much more likely to rise to high
 frequency than deleterious ones, the majority of the synonymous mutations needs
-to be slightly deleterious observe a significant reduction of $P_\text{fix}$.
+to be deleterious to observe a significant reduction of $P_\text{fix}$.
 In order to further quantify the reduction in fixation probability, we look at
@@
 
+\comment{CAN WE DO MORE HERE? WHAT about many simulations while varying all 4
+parameters, pick those that work. Do a PCA on the result and display in two dimensions?? }
+
 \begin{figure}
@@
  along the diagonal (panel B): more dense sweeps can support more deleterious
- linked mutations. The density of sweeps is limited, however, by the
- nonsynonymous fixation probability, which is quite close to neutrality (panel
+ linked mutations. 
+ \comment{Panel C is confusing. we should be able to do smth better} The density
+ of sweeps is limited, however, by the nonsynonymous fixation probability, which is quite close to neutrality (panel
  C). Moreover, strong competition between escape mutants is required, so that
@@
 
-However, if hitchhiking is driven by nonsynonymous mutations that are
+If hitchhiking is driven by nonsynonymous mutations that are
 unconditionally beneficial, we should find that nonsynonymous mutations almost
@@
 mutations are brought to high frequency through linkage to adaptive mutations.
-The latter mutations are only transiently adaptive, either through a
+Most of the latter mutations are only transiently adaptive, either through a
 coevolution with the immune system or redundant escape within an epitope.
@@
 linked to the adaptive mutation. With $s_a=0.01$, $l\approx 100$ bases which is
-consistent with strong linkage between the variable loops and the stems in
-between. Furthermore, we do not expect hitchhiking to extend far beyond
-the variable regions consistent with the lack of signal out side of C1-V5. In
+consistent with strong linkage between the adaptive mutations in variable loops
+and intervening conserved stems. Furthermore, we do not expect hitchhiking to
+extend far beyond the variable regions consistent with the lack of signal out side of C1-V5. In
 case of much stronger selection -- such as observed during early CTL escape or
-drug resistance evolution -- the linked  region is of course much larger.
+drug resistance evolution -- the affected region is of course much larger
+\citep{nijhuis_stochastic_1998}.
 
@@
 
-As far as population genetics models are concerned, our study uncovers the
+Our study uncovers the
 subtle balance of evolutionary forces governing intrapatient HIV evolution. The
 fixation and extinction times and probabilities represent a rich and simple
-summary statistics to test sequencing data and computer simulation upon. A
-similar method has been recently used in a longitudinal study of
+summary statistics useful to characterize sequencing data and compare to
+models via computer simulations.
+A similar method has been recently used in a longitudinal study of
 influenza~\citep{strelkowa_clonal_2012}. The propagators suggested in that
-paper, however, represent ratios between (certain kinds of) nonsynonymous
+paper, however, represent ratios between nonsynonymous
 mutations and synonymous ones, hence they are inadequate to investigate
-synonymous changes themselves. Those authors also conclude that several
+synonymous changes themselves. The authors also conclude that several
 beneficial mutations segregate simultaneously in influenza, a scenario
-remarkably similar to our within-epitope competition picture. These results
-jointly suggest that viral evolution proceeds by multiple concurrent sweeps
-rather then by successive fixation~\citep{desai_beneficial_2007, neher_rate_2010}.
+remarkably similar to our within-epitope competition picture. While
+recombination hardly affects evolution within an epitope, it might facilitate
+immune escape by easing competition between different epitopes
+\citep{neher_rate_2010,Rouzine:2005p17398}.
 
-Finally, our results emphasize the inadequacy of independent site
-models of HIV evolution, especially in the light of transient effects on
-sweeping sites, such as time-dependent selection and within-epitope negative
-epistasis. Although a final word about which mechanism is more
-widespread is yet to be spoken, both intuition and biological evidence from the
-literature support a mixed scenario~\citep{richman_rapid_2003,
-moore_limited_2009, bar_early_2012}. Note also that, unlike influenza, HIV does
-recombine if rarely, hence clonal interference as studied in
-ref.~\citep{strelkowa_clonal_2012} is only a short-term effect.
+Our results emphasize the inadequacy of independent site
+models of HIV evolution and the common assumption that selection is time
+independent. If genetic variation is only transiently beneficial, existing
+methods to quantify selection will yield substantial underestimates
+\citep{williamson_adaptation_2003,neher_rate_2010,OTHER}. To explain the
+observations regarding the fixation probabilities of non-synonymous mutations,
+either transient selection, or substational within-epitope competition are
+necessary. Which mechanism is more widespread is not clear as of now,
+there is evidence for both~\citep{richman_rapid_2003, moore_limited_2009,
+bar_early_2012}.
 
@@
 patients with approximately 6 time points each and 10 sequences per time point
-were analyzed. The time interval or resolution between two ocnsecutive sequences
+were analyzed. The time interval or resolution between two consecutive sequences
 was approximately 6 to 18 months.
 
-\subsection{Sequence analisys}
+\comment{How did you determine who to discard. Maybe we should include the PCA
+plots into the supplement.}
+
+\subsection{Sequence analysis}
 The good sequences were aligned within each patient
 via the translated amino acid sequence, using
-Muscle~\citep{edgar_muscle:_2004}, and to the NL4-3 reference sequence probed
-by \citet{watts_architecture_2009}. Within each patient, a consensus RNA
-sequence at the first time point was used to classify alleles as ancestral or
+Muscle~\citep{edgar_muscle:_2004}, and to the NL4-3 reference sequence used
+by \citet{watts_architecture_2009} in the SHAPE assay. Within each patient, a
+consensus RNA sequence at the first time point was used to classify alleles as ancestral or
 derived at all sites. Problematic sites that included large frequencies of gaps
-were excluded from the analysis, because variable regions are known to be
-subject to frequent indels, while our analysis is limited to nucleotide
-substitutions. Time series of allele frequencies were extracted from the
+were excluded from the analysis to avoid artefactual substitutions due to
+alignment errors. Time series of allele frequencies were extracted from the
 sequences.
@@
 very few polymorphisms hitchhike beyond, say, a frequency of 0.5, this pool is
-enriched for to-be-lost mutations; hence the "lost" curve in \FIG{SHAPEA}
-contains much more points than the "fixed" one.
+enriched for to-be-lost mutations; hence the ``lost" curve in \FIG{SHAPEA}
+contains much more points than the ``fixed" one.
 

### 2013-01-22 synmut.tex
new figures from simulations

--- 2013-01-02 synmut.tex
@@
 Various simulation curves are shown in \FIG{simfixpvar}. Then, in
-\FIGS{simheat1}{simheat2}, we explore the parameter space: the combinations that
-yield areas close to the experimental result are roughly indicated by ellipses.
+\FIG{simsfig}, we explore the parameter space: the combinations that
+yield areas close to the experimental result are shown.
 The two crucial parameters that control the fixation probability are the
@@
 \begin{center}
-\subfloat{\includegraphics[width=0.9\linewidth]{fixation_loss_shortgenome_distance_ada_frac_del_eff_coi_various.pdf}
+\subfloat{\includegraphics[width=\linewidth]{simulations_graduallydel.pdf}
 \label{fig:simfixpvar}}\\
-\subfloat{\includegraphics[width=0.9\linewidth]{fixation_loss_shortgenome_area_ada_frac_del_eff_coi_0_01_nescepi_6_heat.pdf}
-\label{fig:simheat1}}\\
-\subfloat{\includegraphics[width=0.9\linewidth]{fixation_loss_shortgenome_area_ada_frac_del_eff_coi_0_01_nescepi_6_nonsyn_heat.pdf}
-\label{fig:simheat2}}
-\caption{\comment{USE A GOOD FIG WITH THE CORRECT $s_a$!} The depression in $P_\text{fix}$ depends on the deleterious effect size
- of the synonymous alleles (panel A). Simulations on the escape competition
- scenario show that the density of selective sweeps and the size of the
- deleterious effects of synonymous mutations are the main driving forces of the
- phenomenon. A convex fixation probability is recovered, as seen in the data,
- along the diagonal (panel B): more dense sweeps can support more deleterious
- linked mutations. The density of sweeps is limited, however, by the
- nonsynonymous fixation probability, which is quite close to neutrality (panel
- C). Moreover, strong competition between escape mutants is required, so that
- several escape mutants are ``found'' by HIV within a few months of antibody
-production.}
+\subfloat{\includegraphics[width=\linewidth]{simulations_syn.pdf}
+\label{fig:simsfig}}
+\caption{(Panel A) The depression in $P_\text{fix}$ depends on the deleterious
+effect size of the synonymous alleles. This parameter also reduces synonymous
+diversity (first inset). The depression found in HIV data is about 0.2.
+(Panel B) To assess the parameter space that affects synonymous fixation and
+diversity, we run many simulations with random parameters for deleterious effect
+size, fraction of deleterious synonymous sites, average size of escape
+mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
+introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
+generation). Mostly simulations with small deleterious effects and relatively
+high deleterious fractions of synonymous mutations reproduce synonymous HIV-like
+fixation probability and diversity.}
 \label{fig:simheat}

### 2013-01-23 synmut.tex
Text adapted to new figures

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\renewcommand{\thesubfigure}{\Alph{subfigure}}
 \newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
@@
 neutral. In this paper, we show that synonymous changes in epitope-rich regions
-are often deleterious but still reach high frequencies.  We analyze time
+are often deleterious but still reach high frequencies in the viral population.  We analyze time
 series of viral sequences from the \shankaregion~part of {\it env} within individual
-hosts and observe that synonymous derived alleles rarely fix in the
-viral population. Simulations suggest that such synonymous mutations
+hosts and observe that synonymous derived alleles rarely reach fixation.
+Simulations suggest that such synonymous mutations
 have a (Malthusian) selection coefficient of the order of $-0.001$, and that
 they are brought up to high frequency by hitchhiking on neighboring beneficial
-nonsynonymous alleles. We detect a negative correlation between fixation of an allele and
+nonsynonymous alleles. We detect a negative correlation between the fixation of an allele and
 its involvement in RNA stem-loop structures.
 Deleterious synonymous mutations are not observed as abundantly in other parts of the HIV genome, in which
-selective sweeps are less dense and hitch-hiking not as strong.
+selective sweeps are less dense and hitchhiking not as strong; this behaviour is
+confirmed by extensive computer simulations.
 
@@
 
-These escape mutations are strongly selected for their effect on the amino acid
-sequence of the viral proteins. Conversely, synonymous mutations are commonly
+Escape mutations are strongly selected for their effect on the amino acid
+sequence of the HIV proteins. Conversely, synonymous mutations are commonly
 used as approximate neutral markers in studies of viral evolution. Neutral
@@
 well. Recent studies have shown that genetically engineered HIV strains with
-skewed codon usage bias (CUB) patterns towards more or less abundant tRNAs
+a skewed codon usage bias towards more or less abundant tRNAs
 replicate better or worse, respectively~\citep{ngumbela_quantitative_2008,
@@
 live attenuated vaccines in mice~\citep{mueller_live_2010}. Purifying selection
-beyond the protein sequence is therefore expected, while it seems reasonable
+beyond the protein sequence is therefore expected, but it seems reasonable
 that the bulk of positive selection through the immune system be restricted to
@@
 markers, the degree to which they hitchhike with nearby nonsynonymous mutations
-is very informative. Their ability to hitchhike for extended times, which is a
-core requirement for our analysis, is rooted in the small recombination rate of
+is very informative; their ability to hitchhike for extended times itself is
+rooted in the small recombination rate of
 HIV~\citep{neher_recombination_2010, batorsky_estimate_2011}. Extending the
@@
 spread through the population and fix; in the rest of the cases, i.e. with
-probability $1-\nu$, it goes extinct. This is a simple consequence of the fact
-that (a) exactly one of the $N$ individuals in the current population will be
+probability $1-\nu$, it goes extinct. As illustrated in the inset of \FIG{fixp},
+this is a simple consequence of the fact that
+(a) exactly one of the $N$ individuals in the current population will be
 the common ancestor of the entire future population at a particular locus and
-(b) this ancestor has a probability $\nu$ of carrying the mutation, see
-illustration in \FIG{fixp}.  Deleterious or beneficial mutations fix less or
+(b) this ancestor has a probability $\nu$ of carrying the mutation.
+Deleterious or beneficial mutations fix less or
 more often than neutral ones, respectively. Time series sequence data enable a
@@
 \env~, C2-V5, in patient p8~\citep{shankarappa_consistent_1999}. Despite many
-synonymous mutations reaching high frequency (dashed lines), very few fix. This
-observation is quantified in panels \FIG{fixp1} and \ref{fig:fixp2}, which
-stratify the data of 7 (resp. 10) patients according to the frequency at which
+synonymous mutations reaching high frequency, very few fix
+(panel~\ref{fig:aftsyn}); however, quite some nonsynonymous mutations reach
+fixation (panel~\ref{fig:aftnonsyn}).
+These observations are quantified in \FIG{fixp}, which stratifies the data
+of 7 (resp. 10) patients according to the frequency at which
 different mutations are observed (see methods). Considering all mutations in a
@@
 the time interval $t_f-t_i$, we see that most synonymous mutations segregate for
-roughly two years and are lost much more frequently than expected. The long-time
-probability of fixation versus extinction is shown as a function of the initial
-frequency $\nu_0$ in panel~\ref{fig:fixp2} (red line). In contrast to synonymous
-mutations, the nonsynonymous seem to follow more a less the neutral expectation
-(blue line) -- a point to which we will come back below.
+roughly two years and are lost much more frequently than expected (panel
+\ref{fig:fixp1}). The long-time probability of fixation versus extinction of
+synonymous mutations is shown as a function of the initial frequency $\nu_0$
+in panel~\ref{fig:fixp2} (red line). The nonsynonymous seem to follow more or
+less the neutral expectation (blue line) -- a point to which we will come back
+below.
 
@@
 \begin{center}
-\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_nonsynp8.pdf}
-\caption{Synonymous mutations rarely fix in \env, C2-V5: mutation frequency
- trajectories observed in patient 8~\cite{shankarappa_consistent_1999};
- Nonsynonymous and synonymous mutations are shown as solid and dashed lines,
- respectively. Colors indicate the position of the site along the C2-V5 region
- (red to blue) MAYBE MAKE FIGURE WITH SYNONYMOUS AND NONSYN
- SEPARATELY. While nonsynonymous mutations frequently fix, very few synonymous
+\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_p8.pdf}
+\label{fig:aftsyn}}\\
+\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p8.pdf}
+\label{fig:aftnonsyn}}
+\caption{Time series of allele frequencies in \env, C2-V5, from
+patient 8~\cite{shankarappa_consistent_1999},
+divided by synonymous (panel A) and nonsynonymous (panel B) derived alleles.
+While nonsynonymous mutations frequently fix, very few synonymous
 mutations do even though they are frequently observed at intermediate
-frequencies.}
+frequencies. Colors indicate the position of the site along the C2-V5 region
+(blue to red). Inset: the fixation probability $P_\text{fix}(\nu)$ of a neutral mutation
+is simply the likelihood that the future common ancestor is currently carrying
+it, i.e. its frequency $\nu$.}
 \label{fig:aft}
@@
 C2-V5 region studied above and the remainder of \env~ reveal striking
-differences (see \FIG{fixp}). Within C2-V5, this data fully confirms the
+differences (see \FIG{fixp2}). Within C2-V5, this data fully confirms the
 observations made in the data set by \citet{shankarappa_consistent_1999} (red
 line). In the remainder of \env, however, observed synonymous mutations behave
-as if they were neutral (orange line).
-
-%ARE OBSERVED SYNONYMOUS MUTATIONS OUTSIDE C2-V5 NEUTRAL? (?? SOME!)
-%DOES LOSS/FIX CORRELATE WITH CONSERVATION? YES.
-%MAYBE WE COULD HAVE ONE -- COMPLETELY CIRCULAR -- FIGURE SHOWING LOSS/FIX VS CONSERVATION: SUPPLEMENTARY?
-%CAN WE LOOK AT THE AVERAGE LEVEL OF CONSERVATION STRATIFIED BY MAX FREQ? TRICKY: the maximal freq is achieved by hitchhiking...
+neutrally (orange line).
 
@@
 alleles (red line) are systematically larger than of alleles that are doomed to
-extinction (blue line) (Kolmogorov-Smirnov test, $P\approx
-2~\text{\textperthousand}$). In other words, alleles that are likely to be
+extinction (blue line) (Kolmogorov-Smirnov test, $P\approx 0.002$).
+In other words, alleles that are likely to be
 breaking RNA helices are also more likely to revert and finally be lost from the
@@
 In addition to RNA secondary structure, we have considered other possible
-explanations for a fitness effect of synonymous mutations, in particular codon
+explanations for a fitness defect of synonymous mutations, in particular codon
 usage bias (CUB). HIV is known to prefer A-rich codons over highly expressed
@@
 respectively~\citep{li_codon-usage-based_2012, ngumbela_quantitative_2008,
-coleman_virus_2008}. We do not find, however, evidence for any contribution of
+coleman_virus_2008}. We do not find, however, any evidence for a contribution of
 CUB to the ultimate fate of synonymous alleles. Several lines of thought support
-this result. First of all, although codon-optimized HIV seems to perform better
-{\it in vitro}, the distance in CUB between HIV and human genes is not shrinking
-at the macroevolutionary level. Second, within a single patient, we do not
-observe any bias towards more human-like CUB in the synonymous mutations that
-reach fixation rather than extinction. Third, it is a common phenomenon for
+this result. First of all, within a single patient, we do not observe any bias
+towards more human-like CUB in the synonymous mutations that reach fixation
+rather than extinction. Second, although codon-optimized HIV seems to perform
+better {\it in vitro}, the distance in CUB between HIV and human genes is not
+shrinking at the macroevolutionary level. Third, it is a common phenomenon for
 retroviruses to use variously different codons from their hosts, and CUB effects
@@
 deleterious mutations that are linked to adaptive variants can reach high
-frequency~\citep{smith_hitch-hiking_1974}.
+frequency~\citep[genetic hitchhiking,][]{smith_hitch-hiking_1974}.
 
@@
 
-To get a better idea of the range of parameters that are compatible with the
-observations and our interpretation, we perform computer simulations of
-evolving viral populations under selection and rare recombination. For this
-purpose, we use the recently published package FFPopSim, which includes a module
-dedicated to intrapatient HIV evolution~\citep{zanini_ffpopsim:_2012}. We
-analyze many combinations of parameters such as population size, recombination
-rate, selection coefficient and density of escape mutations, and deleterious effect
-of synonymous mutation.
-
-The main result of the simulations is that genetic draft can indeed bring weakly
-deleterious mutations to high frequencies and result in a dependence of the
-fixation probability on initial frequency that is compatible with observations.
-First of all, since neutral mutations are much more likely to rise to high
-frequency than deleterious ones, the majority of the synonymous mutations needs
-to be slightly deleterious observe a significant reduction of $P_\text{fix}$.
-In order to further quantify the reduction in fixation probability, we look at
-the difference between the neutral curve ($P_\text{fix}(\nu) = \nu$) and the
-measured fixation probability and calculate its area (see inset of
-\FIG{simfixpvar}). The minimal and maximal values for this area are zero
-(neutral-like curve) and 0.5 (no fixation at all), respectively. The HIV data
-correspond to an area under the diagonal of approximately 0.2 for synonymous
-changes, and a very small area over the diagonal for nonsynonymous changes.
-Various simulation curves are shown in \FIG{simfixpvar}. Then, in
-\FIG{simsfig}, we explore the parameter space: the combinations that
-yield areas close to the experimental result are shown.
-The two crucial parameters that control the fixation probability are the
-following: (a) the deleterious effects of hitchhikers compared to the beneficial
-effects of escape mutants, and (b) the density of escape mutations. Intuitively,
-a higher density of escape mutations (i.e., epitopes) enables a larger degree of
-genetic draft, because escape mutations from different epitopes start to combine
-and their effects add up. In \FIG{simheat1}, we show that this is indeed the
-case in simulations.
-
 \begin{figure}
@@
 \caption{(Panel A) The depression in $P_\text{fix}$ depends on the deleterious
-effect size of the synonymous alleles. This parameter also reduces synonymous
-diversity (first inset). The depression found in HIV data is about 0.2.
+effect size of the synonymous alleles (in HIV data is about $-0.2$).
+This parameter also reduces synonymous
+diversity, measured by the number of mutations at frequencies $0.25 < \nu < 0.75$,
+normalized by the number of possible mutations (first inset).
 (Panel B) To assess the parameter space that affects synonymous fixation and
@@
 
-However, if hitchhiking is driven by nonsynonymous mutations that are
-unconditionally beneficial, we should find that nonsynonymous mutations almost
-always fix once they reach high frequencies -- in contrast with \FIG{fixp} that
-shows that nonsynonymous mutations fix as if they were neutral. We know,
-however, that nonsynonymous variation in the variable regions is driven by
-positive selection. Inspecting the trajectories of nonsynonymous mutations
-suggest the rapid rise and fall of many alleles.  We test two possible
+To get a better idea of the range of parameters that are compatible with the
+observations and our interpretation, we perform computer simulations of
+evolving viral populations under selection and rare recombination. For this
+purpose, we use the recently published package FFPopSim, which includes a module
+dedicated to intrapatient HIV evolution~\citep{zanini_ffpopsim:_2012}. We
+pick many random combinations of the following four parameters: effect
+size and rate of appearance of beneficial escape mutations, and effect size
+and fraction of deleterious synonymous mutations.
+
+The first result of the simulations is that genetic draft can indeed bring weakly
+deleterious mutations to high frequencies and result in a dependence of the
+fixation probability on initial frequency that is compatible with observations
+(see \FIG{simfixpvar}).
+The fixation probability of synonymous alleles decreases from the neutral
+expectation to zero as their fitness effect worsens (\FIG{simfixpvar},
+from $-0.0003$ in blue to $-0.003$ in red). Contextually,
+the number of alleles that reach high frequencies, which is a proxy for
+synonymous diversity, is reduced as well (\FIG{simfixpvar}, top inset).
+
+To assess the reduction in fixation probability quantitatively, we calculathe
+the area between the measured fixation probability and the diagonal, which is
+the neutral expectation (\FIG{simfixpvar}, lower inset). This area spans a range
+between $-0.5$ (no fixation at all), zero (neutral-like curve), and $+0.5$ (always
+fixed). The HIV data correspond to an area of approximately $-0.2$ for synonymous
+changes, and essentially zero for nonsynonymous changes.
+
+Among all simulations, we select the one that show a large depression in
+fixation probability of synonymous mutations but, simulteneously, a high
+synonymous diversity, as observed in HIV data. This selection imposes a strong
+constraint on the parameters, in particular on the ones related to synonymous
+changes, i.e. their deleterious fraction and effect size (see \FIG{simsfig}).
+A high fraction ($\gtrsim 0.8$) of only slightly deleterious synonymous sites,
+with effect size $|s_d| \lesssim 0.002$, is required. This result fits well
+the expectation based on the fixation/extinction times above (see \FIG{fixp1}).
+Moreover, it corresponds to {\it a priori} theoretical expectations:
+(a) since high frequency bins are enriched in neutral mutations (it is easier for
+them to hitchhike), the hallmark of deleterious mutations will be visible there
+only if they are pervasive; (b) genetic draft stops working if the
+deleterious effect size becomes comparable to the adaptive effect size of escape
+mutations, because the double mutant has little or no fitness advantage.
+
+As long as only features of synonymous mutations are filtered, as performed
+above, no strong correlation of the two nonsynonymous parameters, effect size
+and rate of appearance of epitopes, is observed; in \FIG{simsfig}, the points
+shown possess a variety of shapes and colors. In most of these simulations, 
+however, nonsynonymous mutations almost always fix once they reach high frequencies
+-- their nonsynonymous fixation area is well above zero. This stands in contrast
+with the blue line in \FIG{fixp}: in an HIV infection, nonsynonymous mutations fix as if they
+were neutral. Although there are no doubts that nonsynonymous variation in the
+variable regions is driven by positive selection and not by genetic drift, the
+biological reason for the neutral-like curve in \FIG{fixp} is unclear.
+Inspecting the trajectories of nonsynonymous mutations
+suggests the rapid rise and fall of many alleles. We test two possible
 mechanisms that are biologically plausible and could explain the transient rise
 of nonsynonymous mutations: time-dependent selection and within-epitope
-competition. If the immune system recognizes the escape mutant before its
-fixation, the mutant might cease to be beneficial and disappear despite its
-quick initial rise in frequency.  In support of this idea,
+competition.
+
+The former explanation can be formulated as follows: if the immune system
+recognizes the escape mutant before its fixation, the mutant might cease to be
+beneficial and disappear soon, despite its quick initial rise in frequency. In support of this idea,
 \citet{richman_rapid_2003, bunnik_autologous_2008} report antibody responses to
 escape mutants. These respones are delayed by a few months, roughly matching the
-average sweep time of an escape mutant. Alternatively, several different escape
-mutations in the same epitope can arise almost simultaneously and start to
+average time needed by an escape mutant to cross frequencies of order one.
+In the alternative explanation, several different escape
+mutations within the same epitope might arise almost simultaneously and start to
 spread. Their fitness benefits are not additive, because each of them is
 essentially sufficient to escape. As a consequence, several escape mutations rise to
-high frequency, while the escape with the smallest cost in terms of replication,
-packaging, etc. is most likely to
-eventually fix. In simulations, this kind of epistatic interactions within
-epitopes reduces the fixation probability. The emergence of
+high frequency rapidly, while the one with the smallest cost in terms of replication,
+packaging, etc. is most likely to eventually fix. The emergence of
 multiple sweeping nonsynonymous mutations in real HIV infections has been shown
 previously~\citep{moore_limited_2009, bar_early_2012}.
-See the supplementary material for examples of successful simulations in both scenarios.
+We tested both hypotheses in our simulation framework and they both seem to
+achieve a reduced, neutral-like fixation probability while maintaining a high
+rate of substitutions, compatibly with HIV measurements. In real HIV infections,
+both mechanisms are likely to be playing a role.
 
@@
 that a region of length $l = (\rho t_{sw})^{-1} = s_a / \rho \log \nu_0$ remains
-linked to the adaptive mutation. With $s_a=0.01$, $l\approx 100$ bases which is
-consistent with strong linkage between the variable loops and the stems in
-between. Furthermore, we do not expect hitchhiking to extend far beyond
-the variable regions consistent with the lack of signal out side of C1-V5. In
-case of much stronger selection -- such as observed during early CTL escape or
-drug resistance evolution -- the linked  region is of course much larger.
+linked to the adaptive mutation. With $s_a=0.01$, $l\approx 100$ bases, hence we 
+expect strong linkage between the variable loops and their surrounding stems,
+but none far beyond the variable regions, consistent with the lack of signal
+outside of C1-V5. In case of much stronger selection -- such as observed during
+early CTL escape or drug resistance evolution -- the linked  region is of course
+much larger.
 
@@
 longitudinal data which requires a focus on the the variable regions of \env.
-Conserved RNA structures most likely exist in different parts
-of the HIV genome (several are known). In absence of repeated adaptive substitutions in the vicinity
-that cause hitchhiking, the deleterious synonymous mutations remain at low
-frequencies and can only be observed by deep sequencing methods.
+Conserved RNA structures exist in different parts of the HIV genome (several are
+known). In absence of repeated adaptive substitutions in the vicinity that cause
+hitchhiking, the deleterious synonymous mutations remain at low frequencies and
+can only be observed by deep sequencing methods.
 
@@
 sweeping sites, such as time-dependent selection and within-epitope negative
-epistasis. Although a final word about which mechanism is more
-widespread is yet to be spoken, both intuition and biological evidence from the
-literature support a mixed scenario~\citep{richman_rapid_2003,
+epistasis. With specific regard to these two scenarios, although a final word
+about which mechanism is more widespread is yet to be spoken, both intuition
+and biological evidence from the literature support a mixed scenario~\citep{richman_rapid_2003,
 moore_limited_2009, bar_early_2012}. Note also that, unlike influenza, HIV does

### 2013-01-23 synmut.tex

@@
 variable loops of the envelope protein gp120, which need to avoid recognition by
-neutralizing ABs.  Mutations in \env~, the gene encoding for gp120, spread
+neutralizing ABs.  Mutations in \env, the gene encoding for gp120, spread
 through the population within a few months (see \figurename~\ref{fig:aft}, solid
-lines).  The (Malthusian) effect size of these beneficial mutations is of the
-order of $s_a \sim 0.01$~\citep{neher_recombination_2010}.
+lines).
 
-Escape mutations are strongly selected for their effect on the amino acid
-sequence of the HIV proteins. Conversely, synonymous mutations are commonly
-used as approximate neutral markers in studies of viral evolution. Neutral
-markers are very useful in practice, because they can be used to make inferences
-about the stochastic forces driving evolution~\citep{yang_statistical_2000}.
-The viral genome, however, needs to satisfy further constraints in addition to
-immune escape, such as efficient processing and translation, nuclear export, and
-packaging into the viral capsid: all these processes operate at the RNA level
-and are sensitive to synonymous changes. A few functionally important RNA
+These escape mutations are strongly selected for their effect on the amino acid
+sequence of the viral proteins. Conversely, synonymous mutations are commonly
+used as approximately neutral markers in studies of viral evolution. Neutral
+markers are very useful in practice, because there evolution can be compared to
+that of putatively functional sites to detect purifying or directional selection
+\citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In most species,
+however, it is  known that synonymous codons don't evolve completely neutrally
+but that some codons are favored over others \citep{plotkin_synonymous_2011}.
+In addition to maintaining protein function and avoiding immune
+recognition, the HIV genome has to ensure efficient processing and translation,
+nuclear export, and packaging into the viral capsid: all these processes operate at the RNA
+level and are sensitive to synonymous changes. A few functionally important RNA
 elements are well characterized. For example, a certain RNA sequence in the HIV
@@
 markers, the degree to which they hitchhike with nearby nonsynonymous mutations
-is very informative; their ability to hitchhike for extended times itself is
-rooted in the small recombination rate of
+is very informative both about their on deleterious effect, as well as the
+strength and nature of selection for nearby positively selected sites.
+Their ability to hitchhike for extended times, which is a core requirement for
+our analysis, is rooted in the small recombination rate of
 HIV~\citep{neher_recombination_2010, batorsky_estimate_2011}. Extending the
@@
 direct observation of both the current frequency $\nu$ of any particular
-mutation and its future fate (fixation or extinction). They therefore represent
-a simple way to investigate average properties of different classes of
-mutations.
+mutation and its future fate (fixation or extinction).
 
-\subsection{Synonymous polymorphisms in \env, C2-V5 are mostly deleterious}
+\subsection{Synonymous polymorphisms in \env, C2-V5, are mostly deleterious}
 
@@
 usage bias (CUB). HIV is known to prefer A-rich codons over highly expressed
-human housekeeping genes~\citep{jenkins_extent_2003}. Moreover, codon-optimized
+human codons~\citep{jenkins_extent_2003}. Moreover, codon-optimized
 and -pessimized viruses have recently been generated and shown to replicate
@@
 
-As far as population genetics models are concerned, our study uncovers the
+Our study uncovers the
 subtle balance of evolutionary forces governing intrapatient HIV evolution. The
 fixation and extinction times and probabilities represent a rich and simple
-summary statistics to test sequencing data and computer simulation upon. A
-similar method has been recently used in a longitudinal study of
+summary statistics useful to characterize sequencing data and compare to
+models via computer simulations.
+A similar method has been recently used in a longitudinal study of
 influenza~\citep{strelkowa_clonal_2012}. The propagators suggested in that
-paper, however, represent ratios between (certain kinds of) nonsynonymous
+paper, however, represent ratios between nonsynonymous
 mutations and synonymous ones, hence they are inadequate to investigate
-synonymous changes themselves. Those authors also conclude that several
+synonymous changes themselves. The authors also conclude that several
 beneficial mutations segregate simultaneously in influenza, a scenario
-remarkably similar to our within-epitope competition picture. These results
-jointly suggest that viral evolution proceeds by multiple concurrent sweeps
-rather then by successive fixation~\citep{desai_beneficial_2007, neher_rate_2010}.
+remarkably similar to our within-epitope competition picture. While
+recombination hardly affects evolution within an epitope, it might facilitate
+immune escape by easing competition between different epitopes
+\citep{neher_rate_2010,Rouzine:2005p17398}.
 
-Finally, our results emphasize the inadequacy of independent site
-models of HIV evolution, especially in the light of transient effects on
-sweeping sites, such as time-dependent selection and within-epitope negative
-epistasis. With specific regard to these two scenarios, although a final word
-about which mechanism is more widespread is yet to be spoken, both intuition
-and biological evidence from the literature support a mixed scenario~\citep{richman_rapid_2003,
-moore_limited_2009, bar_early_2012}. Note also that, unlike influenza, HIV does
-recombine if rarely, hence clonal interference as studied in
-ref.~\citep{strelkowa_clonal_2012} is only a short-term effect.
+Our results emphasize the inadequacy of independent site
+models of HIV evolution and the common assumption that selection is time
+independent. If genetic variation is only transiently beneficial, existing
+methods to quantify selection will yield substantial underestimates
+\citep{williamson_adaptation_2003,neher_rate_2010,OTHER}. To explain the
+observations regarding the fixation probabilities of non-synonymous mutations,
+either transient selection, or substational within-epitope competition are
+necessary. Which mechanism is more widespread is not clear as of now,
+there is evidence for both~\citep{richman_rapid_2003, moore_limited_2009,
+bar_early_2012}.
 

### 2013-01-24 synmut.tex
figures adapted

@@
 spread through the population and fix; in the rest of the cases, i.e. with
-probability $1-\nu$, it goes extinct. As illustrated in the inset of \FIG{fixp},
+probability $1-\nu$, it goes extinct. As illustrated in the inset of \FIG{aftsyn},
 this is a simple consequence of the fact that
@@
 line). In the remainder of \env, however, observed synonymous mutations behave
-neutrally (orange line).
+neutrally (green line).
 

### 2013-01-25 synmut.tex
Added bootstrap on Pfix

@@
 \label{fig:fixp2}}
-\caption{Left panel: time course of loss and fixation of synonymous mutations
+\caption{(Panel A) Time course of loss and fixation of synonymous mutations
 observed in a frequency interval $\nu_0$. The ultimate fraction of synonymous
 mutations that fix as a function of intermediate frequency $\nu_0$ is the
- fixation probability.  Right panel: fixation probability of derived synonymous
+fixation probability. (Panel B) Fixation probability of derived synonymous
 alleles is strongly suppressed in C2-V5 versus other parts of the {\it env}
-gene, and of nonsynonymous ones. Data from
+gene. The horizontal error bars on the abscissa are bin sizes, the vertical ones
+the standard deviation after 100 patient bootstraps of the data. Data from
 Refs.~\cite{shankarappa_consistent_1999, bunnik_autologous_2008}.}

### 2013-01-28 synmut.tex
new figure from simulations, broader

@@
 synonymous diversity, is reduced as well (\FIG{simfixpvar}, top inset).
-
 To assess the reduction in fixation probability quantitatively, we calculathe
@@
 Among all simulations, we select the one that show a large depression in
-fixation probability of synonymous mutations but, simulteneously, a high
-synonymous diversity, as observed in HIV data. This selection imposes a strong
+fixation probability of synonymous mutations but, simulteneously, a moderately
+high synonymous diversity, quantified by the probability $P_\text{interm}$ of a
+derived allele to be at intermediate frequencies $0.25 < \nu < 0.75$. HIV data show
+$A_\text{syn} \sim -0.2$ and $P_\text{interm} \sim 0.005$; simulations are gated
+such that $A_\text{syn} < -0.15$ and $0.0025 < P_\text{interm} < 0.010$, hence
+allowing an uncertainty on mutation rates or generation times of about twofold.
+This selection imposes a strong
 constraint on the parameters, in particular on the ones related to synonymous
@@
 A high fraction ($\gtrsim 0.8$) of only slightly deleterious synonymous sites,
-with effect size $|s_d| \lesssim 0.002$, is required. This result fits well
+with effect size $|s_d| \sim 0.002$, is required. This result fits well
 the expectation based on the fixation/extinction times above (see \FIG{fixp1}).
@@
 deleterious effect size becomes comparable to the adaptive effect size of escape
-mutations, because the double mutant has little or no fitness advantage.
+mutations, because the double mutant has little or no fitness advantage; (c)
+mutations with a deleterious effect weaker than approximately $0.001$ are
+indistinguishable from neutral ones and are drafted seamlessly to any frequency.
 

### 2013-01-28 synmut.tex
text revisited around new figure

@@
 \caption{(Panel A) The depression in $P_\text{fix}$ depends on the deleterious
-effect size of the synonymous alleles (in HIV data is about $-0.2$).
-This parameter also reduces synonymous
-diversity, measured by the number of mutations at frequencies $0.25 < \nu < 0.75$,
-normalized by the number of possible mutations (first inset).
+effect size of the synonymous alleles. This parameter also reduces synonymous
+diversity, measured by probability of a derived allele to be found at
+intermediate frequencies $P_\text{interm}$ (first inset).
 (Panel B) To assess the parameter space that affects synonymous fixation and
@@
 (see \FIG{simfixpvar}).
-The fixation probability of synonymous alleles decreases from the neutral
-expectation to zero as their fitness effect worsens (\FIG{simfixpvar},
-from $-0.0003$ in blue to $-0.003$ in red). Contextually,
-the number of alleles that reach high frequencies, which is a proxy for
-synonymous diversity, is reduced as well (\FIG{simfixpvar}, top inset).
-To assess the reduction in fixation probability quantitatively, we calculathe
+To assess the reduction in fixation probability quantitatively, we calculate
 the area between the measured fixation probability and the diagonal, which is
@@
 between $-0.5$ (no fixation at all), zero (neutral-like curve), and $+0.5$ (always
-fixed). The HIV data correspond to an area of approximately $-0.2$ for synonymous
-changes, and essentially zero for nonsynonymous changes.
+fixed). Furthermore, we quantify synonymous diversity via the probability
+$P_\text{interm}$ of a synonymous derived allele to be at intermediate frequencies
+$0.25 < \nu < 0.75$ (\FIG{simfixpvar}, top inset).
+The HIV data correspond to an area $A_\text{syn} \sim -0.2$  for synonymous
+changes and $A_\text{nonsyn} \sim 0$ for nonsynonymous changes,
+and a synonymous diversity $P_\text{interm} \sim 0.005$.
+In the three simulations shown, the fixation probability of synonymous alleles
+decreases from the neutral expectation ($A_\text{syn} \sim 0$) to zero
+($A_\text{syn} \sim -0.5$) as their fitness effect
+worsens; the synonymous diversity plummets as well, as it is harder for
+deleterious mutations to reach the frequency threshold of 0.25
+(\FIG{simfixpvar}, from $-0.0003$ in blue to $-0.003$ in red).
 
-Among all simulations, we select the one that show a large depression in
+Among all simulations, we select the ones that show HIV-like $A_\text{syn}$ and
+$P_\text{interm}$, i.e. a large depression in
 fixation probability of synonymous mutations but, simulteneously, a moderately
-high synonymous diversity, quantified by the probability $P_\text{interm}$ of a
-derived allele to be at intermediate frequencies $0.25 < \nu < 0.75$. HIV data show
-$A_\text{syn} \sim -0.2$ and $P_\text{interm} \sim 0.005$; simulations are gated
-such that $A_\text{syn} < -0.15$ and $0.0025 < P_\text{interm} < 0.010$, hence
-allowing an uncertainty on mutation rates or generation times of about twofold.
-This selection imposes a strong
+high synonymous diversity: simulations are gated such that $A_\text{syn} < -0.15$
+and $0.0025 < P_\text{interm} < 0.010$, hence allowing an uncertainty on mutation
+rates or generation times of about twofold. This selection imposes a strong
 constraint on the parameters, in particular on the ones related to synonymous

### 2013-01-28 synmut.tex

--- 2013-01-25 synmut.tex
@@
 through the population within a few months (see \figurename~\ref{fig:aft}, solid
-lines).
+lines). Consistent with this time scale, it is found that serum from a
+particular time typically neutralizes virus extracted more than 3-6 month
+earlier \citep{richman_rapid_2003}.
 
@@
 a skewed codon usage bias towards more or less abundant tRNAs
-replicate better or worse, respectively~\citep{ngumbela_quantitative_2008,
-li_codon-usage-based_2012}. A similar conclusion has been reached about
+replicate better or worse \comment{replicate better or produce more protein?},
+respectively~\citep{ngumbela_quantitative_2008, li_codon-usage-based_2012}. A similar conclusion has been reached about
 influenza, and codon-pessimized influenza strains have been shown to be good
 live attenuated vaccines in mice~\citep{mueller_live_2010}. Purifying selection
-beyond the protein sequence is therefore expected, but it seems reasonable
-that the bulk of positive selection through the immune system be restricted to
-amino acid sequences.
+beyond the protein sequence is therefore expected
+\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}, but it seems reasonable
+that the bulk of positive selection through the immune system be restricted to amino acid sequences.
 
@@
 segregating at frequency $\nu$ has a probability $P_\text{fix}(\nu) = \nu$ to
-spread through the population and fix; in the rest of the cases, i.e. with
+spread through the population and fix; in the rest of the cases, i.e.~with
 probability $1-\nu$, it goes extinct. As illustrated in the inset of \FIG{aftsyn},
@@
 the common ancestor of the entire future population at a particular locus and
-(b) this ancestor has a probability $\nu$ of carrying the mutation.
+(b) this ancestor has a probability $\nu$ of carrying the mutation (assuming
+the neutral mutation is not preferentially associated with genomes of high or
+low fitness).
 Deleterious or beneficial mutations fix less or
@@
 synonymous mutations reaching high frequency, very few fix
-(panel~\ref{fig:aftsyn}); however, quite some nonsynonymous mutations reach
-fixation (panel~\ref{fig:aftnonsyn}).
+(panel~\ref{fig:aftsyn}); however, many nonsynonymous mutations fix
+(panel~\ref{fig:aftnonsyn}).
 These observations are quantified in \FIG{fixp}, which stratifies the data
-of 7 (resp. 10) patients according to the frequency at which
+of 7 (resp.~10) patients according to the frequency at which
 different mutations are observed (see methods). Considering all mutations in a
@@
 the time interval $t_f-t_i$, we see that most synonymous mutations segregate for
-roughly two years and are lost much more frequently than expected (panel
+roughly one year and are lost much more frequently than expected (panel
 \ref{fig:fixp1}). The long-time probability of fixation versus extinction of
@@
 
-These observations suggest that many of the synonymous polymorphisms in the part
-of \env~that includes the hypervariable regions are deleterious, while outside
-this regions polymorphisms are mostly roughly neutral. Note that this does not imply that all 
-synonymous mutations are neutral -- only those mutations observed at high frequencies tend 
-to be neutral.
+These observations suggest that many of the synonymous polymorphisms at
+intermediate frequencies in the part of \env~that includes the hypervariable
+regions are deleterious, while outside this regions polymorphisms are mostly
+roughly neutral. Note that this does not imply that all synonymous mutations are
+neutral -- only those mutations observed at high frequencies tend to be neutral.
 
@@
 One possible {\it a priori} explanation for lack of fixation of synonymous
-mutations in C2-V5 are  secondary structures in the viral RNA. If any RNA
-secondary structures are relevant for HIV replication, mutations in nucleotides
-involved in those base pairs are expected to be deleterious and to revert
-preferentially. Many functionally important secondary structure elements have
-been characterized, including  the RRE~\citep{fernandes_hiv-1_2012} and the 5'
-UTR pseudoknot interacting with the host
-tRNA$^\text{Lys3}$~\citep{barat_interaction_1991, paillart_vitro_2002}. It has
+mutations in C2-V5 are  secondary structures in the viral RNA. It has
 been suggested early on that parts of the viral genome that has the potential to
 form stems is better conserved than the
-remainder~\citep{forsdyke_reciprocal_1995}.
+remainder~\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}.
 
@@
 exclude an effect of CUB on fitness as a general rule, we expect it to be a
-minor effect in our context.
+minor effect in our context. \comment{Need to read literature}
 
@@
 \subfloat{\includegraphics[width=0.9\linewidth]{Shankarappa_fixmid_syn_V_regions.pdf}\label{fig:SHAPEB}}
-\caption{Watts et al. have measured the reactivity of HIV nucleotides to {\it
-in vitro} chemical attack and shown that some nucleotides are more likely to
-be involved in RNA secondary folds. C1-C5 regions, in particular, show
-conserved stem-loop structures~\citep{watts_architecture_2009}. We show that
-among all derived alleles in those regions reaching frequencies of order one,
-there is a negative correlation between fixation and involvement in a base
-pairing in a RNA stem (left panel). The rest of the genome does not show any
-correlation (right panel). There might be too few silent polymorphisms in the
-first place, or the signal might be masked by non-functional RNA
-structures. Data from Refs.~\cite{shankarappa_consistent_1999,
+\caption{Permissible synonymous mutations have
+high SHAPE reactivities. \comment{Is this C2-V5 only?} 
+Panel A) Synonymous substitions in the C2-V5 region have substantially higher
+SHAPE reactivities than mutations that reach frequencies above 15\%, but are
+susequently lost. This difference is quantified by a cumulative histogram
+($p=0.002$, Kolmogorov-Smirnov two sample test). Mutations that are never
+observered show an intermediate distribution of SHAPE values.
+Panel B shows the fixation probablility of synonymous mutations in C2-V5
+separately for variable loops and the connecting conserved regions, that
+harbor conserved RNA stems. As expected, fixation probability is lower inside
+the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
 bunnik_autologous_2008, liu_selection_2006}.}
@@
 that it undergoes frequent adaptive changes to evade recognition by neutralizing
-antibodies~\cite{williamson_adaptation_2003}. Due to the limited amount of
+antibodies~\cite{williamson_adaptation_2003,richman_rapid_2003}. Due to the limited amount of
 recombination in HIV~\cite{neher_recombination_2010,batorsky_estimate_2011},
@@
 trajectories in \FIG{aft}, where many mutations appear to change rapidly in
-frequency as a flock. Deleterious synonymous mutations can be amplified
-exponentially by selection on linked nonsynonymous sites, a process known as
-{\it genetic draft}~\citep{gillespie_genetic_2000, neher_genetic_2011}. In order
+frequency as a flock. Deleterious synonymous mutations can be brought to high
+frequency by selection on linked nonsynonymous sites, a process
+known as {\it genetic draft}~\citep{gillespie_genetic_2000, neher_genetic_2011}. In order
 to be advected to high frequency by a linked adaptive mutation, the deleterious
@@
 \label{fig:simsfig}}
-\caption{(Panel A) The depression in $P_\text{fix}$ depends on the deleterious
+\caption{Evolutionary parameters compatible with observations.
+Panel A) The depression in $P_\text{fix}$ depends on the deleterious
 effect size of the synonymous alleles (in HIV data is about $-0.2$).
-This parameter also reduces synonymous
+Increasing this parameter also reduces synonymous
 diversity, measured by the number of mutations at frequencies $0.25 < \nu < 0.75$,
 normalized by the number of possible mutations (first inset).
-(Panel B) To assess the parameter space that affects synonymous fixation and
-diversity, we run many simulations with random parameters for deleterious effect
-size, fraction of deleterious synonymous sites, average size of escape
-mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
+Panel B) To assess the parameter space compatible with observations, we run many
+simulations with random parameters for deleterious effect size, fraction of
+deleterious synonymous sites, average size of escape mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
 introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
@@
 high deleterious fractions of synonymous mutations reproduce synonymous HIV-like
-fixation probability and diversity.}
+fixation probability and diversity. Parameters are chosen uniformly on a
+logarithmic scale from the indicated windows. \comment{should we extend the
+parameters space maybe to $10^{-4}$?}}
 \label{fig:simheat}
@@
 observations and our interpretation, we perform computer simulations of
-evolving viral populations under selection and rare recombination. For this
-purpose, we use the recently published package FFPopSim, which includes a module
-dedicated to intrapatient HIV evolution~\citep{zanini_ffpopsim:_2012}. We
-pick many random combinations of the following four parameters: effect
-size and rate of appearance of beneficial escape mutations, and effect size
-and fraction of deleterious synonymous mutations.
+evolving viral populations assuming a mix of positive and purifying selection
+and rare recombination.
+For this purpose, we use the simulation package FFPopSim, which includes a
+module dedicated to intrapatient HIV evolution~\citep{zanini_ffpopsim:_2012}.
 
 The first result of the simulations is that genetic draft can indeed bring weakly
-deleterious mutations to high frequencies and result in a dependence of the
+deleterious mutations to high frequencies and results in a dependence of the
 fixation probability on initial frequency that is compatible with observations
@@
 The fixation probability of synonymous alleles decreases from the neutral
-expectation to zero as their fitness effect worsens (\FIG{simfixpvar},
-from $-0.0003$ in blue to $-0.003$ in red). Contextually,
-the number of alleles that reach high frequencies, which is a proxy for
-synonymous diversity, is reduced as well (\FIG{simfixpvar}, top inset).
+expectation to zero as their deleterious effect increases
+(\FIG{simfixpvar}, from $-0.0003$ in blue to $-0.003$ in red). Along with the
+reduction in fixation probability, the number of alleles that reach high
+frequencies, which is a proxy for synonymous diversity, is reduced as well (\FIG{simfixpvar}, top inset).
 
-To assess the reduction in fixation probability quantitatively, we calculathe
+To assess the reduction in fixation probability quantitatively, we calculate
 the area between the measured fixation probability and the diagonal, which is
@@
 
+We pick many random combinations of the following four parameters: effect size
+and rate of appearance of beneficial escape mutations, and effect size and
+fraction of deleterious synonymous mutations.
 Among all simulations, we select the one that show a large depression in
-fixation probability of synonymous mutations but, simulteneously, a high
-synonymous diversity, as observed in HIV data. This selection imposes a strong
-constraint on the parameters, in particular on the ones related to synonymous
-changes, i.e. their deleterious fraction and effect size (see \FIG{simsfig}).
+fixation probability (>-0.??) of synonymous mutations but, simulteneously, a
+high synonymous diversity (>0.0? per site), as observed in HIV data. This
+selection imposes a strong constraint on the parameters, in particular on the
+ones related to synonymous changes, i.e.~their deleterious fraction and effect
+size (see \FIG{simsfig}).
 A high fraction ($\gtrsim 0.8$) of only slightly deleterious synonymous sites,
-with effect size $|s_d| \lesssim 0.002$, is required. This result fits well
-the expectation based on the fixation/extinction times above (see \FIG{fixp1}).
+with effect size $|s_d| \lesssim 0.002$, is required. This result fits well the
+expectation based on the fixation/extinction times above (see \FIG{fixp1}).
 Moreover, it corresponds to {\it a priori} theoretical expectations:
-(a) since high frequency bins are enriched in neutral mutations (it is easier for
-them to hitchhike), the hallmark of deleterious mutations will be visible there
-only if they are pervasive; (b) genetic draft stops working if the
+(a) since high frequency bins are enriched in neutral mutations (it is easier
+for them to hitchhike), the hallmark of deleterious mutations will be visible
+there only if they are pervasive; (b) genetic draft stops working if the
 deleterious effect size becomes comparable to the adaptive effect size of escape
@@
 however, nonsynonymous mutations almost always fix once they reach high frequencies
--- their nonsynonymous fixation area is well above zero. This stands in contrast
+-- their nonsynonymous fixation area is well above zero. This is incompatible
 with the blue line in \FIG{fixp}: in an HIV infection, nonsynonymous mutations fix as if they
@@
 mutations within the same epitope might arise almost simultaneously and start to
-spread. Their fitness benefits are not additive, because each of them is
+spread. Their benefits are not additive, because each of them is
 essentially sufficient to escape. As a consequence, several escape mutations rise to
 high frequency rapidly, while the one with the smallest cost in terms of replication,
-packaging, etc. is most likely to eventually fix. The emergence of
+packaging, etc.~is most likely to eventually fix. The emergence of
 multiple sweeping nonsynonymous mutations in real HIV infections has been shown

### 2013-01-28 synmut.tex

@@
 \label{fig:simsfig}}
-\caption{Evolutionary parameters compatible with observations.
-Panel A) The depression in $P_\text{fix}$ depends on the deleterious
-effect size of the synonymous alleles (in HIV data is about $-0.2$).
-Increasing this parameter also reduces synonymous
-diversity, measured by the number of mutations at frequencies $0.25 < \nu < 0.75$,
-normalized by the number of possible mutations (first inset).
-Panel B) To assess the parameter space compatible with observations, we run many
-simulations with random parameters for deleterious effect size, fraction of
-deleterious synonymous sites, average size of escape mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
+\caption{(Panel A) The depression in $P_\text{fix}$ depends on the deleterious
+effect size of the synonymous alleles. This parameter also reduces synonymous
+diversity, measured by probability of a derived allele to be found at
+intermediate frequencies $P_\text{interm}$ (first inset).
+(Panel B) To assess the parameter space that affects synonymous fixation and
+diversity, we run many simulations with random parameters for deleterious effect
+size, fraction of deleterious synonymous sites, average size of escape
+mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
 introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
@@
 (see \FIG{simfixpvar}).
-The fixation probability of synonymous alleles decreases from the neutral
-expectation to zero as their deleterious effect increases
-(\FIG{simfixpvar}, from $-0.0003$ in blue to $-0.003$ in red). Along with the
-reduction in fixation probability, the number of alleles that reach high
-frequencies, which is a proxy for synonymous diversity, is reduced as well (\FIG{simfixpvar}, top inset).
-
 To assess the reduction in fixation probability quantitatively, we calculate
@@
 between $-0.5$ (no fixation at all), zero (neutral-like curve), and $+0.5$ (always
-fixed). The HIV data correspond to an area of approximately $-0.2$ for synonymous
-changes, and essentially zero for nonsynonymous changes.
+fixed). Furthermore, we quantify synonymous diversity via the probability
+$P_\text{interm}$ of a synonymous derived allele to be at intermediate frequencies
+$0.25 < \nu < 0.75$ (\FIG{simfixpvar}, top inset).
+The HIV data correspond to an area $A_\text{syn} \sim -0.2$  for synonymous
+changes and $A_\text{nonsyn} \sim 0$ for nonsynonymous changes,
+and a synonymous diversity $P_\text{interm} \sim 0.005$.
+In the three simulations shown, the fixation probability of synonymous alleles
+decreases from the neutral expectation ($A_\text{syn} \sim 0$) to zero
+($A_\text{syn} \sim -0.5$) as their fitness effect
+worsens; the synonymous diversity plummets as well, as it is harder for
+deleterious mutations to reach the frequency threshold of 0.25
+(\FIG{simfixpvar}, from $-0.0003$ in blue to $-0.003$ in red).
 
-We pick many random combinations of the following four parameters: effect size
-and rate of appearance of beneficial escape mutations, and effect size and
-fraction of deleterious synonymous mutations.
-Among all simulations, we select the one that show a large depression in
-fixation probability (>-0.??) of synonymous mutations but, simulteneously, a
-high synonymous diversity (>0.0? per site), as observed in HIV data. This
-selection imposes a strong constraint on the parameters, in particular on the
-ones related to synonymous changes, i.e.~their deleterious fraction and effect
-size (see \FIG{simsfig}).
+Among all simulations, we select the ones that show HIV-like $A_\text{syn}$ and
+$P_\text{interm}$, i.e. a large depression in
+fixation probability of synonymous mutations but, simulteneously, a moderately
+high synonymous diversity: simulations are gated such that $A_\text{syn} < -0.15$
+and $0.0025 < P_\text{interm} < 0.010$, hence allowing an uncertainty on mutation
+rates or generation times of about twofold. This selection imposes a strong
+constraint on the parameters, in particular on the ones related to synonymous
+changes, i.e. their deleterious fraction and effect size (see \FIG{simsfig}).
 A high fraction ($\gtrsim 0.8$) of only slightly deleterious synonymous sites,
-with effect size $|s_d| \lesssim 0.002$, is required. This result fits well the
-expectation based on the fixation/extinction times above (see \FIG{fixp1}).
+with effect size $|s_d| \sim 0.002$, is required. This result fits well
+the expectation based on the fixation/extinction times above (see \FIG{fixp1}).
 Moreover, it corresponds to {\it a priori} theoretical expectations:
-(a) since high frequency bins are enriched in neutral mutations (it is easier
-for them to hitchhike), the hallmark of deleterious mutations will be visible
-there only if they are pervasive; (b) genetic draft stops working if the
+(a) since high frequency bins are enriched in neutral mutations (it is easier for
+them to hitchhike), the hallmark of deleterious mutations will be visible there
+only if they are pervasive; (b) genetic draft stops working if the
 deleterious effect size becomes comparable to the adaptive effect size of escape
-mutations, because the double mutant has little or no fitness advantage.
+mutations, because the double mutant has little or no fitness advantage; (c)
+mutations with a deleterious effect weaker than approximately $0.001$ are
+indistinguishable from neutral ones and are drafted seamlessly to any frequency.
 

### 2013-01-28 synmut.tex
histograms for SHAPE instead of cumulative
combed through the text. Discussion and captions still need work

@@
 \newcommand{\mexpfit}{\langle e^{F}\rangle}
+\newcommand{\pfix}{P_{fix}}
 \newcommand{\ox}{r}
@@
 (ABs)~\citep{rambaut_causes_2004} and facilitated by the high
-mutation rate of HIV~\citep{mansky_lower_1995}. When the host develops a CTL or
-AB response against a particular HIV epitope, mutations in the viral genome that
+mutation rate of HIV~\citep{mansky_lower_1995,abram_nature_2010}. When the host
+develops a CTL or AB response against a particular HIV epitope, mutations in the viral genome that
 reduce or prevent recognition of the epitope frequently emerge. Escape mutations
@@
 used as approximately neutral markers in studies of viral evolution. Neutral
-markers are very useful in practice, because there evolution can be compared to
+markers are very useful since their dynamics can be compared to
 that of putatively functional sites to detect purifying or directional selection
@@
 but that some codons are favored over others \citep{plotkin_synonymous_2011}.
-In addition to maintaining protein function and avoiding immune
+This is expected to be particularly important in compact viruses like HIV. 
+In addition to maintaining protein function and avoiding the adaptive immune
 recognition, the HIV genome has to ensure efficient processing and translation,
@@
 elements are well characterized. For example, a certain RNA sequence in the HIV
-genome, called \rev{} response element (RRE), enhances nuclear export of viral
-transcripts~\citep{fernandes_hiv-1_2012}. Another well studied case is the
-interaction between viral reverse transcriptase, viral ssRNA, and the host
+genome, called \rev{} response element (RRE), enhances nuclear export of
+full length or partially spliced viral transcripts~\citep{fernandes_hiv-1_2012}.
+Another well studied case is the interaction between viral reverse transcriptase, viral ssRNA, and the host
 tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription
-(RT) and bound by a specifical pseudoknotted RNA structure in the viral 5'
+(RT) and is bound by a pseudoknotted RNA structure in the viral 5'
 untranslated region~\citep{barat_interaction_1991, paillart_vitro_2002}.
@@
 well. Recent studies have shown that genetically engineered HIV strains with
-a skewed codon usage bias towards more or less abundant tRNAs
-replicate better or worse \comment{replicate better or produce more protein?},
-respectively~\citep{ngumbela_quantitative_2008, li_codon-usage-based_2012}. A similar conclusion has been reached about
-influenza, and codon-pessimized influenza strains have been shown to be good
-live attenuated vaccines in mice~\citep{mueller_live_2010}. Purifying selection
-beyond the protein sequence is therefore expected
-\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}, but it seems reasonable
-that the bulk of positive selection through the immune system be restricted to amino acid sequences.
+altered codon usage can in some cases produce more viral protein, but in general
+replicate less efficiently~\citep{ngumbela_quantitative_2008,
+li_codon-usage-based_2012,keating_rich_2009}.
+Codon-deoptimization has been suggested as attenuation strategy for polio and 
+influenza~\citep{mueller_live_2010,coleman_virus_2008}. Purifying
+selection beyond the protein sequence is therefore expected
+\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}, but while positive
+selection through the immune system be mainly restricted to changes in the
+amino acid sequence.
 
@@
 In this paper, we characterize the dynamics of synonymous mutations in \env{}
-and show that a substantial fraction of these mutations is deleterious. We
-further show that, although such synonymous mutations cannot be used as neutral
-markers, the degree to which they hitchhike with nearby nonsynonymous mutations
-is very informative both about their on deleterious effect, as well as the
-strength and nature of selection for nearby positively selected sites.
-Their ability to hitchhike for extended times, which is a core requirement for
-our analysis, is rooted in the small recombination rate of
-HIV~\citep{neher_recombination_2010, batorsky_estimate_2011}. Extending the
-analysis of fixation probabilities to the nonsynonymous mutations, we show that
-time dependent selection or strong competition of escape mutations inside the
-same epitope are necessary to explain the observed patterns of fixation and
-loss.
+and show that a substantial fraction of these mutations is deleterious. We then
+compare our observations to computational models of HIV evolution and derive
+estimates for the effect synonymous mutations have on fitness.
+Hitchhiking of synonymous mutations with escape mutations is due to the small
+recombination rate of HIV~\citep{neher_recombination_2010,
+batorsky_estimate_2011}. Extending the analysis of fixation probabilities to the
+nonsynonymous mutations, we show that time dependent selection or strong
+competition of escape mutations inside the same epitope are necessary to explain
+the observed patterns of fixation and loss.
 
@@
 Deleterious or beneficial mutations fix less or
-more often than neutral ones, respectively. Time series sequence data enable a
-direct observation of both the current frequency $\nu$ of any particular
-mutation and its future fate (fixation or extinction).
+more often than neutral ones, respectively. \FIG{aft}A\&B show 
+the time course the frequencies of all synonymous and non-synonymous mutations
+observed \env~, C2-V5, in patient p8~\citep{shankarappa_consistent_1999},
+respectively. Despite many synonymous mutations reaching high frequency, very
+few fix (panel~\ref{fig:aftsyn}); however, many nonsynonymous mutations fix
+(panel~\ref{fig:aftnonsyn}).
 
 \subsection{Synonymous polymorphisms in \env, C2-V5, are mostly deleterious}
-
-\FIG{aft} shows time series data of the frequencies of all mutations observed
-\env~, C2-V5, in patient p8~\citep{shankarappa_consistent_1999}. Despite many
-synonymous mutations reaching high frequency, very few fix
-(panel~\ref{fig:aftsyn}); however, many nonsynonymous mutations fix
-(panel~\ref{fig:aftnonsyn}).
-These observations are quantified in \FIG{fixp}, which stratifies the data
-of 7 (resp.~10) patients according to the frequency at which
-different mutations are observed (see methods). Considering all mutations in a
+We study the dynamics and fate of synonymous mutations more quantitatively by
+analyzing data from 7 patients from
+\citet{shankarappa_consistent_1999,liu_selection_2006} and 3 patients from
+\citet{bunnik_autologous_2008} (see methods). The data set from
+\citet{shankarappa_consistent_1999,liu_selection_2006} is restricted to the
+C2-V5 region of \env, while \citet{bunnik_autologous_2008} covers the
+majority of \env. Considering all mutations in a
 frequency interval around $\nu_0$ at some time $t_i$, we calculate the fraction
@@
 \ref{fig:fixp1}). The long-time probability of fixation versus extinction of
-synonymous mutations is shown as a function of the initial frequency $\nu_0$
-in panel~\ref{fig:fixp2} (red line). The nonsynonymous seem to follow more or
-less the neutral expectation (blue line) -- a point to which we will come back
-below.
+synonymous mutations is shown as a function of the initial frequency $\nu_0$ in
+panel~\ref{fig:fixp2} (red line). Restricted to the region C2-V5, we find that
+$\pfix$ of synonymous mutations is far below the neutral expectation.
+Outside of C2-V5 using data from \citet{bunnik_autologous_2008} only, no such
+reduction in $\pfix$ is found. Restricted to the C2-V5 area, the data from
+\citet{bunnik_autologous_2008} is fully compatible with data from
+\citet{shankarappa_consistent_1999}. The nonsynonymous seem to follow more or
+less the neutral expectation (blue line) -- a point to which we will come back below.
 
@@
 
-\citet{bunnik_autologous_2008} present a longitudinal dataset on the entire
-\env~gene of 3 patients at $\sim 5$ time points with approximately 5-20
-sequences each (see methods). Repeating the above analysis separately on the
-C2-V5 region studied above and the remainder of \env~ reveal striking
-differences (see \FIG{fixp2}). Within C2-V5, this data fully confirms the
-observations made in the data set by \citet{shankarappa_consistent_1999} (red
-line). In the remainder of \env, however, observed synonymous mutations behave
-neutrally (green line).
-
-These observations suggest that many of the synonymous polymorphisms at
-intermediate frequencies in the part of \env~that includes the hypervariable
-regions are deleterious, while outside this regions polymorphisms are mostly
-roughly neutral. Note that this does not imply that all synonymous mutations are
-neutral -- only those mutations observed at high frequencies tend to be neutral.
+When interpreting these results for the fixation probabilities, it is important
+to distinguish random mutations, and polymorphisms at a given frequency that has
+already been filtered by selection.
+A polymorphism could be beneficial to the virus and on its way to fixation. In
+this case, we expect that it fixes almost surely given we see it at high
+frequency. If, on the other hand, the polymorphism is deleterious it must have
+gotten to high frequency by a chance event (genetic drift or hitch-hiking), and
+we expect that selection will drive it out of the population again. Hence our
+observations suggest that many of the synonymous polymorphisms at intermediate
+frequencies in the part of \env~that includes the hypervariable regions are
+deleterious, while outside this regions polymorphisms are mostly roughly
+neutral. Note that this does not imply that all synonymous mutations are neutral
+-- only those mutations observed at high frequencies tend to be neutral.
 
@@
 usage bias (CUB). HIV is known to prefer A-rich codons over highly expressed
-human codons~\citep{jenkins_extent_2003}. Moreover, codon-optimized
-and -pessimized viruses have recently been generated and shown to replicate
-better or worse than wild type strains,
-respectively~\citep{li_codon-usage-based_2012, ngumbela_quantitative_2008,
-coleman_virus_2008}. We do not find, however, any evidence for a contribution of
-CUB to the ultimate fate of synonymous alleles. Several lines of thought support
-this result. First of all, within a single patient, we do not observe any bias
-towards more human-like CUB in the synonymous mutations that reach fixation
-rather than extinction. Second, although codon-optimized HIV seems to perform
-better {\it in vitro}, the distance in CUB between HIV and human genes is not
-shrinking at the macroevolutionary level. Third, it is a common phenomenon for
-retroviruses to use variously different codons from their hosts, and CUB effects
-on fitness are thought to be so small that divergent nucleotide composition has
-been suggested as a possible mechanism for viral
-speciation~\citep{bronson_nucleotide_1994}. Fourth, CUB in the V1-C5 region is
-not very different from other parts of the HIV genome, whereas the reduced
-fixation probability is only observed there. In conclusion, although we cannot
-exclude an effect of CUB on fitness as a general rule, we expect it to be a
-minor effect in our context. \comment{Need to read literature}
+human codons~\citep{jenkins_extent_2003,kuyl_biased_2012}. We do not find,
+however, any evidence for a contribution of CUB to the ultimate fate of
+synonymous alleles consistent with the fact that CUB between HIV and human genes
+is not shrinking at the macroevolutionary level \citep{kuyl_biased_2012}. 
+Furthermore, CUB in the V1-C5 region is not very different from other parts of
+the HIV genome, whereas the reduced fixation probability is only observed there. In
+conclusion, although we cannot exclude an effect of CUB on fitness as a general
+rule, we expect it to be a minor effect in our context.
 
@@
 deleterious mutations that are linked to adaptive variants can reach high
-frequency~\citep[genetic hitchhiking,][]{smith_hitch-hiking_1974}.
+frequency. This process is known as hitch-hiking \citep{smith_hitch-hiking_1974}
+or genetic draft \citep{gillespie_genetic_2000,neher_genetic_2011}.
 
@@
 trajectories in \FIG{aft}, where many mutations appear to change rapidly in
-frequency as a flock. Deleterious synonymous mutations can be brought to high
-frequency by selection on linked nonsynonymous sites, a process
-known as {\it genetic draft}~\citep{gillespie_genetic_2000, neher_genetic_2011}. In order
+frequency as a flock. In order
 to be advected to high frequency by a linked adaptive mutation, the deleterious
@@
 and a synonymous diversity $P_\text{interm} \sim 0.005$.
-In the three simulations shown, the fixation probability of synonymous alleles
-decreases from the neutral expectation ($A_\text{syn} \sim 0$) to zero
+In the three simulations shown in \FIG{simfixpvar}, the fixation probability of
+synonymous alleles decreases from the neutral expectation ($A_\text{syn} \sim 0$) to zero
 ($A_\text{syn} \sim -0.5$) as their fitness effect
 worsens; the synonymous diversity plummets as well, as it is harder for
-deleterious mutations to reach the frequency threshold of 0.25
-(\FIG{simfixpvar}, from $-0.0003$ in blue to $-0.003$ in red).
+deleterious mutations to reach the frequency threshold of 0.25.
 
-Among all simulations, we select the ones that show HIV-like $A_\text{syn}$ and
-$P_\text{interm}$, i.e. a large depression in
-fixation probability of synonymous mutations but, simulteneously, a moderately
-high synonymous diversity: simulations are gated such that $A_\text{syn} < -0.15$
-and $0.0025 < P_\text{interm} < 0.010$, hence allowing an uncertainty on mutation
-rates or generation times of about twofold. This selection imposes a strong
-constraint on the parameters, in particular on the ones related to synonymous
-changes, i.e. their deleterious fraction and effect size (see \FIG{simsfig}).
+To map the parameter range of the model that is compatible with the data, we
+repeatedly simulated the evolution with random choices for the deleterious
+effect of synonymous mutations, the fraction of synonymous mutations that are
+deleterious, the escape rate of adaptive mutations, and the frequency of
+escapes.  Among all simulations, we select the ones that show $A_\text{syn}$ and
+$P_\text{interm}$ as observed in the data, i.e., a large depression in fixation
+probability of synonymous mutations but, simulteneously, a moderately high
+synonymous diversity. Specifically, \FIG{simsfig} shows parameter combinations
+for which we found $A_\text{syn} < -0.15$ and $0.0025 < P_\text{interm} < 0.010$
+(twofold variation around the observed value). This selection imposes a strong
+constraint on the deleterious fraction and effect size (see \FIG{simsfig}).
 A high fraction ($\gtrsim 0.8$) of only slightly deleterious synonymous sites,
-with effect size $|s_d| \sim 0.002$, is required. This result fits well
-the expectation based on the fixation/extinction times above (see \FIG{fixp1}).
-Moreover, it corresponds to {\it a priori} theoretical expectations:
-(a) since high frequency bins are enriched in neutral mutations (it is easier for
-them to hitchhike), the hallmark of deleterious mutations will be visible there
-only if they are pervasive; (b) genetic draft stops working if the
-deleterious effect size becomes comparable to the adaptive effect size of escape
-mutations, because the double mutant has little or no fitness advantage; (c)
-mutations with a deleterious effect weaker than approximately $0.001$ are
-indistinguishable from neutral ones and are drafted seamlessly to any frequency.
+with effect size $|s_d| \sim 0.002$, is required. This result fits well the
+expectation based on the fixation/extinction times above (see \FIG{fixp1}).
+The results are plausible:
+(a) since high frequency bins are enriched in neutral mutations (it is easier
+for them to hitchhike), a depression in $\pfix$ will be
+visible only if deleterious mutations are pervasive; (b) in order to
+hitchhike, the deleterious effect size has to be less than the escape rate, otherwise the
+double mutant has little or no fitness advantage; (c) mutations with a 
+deleterious effect weaker than approximately $0.001$ behave neutrally
+consistent with the typical coalescent times observed in HIV.
 
 As long as only features of synonymous mutations are filtered, as performed
-above, no strong correlation of the two nonsynonymous parameters, effect size
+above, no strong correlation of the two nonsynonymous parameters, escape rate
 and rate of appearance of epitopes, is observed; in \FIG{simsfig}, the points
@@
 
-The former explanation can be formulated as follows: if the immune system
+The former hypothesis can be formulated as follows: if the immune system
 recognizes the escape mutant before its fixation, the mutant might cease to be
@@
 average time needed by an escape mutant to cross frequencies of order one.
-In the alternative explanation, several different escape
+In the alternative hypothesis, several different escape
 mutations within the same epitope might arise almost simultaneously and start to
@@
 achieve a reduced, neutral-like fixation probability while maintaining a high
-rate of substitutions, compatibly with HIV measurements. In real HIV infections,
-both mechanisms are likely to be playing a role.
+rate of substitutions, compatibly with HIV measurements \comment{Supplementary}.
+In real HIV infections, both mechanisms are likely to be playing a role.
 
@@
 josefsson_majority_2011}. The recombination rate has been estimated to be on the
-order of $\rho = 10^{-5}$ per base and day. It takes roughly $t_{sw} = s_a^{-1}
-\log \nu_0$ generations for an adaptive mutation with growth rate $s_a$ to rise
+order of $\rho = 10^{-5}$ per base and day. It takes roughly $t_{sw} =
+\epsilon^{-1} \log \nu_0$ generations for escape mutation with escape rate $\epsilon$ to rise
 from an initially low frequency $\nu_0\sim \mu$ to frequency one. This implies
-that a region of length $l = (\rho t_{sw})^{-1} = s_a / \rho \log \nu_0$ remains
-linked to the adaptive mutation. With $s_a=0.01$, $l\approx 100$ bases, hence we 
-expect strong linkage between the variable loops and their surrounding stems,
-but none far beyond the variable regions, consistent with the lack of signal
+that a region of length $l = (\rho t_{sw})^{-1} = \epsilon / \rho \log \nu_0$
+remains linked to the adaptive mutation. With $\epsilon=0.01$, we have $l\approx
+100$ bases. Hence we expect strong linkage between the variable loops and their
+surrounding stems, but none far beyond the variable regions, consistent with the lack of signal
 outside of C1-V5. In case of much stronger selection -- such as observed during
 early CTL escape or drug resistance evolution -- the linked  region is of course
-much larger.
+much larger \citep{nijhuis_stochastic_1998}.
 
-The functional significance of the insulating RNA structure stems between the
+A functional significance of the insulating RNA structure stems between the
 hyper variable loops has been proposed
@@
 
-Our study uncovers the
-subtle balance of evolutionary forces governing intrapatient HIV evolution. The
-fixation and extinction times and probabilities represent a rich and simple
-summary statistics useful to characterize sequencing data and compare to
-models via computer simulations.
+The fixation and extinction times and probabilities represent a rich and simple
+summary statistics useful to characterize longitudinal sequence data and compare
+to models via computer simulations.
 A similar method has been recently used in a longitudinal study of
 influenza~\citep{strelkowa_clonal_2012}. The propagators suggested in that
-paper, however, represent ratios between nonsynonymous
-mutations and synonymous ones, hence they are inadequate to investigate
-synonymous changes themselves. The authors also conclude that several
-beneficial mutations segregate simultaneously in influenza, a scenario
-remarkably similar to our within-epitope competition picture. While
-recombination hardly affects evolution within an epitope, it might facilitate
-immune escape by easing competition between different epitopes
-\citep{neher_rate_2010,Rouzine:2005p17398}.
+paper, however, represent ratios between nonsynonymous and synonymous mutations.
+The latter is used as an approximately neutral control and this method can
+therefore not be used to investigate synonymous changes themselves.
 
@@
 models of HIV evolution and the common assumption that selection is time
-independent. If genetic variation is only transiently beneficial, existing
-methods to quantify selection will yield substantial underestimates
-\citep{williamson_adaptation_2003,neher_rate_2010,OTHER}. To explain the
+independent or additive. If genetic variation is only transiently beneficial,
+existing methods to quantify selection will yield substantial underestimates
+\citep{williamson_adaptation_2003,neher_rate_2010,batorsky_estimate_2011}. To explain the
 observations regarding the fixation probabilities of non-synonymous mutations,

### 2013-01-29 synmut.tex
edits in captions

@@
 is simply the likelihood that the future common ancestor is currently carrying
-it, i.e. its frequency $\nu$.}
+it, i.e., its frequency $\nu$.}
 \label{fig:aft}
@@
 \label{fig:fixp2}}
-\caption{(Panel A) Time course of loss and fixation of synonymous mutations
-observed in a frequency interval $\nu_0$. The ultimate fraction of synonymous
-mutations that fix as a function of intermediate frequency $\nu_0$ is the
-fixation probability. (Panel B) Fixation probability of derived synonymous
-alleles is strongly suppressed in C2-V5 versus other parts of the {\it env}
-gene. The horizontal error bars on the abscissa are bin sizes, the vertical ones
-the standard deviation after 100 patient bootstraps of the data. Data from
+\caption{Fixation and loss of synonymous mutations. Panel A shows the time
+course of loss and fixation of synonymous mutations observed in a frequency interval $\nu_0$. 
+In each frequency interval, the  fraction of synonymous
+mutations that ultimately fix is the fixation probability conditional on the
+initial frequency.
+Panel B) Fixation probability of derived synonymous
+alleles as a function of $\nu_0$ in C2-V5 is below the neutral
+expectation indicated by the diagonal line. This suppression is not
+oberserved in other parts of {\it env} or for nonsynonymous mutations.
+The horizontal error bars on the abscissa are bin sizes, the vertical ones the
+standard deviation after 100 patient bootstraps of the data. Data from
 Refs.~\cite{shankarappa_consistent_1999, bunnik_autologous_2008}.}
@@
 shown in \FIG{SHAPEA} in a cumulative histogram, the reactivities of fixed
-alleles (red line) are systematically larger than of alleles that are doomed to
-extinction (blue line) (Kolmogorov-Smirnov test, $P\approx 0.002$).
-In other words, alleles that are likely to be
-breaking RNA helices are also more likely to revert and finally be lost from the
-population. As a control, the average over non-observed but potentially
-available polymorphisms lies between the two curves (green line), as expected
-(because only some of these mutations will interfere with stem loop formation).
-To test the hypothesis that mutations in C2-V5 are lost since break stems
-in the conserved between the variable loops, we split the synonymous mutations
-in the extended V1-V5 region further into conserved and variable regions and
-found that the biggest depression in fixation probability is observed in the
-conserved stems, while the variable loops show little deviation from the
-neutral signature, see \FIG{SHAPEB}. This is consistent with important stem
-structures in conservered regions between loops.
+alleles (red line) are systematically larger than of alleles are lost (blue
+line) (Kolmogorov-Smirnov test, $P\approx 0.002$).
+In other words, alleles that are likely to break RNA helices are also more
+likely to revert and finally be lost from the population. As a control, the
+average over non-observed but potentially available polymorphisms lies between
+the two curves (green line), as expected (because only some of these mutations
+will interfere with stem loop formation).
+To test the hypothesis that mutations in C2-V5 are lost since break stems in the
+conserved between the variable loops, we split the synonymous mutations in the
+extended V1-V5 region further into conserved and variable regions and found that
+the biggest depression in fixation probability is observed in the conserved
+stems, while the variable loops show little deviation from the neutral
+signature, see \FIG{SHAPEB}. This is consistent with important stem structures
+in conservered regions between loops.
 
@@
 \begin{center}
-\subfloat{\includegraphics[width=0.9\linewidth]{mixed_Shankarappa_Bunnik2008_Liu_fixation_reactivity_Vandflanking_fromSHAPE}
-\label{fig:SHAPEA}}\\
-\subfloat{\includegraphics[width=0.9\linewidth]{Shankarappa_fixmid_syn_V_regions.pdf}\label{fig:SHAPEB}}
+\subfloat{\includegraphics[width=0.48\linewidth]{reactivities_histograms_syn}
+\label{fig:SHAPEA}}
+\subfloat{\includegraphics[width=0.48\linewidth]{Shankarappa_fixmid_syn_V_regions.pdf}\label{fig:SHAPEB}}
 \caption{Permissible synonymous mutations have
-high SHAPE reactivities. \comment{Is this C2-V5 only?} 
-Panel A) Synonymous substitions in the C2-V5 region have substantially higher
-SHAPE reactivities than mutations that reach frequencies above 15\%, but are
-susequently lost. This difference is quantified by a cumulative histogram
-($p=0.002$, Kolmogorov-Smirnov two sample test). Mutations that are never
+high SHAPE reactivities.
+Panel A) Synonymous substitions have
+substantially higher SHAPE reactivities than mutations that reach frequencies above 15\%, but are
+susequently lost (restricted to the vicinity of variable loops). This difference
+is quantified by a cumulative histogram ($p=0.002$, Kolmogorov-Smirnov two 
+sample test). Mutations that are never
 observered show an intermediate distribution of SHAPE values.
-Panel B shows the fixation probablility of synonymous mutations in C2-V5
+Panel B shows the fixation probability of synonymous mutations in C2-V5
 separately for variable loops and the connecting conserved regions, that
@@
 \label{fig:simsfig}}
-\caption{(Panel A) The depression in $P_\text{fix}$ depends on the deleterious
-effect size of the synonymous alleles. This parameter also reduces synonymous
-diversity, measured by probability of a derived allele to be found at
+\caption{Distribution of selection coefficients on synonymous sites. Panel A)
+The depression in $P_\text{fix}$ depends on the deleterious effect size 
+of the synonymous alleles. This parameter also reduces synonymous
+diversity, measured by the probability of a derived allele to be found at
 intermediate frequencies $P_\text{interm}$ (first inset).
-(Panel B) To assess the parameter space that affects synonymous fixation and
+Panel B) To assess the parameter space that affects synonymous fixation and
 diversity, we run many simulations with random parameters for deleterious effect
@@
 fixation probability and diversity. Parameters are chosen uniformly on a
-logarithmic scale from the indicated windows. \comment{should we extend the
-parameters space maybe to $10^{-4}$?}}
+logarithmic scale from the indicated windows. This simulation assumes 6
+roughly equivalent escape mutations per epitope; see supplement for alternative
+models. }
 \label{fig:simheat}

### 2013-01-30 synmut.tex
corrected time-dep sims and figure adaptations
rewrote discussion and abstract

@@
 \newcommand{\env}{\textit{env}}
-\newcommand{\shankaregion}{C2-C5}
+\newcommand{\shankaregion}{C2-V5}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \noindent
-Intrapatient HIV evolution is governed by selection on the protein level in the
-arms race with the immune system (killer T-cells and antibodies). Synonymous
-mutations do not have an immunity-related phenotype and are often assumed to be
-neutral. In this paper, we show that synonymous changes in epitope-rich regions
-are often deleterious but still reach high frequencies in the viral population.  We analyze time
-series of viral sequences from the \shankaregion~part of {\it env} within individual
-hosts and observe that synonymous derived alleles rarely reach fixation.
-Simulations suggest that such synonymous mutations
-have a (Malthusian) selection coefficient of the order of $-0.001$, and that
-they are brought up to high frequency by hitchhiking on neighboring beneficial
-nonsynonymous alleles. We detect a negative correlation between the fixation of an allele and
-its involvement in RNA stem-loop structures.
-Deleterious synonymous mutations are not observed as abundantly in other parts of the HIV genome, in which
-selective sweeps are less dense and hitchhiking not as strong; this behaviour is
-confirmed by extensive computer simulations.
+
+Intrapatient HIV evolution is dominated by selection on the protein level in the
+arms race with the immune system. Synonymous mutations do not have an
+immunity-related phenotype and are often assumed to be neutral. Here, we show
+that synonymous changes in epitope-rich regions are often deleterious but still
+reach high frequencies in the viral population.
+We analyze longitudinal intra-patient data from the \shankaregion~part of the
+envelope gene (\env) and observe that synonymous derived alleles rarely reach
+fixation.
+Simulations suggest that such synonymous mutations have a (Malthusian) selection
+coefficient of the order of $-0.002$, and that they are brought up to high
+frequency by hitchhiking on neighboring beneficial nonsynonymous alleles. We
+detect a negative correlation between the fixation of an allele and its
+involvement in RNA stem-loop structures.
+The patterns of fixation of non-synonymous mutations suggest that antibody
+escape mutations in \shankaregion~are only transiently beneficial, either since
+the immune system is catching up or because of competition between equivalent
+escapes.
 
@@
 \label{fig:fixp1}}\\
-\subfloat{\includegraphics[width=0.9\linewidth]{Bunnik2008_fixmid_syn_ShankanonShanka}
+\subfloat{\includegraphics[width=0.9\linewidth]{fixation_probabilities}
 \label{fig:fixp2}}
@@
 \begin{center}
-\subfloat{\includegraphics[width=0.48\linewidth]{reactivities_histograms_syn}
+\subfloat{\includegraphics[height=0.46\linewidth]{reactivities_histograms_syn}
 \label{fig:SHAPEA}}
-\subfloat{\includegraphics[width=0.48\linewidth]{Shankarappa_fixmid_syn_V_regions.pdf}\label{fig:SHAPEB}}
+\subfloat{\includegraphics[height=0.46\linewidth]{fixation_probabilities_VnonV}\label{fig:SHAPEB}}
 \caption{Permissible synonymous mutations have
@@
 genome, synonymous mutations are often used as approximately neutral markers in
-evolutionary studies of viruses. We have shown that the majority of synonymous
-mutations in the conserved regions C2-C5 of the \env~gene are deleterious.
+evolutionary studies of viruses. By analyzing the fate of mutations in
+longitudinal data of HIV \env{} evolution, we have shown that the majority of
+synonymous mutations in the conserved regions C2-C5 of the \env~gene are deleterious.
 Comparison with recent biochemical studies of binding propensity of bases in RNA
@@
 mutations are brought to high frequency through linkage to adaptive mutations.
-The latter mutations are only transiently adaptive, either through a
-coevolution with the immune system or redundant escape within an epitope.
 
-Our observations and conclusion rely heavily on longitudinal data in which the
-dynamics of mutations can be explicitly observed. The fact that deleterious
-mutations can be brought to high frequencies through hitchhiking underscores
-the intensity of the coevolution with the immune system. The fact that
-multiple escape mutations in the same epitope -- as is indeed observed in
-studies of antibody escape~\citep{moore_limited_2009, bar_early_2012} -- are
-necessary to explain the patterns of fixation of nonsynonymous mutations points
-towards a large populations size that rapidly discovers adaptive mutations. A
+Contrary to naive expectations, the adaptive escape mutations don't seem to be
+unconditionally beneficial. Otherwise we should observe almost sure fixation of
+non-synonymous mutations once they reach intermediate frequency. Instead, we
+find that the fixation probability of non-synonymous mutations is roughly given
+by its frequency. There are several possible explanations for this observation. 
+Similar to synonymous mutations, the majority of non-synonymous mutations could
+be weakly deleterious and the adaptive and deleterious parts conspire to yield
+the neutral-like averaged fixation probability. We consider this possibility
+inplausible since the amino-acid sequence outside the variable loops is much
+more conserved that the synonymous positions, suggesting that the majority of
+the non-synonymous mutations is much more deleterious.
+
+Alternatively, the lack of fixation could be due to time dependent environment
+through an immune system that is catching up, or competition between mutations
+that mediate escape within the same epitope. We explored both of these
+possibilities and find that both produce the desired effect. Furthermore, there
+is evidence in support of both of these hypothesis. Serum from HIV infected
+individuals tyically neutralizes the virus that dominated the population a few
+(3-6) month ago \citep{richman_rapid_2003}. This suggests that escape mutations
+cease to be beneficial after a few month and might revert if they come with a
+fitness cost. Deep sequencing of regions of \env{} after antibody escape have
+revealed multiple escape mutations in the same epitope
+\citep{moore_limited_2009, bar_early_2012}. Presumably, each one of these
+mutations is sufficient for escape but most combinations of them don't provide
+any additional benefit to the virus. Hence only one of them will spread and the
+others will be driven out of the population even if they transiently reach high
+frequencies. The rapid emergence of multiple escape mutations in the same
+epitope implies a large population size that explores all necessary point mutations rapidly. A
 similar point has been made recently by Boltz {\it et al.} in the context of
@@
 order of $\rho = 10^{-5}$ per base and day. It takes roughly $t_{sw} =
-\epsilon^{-1} \log \nu_0$ generations for escape mutation with escape rate $\epsilon$ to rise
-from an initially low frequency $\nu_0\sim \mu$ to frequency one. This implies
-that a region of length $l = (\rho t_{sw})^{-1} = \epsilon / \rho \log \nu_0$
-remains linked to the adaptive mutation. With $\epsilon=0.01$, we have $l\approx
-100$ bases. Hence we expect strong linkage between the variable loops and their
-surrounding stems, but none far beyond the variable regions, consistent with the lack of signal
-outside of C1-V5. In case of much stronger selection -- such as observed during
-early CTL escape or drug resistance evolution -- the linked  region is of course
-much larger \citep{nijhuis_stochastic_1998}.
+\epsilon^{-1} \log \nu_0$ generations for escape mutation with escape rate
+$\epsilon$ to rise from an initially low frequency $\nu_0\sim \mu$ to frequency
+one. This implies that a region of length $l = (\rho t_{sw})^{-1} = \epsilon /
+\rho \log \nu_0$ remains linked to the adaptive mutation. With $\epsilon=0.01$,
+we have $l\approx 100$ bases. Hence we expect strong linkage between the
+variable loops and the intervening sequence, but none far beyond the variable
+regions, consistent with the lack of signal outside of C1-V5. In case of much
+stronger selection -- such as observed during early CTL escape or drug
+resistance evolution -- the linked  region is of course much larger
+\citep{nijhuis_stochastic_1998}.
 
 A functional significance of the insulating RNA structure stems between the
-hyper variable loops has been proposed
+hyper-variable loops has been proposed
 previously~\citep{watts_architecture_2009, sanjuan_interplay_2011}.
-\citet{sanjuan_interplay_2011} have shown that insulating stems are relevant for
-viral fitness {\it in vivo}. Our analysis is limited by the availability of
+Our analysis is limited by the availability of
 longitudinal data which requires a focus on the the variable regions of \env.
@@
 existing methods to quantify selection will yield substantial underestimates
-\citep{williamson_adaptation_2003,neher_rate_2010,batorsky_estimate_2011}. To explain the
-observations regarding the fixation probabilities of non-synonymous mutations,
-either transient selection, or substational within-epitope competition are
-necessary. Which mechanism is more widespread is not clear as of now,
-there is evidence for both~\citep{richman_rapid_2003, moore_limited_2009,
-bar_early_2012}.
+\citep{williamson_adaptation_2003,neher_rate_2010,batorsky_estimate_2011}.
 
@@
 liu_selection_2006, bunnik_autologous_2008} and downloaded from the Los Alamos
-National Laboratory (LANL) HIV sequence database~\citep{LANL2012}. The sequences from
-some patients showed signs of HIV compartimentalization into subpopulations and
-were discarded; a grand total of 11
+National Laboratory (LANL) HIV sequence database~\citep{LANL2012}. The sequences from some patients
+showed signs of HIV compartimentalization into subpopulations and were discarded; a total of 11
 patients with approximately 6 time points each and 10 sequences per time point
@@
 by \citet{watts_architecture_2009} in the SHAPE assay. Within each patient, a
-consensus RNA sequence at the first time point was used to classify alleles as ancestral or
-derived at all sites. Problematic sites that included large frequencies of gaps
+consensus nucleotide sequence at the first time point was used to classify
+alleles as ancestral or derived at all sites. Problematic sites that included large frequencies of gaps
 were excluded from the analysis to avoid artefactual substitutions due to
 alignment errors. Time series of allele frequencies were extracted from the
-sequences.
+sequences. \comment{The alignments used are available for download}.
 
@@
 
+\comment{All analysis scripts are available for download\ldots.}. 
 \section*{Acknowledgements}
-\comment{to be written\dots}
+We are grateful for stimulating discussions with Jan Albert and Trevor Bedford.
+This work is supported by the ERC starting grant HIVEVO 260686.
 

### 2013-02-01 supplementary.tex
supplementary first draft

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Template for a revtex article
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[rmp]{revtex4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym,textcomp}
\usepackage{color}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage{natbib}
\usepackage{pslatex}\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\comment}[1]{{\color{red}#1}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\thesubfigure}{\Alph{subfigure}}
\newcommand{\Author}{Fabio~Zanini* and Richard~A.~Neher*}
\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV \env~evolution}
\newcommand{\Affiliation}{Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany}
\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{\Title}
\author{\Author}
\affiliation{\textsf{*}\Affiliation}
\date{\today}
\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\tableofcontents

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Selection of the patient data}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.48\linewidth]{Shankarappa_PCA_p7}
\includegraphics[width=0.48\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p7}
\caption{Left: PCA of all sequences from patient p7 (colors indicate time from seroconversion,
from blue to red). Right: Allele frequency trajectories for nonsynonymous
changes in the same patient.}
\label{fig:aftp7}
\end{center}
\end{figure}
Not all of the patients studied in refs.~\citealp{shankarappa_consistent_1999} and
\citealp{bunnik_autologous_2008} were analyzed. The reason is that patients p4,
p7, p8, p9 from ref.~\citealp{shankarappa_consistent_1999} and patients ACH19542
and ACH19768 from ref.~\citealp{bunnik_autologous_2008} showed clear signs of
compartimentalization (population structure) during infection, which would have
been an irreparable confounder for our analysis. The difference is clear if
PCA is performed on all sequences from a ``compartimentalized'' patient, p7.
This is shown in \figurename~\ref{fig:aftp7}: late sequences subdivide in two
well-defined clusters -- this characteristic is absent from all the patient
sequences used in our main analysis. On the right, allele frequency trajectories from the same
patient are plotted: evidently, no derived allele is fixed after 1000 days
from seroconversion, but a lot of mutations swarm to frequencies around 0.5-0.8. Of
course, the fixation probability of new mutations in such a sample is very low,
but this is an effect of population structure rather than any
of the processes discussed in the main text (time-dependent selection, competition
of mutants within the same epitope).

\clearpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Nonsynonymous changes outside of variable regions are deleterious}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.5\linewidth]{synmut_conservation_4fold_synnonsyn}
\caption{Cumulative distribution of synonymous and nonsynonymous diversity in
{\it gag}. Nonsynonymous changes are, overall, much less common than synonymous
ones.}
\label{fig:synnonsyncons}
\end{center}
\end{figure}
In the main text, it is argued that, whereas the deleterious effect of
synonymous changes is very small, most nonsynonymous mutations outside of the
variable regions are much worse. To show this, we take a multiple sequence alignment of
the {\it gag} gene of subtype-B HIV sequences from the LANL website (filtered
sequences only, version 2011)~\cite{LANL2012} and look at diversity at fourfold degenerate
codons. The ratio of observed versus possible mutations is much lower for
nonsynonymous changes than synonymous ones, as shown in
\figurename~\ref{fig:synnonsyncons} in a cumulative histogram. From this
evidence (which is highly statistically significant) we conclude that, as a
general rule, nonsynonymous changes that are not currently escape mutations
are likely to incur in large fitness costs.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Synonymous diversity across the HIV genome}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=\linewidth]{conservation_codons_genome}
\caption{Normalized codon entropy across the HIV genome is clustered around
variable regions in {\it env}.}
\label{fig:syndiv_genome}
\end{center}
\end{figure}

Our analysis relies on the ability of synonymous mutation to reach high
frequencies within the viral population (by hitchhiking). The ability of HIV to
explore the synonymous mutational space can be quantified by the codon entropy.
Using a multiple sequence alignment of subtype-B HIV sequences from the LANL
website (filtered sequences only, version 2011)~\cite{LANL2012}, we show in
\figurename~\ref{fig:syndiv_genome} that the genomic regions within and flanking
V loops contain the highest codon entropy (normalized by the maximal entropy for
each degenerate codon). On the contrary, regions that are under selection either
the the RNA level (the RRE hairpin) or at the protein level but in a different
reading frame (rev second exon), show reduced entropy, i.e. more constrains.

All parts of {\it env} that are part of a different gene have been excluded from
our main analysis, to avoid contamination about synonymity.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Alternative models for suppression of nonsynonymous mutations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.5\linewidth]{simulations_graduallyepitopesandtimeselec}
\caption{Time-dependent selection (the escape mutant is recognized after some
months from its appearance) and competition between escape mutations in the same
epitope both reduce fixation of nonsynonymous mutations, without affecting the
synonymous curve much.}
\label{fig:tds_wec}
\end{center}
\end{figure}
As mentioned in the main text, both competition of different mutants within the
same allele and a slow, catching-up immune system can reduce the fixation
probability of nonsynonymous derived alleles. This is observed in computer
models, an example of which is shown in \figurename~\ref{fig:tds_wec}. In the
curves denoted with ``tds'' (time-dependent selection), the fitness advantage of
an escape mutaition can vanish at each time point with a probability density
\[ P_\text{recognized}(t) = c \cdot \nu(t), \]
where $c$ is a constant coefficient shown in the legend that encodes the overall
efficiency of the host immune system, and $\nu(t)$ is the frequency of the
escape allele at time $t$. The curves denoted with ``mut/epi'' describe the
competition scenario instead, with 3 or 6 escape mutations within the same
epitope of length 9 nucleotides. Increasing either the coefficient $c$ or the
number of escape mutants per epitope reduces fixation of nonsynonymous derived
alleles. Importantly, the synonymous fixation probability is hardly affected, which casts
trust over the main conclusions of the main text independently on the details of
the selection scenario at the protein level.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{natbib}
\bibliography{bib}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2013-02-01 synmut.tex
supplementary first draft

@@
 involvement in RNA stem-loop structures.
-The patterns of fixation of non-synonymous mutations suggest that antibody
+The patterns of fixation of nonsynonymous mutations suggest that antibody
 escape mutations in \shankaregion~are only transiently beneficial, either since
@@
 more often than neutral ones, respectively. \FIG{aft}A\&B show 
-the time course the frequencies of all synonymous and non-synonymous mutations
+the time course the frequencies of all synonymous and nonsynonymous mutations
 observed \env~, C2-V5, in patient p8~\citep{shankarappa_consistent_1999},
@@
 roughly equivalent escape mutations per epitope; see supplement for alternative
-models. }
+models.}
 \label{fig:simheat}
@@
 
-Contrary to naive expectations, the adaptive escape mutations don't seem to be
-unconditionally beneficial. Otherwise we should observe almost sure fixation of
-non-synonymous mutations once they reach intermediate frequency. Instead, we
-find that the fixation probability of non-synonymous mutations is roughly given
+Contrary to na\"ive expectations, the adaptive escape mutations don't seem to be
+unconditionally beneficial. Otherwise we would observe almost sure fixation of
+nonsynonymous mutations once they reach intermediate frequencies. Instead, we
+find that the fixation probability of nonsynonymous mutations is roughly given
 by its frequency. There are several possible explanations for this observation. 
-Similar to synonymous mutations, the majority of non-synonymous mutations could
+Similar to synonymous mutations, the majority of nonsynonymous mutations could
 be weakly deleterious and the adaptive and deleterious parts conspire to yield
@@
 more conserved that the synonymous positions, suggesting that the majority of
-the non-synonymous mutations is much more deleterious.
+the nonsynonymous mutations is much more deleterious.
 
-Alternatively, the lack of fixation could be due to time dependent environment
-through an immune system that is catching up, or competition between mutations
+Alternatively, the lack of fixation could be due to (a) time dependent environment
+through an immune system that is catching up, or (b) competition between mutations
 that mediate escape within the same epitope. We explored both of these
@@
 (3-6) month ago \citep{richman_rapid_2003}. This suggests that escape mutations
-cease to be beneficial after a few month and might revert if they come with a
+cease to be beneficial after a few months and might revert if they come with a
 fitness cost. Deep sequencing of regions of \env{} after antibody escape have
@@
 mutations is sufficient for escape but most combinations of them don't provide
-any additional benefit to the virus. Hence only one of them will spread and the
-others will be driven out of the population even if they transiently reach high
+any additional benefit to the virus. Hence only one mutation will spread and the
+others will be driven out of the population although they transiently reach high
 frequencies. The rapid emergence of multiple escape mutations in the same
@@
 \subsection{Sequence data collection}
+
 Longitudinal intrapatient viral RNA sequences were collected for published
-studies~\citep{shankarappa_consistent_1999,
-liu_selection_2006, bunnik_autologous_2008} and downloaded from the Los Alamos
-National Laboratory (LANL) HIV sequence database~\citep{LANL2012}. The sequences from some patients
-showed signs of HIV compartimentalization into subpopulations and were discarded; a total of 11
-patients with approximately 6 time points each and 10 sequences per time point
-were analyzed. The time interval or resolution between two consecutive sequences
-was approximately 6 to 18 months.
+studies~\citep{shankarappa_consistent_1999, liu_selection_2006,
+bunnik_autologous_2008} and downloaded from the Los Alamos National Laboratory
+(LANL) HIV sequence database~\citep{LANL2012}. The sequences from some patients
+showed signs of HIV compartimentalization into subpopulations and were
+discarded; a total of 11 patients with approximately 6 time points each and 10
+sequences per time point were eventually analyzed. The time interval or
+resolution between two consecutive sequences was approximately 6 to 18 months.
 

### 2013-02-03 synmut.tex
text small changes

@@
 Intrapatient HIV evolution is dominated by selection on the protein level in the
-arms race with the immune system. Synonymous mutations do not have an
-immunity-related phenotype and are often assumed to be neutral. Here, we show
+arms race with the adaptive immune system. Synonymous mutations do not have an
+adaptive immunity-related phenotype and are often assumed to be neutral. Here, we show
 that synonymous changes in epitope-rich regions are often deleterious but still
 reach high frequencies in the viral population.
-We analyze longitudinal intra-patient data from the \shankaregion~part of the
-envelope gene (\env) and observe that synonymous derived alleles rarely reach
-fixation.
-Simulations suggest that such synonymous mutations have a (Malthusian) selection
+We analyze longitudinal intrapatient data from the \shankaregion~part of the
+envelope gene (\env) and observe that synonymous derived alleles are rarely
+fixed.
+Computer models suggest that such synonymous mutations have a (Malthusian) selection
 coefficient of the order of $-0.002$, and that they are brought up to high
-frequency by hitchhiking on neighboring beneficial nonsynonymous alleles. We
-detect a negative correlation between the fixation of an allele and its
-involvement in RNA stem-loop structures.
+frequency by hitchhiking on neighboring beneficial nonsynonymous alleles.
+We find that synonymous mutations that disrupt base pairs in RNA stems flanking
+the V loops are more likely to be lost than other synonymous changes, hinting at
+a direct fitness effect of such stem-loop structures in the HIV RNA.
 The patterns of fixation of nonsynonymous mutations suggest that antibody
@@
 variable loops of the envelope protein gp120, which need to avoid recognition by
-neutralizing ABs.  Mutations in \env, the gene encoding for gp120, spread
-through the population within a few months (see \figurename~\ref{fig:aft}, solid
-lines). Consistent with this time scale, it is found that serum from a
+neutralizing ABs.  Mutations in \env, the gene encoding gp120, spread
+through the population within a few months (see \figurename~\ref{fig:aftnonsyn}).
+Consistent with this time scale, it is found that serum from a
 particular time typically neutralizes virus extracted more than 3-6 month
@@
 that of putatively functional sites to detect purifying or directional selection
-\citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In most species,
-however, it is  known that synonymous codons don't evolve completely neutrally
-but that some codons are favored over others \citep{plotkin_synonymous_2011}.
-This is expected to be particularly important in compact viruses like HIV. 
+\citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}.
 In addition to maintaining protein function and avoiding the adaptive immune
-recognition, the HIV genome has to ensure efficient processing and translation,
+recognition, however, the HIV genome has to ensure efficient processing and translation,
 nuclear export, and packaging into the viral capsid: all these processes operate at the RNA
-level and are sensitive to synonymous changes. A few functionally important RNA
-elements are well characterized. For example, a certain RNA sequence in the HIV
-genome, called \rev{} response element (RRE), enhances nuclear export of
+level and are sensitive to synonymous changes.
+On the one hand, these processes often involve some kind of RNA secondary structure.
+For example, the HIV \rev{} response element (RRE) in \env{} enhances nuclear export of
 full length or partially spliced viral transcripts~\citep{fernandes_hiv-1_2012}.
@@
 untranslated region~\citep{barat_interaction_1991, paillart_vitro_2002}.
-Nucleotide-level fitness effects have been observed beyond RNA structure as
-well. Recent studies have shown that genetically engineered HIV strains with
+On the other hand, nucleotide-level fitness effects have been observed beyond RNA structure as
+well.
+In most species, it is  known that synonymous codons don't evolve completely neutrally
+but that some codons are favored over others \citep{plotkin_synonymous_2011};
+this is expected to be particularly important in compact viruses like HIV. 
+Recent studies have shown that genetically engineered HIV strains with
 altered codon usage can in some cases produce more viral protein, but in general
@@
 selection beyond the protein sequence is therefore expected
-\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}, but while positive
-selection through the immune system be mainly restricted to changes in the
+\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}; nonetheless, positive
+selection via the host adaptive immune system is restricted to changes in the
 amino acid sequence.
 
-%SYNONYMOUS CONSERVATION. DO WE HAVE A PLOT OF GENOME WIDE CONSERVATION, MAYBE
-%FOR SUPPLEMENT? YES
-
 In this paper, we characterize the dynamics of synonymous mutations in \env{}
-and show that a substantial fraction of these mutations is deleterious. We then
+and show that a substantial fraction of these mutations is deleterious.
+We argue that synonymous mutations reach high frequencies via genetic hitchhiking
+thanks to the small recombination rate of HIV~\citep{neher_recombination_2010,
+batorsky_estimate_2011}.
+We then
 compare our observations to computational models of HIV evolution and derive
 estimates for the effect synonymous mutations have on fitness.
-Hitchhiking of synonymous mutations with escape mutations is due to the small
-recombination rate of HIV~\citep{neher_recombination_2010,
-batorsky_estimate_2011}. Extending the analysis of fixation probabilities to the
+Extending the analysis of fixation probabilities to the
 nonsynonymous mutations, we show that time dependent selection or strong
@@
 Deleterious or beneficial mutations fix less or
-more often than neutral ones, respectively. \FIG{aft}A\&B show 
+more often than neutral ones, respectively. \FIG{aft} shows 
 the time course the frequencies of all synonymous and nonsynonymous mutations
@@
 respectively. Despite many synonymous mutations reaching high frequency, very
-few fix (panel~\ref{fig:aftsyn}); however, many nonsynonymous mutations fix
+few fix (panel~\ref{fig:aftsyn}); in constrast, many nonsynonymous mutations fix
 (panel~\ref{fig:aftnonsyn}).
@@
 $\pfix$ of synonymous mutations is far below the neutral expectation.
-Outside of C2-V5 using data from \citet{bunnik_autologous_2008} only, no such
+Outside of C2-V5, using data from \citet{bunnik_autologous_2008} only, no such
 reduction in $\pfix$ is found. Restricted to the C2-V5 area, the data from
@@
 When interpreting these results for the fixation probabilities, it is important
-to distinguish random mutations, and polymorphisms at a given frequency that has
-already been filtered by selection.
+to distinguish between random mutations and polymorphisms at a given frequency,
+which have already been filtered by selection.
 A polymorphism could be beneficial to the virus and on its way to fixation. In
@@
 frequency. If, on the other hand, the polymorphism is deleterious it must have
-gotten to high frequency by a chance event (genetic drift or hitch-hiking), and
+gotten to high frequency by chance (genetic drift or hitchhiking), and
 we expect that selection will drive it out of the population again. Hence our
@@
 neutral. Note that this does not imply that all synonymous mutations are neutral
--- only those mutations observed at high frequencies tend to be neutral.
+-- only those mutations observed at high frequencies, which have been
+experiencing selection for some time already, tend to be neutral.
 
@@
 \label{fig:fixp2}}
-\caption{Fixation and loss of synonymous mutations. Panel A shows the time
+\caption{Fixation and loss of synonymous mutations.
+Panel A) Time
 course of loss and fixation of synonymous mutations observed in a frequency interval $\nu_0$. 
@@
 ref.~\citep{watts_architecture_2009} and assign them SHAPE reactivities. As
-shown in \FIG{SHAPEA} in a cumulative histogram, the reactivities of fixed
+shown in \FIG{SHAPEA}, the reactivities of fixed
 alleles (red line) are systematically larger than of alleles are lost (blue
@@
 the two curves (green line), as expected (because only some of these mutations
-will interfere with stem loop formation).
-To test the hypothesis that mutations in C2-V5 are lost since break stems in the
-conserved between the variable loops, we split the synonymous mutations in the
+would interfere with stem-loop formation).
+To test the hypothesis that mutations in C2-V5 are lost because they break stems in the
+conserved streches between the variable loops, we split the synonymous mutations in the
 extended V1-V5 region further into conserved and variable regions and found that
@@
 observered show an intermediate distribution of SHAPE values.
-Panel B shows the fixation probability of synonymous mutations in C2-V5
+Panel B) Fixation probability of synonymous mutations in C2-V5
 separately for variable loops and the connecting conserved regions, that
@@
 
-\subsection{Deleterious mutations are brought to high frequency by hitch-hiking}
+\subsection{Deleterious mutations are brought to high frequency by hitchhiking}
 
@@
 deleterious mutations that are linked to adaptive variants can reach high
-frequency. This process is known as hitch-hiking \citep{smith_hitch-hiking_1974}
+frequency. This process is known as hitchhiking \citep{smith_hitch-hiking_1974}
 or genetic draft \citep{gillespie_genetic_2000,neher_genetic_2011}.
@@
 effect of synonymous mutations, the fraction of synonymous mutations that are
-deleterious, the escape rate of adaptive mutations, and the frequency of
+deleterious, the fitness effect of adaptive mutations (escape rate) and the frequency of
 escapes. Among all simulations, we select the ones that show $A_\text{syn}$ and
@@
 synonymous diversity. Specifically, \FIG{simsfig} shows parameter combinations
-for which we found $A_\text{syn} < -0.15$ and $0.0025 < P_\text{interm} < 0.010$
-(twofold variation around the observed value). This selection imposes a strong
+for which we found $A_\text{syn} < -0.15$ and $0.0025 < P_\text{interm} <
+0.010$. This selection imposes a strong
 constraint on the deleterious fraction and effect size (see \FIG{simsfig}).
@@
 visible only if deleterious mutations are pervasive; (b) in order to
-hitchhike, the deleterious effect size has to be less than the escape rate, otherwise the
+hitchhike, the deleterious effect size has to be much smaller than the escape
+rate, otherwise the
 double mutant has little or no fitness advantage; (c) mutations with a 
@@
 achieve a reduced, neutral-like fixation probability while maintaining a high
-rate of substitutions, compatibly with HIV measurements \comment{Supplementary}.
-In real HIV infections, both mechanisms are likely to be playing a role.
+rate of substitutions, compatibly with HIV measurements (see supplementary
+materials).
+In real HIV infections, both mechanisms are likely to be playing a role, as they
+are not mutually exclusive.
 
@@
 
-Contrary to na\"ive expectations, the adaptive escape mutations don't seem to be
+Contrary to na\"ive expectations, the adaptive escape mutations do not seem to be
 unconditionally beneficial. Otherwise we would observe almost sure fixation of
@@
 more conserved that the synonymous positions, suggesting that the majority of
-the nonsynonymous mutations is much more deleterious.
+the nonsynonymous mutations is much more deleterious (see supplementary
+materials).
 
@@
 through an immune system that is catching up, or (b) competition between mutations
-that mediate escape within the same epitope. We explored both of these
+that mediate escape within the same epitope. We explore both of these
 possibilities and find that both produce the desired effect. Furthermore, there
-is evidence in support of both of these hypothesis. Serum from HIV infected
+is experimental evidence in support of both of these hypotheses. Serum from HIV infected
 individuals tyically neutralizes the virus that dominated the population a few

### 2013-02-03 synmut.tex
fig.1 refers to p1

@@
 \end{abstract}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \maketitle
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Introduction}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Results}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 The central quantity we investigate is the probability of fixation of a
@@
 the time course the frequencies of all synonymous and nonsynonymous mutations
-observed \env~, C2-V5, in patient p8~\citep{shankarappa_consistent_1999},
+observed \env~, C2-V5, in patient p1~\citep{shankarappa_consistent_1999},
 respectively. Despite many synonymous mutations reaching high frequency, very
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Synonymous polymorphisms in \env, C2-V5, are mostly deleterious}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 We study the dynamics and fate of synonymous mutations more quantitatively by
@@
 \begin{center}
-\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_p8.pdf}
+\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_p1.pdf}
 \label{fig:aftsyn}}\\
-\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p8.pdf}
+ \subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p1.pdf}
 \label{fig:aftnonsyn}}
 \caption{Time series of allele frequencies in \env, C2-V5, from
-patient 8~\cite{shankarappa_consistent_1999},
+patient 1~\cite{shankarappa_consistent_1999},
 divided by synonymous (panel A) and nonsynonymous (panel B) derived alleles.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Synonymous mutations in C2-V5 tend to disrupt conserved RNA stems}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 One possible {\it a priori} explanation for lack of fixation of synonymous
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Deleterious mutations are brought to high frequency by hitchhiking}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 While the observation that some fraction of synonymous mutations is deleterious
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Discussion}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 Despite several known functional roles for RNA secondary structure in the HIV
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Methods}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Sequence data collection}
-
-Longitudinal intrapatient viral RNA sequences were collected for published
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+Longitudinal intrapatient viral RNA sequences are collected for published
 studies~\citep{shankarappa_consistent_1999, liu_selection_2006,
@@
 (LANL) HIV sequence database~\citep{LANL2012}. The sequences from some patients
-showed signs of HIV compartimentalization into subpopulations and were
-discarded; a total of 11 patients with approximately 6 time points each and 10
-sequences per time point were eventually analyzed. The time interval or
-resolution between two consecutive sequences was approximately 6 to 18 months.
+show signs of HIV compartimentalization into subpopulations and are
+discarded (see supplementary materials);
+a total of 11 patients with approximately 6 time points each and 10
+sequences per time point are eventually analyzed. The time interval or
+resolution between two consecutive sequences is approximately 6 to 18 months.
 
-\comment{How did you determine who to discard. Maybe we should include the PCA
-plots into the supplement.}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Sequence analysis}
-The good sequences were aligned within each patient
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+The sequences are aligned within each patient
 via the translated amino acid sequence, using
@@
 by \citet{watts_architecture_2009} in the SHAPE assay. Within each patient, a
-consensus nucleotide sequence at the first time point was used to classify
-alleles as ancestral or derived at all sites. Problematic sites that included large frequencies of gaps
-were excluded from the analysis to avoid artefactual substitutions due to
-alignment errors. Time series of allele frequencies were extracted from the
+consensus nucleotide sequence at the first time point is used to classify
+alleles as ancestral or derived at all sites. Problematic sites that include large frequencies of gaps
+are excluded from the analysis to avoid artefactual substitutions due to
+alignment errors. Time series of allele frequencies are extracted from the
 sequences. \comment{The alignments used are available for download}.
 
-The synonymity of a mutation was assigned if the rest of the codon was
+The synonymity of a mutation is assigned if the rest of the codon was
 in the ancestral state and using the standard genetic code. Cases where more
-than one mutation within the codon was observed were discarded. Slightly
-different criteria for synonymous/nonsynonymous discrimination yielded similar
+than one mutation within the codon is observed are discarded. Slightly
+different criteria for synonymous/nonsynonymous discrimination yield similar
 results.
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Fixation probability and secondary structure}
-For the estimate of times to fixation/extinction, polymorphisms were
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+For the estimate of times to fixation/extinction, polymorphisms are
 binned by frequency and the time to reaching the first boundary (fixation or
-extinction) was stored. For the fixation probability, the long-time limit of the
-resulting curves was used, excluding polymorphisms that arose late in the
-clinical history (and would have had no time to reach either boundary).
+extinction) is stored. For the fixation probability, the long-time limit of the
+resulting curves is used, excluding polymorphisms that arise late in the
+clinical history (hence would have no time to reach either boundary).
 
-For the correlation analysis with RNA secondary structure, the SHAPE scores were
+For the correlation analysis with RNA secondary structure, the SHAPE scores are
 downloaded from the journal website~\citep{watts_architecture_2009}. By virtue
 of the alignment of the longitudinal sequences with the reference used by
-\citet{watts_architecture_2009}, SHAPE reactivities were assigned to most sites.
-Problematic assignments in indel-rich regions were excluded from the analysis.
+\citet{watts_architecture_2009}, SHAPE reactivities are assigned to most sites.
+Problematic assignments in indel-rich regions are excluded from the analysis.
 In order to restrict the analysis to synonymous polymorphisms, a lower frequency
-threshold of 0.15 was used (other thresholds yielded the same results). Since
+threshold of 0.15 was used (other thresholds yield the same results). Since
 very few polymorphisms hitchhike beyond, say, a frequency of 0.5, this pool is
@@
 
-The V loops and flanking regions were identified manually starting from the
+The V loops and flanking regions are identified manually starting from the
 annotated reference HXB2 sequence from the LANL HIV database~\citep{LANL2012}. A
-similar approach was used to label the C2-V5 region sequenced in
+similar approach is used to label the C2-V5 region sequenced in
 ref.~\citep{shankarappa_consistent_1999}.
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Computer simulations}
-Simulations were performed using the recently published software
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+Simulations are performed using the recently published software
 FFPopSim~\citep{zanini_ffpopsim:_2012}. Both full-length HIV genomes and
-\env{}-only simulations were performed and yielded comparable results. For each
-set of parameters, approximately 100 simulation runs were averaged over. In each
+\env{}-only simulations are performed and yield comparable results. For each
+set of parameters, approximately 100 simulation runs are averaged over. In each
 run, a random fitness landscape with specified statistical properties (e.g.
-density of beneficial sites, average deleterious effect of synonymous changes) was generated.
-Although the curves shown in \FIG{simfixpvar} are not very smooth, small
-parameter changes resulted in overall consistent trends across many repetitions.
+density of beneficial sites, average deleterious effect of synonymous changes) is generated.
 
 For the discussion of simulation parameters, the areas below or above the neutral
-diagonal were estimated from the binned fixation probabilities using the linear
+diagonal were estimated from the binned fixation probabilities using linear
 interpolation between the bin centers. This measure is sufficiently precise for
-our purposes, because the HIV data are quite scarse themselves.
+our purposes.
 
-\comment{All analysis scripts are available for download\ldots.}. 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Methods availability}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+All analysis scripts are available for download.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section*{Acknowledgements}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 We are grateful for stimulating discussions with Jan Albert and Trevor Bedford.

### 2013-02-03 scripts/allele_frequency_trajectories.py
added first scripts

@@ -1,0 +1,106 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       01/03/2012
+content:    Trajectories of allele frequencies.
+'''
+# Standard modules
+from operator import *
+import numpy as np
+import matplotlib.pyplot as ppl
+import matplotlib.cm as cm
+
+
+# Custom modules
+import sys
+sys.path.insert(0, '.')
+import modules.parser_Shankarappa as pS
+from modules.helper import is_nonsyn_table
+
+
+
+# Tables
+alpha = pS.alpha
+
+
+
+# Functions
+def expanded_isnongap(afs):
+    '''Restrict the sequence to non-gapped CODONS.'''
+    is_gap = (afs[:,4,:].any(axis=1)).nonzero()[0]
+    is_gap2 = set()
+    for i in is_gap:
+        is_gap2 |= set(range(i - i % 3, i - i % 3 + 3))
+    is_nongap = np.ones(afs.shape[0], bool)
+    is_nongap[list(is_gap2)] = False
+    return is_nongap
+
+
+
+# Script
+if __name__ == '__main__':
+
+    # Define the patients
+    patients = pS.parse_sequences(reference='HXB2')
+
+    for k,p in enumerate(patients[:1]):
+        print p
+
+        p.filter_only_sequenced()
+
+        # Measure the allele frequencies (of all alleles at all positions)
+        paf = p.allele_frequencies
+        afs = np.asarray([paf(seqs=p.seqs_from_visit(v)) for v in p.visit])
+
+        # Reshape so that we get the site as first axis, the nucleotide as
+        # second, the time as third
+        afs = afs.swapaxes(0,2)
+
+        # Eliminate gaps (whole codons are excluded to keep translation possible)
+        # first and translate then!
+        # Note: this has been implemented at the root
+        #is_nongap = expanded_isnongap(afs)
+        is_nongap = (afs[:,4,:] == 0).all(axis=1)
+        is_V_region = p.only_V_regions()
+
+        # Divide into syn/nonsyn to spot hitchhiking
+        consensus = alpha[afs[:,:,0].argmax(axis=1)]
+        is_nonsyn = is_nonsyn_table(consensus)
+
+        # Plot trajectories (no gaps, no conserved)
+        figsyn = ppl.figure(figsize=[9.35, 6.325])
+        fignonsyn = ppl.figure(figsize=[9.35, 6.325])
+        axsyn = figsyn.add_subplot(1,1,1)
[…]

### 2013-02-03 scripts/modules/__init__.py

@@ -1,0 +1,2 @@
+__all__ = ['helper',
+           'parser_Shankarappa']

### 2013-02-03 scripts/modules/helper.py

@@ -1,0 +1,114 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       29/03/12
+content:    Helper functions for all parsers
+'''
+# Standard modules
+import numpy as np
+import matplotlib.pyplot as ppl
+import Bio.Seq
+import re
+
+
+# Functions
+def calculate_distance(smat0, smat1):
+    dM = np.zeros((smat0.shape[0], smat1.shape[0]), int)
+    for i0, s0 in enumerate(smat0):
+        dM[i0] = (s0 != smat1).sum(axis=1)
+    return dM
+
+
+def plot_distance(dM, ax=None):
+    if ax is not None:
+        ppl.sca(ax)
+    ppl.imshow(dM, interpolation='nearest')
+    ppl.ylabel('First seqs')
+    ppl.xlabel('Second seqs')
+    ppl.title('Hamming Distance', fontsize=16)
+    
+    # set ticks to integers
+    dMmax = dM.max()
+    tickwidth = ((dMmax - 1) // 10) + 1
+    tickmax = (dMmax // tickwidth + 1) * (tickwidth)
+    ppl.colorbar(ticks=np.arange(tickmax))
+
+    ppl.ion()
+    ppl.show()
+
+
+def translate(seqdna):
+    '''Translate one or more sequences with gaps.'''
+    if np.rank(seqdna) > 1:
+        aa = []
+        for seq in seqdna:
+            aa.append(translate(seq))
+        return np.asarray(aa)
+    else:
+        tmp = seqdna.reshape((len(seqdna) / 3, 3))
+        aa = np.zeros(tmp.shape[0], str)
+        for i1 in xrange(aa.shape[0]):
+            if not (tmp[i1] == '-').any():
+                aa[i1] = Bio.Seq.translate(''.join(tmp[i1]))
+        return aa
+
+
+def is_syn_table(sequence, *args, **kwargs):
+    '''Find what loci would lead to synonymous mutations.'''
+    return -is_nonsyn_table(sequence, *args, **kwargs)
+
+
+def is_nonsyn_table(sequence, alpha=np.array(['A', 'C', 'G', 'T', '-'])):
+    '''Find what loci would lead to nonsynonymous mutations.'''
+    sequence = np.asarray(sequence)
+    codons = sequence.reshape(sequence.shape[0] / 3, 3)
+    aa = translate(sequence)
+
+    # All possible single mutants
+    is_nonsyn = np.ones((sequence.shape[0], 5), bool)
[…]

### 2013-02-03 scripts/modules/parser_Shankarappa.py

@@ -1,0 +1,659 @@
+# vim: fdm=marker:fileencoding=utf-8
+'''
+author:     Fabio Zanini
+date:       04/01/2012
+content:    Parser for various Shankarappa data files
+'''
+# Standard modules
+import re
+import os
+import glob
+import shutil
+from operator import *
+import numpy as np
+import scipy.linalg as linalg
+import matplotlib.pyplot as ppl
+import matplotlib.cm as cm
+
+from Bio import SeqIO, AlignIO
+from Bio.Seq import Seq
+from Bio.Alphabet.IUPAC import *
+from Bio.Align import MultipleSeqAlignment as MSA
+
+import longitudinal.classes.parser_SHAPE as pSHAPE
+from longitudinal.classes.helper import find_with_gaps
+from longitudinal.classes.parser_reference import seq_patterns
+from longitudinal.classes.parser_generic import patient_generic, alpha
+
+
+# Globals
+datadir = '/home/fabio/university/phd/longitudinal/data/Shankarappa'
+
+
+# Classes
+class clinical_parser(object):
+    '''Parse the TXT file with the clinical info.'''
+    def __init__(self, filename):
+        cols = [['visit', 1, int],
+                ['mo_to_SC', 2, float],
+                ['logRNA', 4, float],
+                ['CD3', 5, float],
+                ['CD8', 6, float],
+                ['CD4', 7, float]]
+
+        with open(filename, 'r') as f:
+            f.readline()    # Header
+            lsts = [l.split('\t') for l in f.readlines()]
+
+        names = set(map(itemgetter(0),lsts)) - set([''])
+        patients = [patient(p) for p in names]
+
+        for p in patients:
+            tmpls = filter(lambda x: x[0] == p.name,lsts)
+            n_points = len(tmpls)
+            tmpdict = {}
+            for n in cols:
+                tmpdict[n[0]] = np.ma.zeros(n_points, dtype=n[2])
+
+            for i,tmp in enumerate(tmpls):
+                for n in cols:
+                    try:
+                        tmpdict[n[0]][i] = tmp[n[1]]
+                    except ValueError:
+                        tmpdict[n[0]][i] = np.ma.masked
+                    
+            for n,c in tmpdict.items():
+                setattr(p,n,c)
+
+            # Try to add days_from_seroconversion
+            p.days_from_seroconversion = p.mo_to_SC * 30.5
+
+        patients.sort(key=(lambda x: int(x.name[1:])))
[…]

### 2013-02-04 scripts/allele_frequency_trajectories.py
all experimental scripts (but no alignments yet, nor remote modules)

@@ -16,11 +16,7 @@ import sys
 sys.path.insert(0, '.')
 import modules.parser_Shankarappa as pS
 from modules.helper import is_nonsyn_table
-
-
-
-# Tables
-alpha = pS.alpha
+from modules.alphabet import alpha
 
 
 

### 2013-02-04 scripts/fixation_loss_synonymous_times.py

@@ -1,0 +1,242 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       15/06/12
+content:    Calculate how many of the synonymous alleles at intermediate nu0
+            reach either boundary after a certain time.
+'''
+# Standard modules
+from sys import stderr
+import re
+from operator import *
+import numpy as np
+import matplotlib.pyplot as ppl
+import matplotlib.cm as cm
+import matplotlib.patches as patches
+
+# Custom modules
+import sys
+sys.path.insert(0, '.')
+import modules.parser_Shankarappa as pS
+from modules.helper import translate
+from modules.alphabet import alpha
+
+# Tables
+# Verbosity level (0-3)
+VERBOSE = 1
+
+# Bad patients (selected manually in a conservative way looking e.g. at PCA)
+bad_patients = ['p4', 'p7', 'p8', 'p11']
+
+# Initial frequency bins
+nu0ss = [[0.15, 0.25],
+         [0.25, 0.4],
+         [0.4, 0.6]]
+
+# Colors for the plot
+colors = ['b', 'g', 'r']
+
+# Maximal time to plot
+maxt = 40 # Months
+
+# Approximate months-to-days conversion
+mo_to_gen = 30.5
+
+
+
+# Script
+if __name__ == '__main__':
+
+    fixed = []
+    lost = []
+    fixlost = [lost, fixed]
+
+    # Counts and times (overall)
+    # The first list of each nu0s is lost, the second fixed, the third floating
+    counts_all = [[0, 0, 0] for nu0s in nu0ss]
+    # The first list of each nu0s is lost, the second fixed
+    times_all = [[[], []] for nu0s in nu0ss]
+    
+    # Define the patients
+    patients = pS.parse_sequences(exclude=bad_patients)
+    
+    # Aggregate information from all patients
+    for k, p in enumerate(patients):
+
+        if VERBOSE >= 1:
+            stderr.write(str(p)+'\n')
+
+        # Filter the time points to sequenced times
+        p.filter_only_sequenced()
+        n = len(p.visit)
+    
+        # Filter away gaps (conservative), keep reading frame
+        is_nongap = ((np.array(p.seqs) == '-').sum(axis=0) == 0)
+        for i in xrange(len(is_nongap)):
+            if not is_nongap[i]:
+                is_nongap[i - i % 3: i - i % 3 + 3] = False
+    
+        # Collect allele frequencies
[…]

### 2013-02-04 scripts/fixation_probability.py

@@ -1,0 +1,281 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       15/06/12
+content:    Calculate how many of the synonymous alleles at intermediate nu0
+            reach either boundary after a certain time.
+'''
+# Standard modules
+from sys import stderr
+import re
+from operator import *
+import numpy as np
+import matplotlib.pyplot as ppl
+import matplotlib.cm as cm
+import matplotlib.patches as patches
+import Bio
+
+# Custom modules
+import sys
+sys.path.insert(0, '.')
+import modules.parser_Shankarappa as pS
+import modules.parser_Bunnik2008 as pB
+from modules.helper import translate
+from modules.alphabet import alpha
+
+
+
+# Globals
+# Verbosity level (0-3)
+VERBOSE = 1
+
+# Initial frequency bins
+nu0ss = [[0.05, 0.15],
+         [0.15, 0.25],
+         [0.25, 0.4],
+         [0.4, 0.6],
+         [0.6, 0.8]]
+
+# Bad patients (selected manually in a conservative way looking e.g. at PCA)
+bad_patients = {'Bunnik': ['ACH19542', 'ACH19768'],
+                'Shankarappa': ['p4', 'p7', 'p8', 'p11']}
+
+# Classes of mutations to be shown
+classes = ['syn_Shanka',
+           'syn_nonShanka',
+           'nonsyn']
+
+# Classes of mutations including script intermediates
+classes_all = ['nonsyn',
+               'syn',
+               'syn_Shanka',
+               'syn_nonShanka']
+
+# Colors for the plot
+colors = {'syn_Shanka': 'red',
+          'syn_nonShanka': 'green',
+          'nonsyn': 'blue'}
+
+# Labels for the plot
+labels = {'syn_Shanka': 'synonymous, C2-V5',
+          'syn_nonShanka': 'synonymous, other',
+          'nonsyn': 'nonsynonymous'}
+
+# Distance from last time point to be considered likely to reach either boundary
+# before the sequencing stops (later sequences will probably stay floating).
+# Note: this is not used in the plot (only for testing).
+maxt = {'Shankarappa': 40,  # Months
+        'Bunnik': 1200}     # Days
+
+
+
+# Script
+if __name__ == '__main__':
+
+    # Define the patients (excluding problematic ones)
+    patientsB = pB.parse_sequences(reference='SHAPE', exclude=bad_patients['Bunnik'])
[…]

### 2013-02-04 scripts/fixation_probability_Vloops_flanking.py

@@ -1,0 +1,268 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       15/06/12
+content:    Calculate the fixation probability of synonymous derived alleles in
+            the V loops and flanking regions.
+'''
+# Standard modules
+from sys import stderr
+import re
+from operator import *
+import numpy as np
+import matplotlib.pyplot as ppl
+import matplotlib.cm as cm
+import matplotlib.patches as patches
+import Bio
+
+# Custom modules
+import sys
+sys.path.insert(0, '.')
+import modules.parser_Shankarappa as pS
+import modules.parser_Bunnik2008 as pB
+from modules.helper import translate
+from modules.alphabet import alpha
+
+
+
+# Globals
+# Verbosity level (0-3)
+VERBOSE = 1
+
+# Initial frequency bins
+nu0ss = [[0.05, 0.15],
+         [0.15, 0.25],
+         [0.25, 0.4],
+         [0.4, 0.6],
+         [0.6, 0.8]]
+
+# Bad patients (selected manually in a conservative way looking e.g. at PCA)
+bad_patients = {'Bunnik': ['ACH19542', 'ACH19768'],
+                'Shankarappa': ['p4', 'p7', 'p8', 'p11']}
+
+# Classes of mutations
+classes = ['loops',
+               'flank']
+
+# Colors for the plot
+colors = {'loops': 'red',
+          'flank': 'blue'}
+
+# Labels for the plot
+labels = {'loops': 'V3-V5 loops',
+          'flank': 'flanking stems'}
+
+# Distance from last time point to be considered likely to reach either boundary
+# before the sequencing stops (later sequences will probably stay floating).
+# Note: this is not used in the plot (only for testing).
+maxt = {'Shankarappa': 40,  # Months
+        'Bunnik': 1200}     # Days
+
+
+
+# Script
+if __name__ == '__main__':
+
+    # Define the patients (excluding problematic ones)
+    patientsB = pB.parse_sequences(reference='SHAPE', exclude=bad_patients['Bunnik'])
+    patientsS = pS.parse_sequences(reference='SHAPE', exclude=bad_patients['Shankarappa'])
+    patients = patientsB + patientsS
+
+    # Counts (overall)
+    # The first list of each nu0s is lost, the second fixed, the third floating
+    # Moreover, record the patient number they came from
+    counts = {x: np.zeros((len(nu0ss), 3, len(patients)), int) for x in classes}
+
[…]

### 2013-02-04 scripts/modules/__init__.py

@@ -1,2 +1,8 @@
-__all__ = ['helper',
-           'parser_Shankarappa']
+__all__ = [
+           'helper',
+           'alphabet',
+           'secondary',
+           'parser_Shankarappa',
+           'parser_Bunnik2008',
+           'parser_LANL',
+          ]

### 2013-02-04 scripts/modules/parser_Bunnik2008.py

@@ -1,0 +1,110 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       25/07/12
+content:    Parser for Bunnik2008 (subparser of LANL)
+'''
+# Standard modules
+from operator import *
+import numpy as np
+
+from Bio.Alphabet.IUPAC import *
+
+from longitudinal.classes.parser_LANL import patient as patientLANL
+from longitudinal.classes.parser_LANL import parse_sequences as parse_sequencesLANL
+from longitudinal.classes.helper import translate, find_with_gaps
+from longitudinal.classes.parser_reference import seq_patterns
+
+
+# Globals
+# Second rev exon in env (tat is included)
+rev2 = {'ACH19999': ['ACCCACCTCCCA', 'TGCTGTTAGCTTG'],
+        'ACH18969': ['ACCCGCCTCCCA', 'TGCTGTTAGCTTG'],
+        'ACH19659': ['ACCCGCCTCCCA', 'TGCTGTTAGCTTG']}
+
+
+
+
+# Classes
+class patient(patientLANL):
+    '''Specialized version of patient from parser_LANL.'''
+
+    aaS = ['EEE', 'RPGGG']
+    '''Shankarappa region'''
+
+    def filter_for_longitudinal_genetics(self):
+        '''Filter only time points with sequences.'''
+        self.filter_seqs_with_attribute('days_from_seroconversion')
+
+
+    def only_Shanka(self):
+        '''Denote the Shankarappa region.
+ 
+        Returns:
+            index: boolean index with the Shankarappa region marked by 'True's.
+
+        *Note*: The region is marked by 'EEE' at the beginning and 'RPGGG' at the end.
+        '''
+
+        cons = self.get_consensus_initial()
+        is_Shanka = np.zeros(len(cons), bool)
+        consaa = ''.join(translate(cons))
+        is_Shanka[consaa.find(self.aaS[0]) * 3: consaa.find(self.aaS[1]) * 3] = True
+        return is_Shanka
+
+
+    def only_rev2(self):
+        '''Denote the second rev exon.
+
+        Returns:
+            index: boolean index with the Shankarappa region marked by 'True's.
+        '''
+        cons = ''.join(self.get_consensus_initial())
+        is_rev2 = np.zeros(len(cons), bool)
+        is_rev2[cons.find(rev2[self.name][0]): cons.find(rev2[self.name][1])] = True
+        return is_rev2
+
+
+    def find_V_region(self, name):
[…]

### 2013-02-04 scripts/modules/parser_LANL.py

@@ -1,0 +1,782 @@
+# vim: fdm=marker
+'''
+author:     Fabio Zanini
+date:       10/02/2012
+content:    Parser for the various files downloaded from the LANL HIV database.
+'''
+# Standard modules
+import os
+import re
+import glob
+from operator import *
+import numpy as np
+import scipy.linalg as linalg
+import matplotlib.pyplot as ppl
+import matplotlib.cm as cm
+
+from Bio import SeqIO, AlignIO
+from Bio.Seq import Seq
+from Bio.SeqRecord import SeqRecord
+from Bio.Alphabet.IUPAC import *
+
+import longitudinal.classes.parser_SHAPE as pSHAPE
+from longitudinal.classes.parser_generic import patient_generic, alpha
+
+
+
+# Globals
+datadir = '/home/fabio/university/phd/longitudinal/data/LANL'
+
+columns = [('#',int),
+           ('patient_code',str),
+           ('patient_id',str),
+           ('accession',str),
+           ('name',str),
+           ('subtype',str),
+           ('country',str),
+           ('sampling_year',int),
+           ('days_from_first_sample',int),
+           ('Fiebig_stage',str),
+           ('days_from_treatment_end',int),
+           ('n_timepoints',int),
+           ('days_from_treatment_start',int),
+           ('days_from_infection',int),
+           ('days_from_seroconversion',int),
+           ('pubmed_ID',str),
+           ('start',int),
+           ('stop',int),
+           ('sequence_length',int),
+           ('organism',str),
+           ('sequence',str)]
+
+
+# Classes
+class patient(patient_generic):
+    def __init__(self,name):
+        self.name = name
+        self.visit = []
+        self.seqs = []
+
+    def __call__(self):
+        for a in vars(self).keys():
+                if a != 'seqs':
+                    print a
+                    print getattr(self,a)
+                elif self.seqs:
+                    print a
+                    print self.seqs[0]
+                    print '[...]'
+                print
+
+
+    def visit_to_time(self,visitnumber):
+        attr = filter(lambda x: 'days' in x, vars(self).keys())[0]
+        return getattr(self, attr)[self.visit == visitnumber][0]
+
+
+    def time_to_visit(self,days):
+        attr = filter(lambda x: 'days' in x, vars(self).keys())[0]
+        return self.visit[getattr(self, attr) == days][0]
+
+
+    def seqs_from_visit(self,visitnumbers):
[…]

### 2013-02-04 supplementary.tex
PCA of good patient in suppl

@@
 \begin{center}
-\includegraphics[width=0.48\linewidth]{Shankarappa_PCA_p7}
-\includegraphics[width=0.48\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p7}
-\caption{Left: PCA of all sequences from patient p7 (colors indicate time from seroconversion,
-from blue to red). Right: Allele frequency trajectories for nonsynonymous
-changes in the same patient.}
-\label{fig:aftp7}
+\includegraphics[width=0.4\linewidth]{Shankarappa_PCA_p1}
+\includegraphics[width=0.4\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p1}
+\includegraphics[width=0.4\linewidth]{Shankarappa_PCA_p7}
+\includegraphics[width=0.4\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p7}
+\caption{Panel A) PCA of all sequences from patient p1 (colors indicate time from seroconversion,
+from blue to red). Panel B) Allele frequency trajectories for nonsynonymous
+changes in the same patient. Panels C and D) the same plots from patient p7, whose
+sequences show signs of compartimentalization.}
+\label{fig:aftp}
 \end{center}
@@
 PCA is performed on all sequences from a ``compartimentalized'' patient, p7.
-This is shown in \figurename~\ref{fig:aftp7}: late sequences subdivide in two
+This is shown in \figurename~\ref{fig:aftp}: late sequences subdivide in two
 well-defined clusters -- this characteristic is absent from all the patient
@@
 of mutants within the same epitope).
-
-\clearpage
 

### 2013-02-04 synmut.tex
all experimental scripts (but no alignments yet, nor remote modules)
PCA of good patient in suppl

@@
 \begin{center}
-\subfloat{\includegraphics[width=0.9\linewidth]{Shankarappa_fix_loss_dt_times}
+\subfloat{\includegraphics[width=0.9\linewidth]{fixation_times}
 \label{fig:fixp1}}\\
@@
 effect of synonymous mutations, the fraction of synonymous mutations that are
-deleterious, the fitness effect of adaptive mutations (escape rate) and the frequency of
-escapes. Among all simulations, we select the ones that show $A_\text{syn}$ and
+deleterious, the escape rate and the frequency of new escapes. Note that the
+escape rate is sum of two factors: (a) the beneficial effect due to the ability to
+evade the immune system minus (b) the fitness side costs in terms of structure,
+stability, etc.
+Among all simulations, we select the ones that show $A_\text{syn}$ and
 $P_\text{interm}$ as observed in the data, i.e., a large depression in fixation

### 2013-02-04 scripts/allele_frequency_trajectories.py
simulation scripts

@@ -20,19 +20,6 @@ from modules.alphabet import alpha
 
 
 
-# Functions
-def expanded_isnongap(afs):
-    '''Restrict the sequence to non-gapped CODONS.'''
-    is_gap = (afs[:,4,:].any(axis=1)).nonzero()[0]
-    is_gap2 = set()
-    for i in is_gap:
-        is_gap2 |= set(range(i - i % 3, i - i % 3 + 3))
-    is_nongap = np.ones(afs.shape[0], bool)
-    is_nongap[list(is_gap2)] = False
-    return is_nongap
-
-
-
 # Script
 if __name__ == '__main__':
 
@@ -55,7 +42,6 @@ if __name__ == '__main__':
         # Eliminate gaps (whole codons are excluded to keep translation possible)
         # first and translate then!
         # Note: this has been implemented at the root
-        #is_nongap = expanded_isnongap(afs)
         is_nongap = (afs[:,4,:] == 0).all(axis=1)
         is_V_region = p.only_V_regions()
 

### 2013-02-04 scripts/simulations_plot_gradually_epitopesandtimeselec.py

@@ -1,0 +1,106 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       22/01/13
+content:    Plot the fixation probability of increasingly deleterious sims and 
+            allele freqs trajectories of some of its simulations.
+'''
+# Modules
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib.patches import Rectangle
+import matplotlib.cm as cm
+
+
+
+# Globals
+datafile = 'parameters_fixation_areas.dat'
+
+
+
+# Functions
+def filterfolders(folders):
+    '''Filter and sort folders'''
+    tmp = folders[0].split('_')
+    if ('nescepi' in tmp) or ('timeselec' in tmp):
+        tmp = tmp[:-2]
+    base = '_'.join(tmp)
+    print tmp
+    print base
+
+    ind = {}
+    for i, f in enumerate(folders):
+        if 'nescepi_3' in f:
+            ind[1] = i
+        elif 'nescepi_6' in f:
+            ind[2] = i
+        elif 'timeselec_0.005' in f:
+            ind[3] = i
+        elif 'timeselec_0.01' in f:
+            ind[4] = i
+        elif ('nescepi' not in f) and ('timeselec' not in f):
+            ind[0] = i
+    print folders
+    print ind
+    return [folders[ind[i]] for i in xrange(5)]
+
+
+
+# Script
+if __name__ == '__main__':
+
+
+    # Input results dir
+    args = sys.argv
+    if len(args) < 2:
+        raise ValueError('Please specify a results folder')
+    results_dir = args[1].rstrip('/')+'/'
+
+    # Folders
+    g = os.walk(results_dir)
+    folders = np.array(g.next()[1])
+
+    # Sort them smartly
+    folders = filterfolders(folders)
+    labels = ['naive', '3 mut/epi', '6 mut/epi', 'tds 0.005', 'tds 0.01']
+    colors = ['b', 'g', 'r', 'cyan', 'magenta']
+
+    # Prepare figure
+    fig = plt.figure()
+
+    # Get the data
+    for i, folder in enumerate(folders):
+
+        fn = results_dir+folder
+        nu0, nu1, Psyn, Pnonsyn, csyn, cnonsyn = np.genfromtxt(fn+'/Pfix.dat',
+                                                               unpack=True,
+                                                               usemask=True,
[…]

### 2013-02-04 scripts/simulations_plot_graduallydel.py

@@ -1,0 +1,125 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       22/01/13
+content:    Plot the fixation probability of increasingly deleterious sims and 
+            allele freqs trajectories of some of its simulations.
+'''
+# Modules
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib.patches import Rectangle
+import matplotlib.cm as cm
+
+
+
+# Globals
+datafile = 'parameters_fixation_areas.dat'
+
+
+
+# Script
+if __name__ == '__main__':
+
+
+    # Input results dir
+    args = sys.argv
+    if len(args) < 2:
+        raise ValueError('Please specify a results folder')
+    results_dir = args[1].rstrip('/')+'/'
+
+    # Load data using record arrays
+    tb = np.genfromtxt(results_dir+datafile, unpack=False,
+                       dtype=zip(('del_eff', 'ada_eff',
+                                  'del_frac', 'ada_rate',
+                                  'trashed', 'good',
+                                  'Asyn', 'Anonsyn',
+                                  'count_syn', 'counts_nonsyn',
+                                  'subs_syn', 'subs_nonsyn',
+                                  'poly_syn', 'poly_nonsyn',
+                                  'foldername'),
+                                 (float, float,
+                                  float, float,
+                                  int, int,
+                                  float, float,
+                                  float, float,
+                                  float, float,
+                                  float, float,
+                                  'S200')))
+
+    # Filter out NaNs
+    is_nonnan = -np.isnan(tb['Asyn'])
+    tb = tb[is_nonnan]
+    ind = np.argsort(tb['del_eff'])
+
+    # Filter only some parameters
+    ind = [ind[1], ind[2], ind[4]]
+
+    # Limits of deleff
+    colors = ['b', 'g', 'r', 'cyan', 'magenta']
+    labels = [r'$10^{-3.5}$', r'$10^{-3}$', r'$10^{-2.5}$']
+
+    # Prepare figure
+    fig = plt.figure()
+
+    # Get the data
+    for ii, i in enumerate(ind):
+
+        folder = tb['foldername'][i]
+        fn = results_dir+folder
[…]

### 2013-02-04 scripts/simulations_scatter_all.py

@@ -1,0 +1,113 @@
+# vim: fdm=marker
+'''
+author:     Fabio Zanini
+date:       14/01/13
+content:    Scatter plot and further visualization of the results of random
+            gradual simulations.
+'''
+# Modules
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib.patches import Rectangle
+import matplotlib.cm as cm
+
+
+
+# Globals
+datafile = 'parameters_fixation_areas.dat'
+
+# Set limits for data-like sims
+Asynmin = -0.5
+Asynmax = -0.15
+poly_synmin = 0.0025
+poly_synmax = 0.010
+
+# (nonsyn gates, if only nonsyn data-like results are wished)
+Anonsynmax = 0.5
+Anonsynmin = -0.5
+tmpsnsmin = 0e-6
+
+
+# Script
+if __name__ == '__main__':
+
+
+    # Input results dir
+    args = sys.argv
+    if len(args) < 2:
+        raise ValueError('Please specify a results folder')
+    results_dir = args[1].rstrip('/')+'/'
+    
+
+
+    # Load data using record arrays
+    tb = np.genfromtxt(results_dir+datafile, unpack=False,
+                       dtype=zip(('del_eff', 'ada_eff',
+                                  'del_frac', 'ada_rate',
+                                  'trashed', 'good',
+                                  'Asyn', 'Anonsyn',
+                                  'count_syn', 'counts_nonsyn',
+                                  'subs_syn', 'subs_nonsyn',
+                                  'poly_syn', 'poly_nonsyn',
+                                  'foldername'),
+                                 (float, float,
+                                  float, float,
+                                  int, int,
+                                  float, float,
+                                  float, float,
+                                  float, float,
+                                  float, float,
+                                  'S200')))
+
+    # Filter out NaNs
+    is_nonnan = -np.isnan(tb['Asyn'])
+    tb = tb[is_nonnan]
+
+    # Select only configurations with the right output (Asyn in the right
+    # window)
+    is_good = (tb['Asyn'] < Asynmax) & (tb['Asyn'] > Asynmin)
+
[…]

### 2013-02-04 scripts/codon_entropy_synonymous_subtypeB.py
scripts for supplementary

@@ -1,0 +1,139 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       07/12/12
+content:    Measure the level of conservation of synonymous mutations in subtype
+            B.
+'''
+# Modules
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib.patches import Rectangle
+
+# Custom modules
+import sys
+sys.path.insert(0, '.')
+import modules.alignment as alignment
+from modules.alphabet import alpha
+from modules.codon_usage import number_of_codons
+from modules.helper import translate
+from modules.parser_reference import seq_patterns, find_with_gaps
+
+
+
+# Globals
+genes = ['gag', 'pol', 'env', 'nef']
+reference = 'SHAPE'
+seq_patterns = seq_patterns[reference]
+
+
+
+# Functions
+def smoothen(x, f):
+    l = len(x)
+    x = x[:l - l % f]
+    x = x.reshape((len(x) / f, f)).mean(axis=1)
+    return x
+
+
+
+# Script
+if __name__ == '__main__':
+
+    # Load data for a certain gene
+    gene = genes[0]
+    ali = alignment.MSA(gene=gene, reference=reference)
+    msa = np.array(ali.MSA)
+    afs = ali.allele_frequencies()
+
+    # Filter only part of pol
+    if gene == 'pol':
+        msa = msa[:, :2718]
+        afs = afs[:, :2718]
+
+    # Find consensus sequence
+    consind = afs.argmax(axis=0)
+    consensus = alpha[consind]
+
+    # Some seqs are not viable and include frameshifts that mess up the
+    # translation, hence we restrict to positions for which gaps are a minority
+    is_gap = consensus == '-'
+    # Exclude full codons
+    tmp = np.unique(is_gap.nonzero()[0] / 3)
+    is_gap[np.concatenate([tmp * 3, tmp * 3 +1, tmp * 3 + 2])] = True
+
+    # Exclude stop codons
+    is_stop = np.zeros_like(is_gap)
+    tmp = (translate(consensus) == '*').nonzero()[0]
+    is_stop[np.concatenate([tmp * 3, tmp * 3 +1, tmp * 3 + 2])] = True
+
+    ## Plot base prevalence
+    #for i in xrange(4):
+    #    plt.plot(np.arange(len(consensus)), afs[i],
+    #             lw=1.5, alpha=0.5)
[…]

### 2013-02-04 scripts/conservation_syn_nonsyn_subtypeB.py

@@ -1,0 +1,168 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       07/12/12
+content:    Measure the level of conservation of synonymous and nonsynonymous 
+            mutations in subtype B. Nonsynonymous mutations are expected to be
+            more broadly conserved, because they impair protein function.
+'''
+# Modules
+import sys
+from collections import Counter
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib.patches import Rectangle
+
+# Custom modules
+import modules.alignment as alignment
+from modules.alphabet import alpha, alphal_ng
+from modules.codon_usage import all_codons
+from modules.codon_usage import translate as transstrict
+from modules.helper import translate
+from modules.parser_reference import seq_patterns, find_with_gaps
+
+
+
+# Globals
+gene = 'gag'
+reference = 'SHAPE'
+seq_patterns = seq_patterns[reference]
+
+
+
+# Functions
+def smooth(x, f):
+    l = len(x)
+    x = x[:l - l % f]
+    x = x.reshape((len(x) / f, f)).mean(axis=1)
+    return x
+
+
+def codon_single_mutants_synnonsyn(codon):
+    codon = list(codon)
+    aa = transstrict(''.join(codon))
+    syn = []
+    nonsyn = []
+    for pos in xrange(3):
+        for mut in alphal_ng:
+            if mut != codon[pos]:
+                tmp = list(codon)
+                tmp[pos] = mut
+                tmp = ''.join(tmp)
+                if transstrict(tmp) == aa:
+                    syn.append(tmp)
+                else:
+                    nonsyn.append(tmp)
+    return {'syn': syn, 'nonsyn': nonsyn}
+
+
+
+
+# Script
+if __name__ == '__main__':
+
+    # Load data for a certain gene
+    ali = alignment.MSA(gene=gene, reference=reference)
+    msaraw = np.array(ali.MSA)
+    afs = ali.allele_frequencies()
+
+    # Filter only part of pol
+    if gene == 'pol':
+        msaraw = msaraw[:, :2718]
+        afs = afs[:, :2718]
+
+    # Find consensus sequence
+    consind = afs.argmax(axis=0)
+    consensus = alpha[consind]
+
[…]

### 2013-02-04 scripts/modules/__init__.py

@@ -1,7 +1,9 @@
-__all__ = [
-           'helper',
+__all__ = ['alignment',
            'alphabet',
+           'codon_usage',
+           'helper',
            'secondary',
+           'parser_reference',
            'parser_Shankarappa',
            'parser_Bunnik2008',
            'parser_LANL',

### 2013-02-04 scripts/modules/alignment.py

@@ -1,0 +1,82 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       12/11/12
+content:    Tools for managing sequences and alignments of general nature.
+'''
+# Modules
+import os
+import glob
+import numpy as np
+from Bio import SeqIO, AlignIO
+import general.classes.alphabet as abc
+
+
+# Globals
+datadir = '/home/fabio/university/phd/general/data/'
+genes = ['gag', 'pol', 'env', 'nef', 'vpu', 'vpr', 'vif', 'tat']
+
+
+
+# Classes
+class MSA(object):
+    '''Multiple sequence alignment with general properties'''
+    def __repr__(self):
+        s = 'MSA(kind="'+self.kind+'", gene="'+self.gene+'"'
+        if hasattr(self, 'reference'):
+            s += ', reference="'+self.reference+'"'
+        s += ')'
+        return s
+
+    def __init__(self, kind='subtypeB', reference=None, gene=None):
+
+        kinddir = (datadir + kind).rstrip('/') + '/'
+        self.kind = kind.rstrip('/')
+        if reference is not None:
+            kinddir = (kinddir + 'to_reference/')
+            self.reference = reference
+        if not os.path.isdir(kinddir):
+            raise IOError('Directory not found')
+
+        found = False
+        for fn in glob.glob(kinddir+'*.fasta'):
+            if gene in fn:
+                MSA = AlignIO.read(fn, 'fasta')
+                if reference is None:
+                    self.MSA = MSA
+                else:
+                    self.reference_seq = MSA[0]
+                    if reference not in self.reference_seq.id:
+                        raise ValueError('Reference name not found in first sequence of the MSA')
+                    self.MSA = MSA[1:]
+                self.gene = gene
+                found = True
+                break
+        if not found:
+            raise IOError('Alignment file not found')
+
+
+    def allele_frequencies(self, alpha=abc.alpha):
+        '''Get the allele frequencies for an alphabet
+        
+        Parameters:
+            - alpha: an alphabet
+        '''
+        alpha = list(alpha)
+        L = len(self.MSA[0])
+        afs = np.zeros((len(alpha), L))
+        mat = np.array(self.MSA)
+        for i, a in enumerate(alpha):
+            afs[i] = (mat == a).sum(axis=0)
+        afs /= afs.sum(axis=0)
+
+        return afs
+
+
+
+
+# Script
+if __name__ == '__main__':
+
+    msa = MSA('subtypeB', gene='nef')
+

### 2013-02-04 scripts/modules/codon_usage.py

@@ -1,0 +1,143 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       16/10/12
+content:    Read codon usage tables
+'''
+# Modules
+import re
+from longitudinal.classes.parser_generic import alpha
+from Bio.Seq import translate
+
+
+
+# Globals
+datadir = '/home/fabio/university/phd/general/data/codon_usage/'
+alpha = list(alpha)
+
+
+
+# Classes
+class codon_table(object):
+
+    def __init__(self, organism='homo sapiens'):
+
+        self.organism = organism
+        self.datafile = datadir + re.sub(' ', '_', organism) +'.dat'
+        
+        table_within = {}
+        table_abs = {}
+        with open(self.datafile, 'r') as f:
+            for line in f:
+                if line[0] != '#':
+                    fields = line.rstrip('\n').split('\t')
+                    key = fields[0]
+                    aa = fields[1]
+                    freqwithin = float(fields[2])
+                    freqabs = 1e-3 * float(fields[3])
+
+                    table_within[key] = (aa, freqwithin)
+                    table_abs[key] = freqabs
+
+        self.within = table_within
+        self.absolute = table_abs
+
+
+
+# Functions
+def all_codons():
+    '''Generate a list with all codons'''
+    return ('AAA', 'AAC', 'AAG', 'AAT', 'ACA', 'ACC',
+            'ACG', 'ACT', 'AGA', 'AGC', 'AGG', 'AGT',
+            'ATA', 'ATC', 'ATG', 'ATT', 'CAA', 'CAC',
+            'CAG', 'CAT', 'CCA', 'CCC', 'CCG', 'CCT',
+            'CGA', 'CGC', 'CGG', 'CGT', 'CTA', 'CTC',
+            'CTG', 'CTT', 'GAA', 'GAC', 'GAG', 'GAT',
+            'GCA', 'GCC', 'GCG', 'GCT', 'GGA', 'GGC',
+            'GGG', 'GGT', 'GTA', 'GTC', 'GTG', 'GTT',
+            'TAA', 'TAC', 'TAG', 'TAT', 'TCA', 'TCC',
+            'TCG', 'TCT', 'TGA', 'TGC', 'TGG', 'TGT',
+            'TTA', 'TTC', 'TTG', 'TTT')
+
+
+def all_amino_acids():
+    '''All amino acids'''
+    return ('A', 'C', 'E', 'D', 'G', 'F',
+            'I', 'H', 'K', '*', 'M', 'L',
[…]

### 2013-02-04 scripts/modules/parser_reference.py

@@ -1,0 +1,109 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       04/09/12
+content:    Implement some globals used by many parsers that interact with a
+            reference.
+'''
+# Modules
+import numpy as np
+import Bio.AlignIO as AlignIO
+from itertools import permutations
+
+from longitudinal.classes.helper import find_with_gaps
+from longitudinal.classes.parser_generic import alpha
+
+
+
+# Globals
+# these are derived from each other using MSA.corresponding_pattern
+# (and ultimately by hand from the horrible XLS file for HXB2)
+seq_patterns = {'SHAPE':{'V1': ['TGCACTGATTTGAA', 'TGCTCTTTCAATATCAGC'],
+                         'V2': ['TGCTCTTTCAATATCAGC', 'TGTAACACCTCAGTCAT'],
+                         'V3': ['TGTACAAGACCCAACAACA', 'TGTAACATTAGTAGA'],
+                         'V4': ['TGTAATTCAACACAACTGTTTAA', 'TGCAGAATAAAACAATT'],
+                         'V5': ['TAATAACAACAATGGGTCCGAGAT', 'CCTGGAGGA'],
+                         'env_signal': ['ATGAGAGTGAAGGAGAA', 'ACAGAAAAATTGTGGGTC']},
+                'NL4-3':{'V1': ['TGCACTGATTTGAA', 'TGCTCTTTCAATATCAGC'],
+                         'V2': ['TGCTCTTTCAATATCAGC', 'TGTAACACCTCAGTCAT'],
+                         'V3': ['TGTACAAGACCCAACAACA', 'TGTAACATTAGTAGA'],
+                         'V4': ['TGTAATTCAACACAACTGTTTAA', 'TGCAGAATAAAACAATT'],
+                         'V5': ['TAATAACAACAATGGGTCCGAGAT', 'CCTGGAGGA']},
+                'HXB2': {'V1': ['TGCACTGATTTGAA', 'TGCTCTTTCAATATCAGC'],
+                         'V2': ['TGCTCTTTCAATATCAGC', 'TGTAACACCTCAGTCAT'],
+                         'V3': ['TGTACAAGACCCAACAACA', 'TGTAACATTAGTAGA'],
+                         'V4': ['TGTAATTCAACACAACTGTTTAA', 'TGCAGAATAAAACAAAT'],
[…]

### 2013-02-04 data/longitudinal/Shankarappa/initial_consensus/split.py
alignments added

@@ -1,0 +1,18 @@
+import re
+import numpy as np
+
+with open('nuc_1-11_aligned.dbn', 'r') as f:
+    for i, line in enumerate(f):
+        if i % 3 == 0:
+            tmp = []
+        tmp.append(line)
+        if i % 3 == 2:
+            pname = re.sub(' ', '', tmp[0][2:5])
+
+#            # Try to feed the gaps into jViz
+#            tmp[1] = re.sub('-', 'A', tmp[1])
+#            tmp[2] = ''.join([tmp[2][i] for i, p in enumerate(tmp[1]) if p != '-'])
+#            tmp[1] = ''.join([p for i, p in enumerate(tmp[1]) if p != '-'])
+            with open('nuc_'+pname+'.dbn', 'w') as ff:
+                for lline in tmp:
+                    ff.write(lline)

### 2013-02-04 scripts/modules/__init__.py

@@ -4,6 +4,8 @@ __all__ = ['alignment',
            'helper',
            'secondary',
            'parser_reference',
+           'parser_generic',
+           'parser_SHAPE',
            'parser_Shankarappa',
            'parser_Bunnik2008',
            'parser_LANL',

### 2013-02-04 scripts/modules/alignment.py

@@ -13,7 +13,7 @@ import general.classes.alphabet as abc
 
 
 # Globals
-datadir = '/home/fabio/university/phd/general/data/'
+datadir = '../data/'
 genes = ['gag', 'pol', 'env', 'nef', 'vpu', 'vpr', 'vif', 'tat']
 
 

### 2013-02-04 scripts/modules/codon_usage.py

@@ -12,7 +12,7 @@ from Bio.Seq import translate
 
 
 # Globals
-datadir = '/home/fabio/university/phd/general/data/codon_usage/'
+datadir = '../data/codon_usage/'
 alpha = list(alpha)
 
 

### 2013-02-04 scripts/modules/parser_Bunnik2008.py

@@ -10,10 +10,10 @@ import numpy as np
 
 from Bio.Alphabet.IUPAC import *
 
-from longitudinal.classes.parser_LANL import patient as patientLANL
-from longitudinal.classes.parser_LANL import parse_sequences as parse_sequencesLANL
-from longitudinal.classes.helper import translate, find_with_gaps
-from longitudinal.classes.parser_reference import seq_patterns
+from parser_LANL import patient as patientLANL
+from parser_LANL import parse_sequences as parse_sequencesLANL
+from helper import translate, find_with_gaps
+from parser_reference import seq_patterns
 
 
 # Globals

### 2013-02-04 scripts/modules/parser_LANL.py

@@ -19,13 +19,13 @@ from Bio.Seq import Seq
 from Bio.SeqRecord import SeqRecord
 from Bio.Alphabet.IUPAC import *
 
-import longitudinal.classes.parser_SHAPE as pSHAPE
-from longitudinal.classes.parser_generic import patient_generic, alpha
+import parser_SHAPE as pSHAPE
+from parser_generic import patient_generic, alpha
 
 
 
 # Globals
-datadir = '/home/fabio/university/phd/longitudinal/data/LANL'
+datadir = '../data/longitudinal/LANL'
 
 columns = [('#',int),
            ('patient_code',str),

### 2013-02-04 scripts/modules/parser_SHAPE.py

@@ -1,0 +1,205 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       03/07/12
+content:    Parser for the SHAPE reactivity data, from doi:10.1038/nature08237.
+'''
+# Modules
+import numpy as np
+import re
+
+
+# Globals
+helixfile = '/home/fabio/university/phd/longitudinal/data/SHAPE/nature08237-s2.txt'
+bpseqfile = '/home/fabio/university/phd/longitudinal/data/SHAPE/nature08237-s2_bpseq.txt'
+ppfile = '/home/fabio/university/phd/longitudinal/data/SHAPE/nature08237-s3.txt'
+
+
+
+# Classes
+class secondary(object):
+    '''Store the SHAPE reactivities and pairing probs.'''
+
+    Shankarappa_indices = [6558, 7173]
+    
+
+    def __init__(self):
+        '''Store sequence, reactivities, pairing probs, and map.'''
+        self.r, self.p = np.genfromtxt(ppfile,
+                                       usecols=(2,3),
+                                       delimiter='\t',
+                                       unpack=True,
+                                       usemask=True)
+    
+        self.sequence = np.genfromtxt(ppfile,
+                                      dtype='S1',
+                                      usecols=[1],
+                                      delimiter='\t',
+                                      unpack=True,
+                                      usemask=False)
+
+        # Convert to DNA notation
+        self.sequence[self.sequence == 'U'] = 'T'
+
+        # Save map in a masked array
+        bp1, bp2 = np.loadtxt(bpseqfile, unpack=True, usecols=(0,2), dtype=int)
+        self.map = np.ma.masked_all(len(self.sequence), int)
+        self.map[bp1 - 1] = bp2 - 1
+
+
+    def __repr__(self):
+        return 'secondary structure class of HIV-1 NL4-3'
+
+
+
+# Functions
+def assign_prob(seq, include_reactivity=False):
+    '''Create a masked array with pairing probability and reactivity.
+    
+    Gap positions in the input sequence will stay masked.
+    Sites where the reactivity has not been measured will stay masked as well.
+    '''
+    # Create secondary structure object
+    sec = secondary()
+    s = ''.join(sec.sequence)
+
+    # Initialize
+    assp = np.ma.masked_all(len(seq))
+    assr = np.ma.masked_all(len(seq))
+
[…]

### 2013-02-04 scripts/modules/parser_Shankarappa.py

@@ -20,14 +20,16 @@ from Bio.Seq import Seq
 from Bio.Alphabet.IUPAC import *
 from Bio.Align import MultipleSeqAlignment as MSA
 
-import longitudinal.classes.parser_SHAPE as pSHAPE
-from longitudinal.classes.helper import find_with_gaps
-from longitudinal.classes.parser_reference import seq_patterns
-from longitudinal.classes.parser_generic import patient_generic, alpha
+import parser_SHAPE as pSHAPE
+from helper import find_with_gaps
+from parser_reference import seq_patterns
+from parser_generic import patient_generic, alpha
+
 
 
 # Globals
-datadir = '/home/fabio/university/phd/longitudinal/data/Shankarappa'
+datadir = '../data/longitudinal/Shankarappa'
+
 
 
 # Classes

### 2013-02-04 scripts/modules/parser_generic.py

@@ -1,0 +1,148 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       11/10/12
+content:    Base class for patients and such with a few methods shared by all
+            parsers.
+'''
+# Standard modules
+import os
+import re
+import glob
+from operator import *
+import numpy as np
+import scipy.linalg as linalg
+import matplotlib.pyplot as ppl
+import matplotlib.cm as cm
+
+from Bio import SeqIO, AlignIO
+from Bio.Seq import Seq
+from Bio.SeqRecord import SeqRecord
+from Bio.Alphabet.IUPAC import *
+
+
+
+# Globals
+alpha = np.array(['A', 'C', 'G', 'T', '-'])
+alphal = list(alpha)
+
+
+
+# Classes
+class patient_generic(object):
+    '''Generic class for a patient'''
+
+    def __repr__(self):
+        return self.name
+    
+
+    def L(self):
+        return len(self.seqs[0])
+
+   
+    def allele_frequency_trajectories(self, first_visit=None, last_visit=None,
+                                      alpha=alpha):
+        '''Allele frequency trajectories.
+        
+        Parameters:
+        -----------------
+        first_visit: the first visit to track. Default is the first visit found.
+        last_visit: the last visit to track. Default is the last visit found.
+        alpha: dictionary for possible alleles. Default is usual DNA code.
+
+        Returns:
+        -----------------
+        A tensor of size La x L x n_visits.
+        '''
+        if first_visit == None:
+            first_visit = self.visit[0]
+        if last_visit == None:
+            last_visit = self.visit[-1]
+
+        vs = self.visit[(self.visit >= first_visit) & (self.visit <= last_visit)]
+        n = len(vs)
+        L = self.L()
+        La = len(alpha)
+        aft = np.zeros((n, La, L))
+        for i, v in enumerate(vs):
+            aft[i] = self.allele_frequencies(seqs=self.seqs_from_visit(v),
+                                             alpha=alpha)
+        return aft.swapaxes(0,1).swapaxes(1,2)
+
+
+    def allele_frequencies(self, seqs=None, alpha=alpha):
+        '''Allele freqs in the choses seqs.
+
+        Parameters:
+        -----------------
+        seqs: which seqs to calculate from. Default is all seqs of the patient.
+        alpha: dictionary for possible alleles. Default is usual DNA code.
+        '''
+        if seqs is None:
+            seqs = self.seqs
+        smat = np.asarray(seqs)
[…]

### 2013-02-04 scripts/modules/parser_reference.py

@@ -10,8 +10,8 @@ import numpy as np
 import Bio.AlignIO as AlignIO
 from itertools import permutations
 
-from longitudinal.classes.helper import find_with_gaps
-from longitudinal.classes.parser_generic import alpha
+from helper import find_with_gaps
+from parser_generic import alpha
 
 
 

### 2013-02-04 scripts/simulations_plot_gradually_epitopesandtimeselec.py

@@ -16,6 +16,7 @@ import matplotlib.cm as cm
 
 
 # Globals
+results_dir = '../data/simulations/gradually_epitopesandtimeselec/'
 datafile = 'parameters_fixation_areas.dat'
 
 
@@ -27,8 +28,6 @@ def filterfolders(folders):
     if ('nescepi' in tmp) or ('timeselec' in tmp):
         tmp = tmp[:-2]
     base = '_'.join(tmp)
-    print tmp
-    print base
 
     ind = {}
     for i, f in enumerate(folders):
@@ -42,21 +41,12 @@ def filterfolders(folders):
             ind[4] = i
         elif ('nescepi' not in f) and ('timeselec' not in f):
             ind[0] = i
-    print folders
-    print ind
     return [folders[ind[i]] for i in xrange(5)]
 
 
 
 # Script
 if __name__ == '__main__':
-
-
-    # Input results dir
-    args = sys.argv
-    if len(args) < 2:
-        raise ValueError('Please specify a results folder')
-    results_dir = args[1].rstrip('/')+'/'
 
     # Folders
     g = os.walk(results_dir)

### 2013-02-04 scripts/simulations_plot_graduallydel.py

@@ -16,19 +16,13 @@ import matplotlib.cm as cm
 
 
 # Globals
+results_dir = '../data/simulations/gradually_more_del_6epitopes/'
 datafile = 'parameters_fixation_areas.dat'
 
 
 
 # Script
 if __name__ == '__main__':
-
-
-    # Input results dir
-    args = sys.argv
-    if len(args) < 2:
-        raise ValueError('Please specify a results folder')
-    results_dir = args[1].rstrip('/')+'/'
 
     # Load data using record arrays
     tb = np.genfromtxt(results_dir+datafile, unpack=False,

### 2013-02-04 scripts/simulations_scatter_all.py

@@ -16,6 +16,7 @@ import matplotlib.cm as cm
 
 
 # Globals
+results_dir = '../data/simulations/random_gradual_nonsyndead_6epitopes/'
 datafile = 'parameters_fixation_areas.dat'
 
 # Set limits for data-like sims
@@ -30,17 +31,9 @@ Anonsynmin = -0.5
 tmpsnsmin = 0e-6
 
 
+
 # Script
 if __name__ == '__main__':
-
-
-    # Input results dir
-    args = sys.argv
-    if len(args) < 2:
-        raise ValueError('Please specify a results folder')
-    results_dir = args[1].rstrip('/')+'/'
-    
-
 
     # Load data using record arrays
     tb = np.genfromtxt(results_dir+datafile, unpack=False,

### 2013-02-04 scripts/modules/alignment.py
lifted remote external dependencies (should be self-contained now)

@@ -8,8 +8,10 @@ content:    Tools for managing sequences and alignments of general nature.
 import os
 import glob
 import numpy as np
+
 from Bio import SeqIO, AlignIO
-import general.classes.alphabet as abc
+
+import alphabet as abc
 
 
 # Globals

### 2013-02-04 scripts/modules/codon_usage.py

@@ -6,8 +6,10 @@ content:    Read codon usage tables
 '''
 # Modules
 import re
-from longitudinal.classes.parser_generic import alpha
+
 from Bio.Seq import translate
+
+from parser_generic import alpha
 
 
 

### 2013-02-04 scripts/modules/helper.py

@@ -7,8 +7,9 @@ content:    Helper functions for all parsers
 # Standard modules
 import numpy as np
 import matplotlib.pyplot as ppl
-import Bio.Seq
 import re
+
+import Bio.Seq
 
 
 # Functions

### 2013-02-04 scripts/modules/parser_reference.py

@@ -7,8 +7,9 @@ content:    Implement some globals used by many parsers that interact with a
 '''
 # Modules
 import numpy as np
-import Bio.AlignIO as AlignIO
 from itertools import permutations
+
+import Bio.AlignIO as AlignIO
 
 from helper import find_with_gaps
 from parser_generic import alpha

### 2013-02-06 scripts/modules/parser_SHAPE.py
SHAPE files plus simulation script

@@ -10,9 +10,9 @@ import re
 
 
 # Globals
-helixfile = '/home/fabio/university/phd/longitudinal/data/SHAPE/nature08237-s2.txt'
-bpseqfile = '/home/fabio/university/phd/longitudinal/data/SHAPE/nature08237-s2_bpseq.txt'
-ppfile = '/home/fabio/university/phd/longitudinal/data/SHAPE/nature08237-s3.txt'
+helixfile = '../data/SHAPE/nature08237-s2.txt'
+bpseqfile = '../data/SHAPE/nature08237-s2_bpseq.txt'
+ppfile = '../data/SHAPE/nature08237-s3.txt'
 
 
 

### 2013-02-06 scripts/simulations_evolution.py

@@ -1,0 +1,311 @@
+#!/ebio/ag-neher/share/programs/EPD/bin/python
+# vim: fdm=marker
+'''
+author:     Fabio Zanini
+date:       07/08/12
+content:    Study fixation and loss of synonymous/nonsynonymous changes by a
+            short-genome model (1000 sites). This script is the one eventually
+            used to produce the simulations for the paper.
+'''
+# Standard modules
+import os
+import sys
+from operator import *
+import numpy as np
+from itertools import combinations
+
+# Env vars
+JOBSSCRIPT = os.path.realpath(__file__)      # This file
+JOBDIR = os.path.dirname(JOBSSCRIPT)+'/'
+JOBMODULES = JOBDIR+'modules'    
+if JOBMODULES not in sys.path: sys.path.insert(0,JOBMODULES)
+
+# Custom modules (import local FFPopSim)
+import FFPopSim as h
+
+
+
+# Globals
+VERBOSE = 0
+L = 1000
+mu = 2e-5
+crossover = 1e-3
+epitope_length = 9
+
+# Simulation length
+dt = 10
+n_time_points = 600
+dt_sample = 20
+
+
+
+# Classes
+class haploid_highd_gradual(h.haploid_highd):
+    '''Population with dynamic fitness landscape'''
+    
+
+    def __init__(self, *args, **kwargs):
+        if 'nescepi' not in kwargs:
+            raise ValueError('Please add the "nescepi" argument.')
+        newkwargs = {key: value for key, value in kwargs.iteritems() if key != 'nescepi'}
+        super(haploid_highd_gradual, self).__init__(*args, **newkwargs)
+        self.nescepi = kwargs['nescepi']
+        self.epitopes = []
+        self.epitope_times = []
+        self.epitope_fitness = []
+
+
+    def add_epitope(self, value):
+
+        # Complex epitope creation
+        if self.nescepi > 1:
+            # Choose a random good position
+            pos = np.random.randint(self.L - epitope_length)
+            epitope = [p for p in range(pos, pos + epitope_length) if (p + 1) % 3]
+            np.random.shuffle(epitope)
+            epitope = epitope[:self.nescepi]
+
+            add, epis = create_epitope(epitope, value)
+            fitness = self.get_trait_additive()
+            for i, v in add.iteritems():
+                fitness[i] = v
+            self.set_trait_additive(fitness)
+            for loci, v in epis.iteritems():
+                self.add_trait_coefficient(v, loci)
+
+            # Store changes
[…]

### 2013-02-09 supplementary.tex
Edits in main text and supplement. Supplement needs more work. Methods need to be somewhat more detailed.

@@
 \usepackage{natbib}
-\usepackage{pslatex}\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
+\usepackage{pslatex}
+\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \newcommand{\Affiliation}{Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany}
-\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
+%\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[h]
+\begin{figure}[ht]
 \begin{center}
@@
 from blue to red). Panel B) Allele frequency trajectories for nonsynonymous
-changes in the same patient. Panels C and D) the same plots from patient p7, whose
-sequences show signs of compartimentalization.}
+changes in the same patient. Panels C and D) show analogous plots for data from
+patient p7. Samples after day 1000 split into two clusters in the PCA and no
+mutations that arise after day 1000 fix, presumably because they are restricted
+to one subpopulation. }
 \label{fig:aftp}
@@
 \end{figure}
-Not all of the patients studied in refs.~\citealp{shankarappa_consistent_1999} and
-\citealp{bunnik_autologous_2008} were analyzed. The reason is that patients p4,
-p7, p8, p9 from ref.~\citealp{shankarappa_consistent_1999} and patients ACH19542
-and ACH19768 from ref.~\citealp{bunnik_autologous_2008} showed clear signs of
-compartimentalization (population structure) during infection, which would have
-been an irreparable confounder for our analysis. The difference is clear if
-PCA is performed on all sequences from a ``compartimentalized'' patient, p7.
-This is shown in \figurename~\ref{fig:aftp}: late sequences subdivide in two
-well-defined clusters -- this characteristic is absent from all the patient
-sequences used in our main analysis. On the right, allele frequency trajectories from the same
-patient are plotted: evidently, no derived allele is fixed after 1000 days
-from seroconversion, but a lot of mutations swarm to frequencies around 0.5-0.8. Of
-course, the fixation probability of new mutations in such a sample is very low,
-but this is an effect of population structure rather than any
-of the processes discussed in the main text (time-dependent selection, competition
-of mutants within the same epitope).
+Patients p4, p7, p8, p9 from ref.~\citealp{shankarappa_consistent_1999} and
+patients ACH19542 and ACH19768 from ref.~\citealp{bunnik_autologous_2008} showed clear signs of
+compartimentalization (population structure) during infection. This population
+structure is easy to spot after a principle component analysis (PCA) of all
+sequences from all samples from one patients. When the projections onto the
+first 3 PCs for each sequence, the above mentioned patients show bifurcation
+into two clusters at later time points. Since mutations that arise in one subpopulation cannot fix population wide, such population
+structure confounds our analysis and these patients were exclude.
+
+Similar, but less obvious, conclusions can be drawn from the allele frequency
+trajectories plotted next to the PCA plots: no allele that arose after 1000
+days from seroconversion ever fixes, but a most mutations hover at frequencies
+around 0.5-0.8. Fluctuations of these allele frequencies probably reflect
+changing relative sizes of the subpopulations.
 
@@
 \caption{Cumulative distribution of synonymous and nonsynonymous diversity in
-{\it gag}. Nonsynonymous changes are, overall, much less common than synonymous
-ones.}
+{\it gag} in the LANL reference panel. Nonsynonymous sites are much
+less diverse than 4-fold synonymous sites. \comment{The title is wrong. How
+do you decide on 4-fold degeneracy? (consensus amino acid???)}.}
 \label{fig:synnonsyncons}
@@
 \end{figure}
+\comment{this can use streamlining. Why not put is all into the figure caption}
 In the main text, it is argued that, whereas the deleterious effect of

### 2013-02-09 synmut.tex
small change in the methods
Edits in main text and supplement. Supplement needs more work. Methods need to be somewhat more detailed.

--- 2013-02-03 synmut.tex
@@
 Intrapatient HIV evolution is dominated by selection on the protein level in the
-arms race with the adaptive immune system. Synonymous mutations do not have an
-adaptive immunity-related phenotype and are often assumed to be neutral. Here, we show
-that synonymous changes in epitope-rich regions are often deleterious but still
-reach high frequencies in the viral population.
+arms race with the adaptive immune system. Synonymous mutations do not affect 
+the interaction with the adaptive immune system and are often
+assumed to be neutral.
 We analyze longitudinal intrapatient data from the \shankaregion~part of the
-envelope gene (\env) and observe that synonymous derived alleles are rarely
-fixed.
-Computer models suggest that such synonymous mutations have a (Malthusian) selection
-coefficient of the order of $-0.002$, and that they are brought up to high
+envelope gene (\env) and observe that synonymous derived alleles  rarely
+fix even though they often reach high frequencies in the viral population.
+We find that synonymous mutations that disrupt base pairs in RNA stems flanking
+the variable loops of gp120 are more likely to be lost than other synonymous
+changes, hinting at a direct fitness effect of these stem-loop structures in the
+HIV RNA.
+Computational modeling indicates that these synonymous mutations have a
+(Malthusian) selection coefficient of the order of $-0.002$, and that they are brought up to high
 frequency by hitchhiking on neighboring beneficial nonsynonymous alleles.
-We find that synonymous mutations that disrupt base pairs in RNA stems flanking
-the V loops are more likely to be lost than other synonymous changes, hinting at
-a direct fitness effect of such stem-loop structures in the HIV RNA.
 The patterns of fixation of nonsynonymous mutations suggest that antibody
@@
 nuclear export, and packaging into the viral capsid: all these processes operate at the RNA
-level and are sensitive to synonymous changes.
-On the one hand, these processes often involve some kind of RNA secondary structure.
+level and are sensitive to synonymous changes since these processes often
+depend on RNA folding.
 For example, the HIV \rev{} response element (RRE) in \env{} enhances nuclear export of
@@
 untranslated region~\citep{barat_interaction_1991, paillart_vitro_2002}.
-On the other hand, nucleotide-level fitness effects have been observed beyond RNA structure as
-well.
-In most species, it is  known that synonymous codons don't evolve completely neutrally
-but that some codons are favored over others \citep{plotkin_synonymous_2011};
-this is expected to be particularly important in compact viruses like HIV. 
+
+Even in absense of important RNA structure synonymous codons don't
+evolve completely neutrally but that some codons are favored over others in many
+species \citep{plotkin_synonymous_2011}.
 Recent studies have shown that genetically engineered HIV strains with
@@
 selection beyond the protein sequence is therefore expected
-\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}; nonetheless, positive
-selection via the host adaptive immune system is restricted to changes in the
-amino acid sequence.
+\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}. Positive
+selection through the host's immune system, however, is restricted to
+changes in the amino acid sequence.
 
@@
 We argue that synonymous mutations reach high frequencies via genetic hitchhiking
-thanks to the small recombination rate of HIV~\citep{neher_recombination_2010,
+due to the small recombination rate of HIV~\citep{neher_recombination_2010,
 batorsky_estimate_2011}.
@@
 this is a simple consequence of the fact that
-(a) exactly one of the $N$ individuals in the current population will be
+(i) exactly one individual in the current population will be
 the common ancestor of the entire future population at a particular locus and
-(b) this ancestor has a probability $\nu$ of carrying the mutation (assuming
+(ii) this ancestor has a probability $\nu$ of carrying the mutation (assuming
 the neutral mutation is not preferentially associated with genomes of high or
@@
 the time course the frequencies of all synonymous and nonsynonymous mutations
-observed \env~, C2-V5, in patient p1~\citep{shankarappa_consistent_1999},
-respectively. Despite many synonymous mutations reaching high frequency, very
+observed \env, C2-V5, in patient p1~\citep{shankarappa_consistent_1999},
+respectively. Despite many synonymous mutations reaching high frequency, 
 few fix (panel~\ref{fig:aftsyn}); in constrast, many nonsynonymous mutations fix
@@
 \citet{bunnik_autologous_2008} is fully compatible with data from
-\citet{shankarappa_consistent_1999}. The nonsynonymous seem to follow more or
-less the neutral expectation (blue line) -- a point to which we will come back below.
+\citet{shankarappa_consistent_1999}. The nonsynonymous mutations seem to follow
+more or less the neutral expectation (blue line) -- a point to which we will come back below.
 
@@
 \label{fig:aftnonsyn}}
-\caption{Time series of allele frequencies in \env, C2-V5, from
-patient 1~\cite{shankarappa_consistent_1999},
-divided by synonymous (panel A) and nonsynonymous (panel B) derived alleles.
-While nonsynonymous mutations frequently fix, very few synonymous
+\caption{Time series of frequencies of synonymous (A) and non-synonymous (B)
+derived alleles in \env, C2-V5, from patient
+1~\cite{shankarappa_consistent_1999}.
+While many nonsynonymous mutations  fix, few synonymous
 mutations do even though they are frequently observed at intermediate
@@
 When interpreting these results for the fixation probabilities, it is important
-to distinguish between random mutations and polymorphisms at a given frequency,
-which have already been filtered by selection.
+to distinguish between random mutations and polymorphisms observed at a certain
+frequency since the latter have already been filtered by selection.
 A polymorphism could be beneficial to the virus and on its way to fixation. In
@@
 deleterious, while outside this regions polymorphisms are mostly roughly
-neutral. Note that this does not imply that all synonymous mutations are neutral
--- only those mutations observed at high frequencies, which have been
-experiencing selection for some time already, tend to be neutral.
+neutral. Note that this does not imply that all synonymous mutations in this
+region are neutral -- only those mutations observed at high frequencies, which
+have been experiencing selection for some time already, tend to be neutral.
 
@@
 \label{fig:fixp2}}
-\caption{Fixation and loss of synonymous mutations.
+\caption{Fixation and loss of synonymous mutations. \comment{fix x-axis label}
 Panel A) Time
@@
 One possible {\it a priori} explanation for lack of fixation of synonymous
-mutations in C2-V5 are  secondary structures in the viral RNA. It has
-been suggested early on that parts of the viral genome that has the potential to
-form stems is better conserved than the
+mutations in C2-V5 are  secondary structures in the viral RNA whose disruption
+is deleterious to the virus.
+It has been suggested early on that parts of the viral genome that have the
+potential to form RNA stems are better conserved than the
 remainder~\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}.
@@
 shown in \FIG{SHAPEA}, the reactivities of fixed
-alleles (red line) are systematically larger than of alleles are lost (blue
-line) (Kolmogorov-Smirnov test, $P\approx 0.002$).
+alleles (red histogram) are systematically larger than of alleles are lost (blue
+) (Kolmogorov-Smirnov test, $P\approx 0.002$).
 In other words, alleles that are likely to break RNA helices are also more
-likely to revert and finally be lost from the population. As a control, the
-average over non-observed but potentially available polymorphisms lies between
-the two curves (green line), as expected (because only some of these mutations
-would interfere with stem-loop formation).
+likely to revert and finally be lost from the population. The
+average over all mutations that rare not observed  (green) lies between the
+those that fix or get lost.
 To test the hypothesis that mutations in C2-V5 are lost because they break stems in the
-conserved streches between the variable loops, we split the synonymous mutations in the
-extended V1-V5 region further into conserved and variable regions and found that
-the biggest depression in fixation probability is observed in the conserved
-stems, while the variable loops show little deviation from the neutral
-signature, see \FIG{SHAPEB}. This is consistent with important stem structures
-in conservered regions between loops.
+conserved streches between the variable loops, we consider mutations in
+variable loops and conserved parts separately. The biggest depression in
+fixation probability is observed in the conserved stems, while the variable
+loops show little deviation from the neutral signature, see \FIG{SHAPEB}. This
+is consistent with important stem structures in conservered regions between loops.
 
@@
 human codons~\citep{jenkins_extent_2003,kuyl_biased_2012}. We do not find,
-however, any evidence for a contribution of CUB to the ultimate fate of
-synonymous alleles consistent with the fact that CUB between HIV and human genes
-is not shrinking at the macroevolutionary level \citep{kuyl_biased_2012}. 
-Furthermore, CUB in the V1-C5 region is not very different from other parts of
-the HIV genome, whereas the reduced fixation probability is only observed there. In
-conclusion, although we cannot exclude an effect of CUB on fitness as a general
-rule, we expect it to be a minor effect in our context.
+however, any evidence for a contribution of average CUB to the ultimate fate of
+synonymous alleles consistent with that HIV does not seem to adapt its codon
+usage to its human host cells at the macroevolutionary level
+\citep{kuyl_biased_2012}.
 
@@
 substantially higher SHAPE reactivities than mutations that reach frequencies above 15\%, but are
-susequently lost (restricted to the vicinity of variable loops). This difference
+susequently lost (restricted to the region V1-V5$\pm 50bp$)????. This difference
 is quantified by a cumulative histogram ($p=0.002$, Kolmogorov-Smirnov two 
@@
 is not unexpected, it seems odd that we observe them at high population
-frequency -- at least in some regions of the genome. The region of \env~ in
-which we observe deleterious mutations at high frequency, however, is special in
-that it undergoes frequent adaptive changes to evade recognition by neutralizing
-antibodies~\cite{williamson_adaptation_2003,richman_rapid_2003}. Due to the limited amount of
-recombination in HIV~\cite{neher_recombination_2010,batorsky_estimate_2011},
+frequency -- at least in some regions of the genome. The region of \env~in
+which we observe deleterious mutations at high frequency undergoes frequent
+adaptive changes to evade recognition by neutralizing 
+antibodies~\cite{williamson_adaptation_2003,richman_rapid_2003}. Due to the 
+limited amount of recombination in HIV~\cite{neher_recombination_2010,batorsky_estimate_2011},
 deleterious mutations that are linked to adaptive variants can reach high
@@
 or genetic draft \citep{gillespie_genetic_2000,neher_genetic_2011}.
+Hitchhiking is  apparent in \FIG{aft}, that shows that many mutations
+change rapidly in frequency as a flock.
 
-The potential for hitchhiking is already apparent from the allele frequency
-trajectories in \FIG{aft}, where many mutations appear to change rapidly in
-frequency as a flock. In order
-to be advected to high frequency by a linked adaptive mutation, the deleterious
-effect of the mutation has to be substantially smaller than the adaptive effect.
-The latter was estimated to be on the order of $s_a = 0.01$ per day~\citep{neher_recombination_2010}.
 The approximate magnitude of the deleterious effects can be estimated from
@@
 fixation probability and diversity. Parameters are chosen uniformly on a
-logarithmic scale from the indicated windows. This simulation assumes 6
-roughly equivalent escape mutations per epitope; see supplement for alternative
-models.}
+logarithmic scale from the indicated windows. \comment{Can't we show this
+for simple epitopes? I can't seem to find a way to describe this from the
+outset}.
+This simulation assumes 6 roughly equivalent escape mutations per epitope; see supplement for alternative models.}
 \label{fig:simheat}
@@
 To get a better idea of the range of parameters that are compatible with the
-observations and our interpretation, we perform computer simulations of
+observations and our interpretation, we performed computer simulations of
 evolving viral populations assuming a mix of positive and purifying selection
@@
 module dedicated to intrapatient HIV evolution~\citep{zanini_ffpopsim:_2012}. 
+For each simulation run, we specify the deleterious effect of synonymous
+mutations, the fraction of synonymous mutations that are deleterious, the
+escape rate of adaptive non-synonymous mutations and the frequency of new
+escapes. Note that the escape rate is sum of two factors: (i) the beneficial
+effect due to the ability to evade the immune system minus (ii) the fitness
+cost of the mutation in terms of structure, stability, etc. Net escape rates
+in chronic infections have been estimated to be on the order of 
+$s_a = 0.01$ per day~\citep{neher_recombination_2010,Asquith}.
 
-The first result of the simulations is that genetic draft can indeed bring weakly
-deleterious mutations to high frequencies and results in a dependence of the
-fixation probability on initial frequency that is compatible with observations
-(see \FIG{simfixpvar}).
-To assess the reduction in fixation probability quantitatively, we calculate
+\FIG{simfixpvar} shows simulations results for the fixation probability and the
+synonymous diversity for different deleterious effects of synonymous mutations. 
+We quantify synonymous diversity,$P_\text{interm}$, as the fraction of sites
+with an allele at frequency $0.25 < \nu < 0.75$. The synonymous diversity
+observed in patient data is indicated in the figure.
+To quantify the depression of the fixation probability, we calculate
 the area between the measured fixation probability and the diagonal, which is
-the neutral expectation (\FIG{simfixpvar}, lower inset). This area spans a range
-between $-0.5$ (no fixation at all), zero (neutral-like curve), and $+0.5$ (always
-fixed). Furthermore, we quantify synonymous diversity via the probability
-$P_\text{interm}$ of a synonymous derived allele to be at intermediate frequencies
-$0.25 < \nu < 0.75$ (\FIG{simfixpvar}, top inset).
-The HIV data correspond to an area $A_\text{syn} \sim -0.2$  for synonymous
-changes and $A_\text{nonsyn} \sim 0$ for nonsynonymous changes,
-and a synonymous diversity $P_\text{interm} \sim 0.005$.
+the neutral expectation (\FIG{simfixpvar}, lower inset). If no fixation happens,
+the area will be $-0.5$; if every mutation fixes, we will measure 
+$A= 0.5$. In HIV, we find $A_\text{syn} \sim -0.2$  for
+synonymous changes and $A_\text{nonsyn} \sim 0$ for nonsynonymous changes.
+
 In the three simulations shown in \FIG{simfixpvar}, the fixation probability of
@@
 To map the parameter range of the model that is compatible with the data, we
-repeatedly simulated the evolution with random choices for the deleterious
-effect of synonymous mutations, the fraction of synonymous mutations that are
-deleterious, the fitness effect of adaptive mutations (escape rate) and the frequency of
-escapes. Among all simulations, we select the ones that show $A_\text{syn}$ and
+repeatedly simulated the evolution with random choices for the parameters in
+certain bounds, see \FIG{simsfig}.
+Among all simulations, we select the ones that show $A_\text{syn}$ and
 $P_\text{interm}$ as observed in the data, i.e., a large depression in fixation
@@
 for which we found $A_\text{syn} < -0.15$ and $0.0025 < P_\text{interm} <
-0.010$. This selection imposes a strong
-constraint on the deleterious fraction and effect size (see \FIG{simsfig}).
-A high fraction ($\gtrsim 0.8$) of only slightly deleterious synonymous sites,
-with effect size $|s_d| \sim 0.002$, is required. This result fits well the
-expectation based on the fixation/extinction times above (see \FIG{fixp1}).
+0.010$. These conditions indicate that a high fraction ($\gtrsim 0.8$) of
+deleterious sites has to be deleterious with effect size $|s_d| \sim 0.002$. 
+This result fits well the expectation based on the fixation/extinction times above (see \FIG{fixp1}).
 The results are plausible:
-(a) since high frequency bins are enriched in neutral mutations (it is easier
-for them to hitchhike), a depression in $\pfix$ will be
-visible only if deleterious mutations are pervasive; (b) in order to
-hitchhike, the deleterious effect size has to be much smaller than the escape
-rate, otherwise the
-double mutant has little or no fitness advantage; (c) mutations with a 
-deleterious effect weaker than approximately $0.001$ behave neutrally
-consistent with the typical coalescent times observed in HIV.
+(i) a substantial depression in $\pfix$ requires pervasive deleterious
+mutations, otherwise the majority of observed polymorphisms are neutral and no
+depression is observed; (ii) in order to hitchhike, the deleterious effect size
+has to be much smaller than the escape rate, otherwise the double mutant has
+little or no fitness advantage; (iii) mutations with a deleterious effect
+smaller than approximately $0.001$ behave neutrally consistent with the typical
+coalescent times observed in HIV.
 
 As long as only features of synonymous mutations are filtered, as performed
-above, no strong correlation of the two nonsynonymous parameters, escape rate
-and rate of appearance of epitopes, is observed; in \FIG{simsfig}, the points
-shown possess a variety of shapes and colors. In most of these simulations, 
-however, nonsynonymous mutations almost always fix once they reach high frequencies
+above, the two nonsynonymous parameters, escape rate
+and rate of appearance of epitopes, are little constrained as is apparent in
+\FIG{simsfig} from the variety of sizes and colors indicating these parameters.
+In most of these simulations, however, nonsynonymous mutations almost always fix once they reach high frequencies
 -- their nonsynonymous fixation area is well above zero. This is incompatible
-with the blue line in \FIG{fixp}: in an HIV infection, nonsynonymous mutations fix as if they
-were neutral. Although there are no doubts that nonsynonymous variation in the
-variable regions is driven by positive selection and not by genetic drift, the
-biological reason for the neutral-like curve in \FIG{fixp} is unclear.
-Inspecting the trajectories of nonsynonymous mutations
-suggests the rapid rise and fall of many alleles. We test two possible
+with the blue line in \FIG{fixp}: in an HIV infection, nonsynonymous
+mutations at high frequency often disappear again, even though many are at
+least transiently beneficial. Inspecting the trajectories of nonsynonymous
+mutations suggests the rapid rise and fall of many alleles. We test two possible
 mechanisms that are biologically plausible and could explain the transient rise
@@
 escape mutants. These respones are delayed by a few months, roughly matching the
-average time needed by an escape mutant to cross frequencies of order one.
+average time needed by an escape mutant to rise from low to high frequency.
 In the alternative hypothesis, several different escape
@@
 packaging, etc.~is most likely to eventually fix. The emergence of
-multiple sweeping nonsynonymous mutations in real HIV infections has been shown
-previously~\citep{moore_limited_2009, bar_early_2012}.
-We tested both hypotheses in our simulation framework and they both seem to
-achieve a reduced, neutral-like fixation probability while maintaining a high
-rate of substitutions, compatibly with HIV measurements (see supplementary
-materials).
-In real HIV infections, both mechanisms are likely to be playing a role, as they
-are not mutually exclusive.
+multiple sweeping nonsynonymous mutations in real HIV infections has been 
+shown~\citep{moore_limited_2009, bar_early_2012}.
+We tested both hypotheses in our simulation framework and they both result
+in a reduced, neutral-like fixation probability while maintaining a high
+rate of substitutions, compatible with observations in patient data (see
+supplementary materials).
+The two scenarios are not exclusive and possibly both important in HIV
+evolution.
 
@@
 Comparison with recent biochemical studies of binding propensity of bases in RNA
-genome suggest that these mutations are deleterious, at least in part, because they disrupt
-stems in RNA secondary structures. Furthermore, we provide evidence that these
-mutations are brought to high frequency through linkage to adaptive mutations.
+genome indicates that these mutations are deleterious, at least in part, because
+they disrupt stems in RNA secondary structures. Computational modelling
+shows that these mutations can be brought to high frequency through
+linkage to adaptive mutations.
 
@@
 
-Alternatively, the lack of fixation could be due to (a) time dependent environment
-through an immune system that is catching up, or (b) competition between mutations
+Alternatively, the lack of fixation could be due to time dependent environment
+through an immune system that is catching up, or competition between mutations
 that mediate escape within the same epitope. We explore both of these
@@
 previously~\citep{watts_architecture_2009, sanjuan_interplay_2011}.
-Our analysis is limited by the availability of
-longitudinal data which requires a focus on the the variable regions of \env.
 Conserved RNA structures exist in different parts of the HIV genome (several are
-known). In absence of repeated adaptive substitutions in the vicinity that cause
-hitchhiking, the deleterious synonymous mutations remain at low frequencies and
-can only be observed by deep sequencing methods.
+known), but our analysis requires longitudinal data most of which focusses
+on \env.
+Furthermore, we only expect to see deleterious synonymous
+mutations at high frequencies if they are linked to repeated adaptive
+substitions. Otherwise, they remain at low frequencies and can only be observed
+by deep sequencing methods.
 
@@
 A similar method has been recently used in a longitudinal study of
-influenza~\citep{strelkowa_clonal_2012}. The propagators suggested in that
-paper, however, represent ratios between nonsynonymous and synonymous mutations.
+influenza evolution~\citep{strelkowa_clonal_2012}. The propagators suggested in
+that paper, however, represent ratios between nonsynonymous and synonymous mutations.
 The latter is used as an approximately neutral control and this method can
@@
 independent or additive. If genetic variation is only transiently beneficial,
-existing methods to quantify selection will yield substantial underestimates
-\citep{williamson_adaptation_2003,neher_rate_2010,batorsky_estimate_2011}.
+existing estimates  of the strength of selection
+\citep{neher_rate_2010,batorsky_estimate_2011} could
+be substantial underestimates.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-Longitudinal intrapatient viral RNA sequences are collected for published
+Longitudinal intrapatient viral RNA sequences were collected from published
 studies~\citep{shankarappa_consistent_1999, liu_selection_2006,
 bunnik_autologous_2008} and downloaded from the Los Alamos National Laboratory
-(LANL) HIV sequence database~\citep{LANL2012}. The sequences from some patients
-show signs of HIV compartimentalization into subpopulations and are
-discarded (see supplementary materials);
-a total of 11 patients with approximately 6 time points each and 10
-sequences per time point are eventually analyzed. The time interval or
-resolution between two consecutive sequences is approximately 6 to 18 months.
+(LANL) HIV sequence database~\citep{LANL2012}. The samples from some patients
+show substantial population structure and were discarded (see supplementary
+materials); a total of 11 patients with 6-?? time points each and 10
+sequences per time point were analyzed. The time intervals between two
+consecutive sequences ranged from 6 to 18 months. \comment{CHECK VALUES}.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Sequence analysis}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-The sequences are aligned within each patient
-via the translated amino acid sequence, using
-Muscle~\citep{edgar_muscle:_2004}, and to the NL4-3 reference sequence used
-by \citet{watts_architecture_2009} in the SHAPE assay. Within each patient, a
-consensus nucleotide sequence at the first time point is used to classify
-alleles as ancestral or derived at all sites. Problematic sites that include large frequencies of gaps
-are excluded from the analysis to avoid artefactual substitutions due to
-alignment errors. Time series of allele frequencies are extracted from the
-sequences. \comment{The alignments used are available for download}.
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+The sequences were translated and the resulting amino-acid sequences aligned
+using Muscle~\citep{edgar_muscle:_2004} to each other and the NL4-3 reference
+sequences separately for each patient. Within each patient, the consensus
+nucleotide sequence at the first time point was used to classify alleles as
+``ancestral'' or ``derived'' at all sites. Sites that include large
+frequencies of gaps were excluded from the analysis to avoid artefactual
+substitutions due to alignment errors. Allele frequencies at different time
+points were extracted from the multiple sequence alignment.
 
-The synonymity of a mutation is assigned if the rest of the codon was
-in the ancestral state and using the standard genetic code. Cases where more
-than one mutation within the codon is observed are discarded. Slightly
-different criteria for synonymous/nonsynonymous discrimination yield similar
-results.
+A mutation was considered synonymous if it did not change the amino acid
+corresponding to the codon, and if the rest of the codon was in the ancestral
+state. Codons with more than one mutation were discarded. Slightly different
+criteria for synonymous/nonsynonymous discrimination yield similar results.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-For the estimate of times to fixation/extinction, polymorphisms are
+For the estimate of times to fixation/extinction, polymorphisms were
 binned by frequency and the time to reaching the first boundary (fixation or
-extinction) is stored. For the fixation probability, the long-time limit of the
-resulting curves is used, excluding polymorphisms that arise late in the
-clinical history (hence would have no time to reach either boundary).
+extinction) was stored. The fixation probability was determined as the long-time
+limit of the resulting curves. Samples less than ???days
+prior the last time point were omitted since there was not enough information
+on their future fate, i.e., whether they fix or are lost.
 
-For the correlation analysis with RNA secondary structure, the SHAPE scores are
-downloaded from the journal website~\citep{watts_architecture_2009}. By virtue
-of the alignment of the longitudinal sequences with the reference used by
-\citet{watts_architecture_2009}, SHAPE reactivities are assigned to most sites.
-Problematic assignments in indel-rich regions are excluded from the analysis.
+The SHAPE scores quantifying the degree of base pairing of individuals sites in
+the HIV genome were downloaded from the journal
+website~\citep{watts_architecture_2009}.
+Where possible, SHAPE reactivities were assigned to sites in the multiple
+sequence alignments for each patient through the alignment to the sequence
+of the NL4.3 virus used in \citep{watts_architecture_2009}. Problematic
+assignments in indel-rich regions are excluded from the analysis.
+\comment{The following is super confusing}
 In order to restrict the analysis to synonymous polymorphisms, a lower frequency
-threshold of 0.15 was used (other thresholds yield the same results). Since
-very few polymorphisms hitchhike beyond, say, a frequency of 0.5, this pool is
+threshold of 0.15 was used (other thresholds yield the same results). Since very
+few polymorphisms hitchhike beyond, say, a frequency of 0.5, this pool is
 enriched for to-be-lost mutations; hence the ``lost" curve in \FIG{SHAPEA}
@@
 
-The V loops and flanking regions are identified manually starting from the
-annotated reference HXB2 sequence from the LANL HIV database~\citep{LANL2012}. A
+The variable loops and flanking regions are identified manually starting from
+the annotated reference HXB2 sequence from the LANL HIV database~\citep{LANL2012}. A
 similar approach is used to label the C2-V5 region sequenced in
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-Simulations are performed using the recently published software
-FFPopSim~\citep{zanini_ffpopsim:_2012}. Both full-length HIV genomes and
-\env{}-only simulations are performed and yield comparable results. For each
-set of parameters, approximately 100 simulation runs are averaged over. In each
-run, a random fitness landscape with specified statistical properties (e.g.
-density of beneficial sites, average deleterious effect of synonymous changes) is generated.
-
+Simulations are performed using FFPopSim~\citep{zanini_ffpopsim:_2012}. Both full-length HIV genomes and
+\env{}-only simulations are performed and yield comparable results. In each
+run, a random fitness landscape with specified statistical properties 
+(e.g.~density of beneficial sites, average deleterious effect of synonymous changes) is generated.
+For each set of parameters, approximately 100 simulation runs are averaged over. 
+\comment{needs a little bit more detail}.
 For the discussion of simulation parameters, the areas below or above the neutral
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-All analysis scripts are available for download.
+All analysis scripts, as well as the sequence alignment used, are available for download.
 

### 2013-02-10 supplementary.tex
supplement only captions, more details about simulations. need to redo simulations without complex epitopes for 4B

@@
 \newcommand{\Affiliation}{Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany}
-%\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{center}
-\includegraphics[width=0.4\linewidth]{Shankarappa_PCA_p1}
-\includegraphics[width=0.4\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p1}
-\includegraphics[width=0.4\linewidth]{Shankarappa_PCA_p7}
-\includegraphics[width=0.4\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p7}
+\includegraphics[width=0.35\linewidth]{Shankarappa_PCA_p1}
+\includegraphics[width=0.35\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p1}
+\includegraphics[width=0.35\linewidth]{Shankarappa_PCA_p7}
+\includegraphics[width=0.35\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p7}
 \caption{Panel A) PCA of all sequences from patient p1 (colors indicate time from seroconversion,
@@
 mutations that arise after day 1000 fix, presumably because they are restricted
-to one subpopulation. }
+to one subpopulation. All patients like p7 (p4, p7, p8, p9 from ref.~\citealp{shankarappa_consistent_1999} and
+ACH19542 and ACH19768 from ref.~\citealp{bunnik_autologous_2008}) were excluded
+from our analysis.}
 \label{fig:aftp}
@@
 \end{figure}
-Patients p4, p7, p8, p9 from ref.~\citealp{shankarappa_consistent_1999} and
-patients ACH19542 and ACH19768 from ref.~\citealp{bunnik_autologous_2008} showed clear signs of
-compartimentalization (population structure) during infection. This population
-structure is easy to spot after a principle component analysis (PCA) of all
-sequences from all samples from one patients. When the projections onto the
-first 3 PCs for each sequence, the above mentioned patients show bifurcation
-into two clusters at later time points. Since mutations that arise in one subpopulation cannot fix population wide, such population
-structure confounds our analysis and these patients were exclude.
-
-Similar, but less obvious, conclusions can be drawn from the allele frequency
-trajectories plotted next to the PCA plots: no allele that arose after 1000
-days from seroconversion ever fixes, but a most mutations hover at frequencies
-around 0.5-0.8. Fluctuations of these allele frequencies probably reflect
-changing relative sizes of the subpopulations.
 
@@
 \caption{Cumulative distribution of synonymous and nonsynonymous diversity in
-{\it gag} in the LANL reference panel. Nonsynonymous sites are much
-less diverse than 4-fold synonymous sites. \comment{The title is wrong. How
-do you decide on 4-fold degeneracy? (consensus amino acid???)}.}
+{\it gag} in the LANL reference panel(filtered
+sequences only, version 2011)~\cite{LANL2012}. Sites such that the consensus codon
+has three synonymous and six nonsynonymous single mutants were used, and the number of observed mutants
+of a certain type over the number of possible mutants is plotted. Nonsynonymous
+changes are observed less often, {\it ergo} are more conserved, than
+synonymous changes. It can be therefore assumed, as mentioned in the main text,
+that non-escape nonsynonymous changes involve a large fitness cost.}
 \label{fig:synnonsyncons}
@@
 \end{figure}
-\comment{this can use streamlining. Why not put is all into the figure caption}
-In the main text, it is argued that, whereas the deleterious effect of
-synonymous changes is very small, most nonsynonymous mutations outside of the
-variable regions are much worse. To show this, we take a multiple sequence alignment of
-the {\it gag} gene of subtype-B HIV sequences from the LANL website (filtered
-sequences only, version 2011)~\cite{LANL2012} and look at diversity at fourfold degenerate
-codons. The ratio of observed versus possible mutations is much lower for
-nonsynonymous changes than synonymous ones, as shown in
-\figurename~\ref{fig:synnonsyncons} in a cumulative histogram. From this
-evidence (which is highly statistically significant) we conclude that, as a
-general rule, nonsynonymous changes that are not currently escape mutations
-are likely to incur in large fitness costs.
 
@@
 \caption{Normalized codon entropy across the HIV genome is clustered around
-variable regions in {\it env}.}
+variable regions in {\it env}. hence the synonymous mutational space is explored
+deeper by HIV in that region, and even deleterious synonymous mutations can
+reach high frequencies.
+On the contrary, regions that are under purifying selection either
+the the RNA level (the RRE hairpin) or at the protein level but in a different
+reading frame (second {\it rev} exon, which includes the second {\it tat} exon),
+show reduced entropy, i.e. more constrains.
+All parts of {\it env} that are part of a different gene (signaling peptide,
+second {\it rev} exon) have been excluded from our main analysis, to avoid
+contamination about synonymity.
+Data from a multiple sequence alignment of subtype-B HIV sequences from the LANL
+website (filtered sequences only, version 2011)~\cite{LANL2012}.}
 \label{fig:syndiv_genome}
@@
 
-Our analysis relies on the ability of synonymous mutation to reach high
-frequencies within the viral population (by hitchhiking). The ability of HIV to
-explore the synonymous mutational space can be quantified by the codon entropy.
-Using a multiple sequence alignment of subtype-B HIV sequences from the LANL
-website (filtered sequences only, version 2011)~\cite{LANL2012}, we show in
-\figurename~\ref{fig:syndiv_genome} that the genomic regions within and flanking
-V loops contain the highest codon entropy (normalized by the maximal entropy for
-each degenerate codon). On the contrary, regions that are under selection either
-the the RNA level (the RRE hairpin) or at the protein level but in a different
-reading frame (rev second exon), show reduced entropy, i.e. more constrains.
-
-All parts of {\it env} that are part of a different gene have been excluded from
-our main analysis, to avoid contamination about synonymity.
+\clearpage
 
@@
 epitope both reduce fixation of nonsynonymous mutations, without affecting the
-synonymous curve much.}
-\label{fig:tds_wec}
-\end{center}
-\end{figure}
-As mentioned in the main text, both competition of different mutants within the
-same allele and a slow, catching-up immune system can reduce the fixation
-probability of nonsynonymous derived alleles. This is observed in computer
-models, an example of which is shown in \figurename~\ref{fig:tds_wec}. In the
+synonymous curve much. In the
 curves denoted with ``tds'' (time-dependent selection), the fitness advantage of
@@
 competition scenario instead, with 3 or 6 escape mutations within the same
-epitope of length 9 nucleotides. Increasing either the coefficient $c$ or the
-number of escape mutants per epitope reduces fixation of nonsynonymous derived
-alleles. Importantly, the synonymous fixation probability is hardly affected, which casts
-trust over the main conclusions of the main text independently on the details of
-the selection scenario at the protein level.
-
+epitope of length 9 nucleotides.}
+\label{fig:tds_wec}
+\end{center}
+\end{figure}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2013-02-10 synmut.tex
supplement only captions, more details about simulations. need to redo simulations without complex epitopes for 4B

@@
 \label{fig:fixp2}}
-\caption{Fixation and loss of synonymous mutations. \comment{fix x-axis label}
+\caption{Fixation and loss of synonymous mutations.
 Panel A) Time
@@
 likely to revert and finally be lost from the population. The
-average over all mutations that rare not observed  (green) lies between the
-those that fix or get lost.
+average over all mutations that are not observed (green) lies between the
+those that fix and those that get lost.
 To test the hypothesis that mutations in C2-V5 are lost because they break stems in the
@@
 substantially higher SHAPE reactivities than mutations that reach frequencies above 15\%, but are
-susequently lost (restricted to the region V1-V5$\pm 50bp$)????. This difference
+susequently lost (restricted to the regions V1-V5$\pm 100bp$ from each). This difference
 is quantified by a cumulative histogram ($p=0.002$, Kolmogorov-Smirnov two 
@@
 for simple epitopes? I can't seem to find a way to describe this from the
-outset}.
+outset}. \comment{FIXME: (i) only x acis is logarithmic; (ii) redo without
+epitopes (less representative in the end, but easier to introduce).}
 This simulation assumes 6 roughly equivalent escape mutations per epitope; see supplement for alternative models.}
@@
 Conserved RNA structures exist in different parts of the HIV genome (several are
-known), but our analysis requires longitudinal data most of which focusses
-on \env.
+known), but our analysis requires longitudinal data, most of which focuses
+on \env{}.
 Furthermore, we only expect to see deleterious synonymous
@@
 show substantial population structure and were discarded (see supplementary
-materials); a total of 11 patients with 6-?? time points each and 10
+materials); a total of 11 patients with 4-23 time points each and approximately 10
 sequences per time point were analyzed. The time intervals between two
-consecutive sequences ranged from 6 to 18 months. \comment{CHECK VALUES}.
+consecutive sequences ranged from 1 to 34 months, most of them approximately 6 to 10 months.
 
@@
 extinction) was stored. The fixation probability was determined as the long-time
-limit of the resulting curves. Samples less than ???days
-prior the last time point were omitted since there was not enough information
-on their future fate, i.e., whether they fix or are lost.
+limit of the resulting curves. Mutations that reached high frequency but neither
+fixed nor got extinct were classified as ``floating'' and not analyzed further.
 
@@
 of the NL4.3 virus used in \citep{watts_architecture_2009}. Problematic
-assignments in indel-rich regions are excluded from the analysis.
-\comment{The following is super confusing}
-In order to restrict the analysis to synonymous polymorphisms, a lower frequency
-threshold of 0.15 was used (other thresholds yield the same results). Since very
-few polymorphisms hitchhike beyond, say, a frequency of 0.5, this pool is
-enriched for to-be-lost mutations; hence the ``lost" curve in \FIG{SHAPEA}
-contains much more points than the ``fixed" one.
+assignments in indel-rich regions were excluded from the analysis.
 
-The variable loops and flanking regions are identified manually starting from
+The variable loops and flanking regions were identified manually starting from
 the annotated reference HXB2 sequence from the LANL HIV database~\citep{LANL2012}. A
-similar approach is used to label the C2-V5 region sequenced in
+similar approach was used to label the C2-V5 region sequenced in
 ref.~\citep{shankarappa_consistent_1999}.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-Simulations are performed using FFPopSim~\citep{zanini_ffpopsim:_2012}. Both full-length HIV genomes and
-\env{}-only simulations are performed and yield comparable results. In each
-run, a random fitness landscape with specified statistical properties 
-(e.g.~density of beneficial sites, average deleterious effect of synonymous changes) is generated.
-For each set of parameters, approximately 100 simulation runs are averaged over. 
-\comment{needs a little bit more detail}.
-For the discussion of simulation parameters, the areas below or above the neutral
-diagonal were estimated from the binned fixation probabilities using linear
+Computer simulations were performed using
+FFPopSim~\citep{zanini_ffpopsim:_2012}. Both full-length HIV genomes and
+\env{}-only simulations were performed and yielded comparable results. In each
+simulation, four random parameters were drawn form prior distributions: (i)
+fraction of deleterious synonymous sites (flat between 0.75 and 1.0); (ii)
+deleterious effect at those sites (logarithmic between $10^{-4}$ and $10^{-2}$);
+(iii) average beneficial effect of an escape mutation (logarithmic between
+$10^{-2.5}$ and $10^{-1.5}$); (iv) rate of introduction of new epitopes
+(logarithmic between $10^{-3}$ and $10^{-2}$ per generation). After these four
+parameters were fixed, the simulation started. All the deleterious synonymous
+sites had the same fitness effect (variations were tested and produced the same
+results), and no epitopes (i.e. no escape mutations) were initially present.
+After 30 generations of burn-in to create some genetic diversity, new epitopes
+were introduced at a constant rate (as specified above). Each time such an
+epitope was actually introduced, its fitness advantage was drawn from an
+exponential distribution with fixed average value (as specified above).
+For the models with competition within epitopes, a complex epistatic fitness
+landscape was manually crafted, in which each single mutant provides the whole
+beneficial effect of escape. For the models with time-dependent selection due to
+the immune system catching up, the beneficial effect of an escape mutation was
+reverted at random after it had appeared, according to its integrated allele
+frequency (see supplementary materials for more details).
+
+For each set of parameters, fixation probabilities and probabilities of
+synonymous polymorphisms $P_\text{interm}$ were calculated as averages over
+approximately 100 repetitions (with different random seeds).
+
+The areas below or above the neutral fixation probability (diagonal line)
+were estimated from the binned fixation probabilities using linear
 interpolation between the bin centers. This measure is sufficiently precise for
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-All analysis scripts, as well as the sequence alignment used, are available for download.
+All analysis and computer simulation scripts, as well as the sequence alignments
+used, are available for download.
 

### 2013-02-11 supplementary.tex
new Fig 4B plus some editing

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Synonymous diversity across the HIV genome}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=\linewidth]{conservation_codons_genome}
+\caption{Synonymous diversity across the HIV genome, as quantified by the
+normalized codon entropy. The synonymous diversity peaks at the variable
+regions in {\it env} and is reduced in regions under purifying selection (RRE
+hairpin, second {\it tat}/{\it rev} exons). Since the variable regions are under
+positive selection at the protein level, synonymous hitchhiking is stronger
+there, and even deleterious synonymous mutations can be brought to high
+frequency. This gives to our signal in the fixation probability, Figs. 2A and
+2B of the main text.
+The normalized codon entropy is calculated as follows (see the script
+for the full algorithm): (i) from a subtype B multiple sequence alignment
+(MSA) from the LANL website (filtered sequences only, version 2011)~\cite{LANL2012},
+we calculate the consensus amino acid at each position in the HIV genome; (ii) we count
+how often each codon coding for the consensus amino acid appears in the MSA;
+(iii) at each amino acid position, we divide by the number of sequences in the
+MSA that had the consensus amino acid at that position, obtaining {\it codon
+frequencies} $\nu_c$; (iv) we calculate the codon entropy from each position as:
+$S := - \sum_{c \in \Sigma} \nu_c \log \nu_c$, where $\Sigma$ is the codon
+alphabet for that particular amino acid; (v) we divide by the maximal codon entropy of
+that amino acid (e.g. $\log 2$ for twofold degenerate codons).
+All parts of {\it env} that are part of a different gene (signaling peptide,
+second {\it rev} exon) have been excluded from our main analysis, to avoid
+contamination about synonymity.}
+\label{fig:syndiv_genome}
+\end{center}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Nonsynonymous changes outside of variable regions are deleterious}
@@
 \begin{center}
-\includegraphics[width=0.5\linewidth]{synmut_conservation_4fold_synnonsyn}
+\includegraphics[width=0.4\linewidth]{synmut_conservation_4fold_synnonsyn}
 \caption{Cumulative distribution of synonymous and nonsynonymous diversity in
-{\it gag} in the LANL reference panel(filtered
+{\it gag} in the LANL reference panel (filtered
 sequences only, version 2011)~\cite{LANL2012}. Sites such that the consensus codon
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Synonymous diversity across the HIV genome}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=\linewidth]{conservation_codons_genome}
-\caption{Normalized codon entropy across the HIV genome is clustered around
-variable regions in {\it env}. hence the synonymous mutational space is explored
-deeper by HIV in that region, and even deleterious synonymous mutations can
-reach high frequencies.
-On the contrary, regions that are under purifying selection either
-the the RNA level (the RRE hairpin) or at the protein level but in a different
-reading frame (second {\it rev} exon, which includes the second {\it tat} exon),
-show reduced entropy, i.e. more constrains.
-All parts of {\it env} that are part of a different gene (signaling peptide,
-second {\it rev} exon) have been excluded from our main analysis, to avoid
-contamination about synonymity.
-Data from a multiple sequence alignment of subtype-B HIV sequences from the LANL
-website (filtered sequences only, version 2011)~\cite{LANL2012}.}
-\label{fig:syndiv_genome}
-\end{center}
-\end{figure}
-
-\clearpage
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Alternative models for suppression of nonsynonymous mutations}
@@
 months from its appearance) and competition between escape mutations in the same
-epitope both reduce fixation of nonsynonymous mutations, without affecting the
-synonymous curve much. In the
-curves denoted with ``tds'' (time-dependent selection), the fitness advantage of
+epitope both reduce fixation of nonsynonymous mutations. A ``na\"ive'' computer
+simulation shows a depression in synonymous fixation probability, as observed in
+patients, but the nonsynonymous fixation probability is much higher than the
+(neutral) diagonal: a nonsynonymous mutation that reaches a frequency around
+0.2 is already very likely to fix. In data from infected patients, however,
+nonsynonymous mutations show a neutral-like curve,
+$P_\text{fix}(\nu) \approx \nu$. In the other curves, we show that the
+nonsynonymous fixation probability can be reduced by either time-dependent
+selection or competition between equivalent escapes, without affecting much the
+synonymous fixation probability itself.
+In the curves denoted with ``tds'' (time-dependent selection), the fitness advantage of
 an escape mutaition can vanish at each time point with a probability density
@@
 efficiency of the host immune system, and $\nu(t)$ is the frequency of the
-escape allele at time $t$. The curves denoted with ``mut/epi'' describe the
+escape allele at time $t$.
+In the curves denoted with ``mut/epi'' we test the
 competition scenario instead, with 3 or 6 escape mutations within the same
-epitope of length 9 nucleotides.}
+epitope of length 9 nucleotides. The fitness landscape of each such epitope
+include negative epistatic terms, so that the joint presence of more than one
+escape mutation is not any more beneficial for the virus than a single mutation.
+In this latter scenario, many escape mutations start to sweep on different
+backgrounds within the viral population, but eventually compete and only one of
+them if fixed.}
 \label{fig:tds_wec}

### 2013-02-11 synmut.tex
new Fig 4B plus some editing

@@
 assumed to be neutral.
-We analyze longitudinal intrapatient data from the \shankaregion~part of the
-envelope gene (\env) and observe that synonymous derived alleles rarely
+We analyze longitudinal intrapatient data from the \shankaregion{} part of the
+envelope gene (\env{}) and observe that synonymous derived alleles rarely
 fix even though they often reach high frequencies in the viral population.
@@
 The patterns of fixation of nonsynonymous mutations suggest that antibody
-escape mutations in \shankaregion~are only transiently beneficial, either since
-the immune system is catching up or because of competition between equivalent
+escape mutations in \shankaregion{} are only transiently beneficial, either
+because the immune system is catching up or because of competition between equivalent
 escapes.
@@
 One possible {\it a priori} explanation for lack of fixation of synonymous
-mutations in C2-V5 are  secondary structures in the viral RNA whose disruption
-is deleterious to the virus.
+mutations in C2-V5 are secondary structures in the viral RNA, the disruption of
+which is deleterious to the virus.
 It has been suggested early on that parts of the viral genome that have the
@@
 those that fix and those that get lost.
+Note that, because of small differences in base pairing between the reference
+strain NL4-3 and each patient's initial consensus sequence, this analysis is not
+expected to be very sensitive, but very conservative.
+
 To test the hypothesis that mutations in C2-V5 are lost because they break stems in the
@@
 generation). Mostly simulations with small deleterious effects and relatively
-high deleterious fractions of synonymous mutations reproduce synonymous HIV-like
-fixation probability and diversity. Parameters are chosen uniformly on a
-logarithmic scale from the indicated windows. \comment{Can't we show this
-for simple epitopes? I can't seem to find a way to describe this from the
-outset}. \comment{FIXME: (i) only x acis is logarithmic; (ii) redo without
-epitopes (less representative in the end, but easier to introduce).}
-This simulation assumes 6 roughly equivalent escape mutations per epitope; see supplement for alternative models.}
+high deleterious fractions of synonymous mutations reproduce synonymous the
+fixation probability and diversity observed in patients. Parameters are chosen
+from uninformed prior distributions in the indicated window (see methods).}
 \label{fig:simheat}
@@
 in chronic infections have been estimated to be on the order of 
-$s_a = 0.01$ per day~\citep{neher_recombination_2010,Asquith}.
+$s_a = 0.01$ per day~\citep{neher_recombination_2010, Asquith}.
+\comment{Asquith??}
 
@@
 sequences per time point were analyzed. The time intervals between two
-consecutive sequences ranged from 1 to 34 months, most of them approximately 6 to 10 months.
+consecutive sequences ranged from 1 to 34 months, most of them between 6 and 10
+months.
 
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Sequence analysis}
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 The sequences were translated and the resulting amino-acid sequences aligned
@@
 state. Codons with more than one mutation were discarded. Slightly different
-criteria for synonymous/nonsynonymous discrimination yield similar results.
+criteria for synonymous/nonsynonymous discrimination yielded similar results.
 
@@
 sequence alignments for each patient through the alignment to the sequence
-of the NL4.3 virus used in \citep{watts_architecture_2009}. Problematic
+of the NL4.3 virus used in ref.~\citep{watts_architecture_2009}. Problematic
 assignments in indel-rich regions were excluded from the analysis.
@@
 fraction of deleterious synonymous sites (flat between 0.75 and 1.0); (ii)
-deleterious effect at those sites (logarithmic between $10^{-4}$ and $10^{-2}$);
-(iii) average beneficial effect of an escape mutation (logarithmic between
+deleterious effect at those sites (exponential between $10^{-4}$ and $10^{-2}$);
+(iii) average beneficial effect of an escape mutation (exponential between
 $10^{-2.5}$ and $10^{-1.5}$); (iv) rate of introduction of new epitopes
-(logarithmic between $10^{-3}$ and $10^{-2}$ per generation). After these four
+(exponential between $10^{-3}$ and $10^{-2}$ per generation). After these four
 parameters were fixed, the simulation started. All the deleterious synonymous
 sites had the same fitness effect (variations were tested and produced the same
-results), and no epitopes (i.e. no escape mutations) were initially present.
-After 30 generations of burn-in to create some genetic diversity, new epitopes
+results), no epitopes (i.e. no escape mutations) were initially present, and the
+population was isomorphic.
+After 30 generations of burn-in to create genetic diversity, new epitopes
 were introduced at a constant rate (as specified above). Each time such an
@@
 the immune system catching up, the beneficial effect of an escape mutation was
-reverted at random after it had appeared, according to its integrated allele
+reverted at random times after it had appeared, according to its integrated allele
 frequency (see supplementary materials for more details).

### 2013-02-11 scripts/codon_entropy_synonymous_subtypeB.py

@@ -6,6 +6,7 @@ content:    Measure the level of conservation of synonymous mutations in subtype
             B.
 '''
 # Modules
+import re
 import numpy as np
 import matplotlib.pyplot as plt
 from matplotlib.patches import Rectangle
@@ -40,97 +41,106 @@ def smoothen(x, f):
 # Script
 if __name__ == '__main__':
 
-    # Load data for a certain gene
-    gene = genes[0]
-    ali = alignment.MSA(gene=gene, reference=reference)
-    msa = np.array(ali.MSA)
-    afs = ali.allele_frequencies()
+    # Store data
+    data = {}
 
-    # Filter only part of pol
-    if gene == 'pol':
-        msa = msa[:, :2718]
-        afs = afs[:, :2718]
+    # Make figure for all four genes
+    fig, axs = plt.subplots(1, 4)
 
-    # Find consensus sequence
-    consind = afs.argmax(axis=0)
-    consensus = alpha[consind]
+    for k, gene in enumerate(genes):
 
-    # Some seqs are not viable and include frameshifts that mess up the
-    # translation, hence we restrict to positions for which gaps are a minority
-    is_gap = consensus == '-'
-    # Exclude full codons
-    tmp = np.unique(is_gap.nonzero()[0] / 3)
-    is_gap[np.concatenate([tmp * 3, tmp * 3 +1, tmp * 3 + 2])] = True
+        print gene
 
-    # Exclude stop codons
-    is_stop = np.zeros_like(is_gap)
-    tmp = (translate(consensus) == '*').nonzero()[0]
-    is_stop[np.concatenate([tmp * 3, tmp * 3 +1, tmp * 3 + 2])] = True
+        # Select the axis
+        ax = axs[k]
+        plt.sca(ax)
 
-    ## Plot base prevalence
-    #for i in xrange(4):
-    #    plt.plot(np.arange(len(consensus)), afs[i],
-    #             lw=1.5, alpha=0.5)
-    #plt.xlim(0, 2600)
-    #plt.ylim(-0.05, 1.25)
-    #plt.xlabel('position in '+gene)
-    #plt.ylabel('allele frequency')
-    #plt.legend(alpha, loc=9)
-
-    # Good are codons with no gaps and no stops
-    is_good = (-is_gap) & (-is_stop)
-
-    # For each codon, calculate the entropy
-    msa = msa[:, is_good]
-    consaa = translate(consensus[is_good])
-    entropy = np.zeros(len(consaa))
-    from collections import Counter
[…]

### 2013-02-11 scripts/simulations_scatter_all.py

@@ -16,7 +16,7 @@ import matplotlib.cm as cm
 
 
 # Globals
-results_dir = '../data/simulations/random_gradual_nonsyndead_6epitopes/'
+results_dir = '../data/simulations/random_gradual_nonsyndead/'
 datafile = 'parameters_fixation_areas.dat'
 
 # Set limits for data-like sims
@@ -82,25 +82,26 @@ if __name__ == '__main__':
     colors = cm.jet([int(255.0 * (i + 2.5)) for i in np.log10(tbg['ada_eff'])])
     sizes = 10 + 120 * (np.log10(tbg['ada_rate']) + 3)
     ind = sizes.argsort()[::-1]
-    plt.scatter(np.log10(tbg['del_eff'])[ind], tbg['del_frac'][ind],
+    plt.scatter(tbg['del_eff'][ind], tbg['del_frac'][ind],
                 color=colors[ind], s=sizes[ind])
-    plt.xlabel(r'$\log_{10}( \text{deleterious effect} )$', fontsize=16)
+    plt.xlabel(r'deleterious effect', fontsize=16)
     plt.ylabel(r'deleterious fraction', fontsize=16)
     plt.gca().xaxis.set_label_coords(0.5, -0.07)
     plt.gca().yaxis.set_label_coords(-0.1, 0.5)
-    plt.xlim(-4.1, -1.9)
+    plt.xscale('log')
+    plt.xlim(10**(-4.1), 10**(-1.9))
     plt.ylim(0.72, 1.02)
     plt.title(('color is adaptive effect (1e-2.5 to 1e-1.5)\n'+
                'size is rate of new epitopes (1e-3 to 1e-2)'), fontsize=14)
     
     # Draw a rectangle around the parameter priors
-    rec = Rectangle(xy=(-4, 0.75),
-                    width=2,
+    rec = Rectangle(xy=(10**(-4), 0.75),
+                    width=0.0099,
                     height=0.25,
                     facecolor='none', edgecolor='red',
                     lw=1.5)
     plt.gca().add_patch(rec)
-    txt2 = plt.text(-2.25, 0.735, 'flat prior', color='red')
+    txt2 = plt.text(10**(-2.25), 0.735, 'flat prior', color='red')
 
     plt.ion()
     plt.show()

### 2013-02-11 synmut.tex
discussion reshuffled

@@
 
+The fixation and extinction times and probabilities represent a rich and simple
+summary statistics useful to characterize longitudinal sequence data and compare
+to models via computer simulations. Although several recent publications have
+pointed out that synonymous changes in HIV can be
+deleterious~\citep{lemey_synonymous_2007, ngandu_extensive_2008}, our analysis
+is much more direct, as we rely neither on phylogenetic tree reconstruction nor
+on macroevolutionary models. A method that is similar to ours {\it in spiritu}
+has been recently used in a longitudinal study of influenza
+evolution~\citep{strelkowa_clonal_2012}. The propagators suggested in that
+article, however, represent ratios between nonsynonymous and synonymous
+mutations.  The latter is used as an approximately neutral control; this method
+can therefore not be used to investigate synonymous changes themselves. A
+functional significance of the insulating RNA structure stems between the
+hyper-variable loops has also been proposed
+previously~\citep{watts_architecture_2009, sanjuan_interplay_2011} and conserved
+RNA structures exist in different parts of the HIV
+genome~\citep{ngandu_extensive_2008}. Our analysis is nonetheless much more
+direct and is able to quantify the fitness effect of RNA structure within single
+infections. The main limitations of our approach are the following: (i) it
+requires longitudinal data, most of which focuses on \env{} only; and (ii) we
+only expect to see deleterious synonymous mutations at high frequencies if they
+are linked to repeated adaptive substitions. Otherwise, most synonymous
+polymorphisms remain at low frequencies and can only be observed by deep
+sequencing methods. We indeed plan to exploit next-generation sequencing
+technologies to assess synonymous diversity at the level of rare variants in
+future studies.
+
 Contrary to na\"ive expectations, the adaptive escape mutations do not seem to be
@@
 possibilities and find that both produce the desired effect. Furthermore, there
-is experimental evidence in support of both of these hypotheses. Serum from HIV infected
-individuals tyically neutralizes the virus that dominated the population a few
-(3-6) month ago \citep{richman_rapid_2003}. This suggests that escape mutations
-cease to be beneficial after a few months and might revert if they come with a
-fitness cost. Deep sequencing of regions of \env{} after antibody escape have
-revealed multiple escape mutations in the same epitope
+is experimental evidence in support of both of these hypotheses. Serum from HIV
+infected individuals tyically neutralizes the virus that dominated the
+population a few (3-6) month ago \citep{richman_rapid_2003}. This suggests that
+escape mutations cease to be beneficial after a few months and might revert if
+they come with a fitness cost. Deep sequencing of regions of \env{} after
+antibody escape have revealed multiple escape mutations in the same epitope
 \citep{moore_limited_2009, bar_early_2012}. Presumably, each one of these
@@
 frequencies. The rapid emergence of multiple escape mutations in the same
-epitope implies a large population size that explores all necessary point mutations rapidly. A
-similar point has been made recently by Boltz {\it et al.} in the context of
-preexisting drug resistance mutations~\citep{boltz_ultrasensitive_2012}.
+epitope implies a large population size that explores all necessary point
+mutations rapidly. A similar point has been made recently by Boltz {\it et al.}
+in the context of preexisting drug resistance
+mutations~\citep{boltz_ultrasensitive_2012}.
 
@@
 
-A functional significance of the insulating RNA structure stems between the
-hyper-variable loops has been proposed
-previously~\citep{watts_architecture_2009, sanjuan_interplay_2011}.
-Conserved RNA structures exist in different parts of the HIV genome (several are
-known), but our analysis requires longitudinal data, most of which focuses
-on \env{}.
-Furthermore, we only expect to see deleterious synonymous
-mutations at high frequencies if they are linked to repeated adaptive
-substitions. Otherwise, they remain at low frequencies and can only be observed
-by deep sequencing methods.
-
-The fixation and extinction times and probabilities represent a rich and simple
-summary statistics useful to characterize longitudinal sequence data and compare
-to models via computer simulations.
-A similar method has been recently used in a longitudinal study of
-influenza evolution~\citep{strelkowa_clonal_2012}. The propagators suggested in
-that paper, however, represent ratios between nonsynonymous and synonymous mutations.
-The latter is used as an approximately neutral control and this method can
-therefore not be used to investigate synonymous changes themselves.
-
-Our results emphasize the inadequacy of independent site
-models of HIV evolution and the common assumption that selection is time
-independent or additive. If genetic variation is only transiently beneficial,
-existing estimates  of the strength of selection
-\citep{neher_rate_2010,batorsky_estimate_2011} could
-be substantial underestimates.
+Our results emphasize the inadequacy of independent site models of HIV evolution
+and the common assumption that selection is time independent or additive. If
+genetic variation is only transiently beneficial, existing estimates of the
+strength of selection \citep{neher_rate_2010,batorsky_estimate_2011} could be
+substantial underestimates.
 

### 2013-02-12 supplementary.tex
edits throughout, reorganized discussion and supplement

@@
 \includegraphics[width=0.35\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p7}
-\caption{Panel A) PCA of all sequences from patient p1 (colors indicate time from seroconversion,
-from blue to red). Panel B) Allele frequency trajectories for nonsynonymous
+\caption{Structure of viral population and patient selection.
+Panel A) shows a PCA of all sequences from patient p1 (colors indicate time from
+seroconversion, from blue to red). Panel B) Allele frequency trajectories for nonsynonymous
 changes in the same patient. Panels C and D) show analogous plots for data from
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Synonymous diversity across the HIV genome}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{figure}[h]
@@
 \caption{Synonymous diversity across the HIV genome, as quantified by the
-normalized codon entropy. The synonymous diversity peaks at the variable
-regions in {\it env} and is reduced in regions under purifying selection (RRE
+normalized codon entropy among sequences coding for the
+consensus amino-acid.
+The synonymous diversity peaks at the variable regions in {\it env} and is reduced in regions under purifying selection (RRE
 hairpin, second {\it tat}/{\it rev} exons). Since the variable regions are under
@@
 frequencies} $\nu_c$; (iv) we calculate the codon entropy from each position as:
-$S := - \sum_{c \in \Sigma} \nu_c \log \nu_c$, where $\Sigma$ is the codon
-alphabet for that particular amino acid; (v) we divide by the maximal codon entropy of
-that amino acid (e.g. $\log 2$ for twofold degenerate codons).
+$S := - \sum_{c} \nu_c \log \nu_c$, where $c$ runs over codons that code for
+the consensus amino acid at this site; (v) we divide by the maximal codon
+entropy of that amino acid (e.g. $\log 2$ for twofold degenerate codons).
 All parts of {\it env} that are part of a different gene (signaling peptide,
@@
 \end{figure}
+\newpage
+% 
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \section{Nonsynonymous changes outside of variable regions are deleterious}
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \begin{figure}[h]
+% \begin{center}
+% \includegraphics[width=0.8\linewidth]{synmut_conservation_4fold_synnonsyn}
+% \caption{Cumulative distribution of synonymous and nonsynonymous diversity in
+% {\it gag} in the LANL reference panel (filtered
+% sequences only, version 2011)~\cite{LANL2012}. Sites such that the consensus codon
+% has three synonymous and six nonsynonymous single mutants were used, and the number of observed mutants
+% of a certain type over the number of possible mutants is plotted. Nonsynonymous
+% changes are observed less often, {\it ergo} are more conserved, than
+% synonymous changes. It can be therefore assumed, as mentioned in the main text,
+% that non-escape nonsynonymous changes involve a large fitness cost.}
+% \label{fig:synnonsyncons}
+% \end{center}
+% \end{figure}
+% \newpage
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Nonsynonymous changes outside of variable regions are deleterious}
+\section{Time dependent selection}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{center}
-\includegraphics[width=0.4\linewidth]{synmut_conservation_4fold_synnonsyn}
-\caption{Cumulative distribution of synonymous and nonsynonymous diversity in
-{\it gag} in the LANL reference panel (filtered
-sequences only, version 2011)~\cite{LANL2012}. Sites such that the consensus codon
-has three synonymous and six nonsynonymous single mutants were used, and the number of observed mutants
-of a certain type over the number of possible mutants is plotted. Nonsynonymous
-changes are observed less often, {\it ergo} are more conserved, than
-synonymous changes. It can be therefore assumed, as mentioned in the main text,
-that non-escape nonsynonymous changes involve a large fitness cost.}
-\label{fig:synnonsyncons}
+\includegraphics[width=0.8\linewidth]{simulations_graduallyepitopesandtimeselec}
+\caption{Time-dependent selection (the escape mutant is recognized after some
+months from its appearance) and competition between escape mutations in the same
+epitope both reduce fixation of nonsynonymous mutations. A ``na\"ive'' computer
+simulation shows a depression in synonymous fixation probability, as observed in
+patients, but the nonsynonymous fixation probability is much higher than the
+(neutral) diagonal: a nonsynonymous mutation that reaches a frequency around
+0.2 is already very likely to fix. In data from infected patients, however,
+nonsynonymous mutations show a neutral-like curve,
+$P_\text{fix}(\nu) \approx \nu$. In the other curves, we show that the
+nonsynonymous fixation probability can be reduced by either time-dependent
+selection or competition between equivalent escapes, without affecting much the
+synonymous fixation probability itself.
+In the curves denoted with ``tds'' (time-dependent selection), the fitness advantage of
+an escape mutaition can vanish at each time point with a probability density
+\[ P_\text{recognized}(t) = c \cdot \nu(t), \]
+where $c$ is a constant coefficient shown in the legend that encodes the overall
+efficiency of the host immune system, and $\nu(t)$ is the frequency of the
+escape allele at time $t$.
+In the curves denoted with ``mut/epi'' we test the
+competition scenario instead, with 3 or 6 escape mutations within the same
+epitope of length 9 nucleotides. The fitness landscape of each such epitope
+include negative epistatic terms, so that the joint presence of more than one
+escape mutation is not any more beneficial for the virus than a single mutation.
+In this latter scenario, many escape mutations start to sweep on different
+backgrounds within the viral population, but eventually compete and only one of
+them if fixed.}
+\label{fig:tds_wec}
 \end{center}
@@
 
+\newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Alternative models for suppression of nonsynonymous mutations}
+\section{Within epitope competition}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{center}
-\includegraphics[width=0.5\linewidth]{simulations_graduallyepitopesandtimeselec}
+\includegraphics[width=0.8\linewidth]{simulations_graduallyepitopesandtimeselec}
 \caption{Time-dependent selection (the escape mutant is recognized after some

### 2013-02-12 synmut.tex
edits throughout, reorganized discussion and supplement

@@
 \newcommand{\mexpfit}{\langle e^{F}\rangle}
-\newcommand{\pfix}{P_{fix}}
+\newcommand{\pfix}{P_{\mathrm{fix}}}
 \newcommand{\ox}{r}
@@
 \newcommand{\shankaregion}{C2-V5}
+\newcommand{\timedependence}{4}
+\newcommand{\withinepi}{4}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 infection, the most rapidly evolving part of the HIV genome are the so called
-variable loops of the envelope protein gp120, which need to avoid recognition by
-neutralizing ABs.  Mutations in \env, the gene encoding gp120, spread
+variable loops of the envelope protein gp120, which change to avoid recognition
+by neutralizing ABs.  Mutations in \env, the gene encoding gp120, spread
 through the population within a few months (see \figurename~\ref{fig:aftnonsyn}).
@@
 
-These escape mutations are strongly selected for their effect on the amino acid
+These escape mutations are selected for their effect on the amino acid
 sequence of the viral proteins. Conversely, synonymous mutations are commonly
@@
 species \citep{plotkin_synonymous_2011}. 
-Recent studies have shown that genetically engineered HIV strains with
-altered codon usage can in some cases produce more viral protein, but in general
-replicate less efficiently~\citep{ngumbela_quantitative_2008,
+Recent studies have shown that genetically engineered HIV strains with altered codon usage can in some cases produce more viral protein, but in general replicate less efficiently~\citep{ngumbela_quantitative_2008,
 li_codon-usage-based_2012,keating_rich_2009}.
@@
 selection beyond the protein sequence is therefore expected
-\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}. Positive
-selection through the host's immune system, however, is restricted to
+\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011} and it has been
+shown that rates of evolution at synonymous sites
+vary along the HIV genome \citep{mayrose_towards_2007}.
+Positive selection through the host's immune system, however, is restricted to
 changes in the amino acid sequence.
@@
 \citet{shankarappa_consistent_1999,liu_selection_2006} and 3 patients from
-\citet{bunnik_autologous_2008} (see methods). The data set from
-\citet{shankarappa_consistent_1999,liu_selection_2006} is restricted to the
-C2-V5 region of \env, while \citet{bunnik_autologous_2008} covers the
-majority of \env. Considering all mutations in a
-frequency interval around $\nu_0$ at some time $t_i$, we calculate the fraction
+\citet{bunnik_autologous_2008} (patients whose viral population was
+structured were excluded from the analysis; see methods).
+The former data set from is restricted to the
+C2-V5 region of \env, while the data from \citet{bunnik_autologous_2008} covers
+the majority of \env. Considering all mutations in a
+frequency interval around $\nu_0$ at some time $t$, we calculate the fraction
 that is found at frequency 1, at frequency 0, or at intermediate frequency at
-later times $t_f$. Plotting these fixed, lost, and polymorphic fraction against
-the time interval $t_f-t_i$, we see that most synonymous mutations segregate for
-roughly one year and are lost much more frequently than expected (panel
+later times $t+\Delta t$. Plotting these fixed, lost, and polymorphic fraction
+against the time interval $\Delta t$, we see that most synonymous mutations
+segregate for roughly one year and are lost much more frequently than expected (panel
 \ref{fig:fixp1}). The long-time probability of fixation versus extinction of
@@
 \label{fig:aftnonsyn}}
-\caption{Time series of frequencies of synonymous (A) and non-synonymous (B)
-derived alleles in \env, C2-V5, from patient
+\caption{\comment{I would like p2 or p10 better here} Time series of frequencies
+of synonymous (A) and non-synonymous (B) derived alleles in \env, C2-V5, from patient
 1~\cite{shankarappa_consistent_1999}.
@@
 \caption{Fixation and loss of synonymous mutations.
-Panel A) Time
+Panel A) shows the time
 course of loss and fixation of synonymous mutations observed in a frequency interval $\nu_0$. 
@@
 Panel B) Fixation probability of derived synonymous
-alleles as a function of $\nu_0$ in C2-V5 is below the neutral
+alleles as a function of $\nu_0$ in C2-V5 is reduced relative to the neutral
 expectation indicated by the diagonal line. This suppression is not
-oberserved in other parts of {\it env} or for nonsynonymous mutations.
+observed in other parts of \env~or for nonsynonymous mutations.
 The horizontal error bars on the abscissa are bin sizes, the vertical ones the
 standard deviation after 100 patient bootstraps of the data. Data from
-Refs.~\cite{shankarappa_consistent_1999, bunnik_autologous_2008}.}
+Refs.~\cite{shankarappa_consistent_1999,liu_selection_2006, bunnik_autologous_2008}.}
 \label{fig:fixp}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-One possible {\it a priori} explanation for lack of fixation of synonymous
+One possible explanation for lack of fixation of synonymous
 mutations in C2-V5 are secondary structures in the viral RNA, the disruption of
-which is deleterious to the virus.
-It has been suggested early on that parts of the viral genome that have the
-potential to form RNA stems are better conserved than the
-remainder~\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}.
+which is deleterious to the virus
+~\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011,sanjuan_interplay_2011}.
 
-Recently, the propensity of nucleotides of the HIV genome to form base pairs has
+The propensity of nucleotides of the HIV genome to form base pairs has
 been measured using the SHAPE assay (a biochemical reaction preferentially
@@
 shown in \FIG{SHAPEA}, the reactivities of fixed
-alleles (red histogram) are systematically larger than of alleles are lost (blue
-) (Kolmogorov-Smirnov test, $P\approx 0.002$).
+alleles (red histogram) are systematically larger than of alleles that are lost
+(blue ) (Kolmogorov-Smirnov test, $P\approx 0.002$).
 In other words, alleles that are likely to break RNA helices are also more
@@
 those that fix and those that get lost.
-Note that, because of small differences in base pairing between the reference
-strain NL4-3 and each patient's initial consensus sequence, this analysis is not
-expected to be very sensitive, but very conservative.
+Note that, that this analysis will be sensitive only at position where the base
+pairing pattern of NL4-3 agrees with that of each patient's initial
+consensus sequence.
 
 To test the hypothesis that mutations in C2-V5 are lost because they break stems in the
-conserved streches between the variable loops, we consider mutations in
+conserved stretches between the variable loops, we consider mutations in
 variable loops and conserved parts separately. The biggest depression in
@@
 \subfloat{\includegraphics[height=0.46\linewidth]{fixation_probabilities_VnonV}\label{fig:SHAPEB}}
-\caption{Permissible synonymous mutations have
-high SHAPE reactivities.
-Panel A) Synonymous substitions have
-substantially higher SHAPE reactivities than mutations that reach frequencies above 15\%, but are
-susequently lost (restricted to the regions V1-V5$\pm 100bp$ from each). This difference
-is quantified by a cumulative histogram ($p=0.002$, Kolmogorov-Smirnov two 
-sample test). Mutations that are never
-observered show an intermediate distribution of SHAPE values.
-Panel B) Fixation probability of synonymous mutations in C2-V5
-separately for variable loops and the connecting conserved regions, that
-harbor conserved RNA stems. As expected, fixation probability is lower inside
-the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
+\caption{Permissible synonymous mutations tend to be unpaired.
+Panel A) shows the distribution of SHAPE reactivities among synonymous mutations
+that fix, mutations that reach frequencies above 15\%, but
+are susequently lost, and random mutations (all categories are restricted to the
+regions V1-V5$\pm 100$bp).
+Fixed mutations tend to have higher higher SHAPE reactivities, corresponding to
+less base pairing ($p=0.002$, Kolmogorov-Smirnov two sample test on the
+corresponding cumulative distribution).
+Mutations that are never observered show an intermediate distribution of SHAPE values.
+Panel B) shows the fixation probability of synonymous mutations in C2-V5
+separately for variable loops and the connecting conserved regions that
+harbor conserved RNA stems. As expected, the fixation probability is lower
+inside the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
 bunnik_autologous_2008, liu_selection_2006}.}
@@
 is not unexpected, it seems odd that we observe them at high population
-frequency -- at least in some regions of the genome. The region of \env~in
-which we observe deleterious mutations at high frequency undergoes frequent
+frequency and that the fixation probability is reduced only in parts
+of the genome.
+This regions, however, undergoes frequent
 adaptive changes to evade recognition by neutralizing 
@@
 fixation probability and diversity observed in patients. Parameters are chosen
-from uninformed prior distributions in the indicated window (see methods).}
+from prior distributions uniform in log-space indicated by the red box (see
+methods).}
 \label{fig:simheat}
@@
 in chronic infections have been estimated to be on the order of 
-$s_a = 0.01$ per day~\citep{neher_recombination_2010, Asquith}.
-\comment{Asquith??}
+$s_a = 0.01$ per day~\citep{neher_recombination_2010,Asquith:2006p28003}.
 
@@
 the area will be $-0.5$; if every mutation fixes, we will measure 
-$A= 0.5$. In HIV, we find $A_\text{syn} \sim -0.2$  for
-synonymous changes and $A_\text{nonsyn} \sim 0$ for nonsynonymous changes.
+$A= 0.5$. In HIV, we find $A_\text{syn} \approx -0.2$  for
+synonymous changes and $A_\text{nonsyn} \approx 0$ for nonsynonymous changes.
 
@@
 ($A_\text{syn} \sim -0.5$) as their fitness effect
-worsens; the synonymous diversity plummets as well, as it is harder for
-deleterious mutations to reach the frequency threshold of 0.25.
+worsens; the synonymous diversity plummets as well, as deleterious mutations
+are selected against.
 
@@
 $P_\text{interm}$ as observed in the data, i.e., a large depression in fixation
-probability of synonymous mutations but, simulteneously, a moderately high
+probability of synonymous mutations but, simultaneously, a moderately high
 synonymous diversity. Specifically, \FIG{simsfig} shows parameter combinations
@@
 0.010$. These conditions indicate that a high fraction ($\gtrsim 0.8$) of
-deleterious sites has to be deleterious with effect size $|s_d| \sim 0.002$. 
+sites has to be deleterious with effect size $|s_d| \sim 0.002$. 
 This result fits well the expectation based on the fixation/extinction times above (see \FIG{fixp1}).
@@
 has to be much smaller than the escape rate, otherwise the double mutant has
-little or no fitness advantage; (iii) mutations with a deleterious effect
-smaller than approximately $0.001$ behave neutrally consistent with the typical
-coalescent times observed in HIV.
+little or no fitness advantage. Consistent with this argument, larger
+deleterious effects  in \FIG{simheat} correspond to larger escapes rates. (iii)
+mutations with a deleterious effect smaller than approximately $0.001$ 
+behave neutrally consistent with the typical coalescent times observed in HIV.
 
-As long as only features of synonymous mutations are filtered, as performed
-above, the two nonsynonymous parameters, escape rate
-and rate of appearance of epitopes, are little constrained as is apparent in
-\FIG{simsfig} from the variety of sizes and colors indicating these parameters.
-In most of these simulations, however, nonsynonymous mutations almost always fix once they reach high frequencies
--- their nonsynonymous fixation area is well above zero. This is incompatible
-with the blue line in \FIG{fixp}: in an HIV infection, nonsynonymous
-mutations at high frequency often disappear again, even though many are at
-least transiently beneficial. Inspecting the trajectories of nonsynonymous
-mutations suggests the rapid rise and fall of many alleles. We test two possible
-mechanisms that are biologically plausible and could explain the transient rise
-of nonsynonymous mutations: time-dependent selection and within-epitope
-competition.
+The above simulation show that hitch-hiking is a plausible explanation for the
+obervation of deleterious mutations that rarely fix. However, in a simple model
+where non-synonymous escape mutations are unconditionally beneficial, they
+almost always fix once they reach high frequencies -- $A_{\mathrm{nonsyn}}$ is
+well above zero. This is incompatible with the blue line in \FIG{fixp}: in an HIV infection, nonsynonymous mutations at high frequency often
+disappear again, even though many are at least transiently beneficial.
+Inspecting the trajectories of nonsynonymous mutations suggests the rapid rise
+and fall of many alleles. We test two possible mechanisms that are biologically
+plausible and could explain the transient rise of nonsynonymous mutations:
+time-dependent selection and within-epitope competition.
 
@@
 \citet{richman_rapid_2003, bunnik_autologous_2008} report antibody responses to
-escape mutants. These respones are delayed by a few months, roughly matching the
+escape mutants. These responses are delayed by a few months, roughly matching the
 average time needed by an escape mutant to rise from low to high frequency.
+To model this type of behavior, we assume that anti-body
+responses against escape mutations arise with a rate proportional to the
+frequency against the escape mutation and abolish the benefit of the escape
+mutations. As expected, this type of time dependent
+fixation retains the potential for hitch-hiking, but reduces fixation of
+non-synonymous mutations. Fig.~S\timedependence~shows that $\pfix$ of
+synonymous mutations is not affected by this chance, while $\pfix$ of non-synonymous
+mutations approaches the diagonal as the rate of recognition of escape mutants
+is increased.
+
 In the alternative hypothesis, several different escape
@@
 shown~\citep{moore_limited_2009, bar_early_2012}.
-We tested both hypotheses in our simulation framework and they both result
-in a reduced, neutral-like fixation probability while maintaining a high
-rate of substitutions, compatible with observations in patient data (see
-supplementary materials).
+Within epitope competition can be implemented in the model through epistasis
+between escape mutations. While each mutation is individually beneficial,
+combining the mutations is deleterious (no extra benefit, but additional costs).
+Again, we find that the potential for hitch-hiking is little affected by within
+epitope competition, but that the fixation probability of non-synonymous
+polymorphisms is reduced. With roughly 6 mutations per epitope, the simulation
+data is compatible with observations; see Fig.~S\withinepi.
 The two scenarios are not exclusive and possibly both important in HIV
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-Despite several known functional roles for RNA secondary structure in the HIV
-genome, synonymous mutations are often used as approximately neutral markers in
-evolutionary studies of viruses. By analyzing the fate of mutations in
-longitudinal data of HIV \env{} evolution, we have shown that the majority of
-synonymous mutations in the conserved regions C2-C5 of the \env~gene are deleterious.
-Comparison with recent biochemical studies of binding propensity of bases in RNA
-genome indicates that these mutations are deleterious, at least in part, because
-they disrupt stems in RNA secondary structures. Computational modelling
-shows that these mutations can be brought to high frequency through
-linkage to adaptive mutations.
+By analyzing the fate of mutations in longitudinal data of HIV \env{} evolution,
+we demonstrate selection against synonymous substitutions in the comparatively
+conserved regions C2-C5 of the \env~gene. Comparison with biochemical studies of binding propensity of bases in RNA
+genome of HIV indicates that these mutations are deleterious, at least in part,
+because they disrupt stems in RNA secondary structures. Computational modelling
+shows that these mutations have deleterious effects on the order of $0.002$
+and that they are brought to high frequency through linkage to adaptive
+mutations.
 
@@
 summary statistics useful to characterize longitudinal sequence data and compare
-to models via computer simulations. Although several recent publications have
-pointed out that synonymous changes in HIV can be
-deleterious~\citep{lemey_synonymous_2007, ngandu_extensive_2008}, our analysis
-is much more direct, as we rely neither on phylogenetic tree reconstruction nor
-on macroevolutionary models. A method that is similar to ours {\it in spiritu}
+to models via computer simulations. A method that is similar to ours {\it in spiritu}
 has been recently used in a longitudinal study of influenza
@@
 mutations.  The latter is used as an approximately neutral control; this method
-can therefore not be used to investigate synonymous changes themselves. A
-functional significance of the insulating RNA structure stems between the
+can therefore not be used to investigate synonymous changes themselves. 
+More generally, evolutionary rates at synonymous sites are often used as a
+baseline to detect purifying or diversifying selection at the protein level
+\cite{Hurst:2002p32608}. It has been pointed out, however, that the rate of
+evolution at synonymous sites varies considerable along the HIV genome
+\citep{mayrose_towards_2007} and that this variation can confound estimates of
+selection on proteins substantially \citep{ngandu_extensive_2008}.
+
+A functional significance of the insulating RNA structure stems between the
 hyper-variable loops has also been proposed
 previously~\citep{watts_architecture_2009, sanjuan_interplay_2011} and conserved
-RNA structures exist in different parts of the HIV
-genome~\citep{ngandu_extensive_2008}. Our analysis is nonetheless much more
-direct and is able to quantify the fitness effect of RNA structure within single
-infections. The main limitations of our approach are the following: (i) it
-requires longitudinal data, most of which focuses on \env{} only; and (ii) we
-only expect to see deleterious synonymous mutations at high frequencies if they
-are linked to repeated adaptive substitions. Otherwise, most synonymous
-polymorphisms remain at low frequencies and can only be observed by deep
-sequencing methods. We indeed plan to exploit next-generation sequencing
-technologies to assess synonymous diversity at the level of rare variants in
-future studies.
+RNA structures exist in different parts of the HIV genome. Our analysis is able
+to quantify the fitness effect of RNA structure within single infections and
+demonstrates how selection at synonymous sites can alter genetic diversity and
+dynamics. The observed hitchhiking highlights the importance of linkage due to
+infrequent recombination for the evolution of HIV
+\citep{neher_recombination_2010,batorsky_estimate_2011,
+josefsson_majority_2011}. The recombination rate has been estimated to be on the
+order of $\rho = 10^{-5}$ per base and day. It takes roughly $t_{sw} =
+\epsilon^{-1} \log \nu_0$ generations for escape mutation with escape rate
+$\epsilon$ to rise from an initially low frequency $\nu_0\sim \mu$ to frequency
+one. This implies that a region of length $l = (\rho t_{sw})^{-1} = \epsilon /
+\rho \log \nu_0$ remains linked to the adaptive mutation. With $\epsilon=0.01$,
+we have $l\approx 100$ bases. Hence we expect strong linkage between the
+variable loops and the intervening sequence, but none far beyond the variable
+regions, consistent with the lack of signal outside of C1-V5. In case of much
+stronger selection -- such as observed during early CTL escape or drug
+resistance evolution -- the linked  region is of course much larger
+\citep{nijhuis_stochastic_1998}.
+
+While classical population genetics
+assumes that the dominant stochastic force is genetic drift, i.e., not heritable
+fluctuations in offspring number, our results show that stochasticity due to
+linked selection is much more important. Such fluctuations have been termed
+\emph{genetic draft} by \citet{gillespie_genetic_2000}. Genetic draft in
+facultatively sexual population like HIV has been characterized in
+\citep{neher_genetic_2011}. Importantly, large population sizes are compatible
+with low diversity and fast coalescence when draft dominates over drift.
 
@@
 more conserved that the synonymous positions, suggesting that the majority of
-the nonsynonymous mutations is much more deleterious (see supplementary
-materials).
+the nonsynonymous mutations is much more deleterious.
 
@@
 
-The observed hitchhiking highlights the importance of linkage due to infrequent
-recombination for the evolution of HIV
-\citep{neher_recombination_2010,batorsky_estimate_2011,
-josefsson_majority_2011}. The recombination rate has been estimated to be on the
-order of $\rho = 10^{-5}$ per base and day. It takes roughly $t_{sw} =
-\epsilon^{-1} \log \nu_0$ generations for escape mutation with escape rate
-$\epsilon$ to rise from an initially low frequency $\nu_0\sim \mu$ to frequency
-one. This implies that a region of length $l = (\rho t_{sw})^{-1} = \epsilon /
-\rho \log \nu_0$ remains linked to the adaptive mutation. With $\epsilon=0.01$,
-we have $l\approx 100$ bases. Hence we expect strong linkage between the
-variable loops and the intervening sequence, but none far beyond the variable
-regions, consistent with the lack of signal outside of C1-V5. In case of much
-stronger selection -- such as observed during early CTL escape or drug
-resistance evolution -- the linked  region is of course much larger
-\citep{nijhuis_stochastic_1998}.
-
 Our results emphasize the inadequacy of independent site models of HIV evolution
@@
 genetic variation is only transiently beneficial, existing estimates of the
-strength of selection \citep{neher_rate_2010,batorsky_estimate_2011} could be
+strength of selection \citep{neher_recombination_2010,batorsky_estimate_2011} could be
 substantial underestimates.
@@
 For the estimate of times to fixation/extinction, polymorphisms were
-binned by frequency and the time to reaching the first boundary (fixation or
-extinction) was stored. The fixation probability was determined as the long-time
+binned by frequency and the time to first reaching either fixation or
+extinction was stored. The fixation probability was determined as the long-time
 limit of the resulting curves. Mutations that reached high frequency but neither
@@
 assignments in indel-rich regions were excluded from the analysis.
+The variable loops and flanking regions were identified manually starting from
+the annotated reference HXB2 sequence from the LANL HIV database~\citep{LANL2012}.
 
-The variable loops and flanking regions were identified manually starting from
-the annotated reference HXB2 sequence from the LANL HIV database~\citep{LANL2012}. A
-similar approach was used to label the C2-V5 region sequenced in
-ref.~\citep{shankarappa_consistent_1999}.
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Computer simulations}
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+Computer simulations were performed using
+FFPopSim~\citep{zanini_ffpopsim:_2012}. Briefly, FFPopSim enables individual
+based simulations where each site in the genome is represented by one bit that
+can be in one of two states. Outcrossing rates, crossover rates, mutations rates
+and arbitrary fitness functions can be specified. We used a generation time of 1
+day, an outcrossing rate of $r=0.01$ per
+day \citep{batorsky_estimate_2011,neher_recombination_2010}, a mutation rate of
+$\mu=10^{-5}$ \citep{mansky_lower_1995,abram_nature_2010} and simulated
+intra-patient evolution for 6000 days. For simplicity, all third position in a
+codon where deemed synonymous and assigned either a selection coefficient $0$
+with probability $1-\alpha$ or a deleterious effect $s_d$ with probability
+$\alpha$.
+First and second positions have strongly deleterious fitness effects
+0.02. At at rate $k_A$, a random place in the genome is
+designated an epitope that can escape by one or several mutations with an
+exponentially distributed escape rate with mean $\epsilon$. Both full-length HIV
+genomes and \env{}-only simulations were performed and yielded comparable results.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Computer simulations}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-Computer simulations were performed using
-FFPopSim~\citep{zanini_ffpopsim:_2012}. Both full-length HIV genomes and
-\env{}-only simulations were performed and yielded comparable results. In each
-simulation, four random parameters were drawn form prior distributions: (i)
-fraction of deleterious synonymous sites (flat between 0.75 and 1.0); (ii)
-deleterious effect at those sites (exponential between $10^{-4}$ and $10^{-2}$);
-(iii) average beneficial effect of an escape mutation (exponential between
-$10^{-2.5}$ and $10^{-1.5}$); (iv) rate of introduction of new epitopes
-(exponential between $10^{-3}$ and $10^{-2}$ per generation). After these four
-parameters were fixed, the simulation started. All the deleterious synonymous
-sites had the same fitness effect (variations were tested and produced the same
-results), no epitopes (i.e. no escape mutations) were initially present, and the
-population was isomorphic.
-After 30 generations of burn-in to create genetic diversity, new epitopes
-were introduced at a constant rate (as specified above). Each time such an
-epitope was actually introduced, its fitness advantage was drawn from an
-exponential distribution with fixed average value (as specified above).
+The simulations are repeated 2400 times with random choices for the following
+parameters:
+The fraction of deleterious sites $\alpha$ was sampled uniformly between 0.75 and 1.0; the average deleterious effect $s_d$ was sampled such that
+its logarithm is uniformly distributed  between $10^{-4}$ and $10^{-2}$; the 
+average escape rate $\epsilon$ of escape mutation such that its logarithm is
+uniform between $10^{-2.5}$ and $10^{-1.5}$; the rate $k_A$ of new antibody
+challenges such that its logarithm is uniform between $10^{-3}$ and $10^{-2}$
+per generation. Populations were initialized with a homogenous founder
+population and were kept at an average size of $N=10^4$ throughout the
+simulation. After 30 generations of burn-in to create genetic diversity, new
+epitopes were introduced at a constant rate $k_A$.
+
 For the models with competition within epitopes, a complex epistatic fitness
-landscape was manually crafted, in which each single mutant provides the whole
-beneficial effect of escape. For the models with time-dependent selection due to
-the immune system catching up, the beneficial effect of an escape mutation was
-reverted at random times after it had appeared, according to its integrated allele
-frequency (see supplementary materials for more details).
+landscape was designed such that each single mutant is sufficient for full
+escape. \comment{need more}
+To model recognition of escape variants by the immune system that is catching
+up, the beneficial effect of an escape mutation was set to zero with a
+probability per generation proportional to the frequency of the escape variant.
 
@@
 synonymous polymorphisms $P_\text{interm}$ were calculated as averages over
-approximately 100 repetitions (with different random seeds).
+100 repetitions (with different random seeds).
 
@@
 interpolation between the bin centers. This measure is sufficiently precise for
-our purposes.
+our purposes. \comment{what happens with empty bins??}
 

### 2013-02-12 scripts/allele_frequency_trajectories.py

@@ -26,7 +26,7 @@ if __name__ == '__main__':
     # Define the patients
     patients = pS.parse_sequences(reference='HXB2')
 
-    for k,p in enumerate(patients[:1]):
+    for k,p in enumerate(patients[:]):
         print p
 
         p.filter_only_sequenced()

### 2013-02-12 scripts/fixation_loss_synonymous_times.py

@@ -232,9 +232,9 @@ if __name__ == '__main__':
                                 color=colors[iji], alpha=0.3)
         ax.add_patch(rec)
 
-    ax.set_xlabel(r'$t_\text{boundary} - t_i$ [days]', fontsize=16)
+    ax.set_xlabel(r'time interval $\Delta t$ [days]', fontsize=16)
     ax.set_ylabel(r'$\nu$', fontsize=18)
-    ax.set_title('Fixation/extinction times')
+    #ax.set_title(r'time interval $\Delta t [days]$')
     ax.legend(loc=1)
     ax.set_xlim(0, 2000)
 

### 2013-02-12 supplementary.tex
Fig.1 has p10; Fig S3 -> S3/S4; more details on epitopes; S2 larger

@@
 \includegraphics[width=\linewidth]{conservation_codons_genome}
-\caption{Synonymous diversity across the HIV genome, as quantified by the
-normalized codon entropy among sequences coding for the
-consensus amino-acid.
-The synonymous diversity peaks at the variable regions in {\it env} and is reduced in regions under purifying selection (RRE
-hairpin, second {\it tat}/{\it rev} exons). Since the variable regions are under
-positive selection at the protein level, synonymous hitchhiking is stronger
-there, and even deleterious synonymous mutations can be brought to high
-frequency. This gives to our signal in the fixation probability, Figs. 2A and
-2B of the main text.
-The normalized codon entropy is calculated as follows (see the script
-for the full algorithm): (i) from a subtype B multiple sequence alignment
-(MSA) from the LANL website (filtered sequences only, version 2011)~\cite{LANL2012},
-we calculate the consensus amino acid at each position in the HIV genome; (ii) we count
-how often each codon coding for the consensus amino acid appears in the MSA;
-(iii) at each amino acid position, we divide by the number of sequences in the
-MSA that had the consensus amino acid at that position, obtaining {\it codon
-frequencies} $\nu_c$; (iv) we calculate the codon entropy from each position as:
-$S := - \sum_{c} \nu_c \log \nu_c$, where $c$ runs over codons that code for
-the consensus amino acid at this site; (v) we divide by the maximal codon
-entropy of that amino acid (e.g. $\log 2$ for twofold degenerate codons).
-All parts of {\it env} that are part of a different gene (signaling peptide,
-second {\it rev} exon) have been excluded from our main analysis, to avoid
-contamination about synonymity.}
+\caption{
+Synonymous diversity across the HIV genome, as quantified by the normalized
+codon entropy among sequences coding for the consensus amino-acid. The
+synonymous diversity peaks at the variable regions in {\it env} and is reduced
+in regions under purifying selection (RRE hairpin, second {\it tat}/{\it rev}
+exons). Since the variable regions are under positive selection at the protein
+level, synonymous hitchhiking is stronger there, and even deleterious synonymous
+mutations can be brought to high frequency. This gives to our signal in the
+fixation probability, Figs. 2A and 2B of the main text. The normalized codon
+entropy is calculated as follows (see the script for the full algorithm): (i)
+from a subtype B multiple sequence alignment (MSA) from the LANL website
+(filtered sequences only, version 2011)~\cite{LANL2012}, we calculate the
+consensus amino acid at each position in the HIV genome; (ii) we count how often
+each codon coding for the consensus amino acid appears in the MSA; (iii) at each
+amino acid position, we divide by the number of sequences in the MSA that had
+the consensus amino acid at that position, obtaining {\it codon frequencies}
+$\nu_c$; (iv) we calculate the codon entropy from each position as: $S := -
+\sum_{c} \nu_c \log \nu_c$, where $c$ runs over codons that code for the
+consensus amino acid at this site; (v) we divide by the maximal codon entropy of
+that amino acid (e.g. $\log 2$ for twofold degenerate codons). All parts of
+{\it env} that are part of a different gene (signaling peptide, second {\it rev}
+exon) have been excluded from our main analysis, to avoid contamination about
+synonymity. Note that all gap-rich columns of the MSA are stripped from this
+figure, hence genes such as {\it env} might appear shorter than they actually
+are.
+}
 \label{fig:syndiv_genome}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Time dependent selection}
+\section{Time-dependent selection}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{center}
-\includegraphics[width=0.8\linewidth]{simulations_graduallyepitopesandtimeselec}
-\caption{Time-dependent selection (the escape mutant is recognized after some
-months from its appearance) and competition between escape mutations in the same
-epitope both reduce fixation of nonsynonymous mutations. A ``na\"ive'' computer
-simulation shows a depression in synonymous fixation probability, as observed in
-patients, but the nonsynonymous fixation probability is much higher than the
-(neutral) diagonal: a nonsynonymous mutation that reaches a frequency around
-0.2 is already very likely to fix. In data from infected patients, however,
-nonsynonymous mutations show a neutral-like curve,
+\includegraphics[width=0.6\linewidth]{simulations_gradually_timeselec_deadagain}
+\caption{
+Time-dependent selection (the escape mutant is recognized after some months from
+its appearance) reduces fixation of nonsynonymous
+mutations. A ``na\"ive'' computer simulation shows a depression in synonymous
+fixation probability, as observed in patients, but the nonsynonymous fixation
+probability is much higher than the (neutral) diagonal: a nonsynonymous mutation
+that reaches a frequency around 0.2 is already very likely to fix. In data from
+infected patients, however, nonsynonymous mutations show a neutral-like curve,
 $P_\text{fix}(\nu) \approx \nu$. In the other curves, we show that the
-nonsynonymous fixation probability can be reduced by either time-dependent
-selection or competition between equivalent escapes, without affecting much the
-synonymous fixation probability itself.
-In the curves denoted with ``tds'' (time-dependent selection), the fitness advantage of
-an escape mutaition can vanish at each time point with a probability density
-\[ P_\text{recognized}(t) = c \cdot \nu(t), \]
-where $c$ is a constant coefficient shown in the legend that encodes the overall
-efficiency of the host immune system, and $\nu(t)$ is the frequency of the
-escape allele at time $t$.
-In the curves denoted with ``mut/epi'' we test the
-competition scenario instead, with 3 or 6 escape mutations within the same
-epitope of length 9 nucleotides. The fitness landscape of each such epitope
-include negative epistatic terms, so that the joint presence of more than one
-escape mutation is not any more beneficial for the virus than a single mutation.
-In this latter scenario, many escape mutations start to sweep on different
-backgrounds within the viral population, but eventually compete and only one of
-them if fixed.}
-\label{fig:tds_wec}
+nonsynonymous fixation probability can be reduced by time-dependent selection
+without affecting much the synonymous fixation probability itself. The fitness
+advantage of an escape mutaition can vanish at each time point with a
+probability density \[ P_\text{recognized}(t) = c \cdot \nu(t), \] where $c$ is
+a constant coefficient shown in the legend that encodes the overall efficiency
+of the host immune system, and $\nu(t)$ is the frequency of the escape allele at
+time $t$. After recognition, the former escape mutation has the same fitness
+cost as before the creation of the epitope, hence it is quickly lost. Note that,
+although the curves shown here show almost no reduction in fixation probability
+up to $\nu \approx 0.15$, this effect is not robust and depends on the exact form of
+$ P_\text{recognized}(t)$. The other parameters for the shown simulations are
+the following: deleterious effect $10^{-3}$, average escape rate $0.016$,
+fraction of deleterious synonymous mutations $0.986$, rate of new epitopes
+$0.0014$ per generation.
+}
+\label{fig:tds}
 \end{center}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Within epitope competition}
+\section{Within-epitope competition}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{center}
-\includegraphics[width=0.8\linewidth]{simulations_graduallyepitopesandtimeselec}
-\caption{Time-dependent selection (the escape mutant is recognized after some
-months from its appearance) and competition between escape mutations in the same
-epitope both reduce fixation of nonsynonymous mutations. A ``na\"ive'' computer
-simulation shows a depression in synonymous fixation probability, as observed in
-patients, but the nonsynonymous fixation probability is much higher than the
-(neutral) diagonal: a nonsynonymous mutation that reaches a frequency around
-0.2 is already very likely to fix. In data from infected patients, however,
-nonsynonymous mutations show a neutral-like curve,
-$P_\text{fix}(\nu) \approx \nu$. In the other curves, we show that the
-nonsynonymous fixation probability can be reduced by either time-dependent
-selection or competition between equivalent escapes, without affecting much the
-synonymous fixation probability itself.
-In the curves denoted with ``tds'' (time-dependent selection), the fitness advantage of
-an escape mutaition can vanish at each time point with a probability density
-\[ P_\text{recognized}(t) = c \cdot \nu(t), \]
-where $c$ is a constant coefficient shown in the legend that encodes the overall
-efficiency of the host immune system, and $\nu(t)$ is the frequency of the
-escape allele at time $t$.
-In the curves denoted with ``mut/epi'' we test the
-competition scenario instead, with 3 or 6 escape mutations within the same
-epitope of length 9 nucleotides. The fitness landscape of each such epitope
-include negative epistatic terms, so that the joint presence of more than one
-escape mutation is not any more beneficial for the virus than a single mutation.
-In this latter scenario, many escape mutations start to sweep on different
-backgrounds within the viral population, but eventually compete and only one of
-them if fixed.}
-\label{fig:tds_wec}
+\includegraphics[width=0.6\linewidth]{simulations_gradually_epitopes}
+\caption{
+Competition between escape mutations in the same epitope also reduces fixation of
+nonsynonymous mutations. Compared to a ``na\"ive'' model,
+simulations with 3 or 6 escape mutations competing within the same epitope show
+reduced fixation probabilities of nonsynonymous changes, whereas the synonymous
+changes behave similarly in all cases. The epitope length is 9 nucleotides. The
+fitness landscape of each such epitope include negative epistatic terms, so that
+the joint presence of more than one escape mutation is not any more beneficial
+for the virus than a single mutation. In detail, each site has two alleles, $\pm
+1$, where $-1$ is the ancestral one and $+1$ the dervied one; the fitness
+coefficient of a $k$-uple of sites within the epitope is $f_k = \pm
+\eta_\epsilon / 2^{L-1}$, where $\eta_\epsilon$ is a random variable for the
+escape rate with average value $\epsilon$ (see methods), $L$ is the number of
+competing escapes in the epitope, and the positive (resp. negative) sign is for
+odd (resp. even) $k$. In this evolutionary scenario, many escape mutations start
+to sweep on different backgrounds within the viral population, but eventually
+compete and only one of them if fixed. The other parameters for the shown simulations are
+the following: deleterious effect $10^{-3}$, average escape rate $0.016$,
+fraction of deleterious synonymous mutations $0.986$, rate of new epitopes
+$0.0014$ per generation.
+}
+\label{fig:wec}
 \end{center}

### 2013-02-12 synmut.tex
Fig.1 has p10; Fig S3 -> S3/S4; more details on epitopes; S2 larger

@@
 
-Even in absense of important RNA structure synonymous codons don't
-evolve completely neutrally but that some codons are favored over others in many
-species \citep{plotkin_synonymous_2011}. 
-Recent studies have shown that genetically engineered HIV strains with altered codon usage can in some cases produce more viral protein, but in general replicate less efficiently~\citep{ngumbela_quantitative_2008,
-li_codon-usage-based_2012,keating_rich_2009}.
-Codon-deoptimization has been suggested as attenuation strategy for polio and 
-influenza~\citep{mueller_live_2010,coleman_virus_2008}. Purifying
-selection beyond the protein sequence is therefore expected
-\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011} and it has been
-shown that rates of evolution at synonymous sites
-vary along the HIV genome \citep{mayrose_towards_2007}.
-Positive selection through the host's immune system, however, is restricted to
-changes in the amino acid sequence.
+Even in absense of important RNA structure synonymous codons don't evolve
+completely neutrally but that some codons are favored over others in many
+species \citep{plotkin_synonymous_2011}.  Recent studies have shown that
+genetically engineered HIV strains with altered codon usage can in some cases
+produce more viral protein, but in general replicate less
+efficiently~\citep{ngumbela_quantitative_2008,
+li_codon-usage-based_2012,keating_rich_2009}.  Codon-deoptimization has been
+suggested as attenuation strategy for polio and
+influenza~\citep{mueller_live_2010,coleman_virus_2008}. Purifying selection
+beyond the protein sequence is therefore expected
+\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011} and it has been shown that
+rates of evolution at synonymous sites vary along the HIV genome
+\citep{mayrose_towards_2007}.  Positive selection through the host's immune
+system, however, is restricted to changes in the amino acid sequence.
 
@@
 the time course the frequencies of all synonymous and nonsynonymous mutations
-observed \env, C2-V5, in patient p1~\citep{shankarappa_consistent_1999},
+observed \env, C2-V5, in patient p10~\citep{shankarappa_consistent_1999},
 respectively. Despite many synonymous mutations reaching high frequency, 
@@
 \begin{center}
-\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_p1.pdf}
+\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_p10.pdf}
 \label{fig:aftsyn}}\\
- \subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p1.pdf}
+ \subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p10.pdf}
 \label{fig:aftnonsyn}}
-\caption{\comment{I would like p2 or p10 better here} Time series of frequencies
+\caption{Time series of frequencies
 of synonymous (A) and non-synonymous (B) derived alleles in \env, C2-V5, from patient
@@
 in chronic infections have been estimated to be on the order of 
-$s_a = 0.01$ per day~\citep{neher_recombination_2010,Asquith:2006p28003}.
+$\epsilon = 0.01$ per day~\citep{neher_recombination_2010,Asquith:2006p28003}.
 
@@
 limit of the resulting curves. Mutations that reached high frequency but neither
-fixed nor got extinct were classified as ``floating'' and not analyzed further.
+fixed nor got extinct were classified as ``floating'', with one exception: if they first
+reached high frequencies within 3 years of the last time point, we assumed
+they had not had sufficient time to settle, so we discarded them.
 
@@
 The simulations are repeated 2400 times with random choices for the following
-parameters:
-The fraction of deleterious sites $\alpha$ was sampled uniformly between 0.75 and 1.0; the average deleterious effect $s_d$ was sampled such that
+parameters: The fraction of deleterious sites $\alpha$ was sampled uniformly
+between 0.75 and 1.0; the average deleterious effect $s_d$ was sampled such that
 its logarithm is uniformly distributed  between $10^{-4}$ and $10^{-2}$; the
@@
 landscape was designed such that each single mutant is sufficient for full
-escape. \comment{need more}
-To model recognition of escape variants by the immune system that is catching
-up, the beneficial effect of an escape mutation was set to zero with a
-probability per generation proportional to the frequency of the escape variant.
+escape. In particular, each mutation had a linear effect equal to the escape,
+but a negative epistatic effect of the same magnitude between each pair of sites
+was included. Higher order terms compensated each other to make sure that not
+only double, but all k-mutants with $k \geq 1$ had the same fitness. To model
+recognition of escape variants by the immune system that is catching up, the
+beneficial effect of an escape mutation was set to zero with a probability per
+generation proportional to the frequency of the escape variant.
 
@@
 interpolation between the bin centers. This measure is sufficiently precise for
-our purposes. \comment{what happens with empty bins??}
+our purposes. In 10 runs out of 2400, the highest frequency bin was empty so the
+fixation probability could not be calculated; those runs were excluded from 
+\FIG{simsfig}.
 

### 2013-02-12 scripts/allele_frequency_trajectories.py

@@ -26,7 +26,9 @@ if __name__ == '__main__':
     # Define the patients
     patients = pS.parse_sequences(reference='HXB2')
 
-    for k,p in enumerate(patients[:]):
+    for k,p in enumerate(patients):
+        if str(p) != 'p10':
+            continue
         print p
 
         p.filter_only_sequenced()
@@ -50,8 +52,8 @@ if __name__ == '__main__':
         is_nonsyn = is_nonsyn_table(consensus)
 
         # Plot trajectories (no gaps, no conserved)
-        figsyn = ppl.figure(figsize=[9.35, 6.325])
-        fignonsyn = ppl.figure(figsize=[9.35, 6.325])
+        figsyn = ppl.figure(figsize=[6.5, 4.3])
+        fignonsyn = ppl.figure(figsize=[6.5, 4.3])
         axsyn = figsyn.add_subplot(1,1,1)
         axnonsyn = fignonsyn.add_subplot(1,1,1)
         x = p.mo_to_SC
@@ -79,8 +81,8 @@ if __name__ == '__main__':
             ax.set_ylabel(r'$\nu$', fontsize=18)
             ax.set_ylim(-0.05,1.05)
 
-        axsyn.set_title('Derived synonymous allele frequency trajectories, '+str(p), fontsize=18)
-        axnonsyn.set_title('Derived nonsynonymous allele frequency trajectories, '+str(p), fontsize=18)
+        #axsyn.set_title('Derived synonymous allele frequency trajectories, '+str(p), fontsize=18)
+        #axnonsyn.set_title('Derived nonsynonymous allele frequency trajectories, '+str(p), fontsize=18)
 
     ppl.ion()
     ppl.show()

### 2013-02-12 scripts/codon_entropy_synonymous_subtypeB.py

@@ -45,7 +45,8 @@ if __name__ == '__main__':
     data = {}
 
     # Make figure for all four genes
-    fig, axs = plt.subplots(1, 4)
+    fig, axs = plt.subplots(2, 2)
+    axs = axs.flatten()
 
     for k, gene in enumerate(genes):
 
@@ -113,7 +114,7 @@ if __name__ == '__main__':
         plt.plot(np.arange(len(entropy)), entropyfrac, c='b', lw=1)
         plt.xlabel('Position in '+gene)
         plt.ylim(0, 1)
-        if k == 0:
+        if k in [0, 2]:
             plt.ylabel('Codon entropy / maximal entropy')
         else:
             ax.set_yticklabels([''])
@@ -125,8 +126,9 @@ if __name__ == '__main__':
         plt.plot(xsmooth, ysmooth, c='r', lw=2, ls='-')
         plt.title(gene, fontsize=18)
     
-        # Mark V loops
+        # Mark V loops and rev 2nd exon
         if gene == 'env':
+            # V loops
             refstr = str(ali.reference_seq.seq)
             tmp = (-is_good).nonzero()[0]
             Vind = {}
@@ -141,6 +143,15 @@ if __name__ == '__main__':
                 h = Rectangle((i0, 0), i1 - i0, 1, color='yellow', alpha=0.5)
                 plt.gca().add_patch(h)
                 plt.text(i0 + 0.2 * (i1 - i0), 0.95, key, fontsize=12)
+
+            # Rev exon
+            s = ''.join(consensus[is_good])
+            i0 = s.find('ACCCGCCTCCCA') // 3
+            i1 = s.find('TGCTGTTAGCTTG') // 3
+            h = Rectangle((i0, 0), i1 - i0, 1, color='magenta', alpha=0.5)
+            plt.gca().add_patch(h)
+            t = plt.text(i1 + 0.15 * (i1 - i0), 0.85, 'rev\n2nd\nexon', fontsize=12)
+
         
 
     plt.ion()

### 2013-02-12 scripts/simulations_evolution.py

@@ -275,7 +275,7 @@ if __name__ == '__main__':
                 if len(neutralize_ind):
                     fitness = pop.get_trait_additive()
                     for ie in neutralize_ind:
-                        fitness[pop.epitopes[ie][0]] = 0
+                        fitness[pop.epitopes[ie][0]] = -0.01
                     pop.set_trait_additive(fitness)
     
             sys.stdout.write(str(100 * i / (n_time_points - 1))+'% completed\n')

### 2013-02-12 scripts/simulations_plot_gradually_epitopesandtimeselec.py

@@ -35,9 +35,9 @@ def filterfolders(folders):
             ind[1] = i
         elif 'nescepi_6' in f:
             ind[2] = i
-        elif 'timeselec_0.005' in f:
+        elif 'timeselec_0.002_deadagain' in f:
             ind[3] = i
-        elif 'timeselec_0.01' in f:
+        elif 'timeselec_0.005_deadagain' in f:
             ind[4] = i
         elif ('nescepi' not in f) and ('timeselec' not in f):
             ind[0] = i
@@ -54,43 +54,52 @@ if __name__ == '__main__':
 
     # Sort them smartly
     folders = filterfolders(folders)
-    labels = ['naive', '3 mut/epi', '6 mut/epi', 'tds 0.005', 'tds 0.01']
+    labels = ['naive', '3 mut/epi', '6 mut/epi',
+              'c = 0.002 / gen', 'c = 0.005 / gen']
     colors = ['b', 'g', 'r', 'cyan', 'magenta']
 
-    # Prepare figure
-    fig = plt.figure()
+    # Make two figures, one for tds, one for competing escapes
+    for k, ind in enumerate([[0, 1, 2], [0, 3, 4]]):
 
-    # Get the data
-    for i, folder in enumerate(folders):
+        # Prepare figure
+        fig = plt.figure()
+    
+        # Get the data
+        for i in ind:
+    
+            folder = folders[i]
+            fn = results_dir+folder
+            nu0, nu1, Psyn, Pnonsyn, csyn, cnonsyn = np.genfromtxt(fn+'/Pfix.dat',
+                                                                   unpack=True,
+                                                                   usemask=True,
+                                                                   missing_values='--')
+            # Plot fixation probabilities
+            x = [0] + list(0.5 * (nu0 + nu1)) + [1]
+            ysyn = [0] + list(Psyn) + [1]
+            ynonsyn = [0] + list(Pnonsyn) + [1]
+            plt.plot(x, ysyn,
+                     color=colors[i],
+                     ls='--',
+                     lw=2)
+            plt.plot(x, ynonsyn,
+                     label=labels[i],
+                     color=colors[i],
+                     ls='-',
+                     lw=2)
+    
+        # Plot diagonal
+        plt.plot(np.linspace(0, 1, 100), np.linspace(0, 1, 100), color='k',
+                 ls='--', lw=1.5)
[…]

### 2013-02-13 supplementary.tex
text small changes

@@
 Synonymous diversity across the HIV genome, as quantified by the normalized
-codon entropy among sequences coding for the consensus amino-acid. The
+codon entropy among sequences coding for the consensus amino acid. The
 synonymous diversity peaks at the variable regions in {\it env} and is reduced

### 2013-02-13 synmut.tex
text small changes

@@
 \newcommand{\shankaregion}{C2-V5}
-\newcommand{\timedependence}{4}
+\newcommand{\PCApat}{1}
+\newcommand{\timedependence}{3}
 \newcommand{\withinepi}{4}
@@
 Intrapatient HIV evolution is dominated by selection on the protein level in the
-arms race with the adaptive immune system. Synonymous mutations do not affect 
-the interaction with the adaptive immune system and are often
-assumed to be neutral.
+arms race with the adaptive immune system. When killer T cells or antibodies
+target a new epitope, the virus often escapes via nonsynonymous mutations that
+impair recognition. Synonymous mutations do not affect this interplay and are
+often assumed to be neutral.
 We analyze longitudinal intrapatient data from the \shankaregion{} part of the
@@
 Computational modeling indicates that these synonymous mutations have a
-(Malthusian) selection coefficient of the order of $-0.002$, and that they are brought up to high
-frequency by hitchhiking on neighboring beneficial nonsynonymous alleles.
-The patterns of fixation of nonsynonymous mutations suggest that antibody
-escape mutations in \shankaregion{} are only transiently beneficial, either
-because the immune system is catching up or because of competition between equivalent
-escapes.
+(Malthusian) selection coefficient of the order of $-0.002$, and that they are
+brought up to high frequency by hitchhiking on neighboring beneficial
+nonsynonymous alleles. The patterns of fixation of nonsynonymous mutations
+suggest that escape mutations in \shankaregion{} are only transiently
+beneficial, either because the immune system is catching up or because of
+competition between equivalent escapes.
 
@@
 via killer T cells (CTLs) and neutralizing antibodies
-(ABs)~\citep{rambaut_causes_2004} and facilitated by the high
-mutation rate of HIV~\citep{mansky_lower_1995,abram_nature_2010}. When the host
-develops a CTL or AB response against a particular HIV epitope, mutations in the viral genome that
+(ABs)~\citep{rambaut_causes_2004} and facilitated by the high mutation rate of
+HIV~\citep{mansky_lower_1995,abram_nature_2010}. When the host develops a CTL or
+AB response against a particular HIV epitope, mutations in the viral genome that
 reduce or prevent recognition of the epitope frequently emerge. Escape mutations
@@
 rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
-infection, the most rapidly evolving part of the HIV genome are the so called
-variable loops of the envelope protein gp120, which change to avoid recognition
-by neutralizing ABs.  Mutations in \env, the gene encoding gp120, spread
-through the population within a few months (see \figurename~\ref{fig:aftnonsyn}).
-Consistent with this time scale, it is found that serum from a
-particular time typically neutralizes virus extracted more than 3-6 month
-earlier \citep{richman_rapid_2003}.
+infection, the most rapidly evolving part of the HIV genome are the variable
+loops V1-V5 in the envelope protein gp120, which change to avoid recognition by
+neutralizing ABs. Mutations in \env, the gene encoding gp120, spread through the
+population within a few months (see \figurename~\ref{fig:aftnonsyn}).
+Consistent with this time scale, it is found that serum from a particular time
+typically neutralizes virus extracted more than 3-6 month earlier
+\citep{richman_rapid_2003}.
 
-These escape mutations are selected for their effect on the amino acid
-sequence of the viral proteins. Conversely, synonymous mutations are commonly
-used as approximately neutral markers in studies of viral evolution. Neutral
-markers are very useful since their dynamics can be compared to
-that of putatively functional sites to detect purifying or directional selection
-\citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}.
-In addition to maintaining protein function and avoiding the adaptive immune
-recognition, however, the HIV genome has to ensure efficient processing and translation,
-nuclear export, and packaging into the viral capsid: all these processes operate at the RNA
-level and are sensitive to synonymous changes since these processes often
-depend on RNA folding.
-For example, the HIV \rev{} response element (RRE) in \env{} enhances nuclear export of
-full length or partially spliced viral transcripts~\citep{fernandes_hiv-1_2012}.
-Another well studied case is the interaction between viral reverse transcriptase, viral ssRNA, and the host
-tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription
-(RT) and is bound by a pseudoknotted RNA structure in the viral 5'
-untranslated region~\citep{barat_interaction_1991, paillart_vitro_2002}.
+These escape mutations are selected for their effect on the amino acid sequence
+of the viral proteins. Conversely, synonymous mutations are commonly used as
+approximately neutral markers in studies of viral evolution. Neutral markers are
+very useful since their dynamics can be compared to that of putatively
+functional sites to detect purifying or directional selection
+\citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In addition to
+maintaining protein function and avoiding the adaptive immune recognition,
+however, the HIV genome has to ensure efficient processing and translation,
+nuclear export, and packaging into the viral capsid: all these processes operate
+at the RNA level and are sensitive to synonymous changes since these processes
+often depend on RNA folding. For example, the HIV \rev{} response element (RRE)
+in \env{} enhances nuclear export of full length or partially spliced viral
+transcripts via a complex stem-loop RNA structure~\citep{fernandes_hiv-1_2012}.
+Another well studied case is the interaction between viral reverse
+transcriptase, viral ssRNA, and the host tRNA$^\text{Lys3}$: the latter is
+required for priming reverse transcription (RT) and is bound by a pseudoknotted
+RNA structure in the viral 5' untranslated region~\citep{barat_interaction_1991,
+paillart_vitro_2002}.
 
-Even in absense of important RNA structure synonymous codons don't evolve
-completely neutrally but that some codons are favored over others in many
-species \citep{plotkin_synonymous_2011}.  Recent studies have shown that
-genetically engineered HIV strains with altered codon usage can in some cases
-produce more viral protein, but in general replicate less
-efficiently~\citep{ngumbela_quantitative_2008,
-li_codon-usage-based_2012,keating_rich_2009}.  Codon-deoptimization has been
-suggested as attenuation strategy for polio and
-influenza~\citep{mueller_live_2010,coleman_virus_2008}. Purifying selection
-beyond the protein sequence is therefore expected
+Even in absence of important RNA structures, synonymous codons do not evolve
+completely neutrally. Some codons are favored over others in many species
+\citep{plotkin_synonymous_2011}. Recent studies have shown that genetically
+engineered HIV strains with altered codon usage can in some cases produce more
+viral protein, but in general replicate less efficiently
+\citep{ngumbela_quantitative_2008, li_codon-usage-based_2012,
+keating_rich_2009}. Codon deoptimization has been suggested as attenuation
+strategy for polio and influenza~\citep{mueller_live_2010,coleman_virus_2008}.
+Purifying selection beyond the protein sequence is therefore expected
 \citep{forsdyke_reciprocal_1995,snoeck_mapping_2011} and it has been shown that
 rates of evolution at synonymous sites vary along the HIV genome
-\citep{mayrose_towards_2007}.  Positive selection through the host's immune
-system, however, is restricted to changes in the amino acid sequence.
+\citep{mayrose_towards_2007}. Positive selection through the host adaptive
+immune system, however, is restricted to changes in the amino acid sequence.
 
 In this paper, we characterize the dynamics of synonymous mutations in \env{}
-and show that a substantial fraction of these mutations is deleterious.
-We argue that synonymous mutations reach high frequencies via genetic hitchhiking
+and show that a substantial fraction of these mutations is deleterious.  We
+argue that synonymous mutations reach high frequencies via genetic hitchhiking
 due to the small recombination rate of HIV~\citep{neher_recombination_2010,
-batorsky_estimate_2011}.
-We then
-compare our observations to computational models of HIV evolution and derive
-estimates for the effect synonymous mutations have on fitness.
-Extending the analysis of fixation probabilities to the
+batorsky_estimate_2011}. We then compare our observations to computational
+models of HIV evolution and derive estimates for the effect synonymous mutations
+have on fitness.  Extending the analysis of fixation probabilities to the
 nonsynonymous mutations, we show that time dependent selection or strong
@@
 segregating at frequency $\nu$ has a probability $P_\text{fix}(\nu) = \nu$ to
-spread through the population and fix; in the rest of the cases, i.e.~with
-probability $1-\nu$, it goes extinct. As illustrated in the inset of \FIG{aftsyn},
-this is a simple consequence of the fact that
-(i) exactly one individual in the current population will be
-the common ancestor of the entire future population at a particular locus and
-(ii) this ancestor has a probability $\nu$ of carrying the mutation (assuming
-the neutral mutation is not preferentially associated with genomes of high or
-low fitness).
-Deleterious or beneficial mutations fix less or
-more often than neutral ones, respectively. \FIG{aft} shows 
-the time course the frequencies of all synonymous and nonsynonymous mutations
-observed \env, C2-V5, in patient p10~\citep{shankarappa_consistent_1999},
-respectively. Despite many synonymous mutations reaching high frequency, 
-few fix (panel~\ref{fig:aftsyn}); in constrast, many nonsynonymous mutations fix
-(panel~\ref{fig:aftnonsyn}).
+spread through the population and fix; in the rest of the cases, i.e. with
+probability $1-\nu$, it goes extinct. As illustrated in the inset of
+\FIG{aftsyn}, this is a simple consequence of the fact that (i) exactly one
+individual in the current population will be the common ancestor of the entire
+future population at a particular locus and (ii) this ancestor has a probability
+$\nu$ of carrying the mutation (assuming the neutral mutation is not
+preferentially associated with genomes of high or low fitness). Deleterious or
+beneficial mutations fix less or more often than neutral ones, respectively.
+\FIG{aft} shows the time course the frequencies of all synonymous and
+nonsynonymous mutations observed \env, \shankaregion, in patient
+p10~\citep{shankarappa_consistent_1999}, respectively. Despite many synonymous
+mutations reaching high frequency, few fix (panel~\ref{fig:aftsyn}); in
+constrast, many nonsynonymous mutations fix (panel~\ref{fig:aftnonsyn}).
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Synonymous polymorphisms in \env, C2-V5, are mostly deleterious}
+\subsection{Synonymous polymorphisms in \env, \shankaregion, are mostly deleterious}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \citet{shankarappa_consistent_1999,liu_selection_2006} and 3 patients from
-\citet{bunnik_autologous_2008} (patients whose viral population was
-structured were excluded from the analysis; see methods).
-The former data set from is restricted to the
-C2-V5 region of \env, while the data from \citet{bunnik_autologous_2008} covers
-the majority of \env. Considering all mutations in a
-frequency interval around $\nu_0$ at some time $t$, we calculate the fraction
-that is found at frequency 1, at frequency 0, or at intermediate frequency at
-later times $t+\Delta t$. Plotting these fixed, lost, and polymorphic fraction
-against the time interval $\Delta t$, we see that most synonymous mutations
-segregate for roughly one year and are lost much more frequently than expected (panel
-\ref{fig:fixp1}). The long-time probability of fixation versus extinction of
-synonymous mutations is shown as a function of the initial frequency $\nu_0$ in
-panel~\ref{fig:fixp2} (red line). Restricted to the region C2-V5, we find that
-$\pfix$ of synonymous mutations is far below the neutral expectation.
-Outside of C2-V5, using data from \citet{bunnik_autologous_2008} only, no such
-reduction in $\pfix$ is found. Restricted to the C2-V5 area, the data from
+\citet{bunnik_autologous_2008} (patients whose viral population was structured
+were excluded from the analysis; see methods and \figurename~S\PCApat).  The
+former data set from is restricted to the \shankaregion{} region of \env, while
+the data from \citet{bunnik_autologous_2008} covers the majority of \env.
+Considering all mutations in a frequency interval around $\nu_0$ at some time
+$t$, we calculate the fraction that is found at frequency 1, at frequency 0, or
+at intermediate frequency at later times $t+\Delta t$. Plotting these fixed,
+lost, and polymorphic fraction against the time interval $\Delta t$, we see that
+most synonymous mutations segregate for roughly one year and are lost much more
+frequently than expected (panel \ref{fig:fixp1}). The long-time probability of
+fixation versus extinction of synonymous mutations is shown as a function of the
+initial frequency $\nu_0$ in panel~\ref{fig:fixp2} (red line). Restricted to the
+region \shankaregion, we find that $\pfix$ of synonymous mutations is far below
+the neutral expectation.  Outside of \shankaregion, using data from
+\citet{bunnik_autologous_2008} only, no such reduction in $\pfix$ is found.
+Restricted to the \shankaregion area, the data from
 \citet{bunnik_autologous_2008} is fully compatible with data from
 \citet{shankarappa_consistent_1999}. The nonsynonymous mutations seem to follow
-more or less the neutral expectation (blue line) -- a point to which we will come back below.
+more or less the neutral expectation (blue line) -- a point to which we will
+come back below.
 
@@
 \caption{Time series of frequencies
-of synonymous (A) and non-synonymous (B) derived alleles in \env, C2-V5, from patient
+of synonymous (A) and nonsynonymous (B) derived alleles in \env, \shankaregion, from patient
 1~\cite{shankarappa_consistent_1999}.
@@
 mutations do even though they are frequently observed at intermediate
-frequencies. Colors indicate the position of the site along the C2-V5 region
+frequencies. Colors indicate the position of the site along the \shankaregion region
 (blue to red). Inset: the fixation probability $P_\text{fix}(\nu)$ of a neutral mutation
@@
 frequency. If, on the other hand, the polymorphism is deleterious it must have
-gotten to high frequency by chance (genetic drift or hitchhiking), and
-we expect that selection will drive it out of the population again. Hence our
+reached a high frequency by chance (genetic drift or hitchhiking), and
+we expect that selection drives it out of the population again. Hence our
 observations suggest that many of the synonymous polymorphisms at intermediate
-frequencies in the part of \env~that includes the hypervariable regions are
+frequencies in the part of \env{} that includes the hypervariable regions are
 deleterious, while outside this regions polymorphisms are mostly roughly
@@
 Panel B) Fixation probability of derived synonymous
-alleles as a function of $\nu_0$ in C2-V5 is reduced relative to the neutral
+alleles as a function of $\nu_0$ in \shankaregion is reduced relative to the neutral
 expectation indicated by the diagonal line. This suppression is not
@@
 standard deviation after 100 patient bootstraps of the data. Data from
-Refs.~\cite{shankarappa_consistent_1999,liu_selection_2006, bunnik_autologous_2008}.}
+refs.~\cite{shankarappa_consistent_1999,liu_selection_2006, bunnik_autologous_2008}.}
 \label{fig:fixp}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Synonymous mutations in C2-V5 tend to disrupt conserved RNA stems}
+\subsection{Synonymous mutations in \shankaregion tend to disrupt conserved RNA stems}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-One possible explanation for lack of fixation of synonymous
-mutations in C2-V5 are secondary structures in the viral RNA, the disruption of
-which is deleterious to the virus
-~\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011,sanjuan_interplay_2011}.
+One possible explanation for lack of fixation of synonymous mutations in
+\shankaregion are secondary structures in the viral RNA, the disruption of which
+is deleterious to the virus \citep{forsdyke_reciprocal_1995,
+snoeck_mapping_2011, sanjuan_interplay_2011}.
 
-The propensity of nucleotides of the HIV genome to form base pairs has
-been measured using the SHAPE assay (a biochemical reaction preferentially
-altering unpaired bases)~\citep{watts_architecture_2009}. The SHAPE assay has
-shown that the variable regions V1 to V5 tend to be unpaired, while the
-conserved regions between those variable regions form stems. We partition all
-synonymous alleles observed at intermediate frequencies above 10-15\% depending
-on their final destiny (fixation or extinction). Subsequently, we align our
-sequences to the reference NL4-3 strain used in
-ref.~\citep{watts_architecture_2009} and assign them SHAPE reactivities. As
-shown in \FIG{SHAPEA}, the reactivities of fixed
-alleles (red histogram) are systematically larger than of alleles that are lost
-(blue ) (Kolmogorov-Smirnov test, $P\approx 0.002$).
-In other words, alleles that are likely to break RNA helices are also more
-likely to revert and finally be lost from the population. The
-average over all mutations that are not observed (green) lies between the
-those that fix and those that get lost.
-Note that, that this analysis will be sensitive only at position where the base
-pairing pattern of NL4-3 agrees with that of each patient's initial
-consensus sequence.
+The propensity of nucleotides in the HIV genome to form base pairs has been
+measured using the SHAPE assay, a biochemical reaction preferentially altering
+unpaired bases (the HIV genome is a single stranded RNA)
+\citep{watts_architecture_2009}. The SHAPE assay has shown that the variable
+regions V1-V5 tend to be unpaired, while the conserved regions between those
+variable regions form stems. We partition all synonymous alleles observed at
+intermediate frequencies above 15\% depending on their final destiny (fixation
+or extinction). Subsequently, we align our sequences to the reference NL4-3
+strain used in ref.~\citep{watts_architecture_2009} and assign them SHAPE
+reactivities. As shown in \FIG{SHAPEA}, the reactivities of fixed alleles (red
+histogram) are systematically larger than of alleles that are lost (blue)
+(Kolmogorov-Smirnov test on the cumulative distribution, $P\approx 0.002$). In
+other words, alleles that are likely to break RNA helices are also more likely
+to revert and finally be lost from the population. The average over all
+mutations that are not observed (green) lies between the those that fix and
+those that get lost. Note that this analysis will be sensitive only at position
+where the base pairing pattern of NL4-3 agrees with that of each patient's
+initial consensus sequence (it is thus statistically conservative).
 
-To test the hypothesis that mutations in C2-V5 are lost because they break stems in the
-conserved stretches between the variable loops, we consider mutations in
-variable loops and conserved parts separately. The biggest depression in
-fixation probability is observed in the conserved stems, while the variable
-loops show little deviation from the neutral signature, see \FIG{SHAPEB}. This
-is consistent with important stem structures in conservered regions between loops.
+To test the hypothesis that mutations in \shankaregion{} are lost because they
+break stems in the conserved stretches between the variable loops, we consider
+mutations in variable loops and conserved parts separately. The biggest
+depression in fixation probability is observed in the conserved stems, while the
+variable loops show little deviation from the neutral signature, see
+\FIG{SHAPEB}. This is consistent with important stem structures in conservered
+regions between loops.
 
@@
 however, any evidence for a contribution of average CUB to the ultimate fate of
-synonymous alleles consistent with that HIV does not seem to adapt its codon
-usage to its human host cells at the macroevolutionary level
-\citep{kuyl_biased_2012}.
+synonymous alleles; consistently, HIV does not seem to adapt its codon usage to
+its human host cells at the macroevolutionary level \citep{kuyl_biased_2012}.
 
@@
 Fixed mutations tend to have higher higher SHAPE reactivities, corresponding to
-less base pairing ($p=0.002$, Kolmogorov-Smirnov two sample test on the
-corresponding cumulative distribution).
+less base pairing.
 Mutations that are never observered show an intermediate distribution of SHAPE values.
-Panel B) shows the fixation probability of synonymous mutations in C2-V5
-separately for variable loops and the connecting conserved regions that
-harbor conserved RNA stems. As expected, the fixation probability is lower
+Panel B) shows the fixation probability of synonymous mutations in
+\shankaregion{} separately for variable loops and the connecting conserved regions that
+harbor RNA stems. As expected, the fixation probability is lower
 inside the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 While the observation that some fraction of synonymous mutations is deleterious
 is not unexpected, it seems odd that we observe them at high population
-frequency and that the fixation probability is reduced only in parts
-of the genome.
-This regions, however, undergoes frequent
-adaptive changes to evade recognition by neutralizing 
-antibodies~\cite{williamson_adaptation_2003,richman_rapid_2003}. Due to the 
-limited amount of recombination in HIV~\cite{neher_recombination_2010,batorsky_estimate_2011},
-deleterious mutations that are linked to adaptive variants can reach high
-frequency. This process is known as hitchhiking \citep{smith_hitch-hiking_1974}
-or genetic draft \citep{gillespie_genetic_2000,neher_genetic_2011}.
-Hitchhiking is  apparent in \FIG{aft}, that shows that many mutations
-change rapidly in frequency as a flock.
+frequency and that the fixation probability is reduced only in parts of the
+genome. This regions, however, undergoes frequent adaptive changes to evade
+recognition by neutralizing antibodies \cite{williamson_adaptation_2003,
+richman_rapid_2003}. Due to the limited amount of recombination in HIV
+\cite{neher_recombination_2010, batorsky_estimate_2011}, deleterious mutations
+that are linked to adaptive variants can reach high frequency. This process is
+known as hitchhiking \citep{smith_hitch-hiking_1974} or genetic draft
+\citep{gillespie_genetic_2000,neher_genetic_2011}. Hitchhiking is  apparent in
+\FIG{aft}, that shows that many mutations change rapidly in frequency as a
+flock.
 
 The approximate magnitude of the deleterious effects can be estimated from
-\FIG{fixp1}, that shows the distribution of times for synonymous
+\FIG{fixp1}, which shows the distribution of times for synonymous
 alleles to reach the fix or get lost starting from intermediate frequencies. The
@@
 Panel B) To assess the parameter space that affects synonymous fixation and
-diversity, we run many simulations with random parameters for deleterious effect
+diversity, we run 2400 simulations with random parameters for deleterious effect
 size, fraction of deleterious synonymous sites, average size of escape
@@
 fixation probability and diversity observed in patients. Parameters are chosen
-from prior distributions uniform in log-space indicated by the red box (see
-methods).}
+from uniform prior distributions as indicated by the red box (see methods).}
 \label{fig:simheat}
@@
 evolving viral populations assuming a mix of positive and purifying selection
-and rare recombination.
-For this purpose, we use the simulation package FFPopSim, which includes a
-module dedicated to intrapatient HIV evolution~\citep{zanini_ffpopsim:_2012}. 
-For each simulation run, we specify the deleterious effect of synonymous
-mutations, the fraction of synonymous mutations that are deleterious, the
-escape rate of adaptive non-synonymous mutations and the frequency of new
-escapes. Note that the escape rate is sum of two factors: (i) the beneficial
-effect due to the ability to evade the immune system minus (ii) the fitness
-cost of the mutation in terms of structure, stability, etc. Net escape rates
-in chronic infections have been estimated to be on the order of 
-$\epsilon = 0.01$ per day~\citep{neher_recombination_2010,Asquith:2006p28003}.
+and rare recombination.  For this purpose, we use the simulation package
+FFPopSim, which includes a module dedicated to intrapatient HIV
+evolution~\citep{zanini_ffpopsim:_2012}. For each simulation run, we specify the
+deleterious effect of synonymous mutations, the fraction of synonymous mutations
+that are deleterious, the escape rate of adaptive nonsynonymous mutations and
+the frequency of new escapes. Note that the escape rate is the sum of two
+factors: (i) the beneficial effect due to the ability to evade the immune system
+minus (ii) the fitness cost of the mutation in terms of structure, stability,
+etc. Net escape rates in chronic infections have been estimated to be on the
+order of $\epsilon = 0.01$ per day \citep{neher_recombination_2010,
+Asquith:2006p28003}.
 
@@
 synonymous diversity for different deleterious effects of synonymous mutations.
-We quantify synonymous diversity,$P_\text{interm}$, as the fraction of sites
+We quantify synonymous diversity via $P_\text{interm}$, the fraction of sites
 with an allele at frequency $0.25 < \nu < 0.75$. The synonymous diversity
-observed in patient data is indicated in the figure.
-To quantify the depression of the fixation probability, we calculate
-the area between the measured fixation probability and the diagonal, which is
-the neutral expectation (\FIG{simfixpvar}, lower inset). If no fixation happens,
-the area will be $-0.5$; if every mutation fixes, we will measure 
-$A= 0.5$. In HIV, we find $A_\text{syn} \approx -0.2$  for
-synonymous changes and $A_\text{nonsyn} \approx 0$ for nonsynonymous changes.
-
-In the three simulations shown in \FIG{simfixpvar}, the fixation probability of
-synonymous alleles decreases from the neutral expectation ($A_\text{syn} \sim 0$) to zero
-($A_\text{syn} \sim -0.5$) as their fitness effect
-worsens; the synonymous diversity plummets as well, as deleterious mutations
-are selected against.
+observed in patient data is indicated in the figure.  To quantify the depression
+of the fixation probability, we calculate the area between the measured fixation
+probability and the diagonal, which is the neutral expectation
+(\FIG{simfixpvar}, lower inset). If no fixation happens, the area will be
+$-0.5$; if every mutation fixes, the area will be $+0.5$. In HIV infected
+patients, we find $A_\text{syn} \approx -0.2$ for synonymous changes and
+$A_\text{nonsyn} \approx 0$ for nonsynonymous changes. In the three simulations
+shown in \FIG{simfixpvar}, the fixation probability of synonymous alleles
+decreases from the neutral expectation ($A_\text{syn} \sim 0$) to zero
+($A_\text{syn} \sim -0.5$) as their fitness effect worsens; the synonymous
+diversity plummets as well, as deleterious mutations are selected against.
 
@@
 repeatedly simulated the evolution with random choices for the parameters in
-certain bounds, see \FIG{simsfig}.
-Among all simulations, we select the ones that show $A_\text{syn}$ and
-$P_\text{interm}$ as observed in the data, i.e., a large depression in fixation
-probability of synonymous mutations but, simultaneously, a moderately high
-synonymous diversity. Specifically, \FIG{simsfig} shows parameter combinations
-for which we found $A_\text{syn} < -0.15$ and $0.0025 < P_\text{interm} <
-0.010$. These conditions indicate that a high fraction ($\gtrsim 0.8$) of
-sites has to be deleterious with effect size $|s_d| \sim 0.002$. 
-This result fits well the expectation based on the fixation/extinction times above (see \FIG{fixp1}).
-The results are plausible:
+certain bounds, see \FIG{simsfig}. Among all simulations, we select the ones
+that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
+large depression in fixation probability of synonymous mutations but,
+simultaneously, a moderately high synonymous diversity. Specifically,
+\FIG{simsfig} shows parameter combinations for which we found $A_\text{syn} <
+-0.15$ and $0.0025 < P_\text{interm} < 0.010$. These conditions indicate that a
+high fraction ($\gtrsim 0.8$) of sites has to be deleterious with effect size
+$|s_d| \sim 0.002$.  This result fits well the expectation based on the
+fixation/extinction times above (see \FIG{fixp1}). The results are plausible:
 (i) a substantial depression in $\pfix$ requires pervasive deleterious
@@
 deleterious effects in \FIG{simheat} correspond to larger escapes rates. (iii)
-mutations with a deleterious effect smaller than approximately $0.001$ 
-behave neutrally consistent with the typical coalescent times observed in HIV.
+mutations with a deleterious effect smaller than approximately $0.001$ behave
+neutrally consistent with the typical coalescent times observed in HIV.
 
-The above simulation show that hitch-hiking is a plausible explanation for the
-obervation of deleterious mutations that rarely fix. However, in a simple model
-where non-synonymous escape mutations are unconditionally beneficial, they
-almost always fix once they reach high frequencies -- $A_{\mathrm{nonsyn}}$ is
-well above zero. This is incompatible with the blue line in \FIG{fixp}: in an HIV infection, nonsynonymous mutations at high frequency often
-disappear again, even though many are at least transiently beneficial.
-Inspecting the trajectories of nonsynonymous mutations suggests the rapid rise
-and fall of many alleles. We test two possible mechanisms that are biologically
-plausible and could explain the transient rise of nonsynonymous mutations:
-time-dependent selection and within-epitope competition.
+The above simulation show that hitchhiking can explain the obervation of
+deleterious mutations that rarely fix. However, in a simple model where
+nonsynonymous escape mutations are unconditionally beneficial, they almost
+always fix once they reach high frequencies -- $A_{\mathrm{nonsyn}}$ is well
+above zero. This is incompatible with the blue line in \FIG{fixp}: in an HIV
+infection, nonsynonymous mutations at high frequency often disappear again, even
+though many are at least transiently beneficial. Inspecting the trajectories of
+nonsynonymous mutations suggests the rapid rise and fall of many alleles. We
+test two possible mechanisms that are biologically sound and could explain
+the transient rise of nonsynonymous mutations: time-dependent selection and
+within-epitope competition.
 
@@
 recognizes the escape mutant before its fixation, the mutant might cease to be
-beneficial and disappear soon, despite its quick initial rise in frequency. In support of this idea,
-\citet{richman_rapid_2003, bunnik_autologous_2008} report antibody responses to
-escape mutants. These responses are delayed by a few months, roughly matching the
-average time needed by an escape mutant to rise from low to high frequency.
-To model this type of behavior, we assume that anti-body
-responses against escape mutations arise with a rate proportional to the
-frequency against the escape mutation and abolish the benefit of the escape
-mutations. As expected, this type of time dependent
-fixation retains the potential for hitch-hiking, but reduces fixation of
-non-synonymous mutations. Fig.~S\timedependence~shows that $\pfix$ of
-synonymous mutations is not affected by this chance, while $\pfix$ of non-synonymous
-mutations approaches the diagonal as the rate of recognition of escape mutants
-is increased.
+beneficial and disappear soon, despite its quick initial rise in frequency. In
+support of this idea, \citet{richman_rapid_2003, bunnik_autologous_2008} report
+antibody responses to escape mutants. These responses are delayed by a few
+months, roughly matching the average time needed by an escape mutant to rise
+from low to high frequency. To model this type of behavior, we assume that
+antibody responses against escape mutations arise with a rate proportional to
+the frequency against the escape mutation and abolish the benefit of the escape
+mutations. As expected, this type of time dependent fixation retains the
+potential for hitchhiking, but reduces fixation of nonsynonymous mutations.
+\figurename~S\timedependence~shows that $\pfix$ of synonymous mutations is not
+affected by this chance, while $\pfix$ of nonsynonymous mutations approaches
+the diagonal as the rate of recognition of escape mutants is increased.
 
-In the alternative hypothesis, several different escape
-mutations within the same epitope might arise almost simultaneously and start to
-spread. Their benefits are not additive, because each of them is
-essentially sufficient to escape. As a consequence, several escape mutations rise to
-high frequency rapidly, while the one with the smallest cost in terms of replication,
-packaging, etc.~is most likely to eventually fix. The emergence of
-multiple sweeping nonsynonymous mutations in real HIV infections has been 
-shown~\citep{moore_limited_2009, bar_early_2012}.
-Within epitope competition can be implemented in the model through epistasis
-between escape mutations. While each mutation is individually beneficial,
-combining the mutations is deleterious (no extra benefit, but additional costs).
-Again, we find that the potential for hitch-hiking is little affected by within
-epitope competition, but that the fixation probability of non-synonymous
-polymorphisms is reduced. With roughly 6 mutations per epitope, the simulation
-data is compatible with observations; see Fig.~S\withinepi.
-The two scenarios are not exclusive and possibly both important in HIV
-evolution.
+In the alternative hypothesis, several different escape mutations within the
+same epitope might arise almost simultaneously and start to spread. Their
+benefits are not additive, because each of them is essentially sufficient to
+escape. As a consequence, several escape mutations rise to high frequency
+rapidly, while the one with the smallest cost in terms of replication,
+packaging, etc. is most likely to eventually fix. The emergence of multiple
+sweeping nonsynonymous mutations in real HIV infections has been shown
+\citep{moore_limited_2009, bar_early_2012}. Within epitope competition can be
+implemented in the model through epistasis between escape mutations. While each
+mutation is individually beneficial, combining the mutations is deleterious (no
+extra benefit, but additional costs). Again, we find that the potential for
+hitchhiking is little affected by within epitope competition, but that the
+fixation probability of nonsynonymous polymorphisms is reduced. With roughly 6
+mutations per epitope, the simulation data is compatible with observations; see
+\figurename~S\withinepi. The two scenarios are not exclusive and possibly both
+important in HIV evolution.
 
@@
 we demonstrate selection against synonymous substitutions in the comparatively
-conserved regions C2-C5 of the \env~gene. Comparison with biochemical studies of binding propensity of bases in RNA
-genome of HIV indicates that these mutations are deleterious, at least in part,
-because they disrupt stems in RNA secondary structures. Computational modelling
-shows that these mutations have deleterious effects on the order of $0.002$
-and that they are brought to high frequency through linkage to adaptive
-mutations.
+conserved regions C2-C5 of the \env{} gene. Comparison with biochemical studies
+of binding propensity of bases in RNA genome of HIV indicates that these
+mutations are deleterious, at least in part, because they disrupt stems in RNA
+secondary structures. Computational modeling shows that these mutations have
+deleterious effects on the order of $0.002$ and that they are brought to high
+frequency through linkage to adaptive mutations.
 
@@
 summary statistics useful to characterize longitudinal sequence data and compare
-to models via computer simulations. A method that is similar to ours {\it in spiritu}
-has been recently used in a longitudinal study of influenza
+to models via computer simulations. A method that is similar to ours {\it in
+spiritu} has been recently used in a longitudinal study of influenza
 evolution~\citep{strelkowa_clonal_2012}. The propagators suggested in that
@@
 mutations. The latter is used as an approximately neutral control; this method
-can therefore not be used to investigate synonymous changes themselves. 
-More generally, evolutionary rates at synonymous sites are often used as a
-baseline to detect purifying or diversifying selection at the protein level
+can therefore not be used to investigate synonymous changes themselves. More
+generally, evolutionary rates at synonymous sites are often used as a baseline
+to detect purifying or diversifying selection at the protein level
 \cite{Hurst:2002p32608}. It has been pointed out, however, that the rate of
@@
 A functional significance of the insulating RNA structure stems between the
-hyper-variable loops has also been proposed
-previously~\citep{watts_architecture_2009, sanjuan_interplay_2011} and conserved
-RNA structures exist in different parts of the HIV genome. Our analysis is able
-to quantify the fitness effect of RNA structure within single infections and
+hypervariable loops has also been proposed previously
+\citep{watts_architecture_2009, sanjuan_interplay_2011} and conserved RNA
+structures exist in different parts of the HIV genome. Our analysis is able to
+quantify the fitness effect of RNA structure within single infections and
 demonstrates how selection at synonymous sites can alter genetic diversity and
@@
 infrequent recombination for the evolution of HIV
-\citep{neher_recombination_2010,batorsky_estimate_2011,
+\citep{neher_recombination_2010, batorsky_estimate_2011,
 josefsson_majority_2011}. The recombination rate has been estimated to be on the
@@
 we have $l\approx 100$ bases. Hence we expect strong linkage between the
-variable loops and the intervening sequence, but none far beyond the variable
-regions, consistent with the lack of signal outside of C1-V5. In case of much
-stronger selection -- such as observed during early CTL escape or drug
+variable loops and the flanking sequences, but none far beyond the variable
+regions, consistent with the lack of signal outside of \shankaregion. In case of
+much stronger selection -- such as observed during early CTL escape or drug
 resistance evolution -- the linked region is of course much larger
@@
 
-While classical population genetics
-assumes that the dominant stochastic force is genetic drift, i.e., not heritable
-fluctuations in offspring number, our results show that stochasticity due to
-linked selection is much more important. Such fluctuations have been termed
-\emph{genetic draft} by \citet{gillespie_genetic_2000}. Genetic draft in
-facultatively sexual population like HIV has been characterized in
-\citep{neher_genetic_2011}. Importantly, large population sizes are compatible
-with low diversity and fast coalescence when draft dominates over drift.
+While classical population genetics assumes that the dominant stochastic force
+is genetic drift, i.e. nonheritable fluctuations in offspring number, our
+results show that stochasticity due to linked selection is much more important.
+Such fluctuations have been termed \emph{genetic draft} by
+\citet{gillespie_genetic_2000}. Genetic draft in facultatively sexual population
+such as HIV has been characterized in \citep{neher_genetic_2011}. Importantly,
+large population sizes are compatible with low diversity and fast coalescence
+when draft dominates over drift.
 
-Contrary to na\"ive expectations, the adaptive escape mutations do not seem to be
-unconditionally beneficial. Otherwise we would observe almost sure fixation of
-nonsynonymous mutations once they reach intermediate frequencies. Instead, we
+Contrary to na\"ive expectations, the adaptive escape mutations do not seem to
+be unconditionally beneficial. Otherwise we would observe almost sure fixation
+of nonsynonymous mutations once they reach intermediate frequencies. Instead, we
 find that the fixation probability of nonsynonymous mutations is roughly given
@@
 Similar to synonymous mutations, the majority of nonsynonymous mutations could
-be weakly deleterious and the adaptive and deleterious parts conspire to yield
-the neutral-like averaged fixation probability. We consider this possibility
-inplausible since the amino-acid sequence outside the variable loops is much
+be weakly deleterious and the adaptive and deleterious parts conspire to yield a
+more neutral-like averaged fixation probability. We consider this possibility
+inplausible since the amino acid sequence outside the variable loops is much
 more conserved that the synonymous positions, suggesting that the majority of
@@
 \citep{moore_limited_2009, bar_early_2012}. Presumably, each one of these
-mutations is sufficient for escape but most combinations of them don't provide
+mutations is sufficient for escape but most combinations of them do not provide
 any additional benefit to the virus. Hence only one mutation will spread and the
@@
 mutations rapidly. A similar point has been made recently by Boltz {\it et al.}
-in the context of preexisting drug resistance
-mutations~\citep{boltz_ultrasensitive_2012}.
+in the context of preexisting drug resistance mutations
+\citep{boltz_ultrasensitive_2012}.
 
@@
 genetic variation is only transiently beneficial, existing estimates of the
-strength of selection \citep{neher_recombination_2010,batorsky_estimate_2011} could be
-substantial underestimates.
+strength of selection \citep{neher_recombination_2010,batorsky_estimate_2011}
+could be substantial underestimates.
 
@@
 Longitudinal intrapatient viral RNA sequences were collected from published
-studies~\citep{shankarappa_consistent_1999, liu_selection_2006,
+studies \citep{shankarappa_consistent_1999, liu_selection_2006,
 bunnik_autologous_2008} and downloaded from the Los Alamos National Laboratory
 (LANL) HIV sequence database~\citep{LANL2012}. The samples from some patients
-show substantial population structure and were discarded (see supplementary
-materials); a total of 11 patients with 4-23 time points each and approximately 10
-sequences per time point were analyzed. The time intervals between two
-consecutive sequences ranged from 1 to 34 months, most of them between 6 and 10
-months.
+show substantial population structure and were discarded (see
+\figurename~S\PCApa); a total of 11 patients with 4-23 time points each and
+approximately 10 sequences per time point were analyzed. The time intervals
+between two consecutive sequences ranged from 1 to 34 months, most of them
+between 6 and 10 months.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-The sequences were translated and the resulting amino-acid sequences aligned
+The sequences were translated and the resulting amino acid sequences aligned
 using Muscle~\citep{edgar_muscle:_2004} to each other and the NL4-3 reference
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-For the estimate of times to fixation/extinction, polymorphisms were
-binned by frequency and the time to first reaching either fixation or
-extinction was stored. The fixation probability was determined as the long-time
-limit of the resulting curves. Mutations that reached high frequency but neither
-fixed nor got extinct were classified as ``floating'', with one exception: if they first
-reached high frequencies within 3 years of the last time point, we assumed
-they had not had sufficient time to settle, so we discarded them.
+For the estimate of times to fixation/extinction, polymorphisms were binned by
+frequency and the time to first reaching either fixation or extinction was
+stored. The fixation probability was determined as the long-time limit of the
+resulting curves. Mutations that reached high frequency but neither fixed nor
+got extinct were classified as ``floating'', with one exception: if they first
+reached high frequencies within 3 years of the last time point, it was assumed
+they had not had sufficient time to settle, so they were discarded.
 
 The SHAPE scores quantifying the degree of base pairing of individuals sites in
-the HIV genome were downloaded from the journal
-website~\citep{watts_architecture_2009}.
-Where possible, SHAPE reactivities were assigned to sites in the multiple
-sequence alignments for each patient through the alignment to the sequence
-of the NL4.3 virus used in ref.~\citep{watts_architecture_2009}. Problematic
-assignments in indel-rich regions were excluded from the analysis.
-The variable loops and flanking regions were identified manually starting from
-the annotated reference HXB2 sequence from the LANL HIV database~\citep{LANL2012}.
+the HIV genome were downloaded from the journal website
+\citep{watts_architecture_2009}. Wherever possible, SHAPE reactivities were
+assigned to sites in the multiple sequence alignments for each patient through
+the alignment to the sequence of the NL4.3 virus used in
+ref.~\citep{watts_architecture_2009}. Problematic assignments in indel-rich
+regions were excluded from the analysis. The variable loops and flanking
+regions were identified manually starting from the annotated reference HXB2
+sequence from the LANL HIV database~\citep{LANL2012}.
 
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Computer simulations}
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-Computer simulations were performed using
-FFPopSim~\citep{zanini_ffpopsim:_2012}. Briefly, FFPopSim enables individual
-based simulations where each site in the genome is represented by one bit that
-can be in one of two states. Outcrossing rates, crossover rates, mutations rates
-and arbitrary fitness functions can be specified. We used a generation time of 1
-day, an outcrossing rate of $r=0.01$ per
-day \citep{batorsky_estimate_2011,neher_recombination_2010}, a mutation rate of
-$\mu=10^{-5}$ \citep{mansky_lower_1995,abram_nature_2010} and simulated
-intra-patient evolution for 6000 days. For simplicity, all third position in a
-codon where deemed synonymous and assigned either a selection coefficient $0$
-with probability $1-\alpha$ or a deleterious effect $s_d$ with probability
-$\alpha$.
-First and second positions have strongly deleterious fitness effects
-0.02. At at rate $k_A$, a random place in the genome is
-designated an epitope that can escape by one or several mutations with an
-exponentially distributed escape rate with mean $\epsilon$. Both full-length HIV
-genomes and \env{}-only simulations were performed and yielded comparable results.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+Computer simulations were performed using FFPopSim
+\citep{zanini_ffpopsim:_2012}. Briefly, FFPopSim enables individual-based
+simulations where each site in the genome is represented by one bit that can be
+in one of two states. Outcrossing rates, crossover rates, mutations rates and
+arbitrary fitness functions can be specified. We used a generation time of 1
+day, an outcrossing rate of $r=0.01$ per day \citep{batorsky_estimate_2011,
+neher_recombination_2010}, a mutation rate of $\mu=10^{-5}$
+\citep{mansky_lower_1995, abram_nature_2010} and simulated intrapatient
+evolution for 6000 days. For simplicity, all third position in a codon where
+deemed synonymous and assigned either a selection coefficient $0$ with
+probability $1-\alpha$ or a deleterious effect $s_d$ with probability $\alpha$.
+First and second positions have strongly deleterious fitness effects 0.02. At at
+rate $k_A$, a random place in the genome is designated an epitope that can
+escape by one or several mutations with an exponentially distributed escape rate
+with mean $\epsilon$. Both full-length HIV genomes and \env{}-only simulations
+were performed and yielded comparable results.
 
-The simulations are repeated 2400 times with random choices for the following
-parameters: The fraction of deleterious sites $\alpha$ was sampled uniformly
+The simulations were repeated 2400 times with random choices for the following
+parameters: the fraction of deleterious sites $\alpha$ was sampled uniformly
 between 0.75 and 1.0; the average deleterious effect $s_d$ was sampled such that
@@
 was included. Higher order terms compensated each other to make sure that not
-only double, but all k-mutants with $k \geq 1$ had the same fitness. To model
-recognition of escape variants by the immune system that is catching up, the
-beneficial effect of an escape mutation was set to zero with a probability per
-generation proportional to the frequency of the escape variant.
+only double, but all k-mutants with $k \geq 1$ had the same fitness (see
+supplementary materials). To model recognition of escape variants by the immune
+system that is catching up, the beneficial effect of an escape mutation was set
+to its previous cost of -0.02 with a probability per generation proportional to
+the frequency of the escape variant.
 
@@
 
-The areas below or above the neutral fixation probability (diagonal line)
-were estimated from the binned fixation probabilities using linear
-interpolation between the bin centers. This measure is sufficiently precise for
-our purposes. In 10 runs out of 2400, the highest frequency bin was empty so the
-fixation probability could not be calculated; those runs were excluded from 
+The areas below or above the neutral fixation probability (diagonal line) were
+estimated from the binned fixation probabilities using linear interpolation
+between the bin centers. This measure is sufficiently precise for our purposes.
+In 10 runs out of 2400, the highest frequency bin was empty so the fixation
+probability could not be calculated; those runs were excluded from
 \FIG{simsfig}.

### 2013-02-13 supplementary.tex
types and edits in figure captions

@@
 \newcommand{\Affiliation}{Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany}
+\newcommand{\pfix}{P_{\mathrm{fix}}}
+\newcommand{\env}{\textit{env}}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 seroconversion, from blue to red). Panel B) Allele frequency trajectories for nonsynonymous
-changes in the same patient. Panels C and D) show analogous plots for data from
-patient p7. Samples after day 1000 split into two clusters in the PCA and no
-mutations that arise after day 1000 fix, presumably because they are restricted
+changes in the same patient. Here, the blue to red color map corresponds to the
+position of the allele in \env{} from 5' to 3'. Panels C and D) show analogous
+plots for data from patient p7. Samples after day 1000 split into two clusters in the PCA and no mutations that arise after day 1000 fix, presumably because they are restricted
 to one subpopulation. All patients like p7 (p4, p7, p8, p9 from ref.~\citealp{shankarappa_consistent_1999} and
@@
 Synonymous diversity across the HIV genome, as quantified by the normalized
-codon entropy among sequences coding for the consensus amino acid. The
-synonymous diversity peaks at the variable regions in {\it env} and is reduced
-in regions under purifying selection (RRE hairpin, second {\it tat}/{\it rev}
-exons). Since the variable regions are under positive selection at the protein
-level, synonymous hitchhiking is stronger there, and even deleterious synonymous
-mutations can be brought to high frequency. This gives to our signal in the
-fixation probability, Figs. 2A and 2B of the main text. The normalized codon
-entropy is calculated as follows (see the script for the full algorithm): (i)
-from a subtype B multiple sequence alignment (MSA) from the LANL website
-(filtered sequences only, version 2011)~\cite{LANL2012}, we calculate the
+codon entropy among sequences coding for the consensus amino acid. In most
+parts of the genome, synonymous sites show little conservation. The synonymous
+diversity peaks at the variable regions in {\it env} and is reduced in regions 
+under purifying selection (RRE hairpin, second {\it tat}/{\it rev} exons). The
+normalized codon entropy is calculated as follows (see the script
+\texttt{codon\_entropy\_synonymous\_subtypeB.py} for the full algorithm): (i)
+from a subtype B multiple sequence alignment (MSA) from the LANL website (filtered sequences only, version 2011)~\cite{LANL2012}, we calculate the
 consensus amino acid at each position in the HIV genome; (ii) we count how often
@@
 {\it env} that are part of a different gene (signaling peptide, second {\it rev}
-exon) have been excluded from our main analysis, to avoid contamination about
-synonymity. Note that all gap-rich columns of the MSA are stripped from this
-figure, hence genes such as {\it env} might appear shorter than they actually
+exon) have been excluded from our main analysis, to avoid contamination by
+protein selection in a different reading frame.
+Note that all gap-rich columns of the MSA are stripped from this figure, hence genes such as {\it env} might appear shorter than they actually
 are.
@@
 \caption{
-Time-dependent selection (the escape mutant is recognized after some months from
-its appearance) reduces fixation of nonsynonymous
-mutations. A ``na\"ive'' computer simulation shows a depression in synonymous
-fixation probability, as observed in patients, but the nonsynonymous fixation
-probability is much higher than the (neutral) diagonal: a nonsynonymous mutation
-that reaches a frequency around 0.2 is already very likely to fix. In data from
-infected patients, however, nonsynonymous mutations show a neutral-like curve,
-$P_\text{fix}(\nu) \approx \nu$. In the other curves, we show that the
-nonsynonymous fixation probability can be reduced by time-dependent selection
-without affecting much the synonymous fixation probability itself. The fitness
-advantage of an escape mutaition can vanish at each time point with a
-probability density \[ P_\text{recognized}(t) = c \cdot \nu(t), \] where $c$ is
-a constant coefficient shown in the legend that encodes the overall efficiency
-of the host immune system, and $\nu(t)$ is the frequency of the escape allele at
-time $t$. After recognition, the former escape mutation has the same fitness
-cost as before the creation of the epitope, hence it is quickly lost. Note that,
-although the curves shown here show almost no reduction in fixation probability
-up to $\nu \approx 0.15$, this effect is not robust and depends on the exact form of
-$ P_\text{recognized}(t)$. The other parameters for the shown simulations are
+Time-dependent selection reduces fixation of nonsynonymous mutations. The figure
+compares the fixation probability in the time independent model (na\"ive) to
+a model with time dependent selection that mimics the an evolving immune system.
+It has been found that virus is typically neutralized by serum from a few month
+earlier\citep{richman_rapid_2003} but not by contemporary serum. We model this
+evolving immune system by assuming that escaped variants loose their beneficial
+effect with a probability proportional to the frequency of the escaped variant
+per generation. Specifically, the selection effect of the escape mutations is
+reset to its fitness cost of $-0.02$ with probability
+\[ P_\text{recognized}(t) = c \cdot \nu(t), \] 
+per generation, where $c$ is a constant coefficient shown in the legend that
+encodes the overall efficiency of the host immune system. With increasing
+probability of recognition, the fixation of frequent escape mutants is reduced,
+while hitch-hiking of synonymous mutations is not affected. The precise
+shape of $\pfix(\nu)$ depends on the details of the $P_\text{recognized}(t)$ and 
+we do not think that the high $\pfix(\nu)$ for $\nu<0.2$ is meaningful.
+The other parameters for the shown simulations are
 the following: deleterious effect $10^{-3}$, average escape rate $0.016$,
@@
 \caption{
-Competition between escape mutations in the same epitope also reduces fixation of
-nonsynonymous mutations. Compared to a ``na\"ive'' model,
-simulations with 3 or 6 escape mutations competing within the same epitope show
-reduced fixation probabilities of nonsynonymous changes, whereas the synonymous
-changes behave similarly in all cases. The epitope length is 9 nucleotides. The
-fitness landscape of each such epitope include negative epistatic terms, so that
+Competition between escape mutations in the same epitope reduces fixation of
+nonsynonymous mutations. The figure compares the fixation probability of models
+with one, three, or six mutually exclusive escape mutations within the
+same epitope. Within epitope competition results reduced fixation
+probabilities of nonsynonymous changes, whereas the synonymous changes behave 
+similarly in all cases. We assume that escape can happen at $n$ sites out of 3
+consecutive codons and vary $n$.
+The fitness landscape of each epitope includes negative epistatic terms, so that
 the joint presence of more than one escape mutation is not any more beneficial
-for the virus than a single mutation. In detail, each site has two alleles, $\pm
-1$, where $-1$ is the ancestral one and $+1$ the dervied one; the fitness
-coefficient of a $k$-uple of sites within the epitope is $f_k = \pm
-\eta_\epsilon / 2^{L-1}$, where $\eta_\epsilon$ is a random variable for the
-escape rate with average value $\epsilon$ (see methods), $L$ is the number of
-competing escapes in the epitope, and the positive (resp. negative) sign is for
-odd (resp. even) $k$. In this evolutionary scenario, many escape mutations start
-to sweep on different backgrounds within the viral population, but eventually
+for the virus than a single mutation. Specifically, each site has two alleles,
+$\pm 1$, where $-1$ is the ancestral one and $+1$ the derived one; the fitness
+coefficient of a $k$-tuple of sites within the epitope is $f_k = (-1)^{k-1}
+2^{1-n}\eta_\epsilon $, where $\eta_\epsilon$ is the escape rate of the epitope
+drawn from an exponential distribution with mean $\epsilon$, 
+$n$ is the number of competing escapes in the epitope. 
+In this evolutionary scenario, many escape mutations start to sweep on different backgrounds within the viral population, but eventually
 compete and only one of them if fixed. The other parameters for the shown simulations are

### 2013-02-13 synmut.tex
types and edits in figure captions

@@
 mutation, conditional on its population frequency. A neutral mutation
-segregating at frequency $\nu$ has a probability $P_\text{fix}(\nu) = \nu$ to
+segregating at frequency $\nu$ has a probability $\pfix(\nu) = \nu$ to
 spread through the population and fix; in the rest of the cases, i.e. with
@@
 \citet{bunnik_autologous_2008} only, no such reduction in $\pfix$ is found.
-Restricted to the \shankaregion area, the data from
-\citet{bunnik_autologous_2008} is fully compatible with data from
+Restricted to the \shankaregion{} region, the sequence samples from
+\citet{bunnik_autologous_2008} are fully compatible with data from
 \citet{shankarappa_consistent_1999}. The nonsynonymous mutations seem to follow
@@
 of synonymous (A) and nonsynonymous (B) derived alleles in \env, \shankaregion, from patient
-1~\cite{shankarappa_consistent_1999}.
+10~\cite{shankarappa_consistent_1999}.
 While many nonsynonymous mutations  fix, few synonymous
 mutations do even though they are frequently observed at intermediate
-frequencies. Colors indicate the position of the site along the \shankaregion region
-(blue to red). Inset: the fixation probability $P_\text{fix}(\nu)$ of a neutral mutation
-is simply the likelihood that the future common ancestor is currently carrying
+frequencies. Colors indicate the position of the site along the \shankaregion{} region
+(blue to red). Inset: the fixation probability $\pfix$ of a neutral
+mutation is simply the likelihood that the future common ancestor is currently carrying
 it, i.e., its frequency $\nu$.}
@@
 frequencies in the part of \env{} that includes the hypervariable regions are
-deleterious, while outside this regions polymorphisms are mostly roughly
+deleterious, while outside this regions most polymorphisms are roughly
 neutral. Note that this does not imply that all synonymous mutations in this
@@
 \caption{Fixation and loss of synonymous mutations.
-Panel A) shows the time
-course of loss and fixation of synonymous mutations observed in a frequency interval $\nu_0$. 
+Panel A) shows the fraction of polymorphisms that have fixed (lower curves),
+are lost (distances between upper curves and 1), or remain polymorphic a time
+$\Delta t$ later for different initial allele frequencies (colors, indicated as
+shaded area).
 In each frequency interval, the fraction of synonymous
@@
 initial frequency.
-Panel B) Fixation probability of derived synonymous
-alleles as a function of $\nu_0$ in \shankaregion is reduced relative to the neutral
-expectation indicated by the diagonal line. This suppression is not
-observed in other parts of \env~or for nonsynonymous mutations.
+Panel B) shows the fixation probability of derived synonymous
+alleles as a function of $\nu_0$. Polymorphisms within \shankaregion{} fix less
+often than expected for neutral mutations indicated by the diagonal line.
+This suppression is not observed in other parts of \env~or for nonsynonymous mutations.
 The horizontal error bars on the abscissa are bin sizes, the vertical ones the
@@
 histogram) are systematically larger than of alleles that are lost (blue)
-(Kolmogorov-Smirnov test on the cumulative distribution, $P\approx 0.002$). In
+(Kolmogorov-Smirnov test on the cumulative distribution, $p\approx 0.002$). In
 other words, alleles that are likely to break RNA helices are also more likely
@@
 variable loops show little deviation from the neutral signature, see
-\FIG{SHAPEB}. This is consistent with important stem structures in conservered
+\FIG{SHAPEB}. This is consistent with important stem structures in conserved
 regions between loops.
@@
 less base pairing.
-Mutations that are never observered show an intermediate distribution of SHAPE values.
+Mutations that are never observed show an intermediate distribution of SHAPE values.
 Panel B) shows the fixation probability of synonymous mutations in
@@
 \caption{Distribution of selection coefficients on synonymous sites. Panel A)
-The depression in $P_\text{fix}$ depends on the deleterious effect size 
+The depression in $\pfix$ depends on the deleterious effect size 
 of the synonymous alleles. This parameter also reduces synonymous
@@
 
-The above simulation show that hitchhiking can explain the obervation of
+The above simulation show that hitchhiking can explain the observation of
 deleterious mutations that rarely fix. However, in a simple model where
@@
 While classical population genetics assumes that the dominant stochastic force
-is genetic drift, i.e. nonheritable fluctuations in offspring number, our
+is genetic drift, i.e. non-heritable fluctuations in offspring number, our
 results show that stochasticity due to linked selection is much more important.
@@
 more neutral-like averaged fixation probability. We consider this possibility
-inplausible since the amino acid sequence outside the variable loops is much
+implausible since the amino acid sequence outside the variable loops is much
 more conserved that the synonymous positions, suggesting that the majority of
@@
 is experimental evidence in support of both of these hypotheses. Serum from HIV
-infected individuals tyically neutralizes the virus that dominated the
+infected individuals typically neutralizes the virus that dominated the
 population a few (3-6) month ago \citep{richman_rapid_2003}. This suggests that
@@
 ``ancestral'' or ``derived'' at all sites. Sites that include large
-frequencies of gaps were excluded from the analysis to avoid artefactual
+frequencies of gaps were excluded from the analysis to avoid artifactual
 substitutions due to alignment errors. Allele frequencies at different time

### 2013-02-15 synmut.tex
word version for Jan

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\documentclass[rmp, twocolumn]{revtex4}
+\documentclass[rmp, onecolumn]{revtex4}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 brought up to high frequency by hitchhiking on neighboring beneficial
-nonsynonymous alleles. The patterns of fixation of nonsynonymous mutations
+nonsynonymous alleles.
+Besides, the patterns of fixation of nonsynonymous mutations estimated from the
+longitudinal data and comparisons with computer models
 suggest that escape mutations in \shankaregion{} are only transiently
@@
 \begin{center}
-\subfloat{\includegraphics[width=\linewidth]{simulations_graduallydel.pdf}
+\subfloat{\includegraphics[width=0.7\linewidth]{simulations_graduallydel.pdf}
 \label{fig:simfixpvar}}\\
-\subfloat{\includegraphics[width=\linewidth]{simulations_syn.pdf}
+\subfloat{\includegraphics[width=0.7\linewidth]{simulations_syn.pdf}
 \label{fig:simsfig}}

### 2013-02-26 synmut.tex
Jan suggestions (fig labels)

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\documentclass[rmp, onecolumn]{revtex4}
+\documentclass[rmp, twocolumn]{revtex4}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{center}
-\subfloat{\includegraphics[width=0.9\linewidth]{fixation_times}
+\subfloat{\includegraphics[width=0.9\linewidth]{fixation_times.pdf}
 \label{fig:fixp1}}\\
-\subfloat{\includegraphics[width=0.9\linewidth]{fixation_probabilities}
+\subfloat{\includegraphics[width=0.9\linewidth]{fixation_probabilities.pdf}
 \label{fig:fixp2}}
@@
 \begin{center}
-\subfloat{\includegraphics[height=0.46\linewidth]{reactivities_histograms_syn}
-\label{fig:SHAPEA}}
-\subfloat{\includegraphics[height=0.46\linewidth]{fixation_probabilities_VnonV}\label{fig:SHAPEB}}
+\subfloat{\includegraphics[height=0.47\linewidth]{reactivities_histograms_syn.pdf}\label{fig:SHAPEA}}
+\subfloat{\includegraphics[height=0.46\linewidth]{fixation_probabilities_VnonV.pdf}\label{fig:SHAPEB}}
 \caption{Permissible synonymous mutations tend to be unpaired.
@@
 show substantial population structure and were discarded (see
-\figurename~S\PCApa); a total of 11 patients with 4-23 time points each and
+\figurename~S\PCApat); a total of 11 patients with 4-23 time points each and
 approximately 10 sequences per time point were analyzed. The time intervals

### 2013-02-26 synmut.tex
incorporated Pleuni's and Jan's comments, still need to do smth about the interpretation/nonsyn deleterious hitchhikers

--- 2013-02-15 synmut.tex
@@
 \newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
-\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV \env~evolution}
+\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV-1 \env~evolution}
 \newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
@@
 \noindent
-
-Intrapatient HIV evolution is dominated by selection on the protein level in the
-arms race with the adaptive immune system. When killer T cells or antibodies
+Intrapatient HIV-1 evolution is dominated by selection on the protein level in the
+arms race with the adaptive immune system. When cytotoxic CD8${}^+$ T-cells or neutralizing antibodies
 target a new epitope, the virus often escapes via nonsynonymous mutations that
@@
 changes, hinting at a direct fitness effect of these stem-loop structures in the
-HIV RNA.
+HIV-1 RNA.
 Computational modeling indicates that these synonymous mutations have a
@@
 nonsynonymous alleles.
-Besides, the patterns of fixation of nonsynonymous mutations estimated from the
+The patterns of fixation of nonsynonymous mutations estimated from the
 longitudinal data and comparisons with computer models
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-HIV evolves rapidly within a single host during the course of the infection.
+HIV-1 evolves rapidly within a single host during the course of the infection.
 This evolution is driven by strong selection imposed by the host immune system
-via killer T cells (CTLs) and neutralizing antibodies
-(ABs)~\citep{rambaut_causes_2004} and facilitated by the high mutation rate of
-HIV~\citep{mansky_lower_1995,abram_nature_2010}. When the host develops a CTL or
-AB response against a particular HIV epitope, mutations in the viral genome that
+via cytotoxic CD8${}^+$ T-cells (CTLs) and neutralizing antibodies
+(nAbs)~\citep{rambaut_causes_2004} and facilitated by the high mutation rate of
+HIV-1 (HIV in the following)~\citep{mansky_lower_1995,abram_nature_2010}. When the host develops a CTL or
+nAb response against a particular HIV epitope, mutations in the viral genome that
 reduce or prevent recognition of the epitope frequently emerge. Escape mutations
@@
 loops V1-V5 in the envelope protein gp120, which change to avoid recognition by
-neutralizing ABs. Mutations in \env, the gene encoding gp120, spread through the
+nAbs. Escape mutations in \env, the gene encoding gp120, spread through the
 population within a few months (see \figurename~\ref{fig:aftnonsyn}).
 Consistent with this time scale, it is found that serum from a particular time
-typically neutralizes virus extracted more than 3-6 month earlier
-\citep{richman_rapid_2003}.
+typically neutralizes virus extracted more than 3-6 month earlier, but not contemporary
+virus \citep{richman_rapid_2003}.
 
-These escape mutations are selected for their effect on the amino acid sequence
+Escape mutations are selected for their effect on the amino acid sequence
 of the viral proteins. Conversely, synonymous mutations are commonly used as
@@
 argue that synonymous mutations reach high frequencies via genetic hitchhiking
-due to the small recombination rate of HIV~\citep{neher_recombination_2010,
+due to limited recombination in HIV populations~\citep{neher_recombination_2010,
 batorsky_estimate_2011}. We then compare our observations to computational
-models of HIV evolution and derive estimates for the effect synonymous mutations
-have on fitness.  Extending the analysis of fixation probabilities to the
+models of HIV evolution and derive estimates for the effect of synonymous mutations
+ on fitness.  Extending the analysis of fixation probabilities to the
 nonsynonymous mutations, we show that time dependent selection or strong
@@
 probability $1-\nu$, it goes extinct. As illustrated in the inset of
-\FIG{aftsyn}, this is a simple consequence of the fact that (i) exactly one
+\FIG{aftsyn}, this is a  consequence of the fact that (i) exactly one
 individual in the current population will be the common ancestor of the entire
-future population at a particular locus and (ii) this ancestor has a probability
+future population at a particular site and (ii) this ancestor has a probability
 $\nu$ of carrying the mutation (assuming the neutral mutation is not
@@
 constrast, many nonsynonymous mutations fix (panel~\ref{fig:aftnonsyn}).
+Strictly speaking, no mutation in the HIV population ever fixes because the 
+mutation rate and the population size are large. We define ``fixation'' or ``loss''
+by not observing the mutation in the sample.
 
@@
 We study the dynamics and fate of synonymous mutations more quantitatively by
-analyzing data from 7 patients from
-\citet{shankarappa_consistent_1999,liu_selection_2006} and 3 patients from
+analyzing data from seven patients from
+\citet{shankarappa_consistent_1999,liu_selection_2006} and three patients from
 \citet{bunnik_autologous_2008} (patients whose viral population was structured
@@
 the data from \citet{bunnik_autologous_2008} covers the majority of \env.
-Considering all mutations in a frequency interval around $\nu_0$ at some time
+Considering all mutations in a frequency interval $[\nu_0-\delta\nu, \nu_0+\delta\nu]$ at some time
 $t$, we calculate the fraction that is found at frequency 1, at frequency 0, or
@@
 A polymorphism could be beneficial to the virus and on its way to fixation. In
-this case, we expect that it fixes almost surely given we see it at high
+this case, we expect that it fixes almost surely given that we see it at high
 frequency. If, on the other hand, the polymorphism is deleterious it must have
@@
 observations suggest that many of the synonymous polymorphisms at intermediate
-frequencies in the part of \env{} that includes the hypervariable regions are
-deleterious, while outside this regions most polymorphisms are roughly
-neutral. Note that this does not imply that all synonymous mutations in this
-region are neutral -- only those mutations observed at high frequencies, which
+frequencies in the part of \env{} that includes the \shankaregion{} are
+deleterious, while outside this region most polymorphisms are roughly
+neutral. Note that this does not imply that all synonymous mutations in outside 
+\shankaregion{} are neutral -- only those mutations observed at high frequencies, which
 have been experiencing selection for some time already, tend to be neutral.
@@
 regions V1-V5 tend to be unpaired, while the conserved regions between those
-variable regions form stems. We partition all synonymous alleles observed at
-intermediate frequencies above 15\% depending on their final destiny (fixation
-or extinction). Subsequently, we align our sequences to the reference NL4-3
-strain used in ref.~\citep{watts_architecture_2009} and assign them SHAPE
-reactivities. As shown in \FIG{SHAPEA}, the reactivities of fixed alleles (red
+variable regions form stems.  We aligned the within-patient sequence samples 
+to the reference NL4-3 strain used in ref.~\citep{watts_architecture_2009} and 
+thereby assigned SHAPE reactivities to most positions in the alignment. 
+We then calculated the distributions of SHAPE reactivities for synonymous 
+polymorphisms that fixed or where subsequently lost (only polymorphisms with 
+frequencies above 15\%).
+As shown in \FIG{SHAPEA}, the reactivities of fixed alleles (red
 histogram) are systematically larger than of alleles that are lost (blue)
@@
 to revert and finally be lost from the population. The average over all
-mutations that are not observed (green) lies between the those that fix and
-those that get lost. Note that this analysis will be sensitive only at position
+mutations that are not observed (green) lies between  those that fix and
+those that get lost. Note that this analysis will be sensitive only at positions
 where the base pairing pattern of NL4-3 agrees with that of each patient's
@@
 In addition to RNA secondary structure, we have considered other possible
-explanations for a fitness defect of synonymous mutations, in particular codon
+explanations for a fitness cost of some synonymous mutations, in particular codon
 usage bias (CUB). HIV is known to prefer A-rich codons over highly expressed
@@
 \caption{Permissible synonymous mutations tend to be unpaired.
-Panel A) shows the distribution of SHAPE reactivities among synonymous mutations
-that fix, mutations that reach frequencies above 15\%, but
-are susequently lost, and random mutations (all categories are restricted to the
-regions V1-V5$\pm 100$bp).
-Fixed mutations tend to have higher higher SHAPE reactivities, corresponding to
-less base pairing.
-Mutations that are never observed show an intermediate distribution of SHAPE values.
+Panel A) shows the distribution of SHAPE reactivities among sites at which synonymous 
+mutations fixed (red), at which mutations reached frequencies above 15\% but
+were susequently lost (blue), and sites at which no mutations were observed (green) 
+(all categories are restricted to the regions V1-V5$\pm 100$bp).
+Sites at which mutations fixed tend to have higher SHAPE reactivities, corresponding to
+less base pairing, than those at which mutations are lost.
+Sites at which no mutations are observed show an intermediate distribution of SHAPE values.
 Panel B) shows the fixation probability of synonymous mutations in
-\shankaregion{} separately for variable loops and the connecting conserved regions that
-harbor RNA stems. As expected, the fixation probability is lower
+\shankaregion{} separately for variable regions V3-V5 and the connecting conserved 
+regions C2-C4 that harbor RNA stems. As expected, the fixation probability is lower
 inside the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
@@
 frequency and that the fixation probability is reduced only in parts of the
-genome. This regions, however, undergoes frequent adaptive changes to evade
+genome. These regions, however, undergoes frequent adaptive changes to evade
 recognition by neutralizing antibodies \cite{williamson_adaptation_2003,
@@
 \citep{gillespie_genetic_2000,neher_genetic_2011}. Hitchhiking is  apparent in
-\FIG{aft}, that shows that many mutations change rapidly in frequency as a
+\FIG{aft}, which shows that many mutations change rapidly in frequency as a
 flock.
@@
 deleterious effect of synonymous mutations, the fraction of synonymous mutations
-that are deleterious, the escape rate of adaptive nonsynonymous mutations and
-the frequency of new escapes. Note that the escape rate is the sum of two
+that are deleterious, the escape rate (selection coefficient) of adaptive 
+nonsynonymous mutations and the rate at which previously untargeted epitopes
+become targeted (the latter determines the number of sites available for escape). 
+Note that the escape rate is the sum of two
 factors: (i) the beneficial effect due to the ability to evade the immune system
@@
 decreases from the neutral expectation ($A_\text{syn} \sim 0$) to zero
-($A_\text{syn} \sim -0.5$) as their fitness effect worsens; the synonymous
+($A_\text{syn} \sim -0.5$) as their fitness cost increases; the synonymous
 diversity plummets as well, as deleterious mutations are selected against.
@@
 nonsynonymous mutations suggests the rapid rise and fall of many alleles. We
-test two possible mechanisms that are biologically sound and could explain
+test two possible mechanisms that are biologically plausible and could explain
 the transient rise of nonsynonymous mutations: time-dependent selection and
@@
 antibody responses against escape mutations arise with a rate proportional to
-the frequency against the escape mutation and abolish the benefit of the escape
-mutations. As expected, this type of time dependent fixation retains the
+the frequency of the escape mutation and abolish the benefit of the escape
+mutations. As expected, this type of time-dependent selection retains the
 potential for hitchhiking, but reduces fixation of nonsynonymous mutations.
 \figurename~S\timedependence~shows that $\pfix$ of synonymous mutations is not
-affected by this chance, while $\pfix$ of nonsynonymous mutations approaches
+affected by this change, while $\pfix$ of nonsynonymous mutations approaches
 the diagonal as the rate of recognition of escape mutants is increased.
@@
 sweeping nonsynonymous mutations in real HIV infections has been shown
-\citep{moore_limited_2009, bar_early_2012}. Within epitope competition can be
+\citep{moore_limited_2009, bar_early_2012}. This scenario has been explicitly 
+observed in evolution the evolution of resistance to 3TC, where the mutation 
+M184V is often preceeded by M184I \citep{hedskog_dynamics_2010}. Similarly, AZT 
+resistance often emerges via the competing TAM and TAM1 pathways.
+Within epitope competition can be
 implemented in the model through epistasis between escape mutations. While each
@@
 hitchhiking is little affected by within epitope competition, but that the
-fixation probability of nonsynonymous polymorphisms is reduced. With roughly 6
+fixation probability of nonsynonymous polymorphisms is reduced. With roughly six
 mutations per epitope, the simulation data is compatible with observations; see
@@
 conserved regions C2-C5 of the \env{} gene. Comparison with biochemical studies
-of binding propensity of bases in RNA genome of HIV indicates that these
+of base pairing propensity in RNA genome of HIV indicates that these
 mutations are deleterious, at least in part, because they disrupt stems in RNA
@@
 infected individuals typically neutralizes the virus that dominated the
-population a few (3-6) month ago \citep{richman_rapid_2003}. This suggests that
+population a few (3-6) month earlier \citep{richman_rapid_2003}. This suggests that
 escape mutations cease to be beneficial after a few months and might revert if
@@
 frequencies. The rapid emergence of multiple escape mutations in the same
-epitope implies a large population size that explores all necessary point
+epitope implies a large effective population size that explores all necessary point
 mutations rapidly. A similar point has been made recently by Boltz {\it et al.}
@@
 Our results emphasize the inadequacy of independent site models of HIV evolution
-and the common assumption that selection is time independent or additive. If
-genetic variation is only transiently beneficial, existing estimates of the
+and the common assumption that selection is time independent or additive. 
+If genetic variation is only transiently beneficial, existing estimates of the
 strength of selection \citep{neher_recombination_2010,batorsky_estimate_2011}
-could be substantial underestimates.
+could be substantial underestimates. Furthermore, weak conservation and
+time dependent selection results in estimates of evolutionary 
+rates that depend on the time interval of observation with lower rates across
+larger intervals. This could mean that deep nodes in phylogenies are older than 
+they appear.
 

### 2013-02-26 synmut.tex
typo

@@
 frequency and that the fixation probability is reduced only in parts of the
-genome. These regions, however, undergoes frequent adaptive changes to evade
+genome. These regions, however, undergo frequent adaptive changes to evade
 recognition by neutralizing antibodies \cite{williamson_adaptation_2003,

### 2013-02-28 synmut.tex
redid Fig 2a, adjusted text

@@
 Considering all mutations in a frequency interval $[\nu_0-\delta\nu, \nu_0+\delta\nu]$ at some time
-$t$, we calculate the fraction that is found at frequency 1, at frequency 0, or
-at intermediate frequency at later times $t+\Delta t$. Plotting these fixed,
-lost, and polymorphic fraction against the time interval $\Delta t$, we see that
+$t$, we calcate the fraction that are still observed at later times $t+\Delta t$. Plotting this fraction against the time interval $\Delta t$, we see that
 most synonymous mutations segregate for roughly one year and are lost much more
 frequently than expected (panel \ref{fig:fixp1}). The long-time probability of
-fixation versus extinction of synonymous mutations is shown as a function of the
+fixation, $\pfix$, is shown as a function of the
 initial frequency $\nu_0$ in panel~\ref{fig:fixp2} (red line). Restricted to the
@@
 \caption{Fixation and loss of synonymous mutations.
-Panel A) shows the fraction of polymorphisms that have fixed (lower curves),
-are lost (distances between upper curves and 1), or remain polymorphic a time
-$\Delta t$ later for different initial allele frequencies (colors, indicated as
-shaded area).
+Panel A) shows how quickly synonymous mutations are purged from the populations. 
+Specifically, the figure shows the fraction of mutations that are still observed
+after $\Delta t$ days, conditional on being observed in one of the three frequency 
+intervals (different colors). 
 In each frequency interval, the fraction of synonymous
-mutations that ultimately fix is the fixation probability conditional on the
-initial frequency.
+mutations that ultimately survive is the fixation probability $\pfix$ conditional on the
+initial frequency. The neutral expectation for $\pfix=\nu_0$ is indicated by 
+dashed horizontal lines.
 Panel B) shows the fixation probability of derived synonymous
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Synonymous mutations in \shankaregion tend to disrupt conserved RNA stems}
+\subsection{Synonymous mutations in \shankaregion{} tend to disrupt conserved RNA stems}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{center}
-\subfloat{\includegraphics[width=0.7\linewidth]{simulations_graduallydel.pdf}
+\subfloat{\includegraphics[width=0.9\linewidth]{simulations_graduallydel.pdf}
 \label{fig:simfixpvar}}\\
-\subfloat{\includegraphics[width=0.7\linewidth]{simulations_syn.pdf}
+\subfloat{\includegraphics[width=0.9\linewidth]{simulations_syn.pdf}
 \label{fig:simsfig}}

### 2013-02-28 scripts/fixation_loss_synonymous_times.py

@@ -13,7 +13,16 @@ import numpy as np
 import matplotlib.pyplot as ppl
 import matplotlib.cm as cm
 import matplotlib.patches as patches
-
+from matplotlib import rcParams
+params = {'backend': 'ps',
+        'axes.labelsize': 20, 
+        'text.fontsize': 20,
+    'legend.fontsize': 20,
+    'xtick.labelsize': 18,
+    'ytick.labelsize': 18,
+    'text.usetex': True}
+    
+rcParams.update(params)
 # Custom modules
 import sys
 sys.path.insert(0, '.')
@@ -225,7 +234,7 @@ if __name__ == '__main__':
                             (1 - np.linspace(0,P0,len(dt_loss)))[::-1]])
         ax.plot(x, y,
                 lw=2, c=colors[iji],
-                label=r'$\nu_0 = ['+str(nu0s[0])+','+str(nu0s[1])+']$')
+                label=r'$\nu_0 \in ['+str(nu0s[0])+','+str(nu0s[1])+']$')
     
         # Draw rectangle
         rec = patches.Rectangle((0,nu0s[0]), 2000, nu0s[1] - nu0s[0],
@@ -235,6 +244,40 @@ if __name__ == '__main__':
     ax.set_xlabel(r'time interval $\Delta t$ [days]', fontsize=16)
     ax.set_ylabel(r'$\nu$', fontsize=18)
     #ax.set_title(r'time interval $\Delta t [days]$')
+    ax.legend(loc=1)
+    ax.set_xlim(0, 2000)
+
+    ppl.ion()
+    ppl.show()
+
+    fig = ppl.figure()
+    ax = fig.add_subplot(1,1,1) 
+    for iji, nu0s in enumerate(nu0ss):
+ 
+        # Define the counts (lost, fixed, neither)
+        counts = counts_all[iji]
+        delta_ts = times_all[iji]
+        P0 = 1.0 * counts[0] / sum(counts)
+        P1 = 1.0 * counts[1] / sum(counts)
+    
+        # Plot the fixation/loss probability together as a function of time,
+        # by simply counting how many alleles (in percentage) had fixed/been lost after time dt
+        totcounts = sum(counts)
+        dt_fix = np.sort(delta_ts[1]) * mo_to_gen
+        dt_loss = np.sort(delta_ts[0]) * mo_to_gen
[…]

### 2013-02-28 scripts/fixation_probability_Vloops_flanking.py

@@ -259,7 +259,7 @@ if __name__ == '__main__':
                     c=colors[cla])
     ax.plot([0, 1], [0, 1], lw=2, ls='--', c='k')
     ax.set_xlabel(r'$\nu$', fontsize=18)
-    ax.set_ylabel(r'$P_\text{fix}(\nu)$', fontsize=18)
+    ax.set_ylabel(r'$P_\mathrm{fix}(\nu)$', fontsize=18)
     ax.set_xlim(-0.05, 1.05)
     ax.set_ylim(-0.05, 1.05)
     ax.legend(loc=2)

### 2013-02-28 synmut.tex
fig2a little aesthetics

@@
 Considering all mutations in a frequency interval $[\nu_0-\delta\nu, \nu_0+\delta\nu]$ at some time
-$t$, we calcate the fraction that are still observed at later times $t+\Delta t$. Plotting this fraction against the time interval $\Delta t$, we see that
+$t$, we calculate the fraction that are still observed at later times $t+\Delta t$. Plotting this fraction against the time interval $\Delta t$, we see that
 most synonymous mutations segregate for roughly one year and are lost much more

### 2013-03-01 supplement_body.tex
little edits here and there, moved supplement body into separate file such that we can include it into the main text as an appendix for submission

--- 2013-02-13 supplementary.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Template for a revtex article
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\documentclass[rmp]{revtex4}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\usepackage[english]{babel}
-\usepackage[utf8x]{inputenc}
-\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym,textcomp}
-\usepackage{color}
-\usepackage{graphicx}
-\usepackage[caption=false]{subfig}
-\usepackage{natbib}
-\usepackage{pslatex}
-\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\graphicspath{{./figures/}}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newcommand{\comment}[1]{{\color{red}#1}}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\renewcommand{\thesubfigure}{\Alph{subfigure}}
-\newcommand{\Author}{Fabio~Zanini* and Richard~A.~Neher*}
-\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV \env~evolution}
-\newcommand{\Affiliation}{Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany}
-\newcommand{\pfix}{P_{\mathrm{fix}}}
-\newcommand{\env}{\textit{env}}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{document}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{\Title}
-\author{\Author}
-\affiliation{\textsf{*}\Affiliation}
-\date{\today}
-\maketitle
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\tableofcontents
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: t
+%%% End:
 
@@
 \includegraphics[width=0.35\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p7}
-\caption{Structure of viral population and patient selection.
+\caption{Structure of viral populations and patient selection.
 Panel A) shows a PCA of all sequences from patient p1 (colors indicate time from
@@
 It has been found that virus is typically neutralized by serum from a few month
-earlier\citep{richman_rapid_2003} but not by contemporary serum. We model this
+earlier~\citep{richman_rapid_2003} but not by contemporary serum. We model this
 evolving immune system by assuming that escaped variants loose their beneficial
-effect with a probability proportional to the frequency of the escaped variant
-per generation. Specifically, the selection effect of the escape mutations is
+effect with a rate proportional to the frequency of the escaped variant. 
+Specifically, the selection effect of the escape mutations is
 reset to its fitness cost of $-0.02$ with probability
@@
 The other parameters for the shown simulations are
-the following: deleterious effect $10^{-3}$, average escape rate $0.016$,
-fraction of deleterious synonymous mutations $0.986$, rate of new epitopes
-$0.0014$ per generation.
+the following: deleterious effect $s_d = 10^{-3}$, average escape rate $\epsilon = 0.016$,
+fraction of deleterious synonymous mutations $\alpha = 0.986$, rate of new epitopes
+$k_A=0.0014$ per generation.
 }
@@
 In this evolutionary scenario, many escape mutations start to sweep on different backgrounds within the viral population, but eventually
-compete and only one of them if fixed. The other parameters for the shown simulations are
-the following: deleterious effect $10^{-3}$, average escape rate $0.016$,
-fraction of deleterious synonymous mutations $0.986$, rate of new epitopes
-$0.0014$ per generation.
+compete and only one of them fixes. The other parameters for the shown simulations are
+the following: deleterious effect $s_d = 10^{-3}$, average escape rate $\epsilon = 0.016$,
+fraction of deleterious synonymous mutations $\alpha = 0.986$, rate of new epitopes
+$k_A=0.0014$ per generation.
 }
@@
 \end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliographystyle{natbib}
-\bibliography{bib}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\end{document}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2013-03-01 supplementary.tex
little edits here and there, moved supplement body into separate file such that we can include it into the main text as an appendix for submission

@@
 \renewcommand{\thesubfigure}{\Alph{subfigure}}
-\newcommand{\Author}{Fabio~Zanini* and Richard~A.~Neher*}
+\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
 \newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV \env~evolution}
@@
 \author{\Author}
-\affiliation{\textsf{*}\Affiliation}
+\affiliation{\Affiliation}
 \date{\today}
@@
 \tableofcontents
+\include{supplement_body}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Selection of the patient data}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[ht]
-\begin{center}
-\includegraphics[width=0.35\linewidth]{Shankarappa_PCA_p1}
-\includegraphics[width=0.35\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p1}
-\includegraphics[width=0.35\linewidth]{Shankarappa_PCA_p7}
-\includegraphics[width=0.35\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p7}
-\caption{Structure of viral population and patient selection.
-Panel A) shows a PCA of all sequences from patient p1 (colors indicate time from
-seroconversion, from blue to red). Panel B) Allele frequency trajectories for nonsynonymous
-changes in the same patient. Here, the blue to red color map corresponds to the
-position of the allele in \env{} from 5' to 3'. Panels C and D) show analogous
-plots for data from patient p7. Samples after day 1000 split into two clusters in the PCA and no mutations that arise after day 1000 fix, presumably because they are restricted
-to one subpopulation. All patients like p7 (p4, p7, p8, p9 from ref.~\citealp{shankarappa_consistent_1999} and
-ACH19542 and ACH19768 from ref.~\citealp{bunnik_autologous_2008}) were excluded
-from our analysis.}
-\label{fig:aftp}
-\end{center}
-\end{figure}
-
-\newpage
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Synonymous diversity across the HIV genome}
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=\linewidth]{conservation_codons_genome}
-\caption{
-Synonymous diversity across the HIV genome, as quantified by the normalized
-codon entropy among sequences coding for the consensus amino acid. In most
-parts of the genome, synonymous sites show little conservation. The synonymous
-diversity peaks at the variable regions in {\it env} and is reduced in regions 
-under purifying selection (RRE hairpin, second {\it tat}/{\it rev} exons). The
-normalized codon entropy is calculated as follows (see the script
-\texttt{codon\_entropy\_synonymous\_subtypeB.py} for the full algorithm): (i)
-from a subtype B multiple sequence alignment (MSA) from the LANL website (filtered sequences only, version 2011)~\cite{LANL2012}, we calculate the
-consensus amino acid at each position in the HIV genome; (ii) we count how often
-each codon coding for the consensus amino acid appears in the MSA; (iii) at each
-amino acid position, we divide by the number of sequences in the MSA that had
-the consensus amino acid at that position, obtaining {\it codon frequencies}
-$\nu_c$; (iv) we calculate the codon entropy from each position as: $S := -
-\sum_{c} \nu_c \log \nu_c$, where $c$ runs over codons that code for the
-consensus amino acid at this site; (v) we divide by the maximal codon entropy of
-that amino acid (e.g. $\log 2$ for twofold degenerate codons). All parts of
-{\it env} that are part of a different gene (signaling peptide, second {\it rev}
-exon) have been excluded from our main analysis, to avoid contamination by
-protein selection in a different reading frame.
-Note that all gap-rich columns of the MSA are stripped from this figure, hence genes such as {\it env} might appear shorter than they actually
-are.
-}
-\label{fig:syndiv_genome}
-\end{center}
-\end{figure}
-\newpage
-% 
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \section{Nonsynonymous changes outside of variable regions are deleterious}
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \begin{figure}[h]
-% \begin{center}
-% \includegraphics[width=0.8\linewidth]{synmut_conservation_4fold_synnonsyn}
-% \caption{Cumulative distribution of synonymous and nonsynonymous diversity in
-% {\it gag} in the LANL reference panel (filtered
-% sequences only, version 2011)~\cite{LANL2012}. Sites such that the consensus codon
-% has three synonymous and six nonsynonymous single mutants were used, and the number of observed mutants
-% of a certain type over the number of possible mutants is plotted. Nonsynonymous
-% changes are observed less often, {\it ergo} are more conserved, than
-% synonymous changes. It can be therefore assumed, as mentioned in the main text,
-% that non-escape nonsynonymous changes involve a large fitness cost.}
-% \label{fig:synnonsyncons}
-% \end{center}
-% \end{figure}
-% \newpage
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Time-dependent selection}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=0.6\linewidth]{simulations_gradually_timeselec_deadagain}
-\caption{
-Time-dependent selection reduces fixation of nonsynonymous mutations. The figure
-compares the fixation probability in the time independent model (na\"ive) to
-a model with time dependent selection that mimics the an evolving immune system.
-It has been found that virus is typically neutralized by serum from a few month
-earlier\citep{richman_rapid_2003} but not by contemporary serum. We model this
-evolving immune system by assuming that escaped variants loose their beneficial
-effect with a probability proportional to the frequency of the escaped variant
-per generation. Specifically, the selection effect of the escape mutations is
-reset to its fitness cost of $-0.02$ with probability
-\[ P_\text{recognized}(t) = c \cdot \nu(t), \] 
-per generation, where $c$ is a constant coefficient shown in the legend that
-encodes the overall efficiency of the host immune system. With increasing
-probability of recognition, the fixation of frequent escape mutants is reduced,
-while hitch-hiking of synonymous mutations is not affected. The precise
-shape of $\pfix(\nu)$ depends on the details of the $P_\text{recognized}(t)$ and 
-we do not think that the high $\pfix(\nu)$ for $\nu<0.2$ is meaningful.
-The other parameters for the shown simulations are
-the following: deleterious effect $10^{-3}$, average escape rate $0.016$,
-fraction of deleterious synonymous mutations $0.986$, rate of new epitopes
-$0.0014$ per generation.
-}
-\label{fig:tds}
-\end{center}
-\end{figure}
-
-\newpage
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Within-epitope competition}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=0.6\linewidth]{simulations_gradually_epitopes}
-\caption{
-Competition between escape mutations in the same epitope reduces fixation of
-nonsynonymous mutations. The figure compares the fixation probability of models
-with one, three, or six mutually exclusive escape mutations within the
-same epitope. Within epitope competition results reduced fixation
-probabilities of nonsynonymous changes, whereas the synonymous changes behave 
-similarly in all cases. We assume that escape can happen at $n$ sites out of 3
-consecutive codons and vary $n$.
-The fitness landscape of each epitope includes negative epistatic terms, so that
-the joint presence of more than one escape mutation is not any more beneficial
-for the virus than a single mutation. Specifically, each site has two alleles,
-$\pm 1$, where $-1$ is the ancestral one and $+1$ the derived one; the fitness
-coefficient of a $k$-tuple of sites within the epitope is $f_k = (-1)^{k-1}
-2^{1-n}\eta_\epsilon $, where $\eta_\epsilon$ is the escape rate of the epitope
-drawn from an exponential distribution with mean $\epsilon$, 
-$n$ is the number of competing escapes in the epitope. 
-In this evolutionary scenario, many escape mutations start to sweep on different backgrounds within the viral population, but eventually
-compete and only one of them if fixed. The other parameters for the shown simulations are
-the following: deleterious effect $10^{-3}$, average escape rate $0.016$,
-fraction of deleterious synonymous mutations $0.986$, rate of new epitopes
-$0.0014$ per generation.
-}
-\label{fig:wec}
-\end{center}
-\end{figure}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2013-03-01 synmut.tex
little edits here and there, moved supplement body into separate file such that we can include it into the main text as an appendix for submission

@@
 \newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV-1 \env~evolution}
+\newcommand{\Affiliation}{Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany}
 \newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
@@
 \author{\Author}
+\affiliation{\Affiliation}
 \date{\today}
@@
 via cytotoxic CD8${}^+$ T-cells (CTLs) and neutralizing antibodies
-(nAbs)~\citep{rambaut_causes_2004} and facilitated by the high mutation rate of
-HIV-1 (HIV in the following)~\citep{mansky_lower_1995,abram_nature_2010}. When the host develops a CTL or
-nAb response against a particular HIV epitope, mutations in the viral genome that
+(nAbs)~\citep{rambaut_causes_2004} and facilitated by the high mutation rate
+~\citep{mansky_lower_1995,abram_nature_2010}. When the host develops a CTL or
+nAb response against a particular HIV-1 epitope, mutations in the viral genome that
 reduce or prevent recognition of the epitope frequently emerge. Escape mutations
@@
 rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
-infection, the most rapidly evolving part of the HIV genome are the variable
+infection, the most rapidly evolving part of the HIV-1 genome are the variable
 loops V1-V5 in the envelope protein gp120, which change to avoid recognition by
 nAbs. Escape mutations in \env, the gene encoding gp120, spread through the
-population within a few months (see \figurename~\ref{fig:aftnonsyn}).
+population within a few months.
 Consistent with this time scale, it is found that serum from a particular time
@@
 
-Escape mutations are selected for their effect on the amino acid sequence
-of the viral proteins. Conversely, synonymous mutations are commonly used as
+Escape mutations are selected because they change the amino acid sequence
+of viral proteins in a way that reduces antibody binding or epitope presentation. 
+Conversely, synonymous mutations are commonly used as
 approximately neutral markers in studies of viral evolution. Neutral markers are
@@
 maintaining protein function and avoiding the adaptive immune recognition,
-however, the HIV genome has to ensure efficient processing and translation,
+however, the HIV-1 genome has to ensure efficient processing and translation,
 nuclear export, and packaging into the viral capsid: all these processes operate
 at the RNA level and are sensitive to synonymous changes since these processes
-often depend on RNA folding. For example, the HIV \rev{} response element (RRE)
+often depend on RNA folding. For example, the HIV-1 \rev{} response element (RRE)
 in \env{} enhances nuclear export of full length or partially spliced viral
@@
 \citep{plotkin_synonymous_2011}. Recent studies have shown that genetically
-engineered HIV strains with altered codon usage can in some cases produce more
+engineered HIV-1 strains with altered codon usage can in some cases produce more
 viral protein, but in general replicate less efficiently
@@
 \citep{forsdyke_reciprocal_1995,snoeck_mapping_2011} and it has been shown that
-rates of evolution at synonymous sites vary along the HIV genome
+rates of evolution at synonymous sites vary along the HIV-1 genome
 \citep{mayrose_towards_2007}. Positive selection through the host adaptive
@@
 argue that synonymous mutations reach high frequencies via genetic hitchhiking
-due to limited recombination in HIV populations~\citep{neher_recombination_2010,
+due to limited recombination in HIV-1 populations~\citep{neher_recombination_2010,
 batorsky_estimate_2011}. We then compare our observations to computational
-models of HIV evolution and derive estimates for the effect of synonymous mutations
+models of HIV-1 evolution and derive estimates for the effect of synonymous mutations
  on fitness.  Extending the analysis of fixation probabilities to the
@@
 segregating at frequency $\nu$ has a probability $\pfix(\nu) = \nu$ to
-spread through the population and fix; in the rest of the cases, i.e. with
+spread through the population and fix; in the rest of the cases, i.e., with
 probability $1-\nu$, it goes extinct. As illustrated in the inset of
@@
 constrast, many nonsynonymous mutations fix (panel~\ref{fig:aftnonsyn}).
-Strictly speaking, no mutation in the HIV population ever fixes because the 
+Strictly speaking, no mutation in the HIV-1 population ever fixes because the 
 mutation rate and the population size are large. We define ``fixation'' or ``loss''
@@
 \begin{center}
-\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_p10.pdf}
+\subfloat{\includegraphics[width=\linewidth]
+{Shankarappa_allele_freqs_trajectories_syn_p10.pdf}
 \label{fig:aftsyn}}\\
- \subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p10.pdf}
+ \subfloat{\includegraphics[width=\linewidth]
+{Shankarappa_allele_freqs_trajectories_nonsyn_p10.pdf}
 \label{fig:aftnonsyn}}
 \caption{Time series of frequencies
-of synonymous (A) and nonsynonymous (B) derived alleles in \env, \shankaregion, from patient
-10~\cite{shankarappa_consistent_1999}.
+of synonymous (A) and nonsynonymous (B) derived alleles in \env, 
+\shankaregion, from patient 10~\cite{shankarappa_consistent_1999}.
 While many nonsynonymous mutations  fix, few synonymous
@@
 (blue to red). Inset: the fixation probability $\pfix$ of a neutral
-mutation is simply the likelihood that the future common ancestor is currently carrying
-it, i.e., its frequency $\nu$.}
+mutation is simply the likelihood that the future common ancestor at this 
+position is currently carrying it, i.e., the mutation frequency $\nu$.}
 \label{fig:aft}
@@
 deleterious, while outside this region most polymorphisms are roughly
-neutral. Note that this does not imply that all synonymous mutations in outside 
+neutral. Note that this does not imply that all synonymous mutations outside 
 \shankaregion{} are neutral -- only those mutations observed at high frequencies, which
-have been experiencing selection for some time already, tend to be neutral.
+have experienced selection for some time, tend to be neutral.
 
@@
 
-The propensity of nucleotides in the HIV genome to form base pairs has been
+The propensity of nucleotides in the HIV-1 genome to form base pairs has been
 measured using the SHAPE assay, a biochemical reaction preferentially altering
-unpaired bases (the HIV genome is a single stranded RNA)
+unpaired bases (the HIV-1 genome is a single stranded RNA)
 \citep{watts_architecture_2009}. The SHAPE assay has shown that the variable
@@
 explanations for a fitness cost of some synonymous mutations, in particular codon
-usage bias (CUB). HIV is known to prefer A-rich codons over highly expressed
+usage bias (CUB). HIV-1 is known to prefer A-rich codons over highly expressed
 human codons~\citep{jenkins_extent_2003,kuyl_biased_2012}. We do not find,
 however, any evidence for a contribution of average CUB to the ultimate fate of
-synonymous alleles; consistently, HIV does not seem to adapt its codon usage to
+synonymous alleles; consistently, HIV-1 does not seem to adapt its codon usage to
 its human host cells at the macroevolutionary level \citep{kuyl_biased_2012}.
@@
 frequency and that the fixation probability is reduced only in parts of the
-genome. These regions, however, undergo frequent adaptive changes to evade
+genome. This region, \shankaregion{} undergoes frequent adaptive changes to evade
 recognition by neutralizing antibodies \cite{williamson_adaptation_2003,
-richman_rapid_2003}. Due to the limited amount of recombination in HIV
+richman_rapid_2003}. Due to the limited amount of recombination in HIV-1
 \cite{neher_recombination_2010, batorsky_estimate_2011}, deleterious mutations
@@
 The approximate magnitude of the deleterious effects can be estimated from
-\FIG{fixp1}, which shows the distribution of times for synonymous
-alleles to reach the fix or get lost starting from intermediate frequencies. The
+\FIG{fixp1}, which shows the distribution of times after which synonymous
+alleles at intermediate frequencies fixed or were lost. The
 typical time to loss is of the order of 500 days. If this loss is driven by the
-deleterious effect of the mutation, this corresponds to deleterious effects of
-roughly $s_d \sim - 0.002$ per day.
+deleterious effect of the mutation, this corresponds to deleterious effects $s_d$ of
+the order of $- 0.002$ per day.
 
@@
 The depression in $\pfix$ depends on the deleterious effect size 
-of the synonymous alleles. This parameter also reduces synonymous
+of  synonymous alleles. This parameter also reduces synonymous
 diversity, measured by the probability of a derived allele to be found at
@@
 diversity, we run 2400 simulations with random parameters for deleterious effect
-size, fraction of deleterious synonymous sites, average size of escape
-mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
+size, fraction of deleterious synonymous sites, average escape rate $\epsilon$
+(color, blue to red corresponds to $10^{-2.5}$ to $10^{-1.5}$ per day), and rate of
 introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
-generation). Mostly simulations with small deleterious effects and relatively
-high deleterious fractions of synonymous mutations reproduce synonymous the
-fixation probability and diversity observed in patients. Parameters are chosen
-from uniform prior distributions as indicated by the red box (see methods).}
+day). Only simulations that reproduce the synonymous diversity and fixation
+patterns observed in data are shown. These simulations demonstrate that
+deleterious effects are around $-0.002$ and a large fraction of the 
+synonymous mutations needs to be deleterious. As expected, larger
+$s_d$ require larger $\epsilon$. Parameters are chosen
+from prior distributions uniform in logspace as indicated by the red box (see methods).}
 \label{fig:simheat}
@@
 (\FIG{simfixpvar}, lower inset). If no fixation happens, the area will be
-$-0.5$; if every mutation fixes, the area will be $+0.5$. In HIV infected
+$-0.5$; if every mutation fixes, the area will be $+0.5$. In HIV-1 infected
 patients, we find $A_\text{syn} \approx -0.2$ for synonymous changes and
@@
 shown in \FIG{simfixpvar}, the fixation probability of synonymous alleles
-decreases from the neutral expectation ($A_\text{syn} \sim 0$) to zero
-($A_\text{syn} \sim -0.5$) as their fitness cost increases; the synonymous
+decreases from the neutral expectation ($A_\text{syn} \approx 0$) to zero
+($A_\text{syn} \approx -0.5$) as their fitness cost increases; the synonymous
 diversity plummets as well, as deleterious mutations are selected against.
@@
 little or no fitness advantage. Consistent with this argument, larger
-deleterious effects in \FIG{simheat} correspond to larger escapes rates. (iii)
+deleterious effects in \FIG{simsfig} correspond to larger escapes rates. (iii)
 mutations with a deleterious effect smaller than approximately $0.001$ behave
-neutrally consistent with the typical coalescent times observed in HIV.
+neutrally consistent with the typical coalescent times observed in HIV-1.
 
@@
 always fix once they reach high frequencies -- $A_{\mathrm{nonsyn}}$ is well
-above zero. This is incompatible with the blue line in \FIG{fixp}: in an HIV
+above zero. This is incompatible with the blue line in \FIG{fixp}: in an HIV-1
 infection, nonsynonymous mutations at high frequency often disappear again, even
@@
 packaging, etc. is most likely to eventually fix. The emergence of multiple
-sweeping nonsynonymous mutations in real HIV infections has been shown
+sweeping nonsynonymous mutations in real HIV-1 infections has been shown
 \citep{moore_limited_2009, bar_early_2012}. This scenario has been explicitly 
@@
 \figurename~S\withinepi. The two scenarios are not exclusive and possibly both
-important in HIV evolution.
+important in HIV-1 evolution.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-By analyzing the fate of mutations in longitudinal data of HIV \env{} evolution,
+By analyzing the fate of mutations in longitudinal data of HIV-1 \env{} evolution,
 we demonstrate selection against synonymous substitutions in the comparatively
 conserved regions C2-C5 of the \env{} gene. Comparison with biochemical studies
-of base pairing propensity in RNA genome of HIV indicates that these
+of base pairing propensity in RNA genome of HIV-1 indicates that these
 mutations are deleterious, at least in part, because they disrupt stems in RNA
@@
 spiritu} has been recently used in a longitudinal study of influenza
-evolution~\citep{strelkowa_clonal_2012}. The propagators suggested in that
-article, however, represent ratios between nonsynonymous and synonymous
+evolution~\citep{strelkowa_clonal_2012}. The central quantity used in this
+article, however, is a ratio between propagators of nonsynonymous and synonymous
 mutations. The latter is used as an approximately neutral control; this method
@@
 \cite{Hurst:2002p32608}. It has been pointed out, however, that the rate of
-evolution at synonymous sites varies considerable along the HIV genome
+evolution at synonymous sites varies considerable along the HIV-1 genome
 \citep{mayrose_towards_2007} and that this variation can confound estimates of
@@
 \citep{watts_architecture_2009, sanjuan_interplay_2011} and conserved RNA
-structures exist in different parts of the HIV genome. Our analysis is able to
+structures exist in different parts of the HIV-1 genome. Our analysis is able to
 quantify the fitness effect of RNA structure within single infections and
@@
 dynamics. The observed hitchhiking highlights the importance of linkage due to
-infrequent recombination for the evolution of HIV
+infrequent recombination for the evolution of HIV-1
 \citep{neher_recombination_2010, batorsky_estimate_2011,
@@
 \epsilon^{-1} \log \nu_0$ generations for escape mutation with escape rate
-$\epsilon$ to rise from an initially low frequency $\nu_0\sim \mu$ to frequency
+$\epsilon$ to rise from an initially low frequency $\nu_0\approx \mu$ to frequency
 one. This implies that a region of length $l = (\rho t_{sw})^{-1} = \epsilon /
@@
 While classical population genetics assumes that the dominant stochastic force
-is genetic drift, i.e. non-heritable fluctuations in offspring number, our
+is genetic drift, i.e., non-heritable fluctuations in offspring number, our
 results show that stochasticity due to linked selection is much more important.
@@
 \citet{gillespie_genetic_2000}. Genetic draft in facultatively sexual population
-such as HIV has been characterized in \citep{neher_genetic_2011}. Importantly,
+such as HIV-1 has been characterized in \citep{neher_genetic_2011}. Importantly,
 large population sizes are compatible with low diversity and fast coalescence
@@
 be weakly deleterious and the adaptive and deleterious parts conspire to yield a
-more neutral-like averaged fixation probability. We consider this possibility
-implausible since the amino acid sequence outside the variable loops is much
-more conserved that the synonymous positions, suggesting that the majority of
-the nonsynonymous mutations is much more deleterious.
+more neutral-like averaged fixation probability. While weakly deleterious 
+nonsynonymous mutation certainly exist and will contribute to a depression of the
+fixation probability, we have seen that a substantial depression requires that
+weakly deleterious non-synonymous polymorphisms at high frequency greatly 
+outnumber escape mutations. This seems unlikely, since non-synonymous diversity
+exceeds synonymous diversity despite the overall much greater constraint on
+the aminoacid sequence.
 
@@
 possibilities and find that both produce the desired effect. Furthermore, there
-is experimental evidence in support of both of these hypotheses. Serum from HIV
+is experimental evidence in support of both of these hypotheses. Serum from HIV-1
 infected individuals typically neutralizes the virus that dominated the
@@
 
-Our results emphasize the inadequacy of independent site models of HIV evolution
+Our results emphasize the inadequacy of independent site models of HIV-1 evolution
 and the common assumption that selection is time independent or additive. 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-For the estimate of times to fixation/extinction, polymorphisms were binned by
+For the estimates of time to fixation/extinction, polymorphisms were binned by
 frequency and the time to first reaching either fixation or extinction was
@@
 resulting curves. Mutations that reached high frequency but neither fixed nor
-got extinct were classified as ``floating'', with one exception: if they first
+were lost were classified as ``floating'', with one exception: if they first
 reached high frequencies within 3 years of the last time point, it was assumed
@@
 The SHAPE scores quantifying the degree of base pairing of individuals sites in
-the HIV genome were downloaded from the journal website
+the HIV-1 genome were downloaded from the journal website
 \citep{watts_architecture_2009}. Wherever possible, SHAPE reactivities were
@@
 \citep{mansky_lower_1995, abram_nature_2010} and simulated intrapatient
-evolution for 6000 days. For simplicity, all third position in a codon where
+evolution for 6000 days. For simplicity, third positions of every codon were
 deemed synonymous and assigned either a selection coefficient $0$ with
 probability $1-\alpha$ or a deleterious effect $s_d$ with probability $\alpha$.
-First and second positions have strongly deleterious fitness effects 0.02. At at
+First and second positions have strongly deleterious fitness effects 0.02. At 
 rate $k_A$, a random place in the genome is designated an epitope that can
 escape by one or several mutations with an exponentially distributed escape rate
-with mean $\epsilon$. Both full-length HIV genomes and \env{}-only simulations
+with mean $\epsilon$. Both full-length HIV-1 genomes and \env{}-only simulations
 were performed and yielded comparable results.
@@
 All analysis and computer simulation scripts, as well as the sequence alignments
-used, are available for download.
+used, are available for download at \url{http://git.tuebingen.mpg.de/synmut}.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-We are grateful for stimulating discussions with Jan Albert and Trevor Bedford.
-This work is supported by the ERC starting grant HIVEVO 260686.
+We would like to thank Jan Albert, Trevor Bedford and Pleuni Pennings for 
+stimulating discussions and critical reading of the manuscript.
+This work is supported by the ERC starting grant HIVEVO 260686 and 
+in part by the National Science Foundation under Grant No.~NSF PHY11-25915.
 
@@
 \bibliography{bib}
+\newpage
+\appendix
+\onecolumngrid
+\include{supplement_body}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2013-03-01 synmut.tex
supplementary figures correct numbering
Little changes here and there

@@
  on fitness.  Extending the analysis of fixation probabilities to the
-nonsynonymous mutations, we show that time dependent selection or strong
+nonsynonymous mutations, we show that time-dependent selection or strong
 competition of escape mutations inside the same epitope are necessary to explain
@@
 segregating at frequency $\nu$ has a probability $\pfix(\nu) = \nu$ to
-spread through the population and fix; in the rest of the cases, i.e., with
+spread through the population and fix; in the rest of the cases, i.e. with
 probability $1-\nu$, it goes extinct. As illustrated in the inset of
@@
 mutation is simply the likelihood that the future common ancestor at this 
-position is currently carrying it, i.e., the mutation frequency $\nu$.}
+position is currently carrying it, i.e. the mutation frequency $\nu$.}
 \label{fig:aft}
@@
 One possible explanation for lack of fixation of synonymous mutations in
-\shankaregion are secondary structures in the viral RNA, the disruption of which
+\shankaregion{} are secondary structures in the viral RNA, the disruption of which
 is deleterious to the virus \citep{forsdyke_reciprocal_1995,
@@
 variable regions form stems.  We aligned the within-patient sequence samples 
-to the reference NL4-3 strain used in ref.~\citep{watts_architecture_2009} and 
+to the reference NL4-3 strain used by \citet{watts_architecture_2009} and 
 thereby assigned SHAPE reactivities to most positions in the alignment. 
@@
 frequency and that the fixation probability is reduced only in parts of the
-genome. This region, \shankaregion{} undergoes frequent adaptive changes to evade
+genome (in \shankaregion{} but not in the rest of \env{}, compare the red
+triangle line versus the green square line in \FIG{fixp2}).
+The region \shankaregion{} undergoes frequent adaptive changes to evade
 recognition by neutralizing antibodies \cite{williamson_adaptation_2003,
@@
 \FIG{fixp1}, which shows the distribution of times after which synonymous
-alleles at intermediate frequencies fixed or were lost. The
-typical time to loss is of the order of 500 days. If this loss is driven by the
-deleterious effect of the mutation, this corresponds to deleterious effects $s_d$ of
-the order of $- 0.002$ per day.
+alleles at intermediate frequencies become fixed or lost. The typical time to
+loss is of the order of 500 days. If this loss is driven by the deleterious
+effect of the mutation, this corresponds to deleterious effects $s_d$ of the
+order of $- 0.002$ per day. (This is only an average estimate: every single
+mutation is expected to have a slightly different fitness effect.)
 
@@
 $s_d$ require larger $\epsilon$. Parameters are chosen
-from prior distributions uniform in logspace as indicated by the red box (see methods).}
+from prior distributions uniform in logspace as indicated by the red rectangle
+(see methods).}
 \label{fig:simheat}
@@
 $-0.5$; if every mutation fixes, the area will be $+0.5$. In HIV-1 infected
-patients, we find $A_\text{syn} \approx -0.2$ for synonymous changes and
-$A_\text{nonsyn} \approx 0$ for nonsynonymous changes. In the three simulations
-shown in \FIG{simfixpvar}, the fixation probability of synonymous alleles
-decreases from the neutral expectation ($A_\text{syn} \approx 0$) to zero
-($A_\text{syn} \approx -0.5$) as their fitness cost increases; the synonymous
-diversity plummets as well, as deleterious mutations are selected against.
+patients, we find $P_\text{interm} \approx 0.005$, $A_\text{syn} \approx -0.2$
+for synonymous changes and $A_\text{nonsyn} \approx 0$ for nonsynonymous
+changes. In the three simulations shown in \FIG{simfixpvar}, the fixation
+probability of synonymous alleles decreases from the neutral expectation
+($A_\text{syn} \approx 0$) to zero ($A_\text{syn} \approx -0.5$) as their
+fitness cost increases; the synonymous diversity plummets as well, as
+deleterious mutations are selected against.
 
@@
 little or no fitness advantage. Consistent with this argument, larger
-deleterious effects in \FIG{simsfig} correspond to larger escapes rates. (iii)
+deleterious effects in \FIG{simsfig} correspond to larger escapes rates; and (iii)
 mutations with a deleterious effect smaller than approximately $0.001$ behave
-neutrally consistent with the typical coalescent times observed in HIV-1.
+neutrally, consistently with the typical coalescent times observed in HIV-1.
 
@@
 always fix once they reach high frequencies -- $A_{\mathrm{nonsyn}}$ is well
-above zero. This is incompatible with the blue line in \FIG{fixp}: in an HIV-1
+above zero. This is incompatible with the blue line in \FIG{fixp2}: in an HIV-1
 infection, nonsynonymous mutations at high frequency often disappear again, even
@@
 we demonstrate selection against synonymous substitutions in the comparatively
-conserved regions C2-C5 of the \env{} gene. Comparison with biochemical studies
+conserved regions C2-C4 of the \env{} gene. Comparison with biochemical studies
 of base pairing propensity in RNA genome of HIV-1 indicates that these
@@
 spiritu} has been recently used in a longitudinal study of influenza
-evolution~\citep{strelkowa_clonal_2012}. The central quantity used in this
+evolution~\citep{strelkowa_clonal_2012}. The central quantity used in that
 article, however, is a ratio between propagators of nonsynonymous and synonymous
@@
 While classical population genetics assumes that the dominant stochastic force
-is genetic drift, i.e., non-heritable fluctuations in offspring number, our
+is genetic drift, i.e. non-heritable fluctuations in offspring number, our
 results show that stochasticity due to linked selection is much more important.
@@
 fixation probability, we have seen that a substantial depression requires that
-weakly deleterious non-synonymous polymorphisms at high frequency greatly 
-outnumber escape mutations. This seems unlikely, since non-synonymous diversity
-exceeds synonymous diversity despite the overall much greater constraint on
-the aminoacid sequence.
+weakly deleterious nonsynonymous polymorphisms at high frequency greatly 
+outnumber escape mutations. This seems unlikely, since nonsynonymous diversity
+exceeds synonymous diversity despite the overall much greater constraints on
+the amino acid sequence.
 
-Alternatively, the lack of fixation could be due to time dependent environment
+Alternatively, the lack of fixation could be due to time-dependent environment
 through an immune system that is catching up, or competition between mutations
 that mediate escape within the same epitope. We explore both of these
-possibilities and find that both produce the desired effect. Furthermore, there
+possibilities and find that both produce the desired effect in computer models. Furthermore, there
 is experimental evidence in support of both of these hypotheses. Serum from HIV-1
@@
 could be substantial underestimates. Furthermore, weak conservation and
-time dependent selection results in estimates of evolutionary 
+time-dependent selection results in estimates of evolutionary 
 rates that depend on the time interval of observation with lower rates across
-larger intervals. This could mean that deep nodes in phylogenies are older than 
+larger intervals. This implies that deep nodes in phylogenies might be older than 
 they appear.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-We would like to thank Jan Albert, Trevor Bedford and Pleuni Pennings for 
+We thank Jan Albert, Trevor Bedford and Pleuni Pennings for 
 stimulating discussions and critical reading of the manuscript.
@@
 \onecolumngrid
+\setcounter{figure}{0}
 \include{supplement_body}

### 2013-03-04 supplement_body.tex
supplementary figures correct numbering
Little changes here and there
fixed typos, thanks Taylor

@@
 %%% End:
+
+% use S1..S4 for figures numbers
+\makeatletter 
+\renewcommand{\thefigure}{S\@arabic\c@figure}
+\makeatother
 
@@
 Panel A) shows a PCA of all sequences from patient p1 (colors indicate time from
-seroconversion, from blue to red). Panel B) Allele frequency trajectories for nonsynonymous
+seroconversion, from blue to red). Panel B) shows allele frequency trajectories for nonsynonymous
 changes in the same patient. Here, the blue to red color map corresponds to the
-position of the allele in \env{} from 5' to 3'. Panels C and D) show analogous
+position of the allele in \env{} from 5' to 3'. Panels C) and D) show analogous
 plots for data from patient p7. Samples after day 1000 split into two clusters in the PCA and no mutations that arise after day 1000 fix, presumably because they are restricted
@@
 protein selection in a different reading frame.
-Note that all gap-rich columns of the MSA are stripped from this figure, hence genes such as {\it env} might appear shorter than they actually
+Note that all gap-rich columns of the MSA are stripped from this figure, so genes such as {\it env} might appear shorter than they actually
 are.
@@
 compares the fixation probability in the time independent model (na\"ive) to
-a model with time dependent selection that mimics the an evolving immune system.
-It has been found that virus is typically neutralized by serum from a few month
+a model with time dependent selection that mimics  an evolving immune system.
+It has been found that virus is typically neutralized by serum from a few months
 earlier~\citep{richman_rapid_2003} but not by contemporary serum. We model this
-evolving immune system by assuming that escaped variants loose their beneficial
+evolving immune system by assuming that escaped variants lose their beneficial
 effect with a rate proportional to the frequency of the escaped variant. 
@@
 while hitch-hiking of synonymous mutations is not affected. The precise
-shape of $\pfix(\nu)$ depends on the details of the $P_\text{recognized}(t)$ and 
+shape of $\pfix(\nu)$ depends on the details of the $P_\text{recognized}(t)$, and 
 we do not think that the high $\pfix(\nu)$ for $\nu<0.2$ is meaningful.
@@
 with one, three, or six mutually exclusive escape mutations within the
-same epitope. Within epitope competition results reduced fixation
+same epitope. Within epitope competition results in reduced fixation
 probabilities of nonsynonymous changes, whereas the synonymous changes behave 
@@
 2^{1-n}\eta_\epsilon $, where $\eta_\epsilon$ is the escape rate of the epitope
-drawn from an exponential distribution with mean $\epsilon$, 
+drawn from an exponential distribution with mean $\epsilon$ and 
 $n$ is the number of competing escapes in the epitope. 

### 2013-03-04 synmut.tex
fixed typos, thanks Taylor

@@
 \usepackage{pslatex}
-\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
+\usepackage[colorlinks,linkcolor=red,citecolor=blue]{hyperref}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Intrapatient HIV-1 evolution is dominated by selection on the protein level in the
-arms race with the adaptive immune system. When cytotoxic CD8${}^+$ T-cells or neutralizing antibodies
+arms race with the adaptive immune system. When cytotoxic CD8${}^+$ T-cells or 
+neutralizing antibodies
 target a new epitope, the virus often escapes via nonsynonymous mutations that
@@
 Computational modeling indicates that these synonymous mutations have a
-(Malthusian) selection coefficient of the order of $-0.002$, and that they are
+(Malthusian) selection coefficient of the order of $-0.002$ and that they are
 brought up to high frequency by hitchhiking on neighboring beneficial
@@
 rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
-infection, the most rapidly evolving part of the HIV-1 genome are the variable
+infection, the most rapidly evolving parts of the HIV-1 genome are the variable
 loops V1-V5 in the envelope protein gp120, which change to avoid recognition by
@@
 Consistent with this time scale, it is found that serum from a particular time
-typically neutralizes virus extracted more than 3-6 month earlier, but not contemporary
+typically neutralizes virus extracted more than 3-6 months earlier but not contemporary
 virus \citep{richman_rapid_2003}.
@@
 
-Even in absence of important RNA structures, synonymous codons do not evolve
+Even in the absence of important RNA structures, synonymous codons do not evolve
 completely neutrally. Some codons are favored over others in many species
@@
 \citep{ngumbela_quantitative_2008, li_codon-usage-based_2012,
-keating_rich_2009}. Codon deoptimization has been suggested as attenuation
+keating_rich_2009}. Codon deoptimization has been suggested as an attenuation
 strategy for polio and influenza~\citep{mueller_live_2010,coleman_virus_2008}.
 Purifying selection beyond the protein sequence is therefore expected
-\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011} and it has been shown that
+\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}, and it has been shown that
 rates of evolution at synonymous sites vary along the HIV-1 genome
@@
 beneficial mutations fix less or more often than neutral ones, respectively.
-\FIG{aft} shows the time course the frequencies of all synonymous and
-nonsynonymous mutations observed \env, \shankaregion, in patient
-p10~\citep{shankarappa_consistent_1999}, respectively. Despite many synonymous
+\FIG{aft} shows the time course of the frequencies of all synonymous and
+nonsynonymous mutations observed in \env, \shankaregion, in patient
+p10~\citep{shankarappa_consistent_1999}. Despite many synonymous
 mutations reaching high frequency, few fix (panel~\ref{fig:aftsyn}); in
@@
 Strictly speaking, no mutation in the HIV-1 population ever fixes because the 
-mutation rate and the population size are large. We define ``fixation'' or ``loss''
+mutation rate and the population size are large. Therefore, we define ``fixation'' or ``loss''
 by not observing the mutation in the sample.
@@
 analyzing data from seven patients from
-\citet{shankarappa_consistent_1999,liu_selection_2006} and three patients from
+\citet{shankarappa_consistent_1999} and \citep{liu_selection_2006} as well as three patients from
 \citet{bunnik_autologous_2008} (patients whose viral population was structured
 were excluded from the analysis; see methods and \figurename~S\PCApat).  The
-former data set from is restricted to the \shankaregion{} region of \env, while
-the data from \citet{bunnik_autologous_2008} covers the majority of \env.
+former data set is restricted to the \shankaregion{} region of \env, while
+the data from \citet{bunnik_autologous_2008} cover the majority of \env.
 Considering all mutations in a frequency interval $[\nu_0-\delta\nu, \nu_0+\delta\nu]$ at some time
@@
 fixation, $\pfix$, is shown as a function of the
-initial frequency $\nu_0$ in panel~\ref{fig:fixp2} (red line). Restricted to the
-region \shankaregion, we find that $\pfix$ of synonymous mutations is far below
-the neutral expectation.  Outside of \shankaregion, using data from
-\citet{bunnik_autologous_2008} only, no such reduction in $\pfix$ is found.
+initial frequency $\nu_0$ in panel~\ref{fig:fixp2} (red line). 
+We find that $\pfix$ of synonymous mutations is far below
+the neutral expectation in  \shankaregion.  Outside of \shankaregion, using data from
+\citet{bunnik_autologous_2008} only, we find no such reduction in $\pfix$.
 Restricted to the \shankaregion{} region, the sequence samples from
@@
 observations suggest that many of the synonymous polymorphisms at intermediate
-frequencies in the part of \env{} that includes the \shankaregion{} are
+frequencies in the part of \env{} that includes \shankaregion{} are
 deleterious, while outside this region most polymorphisms are roughly
@@
 alleles as a function of $\nu_0$. Polymorphisms within \shankaregion{} fix less
-often than expected for neutral mutations indicated by the diagonal line.
+often than expected for neutral mutations (indicated by the diagonal line).
 This suppression is not observed in other parts of \env~or for nonsynonymous mutations.
@@
 One possible explanation for lack of fixation of synonymous mutations in
-\shankaregion{} are secondary structures in the viral RNA, the disruption of which
+\shankaregion{} is secondary structures in the viral RNA, the disruption of which
 is deleterious to the virus \citep{forsdyke_reciprocal_1995,
@@
 We then calculated the distributions of SHAPE reactivities for synonymous 
-polymorphisms that fixed or where subsequently lost (only polymorphisms with 
+polymorphisms that fixed or were subsequently lost (only polymorphisms with 
 frequencies above 15\%).
 As shown in \FIG{SHAPEA}, the reactivities of fixed alleles (red
-histogram) are systematically larger than of alleles that are lost (blue)
+histogram) are systematically larger than those of alleles that are lost (blue)
 (Kolmogorov-Smirnov test on the cumulative distribution, $p\approx 0.002$). In
@@
 break stems in the conserved stretches between the variable loops, we consider
-mutations in variable loops and conserved parts separately. The biggest
+mutations in variable loops and conserved parts separately. The greatest
 depression in fixation probability is observed in the conserved stems, while the
@@
 Panel A) shows the distribution of SHAPE reactivities among sites at which synonymous 
-mutations fixed (red), at which mutations reached frequencies above 15\% but
+mutations fixed (red), sites at which mutations reached frequencies above 15\% but
 were susequently lost (blue), and sites at which no mutations were observed (green) 
@@
 frequency and that the fixation probability is reduced only in parts of the
-genome (in \shankaregion{} but not in the rest of \env{}, compare the red
+genome (in \shankaregion{} but not in the rest of \env{}; compare the red
 triangle line versus the green square line in \FIG{fixp2}).
@@
 
-\FIG{simfixpvar} shows simulations results for the fixation probability and the
+\FIG{simfixpvar} shows simulation results for the fixation probability and the
 synonymous diversity for different deleterious effects of synonymous mutations.
@@
 little or no fitness advantage. Consistent with this argument, larger
-deleterious effects in \FIG{simsfig} correspond to larger escapes rates; and (iii)
+deleterious effects in \FIG{simsfig} correspond to larger escape rates; and (iii)
 mutations with a deleterious effect smaller than approximately $0.001$ behave
-neutrally, consistently with the typical coalescent times observed in HIV-1.
+neutrally, consistent with the typical coalescent times observed in HIV-1.
 
-The above simulation show that hitchhiking can explain the observation of
+The above simulations show that hitchhiking can explain the observation of
 deleterious mutations that rarely fix. However, in a simple model where
@@
 \citep{moore_limited_2009, bar_early_2012}. This scenario has been explicitly 
-observed in evolution the evolution of resistance to 3TC, where the mutation 
+observed in  the evolution of resistance to 3TC, where the mutation 
 M184V is often preceeded by M184I \citep{hedskog_dynamics_2010}. Similarly, AZT 
@@
 extra benefit, but additional costs). Again, we find that the potential for
-hitchhiking is little affected by within epitope competition, but that the
+hitchhiking is little affected by within epitope competition but that the
 fixation probability of nonsynonymous polymorphisms is reduced. With roughly six
-mutations per epitope, the simulation data is compatible with observations; see
+mutations per epitope, the simulation data are compatible with observations; see
 \figurename~S\withinepi. The two scenarios are not exclusive and possibly both
@@
 \citep{watts_architecture_2009, sanjuan_interplay_2011} and conserved RNA
-structures exist in different parts of the HIV-1 genome. Our analysis is able to
+structures exist in different parts of the HIV-1 genome. Since there are
+of course many ways to build an RNA stem in a particular location, we do
+not necessarily expect a strong signal of conservation in cross-sectional data. 
+Our analysis, however, is able to
 quantify the fitness effect of RNA structure within single infections and
@@
 order of $\rho = 10^{-5}$ per base and day. It takes roughly $t_{sw} =
-\epsilon^{-1} \log \nu_0$ generations for escape mutation with escape rate
+\epsilon^{-1} \log \nu_0$ generations for an escape mutation with escape rate
 $\epsilon$ to rise from an initially low frequency $\nu_0\approx \mu$ to frequency
@@
 be unconditionally beneficial. Otherwise we would observe almost sure fixation
-of nonsynonymous mutations once they reach intermediate frequencies. Instead, we
+of a nonsynonymous mutation once they reach intermediate frequencies. Instead, we
 find that the fixation probability of nonsynonymous mutations is roughly given
@@
 Similar to synonymous mutations, the majority of nonsynonymous mutations could
-be weakly deleterious and the adaptive and deleterious parts conspire to yield a
+be weakly deleterious, and the adaptive and deleterious parts could conspire to yield a
 more neutral-like averaged fixation probability. While weakly deleterious 
-nonsynonymous mutation certainly exist and will contribute to a depression of the
+nonsynonymous mutations certainly exist and will contribute to a depression of the
 fixation probability, we have seen that a substantial depression requires that
@@
 infected individuals typically neutralizes the virus that dominated the
-population a few (3-6) month earlier \citep{richman_rapid_2003}. This suggests that
+population a few (3-6) months earlier \citep{richman_rapid_2003}. This suggests that
 escape mutations cease to be beneficial after a few months and might revert if
@@
 could be substantial underestimates. Furthermore, weak conservation and
-time-dependent selection results in estimates of evolutionary 
-rates that depend on the time interval of observation with lower rates across
+time-dependent selection result in estimates of evolutionary 
+rates that depend on the time interval of observation, with lower rates across
 larger intervals. This implies that deep nodes in phylogenies might be older than 
@@
 probability $1-\alpha$ or a deleterious effect $s_d$ with probability $\alpha$.
-First and second positions have strongly deleterious fitness effects 0.02. At 
-rate $k_A$, a random place in the genome is designated an epitope that can
+Mutations at the first and second positions were assigned strongly deleterious 
+fitness effects 0.02. At 
+rate $k_A$, a random locus in the genome is designated an epitope that can
 escape by one or several mutations with an exponentially distributed escape rate
@@
 between 0.75 and 1.0; the average deleterious effect $s_d$ was sampled such that
-its logarithm is uniformly distributed  between $10^{-4}$ and $10^{-2}$; the
-average escape rate $\epsilon$ of escape mutation such that its logarithm is
-uniform between $10^{-2.5}$ and $10^{-1.5}$; the rate $k_A$ of new antibody
-challenges such that its logarithm is uniform between $10^{-3}$ and $10^{-2}$
+its logarithm was uniformly distributed  between $10^{-4}$ and $10^{-2}$; the
+average escape rate $\epsilon$ of escape mutation was sampled such that its logarithm was
+uniform between $10^{-2.5}$ and $10^{-1.5}$ and the rate $k_A$ of new antibody
+challenges such that its logarithm was uniform between $10^{-3}$ and $10^{-2}$
 per generation. Populations were initialized with a homogenous founder
@@
 was included. Higher order terms compensated each other to make sure that not
-only double, but all k-mutants with $k \geq 1$ had the same fitness (see
+only double mutants, but all k-mutants with $k \geq 1$ had the same fitness (see
 supplementary materials). To model recognition of escape variants by the immune
-system that is catching up, the beneficial effect of an escape mutation was set
+system  catching up, the beneficial effect of an escape mutation was set
 to its previous cost of -0.02 with a probability per generation proportional to
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-We thank Jan Albert, Trevor Bedford and Pleuni Pennings for 
+We thank Jan Albert, Trevor Bedford and Pleuni Pennings and members of the lab for 
 stimulating discussions and critical reading of the manuscript.

### 2013-03-04 scripts/PCA_vs_time.py
public

@@ -1,0 +1,72 @@
+# vim: fdm=marker
+'''
+author:     Fabio Zanini
+date:       07/02/2012
+content:    Show the first PCs vs time.
+'''
+# Standard modules
+from operator import *
+import numpy as np
+import matplotlib.pyplot as ppl
+import matplotlib.cm as cm
+import scipy.linalg as linalg
+from mpl_toolkits.mplot3d import Axes3D
+
+# Custom modules
+import modules.parser_Shankarappa as pS
+from modules.alphabet import alpha
+
+
+
+# Functions
+def plot3D(patient, ax=None, show=False):
+    if not hasattr(patient,'U'):
+        raise AttributeError('Please perform PCA before plotting.')
+
+    # Time points
+    visits = np.array(map(attrgetter('visit'), patient.seqs))
+
+    # Plot different time points with different colors
+    n = len(patient.visit)
+    cols = cm.jet([int(255.0 * i / n) for i in xrange(n)])
+    
+    if ax == None:
+        fig = ppl.figure(figsize=(6,5))
+        ax = fig.add_subplot(1,1,1, projection='3d')
+
+    for i,v in enumerate(patient.visit):
+        ind = (visits == v)
+        
+        x = patient.U[ind,0]
+        y = patient.U[ind,1]
+        z = patient.U[ind,2]
+    
+        ax.scatter(x,y,z, color=cols[i], s=60)
+    
+    ax.set_xlabel('PC1', fontsize=18)
+    ax.set_ylabel('PC2', fontsize=18)
+    ax.set_zlabel('PC3', fontsize=18)
+    ax.set_title('Patient '+patient.name)
+
+    if show:
+        ppl.show()
+
+
+
+# Script
+if __name__ == '__main__':
+    
+    patients = pS.parse_sequences()
+
+    for p in patients[0:1]:
+        print p
+        p.filter_only_sequenced()
+    
+        # Perform PCA on all sequences
+        p.PCA()
+
+        # Plot
+        plot3D(p)
+
+    ppl.ion()
+    ppl.show()

### 2013-03-04 scripts/allele_frequency_trajectories.py

@@ -1,0 +1,90 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       01/03/2012
+content:    Trajectories of allele frequencies.
+'''
+# Standard modules
+from operator import *
+import numpy as np
+import matplotlib.pyplot as ppl
+import matplotlib.cm as cm
+
+
+# Custom modules
+import sys
+sys.path.insert(0, '.')
+import modules.parser_Shankarappa as pS
+from modules.helper import is_nonsyn_table
+from modules.alphabet import alpha
+
+
+
+# Script
+if __name__ == '__main__':
+
+    # Define the patients
+    patients = pS.parse_sequences(reference='HXB2')
+
+    for k,p in enumerate(patients):
+        if str(p) != 'p10':
+            continue
+        print p
+
+        p.filter_only_sequenced()
+
+        # Measure the allele frequencies (of all alleles at all positions)
+        paf = p.allele_frequencies
+        afs = np.asarray([paf(seqs=p.seqs_from_visit(v)) for v in p.visit])
+
+        # Reshape so that we get the site as first axis, the nucleotide as
+        # second, the time as third
+        afs = afs.swapaxes(0,2)
+
+        # Eliminate gaps (whole codons are excluded to keep translation possible)
+        # first and translate then!
+        # Note: this has been implemented at the root
+        is_nongap = (afs[:,4,:] == 0).all(axis=1)
+        is_V_region = p.only_V_regions()
+
+        # Divide into syn/nonsyn to spot hitchhiking
+        consensus = alpha[afs[:,:,0].argmax(axis=1)]
+        is_nonsyn = is_nonsyn_table(consensus)
+
+        # Plot trajectories (no gaps, no conserved)
+        figsyn = ppl.figure(figsize=[6.5, 4.3])
+        fignonsyn = ppl.figure(figsize=[6.5, 4.3])
+        axsyn = figsyn.add_subplot(1,1,1)
+        axnonsyn = fignonsyn.add_subplot(1,1,1)
+        x = p.mo_to_SC
+        n_cols = afs.shape[0]
+        cols = cm.jet([int(255.0 * i / n_cols) for i in xrange(n_cols)])
+
+        for i in xrange(n_cols):
+            # Exclude gaps and restrict to regions if requested
+            if is_nongap[i]:# and not is_V_region[i]:
+    
+                # Do not plot consensus allele
+                indmaj = afs[i,:,0].argmax()
[…]

### 2013-03-04 scripts/codon_entropy_synonymous_subtypeB.py

@@ -1,0 +1,160 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       07/12/12
+content:    Measure the level of conservation of synonymous mutations in subtype
+            B.
+'''
+# Modules
+import re
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib.patches import Rectangle
+
+# Custom modules
+import sys
+sys.path.insert(0, '.')
+import modules.alignment as alignment
+from modules.alphabet import alpha
+from modules.codon_usage import number_of_codons
+from modules.helper import translate
+from modules.parser_reference import seq_patterns, find_with_gaps
+
+
+
+# Globals
+genes = ['gag', 'pol', 'env', 'nef']
+reference = 'SHAPE'
+seq_patterns = seq_patterns[reference]
+
+
+
+# Functions
+def smoothen(x, f):
+    l = len(x)
+    x = x[:l - l % f]
+    x = x.reshape((len(x) / f, f)).mean(axis=1)
+    return x
+
+
+
+# Script
+if __name__ == '__main__':
+
+    # Store data
+    data = {}
+
+    # Make figure for all four genes
+    fig, axs = plt.subplots(2, 2)
+    axs = axs.flatten()
+
+    for k, gene in enumerate(genes):
+
+        print gene
+
+        # Select the axis
+        ax = axs[k]
+        plt.sca(ax)
+
+        # Load data for a certain gene
+        ali = alignment.MSA(gene=gene, reference=reference)
+        if gene in data:
+            msa = data[gene]
+        else:
+            msa = np.array(ali.MSA)
+            data[gene] = msa
+        afs = ali.allele_frequencies()
+    
+        # Filter only part of pol
+        if gene == 'pol':
+            msa = msa[:, :2718]
+            afs = afs[:, :2718]
+    
+        # Find consensus sequence
+        consind = afs.argmax(axis=0)
+        consensus = alpha[consind]
+    
+        # Some seqs are not viable and include frameshifts that mess up the
+        # translation, hence we restrict to positions for which gaps are a minority
+        is_gap = consensus == '-'
+        # Exclude full codons
+        tmp = np.unique(is_gap.nonzero()[0] / 3)
+        is_gap[np.concatenate([tmp * 3, tmp * 3 +1, tmp * 3 + 2])] = True
+    
+        # Exclude stop codons
[…]

### 2013-03-04 scripts/fixation_loss_secondary_syn.py

@@ -1,0 +1,338 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       04/07/12
+content:    Correlate the fixation/loss of an allele with the amount of
+            secondary structure and codon usage bias. Use three categories:
+                1. polymorphic above 15% but finally extinct
+                2. polymorphic above 15% and fixed
+                3. never observed.
+            All three refer to synonymous, derived alleles.
+
+            Of course, the 3rd category includes alleles that are deleterious
+            and other that were simply not hit by mutation + hitchhiking.
+'''
+# Standard modules
+from sys import stderr
+from operator import *
+import numpy as np
+import matplotlib.pyplot as ppl
+from scipy import stats
+
+# Custom modules
+import sys
+sys.path.insert(0, '.')
+import modules.parser_Shankarappa as pS
+import modules.parser_Bunnik2008 as pB
+import modules.parser_LANL as pL
+import modules.secondary as sec
+from modules.helper import translate
+from modules.alphabet import alpha
+
+
+# Globals
+# Verbosity level (0-3)
+VERBOSE = 1
+
+# Bad patients (selected manually in a conservative way looking e.g. at PCA)
+bad_patients = {'Bunnik': ['ACH19542', 'ACH19768'],
+                'Shankarappa': ['p4', 'p7', 'p8', 'p11']}
+
+# Liu LANL data set code
+Liu_code = 2432
+
+# Thresholding method to spot polymorphisms ('counts' or 'frequency' are
+# possible: only non-singletons are filtered by the first, only derived alleles
+# in a certain frequency window by the second -- same results).
+tmethod = 'counts'
+# Details of the thresholding methods
+thresholdcount = 2
+thresholds = [0.15, 0.8]
+
+# Level of conservativeness in defining 'synonymous' (one of 'normal', 'most',
+# 'least'). Too strict means not many data, too loose means mixing up classes.
+conservative_level = 'normal'
+
+# Region of the genome to analyze
+regions = ['V1', 'V2', 'V3', 'V4', 'V5']
+only_regions = 1    # 1: only V regions and flanks, 0: all, -1: only nonV
+flank = 100     # length in nucleotides of the flanks around V loops
+
+# Keep only alleles that are really found in the SHAPE array (i.e. mutation from
+# the NL4-3-like allele to a different allele).
+only_from_SHAPE = True
+
+
+
+# Script
+if __name__ == '__main__':
+
[…]

### 2013-03-04 scripts/fixation_loss_synonymous_times.py

@@ -1,0 +1,285 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       15/06/12
+content:    Calculate how many of the synonymous alleles at intermediate nu0
+            reach either boundary after a certain time.
+'''
+# Standard modules
+from sys import stderr
+import re
+from operator import *
+import numpy as np
+import matplotlib.pyplot as ppl
+import matplotlib.cm as cm
+import matplotlib.patches as patches
+from matplotlib import rcParams
+params = {'backend': 'ps',
+        'axes.labelsize': 20, 
+        'text.fontsize': 20,
+    'legend.fontsize': 20,
+    'xtick.labelsize': 18,
+    'ytick.labelsize': 18,
+    'text.usetex': True}
+    
+rcParams.update(params)
+# Custom modules
+import sys
+sys.path.insert(0, '.')
+import modules.parser_Shankarappa as pS
+from modules.helper import translate
+from modules.alphabet import alpha
+
+# Tables
+# Verbosity level (0-3)
+VERBOSE = 1
+
+# Bad patients (selected manually in a conservative way looking e.g. at PCA)
+bad_patients = ['p4', 'p7', 'p8', 'p11']
+
+# Initial frequency bins
+nu0ss = [[0.15, 0.25],
+         [0.25, 0.4],
+         [0.4, 0.6]]
+
+# Colors for the plot
+colors = ['b', 'g', 'r']
+
+# Maximal time to plot
+maxt = 40 # Months
+
+# Approximate months-to-days conversion
+mo_to_gen = 30.5
+
+
+
+# Script
+if __name__ == '__main__':
+
+    fixed = []
+    lost = []
+    fixlost = [lost, fixed]
+
+    # Counts and times (overall)
+    # The first list of each nu0s is lost, the second fixed, the third floating
+    counts_all = [[0, 0, 0] for nu0s in nu0ss]
+    # The first list of each nu0s is lost, the second fixed
+    times_all = [[[], []] for nu0s in nu0ss]
+    
+    # Define the patients
+    patients = pS.parse_sequences(exclude=bad_patients)
+    
+    # Aggregate information from all patients
+    for k, p in enumerate(patients):
+
+        if VERBOSE >= 1:
+            stderr.write(str(p)+'\n')
+
+        # Filter the time points to sequenced times
+        p.filter_only_sequenced()
+        n = len(p.visit)
+    
[…]

### 2013-03-04 scripts/fixation_probability_Vloops_flanking.py

@@ -1,0 +1,268 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       15/06/12
+content:    Calculate the fixation probability of synonymous derived alleles in
+            the V loops and flanking regions.
+'''
+# Standard modules
+from sys import stderr
+import re
+from operator import *
+import numpy as np
+import matplotlib.pyplot as ppl
+import matplotlib.cm as cm
+import matplotlib.patches as patches
+import Bio
+
+# Custom modules
+import sys
+sys.path.insert(0, '.')
+import modules.parser_Shankarappa as pS
+import modules.parser_Bunnik2008 as pB
+from modules.helper import translate
+from modules.alphabet import alpha
+
+
+
+# Globals
+# Verbosity level (0-3)
+VERBOSE = 1
+
+# Initial frequency bins
+nu0ss = [[0.05, 0.15],
+         [0.15, 0.25],
+         [0.25, 0.4],
+         [0.4, 0.6],
+         [0.6, 0.8]]
+
+# Bad patients (selected manually in a conservative way looking e.g. at PCA)
+bad_patients = {'Bunnik': ['ACH19542', 'ACH19768'],
+                'Shankarappa': ['p4', 'p7', 'p8', 'p11']}
+
+# Classes of mutations
+classes = ['loops',
+               'flank']
+
+# Colors for the plot
+colors = {'loops': 'red',
+          'flank': 'blue'}
+
+# Labels for the plot
+labels = {'loops': 'V3-V5 loops',
+          'flank': 'flanking stems'}
+
+# Distance from last time point to be considered likely to reach either boundary
+# before the sequencing stops (later sequences will probably stay floating).
+# Note: this is not used in the plot (only for testing).
+maxt = {'Shankarappa': 40,  # Months
+        'Bunnik': 1200}     # Days
+
+
+
+# Script
+if __name__ == '__main__':
+
+    # Define the patients (excluding problematic ones)
+    patientsB = pB.parse_sequences(reference='SHAPE', exclude=bad_patients['Bunnik'])
+    patientsS = pS.parse_sequences(reference='SHAPE', exclude=bad_patients['Shankarappa'])
+    patients = patientsB + patientsS
+
+    # Counts (overall)
+    # The first list of each nu0s is lost, the second fixed, the third floating
+    # Moreover, record the patient number they came from
+    counts = {x: np.zeros((len(nu0ss), 3, len(patients)), int) for x in classes}
+
[…]

### 2013-03-04 scripts/modules/__init__.py

@@ -1,0 +1,12 @@
+__all__ = ['alignment',
+           'alphabet',
+           'codon_usage',
+           'helper',
+           'secondary',
+           'parser_reference',
+           'parser_generic',
+           'parser_SHAPE',
+           'parser_Shankarappa',
+           'parser_Bunnik2008',
+           'parser_LANL',
+          ]

### 2013-03-04 scripts/modules/alignment.py

@@ -1,0 +1,84 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       12/11/12
+content:    Tools for managing sequences and alignments of general nature.
+'''
+# Modules
+import os
+import glob
+import numpy as np
+
+from Bio import SeqIO, AlignIO
+
+import alphabet as abc
+
+
+# Globals
+datadir = '../data/'
+genes = ['gag', 'pol', 'env', 'nef', 'vpu', 'vpr', 'vif', 'tat']
+
+
+
+# Classes
+class MSA(object):
+    '''Multiple sequence alignment with general properties'''
+    def __repr__(self):
+        s = 'MSA(kind="'+self.kind+'", gene="'+self.gene+'"'
+        if hasattr(self, 'reference'):
+            s += ', reference="'+self.reference+'"'
+        s += ')'
+        return s
+
+    def __init__(self, kind='subtypeB', reference=None, gene=None):
+
+        kinddir = (datadir + kind).rstrip('/') + '/'
+        self.kind = kind.rstrip('/')
+        if reference is not None:
+            kinddir = (kinddir + 'to_reference/')
+            self.reference = reference
+        if not os.path.isdir(kinddir):
+            raise IOError('Directory not found')
+
+        found = False
+        for fn in glob.glob(kinddir+'*.fasta'):
+            if gene in fn:
+                MSA = AlignIO.read(fn, 'fasta')
+                if reference is None:
+                    self.MSA = MSA
+                else:
+                    self.reference_seq = MSA[0]
+                    if reference not in self.reference_seq.id:
+                        raise ValueError('Reference name not found in first sequence of the MSA')
+                    self.MSA = MSA[1:]
+                self.gene = gene
+                found = True
+                break
+        if not found:
+            raise IOError('Alignment file not found')
+
+
+    def allele_frequencies(self, alpha=abc.alpha):
+        '''Get the allele frequencies for an alphabet
+        
+        Parameters:
+            - alpha: an alphabet
+        '''
+        alpha = list(alpha)
+        L = len(self.MSA[0])
+        afs = np.zeros((len(alpha), L))
+        mat = np.array(self.MSA)
+        for i, a in enumerate(alpha):
+            afs[i] = (mat == a).sum(axis=0)
+        afs /= afs.sum(axis=0)
+
+        return afs
+
+
+
+
+# Script
+if __name__ == '__main__':
+
+    msa = MSA('subtypeB', gene='nef')
+

### 2013-03-04 scripts/modules/alphabet.py

@@ -1,0 +1,15 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       29/10/12
+content:    Alphabet utilities
+'''
+# Modules
+import numpy as np
+
+
+
+# Globals
+alphal = ['A', 'C', 'G', 'T', '-']
+alpha = np.array(alphal)
+alphal_ng = ['A', 'C', 'G', 'T']

### 2013-03-04 scripts/modules/codon_usage.py

@@ -1,0 +1,145 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       16/10/12
+content:    Read codon usage tables
+'''
+# Modules
+import re
+
+from Bio.Seq import translate
+
+from parser_generic import alpha
+
+
+
+# Globals
+datadir = '../data/codon_usage/'
+alpha = list(alpha)
+
+
+
+# Classes
+class codon_table(object):
+
+    def __init__(self, organism='homo sapiens'):
+
+        self.organism = organism
+        self.datafile = datadir + re.sub(' ', '_', organism) +'.dat'
+        
+        table_within = {}
+        table_abs = {}
+        with open(self.datafile, 'r') as f:
+            for line in f:
+                if line[0] != '#':
+                    fields = line.rstrip('\n').split('\t')
+                    key = fields[0]
+                    aa = fields[1]
+                    freqwithin = float(fields[2])
+                    freqabs = 1e-3 * float(fields[3])
+
+                    table_within[key] = (aa, freqwithin)
+                    table_abs[key] = freqabs
+
+        self.within = table_within
+        self.absolute = table_abs
+
+
+
+# Functions
+def all_codons():
+    '''Generate a list with all codons'''
+    return ('AAA', 'AAC', 'AAG', 'AAT', 'ACA', 'ACC',
+            'ACG', 'ACT', 'AGA', 'AGC', 'AGG', 'AGT',
+            'ATA', 'ATC', 'ATG', 'ATT', 'CAA', 'CAC',
+            'CAG', 'CAT', 'CCA', 'CCC', 'CCG', 'CCT',
+            'CGA', 'CGC', 'CGG', 'CGT', 'CTA', 'CTC',
+            'CTG', 'CTT', 'GAA', 'GAC', 'GAG', 'GAT',
+            'GCA', 'GCC', 'GCG', 'GCT', 'GGA', 'GGC',
+            'GGG', 'GGT', 'GTA', 'GTC', 'GTG', 'GTT',
+            'TAA', 'TAC', 'TAG', 'TAT', 'TCA', 'TCC',
+            'TCG', 'TCT', 'TGA', 'TGC', 'TGG', 'TGT',
+            'TTA', 'TTC', 'TTG', 'TTT')
+
+
+def all_amino_acids():
+    '''All amino acids'''
+    return ('A', 'C', 'E', 'D', 'G', 'F',
[…]

### 2013-03-04 scripts/modules/helper.py

@@ -1,0 +1,115 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       29/03/12
+content:    Helper functions for all parsers
+'''
+# Standard modules
+import numpy as np
+import matplotlib.pyplot as ppl
+import re
+
+import Bio.Seq
+
+
+# Functions
+def calculate_distance(smat0, smat1):
+    dM = np.zeros((smat0.shape[0], smat1.shape[0]), int)
+    for i0, s0 in enumerate(smat0):
+        dM[i0] = (s0 != smat1).sum(axis=1)
+    return dM
+
+
+def plot_distance(dM, ax=None):
+    if ax is not None:
+        ppl.sca(ax)
+    ppl.imshow(dM, interpolation='nearest')
+    ppl.ylabel('First seqs')
+    ppl.xlabel('Second seqs')
+    ppl.title('Hamming Distance', fontsize=16)
+    
+    # set ticks to integers
+    dMmax = dM.max()
+    tickwidth = ((dMmax - 1) // 10) + 1
+    tickmax = (dMmax // tickwidth + 1) * (tickwidth)
+    ppl.colorbar(ticks=np.arange(tickmax))
+
+    ppl.ion()
+    ppl.show()
+
+
+def translate(seqdna):
+    '''Translate one or more sequences with gaps.'''
+    if np.rank(seqdna) > 1:
+        aa = []
+        for seq in seqdna:
+            aa.append(translate(seq))
+        return np.asarray(aa)
+    else:
+        tmp = seqdna.reshape((len(seqdna) / 3, 3))
+        aa = np.zeros(tmp.shape[0], str)
+        for i1 in xrange(aa.shape[0]):
+            if not (tmp[i1] == '-').any():
+                aa[i1] = Bio.Seq.translate(''.join(tmp[i1]))
+        return aa
+
+
+def is_syn_table(sequence, *args, **kwargs):
+    '''Find what loci would lead to synonymous mutations.'''
+    return -is_nonsyn_table(sequence, *args, **kwargs)
+
+
+def is_nonsyn_table(sequence, alpha=np.array(['A', 'C', 'G', 'T', '-'])):
+    '''Find what loci would lead to nonsynonymous mutations.'''
+    sequence = np.asarray(sequence)
+    codons = sequence.reshape(sequence.shape[0] / 3, 3)
+    aa = translate(sequence)
+
+    # All possible single mutants
+    is_nonsyn = np.ones((sequence.shape[0], 5), bool)
[…]

### 2013-03-04 scripts/modules/parser_Bunnik2008.py

@@ -1,0 +1,110 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       25/07/12
+content:    Parser for Bunnik2008 (subparser of LANL)
+'''
+# Standard modules
+from operator import *
+import numpy as np
+
+from Bio.Alphabet.IUPAC import *
+
+from parser_LANL import patient as patientLANL
+from parser_LANL import parse_sequences as parse_sequencesLANL
+from helper import translate, find_with_gaps
+from parser_reference import seq_patterns
+
+
+# Globals
+# Second rev exon in env (tat is included)
+rev2 = {'ACH19999': ['ACCCACCTCCCA', 'TGCTGTTAGCTTG'],
+        'ACH18969': ['ACCCGCCTCCCA', 'TGCTGTTAGCTTG'],
+        'ACH19659': ['ACCCGCCTCCCA', 'TGCTGTTAGCTTG']}
+
+
+
+
+# Classes
+class patient(patientLANL):
+    '''Specialized version of patient from parser_LANL.'''
+
+    aaS = ['EEE', 'RPGGG']
+    '''Shankarappa region'''
+
+    def filter_for_longitudinal_genetics(self):
+        '''Filter only time points with sequences.'''
+        self.filter_seqs_with_attribute('days_from_seroconversion')
+
+
+    def only_Shanka(self):
+        '''Denote the Shankarappa region.
+ 
+        Returns:
+            index: boolean index with the Shankarappa region marked by 'True's.
+
+        *Note*: The region is marked by 'EEE' at the beginning and 'RPGGG' at the end.
+        '''
+
+        cons = self.get_consensus_initial()
+        is_Shanka = np.zeros(len(cons), bool)
+        consaa = ''.join(translate(cons))
+        is_Shanka[consaa.find(self.aaS[0]) * 3: consaa.find(self.aaS[1]) * 3] = True
+        return is_Shanka
+
+
+    def only_rev2(self):
+        '''Denote the second rev exon.
+
+        Returns:
+            index: boolean index with the Shankarappa region marked by 'True's.
+        '''
+        cons = ''.join(self.get_consensus_initial())
+        is_rev2 = np.zeros(len(cons), bool)
+        is_rev2[cons.find(rev2[self.name][0]): cons.find(rev2[self.name][1])] = True
+        return is_rev2
+
+
+    def find_V_region(self, name):
+        '''Find a variable region based on the reference'''
[…]

### 2013-03-04 scripts/modules/parser_LANL.py

@@ -1,0 +1,782 @@
+# vim: fdm=marker
+'''
+author:     Fabio Zanini
+date:       10/02/2012
+content:    Parser for the various files downloaded from the LANL HIV database.
+'''
+# Standard modules
+import os
+import re
+import glob
+from operator import *
+import numpy as np
+import scipy.linalg as linalg
+import matplotlib.pyplot as ppl
+import matplotlib.cm as cm
+
+from Bio import SeqIO, AlignIO
+from Bio.Seq import Seq
+from Bio.SeqRecord import SeqRecord
+from Bio.Alphabet.IUPAC import *
+
+import parser_SHAPE as pSHAPE
+from parser_generic import patient_generic, alpha
+
+
+
+# Globals
+datadir = '../data/longitudinal/LANL'
+
+columns = [('#',int),
+           ('patient_code',str),
+           ('patient_id',str),
+           ('accession',str),
+           ('name',str),
+           ('subtype',str),
+           ('country',str),
+           ('sampling_year',int),
+           ('days_from_first_sample',int),
+           ('Fiebig_stage',str),
+           ('days_from_treatment_end',int),
+           ('n_timepoints',int),
+           ('days_from_treatment_start',int),
+           ('days_from_infection',int),
+           ('days_from_seroconversion',int),
+           ('pubmed_ID',str),
+           ('start',int),
+           ('stop',int),
+           ('sequence_length',int),
+           ('organism',str),
+           ('sequence',str)]
+
+
+# Classes
+class patient(patient_generic):
+    def __init__(self,name):
+        self.name = name
+        self.visit = []
+        self.seqs = []
+
+    def __call__(self):
+        for a in vars(self).keys():
+                if a != 'seqs':
+                    print a
+                    print getattr(self,a)
+                elif self.seqs:
+                    print a
+                    print self.seqs[0]
+                    print '[...]'
+                print
+
+
+    def visit_to_time(self,visitnumber):
+        attr = filter(lambda x: 'days' in x, vars(self).keys())[0]
+        return getattr(self, attr)[self.visit == visitnumber][0]
+
+
+    def time_to_visit(self,days):
+        attr = filter(lambda x: 'days' in x, vars(self).keys())[0]
+        return self.visit[getattr(self, attr) == days][0]
+
+
+    def seqs_from_visit(self,visitnumbers):
+        if np.isscalar(visitnumbers):
[…]

### 2013-03-04 scripts/modules/parser_SHAPE.py

@@ -1,0 +1,205 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       03/07/12
+content:    Parser for the SHAPE reactivity data, from doi:10.1038/nature08237.
+'''
+# Modules
+import numpy as np
+import re
+
+
+# Globals
+helixfile = '../data/SHAPE/nature08237-s2.txt'
+bpseqfile = '../data/SHAPE/nature08237-s2_bpseq.txt'
+ppfile = '../data/SHAPE/nature08237-s3.txt'
+
+
+
+# Classes
+class secondary(object):
+    '''Store the SHAPE reactivities and pairing probs.'''
+
+    Shankarappa_indices = [6558, 7173]
+    
+
+    def __init__(self):
+        '''Store sequence, reactivities, pairing probs, and map.'''
+        self.r, self.p = np.genfromtxt(ppfile,
+                                       usecols=(2,3),
+                                       delimiter='\t',
+                                       unpack=True,
+                                       usemask=True)
+    
+        self.sequence = np.genfromtxt(ppfile,
+                                      dtype='S1',
+                                      usecols=[1],
+                                      delimiter='\t',
+                                      unpack=True,
+                                      usemask=False)
+
+        # Convert to DNA notation
+        self.sequence[self.sequence == 'U'] = 'T'
+
+        # Save map in a masked array
+        bp1, bp2 = np.loadtxt(bpseqfile, unpack=True, usecols=(0,2), dtype=int)
+        self.map = np.ma.masked_all(len(self.sequence), int)
+        self.map[bp1 - 1] = bp2 - 1
+
+
+    def __repr__(self):
+        return 'secondary structure class of HIV-1 NL4-3'
+
+
+
+# Functions
+def assign_prob(seq, include_reactivity=False):
+    '''Create a masked array with pairing probability and reactivity.
+    
+    Gap positions in the input sequence will stay masked.
+    Sites where the reactivity has not been measured will stay masked as well.
+    '''
+    # Create secondary structure object
+    sec = secondary()
+    s = ''.join(sec.sequence)
+
+    # Initialize
+    assp = np.ma.masked_all(len(seq))
+    assr = np.ma.masked_all(len(seq))
+
+    # Short names
+    p = sec.p
+    r = sec.r
+
+    # Check input string
[…]

### 2013-03-04 scripts/modules/parser_Shankarappa.py

@@ -1,0 +1,661 @@
+# vim: fdm=marker:fileencoding=utf-8
+'''
+author:     Fabio Zanini
+date:       04/01/2012
+content:    Parser for various Shankarappa data files
+'''
+# Standard modules
+import re
+import os
+import glob
+import shutil
+from operator import *
+import numpy as np
+import scipy.linalg as linalg
+import matplotlib.pyplot as ppl
+import matplotlib.cm as cm
+
+from Bio import SeqIO, AlignIO
+from Bio.Seq import Seq
+from Bio.Alphabet.IUPAC import *
+from Bio.Align import MultipleSeqAlignment as MSA
+
+import parser_SHAPE as pSHAPE
+from helper import find_with_gaps
+from parser_reference import seq_patterns
+from parser_generic import patient_generic, alpha
+
+
+
+# Globals
+datadir = '../data/longitudinal/Shankarappa'
+
+
+
+# Classes
+class clinical_parser(object):
+    '''Parse the TXT file with the clinical info.'''
+    def __init__(self, filename):
+        cols = [['visit', 1, int],
+                ['mo_to_SC', 2, float],
+                ['logRNA', 4, float],
+                ['CD3', 5, float],
+                ['CD8', 6, float],
+                ['CD4', 7, float]]
+
+        with open(filename, 'r') as f:
+            f.readline()    # Header
+            lsts = [l.split('\t') for l in f.readlines()]
+
+        names = set(map(itemgetter(0),lsts)) - set([''])
+        patients = [patient(p) for p in names]
+
+        for p in patients:
+            tmpls = filter(lambda x: x[0] == p.name,lsts)
+            n_points = len(tmpls)
+            tmpdict = {}
+            for n in cols:
+                tmpdict[n[0]] = np.ma.zeros(n_points, dtype=n[2])
+
+            for i,tmp in enumerate(tmpls):
+                for n in cols:
+                    try:
+                        tmpdict[n[0]][i] = tmp[n[1]]
+                    except ValueError:
+                        tmpdict[n[0]][i] = np.ma.masked
+                    
+            for n,c in tmpdict.items():
+                setattr(p,n,c)
+
+            # Try to add days_from_seroconversion
+            p.days_from_seroconversion = p.mo_to_SC * 30.5
+
+        patients.sort(key=(lambda x: int(x.name[1:])))
+        self.patients = patients
+
+
+    def plot(self,patients=[]):
+        '''Plot the clinical data as a function of time.'''
+
[…]

### 2013-03-04 scripts/modules/parser_reference.py

@@ -1,0 +1,110 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       04/09/12
+content:    Implement some globals used by many parsers that interact with a
+            reference.
+'''
+# Modules
+import numpy as np
+from itertools import permutations
+
+import Bio.AlignIO as AlignIO
+
+from helper import find_with_gaps
+from parser_generic import alpha
+
+
+
+# Globals
+# these are derived from each other using MSA.corresponding_pattern
+# (and ultimately by hand from the horrible XLS file for HXB2)
+seq_patterns = {'SHAPE':{'V1': ['TGCACTGATTTGAA', 'TGCTCTTTCAATATCAGC'],
+                         'V2': ['TGCTCTTTCAATATCAGC', 'TGTAACACCTCAGTCAT'],
+                         'V3': ['TGTACAAGACCCAACAACA', 'TGTAACATTAGTAGA'],
+                         'V4': ['TGTAATTCAACACAACTGTTTAA', 'TGCAGAATAAAACAATT'],
+                         'V5': ['TAATAACAACAATGGGTCCGAGAT', 'CCTGGAGGA'],
+                         'env_signal': ['ATGAGAGTGAAGGAGAA', 'ACAGAAAAATTGTGGGTC']},
+                'NL4-3':{'V1': ['TGCACTGATTTGAA', 'TGCTCTTTCAATATCAGC'],
+                         'V2': ['TGCTCTTTCAATATCAGC', 'TGTAACACCTCAGTCAT'],
+                         'V3': ['TGTACAAGACCCAACAACA', 'TGTAACATTAGTAGA'],
+                         'V4': ['TGTAATTCAACACAACTGTTTAA', 'TGCAGAATAAAACAATT'],
+                         'V5': ['TAATAACAACAATGGGTCCGAGAT', 'CCTGGAGGA']},
+                'HXB2': {'V1': ['TGCACTGATTTGAA', 'TGCTCTTTCAATATCAGC'],
+                         'V2': ['TGCTCTTTCAATATCAGC', 'TGTAACACCTCAGTCAT'],
+                         'V3': ['TGTACAAGACCCAACAACA', 'TGTAACATTAGTAGA'],
+                         'V4': ['TGTAATTCAACACAACTGTTTAA', 'TGCAGAATAAAACAAAT'],
[…]

### 2013-03-04 scripts/modules/secondary.py

@@ -1,0 +1,98 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       01/10/12
+content:    Module for parsing and handling RNA secondary structures
+'''
+# Modules
+import os
+import numpy as np
+
+
+
+# Classes
+class entry(object):
+    '''RNA secondary structure prediction entry'''
+    
+    def __init__(self, name, seq, contact_map=None, vienna=None):
+        self.name = name
+        self.seq = seq
+        if contact_map is not None:
+            self.contact_map = contact_map
+            self.vienna = contact_map_to_vienna(contact_map, length=len(seq))
+        elif vienna is not None:
+            self.vienna = vienna
+            self.contact_map = vienna_to_contact_map(vienna)
+        else:
+            raise ValueError('Please input a Vienna format or a contact map!')
+
+
+    def adapt_map(self, sequence):
+        '''Adapt the prediction to a gapped sequence'''
+        vv = []
+        ii = 0
+        for s in sequence:
+            if s == '-':
+                vv.append('.')
+            else:
+                vv.append(self.vienna[ii])
+                ii += 1
+        return vienna_to_contact_map(''.join(vv))
+
+
+
+# Functions
+def vienna_to_contact_map(vienna):
+    '''Convert a Vienna format to contact map'''
+    contact_map = {}
+    tmp = []
+    for j, c in enumerate(vienna):
+        if c == '(':
+            tmp.append(j)
+        elif c == ')':
+            p1 = tmp.pop(-1)
+            p2 = j
+            contact_map[p1] = p2
+            contact_map[p2] = p1
+    return contact_map
+
+
+def contact_map_to_vienna(contact_map, length):
+    '''Convert a contact map into Vienna format'''
+
+    if max(contact_map.keys()) >= length:
+        raise ValueError('Length is smaller than maximal contact')
+
+    vv = np.repeat('.', length)
+    for (p1, p2) in contact_map.items():
+        if p1 < p2:
+            vv[p1] = '('
+        else:
+            vv[p1] = ')'
+
+    return ''.join(vv)
+
+
+def read_Vienna(filename):
+    '''Parse a list of sequence-structures in Vienna format'''
+
+    entries = []
+
+    # Check file existance
+    with open(filename, 'r') as f:
+        for i, line in enumerate(f):
+
+            # Read name
+            if i % 3 == 0:
+                name = line.lstrip('> ')[:-1]
+
+            # Read sequence
+            elif i % 3 == 1:
[…]

### 2013-03-04 scripts/simulations_evolution.py

@@ -1,0 +1,311 @@
+#!/ebio/ag-neher/share/programs/EPD/bin/python
+# vim: fdm=marker
+'''
+author:     Fabio Zanini
+date:       07/08/12
+content:    Study fixation and loss of synonymous/nonsynonymous changes by a
+            short-genome model (1000 sites). This script is the one eventually
+            used to produce the simulations for the paper.
+'''
+# Standard modules
+import os
+import sys
+from operator import *
+import numpy as np
+from itertools import combinations
+
+# Env vars
+JOBSSCRIPT = os.path.realpath(__file__)      # This file
+JOBDIR = os.path.dirname(JOBSSCRIPT)+'/'
+JOBMODULES = JOBDIR+'modules'    
+if JOBMODULES not in sys.path: sys.path.insert(0,JOBMODULES)
+
+# Custom modules (import local FFPopSim)
+import FFPopSim as h
+
+
+
+# Globals
+VERBOSE = 0
+L = 1000
+mu = 2e-5
+crossover = 1e-3
+epitope_length = 9
+
+# Simulation length
+dt = 10
+n_time_points = 600
+dt_sample = 20
+
+
+
+# Classes
+class haploid_highd_gradual(h.haploid_highd):
+    '''Population with dynamic fitness landscape'''
+    
+
+    def __init__(self, *args, **kwargs):
+        if 'nescepi' not in kwargs:
+            raise ValueError('Please add the "nescepi" argument.')
+        newkwargs = {key: value for key, value in kwargs.iteritems() if key != 'nescepi'}
+        super(haploid_highd_gradual, self).__init__(*args, **newkwargs)
+        self.nescepi = kwargs['nescepi']
+        self.epitopes = []
+        self.epitope_times = []
+        self.epitope_fitness = []
+
+
+    def add_epitope(self, value):
+
+        # Complex epitope creation
+        if self.nescepi > 1:
+            # Choose a random good position
+            pos = np.random.randint(self.L - epitope_length)
+            epitope = [p for p in range(pos, pos + epitope_length) if (p + 1) % 3]
+            np.random.shuffle(epitope)
+            epitope = epitope[:self.nescepi]
+
+            add, epis = create_epitope(epitope, value)
+            fitness = self.get_trait_additive()
+            for i, v in add.iteritems():
+                fitness[i] = v
+            self.set_trait_additive(fitness)
+            for loci, v in epis.iteritems():
+                self.add_trait_coefficient(v, loci)
+
+            # Store changes
[…]

### 2013-03-04 scripts/simulations_plot_gradually_epitopesandtimeselec.py

@@ -1,0 +1,105 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       22/01/13
+content:    Plot the fixation probability of increasingly deleterious sims and 
+            allele freqs trajectories of some of its simulations.
+'''
+# Modules
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib.patches import Rectangle
+import matplotlib.cm as cm
+
+
+
+# Globals
+results_dir = '../data/simulations/gradually_epitopesandtimeselec/'
+datafile = 'parameters_fixation_areas.dat'
+
+
+
+# Functions
+def filterfolders(folders):
+    '''Filter and sort folders'''
+    tmp = folders[0].split('_')
+    if ('nescepi' in tmp) or ('timeselec' in tmp):
+        tmp = tmp[:-2]
+    base = '_'.join(tmp)
+
+    ind = {}
+    for i, f in enumerate(folders):
+        if 'nescepi_3' in f:
+            ind[1] = i
+        elif 'nescepi_6' in f:
+            ind[2] = i
+        elif 'timeselec_0.002_deadagain' in f:
+            ind[3] = i
+        elif 'timeselec_0.005_deadagain' in f:
+            ind[4] = i
+        elif ('nescepi' not in f) and ('timeselec' not in f):
+            ind[0] = i
+    return [folders[ind[i]] for i in xrange(5)]
+
+
+
+# Script
+if __name__ == '__main__':
+
+    # Folders
+    g = os.walk(results_dir)
+    folders = np.array(g.next()[1])
+
+    # Sort them smartly
+    folders = filterfolders(folders)
+    labels = ['naive', '3 mut/epi', '6 mut/epi',
+              'c = 0.002 / gen', 'c = 0.005 / gen']
+    colors = ['b', 'g', 'r', 'cyan', 'magenta']
+
+    # Make two figures, one for tds, one for competing escapes
+    for k, ind in enumerate([[0, 1, 2], [0, 3, 4]]):
+
+        # Prepare figure
+        fig = plt.figure()
+    
+        # Get the data
+        for i in ind:
+    
+            folder = folders[i]
+            fn = results_dir+folder
+            nu0, nu1, Psyn, Pnonsyn, csyn, cnonsyn = np.genfromtxt(fn+'/Pfix.dat',
+                                                                   unpack=True,
+                                                                   usemask=True,
[…]

### 2013-03-04 scripts/simulations_plot_graduallydel.py

@@ -1,0 +1,119 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       22/01/13
+content:    Plot the fixation probability of increasingly deleterious sims and 
+            allele freqs trajectories of some of its simulations.
+'''
+# Modules
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib.patches import Rectangle
+import matplotlib.cm as cm
+
+
+
+# Globals
+results_dir = '../data/simulations/gradually_more_del_6epitopes/'
+datafile = 'parameters_fixation_areas.dat'
+
+
+
+# Script
+if __name__ == '__main__':
+
+    # Load data using record arrays
+    tb = np.genfromtxt(results_dir+datafile, unpack=False,
+                       dtype=zip(('del_eff', 'ada_eff',
+                                  'del_frac', 'ada_rate',
+                                  'trashed', 'good',
+                                  'Asyn', 'Anonsyn',
+                                  'count_syn', 'counts_nonsyn',
+                                  'subs_syn', 'subs_nonsyn',
+                                  'poly_syn', 'poly_nonsyn',
+                                  'foldername'),
+                                 (float, float,
+                                  float, float,
+                                  int, int,
+                                  float, float,
+                                  float, float,
+                                  float, float,
+                                  float, float,
+                                  'S200')))
+
+    # Filter out NaNs
+    is_nonnan = -np.isnan(tb['Asyn'])
+    tb = tb[is_nonnan]
+    ind = np.argsort(tb['del_eff'])
+
+    # Filter only some parameters
+    ind = [ind[1], ind[2], ind[4]]
+
+    # Limits of deleff
+    colors = ['b', 'g', 'r', 'cyan', 'magenta']
+    labels = [r'$10^{-3.5}$', r'$10^{-3}$', r'$10^{-2.5}$']
+
+    # Prepare figure
+    fig = plt.figure()
+
+    # Get the data
+    for ii, i in enumerate(ind):
+
+        folder = tb['foldername'][i]
+        fn = results_dir+folder
+        nu0, nu1, Psyn, Pnonsyn, csyn, cnonsyn = np.genfromtxt(fn+'/Pfix.dat',
+                                                               unpack=True,
[…]

### 2013-03-04 scripts/simulations_scatter_all.py

@@ -1,0 +1,107 @@
+# vim: fdm=marker
+'''
+author:     Fabio Zanini
+date:       14/01/13
+content:    Scatter plot and further visualization of the results of random
+            gradual simulations.
+'''
+# Modules
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib.patches import Rectangle
+import matplotlib.cm as cm
+
+
+
+# Globals
+results_dir = '../data/simulations/random_gradual_nonsyndead/'
+datafile = 'parameters_fixation_areas.dat'
+
+# Set limits for data-like sims
+Asynmin = -0.5
+Asynmax = -0.15
+poly_synmin = 0.0025
+poly_synmax = 0.010
+
+# (nonsyn gates, if only nonsyn data-like results are wished)
+Anonsynmax = 0.5
+Anonsynmin = -0.5
+tmpsnsmin = 0e-6
+
+
+
+# Script
+if __name__ == '__main__':
+
+    # Load data using record arrays
+    tb = np.genfromtxt(results_dir+datafile, unpack=False,
+                       dtype=zip(('del_eff', 'ada_eff',
+                                  'del_frac', 'ada_rate',
+                                  'trashed', 'good',
+                                  'Asyn', 'Anonsyn',
+                                  'count_syn', 'counts_nonsyn',
+                                  'subs_syn', 'subs_nonsyn',
+                                  'poly_syn', 'poly_nonsyn',
+                                  'foldername'),
+                                 (float, float,
+                                  float, float,
+                                  int, int,
+                                  float, float,
+                                  float, float,
+                                  float, float,
+                                  float, float,
+                                  'S200')))
+
+    # Filter out NaNs
+    is_nonnan = -np.isnan(tb['Asyn'])
+    tb = tb[is_nonnan]
+
+    # Select only configurations with the right output (Asyn in the right
+    # window)
+    is_good = (tb['Asyn'] < Asynmax) & (tb['Asyn'] > Asynmin)
+
+    # only with enough synonymous diversity
+    is_good &= (tb['poly_syn'] > poly_synmin) & (tb['poly_syn'] < poly_synmax)
+
[…]

### 2013-03-12 synmut.tex
no hyperref

@@
 \usepackage{pslatex}
-\usepackage[colorlinks,linkcolor=red,citecolor=blue]{hyperref}
+%\usepackage[colorlinks,linkcolor=red,citecolor=blue]{hyperref}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
-\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
+%\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-We thank Jan Albert, Trevor Bedford and Pleuni Pennings and members of the lab for 
+We thank Jan Albert, Trevor Bedford, Pleuni Pennings and members of the lab for 
 stimulating discussions and critical reading of the manuscript.

### 2013-03-12 supplement_body.tex
Format for PLoS CompBio: missing author summary

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[ht]
+\begin{figure}[!ht]
 \begin{center}
-\includegraphics[width=0.35\linewidth]{Shankarappa_PCA_p1}
-\includegraphics[width=0.35\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p1}
-\includegraphics[width=0.35\linewidth]{Shankarappa_PCA_p7}
-\includegraphics[width=0.35\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p7}
+\includegraphics[width=0.35\linewidth]{Shankarappa_PCA_p1.pdf}
+\includegraphics[width=0.35\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p1.pdf}
+\includegraphics[width=0.35\linewidth]{Shankarappa_PCA_p7.pdf}
+\includegraphics[width=0.35\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p7.pdf}
 \caption{Structure of viral populations and patient selection.
@@
 % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[h]
+\begin{figure}[!ht]
 \begin{center}
-\includegraphics[width=\linewidth]{conservation_codons_genome}
+\includegraphics[width=\linewidth]{conservation_codons_genome.pdf}
 \caption{
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[h]
+\begin{figure}[!ht]
 \begin{center}
-\includegraphics[width=0.6\linewidth]{simulations_gradually_timeselec_deadagain}
+\includegraphics[width=0.6\linewidth]{simulations_gradually_timeselec_deadagain.pdf}
 \caption{
@@
 It has been found that virus is typically neutralized by serum from a few months
-earlier~\citep{richman_rapid_2003} but not by contemporary serum. We model this
+earlier~\cite{richman_rapid_2003} but not by contemporary serum. We model this
 evolving immune system by assuming that escaped variants lose their beneficial
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[h]
+\begin{figure}[!ht]
 \begin{center}
-\includegraphics[width=0.6\linewidth]{simulations_gradually_epitopes}
+\includegraphics[width=0.6\linewidth]{simulations_gradually_epitopes.pdf}
 \caption{

### 2013-03-12 synmut.tex
Format for PLoS CompBio: missing author summary

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\documentclass[rmp, twocolumn]{revtex4}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Template for PLoS
+% Version 1.0 January 2009
+\documentclass[10pt]{article}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{graphicx}
+\usepackage{cite}
+\usepackage{color} 
+% Use doublespacing - comment out for single spacing
+%\usepackage{setspace} 
+%\doublespacing
+
+% Text layout
+\topmargin 0.0cm
+\oddsidemargin 0.5cm
+\evensidemargin 0.5cm
+\textwidth 16cm 
+\textheight 21cm
+
+% Bold the 'Figure #' in the caption and separate it with a period
+% Captions will be left justified
+\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}
+
+% Use the PLoS provided bibtex style
+\bibliographystyle{plos2009}
+
+% Remove brackets from numbering in List of References
+\makeatletter
+\renewcommand{\@biblabel}[1]{\quad#1.}
+\makeatother
+
+% Leave date blank
+\date{}
+
+\pagestyle{myheadings}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \usepackage[utf8x]{inputenc}
-\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym,textcomp}
-\usepackage{color}
-\usepackage{graphicx}
+\usepackage{eucal,eurosym,textcomp}
 \usepackage[caption=false]{subfig}
-\usepackage{natbib}
 \usepackage{pslatex}
-%\usepackage[colorlinks,linkcolor=red,citecolor=blue]{hyperref}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
-%\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{document}
-\title{\Title}
-\author{\Author}
-\affiliation{\Affiliation}
-\date{\today}
+\begin{flushleft}
+{\Large
+\textbf{\Title}
+}
+\\
+Fabio~Zanini$^{1}$ and
+Richard~A.~Neher$^{1,\ast}$
+\\
+\bf{1} \Affiliation
+\\
+$\ast$ E-mail: richard.neher@tuebingen.mpg.de
+\end{flushleft}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{abstract}
-\noindent
+% Please keep the abstract between 250 and 300 words
+\section*{Abstract}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 Intrapatient HIV-1 evolution is dominated by selection on the protein level in the
@@
 
-\end{abstract}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\maketitle
+% Please keep the Author Summary between 150 and 200 words
+% Use first person. PLoS ONE authors please skip this step. 
+% Author Summary not valid for PLoS ONE submissions.   
+\section*{Author Summary}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
+\comment{to be written\dots}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Introduction}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 via cytotoxic CD8${}^+$ T-cells (CTLs) and neutralizing antibodies
-(nAbs)~\citep{rambaut_causes_2004} and facilitated by the high mutation rate
-~\citep{mansky_lower_1995,abram_nature_2010}. When the host develops a CTL or
+(nAbs)~\cite{rambaut_causes_2004} and facilitated by the high mutation rate
+~\cite{mansky_lower_1995,abram_nature_2010}. When the host develops a CTL or
 nAb response against a particular HIV-1 epitope, mutations in the viral genome that
@@
 in epitopes targeted by CTLs typically evolve during early infection and spread
-rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
+rapidly through the population~\cite{mcmichael_immune_2009}. During chronic
 infection, the most rapidly evolving parts of the HIV-1 genome are the variable
@@
 typically neutralizes virus extracted more than 3-6 months earlier but not contemporary
-virus \citep{richman_rapid_2003}.
+virus \cite{richman_rapid_2003}.
 
@@
 functional sites to detect purifying or directional selection
-\citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In addition to
+\cite{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In addition to
 maintaining protein function and avoiding the adaptive immune recognition,
@@
 in \env{} enhances nuclear export of full length or partially spliced viral
-transcripts via a complex stem-loop RNA structure~\citep{fernandes_hiv-1_2012}.
+transcripts via a complex stem-loop RNA structure~\cite{fernandes_hiv-1_2012}.
 Another well studied case is the interaction between viral reverse
@@
 required for priming reverse transcription (RT) and is bound by a pseudoknotted
-RNA structure in the viral 5' untranslated region~\citep{barat_interaction_1991,
+RNA structure in the viral 5' untranslated region~\cite{barat_interaction_1991,
 paillart_vitro_2002}.
@@
 completely neutrally. Some codons are favored over others in many species
-\citep{plotkin_synonymous_2011}. Recent studies have shown that genetically
+\cite{plotkin_synonymous_2011}. Recent studies have shown that genetically
 engineered HIV-1 strains with altered codon usage can in some cases produce more
 viral protein, but in general replicate less efficiently
-\citep{ngumbela_quantitative_2008, li_codon-usage-based_2012,
+\cite{ngumbela_quantitative_2008, li_codon-usage-based_2012,
 keating_rich_2009}. Codon deoptimization has been suggested as an attenuation
-strategy for polio and influenza~\citep{mueller_live_2010,coleman_virus_2008}.
+strategy for polio and influenza~\cite{mueller_live_2010,coleman_virus_2008}.
 Purifying selection beyond the protein sequence is therefore expected
-\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}, and it has been shown that
+\cite{forsdyke_reciprocal_1995,snoeck_mapping_2011}, and it has been shown that
 rates of evolution at synonymous sites vary along the HIV-1 genome
-\citep{mayrose_towards_2007}. Positive selection through the host adaptive
+\cite{mayrose_towards_2007}. Positive selection through the host adaptive
 immune system, however, is restricted to changes in the amino acid sequence.
@@
 argue that synonymous mutations reach high frequencies via genetic hitchhiking
-due to limited recombination in HIV-1 populations~\citep{neher_recombination_2010,
+due to limited recombination in HIV-1 populations~\cite{neher_recombination_2010,
 batorsky_estimate_2011}. We then compare our observations to computational
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Results}
+\section*{Results}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 probability $1-\nu$, it goes extinct. As illustrated in the inset of
-\FIG{aftsyn}, this is a  consequence of the fact that (i) exactly one
+\FIG{aft}A, this is a  consequence of the fact that (i) exactly one
 individual in the current population will be the common ancestor of the entire
@@
 nonsynonymous mutations observed in \env, \shankaregion, in patient
-p10~\citep{shankarappa_consistent_1999}. Despite many synonymous
-mutations reaching high frequency, few fix (panel~\ref{fig:aftsyn}); in
-constrast, many nonsynonymous mutations fix (panel~\ref{fig:aftnonsyn}).
+p10~\cite{shankarappa_consistent_1999}. Despite many synonymous
+mutations reaching high frequency, few fix (panel~\ref{fig:aft}A); in
+constrast, many nonsynonymous mutations fix (panel~\ref{fig:aft}B).
 Strictly speaking, no mutation in the HIV-1 population ever fixes because the 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Synonymous polymorphisms in \env, \shankaregion, are mostly deleterious}
+\subsection*{Synonymous polymorphisms in \env, \shankaregion, are mostly deleterious}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 analyzing data from seven patients from 
-\citet{shankarappa_consistent_1999} and \citep{liu_selection_2006} as well as three patients from
-\citet{bunnik_autologous_2008} (patients whose viral population was structured
+\cite{shankarappa_consistent_1999} and \cite{liu_selection_2006} as well as three patients from
+\cite{bunnik_autologous_2008} (patients whose viral population was structured
 were excluded from the analysis; see methods and \figurename~S\PCApat).  The
 former data set is restricted to the \shankaregion{} region of \env, while
-the data from \citet{bunnik_autologous_2008} cover the majority of \env.
+the data from \cite{bunnik_autologous_2008} cover the majority of \env.
 Considering all mutations in a frequency interval $[\nu_0-\delta\nu, \nu_0+\delta\nu]$ at some time
@@
 most synonymous mutations segregate for roughly one year and are lost much more
-frequently than expected (panel \ref{fig:fixp1}). The long-time probability of
+frequently than expected (panel \ref{fig:fixp}A). The long-time probability of
 fixation, $\pfix$, is shown as a function of the
-initial frequency $\nu_0$ in panel~\ref{fig:fixp2} (red line). 
+initial frequency $\nu_0$ in panel~\ref{fig:fixp}B (red line). 
 We find that $\pfix$ of synonymous mutations is far below
 the neutral expectation in  \shankaregion.  Outside of \shankaregion, using data from
-\citet{bunnik_autologous_2008} only, we find no such reduction in $\pfix$.
+\cite{bunnik_autologous_2008} only, we find no such reduction in $\pfix$.
 Restricted to the \shankaregion{} region, the sequence samples from
-\citet{bunnik_autologous_2008} are fully compatible with data from
-\citet{shankarappa_consistent_1999}. The nonsynonymous mutations seem to follow
+\cite{bunnik_autologous_2008} are fully compatible with data from
+\cite{shankarappa_consistent_1999}. The nonsynonymous mutations seem to follow
 more or less the neutral expectation (blue line) -- a point to which we will
 come back below.
-
-\begin{figure}
-\begin{center}
-\subfloat{\includegraphics[width=\linewidth]
-{Shankarappa_allele_freqs_trajectories_syn_p10.pdf}
-\label{fig:aftsyn}}\\
- \subfloat{\includegraphics[width=\linewidth]
-{Shankarappa_allele_freqs_trajectories_nonsyn_p10.pdf}
-\label{fig:aftnonsyn}}
-\caption{Time series of frequencies
-of synonymous (A) and nonsynonymous (B) derived alleles in \env, 
-\shankaregion, from patient 10~\cite{shankarappa_consistent_1999}.
-While many nonsynonymous mutations  fix, few synonymous
-mutations do even though they are frequently observed at intermediate
-frequencies. Colors indicate the position of the site along the \shankaregion{} region
-(blue to red). Inset: the fixation probability $\pfix$ of a neutral
-mutation is simply the likelihood that the future common ancestor at this 
-position is currently carrying it, i.e. the mutation frequency $\nu$.}
-\label{fig:aft}
-\end{center}
-\end{figure}
 
@@
 
-\begin{figure}
-\begin{center}
-\subfloat{\includegraphics[width=0.9\linewidth]{fixation_times.pdf}
-\label{fig:fixp1}}\\
-\subfloat{\includegraphics[width=0.9\linewidth]{fixation_probabilities.pdf}
-\label{fig:fixp2}}
-\caption{Fixation and loss of synonymous mutations.
-Panel A) shows how quickly synonymous mutations are purged from the populations. 
-Specifically, the figure shows the fraction of mutations that are still observed
-after $\Delta t$ days, conditional on being observed in one of the three frequency 
-intervals (different colors). 
-In each frequency interval, the fraction of synonymous
-mutations that ultimately survive is the fixation probability $\pfix$ conditional on the
-initial frequency. The neutral expectation for $\pfix=\nu_0$ is indicated by 
-dashed horizontal lines.
-Panel B) shows the fixation probability of derived synonymous
-alleles as a function of $\nu_0$. Polymorphisms within \shankaregion{} fix less
-often than expected for neutral mutations (indicated by the diagonal line).
-This suppression is not observed in other parts of \env~or for nonsynonymous mutations.
-The horizontal error bars on the abscissa are bin sizes, the vertical ones the
-standard deviation after 100 patient bootstraps of the data. Data from
-refs.~\cite{shankarappa_consistent_1999,liu_selection_2006, bunnik_autologous_2008}.}
-\label{fig:fixp}
-\end{center}
-\end{figure}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Synonymous mutations in \shankaregion{} tend to disrupt conserved RNA stems}
+\subsection*{Synonymous mutations in \shankaregion{} tend to disrupt conserved RNA stems}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \shankaregion{} is secondary structures in the viral RNA, the disruption of which
-is deleterious to the virus \citep{forsdyke_reciprocal_1995,
+is deleterious to the virus \cite{forsdyke_reciprocal_1995,
 snoeck_mapping_2011, sanjuan_interplay_2011}.
@@
 unpaired bases (the HIV-1 genome is a single stranded RNA)
-\citep{watts_architecture_2009}. The SHAPE assay has shown that the variable
+\cite{watts_architecture_2009}. The SHAPE assay has shown that the variable
 regions V1-V5 tend to be unpaired, while the conserved regions between those
 variable regions form stems.  We aligned the within-patient sequence samples 
-to the reference NL4-3 strain used by \citet{watts_architecture_2009} and 
+to the reference NL4-3 strain used in \cite{watts_architecture_2009} and 
 thereby assigned SHAPE reactivities to most positions in the alignment. 
@@
 frequencies above 15\%).
-As shown in \FIG{SHAPEA}, the reactivities of fixed alleles (red
+As shown in \FIG{SHAPE}A, the reactivities of fixed alleles (red
 histogram) are systematically larger than those of alleles that are lost (blue)
@@
 variable loops show little deviation from the neutral signature, see
-\FIG{SHAPEB}. This is consistent with important stem structures in conserved
+\FIG{SHAPE}B. This is consistent with important stem structures in conserved
 regions between loops.
@@
 usage bias (CUB). HIV-1 is known to prefer A-rich codons over highly expressed
-human codons~\citep{jenkins_extent_2003,kuyl_biased_2012}. We do not find,
+human codons~\cite{jenkins_extent_2003,kuyl_biased_2012}. We do not find,
 however, any evidence for a contribution of average CUB to the ultimate fate of
 synonymous alleles; consistently, HIV-1 does not seem to adapt its codon usage to
-its human host cells at the macroevolutionary level \citep{kuyl_biased_2012}.
-
-\begin{figure}
-\begin{center}
-\subfloat{\includegraphics[height=0.47\linewidth]{reactivities_histograms_syn.pdf}\label{fig:SHAPEA}}
-\subfloat{\includegraphics[height=0.46\linewidth]{fixation_probabilities_VnonV.pdf}\label{fig:SHAPEB}}
-\caption{Permissible synonymous mutations tend to be unpaired.
-Panel A) shows the distribution of SHAPE reactivities among sites at which synonymous 
-mutations fixed (red), sites at which mutations reached frequencies above 15\% but
-were susequently lost (blue), and sites at which no mutations were observed (green) 
-(all categories are restricted to the regions V1-V5$\pm 100$bp).
-Sites at which mutations fixed tend to have higher SHAPE reactivities, corresponding to
-less base pairing, than those at which mutations are lost.
-Sites at which no mutations are observed show an intermediate distribution of SHAPE values.
-Panel B) shows the fixation probability of synonymous mutations in
-\shankaregion{} separately for variable regions V3-V5 and the connecting conserved 
-regions C2-C4 that harbor RNA stems. As expected, the fixation probability is lower
-inside the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
-bunnik_autologous_2008, liu_selection_2006}.}
-\label{fig:SHAPE}
-\end{center}
-\end{figure}
+its human host cells at the macroevolutionary level \cite{kuyl_biased_2012}.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Deleterious mutations are brought to high frequency by hitchhiking}
+\subsection*{Deleterious mutations are brought to high frequency by hitchhiking}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 genome (in \shankaregion{} but not in the rest of \env{}; compare the red
-triangle line versus the green square line in \FIG{fixp2}).
+triangle line versus the green square line in \FIG{fixp}B).
 The region \shankaregion{} undergoes frequent adaptive changes to evade
@@
 that are linked to adaptive variants can reach high frequency. This process is
-known as hitchhiking \citep{smith_hitch-hiking_1974} or genetic draft
-\citep{gillespie_genetic_2000,neher_genetic_2011}. Hitchhiking is  apparent in
+known as hitchhiking \cite{smith_hitch-hiking_1974} or genetic draft
+\cite{gillespie_genetic_2000,neher_genetic_2011}. Hitchhiking is  apparent in
 \FIG{aft}, which shows that many mutations change rapidly in frequency as a
@@
 The approximate magnitude of the deleterious effects can be estimated from
-\FIG{fixp1}, which shows the distribution of times after which synonymous
+\FIG{fixp}A, which shows the distribution of times after which synonymous
 alleles at intermediate frequencies become fixed or lost. The typical time to
@@
 
-\begin{figure}
-\begin{center}
-\subfloat{\includegraphics[width=0.9\linewidth]{simulations_graduallydel.pdf}
-\label{fig:simfixpvar}}\\
-\subfloat{\includegraphics[width=0.9\linewidth]{simulations_syn.pdf}
-\label{fig:simsfig}}
-\caption{Distribution of selection coefficients on synonymous sites. Panel A)
-The depression in $\pfix$ depends on the deleterious effect size 
-of synonymous alleles. This parameter also reduces synonymous
-diversity, measured by the probability of a derived allele to be found at
-intermediate frequencies $P_\text{interm}$ (first inset).
-Panel B) To assess the parameter space that affects synonymous fixation and
-diversity, we run 2400 simulations with random parameters for deleterious effect
-size, fraction of deleterious synonymous sites, average escape rate $\epsilon$
-(color, blue to red corresponds to $10^{-2.5}$ to $10^{-1.5}$ per day), and rate of
-introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
-day). Only simulations that reproduce the synonymous diversity and fixation
-patterns observed in data are shown. These simulations demonstrate that
-deleterious effects are around $-0.002$ and a large fraction of the 
-synonymous mutations needs to be deleterious. As expected, larger
-$s_d$ require larger $\epsilon$. Parameters are chosen
-from prior distributions uniform in logspace as indicated by the red rectangle
-(see methods).}
-\label{fig:simheat}
-\end{center}
-\end{figure}
-
 To get a better idea of the range of parameters that are compatible with the
@@
 FFPopSim, which includes a module dedicated to intrapatient HIV
-evolution~\citep{zanini_ffpopsim:_2012}. For each simulation run, we specify the
+evolution~\cite{zanini_ffpopsim:_2012}. For each simulation run, we specify the
 deleterious effect of synonymous mutations, the fraction of synonymous mutations
@@
 etc. Net escape rates in chronic infections have been estimated to be on the
-order of $\epsilon = 0.01$ per day \citep{neher_recombination_2010,
+order of $\epsilon = 0.01$ per day \cite{neher_recombination_2010,
 Asquith:2006p28003}.
 
-\FIG{simfixpvar} shows simulation results for the fixation probability and the
+\FIG{simheat}A shows simulation results for the fixation probability and the
 synonymous diversity for different deleterious effects of synonymous mutations.
@@
 probability and the diagonal, which is the neutral expectation
-(\FIG{simfixpvar}, lower inset). If no fixation happens, the area will be
+(\FIG{simheat}A, lower inset). If no fixation happens, the area will be
 $-0.5$; if every mutation fixes, the area will be $+0.5$. In HIV-1 infected
@@
 for synonymous changes and $A_\text{nonsyn} \approx 0$ for nonsynonymous
-changes. In the three simulations shown in \FIG{simfixpvar}, the fixation
+changes. In the three simulations shown in \FIG{simheat}A, the fixation
 probability of synonymous alleles decreases from the neutral expectation
@@
 repeatedly simulated the evolution with random choices for the parameters in
-certain bounds, see \FIG{simsfig}. Among all simulations, we select the ones
+certain bounds, see \FIG{simheat}B. Among all simulations, we select the ones
 that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
@@
 simultaneously, a moderately high synonymous diversity. Specifically,
-\FIG{simsfig} shows parameter combinations for which we found $A_\text{syn} <
+\FIG{simheat}B shows parameter combinations for which we found $A_\text{syn} <
 -0.15$ and $0.0025 < P_\text{interm} < 0.010$. These conditions indicate that a
@@
 $|s_d| \sim 0.002$.  This result fits well the expectation based on the
-fixation/extinction times above (see \FIG{fixp1}). The results are plausible:
+fixation/extinction times above (see \FIG{fixp}A). The results are plausible:
 (i) a substantial depression in $\pfix$ requires pervasive deleterious
@@
 little or no fitness advantage. Consistent with this argument, larger
-deleterious effects in \FIG{simsfig} correspond to larger escape rates; and (iii)
+deleterious effects in \FIG{simheat}B correspond to larger escape rates; and (iii)
 mutations with a deleterious effect smaller than approximately $0.001$ behave
@@
 always fix once they reach high frequencies -- $A_{\mathrm{nonsyn}}$ is well
-above zero. This is incompatible with the blue line in \FIG{fixp2}: in an HIV-1
+above zero. This is incompatible with the blue line in \FIG{fixp}B: in an HIV-1
 infection, nonsynonymous mutations at high frequency often disappear again, even
@@
 beneficial and disappear soon, despite its quick initial rise in frequency. In
-support of this idea, \citet{richman_rapid_2003, bunnik_autologous_2008} report
-antibody responses to escape mutants. These responses are delayed by a few
+support of this idea, in \cite{richman_rapid_2003,
+bunnik_autologous_2008}, antibody responses to escape mutants have been reported.
+These responses are delayed by a few
 months, roughly matching the average time needed by an escape mutant to rise
@@
 sweeping nonsynonymous mutations in real HIV-1 infections has been shown
-\citep{moore_limited_2009, bar_early_2012}. This scenario has been explicitly 
+\cite{moore_limited_2009, bar_early_2012}. This scenario has been explicitly 
 observed in  the evolution of resistance to 3TC, where the mutation 
-M184V is often preceeded by M184I \citep{hedskog_dynamics_2010}. Similarly, AZT 
+M184V is often preceeded by M184I \cite{hedskog_dynamics_2010}. Similarly, AZT 
 resistance often emerges via the competing TAM and TAM1 pathways.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Discussion}
+\section*{Discussion}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 spiritu} has been recently used in a longitudinal study of influenza
-evolution~\citep{strelkowa_clonal_2012}. The central quantity used in that
+evolution~\cite{strelkowa_clonal_2012}. The central quantity used in that
 article, however, is a ratio between propagators of nonsynonymous and synonymous
@@
 evolution at synonymous sites varies considerable along the HIV-1 genome
-\citep{mayrose_towards_2007} and that this variation can confound estimates of
-selection on proteins substantially \citep{ngandu_extensive_2008}.
+\cite{mayrose_towards_2007} and that this variation can confound estimates of
+selection on proteins substantially \cite{ngandu_extensive_2008}.
 
@@
 hypervariable loops has also been proposed previously
-\citep{watts_architecture_2009, sanjuan_interplay_2011} and conserved RNA
+\cite{watts_architecture_2009, sanjuan_interplay_2011} and conserved RNA
 structures exist in different parts of the HIV-1 genome. Since there are
@@
 infrequent recombination for the evolution of HIV-1
-\citep{neher_recombination_2010, batorsky_estimate_2011,
+\cite{neher_recombination_2010, batorsky_estimate_2011,
 josefsson_majority_2011}. The recombination rate has been estimated to be on the
@@
 resistance evolution -- the linked region is of course much larger
-\citep{nijhuis_stochastic_1998}.
+\cite{nijhuis_stochastic_1998}.
 
@@
 results show that stochasticity due to linked selection is much more important.
-Such fluctuations have been termed \emph{genetic draft} by
-\citet{gillespie_genetic_2000}. Genetic draft in facultatively sexual population
-such as HIV-1 has been characterized in \citep{neher_genetic_2011}. Importantly,
+Such fluctuations have been termed \emph{genetic draft} in
+\cite{gillespie_genetic_2000}. Genetic draft in facultatively sexual population
+such as HIV-1 has been characterized in \cite{neher_genetic_2011}. Importantly,
 large population sizes are compatible with low diversity and fast coalescence
@@
 infected individuals typically neutralizes the virus that dominated the
-population a few (3-6) months earlier \citep{richman_rapid_2003}. This suggests that
+population a few (3-6) months earlier \cite{richman_rapid_2003}. This suggests that
 escape mutations cease to be beneficial after a few months and might revert if
@@
 antibody escape have revealed multiple escape mutations in the same epitope
-\citep{moore_limited_2009, bar_early_2012}. Presumably, each one of these
+\cite{moore_limited_2009, bar_early_2012}. Presumably, each one of these
 mutations is sufficient for escape but most combinations of them do not provide
@@
 in the context of preexisting drug resistance mutations
-\citep{boltz_ultrasensitive_2012}.
+\cite{boltz_ultrasensitive_2012}.
 
@@
 If genetic variation is only transiently beneficial, existing estimates of the
-strength of selection \citep{neher_recombination_2010,batorsky_estimate_2011}
+strength of selection \cite{neher_recombination_2010,batorsky_estimate_2011}
 could be substantial underestimates. Furthermore, weak conservation and
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Methods}
+\section*{Methods}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Sequence data collection}
+\subsection*{Sequence data collection}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 Longitudinal intrapatient viral RNA sequences were collected from published
-studies \citep{shankarappa_consistent_1999, liu_selection_2006,
+studies \cite{shankarappa_consistent_1999, liu_selection_2006,
 bunnik_autologous_2008} and downloaded from the Los Alamos National Laboratory
-(LANL) HIV sequence database~\citep{LANL2012}. The samples from some patients
+(LANL) HIV sequence database~\cite{LANL2012}. The samples from some patients
 show substantial population structure and were discarded (see
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Sequence analysis}
+\subsection*{Sequence analysis}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 The sequences were translated and the resulting amino acid sequences aligned
-using Muscle~\citep{edgar_muscle:_2004} to each other and the NL4-3 reference
+using Muscle~\cite{edgar_muscle:_2004} to each other and the NL4-3 reference
 sequences separately for each patient. Within each patient, the consensus
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Fixation probability and secondary structure}
+\subsection*{Fixation probability and secondary structure}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 the HIV-1 genome were downloaded from the journal website
-\citep{watts_architecture_2009}. Wherever possible, SHAPE reactivities were
+\cite{watts_architecture_2009}. Wherever possible, SHAPE reactivities were
 assigned to sites in the multiple sequence alignments for each patient through
 the alignment to the sequence of the NL4.3 virus used in
-ref.~\citep{watts_architecture_2009}. Problematic assignments in indel-rich
+\cite{watts_architecture_2009}. Problematic assignments in indel-rich
 regions were excluded from the analysis. The variable loops and flanking
 regions were identified manually starting from the annotated reference HXB2
-sequence from the LANL HIV database~\citep{LANL2012}.
+sequence from the LANL HIV database~\cite{LANL2012}.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Computer simulations}
+\subsection*{Computer simulations}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 Computer simulations were performed using FFPopSim
-\citep{zanini_ffpopsim:_2012}. Briefly, FFPopSim enables individual-based
+\cite{zanini_ffpopsim:_2012}. Briefly, FFPopSim enables individual-based
 simulations where each site in the genome is represented by one bit that can be
@@
 arbitrary fitness functions can be specified. We used a generation time of 1
-day, an outcrossing rate of $r=0.01$ per day \citep{batorsky_estimate_2011,
+day, an outcrossing rate of $r=0.01$ per day \cite{batorsky_estimate_2011,
 neher_recombination_2010}, a mutation rate of $\mu=10^{-5}$
-\citep{mansky_lower_1995, abram_nature_2010} and simulated intrapatient
+\cite{mansky_lower_1995, abram_nature_2010} and simulated intrapatient
 evolution for 6000 days. For simplicity, third positions of every codon were
@@
 probability could not be calculated; those runs were excluded from
-\FIG{simsfig}.
+\FIG{simheat}B.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Methods availability}
+\subsection*{Methods availability}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliographystyle{natbib}
 \bibliography{bib}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Figure Legends}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=\linewidth]{fig1.pdf}
+\caption{Time series of frequencies
+of synonymous (A) and nonsynonymous (B) derived alleles in \env, 
+\shankaregion, from patient 10~\cite{shankarappa_consistent_1999}.
+While many nonsynonymous mutations  fix, few synonymous
+mutations do even though they are frequently observed at intermediate
+frequencies. Colors indicate the position of the site along the \shankaregion{} region
+(blue to red). Inset: the fixation probability $\pfix$ of a neutral
+mutation is simply the likelihood that the future common ancestor at this 
+position is currently carrying it, i.e. the mutation frequency $\nu$.}
+\label{fig:aft}
+\end{center}
+\end{figure}
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=\linewidth]{fig2.pdf}
+\caption{Fixation and loss of synonymous mutations.
+Panel A) shows how quickly synonymous mutations are purged from the populations. 
+Specifically, the figure shows the fraction of mutations that are still observed
+after $\Delta t$ days, conditional on being observed in one of the three frequency 
+intervals (different colors). 
+In each frequency interval, the fraction of synonymous
+mutations that ultimately survive is the fixation probability $\pfix$ conditional on the
+initial frequency. The neutral expectation for $\pfix=\nu_0$ is indicated by 
+dashed horizontal lines.
+Panel B) shows the fixation probability of derived synonymous
+alleles as a function of $\nu_0$. Polymorphisms within \shankaregion{} fix less
+often than expected for neutral mutations (indicated by the diagonal line).
+This suppression is not observed in other parts of \env~or for nonsynonymous mutations.
+The horizontal error bars on the abscissa are bin sizes, the vertical ones the
+standard deviation after 100 patient bootstraps of the data. Data from
+refs.~\cite{shankarappa_consistent_1999,liu_selection_2006, bunnik_autologous_2008}.}
+\label{fig:fixp}
+\end{center}
+\end{figure}
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=\linewidth]{fig3.pdf}
+\caption{Permissible synonymous mutations tend to be unpaired.
+Panel A) shows the distribution of SHAPE reactivities among sites at which synonymous 
+mutations fixed (red), sites at which mutations reached frequencies above 15\% but
+were susequently lost (blue), and sites at which no mutations were observed (green) 
+(all categories are restricted to the regions V1-V5$\pm 100$bp).
+Sites at which mutations fixed tend to have higher SHAPE reactivities, corresponding to
+less base pairing, than those at which mutations are lost.
+Sites at which no mutations are observed show an intermediate distribution of SHAPE values.
+Panel B) shows the fixation probability of synonymous mutations in
+\shankaregion{} separately for variable regions V3-V5 and the connecting conserved 
+regions C2-C4 that harbor RNA stems. As expected, the fixation probability is lower
+inside the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
+bunnik_autologous_2008, liu_selection_2006}.}
+\label{fig:SHAPE}
+\end{center}
+\end{figure}
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=\linewidth]{fig4.pdf}
+\caption{Distribution of selection coefficients on synonymous sites. Panel A)
+The depression in $\pfix$ depends on the deleterious effect size 
+of synonymous alleles. This parameter also reduces synonymous
+diversity, measured by the probability of a derived allele to be found at
+intermediate frequencies $P_\text{interm}$ (first inset).
+Panel B) To assess the parameter space that affects synonymous fixation and
+diversity, we run 2400 simulations with random parameters for deleterious effect
+size, fraction of deleterious synonymous sites, average escape rate $\epsilon$
+(color, blue to red corresponds to $10^{-2.5}$ to $10^{-1.5}$ per day), and rate of
+introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
+day). Only simulations that reproduce the synonymous diversity and fixation
+patterns observed in data are shown. These simulations demonstrate that
+deleterious effects are around $-0.002$ and a large fraction of the 
+synonymous mutations needs to be deleterious. As expected, larger
+$s_d$ require larger $\epsilon$. Parameters are chosen
+from prior distributions uniform in logspace as indicated by the red rectangle
+(see methods).}
+\label{fig:simheat}
+\end{center}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Appendix
 \newpage

### 2013-03-12 supplement_body.tex
remove packages disliked by plos

@@
 plots for data from patient p7. Samples after day 1000 split into two clusters in the PCA and no mutations that arise after day 1000 fix, presumably because they are restricted
-to one subpopulation. All patients like p7 (p4, p7, p8, p9 from ref.~\citealp{shankarappa_consistent_1999} and
-ACH19542 and ACH19768 from ref.~\citealp{bunnik_autologous_2008}) were excluded
+to one subpopulation. All patients like p7 (p4, p7, p8, p9 from ref.~\cite{shankarappa_consistent_1999} and
+ACH19542 and ACH19768 from ref.~\cite{bunnik_autologous_2008}) were excluded
 from our analysis.}
@@
 \newpage
-% 
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \section{Nonsynonymous changes outside of variable regions are deleterious}
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \begin{figure}[h]
-% \begin{center}
-% \includegraphics[width=0.8\linewidth]{synmut_conservation_4fold_synnonsyn}
-% \caption{Cumulative distribution of synonymous and nonsynonymous diversity in
-% {\it gag} in the LANL reference panel (filtered
-% sequences only, version 2011)~\cite{LANL2012}. Sites such that the consensus codon
-% has three synonymous and six nonsynonymous single mutants were used, and the number of observed mutants
-% of a certain type over the number of possible mutants is plotted. Nonsynonymous
-% changes are observed less often, {\it ergo} are more conserved, than
-% synonymous changes. It can be therefore assumed, as mentioned in the main text,
-% that non-escape nonsynonymous changes involve a large fitness cost.}
-% \label{fig:synnonsyncons}
-% \end{center}
-% \end{figure}
-% \newpage
 
@@
 reset to its fitness cost of $-0.02$ with probability
-\[ P_\text{recognized}(t) = c \cdot \nu(t), \] 
+$ P_\text{recognized}(t) = c \cdot \nu(t), $
 per generation, where $c$ is a constant coefficient shown in the legend that

### 2013-03-12 synmut.tex
remove packages disliked by plos

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\usepackage[english]{babel}
-\usepackage[utf8x]{inputenc}
-\usepackage{eucal,eurosym,textcomp}
-\usepackage[caption=false]{subfig}
-\usepackage{pslatex}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \graphicspath{{./figures/}}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\renewcommand{\thesubfigure}{\Alph{subfigure}}
 \newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
@@
 All analysis and computer simulation scripts, as well as the sequence alignments
-used, are available for download at \url{http://git.tuebingen.mpg.de/synmut}.
+used, are available for download at http://git.tuebingen.mpg.de/synmut.
 
@@
 \begin{center}
-\includegraphics[width=\linewidth]{fig1.pdf}
+%\includegraphics[width=\linewidth]{fig1.pdf}
 \caption{Time series of frequencies
@@
 \begin{center}
-\includegraphics[width=\linewidth]{fig2.pdf}
+%\includegraphics[width=\linewidth]{fig2.pdf}
 \caption{Fixation and loss of synonymous mutations.
@@
 \begin{center}
-\includegraphics[width=\linewidth]{fig3.pdf}
+%\includegraphics[width=\linewidth]{fig3.pdf}
 \caption{Permissible synonymous mutations tend to be unpaired.
@@
 \begin{center}
-\includegraphics[width=\linewidth]{fig4.pdf}
+%\includegraphics[width=\linewidth]{fig4.pdf}
 \caption{Distribution of selection coefficients on synonymous sites. Panel A)
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Appendix
-\newpage
-\appendix
-\onecolumngrid
-\setcounter{figure}{0}
-\include{supplement_body}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Appendix
+%\newpage
+%\appendix
+%\setcounter{figure}{0}
+%\include{supplement_body}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2013-03-12 synmut.tex
Author summary

@@
 % Please keep the Author Summary between 150 and 200 words
-% Use first person. PLoS ONE authors please skip this step. 
-% Author Summary not valid for PLoS ONE submissions.   
+% Use first person.   
 \section*{Author Summary}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\comment{to be written\dots}
+The Human Immunodeficiency Virus (HIV) epidemic is a plague of our era. The
+main reason why it is so hard to treat is that, althought one single strain of
+the virus is transmitted, many variants appear quickly within the same patient
+by random mutation. Among all mutations, synonymous ones do not change the amino
+acid sequence of the viral proteins and are often assumed to be of no
+evolutionary consequence. In this article, we show that many synonymous
+mutations imply a slight fitness cost for the virus; this cost is
+linked to the preservation of certain base pairs in the single-stranded HIV RNA.
+We argue that deleterious synonymous mutations can be widespread within an HIV
+infection, despite their cost and for a limited time only, if they arise by
+chance on a very efficient virus. Furthermore, we suggest that even the most
+efficient virus has only a small chance of eventually dominate, because
+(i) it must often compete with incompatible and equally efficient variants, and
+(ii) the host immune systems soon develops countermeasures.
 

### 2013-05-17 synmut.tex
Small changes in text

--- 2013-03-12 synmut.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Template for PLoS
-% Version 1.0 January 2009
-\documentclass[10pt]{article}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{graphicx}
-\usepackage{cite}
-\usepackage{color} 
-% Use doublespacing - comment out for single spacing
-%\usepackage{setspace} 
-%\doublespacing
-
-% Text layout
-\topmargin 0.0cm
-\oddsidemargin 0.5cm
-\evensidemargin 0.5cm
-\textwidth 16cm 
-\textheight 21cm
-
-% Bold the 'Figure #' in the caption and separate it with a period
-% Captions will be left justified
-\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}
-
-% Use the PLoS provided bibtex style
-\bibliographystyle{plos2009}
-
-% Remove brackets from numbering in List of References
-\makeatletter
-\renewcommand{\@biblabel}[1]{\quad#1.}
-\makeatother
-
-% Leave date blank
-\date{}
-
-\pagestyle{myheadings}
+\documentclass[rmp, twocolumn]{revtex4}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \usepackage[utf8x]{inputenc}
-\usepackage{eucal,eurosym,textcomp}
+\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym,textcomp}
+\usepackage{color}
+\usepackage{graphicx}
 \usepackage[caption=false]{subfig}
+\usepackage{natbib}
 \usepackage{pslatex}
+%\usepackage[colorlinks,linkcolor=red,citecolor=blue]{hyperref}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
+%\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{document}
-\begin{flushleft}
-{\Large
-\textbf{\Title}
-}
-\\
-Fabio~Zanini$^{1}$ and
-Richard~A.~Neher$^{1,\ast}$
-\\
-\bf{1} \Affiliation
-\\
-$\ast$ E-mail: richard.neher@tuebingen.mpg.de
-\end{flushleft}
+\title{\Title}
+\author{\Author}
+\affiliation{\Affiliation}
+\date{\today}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Please keep the abstract between 250 and 300 words
-\section*{Abstract}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{abstract}
+\noindent
 Intrapatient HIV-1 evolution is dominated by selection on the protein level in the
@@
 
+\end{abstract}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Please keep the Author Summary between 150 and 200 words
-% Use first person. PLoS ONE authors please skip this step. 
-% Author Summary not valid for PLoS ONE submissions.   
-\section*{Author Summary}
+\maketitle
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\comment{to be written\dots}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Introduction}
+\section{Introduction}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 via cytotoxic CD8${}^+$ T-cells (CTLs) and neutralizing antibodies
-(nAbs)~\cite{rambaut_causes_2004} and facilitated by the high mutation rate
-~\cite{mansky_lower_1995,abram_nature_2010}. When the host develops a CTL or
+(nAbs)~\citep{rambaut_causes_2004} and facilitated by the high mutation rate
+~\citep{mansky_lower_1995,abram_nature_2010}. When the host develops a CTL or
 nAb response against a particular HIV-1 epitope, mutations in the viral genome that
@@
 in epitopes targeted by CTLs typically evolve during early infection and spread
-rapidly through the population~\cite{mcmichael_immune_2009}. During chronic
+rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
 infection, the most rapidly evolving parts of the HIV-1 genome are the variable
@@
 typically neutralizes virus extracted more than 3-6 months earlier but not contemporary
-virus \cite{richman_rapid_2003}.
+virus \citep{richman_rapid_2003}.
 
@@
 functional sites to detect purifying or directional selection
-\cite{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In addition to
+\citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In addition to
 maintaining protein function and avoiding the adaptive immune recognition,
@@
 in \env{} enhances nuclear export of full length or partially spliced viral
-transcripts via a complex stem-loop RNA structure~\cite{fernandes_hiv-1_2012}.
+transcripts via a complex stem-loop RNA structure~\citep{fernandes_hiv-1_2012}.
 Another well studied case is the interaction between viral reverse
@@
 required for priming reverse transcription (RT) and is bound by a pseudoknotted
-RNA structure in the viral 5' untranslated region~\cite{barat_interaction_1991,
+RNA structure in the viral 5' untranslated region~\citep{barat_interaction_1991,
 paillart_vitro_2002}.
@@
 completely neutrally. Some codons are favored over others in many species
-\cite{plotkin_synonymous_2011}. Recent studies have shown that genetically
+\citep{plotkin_synonymous_2011}. Recent studies have shown that genetically
 engineered HIV-1 strains with altered codon usage can in some cases produce more
 viral protein, but in general replicate less efficiently
-\cite{ngumbela_quantitative_2008, li_codon-usage-based_2012,
+\citep{ngumbela_quantitative_2008, li_codon-usage-based_2012,
 keating_rich_2009}. Codon deoptimization has been suggested as an attenuation
-strategy for polio and influenza~\cite{mueller_live_2010,coleman_virus_2008}.
+strategy for polio and influenza~\citep{mueller_live_2010,coleman_virus_2008}.
 Purifying selection beyond the protein sequence is therefore expected
-\cite{forsdyke_reciprocal_1995,snoeck_mapping_2011}, and it has been shown that
+\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}, and it has been shown that
 rates of evolution at synonymous sites vary along the HIV-1 genome
-\cite{mayrose_towards_2007}. Positive selection through the host adaptive
+\citep{mayrose_towards_2007}. Positive selection through the host adaptive
 immune system, however, is restricted to changes in the amino acid sequence.
@@
 argue that synonymous mutations reach high frequencies via genetic hitchhiking
-due to limited recombination in HIV-1 populations~\cite{neher_recombination_2010,
+due to limited recombination in HIV-1 populations~\citep{neher_recombination_2010,
 batorsky_estimate_2011}. We then compare our observations to computational
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Results}
+\section{Results}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 probability $1-\nu$, it goes extinct. As illustrated in the inset of
-\FIG{aft}A, this is a  consequence of the fact that (i) exactly one
+\FIG{aftsyn}, this is a  consequence of the fact that (i) exactly one
 individual in the current population will be the common ancestor of the entire
@@
 nonsynonymous mutations observed in \env, \shankaregion, in patient
-p10~\cite{shankarappa_consistent_1999}. Despite many synonymous
-mutations reaching high frequency, few fix (panel~\ref{fig:aft}A); in
-constrast, many nonsynonymous mutations fix (panel~\ref{fig:aft}B).
+p10~\citep{shankarappa_consistent_1999}. Despite many synonymous
+mutations reaching high frequency, few fix (panel~\ref{fig:aftsyn}); in
+constrast, many nonsynonymous mutations fix (panel~\ref{fig:aftnonsyn}).
 Strictly speaking, no mutation in the HIV-1 population ever fixes because the 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Synonymous polymorphisms in \env, \shankaregion, are mostly deleterious}
+\subsection{Synonymous polymorphisms in \env, \shankaregion, are mostly deleterious}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 analyzing data from seven patients from
-\cite{shankarappa_consistent_1999} and \cite{liu_selection_2006} as well as three patients from
-\cite{bunnik_autologous_2008} (patients whose viral population was structured
+\citet{shankarappa_consistent_1999} and \citep{liu_selection_2006} as well as three patients from
+\citet{bunnik_autologous_2008} (patients whose viral population was structured
 were excluded from the analysis; see methods and \figurename~S\PCApat).  The
 former data set is restricted to the \shankaregion{} region of \env, while
-the data from \cite{bunnik_autologous_2008} cover the majority of \env.
+the data from \citet{bunnik_autologous_2008} cover the majority of \env.
 Considering all mutations in a frequency interval $[\nu_0-\delta\nu, \nu_0+\delta\nu]$ at some time
@@
 most synonymous mutations segregate for roughly one year and are lost much more
-frequently than expected (panel \ref{fig:fixp}A). The long-time probability of
+frequently than expected (panel \ref{fig:fixp1}). The long-time probability of
 fixation, $\pfix$, is shown as a function of the
-initial frequency $\nu_0$ in panel~\ref{fig:fixp}B (red line). 
+initial frequency $\nu_0$ in panel~\ref{fig:fixp2} (red line). 
 We find that $\pfix$ of synonymous mutations is far below
 the neutral expectation in  \shankaregion.  Outside of \shankaregion, using data from
-\cite{bunnik_autologous_2008} only, we find no such reduction in $\pfix$.
+\citet{bunnik_autologous_2008} only, we find no such reduction in $\pfix$.
 Restricted to the \shankaregion{} region, the sequence samples from
-\cite{bunnik_autologous_2008} are fully compatible with data from
-\cite{shankarappa_consistent_1999}. The nonsynonymous mutations seem to follow
+\citet{bunnik_autologous_2008} are fully compatible with data from
+\citet{shankarappa_consistent_1999}. The nonsynonymous mutations seem to follow
 more or less the neutral expectation (blue line) -- a point to which we will
 come back below.
+
+\begin{figure}
+\begin{center}
+\subfloat{\includegraphics[width=\linewidth]
+{Shankarappa_allele_freqs_trajectories_syn_p10.pdf}
+\label{fig:aftsyn}}\\
+ \subfloat{\includegraphics[width=\linewidth]
+{Shankarappa_allele_freqs_trajectories_nonsyn_p10.pdf}
+\label{fig:aftnonsyn}}
+\caption{Time series of frequencies
+of synonymous (A) and nonsynonymous (B) derived alleles in \env, 
+\shankaregion, from patient 10~\cite{shankarappa_consistent_1999}.
+While many nonsynonymous mutations  fix, few synonymous
+mutations do even though they are frequently observed at intermediate
+frequencies. Colors indicate the position of the site along the \shankaregion{} region
+(blue to red). Inset: the fixation probability $\pfix$ of a neutral
+mutation is simply the likelihood that the future common ancestor at this 
+position is currently carrying it, i.e. the mutation frequency $\nu$.}
+\label{fig:aft}
+\end{center}
+\end{figure}
 
@@
 
+\begin{figure}
+\begin{center}
+\subfloat{\includegraphics[width=0.9\linewidth]{fixation_times.pdf}
+\label{fig:fixp1}}\\
+\subfloat{\includegraphics[width=0.9\linewidth]{fixation_probabilities.pdf}
+\label{fig:fixp2}}
+\caption{Fixation and loss of synonymous mutations.
+Panel A) shows how quickly synonymous mutations are purged from the populations. 
+Specifically, the figure shows the fraction of mutations that are still observed
+after $\Delta t$ days, conditional on being observed in one of the three frequency 
+intervals (different colors). 
+In each frequency interval, the fraction of synonymous
+mutations that ultimately survive is the fixation probability $\pfix$ conditional on the
+initial frequency. The neutral expectation for $\pfix=\nu_0$ is indicated by 
+dashed horizontal lines.
+Panel B) shows the fixation probability of derived synonymous
+alleles as a function of $\nu_0$. Polymorphisms within \shankaregion{} fix less
+often than expected for neutral mutations (indicated by the diagonal line).
+This suppression is not observed in other parts of \env~or for nonsynonymous mutations.
+The horizontal error bars on the abscissa are bin sizes, the vertical ones the
+standard deviation after 100 patient bootstraps of the data. Data from
+refs.~\cite{shankarappa_consistent_1999,liu_selection_2006, bunnik_autologous_2008}.}
+\label{fig:fixp}
+\end{center}
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Synonymous mutations in \shankaregion{} tend to disrupt conserved RNA stems}
+\subsection{Synonymous mutations in \shankaregion{} tend to disrupt conserved RNA stems}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \shankaregion{} is secondary structures in the viral RNA, the disruption of which
-is deleterious to the virus \cite{forsdyke_reciprocal_1995,
+is deleterious to the virus \citep{forsdyke_reciprocal_1995,
 snoeck_mapping_2011, sanjuan_interplay_2011}.
@@
 unpaired bases (the HIV-1 genome is a single stranded RNA)
-\cite{watts_architecture_2009}. The SHAPE assay has shown that the variable
+\citep{watts_architecture_2009}. The SHAPE assay has shown that the variable
 regions V1-V5 tend to be unpaired, while the conserved regions between those
 variable regions form stems.  We aligned the within-patient sequence samples 
-to the reference NL4-3 strain used in \cite{watts_architecture_2009} and 
+to the reference NL4-3 strain used by \citet{watts_architecture_2009} and 
 thereby assigned SHAPE reactivities to most positions in the alignment. 
@@
 frequencies above 15\%).
-As shown in \FIG{SHAPE}A, the reactivities of fixed alleles (red
+As shown in \FIG{SHAPEA}, the reactivities of fixed alleles (red
 histogram) are systematically larger than those of alleles that are lost (blue)
@@
 variable loops show little deviation from the neutral signature, see
-\FIG{SHAPE}B. This is consistent with important stem structures in conserved
+\FIG{SHAPEB}. This is consistent with important stem structures in conserved
 regions between loops.
@@
 usage bias (CUB). HIV-1 is known to prefer A-rich codons over highly expressed
-human codons~\cite{jenkins_extent_2003,kuyl_biased_2012}. We do not find,
+human codons~\citep{jenkins_extent_2003,kuyl_biased_2012}. We do not find,
 however, any evidence for a contribution of average CUB to the ultimate fate of
 synonymous alleles; consistently, HIV-1 does not seem to adapt its codon usage to
-its human host cells at the macroevolutionary level \cite{kuyl_biased_2012}.
+its human host cells at the macroevolutionary level \citep{kuyl_biased_2012}.
+
+\begin{figure}
+\begin{center}
+\subfloat{\includegraphics[height=0.47\linewidth]{reactivities_histograms_syn.pdf}\label{fig:SHAPEA}}
+\subfloat{\includegraphics[height=0.46\linewidth]{fixation_probabilities_VnonV.pdf}\label{fig:SHAPEB}}
+\caption{Permissible synonymous mutations tend to be unpaired.
+Panel A) shows the distribution of SHAPE reactivities among sites at which synonymous 
+mutations fixed (red), sites at which mutations reached frequencies above 15\% but
+were susequently lost (blue), and sites at which no mutations were observed (green) 
+(all categories are restricted to the regions V1-V5$\pm 100$bp).
+Sites at which mutations fixed tend to have higher SHAPE reactivities, corresponding to
+less base pairing, than those at which mutations are lost.
+Sites at which no mutations are observed show an intermediate distribution of SHAPE values.
+Panel B) shows the fixation probability of synonymous mutations in
+\shankaregion{} separately for variable regions V3-V5 and the connecting conserved 
+regions C2-C4 that harbor RNA stems. As expected, the fixation probability is lower
+inside the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
+bunnik_autologous_2008, liu_selection_2006}.}
+\label{fig:SHAPE}
+\end{center}
+\end{figure}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Deleterious mutations are brought to high frequency by hitchhiking}
+\subsection{Deleterious mutations are brought to high frequency by hitchhiking}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 genome (in \shankaregion{} but not in the rest of \env{}; compare the red
-triangle line versus the green square line in \FIG{fixp}B).
+triangle line versus the green square line in \FIG{fixp2}).
 The region \shankaregion{} undergoes frequent adaptive changes to evade
@@
 that are linked to adaptive variants can reach high frequency. This process is
-known as hitchhiking \cite{smith_hitch-hiking_1974} or genetic draft
-\cite{gillespie_genetic_2000,neher_genetic_2011}. Hitchhiking is  apparent in
+known as hitchhiking \citep{smith_hitch-hiking_1974} or genetic draft
+\citep{gillespie_genetic_2000,neher_genetic_2011}. Hitchhiking is  apparent in
 \FIG{aft}, which shows that many mutations change rapidly in frequency as a
@@
 The approximate magnitude of the deleterious effects can be estimated from
-\FIG{fixp}A, which shows the distribution of times after which synonymous
+\FIG{fixp1}, which shows the distribution of times after which synonymous
 alleles at intermediate frequencies become fixed or lost. The typical time to
@@
 
+\begin{figure}
+\begin{center}
+\subfloat{\includegraphics[width=0.9\linewidth]{simulations_graduallydel.pdf}
+\label{fig:simfixpvar}}\\
+\subfloat{\includegraphics[width=0.9\linewidth]{simulations_syn.pdf}
+\label{fig:simsfig}}
+\caption{Distribution of selection coefficients on synonymous sites. Panel A)
+The depression in $\pfix$ depends on the deleterious effect size 
+of synonymous alleles. This parameter also reduces synonymous
+diversity, measured by the probability of a derived allele to be found at
+intermediate frequencies $P_\text{interm}$ (first inset).
+Panel B) To assess the parameter space that affects synonymous fixation and
+diversity, we run 2400 simulations with random parameters for deleterious effect
+size, fraction of deleterious synonymous sites, average escape rate $\epsilon$
+(color, blue to red corresponds to $10^{-2.5}$ to $10^{-1.5}$ per day), and rate of
+introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
+day). Only simulations that reproduce the synonymous diversity and fixation
+patterns observed in data are shown. These simulations demonstrate that
+deleterious effects are around $-0.002$ and a large fraction of the 
+synonymous mutations needs to be deleterious. As expected, larger
+$s_d$ require larger $\epsilon$. Parameters are chosen
+from prior distributions uniform in logspace as indicated by the red rectangle
+(see methods).}
+\label{fig:simheat}
+\end{center}
+\end{figure}
+
 To get a better idea of the range of parameters that are compatible with the
@@
 FFPopSim, which includes a module dedicated to intrapatient HIV
-evolution~\cite{zanini_ffpopsim:_2012}. For each simulation run, we specify the
+evolution~\citep{zanini_ffpopsim:_2012}. For each simulation run, we specify the
 deleterious effect of synonymous mutations, the fraction of synonymous mutations
@@
 etc. Net escape rates in chronic infections have been estimated to be on the
-order of $\epsilon = 0.01$ per day \cite{neher_recombination_2010,
+order of $\epsilon = 0.01$ per day \citep{neher_recombination_2010,
 Asquith:2006p28003}.
 
-\FIG{simheat}A shows simulation results for the fixation probability and the
+\FIG{simfixpvar} shows simulation results for the fixation probability and the
 synonymous diversity for different deleterious effects of synonymous mutations.
@@
 probability and the diagonal, which is the neutral expectation
-(\FIG{simheat}A, lower inset). If no fixation happens, the area will be
+(\FIG{simfixpvar}, lower inset). If no fixation happens, the area will be
 $-0.5$; if every mutation fixes, the area will be $+0.5$. In HIV-1 infected
@@
 for synonymous changes and $A_\text{nonsyn} \approx 0$ for nonsynonymous
-changes. In the three simulations shown in \FIG{simheat}A, the fixation
+changes. In the three simulations shown in \FIG{simfixpvar}, the fixation
 probability of synonymous alleles decreases from the neutral expectation
@@
 repeatedly simulated the evolution with random choices for the parameters in
-certain bounds, see \FIG{simheat}B. Among all simulations, we select the ones
+certain bounds, see \FIG{simsfig}. Among all simulations, we select the ones
 that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
@@
 simultaneously, a moderately high synonymous diversity. Specifically,
-\FIG{simheat}B shows parameter combinations for which we found $A_\text{syn} <
+\FIG{simsfig} shows parameter combinations for which we found $A_\text{syn} <
 -0.15$ and $0.0025 < P_\text{interm} < 0.010$. These conditions indicate that a
@@
 $|s_d| \sim 0.002$.  This result fits well the expectation based on the
-fixation/extinction times above (see \FIG{fixp}A). The results are plausible:
+fixation/extinction times above (see \FIG{fixp1}). The results are plausible:
 (i) a substantial depression in $\pfix$ requires pervasive deleterious
@@
 little or no fitness advantage. Consistent with this argument, larger
-deleterious effects in \FIG{simheat}B correspond to larger escape rates; and (iii)
+deleterious effects in \FIG{simsfig} correspond to larger escape rates; and (iii)
 mutations with a deleterious effect smaller than approximately $0.001$ behave
@@
 always fix once they reach high frequencies -- $A_{\mathrm{nonsyn}}$ is well
-above zero. This is incompatible with the blue line in \FIG{fixp}B: in an HIV-1
+above zero. This is incompatible with the blue line in \FIG{fixp2}: in an HIV-1
 infection, nonsynonymous mutations at high frequency often disappear again, even
@@
 beneficial and disappear soon, despite its quick initial rise in frequency. In
-support of this idea, in \cite{richman_rapid_2003,
-bunnik_autologous_2008}, antibody responses to escape mutants have been reported.
+support of this idea, in \citet{richman_rapid_2003, bunnik_autologous_2008},
+antibody responses to escape mutants have been reported.
 These responses are delayed by a few
@@
 sweeping nonsynonymous mutations in real HIV-1 infections has been shown
-\cite{moore_limited_2009, bar_early_2012}. This scenario has been explicitly 
+\citep{moore_limited_2009, bar_early_2012}. This scenario has been explicitly 
 observed in  the evolution of resistance to 3TC, where the mutation 
-M184V is often preceeded by M184I \cite{hedskog_dynamics_2010}. Similarly, AZT 
+M184V is often preceeded by M184I \citep{hedskog_dynamics_2010}. Similarly, AZT 
 resistance often emerges via the competing TAM and TAM1 pathways.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Discussion}
+\section{Discussion}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 spiritu} has been recently used in a longitudinal study of influenza
-evolution~\cite{strelkowa_clonal_2012}. The central quantity used in that
+evolution~\citep{strelkowa_clonal_2012}. The central quantity used in that
 article, however, is a ratio between propagators of nonsynonymous and synonymous
@@
 evolution at synonymous sites varies considerable along the HIV-1 genome
-\cite{mayrose_towards_2007} and that this variation can confound estimates of
-selection on proteins substantially \cite{ngandu_extensive_2008}.
+\citep{mayrose_towards_2007} and that this variation can confound estimates of
+selection on proteins substantially \citep{ngandu_extensive_2008}.
 
@@
 hypervariable loops has also been proposed previously
-\cite{watts_architecture_2009, sanjuan_interplay_2011} and conserved RNA
+\citep{watts_architecture_2009, sanjuan_interplay_2011} and conserved RNA
 structures exist in different parts of the HIV-1 genome. Since there are
@@
 infrequent recombination for the evolution of HIV-1
-\cite{neher_recombination_2010, batorsky_estimate_2011,
+\citep{neher_recombination_2010, batorsky_estimate_2011,
 josefsson_majority_2011}. The recombination rate has been estimated to be on the
@@
 resistance evolution -- the linked region is of course much larger
-\cite{nijhuis_stochastic_1998}.
+\citep{nijhuis_stochastic_1998}.
 
@@
 Such fluctuations have been termed \emph{genetic draft} in
-\cite{gillespie_genetic_2000}. Genetic draft in facultatively sexual population
-such as HIV-1 has been characterized in \cite{neher_genetic_2011}. Importantly,
+\citet{gillespie_genetic_2000}. Genetic draft in facultatively sexual population
+such as HIV-1 has been characterized in \citep{neher_genetic_2011}. Importantly,
 large population sizes are compatible with low diversity and fast coalescence
@@
 infected individuals typically neutralizes the virus that dominated the
-population a few (3-6) months earlier \cite{richman_rapid_2003}. This suggests that
+population a few (3-6) months earlier \citep{richman_rapid_2003}. This suggests that
 escape mutations cease to be beneficial after a few months and might revert if
@@
 antibody escape have revealed multiple escape mutations in the same epitope
-\cite{moore_limited_2009, bar_early_2012}. Presumably, each one of these
+\citep{moore_limited_2009, bar_early_2012}. Presumably, each one of these
 mutations is sufficient for escape but most combinations of them do not provide
@@
 in the context of preexisting drug resistance mutations
-\cite{boltz_ultrasensitive_2012}.
+\citep{boltz_ultrasensitive_2012}.
 
@@
 If genetic variation is only transiently beneficial, existing estimates of the
-strength of selection \cite{neher_recombination_2010,batorsky_estimate_2011}
+strength of selection \citep{neher_recombination_2010,batorsky_estimate_2011}
 could be substantial underestimates. Furthermore, weak conservation and
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Methods}
+\section{Methods}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Sequence data collection}
+\subsection{Sequence data collection}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 Longitudinal intrapatient viral RNA sequences were collected from published
-studies \cite{shankarappa_consistent_1999, liu_selection_2006,
+studies \citep{shankarappa_consistent_1999, liu_selection_2006,
 bunnik_autologous_2008} and downloaded from the Los Alamos National Laboratory
-(LANL) HIV sequence database~\cite{LANL2012}. The samples from some patients
+(LANL) HIV sequence database~\citep{LANL2012}. The samples from some patients
 show substantial population structure and were discarded (see
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Sequence analysis}
+\subsection{Sequence analysis}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 The sequences were translated and the resulting amino acid sequences aligned
-using Muscle~\cite{edgar_muscle:_2004} to each other and the NL4-3 reference
+using Muscle~\citep{edgar_muscle:_2004} to each other and the NL4-3 reference
 sequences separately for each patient. Within each patient, the consensus
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Fixation probability and secondary structure}
+\subsection{Fixation probability and secondary structure}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 the HIV-1 genome were downloaded from the journal website
-\cite{watts_architecture_2009}. Wherever possible, SHAPE reactivities were
+\citep{watts_architecture_2009}. Wherever possible, SHAPE reactivities were
 assigned to sites in the multiple sequence alignments for each patient through
 the alignment to the sequence of the NL4.3 virus used in
-\cite{watts_architecture_2009}. Problematic assignments in indel-rich
+\citep{watts_architecture_2009}. Problematic assignments in indel-rich
 regions were excluded from the analysis. The variable loops and flanking
 regions were identified manually starting from the annotated reference HXB2
-sequence from the LANL HIV database~\cite{LANL2012}.
+sequence from the LANL HIV database~\citep{LANL2012}.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Computer simulations}
+\subsection{Computer simulations}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 Computer simulations were performed using FFPopSim
-\cite{zanini_ffpopsim:_2012}. Briefly, FFPopSim enables individual-based
+\citep{zanini_ffpopsim:_2012}. Briefly, FFPopSim enables individual-based
 simulations where each site in the genome is represented by one bit that can be
@@
 arbitrary fitness functions can be specified. We used a generation time of 1
-day, an outcrossing rate of $r=0.01$ per day \cite{batorsky_estimate_2011,
+day, an outcrossing rate of $r=0.01$ per day \citep{batorsky_estimate_2011,
 neher_recombination_2010}, a mutation rate of $\mu=10^{-5}$
-\cite{mansky_lower_1995, abram_nature_2010} and simulated intrapatient
+\citep{mansky_lower_1995, abram_nature_2010} and simulated intrapatient
 evolution for 6000 days. For simplicity, third positions of every codon were
@@
 probability could not be calculated; those runs were excluded from
-\FIG{simheat}B.
+\FIG{simsfig}.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Methods availability}
+\subsection{Methods availability}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\bibliographystyle{natbib}
 \bibliography{bib}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Figure Legends}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=\linewidth]{fig1.pdf}
-\caption{Time series of frequencies
-of synonymous (A) and nonsynonymous (B) derived alleles in \env, 
-\shankaregion, from patient 10~\cite{shankarappa_consistent_1999}.
-While many nonsynonymous mutations  fix, few synonymous
-mutations do even though they are frequently observed at intermediate
-frequencies. Colors indicate the position of the site along the \shankaregion{} region
-(blue to red). Inset: the fixation probability $\pfix$ of a neutral
-mutation is simply the likelihood that the future common ancestor at this 
-position is currently carrying it, i.e. the mutation frequency $\nu$.}
-\label{fig:aft}
-\end{center}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=\linewidth]{fig2.pdf}
-\caption{Fixation and loss of synonymous mutations.
-Panel A) shows how quickly synonymous mutations are purged from the populations. 
-Specifically, the figure shows the fraction of mutations that are still observed
-after $\Delta t$ days, conditional on being observed in one of the three frequency 
-intervals (different colors). 
-In each frequency interval, the fraction of synonymous
-mutations that ultimately survive is the fixation probability $\pfix$ conditional on the
-initial frequency. The neutral expectation for $\pfix=\nu_0$ is indicated by 
-dashed horizontal lines.
-Panel B) shows the fixation probability of derived synonymous
-alleles as a function of $\nu_0$. Polymorphisms within \shankaregion{} fix less
-often than expected for neutral mutations (indicated by the diagonal line).
-This suppression is not observed in other parts of \env~or for nonsynonymous mutations.
-The horizontal error bars on the abscissa are bin sizes, the vertical ones the
-standard deviation after 100 patient bootstraps of the data. Data from
-refs.~\cite{shankarappa_consistent_1999,liu_selection_2006, bunnik_autologous_2008}.}
-\label{fig:fixp}
-\end{center}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=\linewidth]{fig3.pdf}
-\caption{Permissible synonymous mutations tend to be unpaired.
-Panel A) shows the distribution of SHAPE reactivities among sites at which synonymous 
-mutations fixed (red), sites at which mutations reached frequencies above 15\% but
-were susequently lost (blue), and sites at which no mutations were observed (green) 
-(all categories are restricted to the regions V1-V5$\pm 100$bp).
-Sites at which mutations fixed tend to have higher SHAPE reactivities, corresponding to
-less base pairing, than those at which mutations are lost.
-Sites at which no mutations are observed show an intermediate distribution of SHAPE values.
-Panel B) shows the fixation probability of synonymous mutations in
-\shankaregion{} separately for variable regions V3-V5 and the connecting conserved 
-regions C2-C4 that harbor RNA stems. As expected, the fixation probability is lower
-inside the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
-bunnik_autologous_2008, liu_selection_2006}.}
-\label{fig:SHAPE}
-\end{center}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=\linewidth]{fig4.pdf}
-\caption{Distribution of selection coefficients on synonymous sites. Panel A)
-The depression in $\pfix$ depends on the deleterious effect size 
-of synonymous alleles. This parameter also reduces synonymous
-diversity, measured by the probability of a derived allele to be found at
-intermediate frequencies $P_\text{interm}$ (first inset).
-Panel B) To assess the parameter space that affects synonymous fixation and
-diversity, we run 2400 simulations with random parameters for deleterious effect
-size, fraction of deleterious synonymous sites, average escape rate $\epsilon$
-(color, blue to red corresponds to $10^{-2.5}$ to $10^{-1.5}$ per day), and rate of
-introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
-day). Only simulations that reproduce the synonymous diversity and fixation
-patterns observed in data are shown. These simulations demonstrate that
-deleterious effects are around $-0.002$ and a large fraction of the 
-synonymous mutations needs to be deleterious. As expected, larger
-$s_d$ require larger $\epsilon$. Parameters are chosen
-from prior distributions uniform in logspace as indicated by the red rectangle
-(see methods).}
-\label{fig:simheat}
-\end{center}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Appendix
 \newpage

### 2013-05-17 synmut.tex
Figure labels size, little text

@@
 \usepackage{pslatex}
-%\usepackage[colorlinks,linkcolor=red,citecolor=blue]{hyperref}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
-\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV-1 \env~evolution}
+\newcommand{\Title}{Quantifying selection against synonymous mutations in HIV-1 \env{} evolution}
 \newcommand{\Affiliation}{Max Planck Institute for Developmental Biology, 72076 T\"ubingen, Germany}
 \newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
-%\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage{hyperref}
+\hypersetup{colorlinks,linkcolor=red,citecolor=blue, pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 spread through the population and fix; in the rest of the cases, i.e. with
-probability $1-\nu$, it goes extinct. As illustrated in the inset of
-\FIG{aftsyn}, this is a  consequence of the fact that (i) exactly one
+probability $1-\nu$, it goes extinct. For instance, if a mutation is observed in
+half of the sequences, it will fix with a probability of 50\% (see inset in
+figure \ref{fig:aftsyn}).
+This is a  consequence of the fact that (i) exactly one
 individual in the current population will be the common ancestor of the entire
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Synonymous polymorphisms in \env, \shankaregion, are mostly deleterious}
+\subsection{Many synonymous polymorphisms in \shankaregion{} are deleterious}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 (blue to red). Inset: the fixation probability $\pfix$ of a neutral
-mutation is simply the likelihood that the future common ancestor at this 
-position is currently carrying it, i.e. the mutation frequency $\nu$.}
+mutation that reached 50\% frequency is one half.}
 \label{fig:aft}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Synonymous mutations in \shankaregion{} tend to disrupt conserved RNA stems}
+\subsection{Synonymous mutations in \shankaregion{} tend to disrupt RNA stems}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2013-05-17 synmut.tex
Some aesthetic stuff from Richard which I had forgotten

@@
 variable regions form stems.  We aligned the within-patient sequence samples 
-to the reference NL4-3 strain used by \citet{watts_architecture_2009} and 
+to the reference NL4-3 strain used in \citet{watts_architecture_2009} and 
 thereby assigned SHAPE reactivities to most positions in the alignment. 
@@
 nonsynonymous escape mutations are unconditionally beneficial, they almost
-always fix once they reach high frequencies -- $A_{\mathrm{nonsyn}}$ is well
+always fix once they reach high frequencies, i.e., $A_{\mathrm{nonsyn}}$ is well
 above zero. This is incompatible with the blue line in \FIG{fixp2}: in an HIV-1
@@
 benefits are not additive, because each of them is essentially sufficient to
-escape. As a consequence, several escape mutations rise to high frequency
+escape and no additional benefit is gained from combining them.
+As a consequence, several escape mutations rise to high frequency
 rapidly, while the one with the smallest cost in terms of replication,
-packaging, etc. is most likely to eventually fix. The emergence of multiple
+packaging, etc. is most likely to eventually fix, while all others are lost. The emergence of multiple
 sweeping nonsynonymous mutations in real HIV-1 infections has been shown

### 2013-05-26 synmut.tex
Test changes, still not done with cites

@@
 \newcommand{\PCApat}{1}
+\newcommand{\syndiv}{2}
 \newcommand{\timedependence}{3}
@@
 the variable loops of gp120 are more likely to be lost than other synonymous
-changes, hinting at a direct fitness effect of these stem-loop structures in the
-HIV-1 RNA.
+changes, hinting at a direct fitness effect of these RNA hairpins in HIV-1.
 Computational modeling indicates that these synonymous mutations have a
-(Malthusian) selection coefficient of the order of $-0.002$ and that they are
+selection coefficient of the order of $-0.002$ and that they are
 brought up to high frequency by hitchhiking on neighboring beneficial
@@
 nAbs. Escape mutations in \env, the gene encoding gp120, spread through the
-population within a few months.
+viral population within a few months.
 Consistent with this time scale, it is found that serum from a particular time
@@
 
-Escape mutations are selected because they change the amino acid sequence
-of viral proteins in a way that reduces antibody binding or epitope presentation. 
-Conversely, synonymous mutations are commonly used as
-approximately neutral markers in studies of viral evolution. Neutral markers are
-very useful since their dynamics can be compared to that of putatively
-functional sites to detect purifying or directional selection
+Escape mutations are selected because they change the amino acid sequence of
+viral proteins in a way that reduces antibody binding or epitope presentation.
+Conversely, synonymous mutations do not modify the viral protein and are
+commonly used as approximately neutral markers in studies of viral evolution.
+Neutral markers are very useful since their dynamics can be compared to that of
+putatively functional sites to detect purifying or directional selection
 \citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In addition to
@@
 nuclear export, and packaging into the viral capsid: all these processes operate
-at the RNA level and are sensitive to synonymous changes since these processes
-often depend on RNA folding. For example, the HIV-1 \rev{} response element (RRE)
-in \env{} enhances nuclear export of full length or partially spliced viral
-transcripts via a complex stem-loop RNA structure~\citep{fernandes_hiv-1_2012}.
-Another well studied case is the interaction between viral reverse
-transcriptase, viral ssRNA, and the host tRNA$^\text{Lys3}$: the latter is
-required for priming reverse transcription (RT) and is bound by a pseudoknotted
-RNA structure in the viral 5' untranslated region~\citep{barat_interaction_1991,
-paillart_vitro_2002}.
+at the RNA level and are sensitive to synonymous changes. For example, the HIV-1
+\rev{} response element (RRE) in \env{} enhances nuclear export of full length
+or partially spliced viral transcripts via a complex hairpin RNA
+structure~\citep{fernandes_hiv-1_2012}. Another well studied case is the
+interaction between viral reverse transcriptase, viral ssRNA, and the host
+tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription
+(RT) and is bound by a pseudoknotted RNA structure in the viral 5' untranslated
+region~\citep{barat_interaction_1991, paillart_vitro_2002}.
 
 Even in the absence of important RNA structures, synonymous codons do not evolve
-completely neutrally. Some codons are favored over others in many species
+completely neutrally. Some codons are favored in many species
 \citep{plotkin_synonymous_2011}. Recent studies have shown that genetically
-engineered HIV-1 strains with altered codon usage can in some cases produce more
-viral protein, but in general replicate less efficiently
-\citep{ngumbela_quantitative_2008, li_codon-usage-based_2012,
-keating_rich_2009}. Codon deoptimization has been suggested as an attenuation
-strategy for polio and influenza~\citep{mueller_live_2010,coleman_virus_2008}.
-Purifying selection beyond the protein sequence is therefore expected
+engineered HIV-1 strains with altered codon usage show differences in gene
+expression and replication capacity \citep{ngumbela_quantitative_2008,
+li_codon-usage-based_2012, keating_rich_2009}. Codon deoptimization has been
+suggested as an attenuation strategy for polio and
+influenza~\citep{mueller_live_2010,coleman_virus_2008}. Purifying selection
+beyond the protein sequence is therefore expected
 \citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}, and it has been shown that
 rates of evolution at synonymous sites vary along the HIV-1 genome
-\citep{mayrose_towards_2007}. Positive selection through the host adaptive
-immune system, however, is restricted to changes in the amino acid sequence.
+\citep{mayrose_towards_2007} (see also \figurename~S\syndiv). Positive
+selection through the host adaptive immune system, however, is restricted to
+changes in the amino acid sequence.
 
 In this paper, we characterize the dynamics of synonymous mutations in \env{}
-and show that a substantial fraction of these mutations is deleterious.  We
-argue that synonymous mutations reach high frequencies via genetic hitchhiking
-due to limited recombination in HIV-1 populations~\citep{neher_recombination_2010,
-batorsky_estimate_2011}. We then compare our observations to computational
-models of HIV-1 evolution and derive estimates for the effect of synonymous mutations
- on fitness.  Extending the analysis of fixation probabilities to the
-nonsynonymous mutations, we show that time-dependent selection or strong
-competition of escape mutations inside the same epitope are necessary to explain
-the observed patterns of fixation and loss.
+and show that, in the region of the V loops, a large fraction of these mutations
+is deleterious. We argue that deleterious synonymous mutations
+spread in the viral population via genetic hitchhiking due to limited recombination in HIV-1
+populations~\citep{neher_recombination_2010, batorsky_estimate_2011}. We show a
+strong correlation between the fate of a synonymous mutation and the surrounding RNA
+structure, suggesting an important role of RNA hairpins around V-loops
+in HIV-1 evolution. We then compare our observations to computational models and derive
+estimates for the effect of synonymous mutations on viral fitness. Extending the
+analysis of fixation probabilities to the nonsynonymous mutations, we show that
+time-dependent selection or strong competition of escape mutations inside the
+same epitope are necessary to explain the observed patterns of fixation and
+loss.
 
@@
 figure \ref{fig:aftsyn}).
-This is a  consequence of the fact that (i) exactly one
-individual in the current population will be the common ancestor of the entire
-future population at a particular site and (ii) this ancestor has a probability
-$\nu$ of carrying the mutation (assuming the neutral mutation is not
-preferentially associated with genomes of high or low fitness). Deleterious or
+%This is a  consequence of the fact that (i) exactly one
+%individual in the current population will be the common ancestor of the entire
+%future population at a particular site and (ii) this ancestor has a probability
+%$\nu$ of carrying the mutation (assuming the neutral mutation is not
+%preferentially associated with genomes of high or low fitness).
+Deleterious or
 beneficial mutations fix less or more often than neutral ones, respectively.
 \FIG{aft} shows the time course of the frequencies of all synonymous and
-nonsynonymous mutations observed in \env, \shankaregion, in patient
-p10~\citep{shankarappa_consistent_1999}. Despite many synonymous
+nonsynonymous viral mutations observed in \env, in \shankaregion, in a chronically
+HIV-1 infected patient (p10 from\citet{shankarappa_consistent_1999}). Despite many synonymous
 mutations reaching high frequency, few fix (panel~\ref{fig:aftsyn}); in
@@
 \citet{shankarappa_consistent_1999} and \citep{liu_selection_2006} as well as three patients from
-\citet{bunnik_autologous_2008} (patients whose viral population was structured
-were excluded from the analysis; see methods and \figurename~S\PCApat).  The
+\citet{bunnik_autologous_2008} (patients with strong viral population structure
+were excluded from the analysis, because it is impossible to measure
+frequencies; see methods and \figurename~S\PCApat). The
 former data set is restricted to the \shankaregion{} region of \env, while
-the data from \citet{bunnik_autologous_2008} cover the majority of \env.
-Considering all mutations in a frequency interval $[\nu_0-\delta\nu, \nu_0+\delta\nu]$ at some time
-$t$, we calculate the fraction that are still observed at later times $t+\Delta t$. Plotting this fraction against the time interval $\Delta t$, we see that
+the data from \citet{bunnik_autologous_2008} cover most of \env.
+We consider all mutations in a frequency interval $[\nu_0-\delta\nu, \nu_0+\delta\nu]$ at some time
+$t$, and calculate the fraction that are still observed at later times $t+\Delta t$. Plotting this fraction against the time interval $\Delta t$, we see that
 most synonymous mutations segregate for roughly one year and are lost much more
@@
 fixation, $\pfix$, is shown as a function of the
-initial frequency $\nu_0$ in panel~\ref{fig:fixp2} (red line). 
+initial frequency $\nu_0$ in panel~\ref{fig:fixp2}. 
 We find that $\pfix$ of synonymous mutations is far below
-the neutral expectation in  \shankaregion.  Outside of \shankaregion, using data from
-\citet{bunnik_autologous_2008} only, we find no such reduction in $\pfix$.
+the neutral expectation in  \shankaregion (red line).  Outside of \shankaregion, using data from
+\citet{bunnik_autologous_2008} only, we find no such reduction in $\pfix$ (green
+line).
 Restricted to the \shankaregion{} region, the sequence samples from
@@
 to the reference NL4-3 strain used in \citet{watts_architecture_2009} and 
-thereby assigned SHAPE reactivities to most positions in the alignment. 
+assigned SHAPE reactivities to most positions in the alignment. 
 We then calculated the distributions of SHAPE reactivities for synonymous 
@@
 other words, alleles that are likely to break RNA helices are also more likely
-to revert and finally be lost from the population. The average over all
+to revert and finally be lost from the population (restoring the helix). The average over all
 mutations that are not observed (green) lies between those that fix and
@@
 To test the hypothesis that mutations in \shankaregion{} are lost because they
-break stems in the conserved stretches between the variable loops, we consider
-mutations in variable loops and conserved parts separately. The greatest
+break stems in the conserved stretches between the V loops, we consider
+mutations in V loops and conserved parts separately. The greatest
 depression in fixation probability is observed in the conserved stems, while the
-variable loops show little deviation from the neutral signature, see
-\FIG{SHAPEB}. This is consistent with important stem structures in conserved
+V loops show little deviation from the neutral signature, see
+\FIG{SHAPEB}. This is suggestive of important RNA helices in conserved
 regions between loops.
@@
 \begin{center}
-\subfloat{\includegraphics[height=0.47\linewidth]{reactivities_histograms_syn.pdf}\label{fig:SHAPEA}}
-\subfloat{\includegraphics[height=0.46\linewidth]{fixation_probabilities_VnonV.pdf}\label{fig:SHAPEB}}
+\subfloat{\includegraphics[height=0.50\linewidth]{reactivities_histograms_syn.pdf}\label{fig:SHAPEA}}
+\subfloat{\includegraphics[height=0.49\linewidth]{fixation_probabilities_VnonV.pdf}\label{fig:SHAPEB}}
 \caption{Permissible synonymous mutations tend to be unpaired.
@@
 demonstrates how selection at synonymous sites can alter genetic diversity and
-dynamics. The observed hitchhiking highlights the importance of linkage due to
+dynamics. %FIXME insert citations and more discussion
+
+The observed hitchhiking highlights the importance of linkage due to
 infrequent recombination for the evolution of HIV-1

### 2013-05-29 synmut.tex
More discussion on RNA stuff + citations

@@
 \citep{watts_architecture_2009, sanjuan_interplay_2011} and conserved RNA
-structures exist in different parts of the HIV-1 genome. Since there are
-of course many ways to build an RNA stem in a particular location, we do
-not necessarily expect a strong signal of conservation in cross-sectional data. 
-Our analysis, however, is able to
-quantify the fitness effect of RNA structure within single infections and
-demonstrates how selection at synonymous sites can alter genetic diversity and
-dynamics. %FIXME insert citations and more discussion
+structures exist in different parts of the HIV-1 genome. Since there are of
+course many ways to build an RNA hairpin at a particular location, we do not
+necessarily expect a strong signal of conservation in cross-sectional data.
+Consistently, in a characterization of the SIV RNA structures via the SHAPE
+assay, it was recently shown that only the general pattern is shared with HIV,
+but the single base pairs are often discordant between the two viruses
+\citep{pollom_comparison_2013}. In fact, for each pairing-disrupting mutation in
+our dataset, we have looked for compensatory mutations at the partner site, but
+observed no effect. For these reasons, we do not believe that exact RNA pairings
+are evolutionary relevant, rather only the general architecture of different
+genomic regions (e.g. the V loops). Despite this flexibility in pairing, our
+analysis is able to quantify the fitness effect of RNA structure within single
+infections and demonstrates how selection at synonymous sites can alter genetic
+diversity and dynamics.
+
+In principle, the fitness effect of RNA structures in terms of replication
+capacity could be tested culturing \textit{in vitro} viruses with engineered
+modifications. This has been attempted recently, but no major fitness effect was
+measured \citep{knoepfel_role_2013}. There is no contradiction with our results,
+because we infer a selection coefficient for each mutation of the order of
+$-0.002$, which is only visible after hundreds of generations. Shorter
+experimental tests cannot detect such a subtle effect; our long-term
+longitudinal analysis is more sensitive in this respect.
 
@@
 epitope implies a large effective population size that explores all necessary point
-mutations rapidly. A similar point has been made recently by Boltz {\it et al.}
+mutations rapidly. A similar point has been made recently by several authors
 in the context of preexisting drug resistance mutations
-\citep{boltz_ultrasensitive_2012}.
+\citep{boltz_ultrasensitive_2012, pennings_loss_2013}.
 

### 2013-05-31 supplement_body.tex
Implemented Richard's suggestions (including 'mutation' -> SNV)

--- 2013-03-04 supplement_body.tex
@@
 position of the allele in \env{} from 5' to 3'. Panels C) and D) show analogous
-plots for data from patient p7. Samples after day 1000 split into two clusters in the PCA and no mutations that arise after day 1000 fix, presumably because they are restricted
+plots for data from patient p7. Samples after day 1000 split into two clusters
+in the PCA and no single nucleotide variants (SNVs) that arise after day 1000 fix, presumably because they are restricted
 to one subpopulation. All patients like p7 (p4, p7, p8, p9 from ref.~\citealp{shankarappa_consistent_1999} and
@@
 \caption{
-Time-dependent selection reduces fixation of nonsynonymous mutations. The figure
+Time-dependent selection reduces fixation of nonsynonymous SNVs. The figure
 compares the fixation probability in the time independent model (na\"ive) to
@@
 probability of recognition, the fixation of frequent escape mutants is reduced,
-while hitch-hiking of synonymous mutations is not affected. The precise
+while hitch-hiking of synonymous SNVs is not affected. The precise
 shape of $\pfix(\nu)$ depends on the details of the $P_\text{recognized}(t)$, and 
@@
 \caption{
-Competition between escape mutations in the same epitope reduces fixation of
-nonsynonymous mutations. The figure compares the fixation probability of models
+Competition between escape SNVs in the same epitope reduces fixation of
+nonsynonymous SNVs. The figure compares the fixation probability of models
 with one, three, or six mutually exclusive escape mutations within the
@@
 $n$ is the number of competing escapes in the epitope. 
-In this evolutionary scenario, many escape mutations start to sweep on different backgrounds within the viral population, but eventually
+In this evolutionary scenario, many escape SNVs start to sweep on different backgrounds within the viral population, but eventually
 compete and only one of them fixes. The other parameters for the shown simulations are

### 2013-05-31 synmut.tex
Implemented Richard's suggestions (including 'mutation' -> SNV)

--- 2013-05-26 synmut.tex
@@
 \noindent
-Intrapatient HIV-1 evolution is dominated by selection on the protein level in the
-arms race with the adaptive immune system. When cytotoxic CD8${}^+$ T-cells or 
-neutralizing antibodies
-target a new epitope, the virus often escapes via nonsynonymous mutations that
-impair recognition. Synonymous mutations do not affect this interplay and are
-often assumed to be neutral.
-We analyze longitudinal intrapatient data from the \shankaregion{} part of the
-envelope gene (\env{}) and observe that synonymous derived alleles rarely
-fix even though they often reach high frequencies in the viral population.
-We find that synonymous mutations that disrupt base pairs in RNA stems flanking
-the variable loops of gp120 are more likely to be lost than other synonymous
-changes, hinting at a direct fitness effect of these RNA hairpins in HIV-1.
+Intrapatient HIV-1 evolution is dominated by selection on the protein level in
+the arms race with the adaptive immune system. When cytotoxic CD8${}^+$ T-cells
+or neutralizing antibodies target a new epitope, the virus often escapes via
+nonsynonymous mutations that impair recognition. Synonymous mutations do not
+affect this interplay and are often assumed to be neutral. We analyze
+longitudinal intrapatient data from the \shankaregion{} part of the envelope
+gene (\env{}) and observe that synonymous mutations rarely spread even though
+they often reach high frequencies in the viral population. Using published data
+from the SHAPE assay, we find that synonymous mutations that disrupt base pairs
+in RNA stems flanking the V loops of gp120 are more likely to be lost than other
+synonymous changes, suggesting a function of these RNA hairpins in HIV-1.
 Computational modeling indicates that these synonymous mutations have a
-selection coefficient of the order of $-0.002$ and that they are
-brought up to high frequency by hitchhiking on neighboring beneficial
-nonsynonymous alleles.
-The patterns of fixation of nonsynonymous mutations estimated from the
-longitudinal data and comparisons with computer models
-suggest that escape mutations in \shankaregion{} are only transiently
-beneficial, either because the immune system is catching up or because of
-competition between equivalent escapes.
-
+selection coefficient of the order of $-0.002$ and that they are brought up to
+high frequency by genetic hitchhiking on neighboring beneficial variants. We
+conclude that, contrary to common assumptions, synonymous mutations in the V
+loops region are neither independent of the escape patterns nor neutral; further
+studies are needed to clarify the function of the RNA hairpins found there.
 \end{abstract}
@@
 reduce or prevent recognition of the epitope frequently emerge. Escape mutations
-in epitopes targeted by CTLs typically evolve during early infection and spread
+in epitopes targeted by CTLs typically emerge during early infection and spread
 rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
@@
 commonly used as approximately neutral markers in studies of viral evolution.
-Neutral markers are very useful since their dynamics can be compared to that of
-putatively functional sites to detect purifying or directional selection
-\citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In addition to
+Neutral markers are very useful as a negative control for detecting selected
+sites \citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In addition to
 maintaining protein function and avoiding the adaptive immune recognition,
@@
 \rev{} response element (RRE) in \env{} enhances nuclear export of full length
-or partially spliced viral transcripts via a complex hairpin RNA
-structure~\citep{fernandes_hiv-1_2012}. Another well studied case is the
-interaction between viral reverse transcriptase, viral ssRNA, and the host
+or partially spliced viral transcripts via a complex hairpin RNA structure
+\citep{fernandes_hiv-1_2012}. Another well studied case is the interaction
+between viral reverse transcriptase, viral ssRNA, and the host
 tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription
 (RT) and is bound by a pseudoknotted RNA structure in the viral 5' untranslated
-region~\citep{barat_interaction_1991, paillart_vitro_2002}.
+region~\citep{barat_interaction_1991, paillart_vitro_2002}. These RNA
+structures, but not the exact base pairs, are well conserved between HIV-1 and
+SIV \citep{pollom_comparison_2013}.
 
@@
 and show that, in the region of the V loops, a large fraction of these mutations
-is deleterious. We argue that deleterious synonymous mutations
-spread in the viral population via genetic hitchhiking due to limited recombination in HIV-1
-populations~\citep{neher_recombination_2010, batorsky_estimate_2011}. We show a
-strong correlation between the fate of a synonymous mutation and the surrounding RNA
-structure, suggesting an important role of RNA hairpins around V-loops
-in HIV-1 evolution. We then compare our observations to computational models and derive
-estimates for the effect of synonymous mutations on viral fitness. Extending the
-analysis of fixation probabilities to the nonsynonymous mutations, we show that
-time-dependent selection or strong competition of escape mutations inside the
-same epitope are necessary to explain the observed patterns of fixation and
-loss.
+is deleterious. Contrary to common assumptions, deleterious synonymous mutations
+rise in frequency in the viral population via genetic hitchhiking due to limited
+recombination in HIV-1 populations~\citep{neher_recombination_2010,
+batorsky_estimate_2011}. We show a strong correlation between the fate of a
+synonymous mutation and the surrounding RNA structure, suggesting an important
+role of RNA hairpins around V-loops in HIV-1 evolution. We then compare our
+observations to computational models and derive estimates for the effect of
+synonymous mutations on viral fitness. Extending the analysis of fixation
+probabilities to the nonsynonymous mutations, we show that time-dependent
+selection or strong competition of escape mutations inside the same epitope are
+necessary to explain the observed patterns of fixation and loss.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-The central quantity we investigate is the probability of fixation of a
-mutation, conditional on its population frequency. A neutral mutation
-segregating at frequency $\nu$ has a probability $\pfix(\nu) = \nu$ to
-spread through the population and fix; in the rest of the cases, i.e. with
-probability $1-\nu$, it goes extinct. For instance, if a mutation is observed in
-half of the sequences, it will fix with a probability of 50\% (see inset in
-figure \ref{fig:aftsyn}).
-%This is a  consequence of the fact that (i) exactly one
-%individual in the current population will be the common ancestor of the entire
-%future population at a particular site and (ii) this ancestor has a probability
-%$\nu$ of carrying the mutation (assuming the neutral mutation is not
-%preferentially associated with genomes of high or low fitness).
-Deleterious or
-beneficial mutations fix less or more often than neutral ones, respectively.
-\FIG{aft} shows the time course of the frequencies of all synonymous and
-nonsynonymous viral mutations observed in \env, in \shankaregion, in a chronically
-HIV-1 infected patient (p10 from\citet{shankarappa_consistent_1999}). Despite many synonymous
-mutations reaching high frequency, few fix (panel~\ref{fig:aftsyn}); in
-constrast, many nonsynonymous mutations fix (panel~\ref{fig:aftnonsyn}).
-Strictly speaking, no mutation in the HIV-1 population ever fixes because the 
-mutation rate and the population size are large. Therefore, we define ``fixation'' or ``loss''
-by not observing the mutation in the sample.
+In HIV-1, because of the high mutation rate, single nucleotide variants (SNVs)
+are continuously injected into the viral population. We say that an SNV has a
+frequency $\nu$ at a certain time if it is found in a fraction $\nu$, between
+zero and one, of the sequences from that time point. An SNV rises or falls in
+frequency by virtue of both (i) its own effect on fitness and escape and (ii)
+the effect of its genetic background. When an SNVs is present in all sequences
+at a certain time point, we say it has ``fixed''; when it is completely absent,
+we say it was ``lost'' or is ``extinct''. Eventually, all SNVs will fix or be lost; the
+probability of fixation is higher for beneficial SNVs, and lower for deleterious
+ones. The intermediate case are neutral SNVs: a neutral mutation segregating at
+frequency $\nu$ has a probability $\pfix(\nu) = \nu$ to spread through the
+population and fix; in the rest of the cases it goes extinct. For instance, if
+a neutral SNV is observed in half of the sequences, it will fix with a probability of
+50\% (see inset in figure \ref{fig:aftsyn}).
+
+The problem with the classification based on beneficial/neutral/deleterious is
+that, given an SNV, we do not know its class \textit{a priori}. Synonymous SNVs,
+however, are usually assumed to be neutral. \FIG{aft} shows the time course of
+the frequencies of all synonymous and nonsynonymous viral mutations observed in
+\env, in \shankaregion, in a chronically HIV-1 infected patient (p10
+from \citet{shankarappa_consistent_1999}). Despite many synonymous SNVs
+reaching high frequency, very few fix (panel~\ref{fig:aftsyn}); in constrast, many
+nonsynonymous mutations fix (panel~\ref{fig:aftnonsyn}). This basic observation
+seems at odds with the assumption of neutrality.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-We study the dynamics and fate of synonymous mutations more quantitatively by
-analyzing data from seven patients from
-\citet{shankarappa_consistent_1999} and \citep{liu_selection_2006} as well as three patients from
+We study the dynamics and fate of synonymous SNVs more quantitatively by
+analyzing data from seven patients from \citet{shankarappa_consistent_1999} and
+\citep{liu_selection_2006} as well as three patients from
 \citet{bunnik_autologous_2008} (patients with strong viral population structure
-were excluded from the analysis, because it is impossible to measure
-frequencies; see methods and \figurename~S\PCApat). The
-former data set is restricted to the \shankaregion{} region of \env, while
-the data from \citet{bunnik_autologous_2008} cover most of \env.
-We consider all mutations in a frequency interval $[\nu_0-\delta\nu, \nu_0+\delta\nu]$ at some time
-$t$, and calculate the fraction that are still observed at later times $t+\Delta t$. Plotting this fraction against the time interval $\Delta t$, we see that
-most synonymous mutations segregate for roughly one year and are lost much more
-frequently than expected (panel \ref{fig:fixp1}). The long-time probability of
-fixation, $\pfix$, is shown as a function of the
-initial frequency $\nu_0$ in panel~\ref{fig:fixp2}. 
-We find that $\pfix$ of synonymous mutations is far below
-the neutral expectation in  \shankaregion (red line).  Outside of \shankaregion, using data from
-\citet{bunnik_autologous_2008} only, we find no such reduction in $\pfix$ (green
-line).
-Restricted to the \shankaregion{} region, the sequence samples from
-\citet{bunnik_autologous_2008} are fully compatible with data from
-\citet{shankarappa_consistent_1999}. The nonsynonymous mutations seem to follow
-more or less the neutral expectation (blue line) -- a point to which we will
-come back below.
+were excluded from the analysis; see methods and \figurename~S\PCApat). The
+former data set is restricted to the \shankaregion{} region of \env, while the
+data from \citet{bunnik_autologous_2008} cover most of \env.  We consider all
+SNVs in a frequency interval $[\nu_0-\delta\nu, \nu_0+\delta\nu]$ at some time
+$t$, and calculate the fraction that are still observed at later times $t+\Delta
+t$. Plotting this fraction against the time interval $\Delta t$, we see that
+most synonymous SNVs segregate for roughly one year and are lost much more
+frequently than expected under neutrality (panel \ref{fig:fixp1}). The long-time
+probability of fixation, $\pfix$, is shown as a function of the initial
+frequency $\nu_0$ in panel \ref{fig:fixp2}. We find that $\pfix$ of synonymous
+variants is far below the neutral expectation in \shankaregion (red line).
+Outside of \shankaregion, using data from \citet{bunnik_autologous_2008} only,
+we find no such reduction in $\pfix$ (green line). Restricted to the
+\shankaregion{} region, the sequence samples from \citet{bunnik_autologous_2008}
+are fully compatible with data from \citet{shankarappa_consistent_1999}. The
+nonsynonymous SNVs seem to follow more or less the neutral expectation (blue
+line) -- a point to which we will come back below.
 
@@
 \caption{Time series of frequencies
-of synonymous (A) and nonsynonymous (B) derived alleles in \env, 
-\shankaregion, from patient 10~\cite{shankarappa_consistent_1999}.
-While many nonsynonymous mutations  fix, few synonymous
-mutations do even though they are frequently observed at intermediate
+of synonymous (A) and nonsynonymous (B) single nucleotide variants (SNVs) in \env, 
+\shankaregion, from patient p10~\cite{shankarappa_consistent_1999}.
+While many nonsynonymous SNVs fix, few synonymous
+SNVs do so even though they are frequently observed at high
 frequencies. Colors indicate the position of the site along the \shankaregion{} region
 (blue to red). Inset: the fixation probability $\pfix$ of a neutral
-mutation that reached 50\% frequency is one half.}
+SNV that reached 50\% frequency is one half.}
 \label{fig:aft}
@@
 When interpreting these results for the fixation probabilities, it is important
-to distinguish between random mutations and polymorphisms observed at a certain
-frequency since the latter have already been filtered by selection.
-A polymorphism could be beneficial to the virus and on its way to fixation. In
-this case, we expect that it fixes almost surely given that we see it at high
-frequency. If, on the other hand, the polymorphism is deleterious it must have
-reached a high frequency by chance (genetic drift or hitchhiking), and
-we expect that selection drives it out of the population again. Hence our
-observations suggest that many of the synonymous polymorphisms at intermediate
-frequencies in the part of \env{} that includes \shankaregion{} are
-deleterious, while outside this region most polymorphisms are roughly
-neutral. Note that this does not imply that all synonymous mutations outside 
-\shankaregion{} are neutral -- only those mutations observed at high frequencies, which
-have experienced selection for some time, tend to be neutral.
+to note that we focus on SNVs that have already reached high frequencies. In
+HIV-1 infection, most SNVs remain very rare throughout; they are not considered
+here. Synonymous SNVs can reach high frequencies by two means, either genetic
+drift or genetic hitchhiking on escape variants (see below); very deleterious
+variants will never reach high frequencies in the first place. Hence, our
+analysis indicates that, among all synonymous SNVs that somehow reach high
+frequencies, most of them are deleterious in \shankaregion{}, while they tend to
+be neutral in the rest of \env{}.
 
@@
 \label{fig:fixp2}}
-\caption{Fixation and loss of synonymous mutations.
-Panel A) shows how quickly synonymous mutations are purged from the populations. 
-Specifically, the figure shows the fraction of mutations that are still observed
+\caption{Fixation and loss of SNVs.
+Panel A) shows how quickly synonymous SNVs are purged from the populations. 
+Specifically, the figure shows the fraction of SNVs that are still observed
 after $\Delta t$ days, conditional on being observed in one of the three frequency 
@@
 In each frequency interval, the fraction of synonymous
-mutations that ultimately survive is the fixation probability $\pfix$ conditional on the
+SNVs that ultimately survive is the fixation probability $\pfix$ conditional on the
 initial frequency. The neutral expectation for $\pfix=\nu_0$ is indicated by 
 dashed horizontal lines.
-Panel B) shows the fixation probability of derived synonymous
-alleles as a function of $\nu_0$. Polymorphisms within \shankaregion{} fix less
-often than expected for neutral mutations (indicated by the diagonal line).
-This suppression is not observed in other parts of \env~or for nonsynonymous mutations.
+Panel B) shows the fixation probability of synonymous SNVs as a function of $\nu_0$. Polymorphisms within \shankaregion{} fix less
+often than expected for neutral SNVs (indicated by the diagonal line).
+This suppression is not observed in other parts of \env{} or for nonsynonymous
+SNVs.
 The horizontal error bars on the abscissa are bin sizes, the vertical ones the
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-One possible explanation for lack of fixation of synonymous mutations in
+One possible explanation for lack of fixation of synonymous SNVs in
 \shankaregion{} is secondary structures in the viral RNA, the disruption of which
@@
 regions V1-V5 tend to be unpaired, while the conserved regions between those
-variable regions form stems.  We aligned the within-patient sequence samples 
-to the reference NL4-3 strain used in \citet{watts_architecture_2009} and 
-assigned SHAPE reactivities to most positions in the alignment. 
-We then calculated the distributions of SHAPE reactivities for synonymous 
-polymorphisms that fixed or were subsequently lost (only polymorphisms with 
-frequencies above 15\%).
-As shown in \FIG{SHAPEA}, the reactivities of fixed alleles (red
-histogram) are systematically larger than those of alleles that are lost (blue)
-(Kolmogorov-Smirnov test on the cumulative distribution, $p\approx 0.002$). In
-other words, alleles that are likely to break RNA helices are also more likely
-to revert and finally be lost from the population (restoring the helix). The average over all
-mutations that are not observed (green) lies between those that fix and
-those that get lost. Note that this analysis will be sensitive only at positions
-where the base pairing pattern of NL4-3 agrees with that of each patient's
-initial consensus sequence (it is thus statistically conservative).
+variable regions form stems. We aligned the within-patient sequence samples to
+the reference NL4-3 strain used in \citet{watts_architecture_2009} and assigned
+SHAPE reactivities to most positions in the alignment. We then calculated the
+distributions of SHAPE reactivities for synonymous SNVs that fixed or were
+subsequently lost (only SNVs with frequencies above 15\%). As shown in
+\FIG{SHAPEA}, the reactivities of fixed SNVs (red histogram) are systematically
+larger than those of lost SNVs (blue) (Kolmogorov-Smirnov test on the cumulative
+distribution, $p\approx 0.002$). In other words, SNVs that are likely to
+break RNA helices are also more likely to revert and finally be lost from the
+population, restoring the helix. Note that this analysis will be sensitive only
+at positions where the base pairing pattern of NL4-3 agrees with that of each
+patient's initial consensus sequence -- it is thus statistically conservative.
+As a control, we also calculate the distribution of SHAPE reactivities for SNVs
+that never reach high frequencies (green). This set is a mixture of neutral and
+deleterious SNVs and, as expected, their distribution lies between those of
+fixed and lost high-frequency SNVs.
 
-To test the hypothesis that mutations in \shankaregion{} are lost because they
+To test the hypothesis that synonymous SNVs in \shankaregion{} are lost because they
 break stems in the conserved stretches between the V loops, we consider
-mutations in V loops and conserved parts separately. The greatest
+separately SNVs in V loops and conserved flanks. The greatest
 depression in fixation probability is observed in the conserved stems, while the
@@
 \FIG{SHAPEB}. This is suggestive of important RNA helices in conserved
-regions between loops.
+regions between the V loops.
 
 In addition to RNA secondary structure, we have considered other possible
-explanations for a fitness cost of some synonymous mutations, in particular codon
-usage bias (CUB). HIV-1 is known to prefer A-rich codons over highly expressed
-human codons~\citep{jenkins_extent_2003,kuyl_biased_2012}. We do not find,
-however, any evidence for a contribution of average CUB to the ultimate fate of
-synonymous alleles; consistently, HIV-1 does not seem to adapt its codon usage to
-its human host cells at the macroevolutionary level \citep{kuyl_biased_2012}.
+explanations for a fitness cost of some synonymous mutations, in particular
+codon usage bias (CUB). HIV-1 is known to prefer A-rich codons over highly
+expressed human codons \citep{jenkins_extent_2003, kuyl_biased_2012}. We do not
+find, however, any evidence for a contribution of average CUB to the ultimate
+fate of synonymous SNVs; this agrees with the observation that HIV-1 is not
+adapting its codon usage to its human host cells at the macroevolutionary level
+\citep{kuyl_biased_2012}.
 
@@
 Panel A) shows the distribution of SHAPE reactivities among sites at which synonymous 
-mutations fixed (red), sites at which mutations reached frequencies above 15\% but
-were susequently lost (blue), and sites at which no mutations were observed (green) 
+SNVs fixed (red), sites at which SNVs reached frequencies above 15\% but
+were susequently lost (blue), and sites at which no high-frequency SNVs were observed (green) 
 (all categories are restricted to the regions V1-V5$\pm 100$bp).
-Sites at which mutations fixed tend to have higher SHAPE reactivities, corresponding to
-less base pairing, than those at which mutations are lost.
-Sites at which no mutations are observed show an intermediate distribution of SHAPE values.
-Panel B) shows the fixation probability of synonymous mutations in
+Sites at which SNVs fixed tend to have higher SHAPE reactivities, corresponding to
+less base pairing, than those at which SNVs are lost.
+Sites at which no SNVs are observed show an intermediate distribution of SHAPE values.
+Panel B) shows the fixation probability of synonymous SNVs in
 \shankaregion{} separately for variable regions V3-V5 and the connecting conserved 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Deleterious mutations are brought to high frequency by hitchhiking}
+\subsection{Deleterious SNVs reach high frequency by hitchhiking}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-While the observation that some fraction of synonymous mutations is deleterious
+While the observation that some fraction of synonymous SNVs is deleterious
 is not unexpected, it seems odd that we observe them at high population
@@
 richman_rapid_2003}. Due to the limited amount of recombination in HIV-1
-\cite{neher_recombination_2010, batorsky_estimate_2011}, deleterious mutations
+\cite{neher_recombination_2010, batorsky_estimate_2011}, deleterious SNVs
 that are linked to adaptive variants can reach high frequency. This process is
@@
 \citep{gillespie_genetic_2000,neher_genetic_2011}. Hitchhiking is apparent in
-\FIG{aft}, which shows that many mutations change rapidly in frequency as a
+\FIG{aft}, which shows that many SNVs change rapidly in frequency as a
 flock.
@@
 \FIG{fixp1}, which shows the distribution of times after which synonymous
-alleles at intermediate frequencies become fixed or lost. The typical time to
+SNVs at intermediate frequencies become fixed or lost. The typical time to
 loss is of the order of 500 days. If this loss is driven by the deleterious
@@
 The depression in $\pfix$ depends on the deleterious effect size 
-of synonymous alleles. This parameter also reduces synonymous
-diversity, measured by the probability of a derived allele to be found at
+of synonymous SNVs. This parameter also reduces synonymous
+diversity, measured by the probability of a SNV to be found at
 intermediate frequencies $P_\text{interm}$ (first inset).
@@
 and rare recombination.  For this purpose, we use the simulation package
-FFPopSim, which includes a module dedicated to intrapatient HIV
-evolution~\citep{zanini_ffpopsim:_2012}. For each simulation run, we specify the
+FFPopSim, which includes a module dedicated to intrapatient HIV evolution
+\citep{zanini_ffpopsim:_2012}. For each simulation run, we specify the
 deleterious effect of synonymous mutations, the fraction of synonymous mutations
@@
 nonsynonymous mutations and the rate at which previously untargeted epitopes
-become targeted (the latter determines the number of sites available for escape). 
-Note that the escape rate is the sum of two
-factors: (i) the beneficial effect due to the ability to evade the immune system
-minus (ii) the fitness cost of the mutation in terms of structure, stability,
-etc. Net escape rates in chronic infections have been estimated to be on the
-order of $\epsilon = 0.01$ per day \citep{neher_recombination_2010,
-Asquith:2006p28003}.
+become targeted (the latter determines the number of sites available for
+escape). Note that the escape rate is the sum of two factors: (i) the beneficial
+effect due to the ability to evade the immune system minus (ii) the fitness cost
+of the mutation in terms of structure, stability, etc. Net escape rates in
+chronic infections have been estimated to be on the order of $\epsilon = 0.01$
+per day \citep{neher_recombination_2010, Asquith:2006p28003}.
 
@@
 We quantify synonymous diversity via $P_\text{interm}$, the fraction of sites
-with an allele at frequency $0.25 < \nu < 0.75$. The synonymous diversity
+with an SNV at frequency $0.25 < \nu < 0.75$. The synonymous diversity
 observed in patient data is indicated in the figure. To quantify the depression
@@
 (\FIG{simfixpvar}, lower inset). If no fixation happens, the area will be
-$-0.5$; if every mutation fixes, the area will be $+0.5$. In HIV-1 infected
+$-0.5$; if every SNV fixes, the area will be $+0.5$. In HIV-1 infected
 patients, we find $P_\text{interm} \approx 0.005$, $A_\text{syn} \approx -0.2$
@@
 changes. In the three simulations shown in \FIG{simfixpvar}, the fixation
-probability of synonymous alleles decreases from the neutral expectation
-($A_\text{syn} \approx 0$) to zero ($A_\text{syn} \approx -0.5$) as their
-fitness cost increases; the synonymous diversity plummets as well, as
-deleterious mutations are selected against.
+probability of synonymous SNVs decreases from the neutral expectation
+($A_\text{syn} \approx 0$) to zero ($A_\text{syn} \approx -0.5$) as the
+fitness cost of the SNVs increases; the synonymous diversity plummets as well, as
+deleterious SNVs are selected against.
 
@@
 that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
-large depression in fixation probability of synonymous mutations but,
-simultaneously, a moderately high synonymous diversity. Specifically,
-\FIG{simsfig} shows parameter combinations for which we found $A_\text{syn} <
--0.15$ and $0.0025 < P_\text{interm} < 0.010$. These conditions indicate that a
-high fraction ($\gtrsim 0.8$) of sites has to be deleterious with effect size
-$|s_d| \sim 0.002$.  This result fits well the expectation based on the
-fixation/extinction times above (see \FIG{fixp1}). The results are plausible:
-(i) a substantial depression in $\pfix$ requires pervasive deleterious
-mutations, otherwise the majority of observed polymorphisms are neutral and no
-depression is observed; (ii) in order to hitchhike, the deleterious effect size
-has to be much smaller than the escape rate, otherwise the double mutant has
-little or no fitness advantage. Consistent with this argument, larger
-deleterious effects in \FIG{simsfig} correspond to larger escape rates; and (iii)
-mutations with a deleterious effect smaller than approximately $0.001$ behave
-neutrally, consistent with the typical coalescent times observed in HIV-1.
+large depression in fixation probability of synonymous SNVs but, simultaneously,
+a moderately high synonymous diversity. Specifically, \FIG{simsfig} shows
+parameter combinations for which we found $A_\text{syn} < -0.15$ and $0.0025 <
+P_\text{interm} < 0.010$. These conditions indicate that a high fraction
+($\gtrsim 0.8$) of sites has to be deleterious with effect size $|s_d| \sim
+0.002$.  This result fits well the expectation based on the fixation/extinction
+times above (see \FIG{fixp1}). The results of simulations are plausible: (i) a
+substantial depression in $\pfix$ requires pervasive deleterious SNVs, otherwise
+the majority of SNVs reaching high frequency are neutral and no depression is
+observed; (ii) in order to hitchhike, the deleterious effect size has to be much
+smaller than the escape rate, otherwise the double mutant (with both the escape
+mutation and the deleterious synonymous one) has little or no fitness advantage
+over the wildtype virus. Consistent with this argument, larger deleterious
+effects in \FIG{simsfig} correspond to larger escape rates; and (iii) SNVs
+with a deleterious effect smaller than approximately $0.001$ behave neutrally,
+consistent with the typical coalescent times observed in HIV-1.
 
 The above simulations show that hitchhiking can explain the observation of
-deleterious mutations that rarely fix. However, in a simple model where
+deleterious SNVs that rarely fix. However, in a simple model where
 nonsynonymous escape mutations are unconditionally beneficial, they almost
@@
 above zero. This is incompatible with the blue line in \FIG{fixp2}: in an HIV-1
-infection, nonsynonymous mutations at high frequency often disappear again, even
+infection, nonsynonymous SNVs at high frequency often disappear again, even
 though many are at least transiently beneficial. Inspecting the trajectories of
-nonsynonymous mutations suggests the rapid rise and fall of many alleles. We
+nonsynonymous SNVs suggests the rapid rise and fall of many SNVs. We
 test two possible mechanisms that are biologically plausible and could explain
-the transient rise of nonsynonymous mutations: time-dependent selection and
+the transient rise of nonsynonymous SNVs: time-dependent selection and
 within-epitope competition.
@@
 support of this idea, in \citet{richman_rapid_2003, bunnik_autologous_2008},
-antibody responses to escape mutants have been reported.
-These responses are delayed by a few
-months, roughly matching the average time needed by an escape mutant to rise
-from low to high frequency. To model this type of behavior, we assume that
-antibody responses against escape mutations arise with a rate proportional to
-the frequency of the escape mutation and abolish the benefit of the escape
-mutations. As expected, this type of time-dependent selection retains the
-potential for hitchhiking, but reduces fixation of nonsynonymous mutations.
-\figurename~S\timedependence~shows that $\pfix$ of synonymous mutations is not
-affected by this change, while $\pfix$ of nonsynonymous mutations approaches
-the diagonal as the rate of recognition of escape mutants is increased.
+antibody responses to escape mutants have been reported. These responses are
+delayed by a few months, roughly matching the average time needed by an escape
+mutant to rise from low to high frequency. To model this type of behavior, we
+assume that antibody responses against escape SNVs arise with a rate
+proportional to the frequency of the escape SNV and abolish the benefit of the
+escape mutations. As expected, this type of time-dependent selection retains the
+potential for hitchhiking, but reduces fixation of nonsynonymous SNVs.
+\figurename~S\timedependence~shows that $\pfix$ of synonymous SNVs is not
+affected by this change, while $\pfix$ of nonsynonymous SNVs approaches the
+diagonal as the rate of recognition of escape mutants is increased.
 
-In the alternative hypothesis, several different escape mutations within the
-same epitope might arise almost simultaneously and start to spread. Their
-benefits are not additive, because each of them is essentially sufficient to
-escape and no additional benefit is gained from combining them.
-As a consequence, several escape mutations rise to high frequency
-rapidly, while the one with the smallest cost in terms of replication,
-packaging, etc. is most likely to eventually fix, while all others are lost. The emergence of multiple
-sweeping nonsynonymous mutations in real HIV-1 infections has been shown
-\citep{moore_limited_2009, bar_early_2012}. This scenario has been explicitly 
-observed in  the evolution of resistance to 3TC, where the mutation 
-M184V is often preceeded by M184I \citep{hedskog_dynamics_2010}. Similarly, AZT 
-resistance often emerges via the competing TAM and TAM1 pathways.
-Within epitope competition can be
-implemented in the model through epistasis between escape mutations. While each
-mutation is individually beneficial, combining the mutations is deleterious (no
-extra benefit, but additional costs). Again, we find that the potential for
+In the alternative hypothesis, several different escape SNVs within the same
+epitope might arise almost simultaneously and start to spread. Their benefits
+are not additive, because each of them is essentially sufficient to escape and
+no additional benefit is gained from combining them. As a consequence, several
+escape SNVs rise to high frequency rapidly, while the one with the smallest cost
+in terms of replication, packaging, etc. is most likely to eventually fix, while
+all others are lost. The emergence of multiple competing escape SNVs in HIV-1
+infections has been shown \citep{moore_limited_2009, bar_early_2012}. For
+instance, this scenario has been explicitly observed in the evolution of
+resistance to 3TC, where the mutation M184V is often preceeded by M184I
+\citep{hedskog_dynamics_2010}. Similarly, AZT resistance often emerges via the
+competing TAM and TAM1 pathways. Within epitope competition can be implemented
+in the model through epistasis between escape mutations. While each mutation is
+individually beneficial, combining the mutations is deleterious (no extra
+benefit, but additional costs). Again, we find that the potential for
 hitchhiking is little affected by within epitope competition but that the
-fixation probability of nonsynonymous polymorphisms is reduced. With roughly six
+fixation probability of nonsynonymous SNVs is reduced. With roughly six
 mutations per epitope, the simulation data are compatible with observations; see
-\figurename~S\withinepi. The two scenarios are not exclusive and possibly both
-important in HIV-1 evolution.
+\figurename~S\withinepi. The two scenarios, time-dependent selection and
+competition between equivalent escape pathways, are not exclusive and possibly
+both important in HIV-1 evolution.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-By analyzing the fate of mutations in longitudinal data of HIV-1 \env{} evolution,
-we demonstrate selection against synonymous substitutions in the comparatively
-conserved regions C2-C4 of the \env{} gene. Comparison with biochemical studies
-of base pairing propensity in RNA genome of HIV-1 indicates that these
-mutations are deleterious, at least in part, because they disrupt stems in RNA
-secondary structures. Computational modeling shows that these mutations have
-deleterious effects on the order of $0.002$ and that they are brought to high
-frequency through linkage to adaptive mutations.
+By analyzing the fate of single nucleotide variants (SNVs) in longitudinal data
+of HIV-1 \env{} evolution, we demonstrate selection against synonymous
+substitutions in the relatively conserved regions C2-C4 of the \env{} gene.
+Comparison with biochemical studies of base pairing propensity in RNA genome of
+HIV-1 indicates that these mutations are deleterious, at least in part, because
+they disrupt stems in RNA secondary structures. Computational modeling suggests
+that these SNVs have deleterious effects on the order of $0.002$ and that they
+are brought to high frequency through linkage to adaptive mutations.
 
@@
 article, however, is a ratio between propagators of nonsynonymous and synonymous
-mutations. The latter is used as an approximately neutral control; this method
+SNVs. The latter is used as an approximately neutral control; this method
 can therefore not be used to investigate synonymous changes themselves. More
@@
 
-A functional significance of the insulating RNA structure stems between the
-hypervariable loops has also been proposed previously
-\citep{watts_architecture_2009, sanjuan_interplay_2011} and conserved RNA
-structures exist in different parts of the HIV-1 genome. Since there are
-of course many ways to build an RNA stem in a particular location, we do
-not necessarily expect a strong signal of conservation in cross-sectional data. 
-Our analysis, however, is able to
-quantify the fitness effect of RNA structure within single infections and
-demonstrates how selection at synonymous sites can alter genetic diversity and
-dynamics. %FIXME insert citations and more discussion
+The biological cause behind the deleterious effects of synonymous SNVs seems to
+be, at least in part, the disruption of RNA base pairs in the flanks of the V
+loops. Computer models of RNA folding predict stable hairpins in this region:
+these hairpins have been previously suggested to be functional and termed
+``insulating stems'' \citep{watts_architecture_2009, sanjuan_interplay_2011}. In
+a characterization of the SIV RNA structures via the SHAPE assay, it was
+recently shown that only the general pattern is shared with HIV-1; the single base
+pairs are almost always discordant between the two viruses
+\citep{pollom_comparison_2013}. In fact, for each pairing-disrupting mutation in
+our dataset, we have looked for compensatory mutations at the partner site, but
+observed no effect. A direct approach to test the function of the insulating
+stems would be \textit{in vitro} replication assays on synonymous variant
+viruses with unstable hairpins. This has been attempted recently, but no major
+fitness effect was measured \citep{knoepfel_role_2013}. This result is
+compatible with our findings that synonymous SNVs are \textit{slightly}
+deleterious: it would take hundreds of cell culture passages to detect fitness
+effects of the order of one per mille, which is what we estimate in our
+long-term dataset. In conclusion, despite the flexibility in pairing partners,
+our analysis is able to quantify the subtle fitness effect of RNA structure
+within single infections and demonstrates how selection at synonymous sites can
+alter genetic diversity and dynamics.
 
@@
 order of $\rho = 10^{-5}$ per base and day. It takes roughly $t_{sw} =
-\epsilon^{-1} \log \nu_0$ generations for an escape mutation with escape rate
+\epsilon^{-1} \log \nu_0$ generations for an escape SNV with escape rate
 $\epsilon$ to rise from an initially low frequency $\nu_0\approx \mu$ to frequency
@@
 be unconditionally beneficial. Otherwise we would observe almost sure fixation
-of a nonsynonymous mutation once they reach intermediate frequencies. Instead, we
-find that the fixation probability of nonsynonymous mutations is roughly given
-by its frequency. There are several possible explanations for this observation.
-Similar to synonymous mutations, the majority of nonsynonymous mutations could
-be weakly deleterious, and the adaptive and deleterious parts could conspire to yield a
+of nonsynonymous SNVs once they reach high frequencies. Instead, we find that
+the fixation probability of nonsynonymous SNVs is roughly given by its
+frequency. There are several possible explanations for this observation.
+Similar to synonymous SNVs, the majority of nonsynonymous SNVs could be weakly
+deleterious, and the adaptive and deleterious parts could conspire to yield a
 more neutral-like averaged fixation probability. While weakly deleterious
-nonsynonymous mutations certainly exist and will contribute to a depression of the
+nonsynonymous SNVs certainly exist and will contribute to a depression of the
 fixation probability, we have seen that a substantial depression requires that
-weakly deleterious nonsynonymous polymorphisms at high frequency greatly 
-outnumber escape mutations. This seems unlikely, since nonsynonymous diversity
-exceeds synonymous diversity despite the overall much greater constraints on
-the amino acid sequence.
+weakly deleterious nonsynonymous SNVs at high frequency greatly outnumber escape
+SNVs. This seems unlikely, since nonsynonymous diversity exceeds synonymous
+diversity despite the overall much greater constraints on the amino acid
+sequence.
 
@@
 epitope implies a large effective population size that explores all necessary point
-mutations rapidly. A similar point has been made recently by Boltz {\it et al.}
+mutations rapidly. A similar point has been made recently by several authors
 in the context of preexisting drug resistance mutations

### 2013-06-04 synmut.tex
changes, mostly in discussion

@@
 \noindent
-Intrapatient HIV-1 evolution is dominated by selection on the protein level in
-the arms race with the adaptive immune system. When cytotoxic CD8${}^+$ T-cells
+Intrapatient HIV-1 evolution is driven by the adaptive immune system
+resulting in rapid evolution of HIV-1 proteins. When cytotoxic CD8${}^+$ T-cells
 or neutralizing antibodies target a new epitope, the virus often escapes via
 nonsynonymous mutations that impair recognition. Synonymous mutations do not
-affect this interplay and are often assumed to be neutral. We analyze
+affect this interplay and are often assumed to be neutral. We test this
+assumption by tracking synonymous mutations in 
 longitudinal intrapatient data from the \shankaregion{} part of the envelope
-gene (\env{}) and observe that synonymous mutations rarely spread even though
-they often reach high frequencies in the viral population. Using published data
-from the SHAPE assay, we find that synonymous mutations that disrupt base pairs
-in RNA stems flanking the V loops of gp120 are more likely to be lost than other
+gene (\env{}).  We find that most synonymous mutations are lost even though
+they often reach high frequencies in the viral population suggesting a
+cost to the virus. Using published data
+from SHAPE assays, we find that synonymous mutations that disrupt base pairs
+in RNA stems flanking the variable loops of gp120 are more likely to be lost than other
 synonymous changes, suggesting a function of these RNA hairpins in HIV-1.
 Computational modeling indicates that these synonymous mutations have a
-selection coefficient of the order of $-0.002$ and that they are brought up to
-high frequency by genetic hitchhiking on neighboring beneficial variants. We
-conclude that, contrary to common assumptions, synonymous mutations in the V
-loops region are neither independent of the escape patterns nor neutral; further
-studies are needed to clarify the function of the RNA hairpins found there.
+selection coefficient of the order of $-0.002$ and that they are rise to
+high frequency by genetic hitchhiking on neighboring beneficial
+variants. This weak selection against synonymous substitutions does not
+result in a strong pattern of conservation in cross-sectional data, but
+slows down the rate evolution considerably. Our findings are consistent with the
+notion that large scale patterns of RNA structure are functionally
+relevant, while the precise base pairing pattern is not.
 \end{abstract}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
+%\section{Introduction}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 (nAbs)~\citep{rambaut_causes_2004} and facilitated by the high mutation rate
-~\citep{mansky_lower_1995,abram_nature_2010}. When the host develops a CTL or
-nAb response against a particular HIV-1 epitope, mutations in the viral genome that
-reduce or prevent recognition of the epitope frequently emerge. Escape mutations
-in epitopes targeted by CTLs typically emerge during early infection and spread
+~\citep{mansky_lower_1995,abram_nature_2010}.  Escape mutations
+in epitopes targeted by CTLs are typically observed during early infection and spread
 rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
@@
 viral proteins in a way that reduces antibody binding or epitope presentation.
-Conversely, synonymous mutations do not modify the viral protein and are
+Conversely, synonymous mutations do not modify the viral proteins and are
 commonly used as approximately neutral markers in studies of viral evolution.
@@
 (RT) and is bound by a pseudoknotted RNA structure in the viral 5' untranslated
-region~\citep{barat_interaction_1991, paillart_vitro_2002}. These RNA
-structures, but not the exact base pairs, are well conserved between HIV-1 and
-SIV \citep{pollom_comparison_2013}.
+region~\citep{barat_interaction_1991, paillart_vitro_2002}. 
+Besides these well characterized and conservered structures, large parts of the virus
+genome are folded into structures of unknown function
+\citep{watts_architecture_2009}. These poorely characterized RNA
+structures are conserved to varying degree between HIV-1 and
+SIV in the sense that corresponding regions tend to be part of
+of similar structural elements. Individual base pairings, however, are
+not conserved \citep{pollom_comparison_2013}.
 
@@
 rates of evolution at synonymous sites vary along the HIV-1 genome
-\citep{mayrose_towards_2007} (see also \figurename~S\syndiv). Positive
-selection through the host adaptive immune system, however, is restricted to
-changes in the amino acid sequence.
+\citep{mayrose_towards_2007} (see also \figurename~S\syndiv).
 
@@
 and show that, in the region of the V loops, a large fraction of these mutations
-is deleterious. Contrary to common assumptions, deleterious synonymous mutations
+is deleterious. Despite their fitness cost, deleterious synonymous mutations
 rise in frequency in the viral population via genetic hitchhiking due to limited
@@
 batorsky_estimate_2011}. We show a strong correlation between the fate of a
-synonymous mutation and the surrounding RNA structure, suggesting an important
+synonymous mutation and the surrounding RNA structure, suggesting a
 role of RNA hairpins around V-loops in HIV-1 evolution. We then compare our
@@
 zero and one, of the sequences from that time point. An SNV rises or falls in
-frequency by virtue of both (i) its own effect on fitness and escape and (ii)
-the effect of its genetic background. When an SNVs is present in all sequences
+frequency because of (i) its own effect on fitness and escape and (ii)
+its association to genetic backgrounds and (iii) stochastic fluctuations
+(genetic drift). When an SNVs is present in all sequences
 at a certain time point, we say it has ``fixed''; when it is completely absent,
-we say it was ``lost'' or is ``extinct''. Eventually, all SNVs will fix or be lost; the
-probability of fixation is higher for beneficial SNVs, and lower for deleterious
-ones. The intermediate case are neutral SNVs: a neutral mutation segregating at
-frequency $\nu$ has a probability $\pfix(\nu) = \nu$ to spread through the
-population and fix; in the rest of the cases it goes extinct. For instance, if
-a neutral SNV is observed in half of the sequences, it will fix with a probability of
-50\% (see inset in figure \ref{fig:aftsyn}).
+we say it was ``lost'' or is ``extinct''.
 
-The problem with the classification based on beneficial/neutral/deleterious is
-that, given an SNV, we do not know its class \textit{a priori}. Synonymous SNVs,
-however, are usually assumed to be neutral. \FIG{aft} shows the time course of
-the frequencies of all synonymous and nonsynonymous viral mutations observed in
-\env, in \shankaregion, in a chronically HIV-1 infected patient (p10
-from \citet{shankarappa_consistent_1999}). Despite many synonymous SNVs
-reaching high frequency, very few fix (panel~\ref{fig:aftsyn}); in constrast, many
-nonsynonymous mutations fix (panel~\ref{fig:aftnonsyn}). This basic observation
-seems at odds with the assumption of neutrality.
+Eventually, all SNVs will either fix or be lost. Given the SNV is at a
+certain frequency $\nu$, the probability of fixation is higher for
+beneficial SNVs than for neutral ones, which in turn fix more frequently
+than deleterious ones. Since a neutral mutation, by definition, has no
+effect on the fate of the viruses carrying it, the fixation probability 
+of a neutral SNV at frequency $\nu$ is $\pfix(\nu) = \nu$; 
+it goes extinct with probability $1-\nu$. For instance, if a
+neutral SNV is observed in half of the sequences, it will fix with a
+probability of 50\% (see inset in figure \ref{fig:aftsyn}). The fixation
+probability of neutral SNVs is independent of most model assumptions
+and is only affected if neutral SNVs are associated preferentially with
+either high or low fitness virus.
+
+\FIG{aft} shows the time course of the frequencies of all synonymous
+SNVs (top) and nonsynonymous SNVs (bottom) observed in \env, in
+\shankaregion, in a chronically HIV-1 infected patient (p10 from
+\citet{shankarappa_consistent_1999}). Despite many synonymous SNVs
+reaching high frequency, very few fix (panel~\ref{fig:aftsyn}); in
+constrast, many nonsynonymous mutations fix
+(panel~\ref{fig:aftnonsyn}). This observation seems at odds with
+the assumption of neutrality.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Many synonymous polymorphisms in \shankaregion{} are deleterious}
+\subsection*{Many synonymous SNVs in \shankaregion{} are deleterious}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-We study the dynamics and fate of synonymous SNVs more quantitatively by
+We studied the dynamics and fate of synonymous SNVs more quantitatively by
 analyzing data from seven patients from \citet{shankarappa_consistent_1999} and
@@
 former data set is restricted to the \shankaregion{} region of \env, while the
-data from \citet{bunnik_autologous_2008} cover most of \env.  We consider all
+data from \citet{bunnik_autologous_2008} cover most of \env.  We considered all
 SNVs in a frequency interval $[\nu_0-\delta\nu, \nu_0+\delta\nu]$ at some time
-$t$, and calculate the fraction that are still observed at later times $t+\Delta
+$t$, and calculated the fraction that are still observed at later times $t+\Delta
 t$. Plotting this fraction against the time interval $\Delta t$, we see that
@@
 probability of fixation, $\pfix$, is shown as a function of the initial
-frequency $\nu_0$ in panel \ref{fig:fixp2}. We find that $\pfix$ of synonymous
+frequency $\nu_0$ in panel \ref{fig:fixp2}. We found that $\pfix$ of synonymous
 variants is far below the neutral expectation in \shankaregion (red line).
 Outside of \shankaregion, using data from \citet{bunnik_autologous_2008} only,
-we find no such reduction in $\pfix$ (green line). Restricted to the
+we found no such reduction in $\pfix$ (green line). Restricted to the
 \shankaregion{} region, the sequence samples from \citet{bunnik_autologous_2008}
@@
 When interpreting these results for the fixation probabilities, it is important
-to note that we focus on SNVs that have already reached high frequencies. In
+to note that we focused on SNVs that have already reached high frequencies. In
 HIV-1 infection, most SNVs remain very rare throughout; they are not considered
-here. Synonymous SNVs can reach high frequencies by two means, either genetic
+here. Synonymous SNVs can reach high frequencies either through genetic
 drift or genetic hitchhiking on escape variants (see below); very deleterious
@@
 analysis indicates that, among all synonymous SNVs that somehow reach high
-frequencies, most of them are deleterious in \shankaregion{}, while they tend to
-be neutral in the rest of \env{}.
+frequencies, most of those in \shankaregion{} are deleterious, while
+those in the rest of \env{} tend to be neutral.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Synonymous mutations in \shankaregion{} tend to disrupt RNA stems}
+\subsection*{Synonymous mutations in \shankaregion{} tend to disrupt RNA stems}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 To test the hypothesis that synonymous SNVs in \shankaregion{} are lost because they
-break stems in the conserved stretches between the V loops, we consider
+break stems in the conserved stretches between the V loops, we considered
 separately SNVs in V loops and conserved flanks. The greatest
@@
 codon usage bias (CUB). HIV-1 is known to prefer A-rich codons over highly
-expressed human codons \citep{jenkins_extent_2003, kuyl_biased_2012}. We do not
-find, however, any evidence for a contribution of average CUB to the ultimate
-fate of synonymous SNVs; this agrees with the observation that HIV-1 is not
+expressed human codons \citep{jenkins_extent_2003, kuyl_biased_2012}. We
+did not find any evidence for a contribution of average CUB to the ultimate
+fate of synonymous SNVs; this is consistent with the observation that HIV-1 is not
 adapting its codon usage to its human host cells at the macroevolutionary level
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Deleterious SNVs reach high frequency by hitchhiking}
+\subsection*{Deleterious SNVs reach high frequency by hitchhiking}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 chronic infections have been estimated to be on the order of $\epsilon = 0.01$
-per day \citep{neher_recombination_2010, Asquith:2006p28003}.
+per day \citep{neher_recombination_2010, Asquith:2006p28003}, consistent
+with a lag in neutralization of a few month \citep{richman_rapid_2003}.
 
 \FIG{simfixpvar} shows simulation results for the fixation probability and the
-synonymous diversity for different deleterious effects of synonymous mutations.
+synonymous diversity for different deleterious effects sizes of synonymous mutations.
 We quantify synonymous diversity via $P_\text{interm}$, the fraction of sites
-with an SNV at frequency $0.25 < \nu < 0.75$. The synonymous diversity
+with an synonymous SNV at frequency $0.25 < \nu < 0.75$. The synonymous diversity
 observed in patient data is indicated in the figure. To quantify the depression
@@
 repeatedly simulated the evolution with random choices for the parameters in
-certain bounds, see \FIG{simsfig}. Among all simulations, we select the ones
+certain bounds, see \FIG{simsfig}. Among all simulations, we selected the ones
 that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
@@
 parameter combinations for which we found $A_\text{syn} < -0.15$ and $0.0025 <
-P_\text{interm} < 0.010$. These conditions indicate that a high fraction
+P_\text{interm} < 0.010$. This subset of parameters indicates that a high fraction
 ($\gtrsim 0.8$) of sites has to be deleterious with effect size $|s_d| \sim
-0.002$.  This result fits well the expectation based on the fixation/extinction
-times above (see \FIG{fixp1}). The results of simulations are plausible: (i) a
-substantial depression in $\pfix$ requires pervasive deleterious SNVs, otherwise
+0.002$, consistent with the fixation/extinction times estimated above
+(see \FIG{fixp1}).
+
+The bounds on the fraction of synonymous SNVs that are deleterious and
+their effect sizes are plausible:
+(i) A substantial depression in $\pfix$ requires pervasive deleterious SNVs, otherwise
 the majority of SNVs reaching high frequency are neutral and no depression is
@@
 mutant to rise from low to high frequency. To model this type of behavior, we
-assume that antibody responses against escape SNVs arise with a rate
+assumed that antibody responses against escape SNVs arise with a rate
 proportional to the frequency of the escape SNV and abolish the benefit of the
-escape mutations. As expected, this type of time-dependent selection retains the
-potential for hitchhiking, but reduces fixation of nonsynonymous SNVs.
+escape mutations. As expected, this type of time-dependent selection retained the
+potential for hitchhiking, but reduced fixation of nonsynonymous SNVs.
 \figurename~S\timedependence~shows that $\pfix$ of synonymous SNVs is not
@@
 escape SNVs rise to high frequency rapidly, while the one with the smallest cost
-in terms of replication, packaging, etc. is most likely to eventually fix, while
+in terms of replication, packaging, etc.~is most likely to eventually fix, while
 all others are lost. The emergence of multiple competing escape SNVs in HIV-1
-infections has been shown \citep{moore_limited_2009, bar_early_2012}. For
-instance, this scenario has been explicitly observed in the evolution of
+infections has been shown \citep{moore_limited_2009, bar_early_2012}. 
+Similarly, this scenario has been explicitly observed in the evolution of
 resistance to 3TC, where the mutation M184V is often preceeded by M184I
-\citep{hedskog_dynamics_2010}. Similarly, AZT resistance often emerges via the
-competing TAM and TAM1 pathways. Within epitope competition can be implemented
-in the model through epistasis between escape mutations. While each mutation is
-individually beneficial, combining the mutations is deleterious (no extra
-benefit, but additional costs). Again, we find that the potential for
-hitchhiking is little affected by within epitope competition but that the
+\citep{hedskog_dynamics_2010}. We implemented within epitope competition 
+in the model by allowing multiple escape mutations per epitope that do
+not provide additional benefit to the virus when combined. Again, we found that the potential for
+hitchhiking is retained by within epitope competition but that the
 fixation probability of nonsynonymous SNVs is reduced. With roughly six
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-By analyzing the fate of single nucleotide variants (SNVs) in longitudinal data
-of HIV-1 \env{} evolution, we demonstrate selection against synonymous
-substitutions in the relatively conserved regions C2-C4 of the \env{} gene.
-Comparison with biochemical studies of base pairing propensity in RNA genome of
-HIV-1 indicates that these mutations are deleterious, at least in part, because
-they disrupt stems in RNA secondary structures. Computational modeling suggests
-that these SNVs have deleterious effects on the order of $0.002$ and that they
-are brought to high frequency through linkage to adaptive mutations.
+By analyzing the fate of single nucleotide variants (SNVs) in
+longitudinal data of HIV-1 \env{} evolution, we demonstrated selection
+against synonymous substitutions in the relatively conserved regions
+C2-C4 of the \env{} gene. Computational modeling suggested that these
+SNVs have deleterious effects on the order of $0.002$ and that they are
+brought to high frequency through linkage to adaptive mutations.
+Comparison with biochemical studies of base pairing propensity in RNA
+genome of HIV-1 indicated that these mutations are deleterious, at least
+in part, because they disrupt RNA secondary structures. Computer models
+of RNA folding predict stable hairpins in these regions that
+have been suggested to be functional and termed ``insulating
+stems'' \citep{watts_architecture_2009, sanjuan_interplay_2011}.
 
-The fixation and extinction times and probabilities represent a rich and simple
+The fixation probabilities and the sojourn times of SNVs represent a rich and simple
 summary statistics useful to characterize longitudinal sequence data and compare
-to models via computer simulations. A method that is similar to ours {\it in
-spiritu} has been recently used in a longitudinal study of influenza
-evolution~\citep{strelkowa_clonal_2012}. The central quantity used in that
-article, however, is a ratio between propagators of nonsynonymous and synonymous
-SNVs. The latter is used as an approximately neutral control; this method
-can therefore not be used to investigate synonymous changes themselves. More
-generally, evolutionary rates at synonymous sites are often used as a baseline
-to detect purifying or diversifying selection at the protein level
-\cite{Hurst:2002p32608}. It has been pointed out, however, that the rate of
-evolution at synonymous sites varies considerable along the HIV-1 genome
-\citep{mayrose_towards_2007} and that this variation can confound estimates of
+to models via computer simulations. These statistics are informative
+even in absence of a neutral control and thus suited to analyze
+properties of synonymous sites. 
+We find selection against synonymous substitutions  despite the
+fact that the corresponding sites are not conservered in
+cross-sectional data. This is consistent with a recent comparative
+analysis of SIV and HIV-1 secondary structure using SHAPE assays and
+computational methods \citep{pollom_comparison_2013}. While large scale
+patterns of RNA structures tend to agree in both viruses, the individual
+base pairs forming the structures are almost always
+discordant. Hence selection to maintain these structures reduces the
+fixation probability and hence the rate of evolution at synonymous
+sites, but the molecular architecture of the structure changes over time.
+As expected from this argument, the evolutionary rate
+at synonymous sites varies greatly along the HIV-1 genome
+\citep{mayrose_towards_2007}. This variation can confound estimates of
 selection on proteins substantially \citep{ngandu_extensive_2008}. 
+The dynamic nature of HIV secondary structure makes finding the 
+compensatory mutations that would restore base pairing in the
+longitundinal data difficult, since the exact base pairing pattern is
+most likely different than in the reference sequence.
 
-The biological cause behind the deleterious effects of synonymous SNVs seems to
-be, at least in part, the disruption of RNA base pairs in the flanks of the V
-loops. Computer models of RNA folding predict stable hairpins in this region:
-these hairpins have been previously suggested to be functional and termed
-``insulating stems'' \citep{watts_architecture_2009, sanjuan_interplay_2011}. In
-a characterization of the SIV RNA structures via the SHAPE assay, it was
-recently shown that only the general pattern is shared with HIV-1; the single base
-pairs are almost always discordant between the two viruses
-\citep{pollom_comparison_2013}. In fact, for each pairing-disrupting mutation in
-our dataset, we have looked for compensatory mutations at the partner site, but
-observed no effect. A direct approach to test the function of the insulating
+A direct approach to test the function of the insulating
 stems would be \textit{in vitro} replication assays on synonymous variant
@@
 fitness effect was measured \citep{knoepfel_role_2013}. This result is
-compatible with our findings that synonymous SNVs are \textit{slightly}
+compatible with our findings that synonymous SNVs are weakly
 deleterious: it would take hundreds of cell culture passages to detect fitness
 effects of the order of one per mille, which is what we estimate in our
-long-term dataset. In conclusion, despite the flexibility in pairing partners,
+long-term dataset. The longitudinal data, however, spans many years and 
 our analysis is able to quantify the subtle fitness effect of RNA structure
-within single infections and demonstrates how selection at synonymous sites can
-alter genetic diversity and dynamics.
+within single infections.
+
+Our analysis of fixation and loss of variants also has implications for
+non-synonymous SNVs observed in \env. Nonsynonymous diversity exceeds
+synonymous diversity despite the overall much greater constraints on the
+amino acid sequence, suggesting that the majority of high frequency SNVs
+are escape mutations. If those mutations were
+unconditionally beneficial, we would expect them to fix almost surely.
+However, we found that many of them are lost and that they on average
+show fixation behavior like neutral mutations. We suggest that this
+paradoxical behavior could be due to escape
+mutations that revert after they themselves are recognized by nAbs, and
+the competition between different escape mutations within one epitope. 
+Both mechanisms reduce the overall fixation probability and could
+explain the observed pattern of fixation.
 
@@
 when draft dominates over drift.
-
-Contrary to na\"ive expectations, the adaptive escape mutations do not seem to
-be unconditionally beneficial. Otherwise we would observe almost sure fixation
-of nonsynonymous SNVs once they reach high frequencies. Instead, we find that
-the fixation probability of nonsynonymous SNVs is roughly given by its
-frequency. There are several possible explanations for this observation.
-Similar to synonymous SNVs, the majority of nonsynonymous SNVs could be weakly
-deleterious, and the adaptive and deleterious parts could conspire to yield a
-more neutral-like averaged fixation probability. While weakly deleterious
-nonsynonymous SNVs certainly exist and will contribute to a depression of the
-fixation probability, we have seen that a substantial depression requires that
-weakly deleterious nonsynonymous SNVs at high frequency greatly outnumber escape
-SNVs. This seems unlikely, since nonsynonymous diversity exceeds synonymous
-diversity despite the overall much greater constraints on the amino acid
-sequence.
-
-Alternatively, the lack of fixation could be due to time-dependent environment
-through an immune system that is catching up, or competition between mutations
-that mediate escape within the same epitope. We explore both of these
-possibilities and find that both produce the desired effect in computer models. Furthermore, there
-is experimental evidence in support of both of these hypotheses. Serum from HIV-1
-infected individuals typically neutralizes the virus that dominated the
-population a few (3-6) months earlier \citep{richman_rapid_2003}. This suggests that
-escape mutations cease to be beneficial after a few months and might revert if
-they come with a fitness cost. Deep sequencing of regions of \env{} after
-antibody escape have revealed multiple escape mutations in the same epitope
-\citep{moore_limited_2009, bar_early_2012}. Presumably, each one of these
-mutations is sufficient for escape but most combinations of them do not provide
-any additional benefit to the virus. Hence only one mutation will spread and the
-others will be driven out of the population although they transiently reach high
-frequencies. The rapid emergence of multiple escape mutations in the same
-epitope implies a large effective population size that explores all necessary point
-mutations rapidly. A similar point has been made recently by several authors
-in the context of preexisting drug resistance mutations
-\citep{boltz_ultrasensitive_2012}.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Methods}
+\section*{Methods}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Sequence data collection}
+\subsection*{Sequence data collection}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Sequence analysis}
+\subsection*{Sequence analysis}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Computer simulations}
+\subsection*{Computer simulations}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Methods availability}
+\subsection*{Methods availability}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2013-06-05 synmut.tex
streamlining discussion

@@
 batorsky_estimate_2011}. We show a strong correlation between the fate of a
-synonymous mutation and the surrounding RNA structure, suggesting a
-role of RNA hairpins around V-loops in HIV-1 evolution. We then compare our
-observations to computational models and derive estimates for the effect of
+synonymous mutation and the surrounding RNA structure. We then compare our
+observations to computational models and obtain estimates for the effect of
 synonymous mutations on viral fitness. Extending the analysis of fixation
@@
 
-Eventually, all SNVs will either fix or be lost. Given the SNV is at a
-certain frequency $\nu$, the probability of fixation is higher for
-beneficial SNVs than for neutral ones, which in turn fix more frequently
-than deleterious ones. Since a neutral mutation, by definition, has no
-effect on the fate of the viruses carrying it, the fixation probability 
-of a neutral SNV at frequency $\nu$ is $\pfix(\nu) = \nu$; 
-it goes extinct with probability $1-\nu$. For instance, if a
-neutral SNV is observed in half of the sequences, it will fix with a
-probability of 50\% (see inset in figure \ref{fig:aftsyn}). The fixation
-probability of neutral SNVs is independent of most model assumptions
-and is only affected if neutral SNVs are associated preferentially with
-either high or low fitness virus.
+Eventually, all SNVs will either fix or be lost, at least in small
+samples. Given the SNV is at a certain frequency $\nu$, the probability
+of fixation is higher for beneficial SNVs than for neutral ones, which
+in turn fix more frequently than deleterious ones. Since a neutral
+mutation, by definition, has no effect on the fate of the viruses
+carrying it, the fixation probability of a neutral SNV at frequency
+$\nu$ is $\pfix(\nu) = \nu$; it goes extinct with probability
+$1-\nu$. For instance, if a neutral SNV is observed in half of the
+sequences, it will fix with a probability of 50\% (see inset in figure
+\ref{fig:aftsyn}). The fixation probability of neutral SNVs is
+independent of most model assumptions and is only affected if neutral
+SNVs are associated preferentially with either high or low fitness
+virus.
 
 \FIG{aft} shows the time course of the frequencies of all synonymous
-SNVs (top) and nonsynonymous SNVs (bottom) observed in \env, in
-\shankaregion, in a chronically HIV-1 infected patient (p10 from
+SNVs (top) and nonsynonymous SNVs (bottom) observed in the
+\shankaregion{} region of \env{} in a chronically HIV-1 infected patient (p10 from
 \citet{shankarappa_consistent_1999}). Despite many synonymous SNVs
@@
 measured using the SHAPE assay, a biochemical reaction preferentially altering
-unpaired bases (the HIV-1 genome is a single stranded RNA)
+unpaired bases 
 \citep{watts_architecture_2009}. The SHAPE assay has shown that the variable
@@
 To map the parameter range of the model that is compatible with the data, we
-repeatedly simulated the evolution with random choices for the parameters in
-certain bounds, see \FIG{simsfig}. Among all simulations, we selected the ones
+repeatedly simulated the evolution with random choices for the
+parameters, see \FIG{simsfig}. Among all simulations, we selected the ones
 that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
@@
 though many are at least transiently beneficial. Inspecting the trajectories of
-nonsynonymous SNVs suggests the rapid rise and fall of many SNVs. We
-test two possible mechanisms that are biologically plausible and could explain
-the transient rise of nonsynonymous SNVs: time-dependent selection and
-within-epitope competition.
+nonsynonymous SNVs suggests the rapid rise and fall of many SNVs. Two
+possible mechanisms that could explain the transient rise of
+nonsynonymous SNVs are time-dependent selection and within-epitope competition.
 
-The former hypothesis can be formulated as follows: if the immune system
+If the immune system
 recognizes the escape mutant before its fixation, the mutant might cease to be
@@
 
-In the alternative hypothesis, several different escape SNVs within the same
-epitope might arise almost simultaneously and start to spread. Their benefits
+Several different escape mutations that arise within the 
+same epitope have similar effects. Their benefits
 are not additive, because each of them is essentially sufficient to escape and
@@
 stems'' \citep{watts_architecture_2009, sanjuan_interplay_2011}.
+The weak selection against synonynous variants is compatible with the
+negative results of \textit{In vitro} replication assays investigating
+the fitness effects of small RNA hairpins in HIV-1
+\citep{knoepfel_role_2013}: it would take hundreds of cell culture
+passages to detect fitness effects of the order of one per mille. The longitudinal data,
+however, spans many years and our analysis is able to quantify the
+subtle fitness effect of RNA structure within single infections.
 
@@
 cross-sectional data. This is consistent with a recent comparative
-analysis of SIV and HIV-1 secondary structure using SHAPE assays and
+analysis of SIV and HIV-1 RNA secondary structure using SHAPE assays and
 computational methods \citep{pollom_comparison_2013}. While large scale
@@
 base pairs forming the structures are almost always
-discordant. Hence selection to maintain these structures reduces the
+discordant. Even though the molecular architecture of the structure
+changes over time, selection seems to maintain these structures, which reduces the
 fixation probability and hence the rate of evolution at synonymous
-sites, but the molecular architecture of the structure changes over time.
-As expected from this argument, the evolutionary rate
+sites. As expected from this argument, the evolutionary rate
 at synonymous sites varies greatly along the HIV-1 genome
@@
 
-A direct approach to test the function of the insulating
-stems would be \textit{in vitro} replication assays on synonymous variant
-viruses with unstable hairpins. This has been attempted recently, but no major
-fitness effect was measured \citep{knoepfel_role_2013}. This result is
-compatible with our findings that synonymous SNVs are weakly
-deleterious: it would take hundreds of cell culture passages to detect fitness
-effects of the order of one per mille, which is what we estimate in our
-long-term dataset. The longitudinal data, however, spans many years and 
-our analysis is able to quantify the subtle fitness effect of RNA structure
-within single infections.
-
-Our analysis of fixation and loss of variants also has implications for
-non-synonymous SNVs observed in \env. Nonsynonymous diversity exceeds
+Selection against the majority of synonymous substitutions is probably
+true across the genome, but we only observe deleterious synonymous SNVs
+at high frequency in \shankaregion{} of \env, where they hitch-hike to
+high frequencies on nAb escape mutations. Surprisingly, that non-synonymous
+mutations display a fixation pattern as if they were neutral;
+comp.~\FIG{pfix}. However, nonsynonymous diversity exceeds
 synonymous diversity despite the overall much greater constraints on the
 amino acid sequence, suggesting that the majority of high frequency SNVs
-are escape mutations. If those mutations were
-unconditionally beneficial, we would expect them to fix almost surely.
-However, we found that many of them are lost and that they on average
-show fixation behavior like neutral mutations. We suggest that this
+are escape mutations despite the fact that are often lost again. We suggest that this
 paradoxical behavior could be due to escape
@@
 Such fluctuations have been termed \emph{genetic draft} in
-\citet{gillespie_genetic_2000}. Genetic draft in facultatively sexual population
+\citet{gillespie_genetic_2000} and their effect in facultatively sexual population
 such as HIV-1 has been characterized in \citep{neher_genetic_2011}. Importantly,
@@
 bunnik_autologous_2008} and downloaded from the Los Alamos National Laboratory
-(LANL) HIV sequence database~\citep{LANL2012}. The samples from some patients
-show substantial population structure and were discarded (see
+(LANL) HIV sequence database~\citep{LANL2012}. Some patients
+show substantial population structure and were excluded (see
 \figurename~S\PCApat); a total of 11 patients with 4-23 time points each and
@@
 using Muscle~\citep{edgar_muscle:_2004} to each other and the NL4-3 reference
-sequences separately for each patient. Within each patient, the consensus
+sequence separately for each patient. Within each patient, the consensus
 nucleotide sequence at the first time point was used to classify alleles as
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-For the estimates of time to fixation/extinction, polymorphisms were binned by
+For the estimates of time to fixation/extinction, SNVs were binned by
 frequency and the time to first reaching either fixation or extinction was
@@
 rate $k_A$, a random locus in the genome is designated an epitope that can
-escape by one or several mutations with an exponentially distributed escape rate
+escape by one or several mutations with exponentially distributed escape rates
 with mean $\epsilon$. Both full-length HIV-1 genomes and \env{}-only simulations
@@
 landscape was designed such that each single mutant is sufficient for full
-escape. In particular, each mutation had a linear effect equal to the escape,
-but a negative epistatic effect of the same magnitude between each pair of sites
-was included. Higher order terms compensated each other to make sure that not
+escape. Specifically, each mutation has an additive effect equal to the
+escape rate, but interacts with all other escape mutations in the same
+epitope with a negative effect of the same magnitude. 
+Higher order terms were included to make sure that not
 only double mutants, but all k-mutants with $k \geq 1$ had the same fitness (see
-supplementary materials). To model recognition of escape variants by the immune
-system  catching up, the beneficial effect of an escape mutation was set
+supplementary materials). To model recognition of escape variants by the
+evolving immune system, the beneficial effect of an escape mutation was set
 to its previous cost of -0.02 with a probability per generation proportional to

### 2013-06-05 scripts/fixation_loss_synonymous_times.py
Modified fixation_times script to include nonsynonymous

@@ -56,15 +56,13 @@ mo_to_gen = 30.5
 # Script
 if __name__ == '__main__':
 
-    fixed = []
-    lost = []
-    fixlost = [lost, fixed]
+    fixlost = {'syn': [[], []], 'nonsyn': [[], []]}
 
     # Counts and times (overall)
     # The first list of each nu0s is lost, the second fixed, the third floating
-    counts_all = [[0, 0, 0] for nu0s in nu0ss]
+    counts_all = {key: [[0, 0, 0] for nu0s in nu0ss] for key in ['syn', 'nonsyn']}
     # The first list of each nu0s is lost, the second fixed
-    times_all = [[[], []] for nu0s in nu0ss]
+    times_all = {key: [[[], []] for nu0s in nu0ss] for key in ['syn', 'nonsyn']}
     
     # Define the patients
     patients = pS.parse_sequences(exclude=bad_patients)
@@ -102,11 +100,7 @@ if __name__ == '__main__':
             # of double counting is fine with our scarse data points, since it
             # should not bias the results)
             is_nonused = np.ones(p.L(), bool)
-    
-            # Define the counts (lost, fixed, neither)
-            counts = counts_all[iji]
-            delta_ts = times_all[iji]
-    
+        
             # Initial time point can be everything up to the second-last visit
             for i in xrange(n-1):
 
@@ -124,25 +118,31 @@ if __name__ == '__main__':
                 if len(alleles):
                     is_mut = np.array([alpha[al[0]] != consensus[al[1]] for al in alleles], bool)
                     alleles = alleles[is_mut]
+
+                if not len(alleles):
+                    continue
     
-                # filter only synonymous changes
-                if len(alleles):
-                    is_syn = np.zeros(len(alleles), bool)
-                    for j, al in enumerate(alleles):
-                        pos = al[1]
-                        mut = alpha[al[0]]
-                        codcons = consensus[pos - pos % 3: pos - pos % 3 + 3]
-                        cod = codcons.copy()
-                        cod[pos % 3] = mut
-                        is_syn[j] = (translate(codcons) == translate(cod))
-                    alleles = alleles[is_syn]
+                ###############################################################
+                # Filter only synonymous/nonsynonymous changes
+                ###############################################################
+                # First test
+                is_syn = np.zeros(len(alleles), bool)
[…]

### 2013-06-05 scripts/fixation_loss_synonymous_times.py
More on the same

@@ -165,8 +165,10 @@ if __name__ == '__main__':
 
                 alleles_all = {'syn': alleles_syn, 'nonsyn': alleles_nonsyn}
                 ###############################################################
+                # End of synonymous test
+                ###############################################################
             
-                # Check whether they fix/are lost in at least one time point
+                # Check whether they fix/are lost in at least one later time point
                 for (key, alleles) in alleles_all.iteritems():
                     # Define the counts (lost, fixed, neither)
                     counts = counts_all[key][iji]
@@ -178,6 +180,8 @@ if __name__ == '__main__':
     
                         # Look at later time points
                         for l in xrange(i+1, n):
+
+                            # Later allele frequency at t1
                             alf_sel1 = afs[l][al[0], al[1]]
             
                             # Look at LOST (j=0) and FIXED (j=1)
@@ -188,6 +192,8 @@ if __name__ == '__main__':
                                     # Finally: set the counts
                                     counts[j] += 1
                                     delta_ts[j].append(p.mo_to_SC[l]-p.mo_to_SC[i])
+
+                                    # The position is used now
                                     is_nonused[al[1]] = False
 
                                     # Add the alleles to the list of dicts
@@ -195,6 +201,16 @@ if __name__ == '__main__':
                                                'pos': al[1],
                                                'all': alpha[al[0]],
                                                'all_cons': consensus[al[1]],
+                                               'cod_cons': ''.join(consensus[al[1] - al[1] % 3:
+                                                                             al[1] - al[1] % 3 + 3]),
+                                               'codpos': al[1] % 3,
+                                               'time': p.mo_to_SC[l]-p.mo_to_SC[i],
+                                               'type': key,
+                                               'nu0': alf_sel0,
+                                               'nu1': alf_sel1,
+                                               't0': p.mo_to_SC[i],
+                                               't1': p.mo_to_SC[l],
+                                               'nu0_index': iji,
                                                 }
                                     fixlost[key][j].append(newitem)
     
[…]

### 2013-06-05 synmut.tex
little changes

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-In HIV-1, because of the high mutation rate, single nucleotide variants (SNVs)
-are continuously injected into the viral population. We say that an SNV has a
-frequency $\nu$ at a certain time if it is found in a fraction $\nu$, between
-zero and one, of the sequences from that time point. An SNV rises or falls in
-frequency because of (i) its own effect on fitness and escape and (ii)
-its association to genetic backgrounds and (iii) stochastic fluctuations
-(genetic drift). When an SNVs is present in all sequences
+Due to the large population size and the high mutation rate, every
+possible single nucleotide variant (SNV) is produced multiple time per
+day \citep{coffin_hiv_1995}. Some of these SNVs rise to high enough
+frequency that they are observed in a sample of sequences. SNVs rise or fall in 
+frequency because of (i) their own effect on fitness and escape and (ii)
+their association to genetic backgrounds and (iii) stochastic fluctuations
+(genetic drift). We study the dynamics of sample frequencies, $\nu$, of SNVs
+(SNV), i.e., the fraction $\nu$ of the sequences in a sample carrying
+the SNV. When an SNVs is present in all sequences 
 at a certain time point, we say it has ``fixed''; when it is completely absent,
@@
 
-Eventually, all SNVs will either fix or be lost, at least in small
-samples. Given the SNV is at a certain frequency $\nu$, the probability
-of fixation is higher for beneficial SNVs than for neutral ones, which
-in turn fix more frequently than deleterious ones. Since a neutral
-mutation, by definition, has no effect on the fate of the viruses
-carrying it, the fixation probability of a neutral SNV at frequency
-$\nu$ is $\pfix(\nu) = \nu$; it goes extinct with probability
-$1-\nu$. For instance, if a neutral SNV is observed in half of the
-sequences, it will fix with a probability of 50\% (see inset in figure
+Most positions are only transiently variable and SNVs will either fix or
+are lost -- at least in small samples. Given a SNV is at a certain
+frequency $\nu$, the probability of fixation is higher for beneficial
+SNVs than for neutral ones, which in turn fix more frequently than
+deleterious ones. Since a neutral mutation, by definition, has no effect
+on the fate of the viruses carrying it, the fixation probability of a
+neutral SNV at frequency $\nu$ is the frequency itself, i.e.,
+$\pfix(\nu) = \nu$; it goes extinct with probability $1-\nu$. For
+instance, if a neutral SNV is observed in half of the sequences, it will
+fix with a probability of 50\% (see inset in figure
 \ref{fig:aftsyn}). The fixation probability of neutral SNVs is
@@
 analyzing data from seven patients from \citet{shankarappa_consistent_1999} and
-\citep{liu_selection_2006} as well as three patients from
+\citet{liu_selection_2006} as well as three patients from
 \citet{bunnik_autologous_2008} (patients with strong viral population structure
@@
 
-The above simulations show that hitchhiking can explain the observation of
-deleterious SNVs that rarely fix. However, in a simple model where
+The above simulations show that hitchhiking with favored nonsynonymous
+variants can explain the observation of
+deleterious synonymous SNVs that rarely fix. However, in a simple model where
 nonsynonymous escape mutations are unconditionally beneficial, they almost
@@
 support of this idea, in \citet{richman_rapid_2003, bunnik_autologous_2008},
-antibody responses to escape mutants have been reported. These responses are
+antibody responses against escape mutants have been reported. These responses are
 delayed by a few months, roughly matching the average time needed by an escape
@@
 stems'' \citep{watts_architecture_2009, sanjuan_interplay_2011}.
-The weak selection against synonynous variants is compatible with the
+The weak selection against synonymous variants is compatible with the
 negative results of \textit{In vitro} replication assays investigating
@@
 We find selection against synonymous substitutions  despite the
-fact that the corresponding sites are not conservered in
+fact that the corresponding sites are not conserved in
 cross-sectional data. This is consistent with a recent comparative
@@
 assigned to sites in the multiple sequence alignments for each patient through
-the alignment to the sequence of the NL4.3 virus used in
+the alignment to the sequence of the NL4-3 virus used in
 \citep{watts_architecture_2009}. Problematic assignments in indel-rich

### 2013-06-05 synmut.tex
fixed typos

@@
 region~\citep{barat_interaction_1991, paillart_vitro_2002}. 
-Besides these well characterized and conservered structures, large parts of the virus
+Besides these well characterized and conserved structures, large parts of the virus
 genome are folded into structures of unknown function
-\citep{watts_architecture_2009}. These poorely characterized RNA
+\citep{watts_architecture_2009}. These poorly characterized RNA
 structures are conserved to varying degree between HIV-1 and
@@
 reaching high frequency, very few fix (panel~\ref{fig:aftsyn}); in
-constrast, many nonsynonymous mutations fix
+contrast, many nonsynonymous mutations fix
 (panel~\ref{fig:aftnonsyn}). This observation seems at odds with
@@
 fate of synonymous SNVs; this is consistent with the observation that HIV-1 is not
-adapting its codon usage to its human host cells at the macroevolutionary level
+adapting its codon usage to its human host cells at the macro-evolutionary level
 \citep{kuyl_biased_2012}.
@@
 SNVs fixed (red), sites at which SNVs reached frequencies above 15\% but
-were susequently lost (blue), and sites at which no high-frequency SNVs were observed (green) 
+were subsequently lost (blue), and sites at which no high-frequency SNVs were observed (green) 
 (all categories are restricted to the regions V1-V5$\pm 100$bp).
@@
 Similarly, this scenario has been explicitly observed in the evolution of
-resistance to 3TC, where the mutation M184V is often preceeded by M184I
+resistance to 3TC, where the mutation M184V is often preceded by M184I
 \citep{hedskog_dynamics_2010}. We implemented within epitope competition 
@@
 compensatory mutations that would restore base pairing in the
-longitundinal data difficult, since the exact base pairing pattern is
+longitudinal data difficult, since the exact base pairing pattern is
 most likely different than in the reference sequence.
@@
 challenges such that its logarithm was uniform between $10^{-3}$ and $10^{-2}$
-per generation. Populations were initialized with a homogenous founder
+per generation. Populations were initialized with a homogeneous founder
 population and were kept at an average size of $N=10^4$ throughout the
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Acknowledgements}
+\section*{Acknowledgments}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2013-06-05 synmut.tex
little changes in discussion

@@
 Most positions are only transiently variable and SNVs will either fix or
-are lost -- at least in small samples. Given a SNV is at a certain
+will be lost -- at least in small samples. Given a SNV is at a certain
 frequency $\nu$, the probability of fixation is higher for beneficial
@@
 passages to detect fitness effects of the order of one per mille. The longitudinal data,
-however, spans many years and our analysis is able to quantify the
+however, spans many years, hence our analysis is able to quantify the
 subtle fitness effect of RNA structure within single infections.
@@
 compensatory mutations that would restore base pairing in the
-longitudinal data difficult, since the exact base pairing pattern is
+longitudinal data difficult; in fact, the exact base pairing pattern is
 most likely different than in the reference sequence.
@@
 true across the genome, but we only observe deleterious synonymous SNVs
-at high frequency in \shankaregion{} of \env, where they hitch-hike to
-high frequencies on nAb escape mutations. Surprisingly, that non-synonymous
+at high frequency in \shankaregion{} of \env{}, where they hitchhike to
+high frequencies on nAb escape mutations. Surprisingly, nonsynonymous
 mutations display a fixation pattern as if they were neutral;
-comp.~\FIG{pfix}. However, nonsynonymous diversity exceeds
+see~\FIG{pfix}. However, nonsynonymous diversity exceeds
 synonymous diversity despite the overall much greater constraints on the
@@
 are escape mutations despite the fact that are often lost again. We suggest that this
-paradoxical behavior could be due to escape
-mutations that revert after they themselves are recognized by nAbs, and
+paradoxical behavior could be due to (i) escape
+mutations that revert after they themselves are recognized by nAbs, or (ii)
 the competition between different escape mutations within one epitope. 

### 2013-06-05 synmut.tex
more little changes

@@
 result in a strong pattern of conservation in cross-sectional data, but
-slows down the rate evolution considerably. Our findings are consistent with the
+slows down the rate of evolution considerably. Our findings are consistent with the
 notion that large scale patterns of RNA structure are functionally
@@
 We find selection against synonymous substitutions  despite the
-fact that the corresponding sites are not conserved in
+fact that the corresponding sites are not strongly conserved in
 cross-sectional data. This is consistent with a recent comparative
@@
 base pairs forming the structures are almost always
-discordant. Even though the molecular architecture of the structure
-changes over time, selection seems to maintain these structures, which reduces the
+discordant. Even though the molecular architecture of these structures
+changes over time, selection seems to maintain them, which reduces the
 fixation probability and hence the rate of evolution at synonymous
@@
 mutations display a fixation pattern as if they were neutral;
-see~\FIG{pfix}. However, nonsynonymous diversity exceeds
+see~\FIG{fixp2}. However, nonsynonymous diversity exceeds
 synonymous diversity despite the overall much greater constraints on the

### 2013-06-06 synmut.tex
Text changes + fig 3A for color blind

@@
 Intrapatient HIV-1 evolution is driven by the adaptive immune system
-resulting in rapid evolution of HIV-1 proteins. When cytotoxic CD8${}^+$ T-cells
+resulting in rapid adaptation of HIV-1 proteins. When cytotoxic CD8${}^+$ T-cells
 or neutralizing antibodies target a new epitope, the virus often escapes via
@@
 affect this interplay and are often assumed to be neutral. We test this
-assumption by tracking synonymous mutations in 
-longitudinal intrapatient data from the \shankaregion{} part of the envelope
-gene (\env{}).  We find that most synonymous mutations are lost even though
-they often reach high frequencies in the viral population suggesting a
+assumption by tracking the abundance of synonymous mutations in 
+longitudinal intrapatient data from the \shankaregion{} part of the \env{}
+gene. We find that most synonymous variants are lost even though
+they often reach high frequencies in the viral population, suggesting a
 cost to the virus. Using published data
@@
 in RNA stems flanking the variable loops of gp120 are more likely to be lost than other
-synonymous changes, suggesting a function of these RNA hairpins in HIV-1.
-Computational modeling indicates that these synonymous mutations have a
-selection coefficient of the order of $-0.002$ and that they are rise to
-high frequency by genetic hitchhiking on neighboring beneficial
+synonymous changes; these RNA hairpins might be important for HIV-1.
+Computational modeling indicates that (i) most synonymous mutations in this
+genomic region have a selection coefficient of the order of $-0.002$, and
+(ii) synonymous variants rise to high frequency by genetic hitchhiking on neighboring beneficial
 variants. This weak selection against synonymous substitutions does not
-result in a strong pattern of conservation in cross-sectional data, but
+result in a strong pattern of conservation in cross sectional data, but
 slows down the rate of evolution considerably. Our findings are consistent with the
@@
 via cytotoxic CD8${}^+$ T-cells (CTLs) and neutralizing antibodies
-(nAbs)~\citep{rambaut_causes_2004} and facilitated by the high mutation rate
+(nAbs)~\citep{rambaut_causes_2004} and facilitated by HIV-1's high mutation rate
 ~\citep{mansky_lower_1995,abram_nature_2010}. Escape mutations
@@
 infection, the most rapidly evolving parts of the HIV-1 genome are the variable
-loops V1-V5 in the envelope protein gp120, which change to avoid recognition by
-nAbs. Escape mutations in \env, the gene encoding gp120, spread through the
+loops V1-V5 in the envelope protein gp120 (V loops), which change to avoid recognition by
+nAbs. Escape mutations in \env{}, the gene encoding gp120, spread through the
 viral population within a few months.
 Consistent with this time scale, it is found that serum from a particular time
-typically neutralizes virus extracted more than 3-6 months earlier but not contemporary
+typically neutralizes autologous virus extracted more than 3-6 months earlier but not contemporary
 virus \citep{richman_rapid_2003}.
@@
 Conversely, synonymous mutations do not modify the viral proteins and are
-commonly used as approximately neutral markers in studies of viral evolution.
-Neutral markers are very useful as a negative control for detecting selected
+commonly used as approximately neutral markers in studies of viral evolution,
+i.e. as a negative control for detecting selected
 sites \citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In addition to
@@
 structures are conserved to varying degree between HIV-1 and
-SIV in the sense that corresponding regions tend to be part of
-of similar structural elements. Individual base pairings, however, are
-not conserved \citep{pollom_comparison_2013}.
+SIV: corresponding regions tend to be part of
+of similar structural elements, but individual base pairings differ widely \citep{pollom_comparison_2013}.
 
@@
 and show that, in the region of the V loops, a large fraction of these mutations
-is deleterious. Despite their fitness cost, deleterious synonymous mutations
+is deleterious. Despite their fitness cost, deleterious synonymous variants
 rise in frequency in the viral population via genetic hitchhiking due to limited
@@
 batorsky_estimate_2011}. We show a strong correlation between the fate of a
-synonymous mutation and the surrounding RNA structure. We then compare our
+synonymous variant and the surrounding RNA structure. We then compare our
 observations to computational models and obtain estimates for the effect of
-synonymous mutations on viral fitness. Extending the analysis of fixation
-probabilities to the nonsynonymous mutations, we show that time-dependent
-selection or strong competition of escape mutations inside the same epitope are
-necessary to explain the observed patterns of fixation and loss.
+synonymous mutations on viral fitness.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% DO WE WANT TO KEEP THIS LAST SENTENCE?
+Extending the analysis of fixation
+probabilities to the nonsynonymous variants, we show that time-dependent
+selection or competition of escape mutations inside the same epitope are
+necessary to explain the observed patterns of spread and extinction.
 
@@
 Due to the large population size and the high mutation rate, every
-possible single nucleotide variant (SNV) is produced multiple time per
-day \citep{coffin_hiv_1995}. Some of these SNVs rise to high enough
+possible single nucleotide variant (SNV) is produced multiple times per
+day \citep{coffin_hiv_1995}. Some of these variants rise to high enough
 frequency that they are observed in a sample of sequences. SNVs rise or fall in 
-frequency because of (i) their own effect on fitness and escape and (ii)
-their association to genetic backgrounds and (iii) stochastic fluctuations
-(genetic drift). We study the dynamics of sample frequencies, $\nu$, of SNVs
-(SNV), i.e., the fraction $\nu$ of the sequences in a sample carrying
-the SNV. When an SNVs is present in all sequences 
+frequency because of three reasons: (i) their own effect on fitness and escape; (ii)
+their association to genetic backgrounds; (iii) stochastic fluctuations
+(genetic drift). We study the dynamics of sample frequencies, $\nu$, of SNVs,
+i.e., the fraction $\nu$ of the sequences in a sample carrying
+the variant. When an SNVs is present in all sequences 
 at a certain time point, we say it has ``fixed''; when it is completely absent,
@@
 
-Most positions are only transiently variable and SNVs will either fix or
-will be lost -- at least in small samples. Given a SNV is at a certain
+Most positions are only transiently variable and variants will either fix or
+will be lost -- at least in small samples. Given an SNV is at a certain
 frequency $\nu$, the probability of fixation is higher for beneficial
-SNVs than for neutral ones, which in turn fix more frequently than
-deleterious ones. Since a neutral mutation, by definition, has no effect
-on the fate of the viruses carrying it, the fixation probability of a
+SNVs than for neutral ones; in turn, neutral variants fix more frequently than
+deleterious ones. The fixation probability of a
 neutral SNV at frequency $\nu$ is the frequency itself, i.e.,
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-We studied the dynamics and fate of synonymous SNVs more quantitatively by
+We studied the dynamics and fate of synonymous variants more quantitatively by
 analyzing data from seven patients from \citet{shankarappa_consistent_1999} and
@@
 \citet{bunnik_autologous_2008} (patients with strong viral population structure
-were excluded from the analysis; see methods and \figurename~S\PCApat). The
+were not considered; see methods and \figurename~S\PCApat). The
 former data set is restricted to the \shankaregion{} region of \env, while the
@@
 probability of fixation, $\pfix$, is shown as a function of the initial
-frequency $\nu_0$ in panel \ref{fig:fixp2}. We found that $\pfix$ of synonymous
-variants is far below the neutral expectation in \shankaregion (red line).
+frequency $\nu_0$ in panel \ref{fig:fixp2}. As a control, the neutral result is
+shown as a black dashed line. We found that $\pfix$ of synonymous
+variants is far below the neutral expectation in \shankaregion{} (red triangle line).
 Outside of \shankaregion, using data from \citet{bunnik_autologous_2008} only,
-we found no such reduction in $\pfix$ (green line). Restricted to the
+we found no such reduction in $\pfix$ (green square line). Restricted to the
 \shankaregion{} region, the sequence samples from \citet{bunnik_autologous_2008}
-are fully compatible with data from \citet{shankarappa_consistent_1999}. The
+are fully compatible with data from \citet{shankarappa_consistent_1999}:
+fixation is rare. The
 nonsynonymous SNVs seem to follow more or less the neutral expectation (blue
-line) -- a point to which we will come back below.
+circle line) -- a point to which we come back below.
 
@@
 to note that we focused on SNVs that have already reached high frequencies. In
-HIV-1 infection, most SNVs remain very rare throughout; they are not considered
+HIV-1 infection, most SNVs remain very rare throughout: they are not considered
 here. Synonymous SNVs can reach high frequencies either through genetic
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-One possible explanation for lack of fixation of synonymous SNVs in
+One possible explanation for a reduced fixation of synonymous variants in
 \shankaregion{} is secondary structures in the viral RNA, the disruption of which
@@
 regions V1-V5 tend to be unpaired, while the conserved regions between those
-variable regions form stems. We aligned the within-patient sequence samples to
+stretches form stems. We aligned the sequences from each patient to
 the reference NL4-3 strain used in \citet{watts_architecture_2009} and assigned
@@
 distributions of SHAPE reactivities for synonymous SNVs that fixed or were
-subsequently lost (only SNVs with frequencies above 15\%). As shown in
+lost (only variants reaching frequencies above 15\%). As shown in
 \FIG{SHAPEA}, the reactivities of fixed SNVs (red histogram) are systematically
@@
 that never reach high frequencies (green). This set is a mixture of neutral and
-deleterious SNVs and, as expected, their distribution lies between those of
-fixed and lost high-frequency SNVs.
+deleterious SNVs and, as expected, its distribution lies between those of
+fixed and lost high-frequency variants.
 
-To test the hypothesis that synonymous SNVs in \shankaregion{} are lost because they
+To test the hypothesis that synonymous variants in \shankaregion{} are lost because they
 break stems in the conserved stretches between the V loops, we considered
-separately SNVs in V loops and conserved flanks. The greatest
+separately SNVs in V loops and their flanks. The greatest
 depression in fixation probability is observed in the conserved stems, while the
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-While the observation that some fraction of synonymous SNVs is deleterious
-is not unexpected, it seems odd that we observe them at high population
+While the observation of some deleterious synonymous variants is not fully
+unexpected, it seems odd that we observe them at high population
 frequency and that the fixation probability is reduced only in parts of the
@@
 richman_rapid_2003}. Due to the limited amount of recombination in HIV-1
-\cite{neher_recombination_2010, batorsky_estimate_2011}, deleterious SNVs
+\cite{neher_recombination_2010, batorsky_estimate_2011}, deleterious variants
 that are linked to adaptive variants can reach high frequency. This process is
@@
 ($\gtrsim 0.8$) of sites has to be deleterious with effect size $|s_d| \sim
-0.002$, consistent with the fixation/extinction times estimated above
+0.002$. This is consistent with the fixation/extinction times estimated above
 (see \FIG{fixp1}).
 
-The bounds on the fraction of synonymous SNVs that are deleterious and
-their effect sizes are plausible:
-(i) A substantial depression in $\pfix$ requires pervasive deleterious SNVs, otherwise
+The bounds on the fraction of deleterious synonymous SNVs and
+on their effect sizes are plausible:
+(i) a substantial depression in $\pfix$ requires pervasive deleterious SNVs, otherwise
 the majority of SNVs reaching high frequency are neutral and no depression is
@@
 mutation and the deleterious synonymous one) has little or no fitness advantage
-over the wildtype virus. Consistent with this argument, larger deleterious
+over the wildtype virus. In agreement with this argument, larger deleterious
 effects in \FIG{simsfig} correspond to larger escape rates; and (iii) SNVs
 with a deleterious effect smaller than approximately $0.001$ behave neutrally,
-consistent with the typical coalescent times observed in HIV-1.
+as expected from the typical coalescent times observed in HIV-1.
 
-The above simulations show that hitchhiking with favored nonsynonymous
+The above simulations show that hitchhiking with favoured nonsynonymous
 variants can explain the observation of
 deleterious synonymous SNVs that rarely fix. However, in a simple model where
-nonsynonymous escape mutations are unconditionally beneficial, they almost
+nonsynonymous escape mutations are unconditionally beneficial, escape mutations almost
 always fix once they reach high frequencies, i.e., $A_{\mathrm{nonsyn}}$ is well
-above zero. This is incompatible with the blue line in \FIG{fixp2}: in an HIV-1
+above zero. This is incompatible with the blue circle line in \FIG{fixp2}: in an HIV-1
 infection, nonsynonymous SNVs at high frequency often disappear again, even
@@
 assumed that antibody responses against escape SNVs arise with a rate
-proportional to the frequency of the escape SNV and abolish the benefit of the
+proportional to the frequency of the escape variant and abolish the benefit of the
 escape mutations. As expected, this type of time-dependent selection retained the
@@
 in terms of replication, packaging, etc.~is most likely to eventually fix, while
-all others are lost. The emergence of multiple competing escape SNVs in HIV-1
+all others are lost. The emergence of multiple competing escape variants in HIV-1
 infections has been shown \citep{moore_limited_2009, bar_early_2012}. 
@@
 \citep{hedskog_dynamics_2010}. We implemented within epitope competition 
-in the model by allowing multiple escape mutations per epitope that do
+in the model by allowing for multiple escape mutations per epitope that do
 not provide additional benefit to the virus when combined. Again, we found that the potential for
-hitchhiking is retained by within epitope competition but that the
+hitchhiking is retained by within-epitope competition but that the
 fixation probability of nonsynonymous SNVs is reduced. With roughly six
@@
 brought to high frequency through linkage to adaptive mutations.
-Comparison with biochemical studies of base pairing propensity in RNA
+Comparison with biochemical studies of base pairing propensity in an RNA
 genome of HIV-1 indicated that these mutations are deleterious, at least
@@
 The fixation probabilities and the sojourn times of SNVs represent a rich and simple
-summary statistics useful to characterize longitudinal sequence data and compare
-to models via computer simulations. These statistics are informative
-even in absence of a neutral control and thus suited to analyze
+summary statistics: they are useful to characterize longitudinal sequence data and
+can be compared to models via computer simulations. These statistics are informative
+even in absence of a neutral control and thus appropriate to analyze
 properties of synonymous sites. 
@@
 fact that the corresponding sites are not strongly conserved in
-cross-sectional data. This is consistent with a recent comparative
+cross sectional data. This is consistent with a recent comparative
 analysis of SIV and HIV-1 RNA secondary structure using SHAPE assays and
@@
 selection on proteins substantially \citep{ngandu_extensive_2008}. 
-The dynamic nature of HIV secondary structure makes finding the 
+The dynamic nature of HIV secondary structure makes it hard to find the 
 compensatory mutations that would restore base pairing in the
-longitudinal data difficult; in fact, the exact base pairing pattern is
+longitudinal data; in fact, the exact base pairing pattern is
 most likely different than in the reference sequence.

### 2013-06-06 synmut.tex
intro: dumped nonsyn and shrunk codon usage

@@
 \newcommand{\OO}{\mathcal{O}}
-\newcommand{\rev}{\textit{rev}}
 \newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
@@
 \newcommand{\env}{\textit{env}}
+\newcommand{\rev}{\textit{rev}}
+\newcommand{\pol}{\textit{pol}}
 \newcommand{\shankaregion}{C2-V5}
@@
 Intrapatient HIV-1 evolution is driven by the adaptive immune system
-resulting in rapid adaptation of HIV-1 proteins. When cytotoxic CD8${}^+$ T-cells
+resulting in rapid change of HIV-1 proteins. When cytotoxic CD8${}^+$ T-cells
 or neutralizing antibodies target a new epitope, the virus often escapes via
@@
 nuclear export, and packaging into the viral capsid: all these processes operate
-at the RNA level and are sensitive to synonymous changes. For example, the HIV-1
+at the RNA level and are sensitive to synonymous changes. For instance, the HIV-1
 \rev{} response element (RRE) in \env{} enhances nuclear export of full length
 or partially spliced viral transcripts via a complex hairpin RNA structure
-\citep{fernandes_hiv-1_2012}. Another well studied case is the interaction
-between viral reverse transcriptase, viral ssRNA, and the host
-tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription
-(RT) and is bound by a pseudoknotted RNA structure in the viral 5' untranslated
-region~\citep{barat_interaction_1991, paillart_vitro_2002}. 
-Besides these well characterized and conserved structures, large parts of the virus
-genome are folded into structures of unknown function
-\citep{watts_architecture_2009}. These poorly characterized RNA
-structures are conserved to varying degree between HIV-1 and
+\citep{fernandes_hiv-1_2012}. In fact, the HIV-1 genome is full of RNA
+structures \citep{watts_architecture_2009}, and several of them have been suggested
+to be functional and relevant for viral evolution \citep{keating_rich_2009, snoeck_mapping_2011,
+forsdyke_reciprocal_1995}, although their biology is not fully understood and
+their effects are sometimes hard to distinguish from codon usage bias.
+These poorly characterized RNA structures are conserved to varying degree between HIV-1 and
 SIV: corresponding regions tend to be part of
-of similar structural elements, but individual base pairings differ widely \citep{pollom_comparison_2013}.
-
-Even in the absence of important RNA structures, synonymous codons do not evolve
-completely neutrally. Some codons are favored in many species
-\citep{plotkin_synonymous_2011}. Recent studies have shown that genetically
-engineered HIV-1 strains with altered codon usage show differences in gene
-expression and replication capacity \citep{ngumbela_quantitative_2008,
-li_codon-usage-based_2012, keating_rich_2009}. Codon deoptimization has been
-suggested as an attenuation strategy for polio and
-influenza~\citep{mueller_live_2010,coleman_virus_2008}. Purifying selection
-beyond the protein sequence is therefore expected
-\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}, and it has been shown that
-rates of evolution at synonymous sites vary along the HIV-1 genome
-\citep{mayrose_towards_2007} (see also \figurename~S\syndiv).
+of similar structural elements, but individual base pairings are almost always
+discordant \citep{pollom_comparison_2013}.
 
@@
 synonymous mutations on viral fitness.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% DO WE WANT TO KEEP THIS LAST SENTENCE?
-Extending the analysis of fixation
-probabilities to the nonsynonymous variants, we show that time-dependent
-selection or competition of escape mutations inside the same epitope are
-necessary to explain the observed patterns of spread and extinction.
 
@@
 at synonymous sites varies greatly along the HIV-1 genome
-\citep{mayrose_towards_2007}. This variation can confound estimates of
+\citep{mayrose_towards_2007} (see also \figurename~S\syndiv). This variation can confound estimates of
 selection on proteins substantially \citep{ngandu_extensive_2008}. 

### 2013-06-06 synmut.tex
a few more little changes

@@
 in RNA stems flanking the variable loops of gp120 are more likely to be lost than other
-synonymous changes; these RNA hairpins might be important for HIV-1.
+synonymous changes: these RNA hairpins might be important for HIV-1.
 Computational modeling indicates that (i) most synonymous mutations in this
@@
 neutral SNV at frequency $\nu$ is the frequency itself, i.e.,
-$\pfix(\nu) = \nu$; it goes extinct with probability $1-\nu$. For
+$\pfix(\nu) = \nu$, while it goes extinct with probability $1-\nu$. For
 instance, if a neutral SNV is observed in half of the sequences, it will
@@
 against synonymous substitutions in the relatively conserved regions
-C2-C4 of the \env{} gene. Computational modeling suggested that these
+C2-C4. We suggest through computational modeling that these
 SNVs have deleterious effects on the order of $0.002$ and that they are

### 2013-06-06 synmut.tex
submission to JVI

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\documentclass[rmp, twocolumn]{revtex4}
+\documentclass[pre, twocolumn]{revtex4}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 affect this interplay and are often assumed to be neutral. We test this
-assumption by tracking the abundance of synonymous mutations in 
+assumption by tracking  synonymous mutations in 
 longitudinal intrapatient data from the \shankaregion{} part of the \env{}
@@
 commonly used as approximately neutral markers in studies of viral evolution,
-i.e. as a negative control for detecting selected
+i.e., as a negative control for detecting selected
 sites \citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In addition to
@@
 nuclear export, and packaging into the viral capsid: all these processes operate
-at the RNA level and are sensitive to synonymous changes. For instance, the HIV-1
-\rev{} response element (RRE) in \env{} enhances nuclear export of full length
+at the RNA level and are sensitive to synonymous changes. Several
+important RNA secondary structures have been characterized in detail, including the HIV-1
+\rev{} response element (RRE) in \env{} which enhances nuclear export of full length
 or partially spliced viral transcripts via a complex hairpin RNA structure
 \citep{fernandes_hiv-1_2012}. In fact, the HIV-1 genome is full of RNA
-structures \citep{watts_architecture_2009}, and several of them have been suggested
-to be functional and relevant for viral evolution \citep{keating_rich_2009, snoeck_mapping_2011,
-forsdyke_reciprocal_1995}, although their biology is not fully understood and
-their effects are sometimes hard to distinguish from codon usage bias.
+structures \citep{watts_architecture_2009} with no or unknown
+function. However, large scale modification of secondary structures
+can results in substantial reduction of the replication capacity
+\citep{keating_rich_2009} and the propensity of forming RNA stems
+anticorrelates with the rate of evolution
+\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}. 
 These poorly characterized RNA structures are conserved to varying degree between HIV-1 and
-SIV: corresponding regions tend to be part of
-of similar structural elements, but individual base pairings are almost always
-discordant \citep{pollom_comparison_2013}.
+SIV: corresponding regions tend to be part
+of similar structural elements, but individual base pairings are very
+rarely conserved \citep{pollom_comparison_2013}.
 
@@
 shown as a black dashed line. We found that $\pfix$ of synonymous
-variants is far below the neutral expectation in \shankaregion{} (red triangle line).
+variants is far below the neutral expectation in \shankaregion{} (red line).
 Outside of \shankaregion, using data from \citet{bunnik_autologous_2008} only,
-we found no such reduction in $\pfix$ (green square line). Restricted to the
+we found no such reduction in $\pfix$ (green line). Restricted to the
 \shankaregion{} region, the sequence samples from \citet{bunnik_autologous_2008}
@@
 nonsynonymous SNVs seem to follow more or less the neutral expectation (blue
-circle line) -- a point to which we come back below.
+line) -- a point to which we come back below.
 
@@
 always fix once they reach high frequencies, i.e., $A_{\mathrm{nonsyn}}$ is well
-above zero. This is incompatible with the blue circle line in \FIG{fixp2}: in an HIV-1
-infection, nonsynonymous SNVs at high frequency often disappear again, even
+above zero. This is incompatible with the observed fixation probability
+nonsynonymous SNVs at high frequency often disappear again (\FIG{fixp2}), even
 though many are at least transiently beneficial. Inspecting the trajectories of
@@
 brought to high frequency through linkage to adaptive mutations.
-Comparison with biochemical studies of base pairing propensity in an RNA
-genome of HIV-1 indicated that these mutations are deleterious, at least
-in part, because they disrupt RNA secondary structures. Computer models
+Comparison with biochemical data (SHAPE) of base pairing propensity in the RNA
+genome of HIV-1 indicated that these mutations tend to disrupt RNA secondary structures\citep{watts_architecture_2009}. Computer models
 of RNA folding predict stable hairpins in these regions that
@@
 Selection against the majority of synonymous substitutions is probably
-true across the genome, but we only observe deleterious synonymous SNVs
+common across the genome, but we only observe deleterious synonymous SNVs
 at high frequency in \shankaregion{} of \env{}, where they hitchhike to
@@
 the competition between different escape mutations within one epitope. 
-Both mechanisms reduce the overall fixation probability and could
-explain the observed pattern of fixation.
+Both mechanisms reduce the overall fixation probability and can
+give rise to the observed pattern of fixation in computer simulations.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliographystyle{natbib}
+%\bibliographystyle{unsrtnat}
 \bibliography{bib}

### 2013-06-06 synmut.tex
reformat for JVI

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\documentclass[pre, twocolumn]{revtex4}
+%\documentclass[pre, onecolumn, preprint]{revtex}
+\documentclass[11pt]{article}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \usepackage[caption=false]{subfig}
-\usepackage{natbib}
+\usepackage[numbers]{natbib}
 \usepackage{pslatex}
+\usepackage{lineno}
+\linenumbers
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\linespread{1.5}
 \begin{document}
 \title{\Title}
-\author{\Author}
-\affiliation{\Affiliation}
+\author{\Author\\ \Affiliation}
+%\affiliation{\Affiliation}
 \date{\today}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\maketitle
 
-\begin{abstract}
+\newpage
+%\begin{abstract}
+\section*{Abstract}
 \noindent
@@
 relevant, while the precise base pairing pattern is not.
-\end{abstract}
+%\end{abstract}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\maketitle
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section{Introduction}
+\section*{Introduction}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Results}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-Due to the large population size and the high mutation rate, every
-possible single nucleotide variant (SNV) is produced multiple times per
-day \citep{coffin_hiv_1995}. Some of these variants rise to high enough
-frequency that they are observed in a sample of sequences. SNVs rise or fall in 
-frequency because of three reasons: (i) their own effect on fitness and escape; (ii)
-their association to genetic backgrounds; (iii) stochastic fluctuations
-(genetic drift). We study the dynamics of sample frequencies, $\nu$, of SNVs,
-i.e., the fraction $\nu$ of the sequences in a sample carrying
-the variant. When an SNVs is present in all sequences 
-at a certain time point, we say it has ``fixed''; when it is completely absent,
-we say it was ``lost'' or is ``extinct''.
-
-Most positions are only transiently variable and variants will either fix or
-will be lost -- at least in small samples. Given an SNV is at a certain
-frequency $\nu$, the probability of fixation is higher for beneficial
-SNVs than for neutral ones; in turn, neutral variants fix more frequently than
-deleterious ones. The fixation probability of a
-neutral SNV at frequency $\nu$ is the frequency itself, i.e.,
-$\pfix(\nu) = \nu$, while it goes extinct with probability $1-\nu$. For
-instance, if a neutral SNV is observed in half of the sequences, it will
-fix with a probability of 50\% (see inset in figure
-\ref{fig:aftsyn}). The fixation probability of neutral SNVs is
-independent of most model assumptions and is only affected if neutral
-SNVs are associated preferentially with either high or low fitness
-virus.
-
-\FIG{aft} shows the time course of the frequencies of all synonymous
-SNVs (top) and nonsynonymous SNVs (bottom) observed in the
-\shankaregion{} region of \env{} in a chronically HIV-1 infected patient (p10 from
-\citet{shankarappa_consistent_1999}). Despite many synonymous SNVs
-reaching high frequency, very few fix (panel~\ref{fig:aftsyn}); in
-contrast, many nonsynonymous mutations fix
-(panel~\ref{fig:aftnonsyn}). This observation seems at odds with
-the assumption of neutrality.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Many synonymous SNVs in \shankaregion{} are deleterious}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-We studied the dynamics and fate of synonymous variants more quantitatively by
-analyzing data from seven patients from \citet{shankarappa_consistent_1999} and
-\citet{liu_selection_2006} as well as three patients from
-\citet{bunnik_autologous_2008} (patients with strong viral population structure
-were not considered; see methods and \figurename~S\PCApat). The
-former data set is restricted to the \shankaregion{} region of \env, while the
-data from \citet{bunnik_autologous_2008} cover most of \env.  We considered all
-SNVs in a frequency interval $[\nu_0-\delta\nu, \nu_0+\delta\nu]$ at some time
-$t$, and calculated the fraction that are still observed at later times $t+\Delta
-t$. Plotting this fraction against the time interval $\Delta t$, we see that
-most synonymous SNVs segregate for roughly one year and are lost much more
-frequently than expected under neutrality (panel \ref{fig:fixp1}). The long-time
-probability of fixation, $\pfix$, is shown as a function of the initial
-frequency $\nu_0$ in panel \ref{fig:fixp2}. As a control, the neutral result is
-shown as a black dashed line. We found that $\pfix$ of synonymous
-variants is far below the neutral expectation in \shankaregion{} (red line).
-Outside of \shankaregion, using data from \citet{bunnik_autologous_2008} only,
-we found no such reduction in $\pfix$ (green line). Restricted to the
-\shankaregion{} region, the sequence samples from \citet{bunnik_autologous_2008}
-are fully compatible with data from \citet{shankarappa_consistent_1999}:
-fixation is rare. The
-nonsynonymous SNVs seem to follow more or less the neutral expectation (blue
-line) -- a point to which we come back below.
-
-\begin{figure}
-\begin{center}
-\subfloat{\includegraphics[width=\linewidth]
-{Shankarappa_allele_freqs_trajectories_syn_p10.pdf}
-\label{fig:aftsyn}}\\
- \subfloat{\includegraphics[width=\linewidth]
-{Shankarappa_allele_freqs_trajectories_nonsyn_p10.pdf}
-\label{fig:aftnonsyn}}
-\caption{Time series of frequencies
-of synonymous (A) and nonsynonymous (B) single nucleotide variants (SNVs) in \env, 
-\shankaregion, from patient p10~\cite{shankarappa_consistent_1999}.
-While many nonsynonymous SNVs fix, few synonymous
-SNVs do so even though they are frequently observed at high
-frequencies. Colors indicate the position of the site along the \shankaregion{} region
-(blue to red). Inset: the fixation probability $\pfix$ of a neutral
-SNV that reached 50\% frequency is one half.}
-\label{fig:aft}
-\end{center}
-\end{figure}
-
-When interpreting these results for the fixation probabilities, it is important
-to note that we focused on SNVs that have already reached high frequencies. In
-HIV-1 infection, most SNVs remain very rare throughout: they are not considered
-here. Synonymous SNVs can reach high frequencies either through genetic
-drift or genetic hitchhiking on escape variants (see below); very deleterious
-variants will never reach high frequencies in the first place. Hence, our
-analysis indicates that, among all synonymous SNVs that somehow reach high
-frequencies, most of those in \shankaregion{} are deleterious, while
-those in the rest of \env{} tend to be neutral.
-
-\begin{figure}
-\begin{center}
-\subfloat{\includegraphics[width=0.9\linewidth]{fixation_times.pdf}
-\label{fig:fixp1}}\\
-\subfloat{\includegraphics[width=0.9\linewidth]{fixation_probabilities.pdf}
-\label{fig:fixp2}}
-\caption{Fixation and loss of SNVs.
-Panel A) shows how quickly synonymous SNVs are purged from the populations. 
-Specifically, the figure shows the fraction of SNVs that are still observed
-after $\Delta t$ days, conditional on being observed in one of the three frequency 
-intervals (different colors). 
-In each frequency interval, the fraction of synonymous
-SNVs that ultimately survive is the fixation probability $\pfix$ conditional on the
-initial frequency. The neutral expectation for $\pfix=\nu_0$ is indicated by 
-dashed horizontal lines.
-Panel B) shows the fixation probability of synonymous SNVs as a function of $\nu_0$. Polymorphisms within \shankaregion{} fix less
-often than expected for neutral SNVs (indicated by the diagonal line).
-This suppression is not observed in other parts of \env{} or for nonsynonymous
-SNVs.
-The horizontal error bars on the abscissa are bin sizes, the vertical ones the
-standard deviation after 100 patient bootstraps of the data. Data from
-refs.~\cite{shankarappa_consistent_1999,liu_selection_2006, bunnik_autologous_2008}.}
-\label{fig:fixp}
-\end{center}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Synonymous mutations in \shankaregion{} tend to disrupt RNA stems}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-One possible explanation for a reduced fixation of synonymous variants in
-\shankaregion{} is secondary structures in the viral RNA, the disruption of which
-is deleterious to the virus \citep{forsdyke_reciprocal_1995,
-snoeck_mapping_2011, sanjuan_interplay_2011}.
-
-The propensity of nucleotides in the HIV-1 genome to form base pairs has been
-measured using the SHAPE assay, a biochemical reaction preferentially altering
-unpaired bases 
-\citep{watts_architecture_2009}. The SHAPE assay has shown that the variable
-regions V1-V5 tend to be unpaired, while the conserved regions between those
-stretches form stems. We aligned the sequences from each patient to
-the reference NL4-3 strain used in \citet{watts_architecture_2009} and assigned
-SHAPE reactivities to most positions in the alignment. We then calculated the
-distributions of SHAPE reactivities for synonymous SNVs that fixed or were
-lost (only variants reaching frequencies above 15\%). As shown in
-\FIG{SHAPEA}, the reactivities of fixed SNVs (red histogram) are systematically
-larger than those of lost SNVs (blue) (Kolmogorov-Smirnov test on the cumulative
-distribution, $p\approx 0.002$). In other words, SNVs that are likely to
-break RNA helices are also more likely to revert and finally be lost from the
-population, restoring the helix. Note that this analysis will be sensitive only
-at positions where the base pairing pattern of NL4-3 agrees with that of each
-patient's initial consensus sequence -- it is thus statistically conservative.
-As a control, we also calculate the distribution of SHAPE reactivities for SNVs
-that never reach high frequencies (green). This set is a mixture of neutral and
-deleterious SNVs and, as expected, its distribution lies between those of
-fixed and lost high-frequency variants.
-
-To test the hypothesis that synonymous variants in \shankaregion{} are lost because they
-break stems in the conserved stretches between the V loops, we considered
-separately SNVs in V loops and their flanks. The greatest
-depression in fixation probability is observed in the conserved stems, while the
-V loops show little deviation from the neutral signature, see
-\FIG{SHAPEB}. This is suggestive of important RNA helices in conserved
-regions between the V loops.
-
-In addition to RNA secondary structure, we have considered other possible
-explanations for a fitness cost of some synonymous mutations, in particular
-codon usage bias (CUB). HIV-1 is known to prefer A-rich codons over highly
-expressed human codons \citep{jenkins_extent_2003, kuyl_biased_2012}. We
-did not find any evidence for a contribution of average CUB to the ultimate
-fate of synonymous SNVs; this is consistent with the observation that HIV-1 is not
-adapting its codon usage to its human host cells at the macro-evolutionary level
-\citep{kuyl_biased_2012}.
-
-\begin{figure}
-\begin{center}
-\subfloat{\includegraphics[height=0.50\linewidth]{reactivities_histograms_syn.pdf}\label{fig:SHAPEA}}
-\subfloat{\includegraphics[height=0.49\linewidth]{fixation_probabilities_VnonV.pdf}\label{fig:SHAPEB}}
-\caption{Permissible synonymous mutations tend to be unpaired.
-Panel A) shows the distribution of SHAPE reactivities among sites at which synonymous 
-SNVs fixed (red), sites at which SNVs reached frequencies above 15\% but
-were subsequently lost (blue), and sites at which no high-frequency SNVs were observed (green) 
-(all categories are restricted to the regions V1-V5$\pm 100$bp).
-Sites at which SNVs fixed tend to have higher SHAPE reactivities, corresponding to
-less base pairing, than those at which SNVs are lost.
-Sites at which no SNVs are observed show an intermediate distribution of SHAPE values.
-Panel B) shows the fixation probability of synonymous SNVs in
-\shankaregion{} separately for variable regions V3-V5 and the connecting conserved 
-regions C2-C4 that harbor RNA stems. As expected, the fixation probability is lower
-inside the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
-bunnik_autologous_2008, liu_selection_2006}.}
-\label{fig:SHAPE}
-\end{center}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Deleterious SNVs reach high frequency by hitchhiking}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-While the observation of some deleterious synonymous variants is not fully
-unexpected, it seems odd that we observe them at high population
-frequency and that the fixation probability is reduced only in parts of the
-genome (in \shankaregion{} but not in the rest of \env{}; compare the red
-triangle line versus the green square line in \FIG{fixp2}).
-The region \shankaregion{} undergoes frequent adaptive changes to evade
-recognition by neutralizing antibodies \cite{williamson_adaptation_2003,
-richman_rapid_2003}. Due to the limited amount of recombination in HIV-1
-\cite{neher_recombination_2010, batorsky_estimate_2011}, deleterious variants
-that are linked to adaptive variants can reach high frequency. This process is
-known as hitchhiking \citep{smith_hitch-hiking_1974} or genetic draft
-\citep{gillespie_genetic_2000,neher_genetic_2011}. Hitchhiking is apparent in
-\FIG{aft}, which shows that many SNVs change rapidly in frequency as a
-flock.
-
-The approximate magnitude of the deleterious effects can be estimated from
-\FIG{fixp1}, which shows the distribution of times after which synonymous
-SNVs at intermediate frequencies become fixed or lost. The typical time to
-loss is of the order of 500 days. If this loss is driven by the deleterious
-effect of the mutation, this corresponds to deleterious effects $s_d$ of the
-order of $- 0.002$ per day. (This is only an average estimate: every single
-mutation is expected to have a slightly different fitness effect.)
-
-\begin{figure}
-\begin{center}
-\subfloat{\includegraphics[width=0.9\linewidth]{simulations_graduallydel.pdf}
-\label{fig:simfixpvar}}\\
-\subfloat{\includegraphics[width=0.9\linewidth]{simulations_syn.pdf}
-\label{fig:simsfig}}
-\caption{Distribution of selection coefficients on synonymous sites. Panel A)
-The depression in $\pfix$ depends on the deleterious effect size 
-of synonymous SNVs. This parameter also reduces synonymous
-diversity, measured by the probability of a SNV to be found at
-intermediate frequencies $P_\text{interm}$ (first inset).
-Panel B) To assess the parameter space that affects synonymous fixation and
-diversity, we run 2400 simulations with random parameters for deleterious effect
-size, fraction of deleterious synonymous sites, average escape rate $\epsilon$
-(color, blue to red corresponds to $10^{-2.5}$ to $10^{-1.5}$ per day), and rate of
-introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
-day). Only simulations that reproduce the synonymous diversity and fixation
-patterns observed in data are shown. These simulations demonstrate that
-deleterious effects are around $-0.002$ and a large fraction of the 
-synonymous mutations needs to be deleterious. As expected, larger
-$s_d$ require larger $\epsilon$. Parameters are chosen
-from prior distributions uniform in logspace as indicated by the red rectangle
-(see methods).}
-\label{fig:simheat}
-\end{center}
-\end{figure}
-
-To get a better idea of the range of parameters that are compatible with the
-observations and our interpretation, we performed computer simulations of
-evolving viral populations assuming a mix of positive and purifying selection
-and rare recombination.  For this purpose, we use the simulation package
-FFPopSim, which includes a module dedicated to intrapatient HIV evolution
-\citep{zanini_ffpopsim:_2012}. For each simulation run, we specify the
-deleterious effect of synonymous mutations, the fraction of synonymous mutations
-that are deleterious, the escape rate (selection coefficient) of adaptive
-nonsynonymous mutations and the rate at which previously untargeted epitopes
-become targeted (the latter determines the number of sites available for
-escape). Note that the escape rate is the sum of two factors: (i) the beneficial
-effect due to the ability to evade the immune system minus (ii) the fitness cost
-of the mutation in terms of structure, stability, etc. Net escape rates in
-chronic infections have been estimated to be on the order of $\epsilon = 0.01$
-per day \citep{neher_recombination_2010, Asquith:2006p28003}, consistent
-with a lag in neutralization of a few month \citep{richman_rapid_2003}.
-
-\FIG{simfixpvar} shows simulation results for the fixation probability and the
-synonymous diversity for different deleterious effects sizes of synonymous mutations.
-We quantify synonymous diversity via $P_\text{interm}$, the fraction of sites
-with an synonymous SNV at frequency $0.25 < \nu < 0.75$. The synonymous diversity
-observed in patient data is indicated in the figure. To quantify the depression
-of the fixation probability, we calculate the area between the measured fixation
-probability and the diagonal, which is the neutral expectation
-(\FIG{simfixpvar}, lower inset). If no fixation happens, the area will be
-$-0.5$; if every SNV fixes, the area will be $+0.5$. In HIV-1 infected
-patients, we find $P_\text{interm} \approx 0.005$, $A_\text{syn} \approx -0.2$
-for synonymous changes and $A_\text{nonsyn} \approx 0$ for nonsynonymous
-changes. In the three simulations shown in \FIG{simfixpvar}, the fixation
-probability of synonymous SNVs decreases from the neutral expectation
-($A_\text{syn} \approx 0$) to zero ($A_\text{syn} \approx -0.5$) as the
-fitness cost of the SNVs increases; the synonymous diversity plummets as well, as
-deleterious SNVs are selected against.
-
-To map the parameter range of the model that is compatible with the data, we
-repeatedly simulated the evolution with random choices for the
-parameters, see \FIG{simsfig}. Among all simulations, we selected the ones
-that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
-large depression in fixation probability of synonymous SNVs but, simultaneously,
-a moderately high synonymous diversity. Specifically, \FIG{simsfig} shows
-parameter combinations for which we found $A_\text{syn} < -0.15$ and $0.0025 <
-P_\text{interm} < 0.010$. This subset of parameters indicates that a high fraction
-($\gtrsim 0.8$) of sites has to be deleterious with effect size $|s_d| \sim
-0.002$. This is consistent with the fixation/extinction times estimated above
-(see \FIG{fixp1}).
-
-The bounds on the fraction of deleterious synonymous SNVs and
-on their effect sizes are plausible:
-(i) a substantial depression in $\pfix$ requires pervasive deleterious SNVs, otherwise
-the majority of SNVs reaching high frequency are neutral and no depression is
-observed; (ii) in order to hitchhike, the deleterious effect size has to be much
-smaller than the escape rate, otherwise the double mutant (with both the escape
-mutation and the deleterious synonymous one) has little or no fitness advantage
-over the wildtype virus. In agreement with this argument, larger deleterious
-effects in \FIG{simsfig} correspond to larger escape rates; and (iii) SNVs
-with a deleterious effect smaller than approximately $0.001$ behave neutrally,
-as expected from the typical coalescent times observed in HIV-1.
-
-The above simulations show that hitchhiking with favoured nonsynonymous
-variants can explain the observation of
-deleterious synonymous SNVs that rarely fix. However, in a simple model where
-nonsynonymous escape mutations are unconditionally beneficial, escape mutations almost
-always fix once they reach high frequencies, i.e., $A_{\mathrm{nonsyn}}$ is well
-above zero. This is incompatible with the observed fixation probability
-nonsynonymous SNVs at high frequency often disappear again (\FIG{fixp2}), even
-though many are at least transiently beneficial. Inspecting the trajectories of
-nonsynonymous SNVs suggests the rapid rise and fall of many SNVs. Two
-possible mechanisms that could explain the transient rise of
-nonsynonymous SNVs are time-dependent selection and within-epitope competition.
-
-If the immune system
-recognizes the escape mutant before its fixation, the mutant might cease to be
-beneficial and disappear soon, despite its quick initial rise in frequency. In
-support of this idea, in \citet{richman_rapid_2003, bunnik_autologous_2008},
-antibody responses against escape mutants have been reported. These responses are
-delayed by a few months, roughly matching the average time needed by an escape
-mutant to rise from low to high frequency. To model this type of behavior, we
-assumed that antibody responses against escape SNVs arise with a rate
-proportional to the frequency of the escape variant and abolish the benefit of the
-escape mutations. As expected, this type of time-dependent selection retained the
-potential for hitchhiking, but reduced fixation of nonsynonymous SNVs.
-\figurename~S\timedependence~shows that $\pfix$ of synonymous SNVs is not
-affected by this change, while $\pfix$ of nonsynonymous SNVs approaches the
-diagonal as the rate of recognition of escape mutants is increased.
-
-Several different escape mutations that arise within the 
-same epitope have similar effects. Their benefits
-are not additive, because each of them is essentially sufficient to escape and
-no additional benefit is gained from combining them. As a consequence, several
-escape SNVs rise to high frequency rapidly, while the one with the smallest cost
-in terms of replication, packaging, etc.~is most likely to eventually fix, while
-all others are lost. The emergence of multiple competing escape variants in HIV-1
-infections has been shown \citep{moore_limited_2009, bar_early_2012}. 
-Similarly, this scenario has been explicitly observed in the evolution of
-resistance to 3TC, where the mutation M184V is often preceded by M184I
-\citep{hedskog_dynamics_2010}. We implemented within epitope competition 
-in the model by allowing for multiple escape mutations per epitope that do
-not provide additional benefit to the virus when combined. Again, we found that the potential for
-hitchhiking is retained by within-epitope competition but that the
-fixation probability of nonsynonymous SNVs is reduced. With roughly six
-mutations per epitope, the simulation data are compatible with observations; see
-\figurename~S\withinepi. The two scenarios, time-dependent selection and
-competition between equivalent escape pathways, are not exclusive and possibly
-both important in HIV-1 evolution.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Discussion}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-By analyzing the fate of single nucleotide variants (SNVs) in
-longitudinal data of HIV-1 \env{} evolution, we demonstrated selection
-against synonymous substitutions in the relatively conserved regions
-C2-C4. We suggest through computational modeling that these
-SNVs have deleterious effects on the order of $0.002$ and that they are
-brought to high frequency through linkage to adaptive mutations.
-Comparison with biochemical data (SHAPE) of base pairing propensity in the RNA
-genome of HIV-1 indicated that these mutations tend to disrupt RNA secondary structures\citep{watts_architecture_2009}. Computer models
-of RNA folding predict stable hairpins in these regions that
-have been suggested to be functional and termed ``insulating
-stems'' \citep{watts_architecture_2009, sanjuan_interplay_2011}.
-The weak selection against synonymous variants is compatible with the
-negative results of \textit{In vitro} replication assays investigating
-the fitness effects of small RNA hairpins in HIV-1
-\citep{knoepfel_role_2013}: it would take hundreds of cell culture
-passages to detect fitness effects of the order of one per mille. The longitudinal data,
-however, spans many years, hence our analysis is able to quantify the
-subtle fitness effect of RNA structure within single infections.
-
-The fixation probabilities and the sojourn times of SNVs represent a rich and simple
-summary statistics: they are useful to characterize longitudinal sequence data and
-can be compared to models via computer simulations. These statistics are informative
-even in absence of a neutral control and thus appropriate to analyze
-properties of synonymous sites. 
-We find selection against synonymous substitutions despite the
-fact that the corresponding sites are not strongly conserved in
-cross sectional data. This is consistent with a recent comparative
-analysis of SIV and HIV-1 RNA secondary structure using SHAPE assays and
-computational methods \citep{pollom_comparison_2013}. While large scale
-patterns of RNA structures tend to agree in both viruses, the individual
-base pairs forming the structures are almost always
-discordant. Even though the molecular architecture of these structures
-changes over time, selection seems to maintain them, which reduces the
-fixation probability and hence the rate of evolution at synonymous
-sites. As expected from this argument, the evolutionary rate
-at synonymous sites varies greatly along the HIV-1 genome
-\citep{mayrose_towards_2007} (see also \figurename~S\syndiv). This variation can confound estimates of
-selection on proteins substantially \citep{ngandu_extensive_2008}. 
-The dynamic nature of HIV secondary structure makes it hard to find the 
-compensatory mutations that would restore base pairing in the
-longitudinal data; in fact, the exact base pairing pattern is
-most likely different than in the reference sequence.
-
-Selection against the majority of synonymous substitutions is probably
-common across the genome, but we only observe deleterious synonymous SNVs
-at high frequency in \shankaregion{} of \env{}, where they hitchhike to
-high frequencies on nAb escape mutations. Surprisingly, nonsynonymous
-mutations display a fixation pattern as if they were neutral;
-see~\FIG{fixp2}. However, nonsynonymous diversity exceeds
-synonymous diversity despite the overall much greater constraints on the
-amino acid sequence, suggesting that the majority of high frequency SNVs
-are escape mutations despite the fact that are often lost again. We suggest that this
-paradoxical behavior could be due to (i) escape
-mutations that revert after they themselves are recognized by nAbs, or (ii)
-the competition between different escape mutations within one epitope. 
-Both mechanisms reduce the overall fixation probability and can
-give rise to the observed pattern of fixation in computer simulations.
-
-The observed hitchhiking highlights the importance of linkage due to
-infrequent recombination for the evolution of HIV-1
-\citep{neher_recombination_2010, batorsky_estimate_2011,
-josefsson_majority_2011}. The recombination rate has been estimated to be on the
-order of $\rho = 10^{-5}$ per base and day. It takes roughly $t_{sw} =
-\epsilon^{-1} \log \nu_0$ generations for an escape SNV with escape rate
-$\epsilon$ to rise from an initially low frequency $\nu_0\approx \mu$ to frequency
-one. This implies that a region of length $l = (\rho t_{sw})^{-1} = \epsilon /
-\rho \log \nu_0$ remains linked to the adaptive mutation. With $\epsilon=0.01$,
-we have $l\approx 100$ bases. Hence we expect strong linkage between the
-variable loops and the flanking sequences, but none far beyond the variable
-regions, consistent with the lack of signal outside of \shankaregion. In case of
-much stronger selection -- such as observed during early CTL escape or drug
-resistance evolution -- the linked region is of course much larger
-\citep{nijhuis_stochastic_1998}.
-
-While classical population genetics assumes that the dominant stochastic force
-is genetic drift, i.e., non-heritable fluctuations in offspring number, our
-results show that stochasticity due to linked selection is much more important.
-Such fluctuations have been termed \emph{genetic draft} in
-\citet{gillespie_genetic_2000} and their effect in facultatively sexual population
-such as HIV-1 has been characterized in \citep{neher_genetic_2011}. Importantly,
-large population sizes are compatible with low diversity and fast coalescence
-when draft dominates over drift.
-
-Our results emphasize the inadequacy of independent site models of HIV-1 evolution
-and the common assumption that selection is time independent or additive. 
-If genetic variation is only transiently beneficial, existing estimates of the
-strength of selection \citep{neher_recombination_2010,batorsky_estimate_2011}
-could be substantial underestimates. Furthermore, weak conservation and
-time-dependent selection result in estimates of evolutionary 
-rates that depend on the time interval of observation, with lower rates across
-larger intervals. This implies that deep nodes in phylogenies might be older than 
-they appear.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Methods}
+\section*{Materials \& Methods}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Results}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+Due to the large population size and the high mutation rate, every
+possible single nucleotide variant (SNV) is produced multiple times per
+day \citep{coffin_hiv_1995}. Some of these variants rise to high enough
+frequency that they are observed in a sample of sequences. SNVs rise or fall in 
+frequency because of three reasons: (i) their own effect on fitness and escape; (ii)
+their association to genetic backgrounds; (iii) stochastic fluctuations
+(genetic drift). We study the dynamics of sample frequencies, $\nu$, of SNVs,
+i.e., the fraction $\nu$ of the sequences in a sample carrying
+the variant. When an SNVs is present in all sequences 
+at a certain time point, we say it has ``fixed''; when it is completely absent,
+we say it was ``lost'' or is ``extinct''.
+
+Most positions are only transiently variable and variants will either fix or
+will be lost -- at least in small samples. Given an SNV is at a certain
+frequency $\nu$, the probability of fixation is higher for beneficial
+SNVs than for neutral ones; in turn, neutral variants fix more frequently than
+deleterious ones. The fixation probability of a
+neutral SNV at frequency $\nu$ is the frequency itself, i.e.,
+$\pfix(\nu) = \nu$, while it goes extinct with probability $1-\nu$. For
+instance, if a neutral SNV is observed in half of the sequences, it will
+fix with a probability of 50\% (see inset in figure
+\ref{fig:aftsyn}). The fixation probability of neutral SNVs is
+independent of most model assumptions and is only affected if neutral
+SNVs are associated preferentially with either high or low fitness
+virus.
+
+\FIG{aft} shows the time course of the frequencies of all synonymous
+SNVs (top) and nonsynonymous SNVs (bottom) observed in the
+\shankaregion{} region of \env{} in a chronically HIV-1 infected patient (p10 from
+\citet{shankarappa_consistent_1999}). Despite many synonymous SNVs
+reaching high frequency, very few fix (panel~\ref{fig:aftsyn}); in
+contrast, many nonsynonymous mutations fix
+(panel~\ref{fig:aftnonsyn}). This observation seems at odds with
+the assumption of neutrality.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Many synonymous SNVs in \shankaregion{} are deleterious}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+We studied the dynamics and fate of synonymous variants more quantitatively by
+analyzing data from seven patients from \citet{shankarappa_consistent_1999} and
+\citet{liu_selection_2006} as well as three patients from
+\citet{bunnik_autologous_2008} (patients with strong viral population structure
+were not considered; see methods and \figurename~S\PCApat). The
+former data set is restricted to the \shankaregion{} region of \env, while the
+data from \citet{bunnik_autologous_2008} cover most of \env.  We considered all
+SNVs in a frequency interval $[\nu_0-\delta\nu, \nu_0+\delta\nu]$ at some time
+$t$, and calculated the fraction that are still observed at later times $t+\Delta
+t$. Plotting this fraction against the time interval $\Delta t$, we see that
+most synonymous SNVs segregate for roughly one year and are lost much more
+frequently than expected under neutrality (panel \ref{fig:fixp1}). The long-time
+probability of fixation, $\pfix$, is shown as a function of the initial
+frequency $\nu_0$ in panel \ref{fig:fixp2}. As a control, the neutral result is
+shown as a black dashed line. We found that $\pfix$ of synonymous
+variants is far below the neutral expectation in \shankaregion{} (red line).
+Outside of \shankaregion, using data from \citet{bunnik_autologous_2008} only,
+we found no such reduction in $\pfix$ (green line). Restricted to the
+\shankaregion{} region, the sequence samples from \citet{bunnik_autologous_2008}
+are fully compatible with data from \citet{shankarappa_consistent_1999}:
+fixation is rare. The
+nonsynonymous SNVs seem to follow more or less the neutral expectation (blue
+line) -- a point to which we come back below.
+
+When interpreting these results for the fixation probabilities, it is important
+to note that we focused on SNVs that have already reached high frequencies. In
+HIV-1 infection, most SNVs remain very rare throughout: they are not considered
+here. Synonymous SNVs can reach high frequencies either through genetic
+drift or genetic hitchhiking on escape variants (see below); very deleterious
+variants will never reach high frequencies in the first place. Hence, our
+analysis indicates that, among all synonymous SNVs that somehow reach high
+frequencies, most of those in \shankaregion{} are deleterious, while
+those in the rest of \env{} tend to be neutral.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Synonymous mutations in \shankaregion{} tend to disrupt RNA stems}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+One possible explanation for a reduced fixation of synonymous variants in
+\shankaregion{} is secondary structures in the viral RNA, the disruption of which
+is deleterious to the virus \citep{forsdyke_reciprocal_1995,
+snoeck_mapping_2011, sanjuan_interplay_2011}.
+
+The propensity of nucleotides in the HIV-1 genome to form base pairs has been
+measured using the SHAPE assay, a biochemical reaction preferentially altering
+unpaired bases 
+\citep{watts_architecture_2009}. The SHAPE assay has shown that the variable
+regions V1-V5 tend to be unpaired, while the conserved regions between those
+stretches form stems. We aligned the sequences from each patient to
+the reference NL4-3 strain used in \citet{watts_architecture_2009} and assigned
+SHAPE reactivities to most positions in the alignment. We then calculated the
+distributions of SHAPE reactivities for synonymous SNVs that fixed or were
+lost (only variants reaching frequencies above 15\%). As shown in
+\FIG{SHAPEA}, the reactivities of fixed SNVs (red histogram) are systematically
+larger than those of lost SNVs (blue) (Kolmogorov-Smirnov test on the cumulative
+distribution, $p\approx 0.002$). In other words, SNVs that are likely to
+break RNA helices are also more likely to revert and finally be lost from the
+population, restoring the helix. Note that this analysis will be sensitive only
+at positions where the base pairing pattern of NL4-3 agrees with that of each
+patient's initial consensus sequence -- it is thus statistically conservative.
+As a control, we also calculate the distribution of SHAPE reactivities for SNVs
+that never reach high frequencies (green). This set is a mixture of neutral and
+deleterious SNVs and, as expected, its distribution lies between those of
+fixed and lost high-frequency variants.
+
+To test the hypothesis that synonymous variants in \shankaregion{} are lost because they
+break stems in the conserved stretches between the V loops, we considered
+separately SNVs in V loops and their flanks. The greatest
+depression in fixation probability is observed in the conserved stems, while the
+V loops show little deviation from the neutral signature, see
+\FIG{SHAPEB}. This is suggestive of important RNA helices in conserved
+regions between the V loops.
+
+In addition to RNA secondary structure, we have considered other possible
+explanations for a fitness cost of some synonymous mutations, in particular
+codon usage bias (CUB). HIV-1 is known to prefer A-rich codons over highly
+expressed human codons \citep{jenkins_extent_2003, kuyl_biased_2012}. We
+did not find any evidence for a contribution of average CUB to the ultimate
+fate of synonymous SNVs; this is consistent with the observation that HIV-1 is not
+adapting its codon usage to its human host cells at the macro-evolutionary level
+\citep{kuyl_biased_2012}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Deleterious SNVs reach high frequency by hitchhiking}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+While the observation of some deleterious synonymous variants is not fully
+unexpected, it seems odd that we observe them at high population
+frequency and that the fixation probability is reduced only in parts of the
+genome (in \shankaregion{} but not in the rest of \env{}; compare the red
+triangle line versus the green square line in \FIG{fixp2}).
+The region \shankaregion{} undergoes frequent adaptive changes to evade
+recognition by neutralizing antibodies \cite{williamson_adaptation_2003,
+richman_rapid_2003}. Due to the limited amount of recombination in HIV-1
+\cite{neher_recombination_2010, batorsky_estimate_2011}, deleterious variants
+that are linked to adaptive variants can reach high frequency. This process is
+known as hitchhiking \citep{smith_hitch-hiking_1974} or genetic draft
+\citep{gillespie_genetic_2000,neher_genetic_2011}. Hitchhiking is apparent in
+\FIG{aft}, which shows that many SNVs change rapidly in frequency as a
+flock.
+
+The approximate magnitude of the deleterious effects can be estimated from
+\FIG{fixp1}, which shows the distribution of times after which synonymous
+SNVs at intermediate frequencies become fixed or lost. The typical time to
+loss is of the order of 500 days. If this loss is driven by the deleterious
+effect of the mutation, this corresponds to deleterious effects $s_d$ of the
+order of $- 0.002$ per day. (This is only an average estimate: every single
+mutation is expected to have a slightly different fitness effect.)
+
+To get a better idea of the range of parameters that are compatible with the
+observations and our interpretation, we performed computer simulations of
+evolving viral populations assuming a mix of positive and purifying selection
+and rare recombination.  For this purpose, we use the simulation package
+FFPopSim, which includes a module dedicated to intrapatient HIV evolution
+\citep{zanini_ffpopsim:_2012}. For each simulation run, we specify the
+deleterious effect of synonymous mutations, the fraction of synonymous mutations
+that are deleterious, the escape rate (selection coefficient) of adaptive
+nonsynonymous mutations and the rate at which previously untargeted epitopes
+become targeted (the latter determines the number of sites available for
+escape). Note that the escape rate is the sum of two factors: (i) the beneficial
+effect due to the ability to evade the immune system minus (ii) the fitness cost
+of the mutation in terms of structure, stability, etc. Net escape rates in
+chronic infections have been estimated to be on the order of $\epsilon = 0.01$
+per day \citep{neher_recombination_2010, Asquith:2006p28003}, consistent
+with a lag in neutralization of a few month \citep{richman_rapid_2003}.
+
+\FIG{simfixpvar} shows simulation results for the fixation probability and the
+synonymous diversity for different deleterious effects sizes of synonymous mutations.
+We quantify synonymous diversity via $P_\text{interm}$, the fraction of sites
+with an synonymous SNV at frequency $0.25 < \nu < 0.75$. The synonymous diversity
+observed in patient data is indicated in the figure. To quantify the depression
+of the fixation probability, we calculate the area between the measured fixation
+probability and the diagonal, which is the neutral expectation
+(\FIG{simfixpvar}, lower inset). If no fixation happens, the area will be
+$-0.5$; if every SNV fixes, the area will be $+0.5$. In HIV-1 infected
+patients, we find $P_\text{interm} \approx 0.005$, $A_\text{syn} \approx -0.2$
+for synonymous changes and $A_\text{nonsyn} \approx 0$ for nonsynonymous
+changes. In the three simulations shown in \FIG{simfixpvar}, the fixation
+probability of synonymous SNVs decreases from the neutral expectation
+($A_\text{syn} \approx 0$) to zero ($A_\text{syn} \approx -0.5$) as the
+fitness cost of the SNVs increases; the synonymous diversity plummets as well, as
+deleterious SNVs are selected against.
+
+To map the parameter range of the model that is compatible with the data, we
+repeatedly simulated the evolution with random choices for the
+parameters, see \FIG{simsfig}. Among all simulations, we selected the ones
+that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
+large depression in fixation probability of synonymous SNVs but, simultaneously,
+a moderately high synonymous diversity. Specifically, \FIG{simsfig} shows
+parameter combinations for which we found $A_\text{syn} < -0.15$ and $0.0025 <
+P_\text{interm} < 0.010$. This subset of parameters indicates that a high fraction
+($\gtrsim 0.8$) of sites has to be deleterious with effect size $|s_d| \sim
+0.002$. This is consistent with the fixation/extinction times estimated above
+(see \FIG{fixp1}).
+
+The bounds on the fraction of deleterious synonymous SNVs and
+on their effect sizes are plausible:
+(i) a substantial depression in $\pfix$ requires pervasive deleterious SNVs, otherwise
+the majority of SNVs reaching high frequency are neutral and no depression is
+observed; (ii) in order to hitchhike, the deleterious effect size has to be much
+smaller than the escape rate, otherwise the double mutant (with both the escape
+mutation and the deleterious synonymous one) has little or no fitness advantage
+over the wildtype virus. In agreement with this argument, larger deleterious
+effects in \FIG{simsfig} correspond to larger escape rates; and (iii) SNVs
+with a deleterious effect smaller than approximately $0.001$ behave neutrally,
+as expected from the typical coalescent times observed in HIV-1.
+
+The above simulations show that hitchhiking with favoured nonsynonymous
+variants can explain the observation of
+deleterious synonymous SNVs that rarely fix. However, in a simple model where
+nonsynonymous escape mutations are unconditionally beneficial, escape mutations almost
+always fix once they reach high frequencies, i.e., $A_{\mathrm{nonsyn}}$ is well
+above zero. This is incompatible with the observed fixation probability
+nonsynonymous SNVs at high frequency often disappear again (\FIG{fixp2}), even
+though many are at least transiently beneficial. Inspecting the trajectories of
+nonsynonymous SNVs suggests the rapid rise and fall of many SNVs. Two
+possible mechanisms that could explain the transient rise of
+nonsynonymous SNVs are time-dependent selection and within-epitope competition.
+
+If the immune system
+recognizes the escape mutant before its fixation, the mutant might cease to be
+beneficial and disappear soon, despite its quick initial rise in frequency. In
+support of this idea, in \citet{richman_rapid_2003, bunnik_autologous_2008},
+antibody responses against escape mutants have been reported. These responses are
+delayed by a few months, roughly matching the average time needed by an escape
+mutant to rise from low to high frequency. To model this type of behavior, we
+assumed that antibody responses against escape SNVs arise with a rate
+proportional to the frequency of the escape variant and abolish the benefit of the
+escape mutations. As expected, this type of time-dependent selection retained the
+potential for hitchhiking, but reduced fixation of nonsynonymous SNVs.
+\figurename~S\timedependence~shows that $\pfix$ of synonymous SNVs is not
+affected by this change, while $\pfix$ of nonsynonymous SNVs approaches the
+diagonal as the rate of recognition of escape mutants is increased.
+
+Several different escape mutations that arise within the 
+same epitope have similar effects. Their benefits
+are not additive, because each of them is essentially sufficient to escape and
+no additional benefit is gained from combining them. As a consequence, several
+escape SNVs rise to high frequency rapidly, while the one with the smallest cost
+in terms of replication, packaging, etc.~is most likely to eventually fix, while
+all others are lost. The emergence of multiple competing escape variants in HIV-1
+infections has been shown \citep{moore_limited_2009, bar_early_2012}. 
+Similarly, this scenario has been explicitly observed in the evolution of
+resistance to 3TC, where the mutation M184V is often preceded by M184I
+\citep{hedskog_dynamics_2010}. We implemented within epitope competition 
+in the model by allowing for multiple escape mutations per epitope that do
+not provide additional benefit to the virus when combined. Again, we found that the potential for
+hitchhiking is retained by within-epitope competition but that the
+fixation probability of nonsynonymous SNVs is reduced. With roughly six
+mutations per epitope, the simulation data are compatible with observations; see
+\figurename~S\withinepi. The two scenarios, time-dependent selection and
+competition between equivalent escape pathways, are not exclusive and possibly
+both important in HIV-1 evolution.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Discussion}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+By analyzing the fate of single nucleotide variants (SNVs) in
+longitudinal data of HIV-1 \env{} evolution, we demonstrated selection
+against synonymous substitutions in the relatively conserved regions
+C2-C4. We suggest through computational modeling that these
+SNVs have deleterious effects on the order of $0.002$ and that they are
+brought to high frequency through linkage to adaptive mutations.
+Comparison with biochemical data (SHAPE) of base pairing propensity in the RNA
+genome of HIV-1 indicated that these mutations tend to disrupt RNA secondary structures\citep{watts_architecture_2009}. Computer models
+of RNA folding predict stable hairpins in these regions that
+have been suggested to be functional and termed ``insulating
+stems'' \citep{watts_architecture_2009, sanjuan_interplay_2011}.
+The weak selection against synonymous variants is compatible with the
+negative results of \textit{In vitro} replication assays investigating
+the fitness effects of small RNA hairpins in HIV-1
+\citep{knoepfel_role_2013}: it would take hundreds of cell culture
+passages to detect fitness effects of the order of one per mille. The longitudinal data,
+however, spans many years, hence our analysis is able to quantify the
+subtle fitness effect of RNA structure within single infections.
+
+The fixation probabilities and the sojourn times of SNVs represent a rich and simple
+summary statistics: they are useful to characterize longitudinal sequence data and
+can be compared to models via computer simulations. These statistics are informative
+even in absence of a neutral control and thus appropriate to analyze
+properties of synonymous sites. 
+We find selection against synonymous substitutions despite the
+fact that the corresponding sites are not strongly conserved in
+cross sectional data. This is consistent with a recent comparative
+analysis of SIV and HIV-1 RNA secondary structure using SHAPE assays and
+computational methods \citep{pollom_comparison_2013}. While large scale
+patterns of RNA structures tend to agree in both viruses, the individual
+base pairs forming the structures are almost always
+discordant. Even though the molecular architecture of these structures
+changes over time, selection seems to maintain them, which reduces the
+fixation probability and hence the rate of evolution at synonymous
+sites. As expected from this argument, the evolutionary rate
+at synonymous sites varies greatly along the HIV-1 genome
+\citep{mayrose_towards_2007} (see also \figurename~S\syndiv). This variation can confound estimates of
+selection on proteins substantially \citep{ngandu_extensive_2008}. 
+The dynamic nature of HIV secondary structure makes it hard to find the 
+compensatory mutations that would restore base pairing in the
+longitudinal data; in fact, the exact base pairing pattern is
+most likely different than in the reference sequence.
+
+Selection against the majority of synonymous substitutions is probably
+common across the genome, but we only observe deleterious synonymous SNVs
+at high frequency in \shankaregion{} of \env{}, where they hitchhike to
+high frequencies on nAb escape mutations. Surprisingly, nonsynonymous
+mutations display a fixation pattern as if they were neutral;
+see~\FIG{fixp2}. However, nonsynonymous diversity exceeds
+synonymous diversity despite the overall much greater constraints on the
+amino acid sequence, suggesting that the majority of high frequency SNVs
+are escape mutations despite the fact that are often lost again. We suggest that this
+paradoxical behavior could be due to (i) escape
+mutations that revert after they themselves are recognized by nAbs, or (ii)
+the competition between different escape mutations within one epitope. 
+Both mechanisms reduce the overall fixation probability and can
+give rise to the observed pattern of fixation in computer simulations.
+
+The observed hitchhiking highlights the importance of linkage due to
+infrequent recombination for the evolution of HIV-1
+\citep{neher_recombination_2010, batorsky_estimate_2011,
+josefsson_majority_2011}. The recombination rate has been estimated to be on the
+order of $\rho = 10^{-5}$ per base and day. It takes roughly $t_{sw} =
+\epsilon^{-1} \log \nu_0$ generations for an escape SNV with escape rate
+$\epsilon$ to rise from an initially low frequency $\nu_0\approx \mu$ to frequency
+one. This implies that a region of length $l = (\rho t_{sw})^{-1} = \epsilon /
+\rho \log \nu_0$ remains linked to the adaptive mutation. With $\epsilon=0.01$,
+we have $l\approx 100$ bases. Hence we expect strong linkage between the
+variable loops and the flanking sequences, but none far beyond the variable
+regions, consistent with the lack of signal outside of \shankaregion. In case of
+much stronger selection -- such as observed during early CTL escape or drug
+resistance evolution -- the linked region is of course much larger
+\citep{nijhuis_stochastic_1998}.
+
+While classical population genetics assumes that the dominant stochastic force
+is genetic drift, i.e., non-heritable fluctuations in offspring number, our
+results show that stochasticity due to linked selection is much more important.
+Such fluctuations have been termed \emph{genetic draft} in
+\citet{gillespie_genetic_2000} and their effect in facultatively sexual population
+such as HIV-1 has been characterized in \citep{neher_genetic_2011}. Importantly,
+large population sizes are compatible with low diversity and fast coalescence
+when draft dominates over drift.
+
+Our results emphasize the inadequacy of independent site models of HIV-1 evolution
+and the common assumption that selection is time independent or additive. 
+If genetic variation is only transiently beneficial, existing estimates of the
+strength of selection \citep{neher_recombination_2010,batorsky_estimate_2011}
+could be substantial underestimates. Furthermore, weak conservation and
+time-dependent selection result in estimates of evolutionary 
+rates that depend on the time interval of observation, with lower rates across
+larger intervals. This implies that deep nodes in phylogenies might be older than 
+they appear.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section*{Acknowledgments}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\bibliographystyle{unsrtnat}
+\bibliographystyle{unsrtnat}
 \bibliography{bib}
-\newpage
-\appendix
-\onecolumngrid
-\setcounter{figure}{0}
-\include{supplement_body}
+%\onecolumngrid
+
+\begin{figure}
+\begin{center}
+\subfloat{\includegraphics[width=0.6\linewidth]
+{Shankarappa_allele_freqs_trajectories_syn_p10.pdf}
+\label{fig:aftsyn}}\\
+ \subfloat{\includegraphics[width=0.6\linewidth]
+{Shankarappa_allele_freqs_trajectories_nonsyn_p10.pdf}
+\label{fig:aftnonsyn}}
+\caption{Time series of frequencies
+of synonymous (A) and nonsynonymous (B) single nucleotide variants (SNVs) in \env, 
+\shankaregion, from patient p10~\cite{shankarappa_consistent_1999}.
+While many nonsynonymous SNVs fix, few synonymous
+SNVs do so even though they are frequently observed at high
+frequencies. Colors indicate the position of the site along the \shankaregion{} region
+(blue to red). Inset: the fixation probability $\pfix$ of a neutral
+SNV that reached 50\% frequency is one half.}
+\label{fig:aft}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\subfloat{\includegraphics[width=0.6\linewidth]{fixation_times.pdf}
+\label{fig:fixp1}}\\
+\subfloat{\includegraphics[width=0.6\linewidth]{fixation_probabilities.pdf}
+\label{fig:fixp2}}
+\caption{Fixation and loss of SNVs.
+Panel A) shows how quickly synonymous SNVs are purged from the populations. 
+Specifically, the figure shows the fraction of SNVs that are still observed
+after $\Delta t$ days, conditional on being observed in one of the three frequency 
+intervals (different colors). 
+In each frequency interval, the fraction of synonymous
+SNVs that ultimately survive is the fixation probability $\pfix$ conditional on the
+initial frequency. The neutral expectation for $\pfix=\nu_0$ is indicated by 
+dashed horizontal lines.
+Panel B) shows the fixation probability of synonymous SNVs as a function of $\nu_0$. Polymorphisms within \shankaregion{} fix less
+often than expected for neutral SNVs (indicated by the diagonal line).
+This suppression is not observed in other parts of \env{} or for nonsynonymous
+SNVs.
+The horizontal error bars on the abscissa are bin sizes, the vertical ones the
+standard deviation after 100 patient bootstraps of the data. Data from
+refs.~\cite{shankarappa_consistent_1999,liu_selection_2006, bunnik_autologous_2008}.}
+\label{fig:fixp}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\subfloat{\includegraphics[height=0.3\linewidth]{reactivities_histograms_syn.pdf}\label{fig:SHAPEA}}
+\subfloat{\includegraphics[height=0.3\linewidth]{fixation_probabilities_VnonV.pdf}\label{fig:SHAPEB}}
+\caption{Permissible synonymous mutations tend to be unpaired.
+Panel A) shows the distribution of SHAPE reactivities among sites at which synonymous 
+SNVs fixed (red), sites at which SNVs reached frequencies above 15\% but
+were subsequently lost (blue), and sites at which no high-frequency SNVs were observed (green) 
+(all categories are restricted to the regions V1-V5$\pm 100$bp).
+Sites at which SNVs fixed tend to have higher SHAPE reactivities, corresponding to
+less base pairing, than those at which SNVs are lost.
+Sites at which no SNVs are observed show an intermediate distribution of SHAPE values.
+Panel B) shows the fixation probability of synonymous SNVs in
+\shankaregion{} separately for variable regions V3-V5 and the connecting conserved 
+regions C2-C4 that harbor RNA stems. As expected, the fixation probability is lower
+inside the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
+bunnik_autologous_2008, liu_selection_2006}.}
+\label{fig:SHAPE}
+\end{center}
+\end{figure}
+\begin{figure}
+\begin{center}
+\subfloat{\includegraphics[width=0.6\linewidth]{simulations_graduallydel.pdf}
+\label{fig:simfixpvar}}\\
+\subfloat{\includegraphics[width=0.6\linewidth]{simulations_syn.pdf}
+\label{fig:simsfig}}
+\caption{Distribution of selection coefficients on synonymous sites. Panel A)
+The depression in $\pfix$ depends on the deleterious effect size 
+of synonymous SNVs. This parameter also reduces synonymous
+diversity, measured by the probability of a SNV to be found at
+intermediate frequencies $P_\text{interm}$ (first inset).
+Panel B) To assess the parameter space that affects synonymous fixation and
+diversity, we run 2400 simulations with random parameters for deleterious effect
+size, fraction of deleterious synonymous sites, average escape rate $\epsilon$
+(color, blue to red corresponds to $10^{-2.5}$ to $10^{-1.5}$ per day), and rate of
+introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
+day). Only simulations that reproduce the synonymous diversity and fixation
+patterns observed in data are shown. These simulations demonstrate that
+deleterious effects are around $-0.002$ and a large fraction of the 
+synonymous mutations needs to be deleterious. As expected, larger
+$s_d$ require larger $\epsilon$. Parameters are chosen
+from prior distributions uniform in logspace as indicated by the red rectangle
+(see methods).}
+\label{fig:simheat}
+\end{center}
+\end{figure}
+
+%\setcounter{figure}{0}
+%\include{supplement_body}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2013-07-31 supplement_body.tex
first inclusions of revision

@@
 The other parameters for the shown simulations are
-the following: deleterious effect $s_d = 10^{-3}$, average escape rate $\epsilon = 0.016$,
-fraction of deleterious synonymous mutations $\alpha = 0.986$, rate of new epitopes
+the following: deleterious effect $s_d = 2 \cdot 10^{-3}$, average escape rate $\epsilon
+= 0.032$, fraction of deleterious synonymous mutations $\alpha = 0.986$, rate of new epitopes
 $k_A=0.0014$ per generation.
@@
 compete and only one of them fixes. The other parameters for the shown simulations are
-the following: deleterious effect $s_d = 10^{-3}$, average escape rate $\epsilon = 0.016$,
-fraction of deleterious synonymous mutations $\alpha = 0.986$, rate of new epitopes
+the following: deleterious effect $s_d = 2 \cdot 10^{-3}$, average escape rate
+$\epsilon = 0.032$, fraction of deleterious synonymous mutations $\alpha = 0.986$, rate of new epitopes
 $k_A=0.0014$ per generation.
@@
 \end{figure}
+
+\newpage
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Simulations with different parameters}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=0.6\linewidth]{simulations_syn_gentime_2days.pdf}
+\caption{
+Results of simulations with alternative parameters.
+Panel A) Like \figurename~4B in the main text, but simulations are performed using a
+generation time of two days. Two selection coefficient of synonymous sites
+doubles in units of 1/generations, hence keeps constants in units of 1/days.
+}
+\label{fig:sims_altparameters}
+\end{center}
+\end{figure}

### 2013-07-31 synmut.tex
first inclusions of revision

@@
 \newcommand{\withinepi}{4}
+\newcommand{\gentimetwo}{5}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Computational modeling indicates that (i) most synonymous mutations in this
-genomic region have a selection coefficient of the order of $-0.002$, and
+genomic region have a selection coefficient of the order of $-0.004$, and
 (ii) synonymous variants rise to high frequency by genetic hitchhiking on neighboring beneficial
@@
 in one of two states. Outcrossing rates, crossover rates, mutations rates and
-arbitrary fitness functions can be specified. We used a generation time of 1
-day, an outcrossing rate of $r=0.01$ per day \citep{batorsky_estimate_2011,
-neher_recombination_2010}, a mutation rate of $\mu=10^{-5}$
-\citep{mansky_lower_1995, abram_nature_2010} and simulated intrapatient
+arbitrary fitness functions can be specified. We used an outcrossing rate of
+$r=0.01$ per day \citep{batorsky_estimate_2011, neher_recombination_2010},
+a mutation rate of $\mu=2 \times 10^{-5}$ \citep{mansky_lower_1995, abram_nature_2010}.
+We performed the simulations with
+generation times of both 1 and 2 days, because this parameter varies across patients
+\citep{perelson_hiv-1_1996, perelson_decay_1997}; our conclusions are shown in
+did not change (see \FIG{simheat} and \figurename~S\gentimetwo, respectively). We simulated intrapatient
 evolution for 6000 days. For simplicity, third positions of every codon were
@@
 between 0.75 and 1.0; the average deleterious effect $s_d$ was sampled such that
-its logarithm was uniformly distributed  between $10^{-4}$ and $10^{-2}$; the
+its logarithm was uniformly distributed  between $2 \cdot 10^{-4}$ and $2 \cdot 10^{-2}$; the
 average escape rate $\epsilon$ of escape mutation was sampled such that its logarithm was
-uniform between $10^{-2.5}$ and $10^{-1.5}$ and the rate $k_A$ of new antibody
+uniform between $2 \cdot 10^{-2.5}$ and $2 \cdot 10^{-1.5}$ and the rate $k_A$ of new antibody
 challenges such that its logarithm was uniform between $10^{-3}$ and $10^{-2}$
@@
 \FIG{simsfig}.
+
+% TODO change mu and rho and N
 
@@
 effect of the mutation, this corresponds to deleterious effects $s_d$ of the
-order of $- 0.002$ per day. (This is only an average estimate: every single
+order of $-0.002$ per day. (This is only an average estimate: every single
 mutation is expected to have a slightly different fitness effect.)
@@
 ($\gtrsim 0.8$) of sites has to be deleterious with effect size $|s_d| \sim
-0.002$. This is consistent with the fixation/extinction times estimated above
-(see \FIG{fixp1}).
+0.004$. The inverse of $|s_d|$ is of the same order of magnitude as the
+fixation/extinction times estimated above (see \FIG{fixp1}).
 
@@
 C2-C4. We suggest through computational modeling that these
-SNVs have deleterious effects on the order of $0.002$ and that they are
+SNVs have deleterious effects on the order of $0.004$ and that they are
 brought to high frequency through linkage to adaptive mutations.
@@
 patterns observed in data are shown. These simulations demonstrate that
-deleterious effects are around $-0.002$ and a large fraction of the 
+deleterious effects are around $-0.004$ and a large fraction of the 
 synonymous mutations needs to be deleterious. As expected, larger

### 2013-08-02 supplement_body.tex
contour

@@
 \newpage
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\section{Simulations with different parameters}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\begin{figure}[h]
+%\begin{center}
+%\includegraphics[width=0.4\linewidth]{simulations_syn_gentime_2days.pdf}\\
+%\includegraphics[width=0.24\linewidth]{simulations_syn_mu_1e-5.pdf}
+%\includegraphics[width=0.24\linewidth]{simulations_syn_latin_suppl.pdf}
+%\includegraphics[width=0.24\linewidth]{simulations_syn_mu_3e-5.pdf}
+%\includegraphics[width=0.24\linewidth]{simulations_syn_mu_4e-5.pdf}\\
+%\includegraphics[width=0.32\linewidth]{simulations_syn_coi_5e-3.pdf}
+%\includegraphics[width=0.32\linewidth]{simulations_syn_latin_suppl2.pdf}
+%\includegraphics[width=0.32\linewidth]{simulations_syn_coi_2e-2.pdf}
+%\caption{
+%Results of simulations with alternative parameters.
+%\textbf{Panel A)} Like \figurename~4B in the main text, but simulations are performed using a
+%generation time of two days. Two selection coefficient of synonymous sites
+%doubles in units of 1/generations, hence keeps constants in units of 1/days.
+%\textbf{Panels B, C, D, and E)} Effect of mutation rate on the simulation
+%results. The mutation rates are respectively $1 \times 10^{-5}$ (B), $2 \times
+%10^{-5}$ (C), $3 \times 10^{-5}$ (D), and $4 \times
+%10^{-5}$ (E) per generation per site. Increasing the mutation rate tightens the
+%requirement that almost all synonymous mutations must be deleterious. This can
+%be rationalized as follows: a higher $\mu$ increases the neutral synonymous
+%diversity (we have measured this directly in our simulations), hence more and
+%more synonymous SNVs that reach high frequency are neutral and the
+%effect of deleterious SNVs, although present, becomes too weak to be observed as
+%a depression in $P_\text{fix}$. As a consequence, only simulations with almost
+%no neutral sites still show a reduced $P_\text{fix}$ and are shown in the
+%figures.
+%\textbf{Panels F, G, H)} Effect of recombination rate on the simulation results.
+%The recombination rates are respectively $0.5 \times 10^{-5}$ (F), $10^{-5}$
+%(G), and $2 \times 10^{-5}$ (H) per generation per site. Recombination does not
+%change the results qualitatively in terms of fixation probability. Simulations
+%with high recombination rates, however, have an excessively high synonymous
+%diversity compared to the data (dots at the bottom and right side of panel H).
+%More stringent gating around the data solves this issue.
+%}
+%\label{fig:sims_altparameters}
+%\end{center}
+%\end{figure}
+
+\newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Simulations with different parameters}
+\section{Marginal distributions in simulations}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{center}
-\includegraphics[width=0.6\linewidth]{simulations_syn_gentime_2days.pdf}
+\includegraphics[width=0.99\linewidth]{simulations_syn_marginals.pdf}
 \caption{
-Results of simulations with alternative parameters.
-Panel A) Like \figurename~4B in the main text, but simulations are performed using a
-generation time of two days. Two selection coefficient of synonymous sites
-doubles in units of 1/generations, hence keeps constants in units of 1/days.
+Marginal distributions of parameters in simulations that reproduce the
+synonymous fixation probability and diversity seen in the data. The prior on all
+parameters are flat lines (see methods). Whereas some parameters do not seem to
+affect the fixation of synonymous deleterious SNVs (e.g. the recombination
+rate), the effect of deleterious synonymous changes and their abundance are
+crucial, as discussed in the main text.
 }
@@
 \end{figure}
+
+\newpage
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Divergence and diversity over time in simulations}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=0.6\linewidth]{simulations_diversity_divergence_example.pdf}
+\caption{
+Diversity and divergence over time in an example simulation with realistic
+parameters. Divergence from the founder genotype accumulates linearly. On the
+contrary, because of the frequent selective sweeps of escape mutations, diversity saturates
+relatively quickly. The population dynamics after that point is in a steady
+state.
+}
+\label{fig:sims_altparameters}
+\end{center}
+\end{figure}

### 2013-08-02 response_to_reviewers.txt
added a response to reviewers file, changed discussion of Landscape paper

##############################################################################
# Reviewer comments:
##############################################################################
# Reviewer #1 (Comments for Author):
##############################################################################

>The text is clearly written and the written arguments are well spelled out.
>They observe that synonymous mutations in the C2-V5 region that disrupt
>modelled base pairs in SHAPE assays are more likely to be lost than other
>mutations, implying clear fitness costs to the virus, although they point out
>that those with major costs will never reach high frequencies. They argue that
>less deleterious mutations, but still with a fitness cost, can rise to high
>frequency by 'hitch hiking' with immune selected non-synonymous mutations, in
>other words simple linkage. The selection against synonymous slows down
>evolution of the virus so selective forces may have been underestimated in the
>past. These are important concepts and show that immune escape mutations
>should not be considered in isolation. They support the view that 'landscaping'
>approaches such as that described recently by Chakraborty in Immunity (Immunity
>38:606-617, 2013), need to be applied.

We thank the reviewer for this thoughtful assessment. We have included a
brief discussion of the landscape aspect of HIV fitness and now cite
Ferguson et al, Immunity, 2013.

##############################################################################
# Reviewer #2 (Comments for Author):
##############################################################################

>These are interesting results and the manuscript is clearly written. The second
>observation is not novel, but underappreciated, whereas this is the first
>concrete calculation of the average selective disadvantage of silent mutations
>in HIV taking draft into account. As such, it is necessary for this work to
>establish proper error estimates so that the community can make valid use of
>the results.  With this in mind, the manuscript can be strengthened in the
>following respects:

We thank the reviewer for the constructive comments, which have helped
us to improve the manuscript. We have resorted to a more sophisticated
sampling scheme for parameters in the computer simulations and now scan
a larger parameter space. All our conclusions remain valid, but are more
reliable now. We detail these improvements in our point to point
response below.

##############################################################################
>1) The modeling generation time of 1 day is odd.  It takes about a day for the
>virion to integrate and start transcription.  One usually takes 2 days as the
>mean generation time, but there is individual variation. See next point.

We agree that one generation in HIV takes closer to 2 days than 1
day. Most escape rates and selection coefficients, however, are more
readily measured in days and we have chosen to model HIV evolution with
generation of one day for simplicity. Factor of 2 changes in the
generation have next to now influence on the dynamics and only
renormalize genetic drift. Genetic drift is not a dominant force in HIV
evolution and the population size, the main determinant of drift, is not
known to this accuracy anyway. We have repeated a subset of the
simulations with a generation time of one day, finding now
difference. IS THIS SHOWN SOMEWHERE?

##############################################################################
>2) Related to the above: the simulation does not address how sensitive the
>results are to varying the parameters (in particular, the generation time, the
>outcrossing rate, the mutation rate, and the assigned deleterious fitness of
>the "non-synonymous" changes).  In the absence of described theoretical
>analysis shaping our expectations, I strongly feel that this sensitivity
>analysis needs to be part of the results, and folded into the error bars.

We have greatly extended the parameter space that we survey and now test
different values for the mutation rate, the recombination rate, the
generation time, the fitness effects, and the population size. To
accomodate this high dimensional data, we have changed figure 4. The
figure now shows the joint distribution of the deleterious effect and
the fraction of synonymous mutations that are deleterious that are
compatible with observations as a heat map. The marginal distribution of
the individual parameters consistent with the data are shown in the supplement.

The distribution of allowed deleterious is sharply peaked around
???. DISCUSS THE DIVERSITY ISSUES.

##############################################################################
>3) The reasonableness of the ranges for the various parameters should be argued
>with citation to existing literature.

We have added a discussion of the parameter ranges to the methods
section.

##############################################################################
>4) I am a bit surprised that the authors chose to do random sampling in their
>multidimensional parameter space.  One would have expected a Latin hypercube
>sampling to provide a better coverage.

Indeed. Our previous simulations where not limited by coverage of the
parameter space. Now that we cover a much larger (higher dimensional)
space, we did implement a Latin hypercube sampling scheme. We thank the
reviewer for this very useful suggestion. WE MIGHT WANT TO SHOW THE RANK
CORRELATION MATRIX BETWEEN THE DIFFERENT VARIABLES.

##############################################################################
>5) The actual dynamics of HIV in the body is presumably more complex: it is
>likely that the virions are produced in large bursts of virtually genetically
>identical virions most of which are then cleared by the body fast on the
>time-scale of HIV generations. There is no discussion whether this leads to
>increased stochasticity in the dynamics, nor whether this can influence the
>quantitative evaluations presented here. It is possible that the authors mean
>to absorb these into some kind of reduced effective population size, but a
>short discussion would be beneficial.

We agree with reviewer that HIV evolution is bursty, but we would like
to mention that actual number of cells that get infected every day is
around 10^7 and hence much larger than we are able to simulate. The
population size in our simulation can therefore be viewed as an
effective population size. We have investigated the sensitivity of our
results to changes in the population size and find that our conclusions
are insensitive to the population size as long as the product of
population size and mutation rate is not very much smaller than 1.

WE SHOULD ADD A DISCUSSION AND POINT OUT HERE.

##############################################################################
>6) In the discussions, the rise of synonymous mutations is attributed to draft,
>but the fall to selective effects. In general, some of the fall might also be
>due to overlapping selective sweeps: the manuscript does not discuss this
>possibility.

Yes, much of the stochastic component in the dynamics of the frequencies
of synonymous mutations will be due to draft. The directed component can
be attributed to selection. WE ADDED A DISCUSSION OF THIS ISSUE.

##############################################################################
>7) The observed neutral fixation pattern of the nonsynonymous mutations is
>mentioned, and two explanations have been proposed to duet of increased
>diversity over synonymous mutations with no increase of fixation over neutral
>expectations.  Neither of the explanations, however, are plausible (at least,
>without argument) absent experimental evidence not mentioned in the text. It
>would perhaps be better to either strengthen this discussion with citation to
>literature, or to weaken it by mentioning these as mere possibilities amongst
>others.

We agree with the reviewer that the both hypothesis we put forward to
explain the fixation pattern of non-synonymous mutations are somewhat
tentative. Qualitatively, however, the processes we discuss have been
observed and we do cite experimental evidence. NEED TO ELABORATE.

### 2013-08-02 synmut.tex
added a response to reviewers file, changed discussion of Landscape paper

@@
 \newcommand{\withinepi}{4}
-\newcommand{\gentimetwo}{5}
+\newcommand{\marginals}{5}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \author{\Author\\ \Affiliation}
-%\affiliation{\Affiliation}
 \date{\today}
@@
 \newpage
-%\begin{abstract}
 \section*{Abstract}
@@
 Computational modeling indicates that (i) most synonymous mutations in this
-genomic region have a selection coefficient of the order of $-0.004$, and
+genomic region have a selection coefficient of the order of $-0.003$, and
 (ii) synonymous variants rise to high frequency by genetic hitchhiking on neighboring beneficial
@@
 relevant, while the precise base pairing pattern is not.
-%\end{abstract}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \citep{perelson_hiv-1_1996, perelson_decay_1997}; our conclusions are shown in
-did not change (see \FIG{simheat} and \figurename~S\gentimetwo, respectively). We simulated intrapatient
+did not change. We simulated intrapatient
 evolution for 6000 days. For simplicity, third positions of every codon were
@@
 
-The simulations were repeated 2400 times with random choices for the following
-parameters: the fraction of deleterious sites $\alpha$ was sampled uniformly
+The simulations were repeated 1000 times with a four dimensional Latin hypercube sampling scheme.
+For two parameters, ranges were chosen around estimates from the recent literature:
+the average escape rate $\epsilon$ of escape mutation was sampled such that its logarithm was
+uniform between $2 \cdot 10^{-2.5}$ and $2 \cdot 10^{-1.5}$
+\citep{neher_recombination_2010} and the rate $k_A$ of new antibody
+challenges such that its logarithm was uniform between $10^{-3}$ and $10^{-2}$
+per generation \citep{richman_rapid_2003, moore_limited_2009}.
+For the other two parameters, ranges were determined to include all simulations
+that reproduced the experimental data: the fraction of deleterious sites $\alpha$ was sampled uniformly
 between 0.75 and 1.0; the average deleterious effect $s_d$ was sampled such that
-its logarithm was uniformly distributed  between $2 \cdot 10^{-4}$ and $2 \cdot 10^{-2}$; the
-average escape rate $\epsilon$ of escape mutation was sampled such that its logarithm was
-uniform between $2 \cdot 10^{-2.5}$ and $2 \cdot 10^{-1.5}$ and the rate $k_A$ of new antibody
-challenges such that its logarithm was uniform between $10^{-3}$ and $10^{-2}$
-per generation. Populations were initialized with a homogeneous founder
+its logarithm was uniformly distributed  between $2 \cdot 10^{-4}$ and $2 \cdot
+10^{-2}$. Populations were initialized with a homogeneous founder
 population and were kept at an average size of $N=10^4$ throughout the
@@
 between the bin centers. This measure is sufficiently precise for our purposes.
-In 10 runs out of 2400, the highest frequency bin was empty so the fixation
+In 4 runs out of 1000, the highest frequency bin was empty so the fixation
 probability could not be calculated; those runs were excluded from
@@
 ($\gtrsim 0.8$) of sites has to be deleterious with effect size $|s_d| \sim
-0.004$. The inverse of $|s_d|$ is of the same order of magnitude as the
+0.003$. The inverse of $|s_d|$ is of the same order of magnitude as the
 fixation/extinction times estimated above (see \FIG{fixp1}).
@@
 C2-C4. We suggest through computational modeling that these
-SNVs have deleterious effects on the order of $0.004$ and that they are
+SNVs have deleterious effects on the order of $0.003$ and that they are
 brought to high frequency through linkage to adaptive mutations.
@@
 longitudinal data; in fact, the exact base pairing pattern is
-most likely different than in the reference sequence.
+most likely different than in the reference sequence. Nevertheless, the
+importance of secondary structure underscores the realization that viral
+fitness is a complicated function (fitness landscape) in the
+high-dimensional space of possible genotypes
+\citep{ferguson_translating_2013}. We show, that this landscape is shaped not
+only by interactions between amino-acids, but also by ubiquitous
+interactions between nucleotides in RNA structures.
 
@@
 \end{figure}
+
 \begin{figure}
@@
 \label{fig:simfixpvar}}\\
-\subfloat{\includegraphics[width=0.6\linewidth]{simulations_syn.pdf}
+\subfloat{\includegraphics[width=0.6\linewidth]{simulations_syn_latin_contour.pdf}
 \label{fig:simsfig}}
@@
 Panel B) To assess the parameter space that affects synonymous fixation and
-diversity, we run 2400 simulations with random parameters for deleterious effect
-size, fraction of deleterious synonymous sites, average escape rate $\epsilon$
-(color, blue to red corresponds to $10^{-2.5}$ to $10^{-1.5}$ per day), and rate of
-introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
-day). Only simulations that reproduce the synonymous diversity and fixation
-patterns observed in data are shown. These simulations demonstrate that
-deleterious effects are around $-0.004$ and a large fraction of the 
-synonymous mutations needs to be deleterious. As expected, larger
-$s_d$ require larger $\epsilon$. Parameters are chosen
-from prior distributions uniform in logspace as indicated by the red rectangle
-(see methods).}
+diversity, we run 3000 simulations with random parameters for deleterious effect
+size, fraction of deleterious synonymous sites, average escape rate $\epsilon$,
+rate of introduction of new epitopes, population size, mutation rate, and
+recombination rate (see methods). The density of simulations that reproduce
+the synonymous diversity and fixation patterns observed in data are shown.
+These simulations demonstrate that deleterious effects are around $-0.003$
+and a large fraction of the synonymous mutations needs to be deleterious. This
+picture is also confirmed by the analysis of the marginal distributions of the
+single parameters (see \figurename~S\marginals).}
 \label{fig:simheat}

### 2013-08-05 synmut.tex
added a few references and discussions of issues ref 2 raised. worked on response

@@
 in one of two states. Outcrossing rates, crossover rates, mutations rates and
-arbitrary fitness functions can be specified. We used an outcrossing rate of
-$r=0.01$ per day \citep{batorsky_estimate_2011, neher_recombination_2010},
-a mutation rate of $\mu=2 \times 10^{-5}$ \citep{mansky_lower_1995, abram_nature_2010}.
-We performed the simulations with
-generation times of both 1 and 2 days, because this parameter varies across patients
-\citep{perelson_hiv-1_1996, perelson_decay_1997}; our conclusions are shown in
-did not change. We simulated intrapatient
-evolution for 6000 days. For simplicity, third positions of every codon were
+arbitrary fitness functions can be specified. While the best estimates
+for the HIV generation time is roughly two days
+\citep{perelson_hiv-1_1996, markowitz_novel_2003}, the generation time
+has very little influence on the result and basically sets the unit of
+time in which other parameters are measured. For simplicity, we use one
+generation per day and have checked that the results obtained are
+indistinguishable from simulations with a generation time of two days.
+
+\subsubsection*{Parameters}
+In order to simulate HIV, we have to specify several parameters that are
+known to various degrees. The mutation rate has been measured using
+exogenous LacZ constructs and the average nucleotide substitution rate is 
+estimated to be $\mu\approx 2 \times 10^{-5}$ per generation \citep{mansky_lower_1995,
+abram_nature_2010}. In our simulations, we vary the mutations rate
+from $?? -?? \times 10^{-5}$ per day. Different virions within an
+infected person can recombine their genome by coinfection of the same
+cell, followed by template switching during reverse transcription. 
+Template switching happens between around ten times per reverse
+transcription \citep{levy_dynamics_2004}. The combined recombination rate of template
+switching and coinfection has been estimated based on modeling to be around $\rho\approx
+10^{-5}$ per nucleotide and day
+\citep{neher_recombination_2010,batorsky_estimate_2011}, implying a
+coinfection rate of a few percent or less, as has been recently
+confirmed experimentally \citep{josefsson_single_2013}. We assume a
+template switching rate of $10^{3}$ per nucleotide and vary the
+recombination rate between $??-??\times 10^{-5}$ by adjusting the
+coinfection rate. The population size relevant to evolution in chronic
+infection has been hotly debated. The number of cells infected by HIV-1
+per day is on the order of $10^{7}$ \citep{perelson_hiv-1_1996}, but the
+number of viruses contributing to the next generation might be
+considerably smaller due to the burstiness of replication. 
+Recent evidence points to a relevant population
+size in excess of the inverse mutation rate
+\citep{boltz_ultrasensitive_2012}. In fact, the evolutionary dynamics
+depends only weakly on the actual value of the population size once the
+product $N\mu$ is of order one or larger \citep{neher_genetic_2011}. We
+simulate populations between $N=10^4$ and $10^5$. Larger populations
+become prohibitively costly to simulate.
+
+Another crucial ingredient for our simulations is the fitness landscape,
+i.e., the effects of mutations on fitness and possible interactions
+between mutations. For simplicity, third positions of every codon were
 deemed synonymous and assigned either a selection coefficient $0$ with
@@
 Mutations at the first and second positions were assigned strongly deleterious 
-fitness effects 0.02. At 
+fitness effects ??? COULD MAKE THIS STRONGER, NEED CITATION. At 
 rate $k_A$, a random locus in the genome is designated an epitope that can
 escape by one or several mutations with exponentially distributed escape rates
-with mean $\epsilon$. Both full-length HIV-1 genomes and \env{}-only simulations
-were performed and yielded comparable results.
-
-The simulations were repeated 1000 times with a four dimensional Latin hypercube sampling scheme.
-For two parameters, ranges were chosen around estimates from the recent literature:
-the average escape rate $\epsilon$ of escape mutation was sampled such that its logarithm was
-uniform between $2 \cdot 10^{-2.5}$ and $2 \cdot 10^{-1.5}$
-\citep{neher_recombination_2010} and the rate $k_A$ of new antibody
-challenges such that its logarithm was uniform between $10^{-3}$ and $10^{-2}$
-per generation \citep{richman_rapid_2003, moore_limited_2009}.
-For the other two parameters, ranges were determined to include all simulations
-that reproduced the experimental data: the fraction of deleterious sites $\alpha$ was sampled uniformly
-between 0.75 and 1.0; the average deleterious effect $s_d$ was sampled such that
-its logarithm was uniformly distributed  between $2 \cdot 10^{-4}$ and $2 \cdot
-10^{-2}$. Populations were initialized with a homogeneous founder
-population and were kept at an average size of $N=10^4$ throughout the
-simulation. After 30 generations of burn-in to create genetic diversity, new
-epitopes were introduced at a constant rate $k_A$.
+with mean $\epsilon$. The rate of escape in chronic infection is
+reported to be on the order of a few percent
+\citep{Asquith:2006p28003,moore_limited_2009}, consistent with the
+finding that virus is neutralized by serum from 3-6 month later
+\citep{richman_rapid_2003}. We simulated escape rates between $2 \cdot
+10^{-2.5}$ and $2 \cdot 10^{-1.5}$ per day. 
+The rate at which new anti-body challenges arise is more difficult to
+quantify, but the majority of non-synonymous substitutions in gp120 are
+probably driven by escape. Since the sequence of the C2-V5 region of
+gp120  changes at a rate of about 1\% per year
+\citep{shankarappa_consistent_1999}, this rate should be on the order of
+$k_A=0.003$ per day.
 
@@
 
-For each set of parameters, fixation probabilities and probabilities of
-synonymous polymorphisms $P_\text{interm}$ were calculated as averages over
-100 repetitions (with different random seeds).
+\subsubsection*{Sampling and analysis}
+To obtain reasonable sampling of a particular parameter combination we
+ran simulation for 6000 days and we repeated each run 100 times with
+different seeds for the random number generator. Both full-length HIV-1
+genomes and \env{}-only simulations were performed and yielded
+comparable results.  Populations were initialized with a homogeneous
+founder population and were kept at an average size of $N=10^4$
+throughout the simulation. After 30 generations of burn-in to create
+genetic diversity, new epitopes were introduced at a constant rate
+$k_A$.  The simulations were repeated 1000 times with a four dimensional
+Latin hypercube sampling scheme \citep{mckay_comparison_1979}.  For each
+parameter, 1000 random values were chosen from a distribution uniform in
+log-space and bounded by the ranges given for each of these parameters
+above. These 1000 values for each parameter were combined using a Latin
+hypercube sampling scheme.
 
@@
 
-If the immune system
-recognizes the escape mutant before its fixation, the mutant might cease to be
-beneficial and disappear soon, despite its quick initial rise in frequency. In
-support of this idea, in \citet{richman_rapid_2003, bunnik_autologous_2008},
-antibody responses against escape mutants have been reported. These responses are
-delayed by a few months, roughly matching the average time needed by an escape
-mutant to rise from low to high frequency. To model this type of behavior, we
-assumed that antibody responses against escape SNVs arise with a rate
-proportional to the frequency of the escape variant and abolish the benefit of the
-escape mutations. As expected, this type of time-dependent selection retained the
-potential for hitchhiking, but reduced fixation of nonsynonymous SNVs.
-\figurename~S\timedependence~shows that $\pfix$ of synonymous SNVs is not
-affected by this change, while $\pfix$ of nonsynonymous SNVs approaches the
-diagonal as the rate of recognition of escape mutants is increased.
+If the immune system recognizes the escape mutant before its fixation,
+the mutant might cease to be beneficial and disappear soon, despite its
+quick initial rise in frequency. In support of this idea, in
+\citet{richman_rapid_2003,wei_antibody_2003,bunnik_autologous_2008},
+antibody responses against escape mutants have been reported. These
+responses are delayed by a few months, roughly matching the average time
+needed by an escape mutant to rise from low to high
+frequency. Furthermore, neutralization responses against early virus
+sometimes decline later in infection, which could lead to a reversion of
+some sites. To model
+this type of behavior, we assumed that antibody responses against escape
+SNVs arise with a rate proportional to the frequency of the escape
+variant and abolish the benefit of the escape mutations. As expected,
+this type of time-dependent selection retained the potential for
+hitchhiking, but reduced fixation of nonsynonymous SNVs.
+\figurename~S\timedependence~shows that $\pfix$ of synonymous SNVs is
+not affected by this change, while $\pfix$ of nonsynonymous SNVs
+approaches the diagonal as the rate of recognition of escape mutants is
+increased.
 
-Several different escape mutations that arise within the 
-same epitope have similar effects. Their benefits
-are not additive, because each of them is essentially sufficient to escape and
-no additional benefit is gained from combining them. As a consequence, several
-escape SNVs rise to high frequency rapidly, while the one with the smallest cost
-in terms of replication, packaging, etc.~is most likely to eventually fix, while
-all others are lost. The emergence of multiple competing escape variants in HIV-1
-infections has been shown \citep{moore_limited_2009, bar_early_2012}. 
-Similarly, this scenario has been explicitly observed in the evolution of
-resistance to 3TC, where the mutation M184V is often preceded by M184I
-\citep{hedskog_dynamics_2010}. We implemented within epitope competition 
-in the model by allowing for multiple escape mutations per epitope that do
-not provide additional benefit to the virus when combined. Again, we found that the potential for
-hitchhiking is retained by within-epitope competition but that the
-fixation probability of nonsynonymous SNVs is reduced. With roughly six
-mutations per epitope, the simulation data are compatible with observations; see
+Several different escape mutations that arise within the same epitope
+have similar effects. Their benefits are not additive, because each of
+them is essentially sufficient to escape and no additional benefit is
+gained from combining them. As a consequence, several escape SNVs rise
+to high frequency rapidly, while the one with the smallest cost in terms
+of replication, packaging, etc.~is most likely to eventually fix, while
+all others are lost. The emergence of multiple competing escape variants
+in HIV-1 infections has been shown \citep{moore_limited_2009,
+  bar_early_2012,wei_antibody_2003}.  This scenario has been explicitly
+observed in the evolution of resistance to 3TC, where the mutation M184V
+is often preceded by M184I \citep{hedskog_dynamics_2010} or in the
+context of CTL escape mutations where multiple variants rise to high
+frequencies rapidly but most of the are lost again
+\citep{fischer_transmission_2010}. We implemented within epitope
+competition in the model by allowing for multiple escape mutations per
+epitope that do not provide additional benefit to the virus when
+combined. Again, we found that the potential for hitchhiking is retained
+by within-epitope competition but that the fixation probability of
+nonsynonymous SNVs is reduced. With roughly six mutations per epitope,
+the simulation data are compatible with observations; see
 \figurename~S\withinepi. The two scenarios, time-dependent selection and
-competition between equivalent escape pathways, are not exclusive and possibly
-both important in HIV-1 evolution.
+competition between equivalent escape pathways, are not exclusive and
+possibly both important in HIV-1 evolution.
 
@@
 brought to high frequency through linkage to adaptive mutations.
+While both the rise and fall of synonymous SNV is subject to hichhiking,
+the consistent loss of these variants conditional on being at high
+frequency implies selection against these SNVs.
 Comparison with biochemical data (SHAPE) of base pairing propensity in the RNA
@@
 subtle fitness effect of RNA structure within single infections. 
+The fixation probabilities and the sojourn times of SNVs represent a
+rich and simple summary statistics of longitudinal sequence data. Most
+importantly, these statistics are informative even in absence of a
+neutral control and thus appropriate to analyze properties of synonymous
+sites.
 
-The fixation probabilities and the sojourn times of SNVs represent a rich and simple
-summary statistics: they are useful to characterize longitudinal sequence data and
-can be compared to models via computer simulations. These statistics are informative
-even in absence of a neutral control and thus appropriate to analyze
-properties of synonymous sites. 
 We find selection against synonymous substitutions despite the

### 2013-08-08 response_to_reviewers.txt

@@
 known to this accuracy anyway. We have repeated a subset of the
-simulations with a generation time of one day, finding now
+simulations with a generation time of one day, finding no
 difference. IS THIS SHOWN SOMEWHERE?
@@
 
-WE SHOULD ADD A DISCUSSION AND POINT OUT HERE.
+We have added a discussion of our choice of population size to the
+methods section next the part where the other parameter choices are
+discussed.
 
@@
 Yes, much of the stochastic component in the dynamics of the frequencies
-of synonymous mutations will be due to draft. The directed component can
-be attributed to selection. WE ADDED A DISCUSSION OF THIS ISSUE.
+of synonymous mutations will be due to draft. The consistent loss of
+variants, conditional on them being at high frequency earlier, implies
+selection against these variants. We have added a sentence to the
+discussion to make this point clear.
 
@@
 tentative. Qualitatively, however, the processes we discuss have been
-observed and we do cite experimental evidence. NEED TO ELABORATE.
+observed and we do cite experimental evidence. It is common for escape
+to happen in multiple equivalent ways and this has been well documented
+fro early CTL escape and early nAb escape (Wei et al, Nature, 2003. Bar
+et al, PPath, 2012). Data for chronic infection is less conclusive and
+selection pressures are generally smaller, but turnover at targeted
+epitopes and the replacement of variants that where previously at high
+frequencies was observed 3y into the infection in Wei et al, Nature, 2003.
+Whether this fall in frequency of some non-synonymous variants happens
+because they are replaced by a less costly escape variant or because the
+escape variant itself is targeted is hard to disentangle. We think it is
+reasonable that both of these effects contribute. Whether they are
+sufficient to produce the observed fixation patterns of non-synonymous
+mutations is largely an empirical question. We simply explored using
+numerical experiments whether these effects push the curve in the
+"right" direction.

### 2013-08-08 response_to_reviewers.txt
few comments on response and manuscript

@@
 generation of one day for simplicity. Factor of 2 changes in the
-generation have next to now influence on the dynamics and only
+generation have little influence on the dynamics and only
 renormalize genetic drift. Genetic drift is not a dominant force in HIV
@@
 simulations with a generation time of one day, finding no
-difference. IS THIS SHOWN SOMEWHERE?
+difference.
+
+IS THIS SHOWN SOMEWHERE? (Fabio is redoing the simulations)
 
@@
 The distribution of allowed deleterious is sharply peaked around
-???. DISCUSS THE DIVERSITY ISSUES.
+0.02. DISCUSS THE DIVERSITY ISSUES.
 
@@
 
-Indeed. Our previous simulations where not limited by coverage of the
+Indeed. Our previous simulations were not limited by coverage of the
 parameter space. Now that we cover a much larger (higher dimensional)
@@
 
+Fabio: I DO NOT THINK THAT ADRESSES THE QUESTION ("IN GENERAL" --> "BUT").
+MAYBE WE COULD STATE SOMETHING LIKE: "IN ABSENCE OF DELETERIOUS EFFECTS, THE
+FIXATION PROBABILITY OF HITCHHIKERS IS NOT AFFECTED BY DRAFT".
+
 ##############################################################################

### 2013-08-08 synmut.tex
few comments on response and manuscript

@@
 \citep{ferguson_translating_2013}. We show, that this landscape is shaped not
-only by interactions between amino-acids, but also by ubiquitous
+only by interactions between amino acids, but also by ubiquitous
 interactions between nucleotides in RNA structures.

### 2013-08-15 scripts/divergence_diversity_syn_nonsyn.py
new figs

@@ -1,0 +1,126 @@
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       01/03/2012
+content:    Trajectories of allele frequencies.
+'''
+# Standard modules
+from operator import *
+import numpy as np
+import matplotlib.cm as cm
+from Bio.Seq import translate
+import matplotlib.pyplot as plt
+
+
+# Custom modules
+import sys
+sys.path.insert(0, '.')
+import modules.parser_Shankarappa as pS
+from modules.helper import is_nonsyn_table
+from modules.alphabet import alpha
+from conservation_syn_nonsyn_subtypeB import codon_single_mutants_synnonsyn
+
+
+def get_is_mutation(consensus):
+    is_mutation = np.ones((len(alpha), len(consensus)), bool)
+    alphal = list(alpha)
+    for i, a in enumerate(consensus):
+        is_mutation[alphal.index(a), i] = False
+    return is_mutation.T
+
+
+
+# Script
+if __name__ == '__main__':
+
+    # Define the patients
+    patients = pS.parse_sequences(exclude=['p4', 'p7', 'p8', 'p11'])
+
+    # Iterate over patients
+    for k,p in enumerate(patients[:1]):
+        p.filter_only_sequenced()
+
+        # Measure the allele frequencies (of all alleles at all positions)
+        paf = p.allele_frequencies
+        afs = np.asarray([paf(seqs=p.seqs_from_visit(v)) for v in p.visit])
+
+        # Reshape so that we get the site as first axis, the nucleotide as
+        # second, the time as third
+        afs = afs.swapaxes(0,2)
+
+        # Eliminate gaps (whole codons are excluded to keep translation possible)
+        # first and translate then!
+        # Note: this has been implemented at the root
+        is_nongap = (afs[:,4,:] == 0).all(axis=1)
+
+        # Divide into syn/nonsyn to spot hitchhiking
+        consensus = alpha[afs[:,:,0].argmax(axis=1)]
+        conscods = map(''.join, consensus.reshape((len(consensus) / 3, 3)))
+        is_mut = get_is_mutation(consensus)
+        is_nonsyn = is_mut & is_nonsyn_table(consensus)
+        is_syn = is_mut & (-is_nonsyn)
+
+        # Times (in months?)
+        ts = np.array([p.visit_to_time(v) for v in p.visit]) * 30.5
+
+        # Divergence
[…]

### 2013-08-15 scripts/fixation_probability.py

@@ -259,6 +259,16 @@ if __name__ == '__main__':
         for i in xrange(counts[cla].shape[0]):
             Pfstd[cla][i] = np.std(Pfboot[cla][i], axis=1)
 
+    # Calculate areas
+    # Note: skip the last point (via i_final < len(x) - 1) because many simulations
+    # have nothing there (not enough time or very noisy)
+    x = [0] + list(nu0mid) + [1]
+    ysyn = [0] + list(Pf['syn_Shanka'][:, 0]) + [1]
+    i_final_offset = 2              # Skip last bin
+    i_final = len(x) - 1 - i_final_offset
+    Atot = 0.5 * x[i_final]**2
+    Asyn = - Atot + sum([0.5 * (ysyn[i+1] + ysyn[i]) * (x[i+1] - x[i]) for i in xrange(i_final)])
+
     # Prepare figure and plot results
     nu0mid = [0.5 * sum(nu0s) for nu0s in nu0ss]
     nu0err = [0.5 * (nu0s[1] - nu0s[0]) for nu0s in nu0ss]

### 2013-08-15 scripts/kimura_demonstration.py

@@ -9,7 +9,7 @@ def fix_prob(s, af):
 
 def area(sd, alpha, af_vec=np.linspace(0.05,0.95,19)):
     sfs_neutral = (1-alpha)/af_vec
-    sfs_dele  = alpha*np.exp(sd)/af_vec
+    sfs_dele  = alpha*np.exp(sd*af_vec)/af_vec
 
     frac_fixed = (af_vec*sfs_neutral+sfs_dele*fix_prob(sd,af_vec))\
         /(sfs_neutral+sfs_dele)

### 2013-08-15 scripts/synonymous_diversity.py

@@ -7,7 +7,7 @@ content:    Trajectories of allele frequencies.
 # Standard modules
 from operator import *
 import numpy as np
-import matplotlib.pyplot as ppl
+import matplotlib.pyplot as plt
 import matplotlib.cm as cm
 from Bio.Seq import translate
 
@@ -18,6 +18,7 @@ sys.path.insert(0, '.')
 import modules.parser_Shankarappa as pS
 from modules.helper import is_nonsyn_table
 from modules.alphabet import alpha
+from conservation_syn_nonsyn_subtypeB import codon_single_mutants_synnonsyn
 
 
 
@@ -46,28 +47,46 @@ if __name__ == '__main__':
 
         # Divide into syn/nonsyn to spot hitchhiking
         consensus = alpha[afs[:,:,0].argmax(axis=1)]
-        is_nonsyn = is_nonsyn_table(consensus)
+        conscods = map(''.join, consensus.reshape((len(consensus) / 3, 3)))
+        is_nongap_cod = np.array(map(lambda x: '-' not in x, conscods))
+
+        # Times (in months?)
+        ts = np.array([p.visit_to_time(v) for v in p.visit]) * 30.5
 
         # Scan the genome codon-wise
-        counts = np.zeros(afs.shape[0], int)
-        chances = np.zeros(afs.shape[0], int)
-        for i in xrange(len(afs)):
-            non_double_hit = ((afs[i - i % 3: i - i % 3 + 3] > 0).sum(axis=0).sum(axis=0) <= 4).all()
-            if (not is_nongap[i]):# or (not non_double_hit):
-                continue
-            af = afs[i]
-            for j, a in enumerate(alpha):
-                # Gate for nonsynonymous changes
-                if ((not is_nonsyn[i, j]) and       # Not nonsyn
-                    (consensus[i] != a)):       # An actual mutation
+        counts = np.zeros((len(ts), afs.shape[0] // 3), int)
+        chances = np.zeros(afs.shape[0] // 3, int)
+        # Proceed codon by codon
+        for i in xrange(len(ts)):
+            v = p.visit[i]
+            seqs = np.array(p.seqs_from_visit(v))
 
-                    chances[i] += 1
-                    counts[i] += ((af[j] >= 0.25) & (af[j] <= 0.75)).sum()
+            for j in xrange(afs.shape[0] // 3):
+                # Exclude gaps
+                if not is_nongap_cod[j]:
+                    continue
[…]

### 2013-08-15 scripts/script_fixation_loss_simulations_final.py
simulation script

@@ -1,0 +1,573 @@
+#!/ebio/ag-neher/share/programs/EPD/bin/python
+# vim: fdm=indent
+'''
+author:     Fabio Zanini
+date:       28/07/13
+content:    Perform the simulations for the depression in fixation probability
+            and analyze them at once, to avoid occupying lots of disk space.
+            This script uses a short genome model (1000 sites).
+
+'''
+# Standard modules
+import os
+import sys
+from operator import *
+import numpy as np
+from itertools import combinations
+import time
+
+# Env vars
+JOBSSCRIPT = os.path.realpath(__file__)      # This file
+JOBDIR = os.path.dirname(JOBSSCRIPT)+'/'
+JOBMODULES = JOBDIR+'modules'    
+# Set the modules path for the cluster nodes (as opposed to my laptop)
+on_laptop = any(map(lambda x: '/home/fabio' in x, os.sys.path))
+
+#if (not on_laptop) and (JOBMODULES not in sys.path):
+#    sys.path.insert(0, JOBMODULES)
+
+# Custom modules (import local FFPopSim)
+import FFPopSim as h
+
+
+
+# Globals
+VERBOSE = 2
+L = 1000
+crossover = 1e-3
+epitope_length = 9
+
+# Simulation length
+dt = 10
+n_time_points = 600
+dt_sample = 20
+
+# Analysis parameters
+t_max = 3000
+IS_ZERO = 0.000001
+max_counts = 10000
+nirands = 10
+sfs_bins = np.logspace(-3, -0.1, 30)
+savefig = True
+sep = '\t'
+
+
+
+# Classes
+class haploid_highd_gradual(h.haploid_highd):
+    '''Population with dynamic fitness landscape'''
+    
+
+    def __init__(self, *args, **kwargs):
+        if 'nescepi' not in kwargs:
+            raise ValueError('Please add the "nescepi" argument.')
+        newkwargs = {key: value for key, value in kwargs.iteritems() if key != 'nescepi'}
+        super(haploid_highd_gradual, self).__init__(*args, **newkwargs)
+        self.nescepi = kwargs['nescepi']
+        self.epitopes = []
+        self.epitope_times = []
+        self.epitope_fitness = []
+
+
+    def add_epitope(self, value):
+
+        # Complex epitope creation
+        if self.nescepi > 1:
+            # Choose a random good position
+            pos = np.random.randint(self.L - epitope_length)
+            epitope = [p for p in range(pos, pos + epitope_length) if (p + 1) % 3]
[…]

### 2013-08-16 scripts/fixation_probability.py
Areas of data, more figs on correlations

@@ -262,6 +262,7 @@ if __name__ == '__main__':
     # Calculate areas
     # Note: skip the last point (via i_final < len(x) - 1) because many simulations
     # have nothing there (not enough time or very noisy)
+    nu0mid = [0.5 * sum(nu0s) for nu0s in nu0ss]
     x = [0] + list(nu0mid) + [1]
     ysyn = [0] + list(Pf['syn_Shanka'][:, 0]) + [1]
     i_final_offset = 2              # Skip last bin
@@ -270,7 +271,6 @@ if __name__ == '__main__':
     Asyn = - Atot + sum([0.5 * (ysyn[i+1] + ysyn[i]) * (x[i+1] - x[i]) for i in xrange(i_final)])
 
     # Prepare figure and plot results
-    nu0mid = [0.5 * sum(nu0s) for nu0s in nu0ss]
     nu0err = [0.5 * (nu0s[1] - nu0s[0]) for nu0s in nu0ss]
     fig, ax = ppl.subplots(1,1)
     for cla in ['nonsyn', 'syn_nonShanka', 'syn_Shanka']:

### 2013-08-16 scripts/script_fixation_loss_simulations_final.py
Fig 4A

@@ -48,7 +48,7 @@ IS_ZERO = 0.000001
 max_counts = 10000
 nirands = 10
 sfs_bins = np.logspace(-3, -0.1, 30)
-savefig = True
+savefig = False
 sep = '\t'
 
 
@@ -553,15 +553,22 @@ if __name__ == '__main__':
                          'ytick.labelsize': 17,
                          'text.usetex': True})
         import matplotlib.pyplot as plt
+        #plt.figure()
+        #plt.plot(gs, divergence, c='b', lw=2, label='Divergence')
+        #plt.plot(gs, diversity, c='r', lw=2, label='Diversity')
+        #plt.xlabel('Time [generations]')
+        #plt.legend(loc=2)
+        #plt.tight_layout()
+        #if savefig:
+        #    for ext in ['svg', 'pdf']:
+        #        plt.savefig('/home/fabio/publications/synmut/figures/simulations_diversity_divergence_example.'+ext)
+
+        # Plot the fixation probability
         plt.figure()
-        plt.plot(gs, divergence, c='b', lw=2, label='Divergence')
-        plt.plot(gs, diversity, c='r', lw=2, label='Diversity')
-        plt.xlabel('Time [generations]')
-        plt.legend(loc=2)
-        plt.tight_layout()
-        if savefig:
-            for ext in ['svg', 'pdf']:
-                plt.savefig('/home/fabio/publications/synmut/figures/simulations_diversity_divergence_example.'+ext)
+        plt.plot([0, 1], [0, 1], lw=1.5, ls='--', c='k')
+        plt.plot([0] + list(nu0ss.mean(axis=1)) + [1], [0] + list(Psyn) + [1], lw=2)
+        plt.xlabel('initial frequency')
+        plt.ylabel(r'$P_{fix}$')
 
         plt.ion()
         plt.show()

### 2013-08-16 scripts/simulations_plot_graduallydel.py

@@ -9,6 +9,15 @@ content:    Plot the fixation probability of increasingly deleterious sims and
 import sys
 import os
 import numpy as np
+
+from matplotlib import rcParams
+rcParams.update({'axes.labelsize': 20, 
+                 'text.fontsize': 20,
+                 'legend.fontsize': 20,
+                 'xtick.labelsize': 20,
+                 'ytick.labelsize': 20,
+                 'text.usetex': True})
+
 import matplotlib.pyplot as plt
 from matplotlib.patches import Rectangle
 import matplotlib.cm as cm
@@ -16,104 +25,85 @@ import matplotlib.cm as cm
 
 
 # Globals
-results_dir = '../data/simulations/gradually_more_del_6epitopes/'
-datafile = 'parameters_fixation_areas.dat'
-
+results_dir = '../data/simulations/gradually_del/'
+subdirs = ['low', 'medium', 'high']
+del_effs = {'low': 0.003, 'medium': 0.001, 'high': 0.0003,
+            'very_high': 0.0001, 'very_low': 0.01} # Multiply by 2!
+colors = ['b', 'g', 'r']
+symbols = ['o', 's', '^']
 
 
 # Script
 if __name__ == '__main__':
 
-    # Load data using record arrays
-    tb = np.genfromtxt(results_dir+datafile, unpack=False,
-                       dtype=zip(('del_eff', 'ada_eff',
-                                  'del_frac', 'ada_rate',
-                                  'trashed', 'good',
-                                  'Asyn', 'Anonsyn',
-                                  'count_syn', 'counts_nonsyn',
-                                  'subs_syn', 'subs_nonsyn',
-                                  'poly_syn', 'poly_nonsyn',
-                                  'foldername'),
-                                 (float, float,
-                                  float, float,
-                                  int, int,
-                                  float, float,
-                                  float, float,
-                                  float, float,
-                                  float, float,
-                                  'S200')))
-
-    # Filter out NaNs
-    is_nonnan = -np.isnan(tb['Asyn'])
-    tb = tb[is_nonnan]
[…]

### 2013-08-16 scripts/simulations_plot_graduallydel.py
fixed supplementary marginals

@@ -105,5 +105,3 @@ if __name__ == '__main__':
             plt.savefig('../figures/simulations_graduallydel.'+ext)
 
     plt.ion()
-    plt.show()
-

### 2013-08-16 scripts/simulations_plot_graduallydel.py
fixup for fig 4A

@@ -36,7 +36,7 @@ symbols = ['o', 's', '^']
 # Script
 if __name__ == '__main__':
 
-    fig = plt.figure()
+    fig, ax0 = plt.subplots(1, 1)
 
     # Plot fixation probability
     for i, subdir in enumerate(subdirs):
@@ -66,13 +66,12 @@ if __name__ == '__main__':
               width=0.007, head_width=0.05, head_length=0.12,
               overhang=0.3)
     plt.xlim(-0.05, 1.05)
-    plt.ylim(-0.05, 1.05)
+    plt.ylim(-0.05, 1.15)
 
     # Inset: plot synonymous diversity
-    from matplotlib.patches import Rectangle
-    plt.gca().add_patch(Rectangle((-0.03, 0.50), 0.52, 0.53,
-                                  edgecolor='k', facecolor='none', lw=1.2))
-    ax = fig.add_axes([0.25, 0.67, 0.25, 0.25])
+    ax0.add_patch(Rectangle((-0.03, 0.66), 0.52, 0.48,
+                            edgecolor='k', facecolor='none', lw=1.2))
+    ax = fig.add_axes([0.25, 0.73, 0.25, 0.22])
     div = []
     for i, subdir in enumerate(subdirs):
         div.append(np.loadtxt(results_dir+subdir+'/poly_25to75.dat',
@@ -81,7 +80,7 @@ if __name__ == '__main__':
     del_eff_x = [del_effs[l] for l in subdirs]
     ax.plot(del_eff_x, div, lw=2, c='k')
     ax.plot([3e-4, 2.5e-4], [9e-3, 1.5e-3], lw=1.2, c='purple')
-    ax.text(1.5e-4, 2e-4, 'observed\ndiversity', fontsize=16)
+    ax.text(1.5e-4, 1.8e-4, 'observed\ndiversity', fontsize=16)
 
     ax.set_xlabel(r'$s_d$', fontsize=18)
     ax.set_ylabel(r'$P_{interm}$', fontsize=18)
@@ -93,10 +92,30 @@ if __name__ == '__main__':
     ax.set_xscale('log')
     ax.set_yscale('log')
     ax.set_xlim(1e-4, 1e-2)
+    ax.set_ylim(1e-4, 8e-2)
+
+    # Second inset: data plot
[…]

### 2013-08-16 scripts/simulations_plot_graduallydel.py
fixup for fig 4A

@@ -71,7 +71,7 @@ if __name__ == '__main__':
     # Inset: plot synonymous diversity
     ax0.add_patch(Rectangle((-0.03, 0.66), 0.52, 0.48,
                             edgecolor='k', facecolor='none', lw=1.2))
-    ax = fig.add_axes([0.25, 0.73, 0.25, 0.22])
+    ax = fig.add_axes([0.25, 0.74, 0.25, 0.21])
     div = []
     for i, subdir in enumerate(subdirs):
         div.append(np.loadtxt(results_dir+subdir+'/poly_25to75.dat',
@@ -79,7 +79,7 @@ if __name__ == '__main__':
     div = np.array(div)
     del_eff_x = [del_effs[l] for l in subdirs]
     ax.plot(del_eff_x, div, lw=2, c='k')
-    ax.plot([3e-4, 2.5e-4], [9e-3, 1.5e-3], lw=1.2, c='purple')
+    ax.plot([3e-4, 2.5e-4], [9e-3, 2e-3], lw=1.2, c='purple')
     ax.text(1.5e-4, 1.8e-4, 'observed\ndiversity', fontsize=16)
 
     ax.set_xlabel(r'$s_d$', fontsize=18)
@@ -95,9 +95,9 @@ if __name__ == '__main__':
     ax.set_ylim(1e-4, 8e-2)
 
     # Second inset: data plot
-    ax0.add_patch(Rectangle((-0.03, 0.3), 0.32, 0.35,
+    ax0.add_patch(Rectangle((-0.03, 0.31), 0.32, 0.35,
                             edgecolor='k', facecolor='none', lw=1.2))
-    ax2 = fig.add_axes([0.18, 0.44, 0.195, 0.18])
+    ax2 = fig.add_axes([0.18, 0.445, 0.195, 0.18])
     x = np.linspace(0, 1, 1000)
     y = (1.4 * x)**3
     ind = x >= y

### 2013-08-16 synmut.tex
edited discussion and other parts

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\linespread{1.5}
+%\linespread{1.5}
 \begin{document}
@@
 Intrapatient HIV-1 evolution is driven by the adaptive immune system
-resulting in rapid change of HIV-1 proteins. When cytotoxic CD8${}^+$ T-cells
-or neutralizing antibodies target a new epitope, the virus often escapes via
-nonsynonymous mutations that impair recognition. Synonymous mutations do not
-affect this interplay and are often assumed to be neutral. We test this
-assumption by tracking  synonymous mutations in 
-longitudinal intrapatient data from the \shankaregion{} part of the \env{}
-gene. We find that most synonymous variants are lost even though
+resulting in rapid change of HIV-1 proteins. When cytotoxic CD8${}^+$
+T-cells or neutralizing antibodies target a new epitope, the virus often
+escapes via nonsynonymous mutations that impair recognition. Synonymous
+mutations do not affect this interplay and are often assumed to be
+neutral. We test this assumption by tracking synonymous mutations in
+longitudinal intrapatient data from the \shankaregion{} part of the
+\env{} gene. We find that most synonymous variants are lost even though
 they often reach high frequencies in the viral population, suggesting a
-cost to the virus. Using published data
-from SHAPE assays, we find that synonymous mutations that disrupt base pairs
-in RNA stems flanking the variable loops of gp120 are more likely to be lost than other
-synonymous changes: these RNA hairpins might be important for HIV-1.
-Computational modeling indicates that (i) most synonymous mutations in this
-genomic region have a selection coefficient of the order of $-0.003$, and
-(ii) synonymous variants rise to high frequency by genetic hitchhiking on neighboring beneficial
-variants. This weak selection against synonymous substitutions does not
-result in a strong pattern of conservation in cross sectional data, but
-slows down the rate of evolution considerably. Our findings are consistent with the
-notion that large scale patterns of RNA structure are functionally
-relevant, while the precise base pairing pattern is not.
+cost to the virus. Using published data from SHAPE assays, we find that
+synonymous mutations that disrupt base pairs in RNA stems flanking the
+variable loops of gp120 are more likely to be lost than other synonymous
+changes: these RNA hairpins might be important for HIV-1.  Computational
+modeling indicates that most synonymous mutations in this genomic region
+have a selection coefficient of the order of $-0.003$, and that the
+majority of synonymous mutations need to be slightly deleterious to be
+consistent with the data. This weak selection against synonymous
+substitutions does not result in a strong pattern of conservation in
+cross sectional data, but slows down the rate of evolution
+considerably. Our findings are consistent with the notion that large
+scale patterns of RNA structure are functionally relevant, while the
+precise base pairing pattern is not.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 abram_nature_2010}. In our simulations, we vary the mutations rate
-from $?? -?? \times 10^{-5}$ per day. Different virions within an
+from $10^{-5} - 3\times 10^{-5}$ per day. Different virions within an
 infected person can recombine their genome by coinfection of the same
@@
 confirmed experimentally \citep{josefsson_single_2013}. We assume a
-template switching rate of $10^{3}$ per nucleotide and vary the
-recombination rate between $??-??\times 10^{-5}$ by adjusting the
+template switching rate of $10^{-3}$ per nucleotide and vary the
+recombination rate between $7\times 10^{-6}- 3\times 10^{-5} 10^{-5}$ by adjusting the
 coinfection rate. The population size relevant to evolution in chronic
@@
 deemed synonymous and assigned either a selection coefficient $0$ with
-probability $1-\alpha$ or a deleterious effect $s_d$ with probability $\alpha$.
+probability $\alpha$ or a deleterious effect $s_d$ with probability
+$1-\alpha$. In our simulations, $\alpha$ varies from 0 to 1.
 Mutations at the first and second positions were assigned strongly deleterious 
-fitness effects ??? COULD MAKE THIS STRONGER, NEED CITATION. At 
+fitness effects of $-???$. At 
 rate $k_A$, a random locus in the genome is designated an epitope that can
@@
 The rate at which new anti-body challenges arise is more difficult to
-quantify, but the majority of non-synonymous substitutions in gp120 are
+quantify, but a substantial fraction of non-synonymous substitutions in gp120 are
 probably driven by escape. Since the sequence of the C2-V5 region of
@@
 \citep{shankarappa_consistent_1999}, this rate should be on the order of
-$k_A=0.003$ per day.
+$k_A=0.005$ per day. We simulate a range from $10^{-4}$ to $10^{-2}$ per day.
 
@@
 evolving immune system, the beneficial effect of an escape mutation was set
-to its previous cost of -0.02 with a probability per generation proportional to
+to its previous cost of -???? with a probability per generation proportional to
 the frequency of the escape variant.
+TODO
 
@@
 genetic diversity, new epitopes were introduced at a constant rate
-$k_A$.  The simulations were repeated 1000 times with a four dimensional
+$k_A$.  The simulations were repeated 1000 times with a seven dimensional
 Latin hypercube sampling scheme \citep{mckay_comparison_1979}.  For each
@@
 estimated from the binned fixation probabilities using linear interpolation
-between the bin centers. This measure is sufficiently precise for our purposes.
-In 4 runs out of 1000, the highest frequency bin was empty so the fixation
-probability could not be calculated; those runs were excluded from
-\FIG{simsfig}.
-
-% TODO change mu and rho and N
+between the bin centers. The bins used were $[0.1, 0.2], [0.2, 0.4],
+[0.4, 0.6]$, which is sufficiently precise for our purposes. Note that
+we did not consider the frequency interval between 0.6 and 1, since very
+few variants are observed in this window and its inclusion generates
+more noise than signal.
 
@@
 probability of fixation, $\pfix$, is shown as a function of the initial
-frequency $\nu_0$ in panel \ref{fig:fixp2}. As a control, the neutral result is
+frequency $\nu_0$ in panel \ref{fig:fixp2}. The neutral result is
 shown as a black dashed line. We found that $\pfix$ of synonymous
@@
 \shankaregion{} region, the sequence samples from \citet{bunnik_autologous_2008}
-are fully compatible with data from \citet{shankarappa_consistent_1999}:
-fixation is rare. The
+are fully compatible with data from \citet{shankarappa_consistent_1999}
+and show that synonymous mutations fix less often then expected under neutrality. The
 nonsynonymous SNVs seem to follow more or less the neutral expectation (blue
@@
 The region \shankaregion{} undergoes frequent adaptive changes to evade
-recognition by neutralizing antibodies \cite{williamson_adaptation_2003,
+recognition by neutralizing antibodies \cite{williamson_adaptation_2003,wei_antibody_2003,
 richman_rapid_2003}. Due to the limited amount of recombination in HIV-1
@@
 \FIG{aft}, which shows that many SNVs change rapidly in frequency as a
-flock.
+cohort.
 
@@
 deleterious effect of synonymous mutations, the fraction of synonymous mutations
-that are deleterious, the escape rate (selection coefficient) of adaptive
+that are neutral, the escape rate (selection coefficient) of adaptive
 nonsynonymous mutations and the rate at which previously untargeted epitopes
@@
 $-0.5$; if every SNV fixes, the area will be $+0.5$. In HIV-1 infected
-patients, we find $P_\text{interm} \approx 0.005$, $A_\text{syn} \approx -0.2$
+patients, we find $P_\text{interm} \approx 0.01$, $A_\text{syn} \approx -0.2$
 for synonymous changes and $A_\text{nonsyn} \approx 0$ for nonsynonymous
@@
 To map the parameter range of the model that is compatible with the data, we
-repeatedly simulated the evolution with random choices for the
-parameters, see \FIG{simsfig}. Among all simulations, we selected the ones
+repeatedly simulated the evolution for many different combinations of
+effects of synonymous mutations, $s_d$, and the fraction of synonymous
+sites that are neutral, $\alpha$, escape rates and number of targeted epitopes. In
+addition, we vary the mutation rate, recombination rate and population
+size, see methods.  Among all simulations, we selected the ones
 that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
 large depression in fixation probability of synonymous SNVs but, simultaneously,
-a moderately high synonymous diversity. Specifically, \FIG{simsfig} shows
-parameter combinations for which we found $A_\text{syn} < -0.15$ and $0.0025 <
-P_\text{interm} < 0.010$. This subset of parameters indicates that a high fraction
-($\gtrsim 0.8$) of sites has to be deleterious with effect size $|s_d| \sim
-0.003$. The inverse of $|s_d|$ is of the same order of magnitude as the
-fixation/extinction times estimated above (see \FIG{fixp1}).
+a moderately high synonymous diversity. Figure \ref{fig:simsfig} shows
+the distribution of deleterious effects $s_d$ and neutral fraction
+$\alpha$ for which we found $A_\text{syn} < -0.15$ and $0.0025 <
+P_\text{interm} < 0.010$. This subset of parameters indicates that in
+order to be consistent with the data only
+a minority of synonyous sites ($\lesssim 0.2$) can be neutral  and that
+the deleterious effects of the remainder is on the order of $|s_d| \sim
+0.003$.
 
-The bounds on the fraction of deleterious synonymous SNVs and
-on their effect sizes are plausible:
-(i) a substantial depression in $\pfix$ requires pervasive deleterious SNVs, otherwise
+The observed distribution of $\alpha$ and $s_d$ is plausible:
+(i) A substantial depression in $\pfix$ requires pervasive deleterious SNVs, otherwise
 the majority of SNVs reaching high frequency are neutral and no depression is
-observed; (ii) in order to hitchhike, the deleterious effect size has to be much
+observed. (ii) In order to hitchhike, the deleterious effect size has to be 
 smaller than the escape rate, otherwise the double mutant (with both the escape
 mutation and the deleterious synonymous one) has little or no fitness advantage
-over the wildtype virus. In agreement with this argument, larger deleterious
-effects in \FIG{simsfig} correspond to larger escape rates; and (iii) SNVs
-with a deleterious effect smaller than approximately $0.001$ behave neutrally,
+over the wildtype virus. (iii) The inverse of $|s_d|$ is of the same order of magnitude as the
+fixation/extinction times estimated above (see \FIG{fixp1}). (iv)
+The selection coefficient $s_d$ tends to be larger than $0.001$, since mutations with smaller
+selection coefficients behave neutrally and often go to fixation
 as expected from the typical coalescent times observed in HIV-1.
@@
 The above simulations show that hitchhiking with favoured nonsynonymous
-variants can explain the observation of
-deleterious synonymous SNVs that rarely fix. However, in a simple model where
-nonsynonymous escape mutations are unconditionally beneficial, escape mutations almost
+variants can explain the observated pattern of deleterious synonymous SNVs that
+rarely fix. While our model is already quite complicated with many
+poorly known parameters that make interpretation of simulation results
+challenging, we know that the actual evolutionary dynamics within an
+infected person is much more complicated still. In our model, 
+ escape mutations are unconditionally beneficial and almost
 always fix once they reach high frequencies, i.e., $A_{\mathrm{nonsyn}}$ is well
-above zero. This is incompatible with the observed fixation probability
-nonsynonymous SNVs at high frequency often disappear again (\FIG{fixp2}), even
-though many are at least transiently beneficial. Inspecting the trajectories of
+above zero. This is incompatible with the observed fixation probability of
+nonsynonymous SNVs at high frequency often disappear again
+(\FIG{fixp2}). One possible reason for this discrepancy is the fact that
+the fitness landscape in our model is static: Once a site is designated
+as targeted by the immune system, mutations at this site are beneficial
+forever. Inspecting the trajectories of
 nonsynonymous SNVs suggests the rapid rise and fall of many SNVs. Two
@@
 
-Several different escape mutations that arise within the same epitope
-have similar effects. Their benefits are not additive, because each of
+Furthermore, several different escape mutations might arise within the same epitope.
+Their benefits are not additive, because each of
 them is essentially sufficient to escape and no additional benefit is
-gained from combining them. As a consequence, several escape SNVs rise
+gained from combining them. As a consequence, several escape SNVs can rise
 to high frequency rapidly, while the one with the smallest cost in terms
@@
 frequencies rapidly but most of the are lost again
-\citep{fischer_transmission_2010}. We implemented within epitope
+\citep{fischer_transmission_2010}. Consequently, there could be many
+non-synonymous mutations are only beneficial until the viral population
+has found a ``better solution'' and subsequently disappear.
+
+We implemented within epitope
 competition in the model by allowing for multiple escape mutations per
 epitope that do not provide additional benefit to the virus when
-combined. Again, we found that the potential for hitchhiking is retained
-by within-epitope competition but that the fixation probability of
-nonsynonymous SNVs is reduced. With roughly six mutations per epitope,
+combined. Again, we found that the rapid rise of several non-synonymous
+mutations to intermediate frequencies allows deleterious synonymous
+mutations to hitch-hike, while the fixation probability of non-synonymous
+mutations is reduced. With roughly six mutations per epitope,
 the simulation data are compatible with observations; see
@@
 competition between equivalent escape pathways, are not exclusive and
-possibly both important in HIV-1 evolution.
+possibly both important in HIV-1 evolution. Our modeling merely suggest
+possible scenarios and to determine to what extend
+either of the of these effects contribute to the observed pattern, more
+data is required.
 
@@
 against synonymous substitutions in the relatively conserved regions
-C2-C4. We suggest through computational modeling that these
-SNVs have deleterious effects on the order of $0.003$ and that they are
-brought to high frequency through linkage to adaptive mutations.
+C2-C4. We suggest through computational modeling that these SNVs have
+deleterious effects on the order of $s_d = 0.003$ and that the majority
+of all synonymous mutations is deleterious. In absense of hitch-hiking
+in a large population, deleterious mutations with effect $s_d$ should be
+at a frequency $\mu/s_d\approx 0.01$. The fact that we see many
+synonymous mutation at high frequency that nevertheless disappears over
+a time scale of a year suggests that they are brought to high frequency
+by hitch-hiking on favorable genetic backgrounds. 
 While both the rise and fall of synonymous SNV is subject to hichhiking,
@@
 While classical population genetics assumes that the dominant stochastic force
-is genetic drift, i.e., non-heritable fluctuations in offspring number, our
-results show that stochasticity due to linked selection is much more important.
+is genetic drift, i.e., non-heritable fluctuations in offspring number,
+in large populations like HIV stochasticity due to linked selection is much more important.
 Such fluctuations have been termed \emph{genetic draft} in

### 2013-08-17 synmut.tex
Fixes in the bulk, ready for me (Fabio)

@@
 changes: these RNA hairpins might be important for HIV-1.  Computational
-modeling indicates that most synonymous mutations in this genomic region
-have a selection coefficient of the order of $-0.003$, and that the
-majority of synonymous mutations need to be slightly deleterious to be
-consistent with the data. This weak selection against synonymous
+modeling indicates that, to be consistent with the data, a large fraction
+of synonymous mutations in this genomic region needs to be deleterious,
+with a selection coefficient of the order of $-0.002$. This weak selection against synonymous
 substitutions does not result in a strong pattern of conservation in
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-For the estimates of time to fixation/extinction, SNVs were binned by
+For the estimates of time to fixation/extinction, single nucleotide variants
+(SNVs) were binned by
 frequency and the time to first reaching either fixation or extinction was
@@
 Mutations at the first and second positions were assigned strongly deleterious 
-fitness effects of $-???$. At 
+fitness effects of $-0.01$ (simulations with a larger cost of $-0.1$ were also
+performed and produced similar results). At 
 rate $k_A$, a random locus in the genome is designated an epitope that can
@@
 10^{-2.5}$ and $2 \cdot 10^{-1.5}$ per day. 
-The rate at which new anti-body challenges arise is more difficult to
-quantify, but a substantial fraction of non-synonymous substitutions in gp120 are
+The rate at which new antibody challenges arise is more difficult to
+quantify, but a substantial fraction of nonsynonymous substitutions in gp120 are
 probably driven by escape. Since the sequence of the C2-V5 region of
@@
 evolving immune system, the beneficial effect of an escape mutation was set
-to its previous cost of -???? with a probability per generation proportional to
+to its previous cost of $-0.01$ with a probability per generation proportional to
 the frequency of the escape variant.
-TODO
 
@@
 throughout the simulation. After 30 generations of burn-in to create
-genetic diversity, new epitopes were introduced at a constant rate
-$k_A$.  The simulations were repeated 1000 times with a seven dimensional
-Latin hypercube sampling scheme \citep{mckay_comparison_1979}.  For each
-parameter, 1000 random values were chosen from a distribution uniform in
-log-space and bounded by the ranges given for each of these parameters
-above. These 1000 values for each parameter were combined using a Latin
-hypercube sampling scheme.
+genetic diversity, new epitopes were introduced at later times according 
+to an exponential distribution with rate
+$k_A$. The simulations were repeated 3000 times with a seven dimensional
+Latin hypercube sampling scheme \citep{mckay_comparison_1979}, bounded by the ranges
+given for each of these parameters above.
 
@@
 genome (in \shankaregion{} but not in the rest of \env{}; compare the red
-triangle line versus the green square line in \FIG{fixp2}).
+and green lines in \FIG{fixp2}).
 The region \shankaregion{} undergoes frequent adaptive changes to evade
@@
 addition, we vary the mutation rate, recombination rate and population
-size, see methods.  Among all simulations, we selected the ones
+size (see methods).  Among all simulations, we selected the ones
 that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
@@
 the deleterious effects of the remainder is on the order of $|s_d| \sim
-0.003$.
+0.002$.
 
@@
 
-The above simulations show that hitchhiking with favoured nonsynonymous
-variants can explain the observated pattern of deleterious synonymous SNVs that
+The above simulations show that hitchhiking on favoured nonsynonymous
+variants can recapitulate the observated pattern of deleterious synonymous SNVs that
 rarely fix. While our model is already quite complicated with many
@@
 \citep{fischer_transmission_2010}. Consequently, there could be many
-non-synonymous mutations are only beneficial until the viral population
+nonsynonymous mutations are only beneficial until the viral population
 has found a ``better solution'' and subsequently disappear.
@@
 epitope that do not provide additional benefit to the virus when
-combined. Again, we found that the rapid rise of several non-synonymous
+combined. Again, we found that the rapid rise of several nonsynonymous
 mutations to intermediate frequencies allows deleterious synonymous
-mutations to hitch-hike, while the fixation probability of non-synonymous
+mutations to hitchhike, while the fixation probability of nonsynonymous
 mutations is reduced. With roughly six mutations per epitope,
@@
 C2-C4. We suggest through computational modeling that these SNVs have
-deleterious effects on the order of $s_d = 0.003$ and that the majority
-of all synonymous mutations is deleterious. In absense of hitch-hiking
+deleterious effects on the order of $s_d = 0.002$ and that the majority
+of all synonymous mutations is deleterious. In absense of hitchhiking
 in a large population, deleterious mutations with effect $s_d$ should be
@@
 a time scale of a year suggests that they are brought to high frequency
-by hitch-hiking on favorable genetic backgrounds. 
+by hitchhiking on favorable genetic backgrounds. 
 While both the rise and fall of synonymous SNV is subject to hichhiking,
@@
 The weak selection against synonymous variants is compatible with the
-negative results of \textit{In vitro} replication assays investigating
+negative results of \textit{in vitro} replication assays investigating
 the fitness effects of small RNA hairpins in HIV-1
@@
 diversity, measured by the probability of a SNV to be found at
-intermediate frequencies $P_\text{interm}$ (first inset).
+intermediate frequencies $P_\text{interm}$ (first inset). The area observed in
+the data is also shown (second inset).
 Panel B) To assess the parameter space that affects synonymous fixation and
-diversity, we run 3000 simulations with random parameters for deleterious effect
+diversity, we run 3000 simulations with random combinations of parameters for deleterious effect
 size, fraction of deleterious synonymous sites, average escape rate $\epsilon$,
@@
 the synonymous diversity and fixation patterns observed in data are shown.
-These simulations demonstrate that deleterious effects are around $-0.003$
+These simulations demonstrate that deleterious effects are around $-0.002$
 and a large fraction of the synonymous mutations needs to be deleterious. This

### 2013-08-17 response_to_reviewers.txt
minor edits

@@
 sampling scheme for parameters in the computer simulations and now scan
-a larger parameter space. All our conclusions remain valid, but are more
-reliable now. We detail these improvements in our point to point
+a larger parameter space. All our conclusions remain valid, but are based
+on more exhaustive sampling. We detail these improvements in our point to point
 response below.
@@
 generation of one day for simplicity. Factor of 2 changes in the
-generation have little influence on the dynamics and only
-renormalize genetic drift. Genetic drift is not a dominant force in HIV
-evolution and the population size, the main determinant of drift, is not
-known to this accuracy anyway. We have repeated a subset of the
+generation have little influence on the dynamics and only renormalize
+genetic drift. Genetic drift is not a dominant force in evolution of the
+HIV envelope gene. The (effective) population size, the main determinant of drift,
+is not known to this accuracy anyway. We have repeated a subset of the
 simulations with a generation time of one day, finding no
-difference.
-
-IS THIS SHOWN SOMEWHERE? (Fabio is redoing the simulations)
+difference. Our more exhaustive survey of the parameter space includes a
+10-fold variation of the population size. Our results are very
+insensitive and hence also insensitive to the generation time.
 
@@
 
-The distribution of allowed deleterious is sharply peaked around
-0.02. DISCUSS THE DIVERSITY ISSUES.
+The distribution of allowed deleterious effects is again peaked around
+0.002 and the fraction of "synonymous" mutations that can be neutral is
+low, as reported in the earlier version of our manuscript.
 
@@
 space, we did implement a Latin hypercube sampling scheme. We thank the
-reviewer for this very useful suggestion. WE MIGHT WANT TO SHOW THE RANK
-CORRELATION MATRIX BETWEEN THE DIFFERENT VARIABLES.
+reviewer for this very useful suggestion.
 
@@
 Yes, much of the stochastic component in the dynamics of the frequencies
-of synonymous mutations will be due to draft. The consistent loss of
-variants, conditional on them being at high frequency earlier, implies
-selection against these variants. We have added a sentence to the
-discussion to make this point clear.
-
-Fabio: I DO NOT THINK THAT ADRESSES THE QUESTION ("IN GENERAL" --> "BUT").
-MAYBE WE COULD STATE SOMETHING LIKE: "IN ABSENCE OF DELETERIOUS EFFECTS, THE
-FIXATION PROBABILITY OF HITCHHIKERS IS NOT AFFECTED BY DRAFT".
+of synonymous mutations will be due to draft. However, the consistent
+loss of variants, conditional on them being at high frequency earlier,
+implies selection against these variants. Without a deleterious effect,
+the mutations would fix with a probability equal to their frequency,
+regardless of whether they are subject to drift of draft. We have added
+a sentence to the discussion to make this point clear.
 
@@
 tentative. Qualitatively, however, the processes we discuss have been
-observed and we do cite experimental evidence. It is common for escape
-to happen in multiple equivalent ways and this has been well documented
-fro early CTL escape and early nAb escape (Wei et al, Nature, 2003. Bar
-et al, PPath, 2012). Data for chronic infection is less conclusive and
-selection pressures are generally smaller, but turnover at targeted
-epitopes and the replacement of variants that where previously at high
-frequencies was observed 3y into the infection in Wei et al, Nature, 2003.
-Whether this fall in frequency of some non-synonymous variants happens
-because they are replaced by a less costly escape variant or because the
-escape variant itself is targeted is hard to disentangle. We think it is
-reasonable that both of these effects contribute. Whether they are
-sufficient to produce the observed fixation patterns of non-synonymous
-mutations is largely an empirical question. We simply explored using
-numerical experiments whether these effects push the curve in the
-"right" direction.
+observed. We now cite additional experimental evidence (Wei et al,
+2003). It is common for escape to happen in multiple equivalent ways and
+this has been well documented fro early CTL escape and early nAb escape
+(Wei et al, Nature, 2003. Bar et al, PPath, 2012). Data for chronic
+infection is less conclusive and selection pressures are generally
+smaller, but turnover at targeted epitopes and the replacement of
+variants that where previously at high frequencies was observed 3y into
+the infection in Wei et al, Nature, 2003.  Whether this fall in
+frequency of some non-synonymous variants happens because they are
+replaced by a less costly escape variant or because the escape variant
+itself is targeted is hard to disentangle. We think it is reasonable
+that both of these effects contribute. Whether they are sufficient to
+produce the observed fixation patterns of non-synonymous mutations is
+largely an empirical question. We simply explored using numerical
+experiments whether these effects push the curve in the "right"
+direction. We have added a disclaimer that more data is needed to fully
+explain the pattern of fixation observed for non-synonymous mutations.

### 2013-08-17 synmut.tex
minor edits

@@
 of synonymous mutations in this genomic region needs to be deleterious,
-with a selection coefficient of the order of $-0.002$. This weak selection against synonymous
+with a cost of the order of $0.002$ per day. This weak selection against synonymous
 substitutions does not result in a strong pattern of conservation in
@@
 function. However, large scale modification of secondary structures
-can results in substantial reduction of the replication capacity
+can result in substantial reduction of the replication capacity
 \citep{keating_rich_2009} and the propensity of forming RNA stems
@@
 resulting curves. Mutations that reached high frequency but neither fixed nor
-were lost were classified as ``floating'', with one exception: if they first
-reached high frequencies within 3 years of the last time point, it was assumed
-they had not had sufficient time to settle, so they were discarded.
+were lost were classified as ``floating'' unless they first
+reached high frequencies within 3 years of the last time point. In that case, it was assumed
+they had not had sufficient time to fix and were discarded.
 
@@
 template switching rate of $10^{-3}$ per nucleotide and vary the
-recombination rate between $7\times 10^{-6}- 3\times 10^{-5} 10^{-5}$ by adjusting the
+recombination rate between $5\times 10^{-6}- 5\times 10^{-5}$ by adjusting the
 coinfection rate. The population size relevant to evolution in chronic
@@
 $1-\alpha$. In our simulations, $\alpha$ varies from 0 to 1.
-Mutations at the first and second positions were assigned strongly deleterious 
+Mutations at the first and second positions were assigned deleterious 
 fitness effects of $-0.01$ (simulations with a larger cost of $-0.1$ were also
@@
 \citep{shankarappa_consistent_1999}, this rate should be on the order of
-$k_A=0.005$ per day. We simulate a range from $10^{-4}$ to $10^{-2}$ per day.
+$k_A=0.005$ per day. We simulate a range from $10^{-4}$ to $2\times 10^{-2}$ per day.
 
@@
 comparable results. Populations were initialized with a homogeneous
-founder population and were kept at an average size of $N=10^4$
-throughout the simulation. After 30 generations of burn-in to create
-genetic diversity, new epitopes were introduced at later times according 
-to an exponential distribution with rate
+founder population. After 30 generations of burn-in to create
+genetic diversity, new epitopes were introduced at random with rate
 $k_A$. The simulations were repeated 3000 times with a seven dimensional
 Latin hypercube sampling scheme \citep{mckay_comparison_1979}, bounded by the ranges
-given for each of these parameters above.
+given for each of these parameters above. For all parameter but the
+fraction $\alpha$ of neutral sites,  parameters were sampled uniformly in
+log-space.
 
@@
 probability of fixation, $\pfix$, is shown as a function of the initial
-frequency $\nu_0$ in panel \ref{fig:fixp2}. The neutral result is
+frequency $\nu_0$ in panel \ref{fig:fixp2}. The neutral expectation is
 shown as a black dashed line. We found that $\pfix$ of synonymous
@@
 One possible explanation for a reduced fixation of synonymous variants in
-\shankaregion{} is secondary structures in the viral RNA, the disruption of which
+\shankaregion{} is secondary structure in the viral RNA, the disruption of which
 is deleterious to the virus \citep{forsdyke_reciprocal_1995,
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Deleterious SNVs reach high frequency by hitchhiking}
+\subsection*{Deleterious SNVs can reach high frequency by hitchhiking}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 effect of the mutation, this corresponds to deleterious effects $s_d$ of the
-order of $-0.002$ per day. (This is only an average estimate: every single
+order of $0.002$ per day. (This is only an average estimate: every single
 mutation is expected to have a slightly different fitness effect.)
@@
 probability and the diagonal, which is the neutral expectation
-(\FIG{simfixpvar}, lower inset). If no fixation happens, the area will be
-$-0.5$; if every SNV fixes, the area will be $+0.5$. In HIV-1 infected
-patients, we find $P_\text{interm} \approx 0.01$, $A_\text{syn} \approx -0.2$
+(\FIG{simfixpvar}, lower inset). We restrict this area up to frequency
+$\nu=0.5$ since the high frequency part is rather noisy.
+If no fixation happens, this restricted area will be
+$0.125$; if every SNV fixes, the area will be $0.125$. In HIV-1 infected
+patients, we find $P_\text{interm} \approx 0.01$, $A_\text{syn} \approx -0.08$
 for synonymous changes and $A_\text{nonsyn} \approx 0$ for nonsynonymous
@@
 probability of synonymous SNVs decreases from the neutral expectation
-($A_\text{syn} \approx 0$) to zero ($A_\text{syn} \approx -0.5$) as the
+($A_\text{syn} \approx 0$) to zero ($A_\text{syn} \approx -0.125$) as the
 fitness cost of the SNVs increases; the synonymous diversity plummets as well, as
@@
 the distribution of deleterious effects $s_d$ and neutral fraction
-$\alpha$ for which we found $A_\text{syn} < -0.15$ and $0.0025 <
-P_\text{interm} < 0.010$. This subset of parameters indicates that in
+$\alpha$ for which we found $A_\text{syn} < -0.15$ and $0.005 <
+P_\text{interm} < 0.015$. This subset of parameters indicates that in
 order to be consistent with the data only
-a minority of synonyous sites ($\lesssim 0.2$) can be neutral  and that
-the deleterious effects of the remainder is on the order of $|s_d| \sim
+a minority of synonymous sites ($\alpha \lesssim 0.2$) can be neutral  and that
+the deleterious effects of the remainder is on the order of $s_d \sim
 0.002$.
@@
 mutation and the deleterious synonymous one) has little or no fitness advantage
-over the wildtype virus. (iii) The inverse of $|s_d|$ is of the same order of magnitude as the
+over the wildtype virus. (iii) The inverse of $s_d$ is of the same order of magnitude as the
 fixation/extinction times estimated above (see \FIG{fixp1}). (iv)
-The selection coefficient $s_d$ tends to be larger than $0.001$, since mutations with smaller
-selection coefficients behave neutrally and often go to fixation
+The cost $s_d$ tends to be larger than $0.001$, since mutations with smaller
+cost behave neutrally and often go to fixation
 as expected from the typical coalescent times observed in HIV-1.
 
-The above simulations show that hitchhiking on favoured nonsynonymous
-variants can recapitulate the observated pattern of deleterious synonymous SNVs that
+The above simulations show that hitchhiking on favored nonsynonymous
+variants can reproduce the observed pattern of deleterious synonymous SNVs that
 rarely fix. While our model is already quite complicated with many
@@
 context of CTL escape mutations where multiple variants rise to high
-frequencies rapidly but most of the are lost again
+frequencies rapidly but most of them are lost again
 \citep{fischer_transmission_2010}. Consequently, there could be many
@@
 competition between equivalent escape pathways, are not exclusive and
-possibly both important in HIV-1 evolution. Our modeling merely suggest
-possible scenarios and to determine to what extend
-either of the of these effects contribute to the observed pattern, more
-data is required.
+possibly both important in HIV-1 evolution. We note that our modeling merely suggest
+possible scenarios. More data is required to determine to what extend
+either of these mechanisms contribute to the observed pattern.
 
@@
 deleterious effects on the order of $s_d = 0.002$ and that the majority
-of all synonymous mutations is deleterious. In absense of hitchhiking
+of all synonymous mutations is deleterious. In absence of hitchhiking
 in a large population, deleterious mutations with effect $s_d$ should be
 at a frequency $\mu/s_d\approx 0.01$. The fact that we see many
-synonymous mutation at high frequency that nevertheless disappears over
+synonymous mutations at high frequency that nevertheless disappears over
 a time scale of a year suggests that they are brought to high frequency
 by hitchhiking on favorable genetic backgrounds. 
-While both the rise and fall of synonymous SNV is subject to hichhiking,
+While both the rise and fall of synonymous SNV is subject to hitchhiking,
 the consistent loss of these variants conditional on being at high
@@
 Comparison with biochemical data (SHAPE) of base pairing propensity in the RNA
-genome of HIV-1 indicated that these mutations tend to disrupt RNA secondary structures\citep{watts_architecture_2009}. Computer models
+genome of HIV-1 indicated that these mutations tend to disrupt RNA 
+secondary structures\citep{watts_architecture_2009}. Computer models
 of RNA folding predict stable hairpins in these regions that
@@
 passages to detect fitness effects of the order of one per mille. The longitudinal data,
-however, spans many years, hence our analysis is able to quantify the
-subtle fitness effect of RNA structure within single infections. 
+however, spans many years and our analysis is able to quantify the
+subtle fitness effect of RNA structure in intra-patient evolution. 
 The fixation probabilities and the sojourn times of SNVs represent a
@@
 Both mechanisms reduce the overall fixation probability and can
-give rise to the observed pattern of fixation in computer simulations.
+give rise to the observed pattern of fixation in computer
+simulations. More data is required to investigate this behavior further.
 
@@
 \epsilon^{-1} \log \nu_0$ generations for an escape SNV with escape rate
-$\epsilon$ to rise from an initially low frequency $\nu_0\approx \mu$ to frequency
+$\epsilon$ to rise from an initially low frequency $\nu_0$ to frequency
 one. This implies that a region of length $l = (\rho t_{sw})^{-1} = \epsilon /
-\rho \log \nu_0$ remains linked to the adaptive mutation. With $\epsilon=0.01$,
+\rho \log \nu_0$ remains linked to the adaptive mutation. With
+$\epsilon=0.01$, and $\nu_0\approx 10^{3}$, 
 we have $l\approx 100$ bases. Hence we expect strong linkage between the
@@
 the synonymous diversity and fixation patterns observed in data are shown.
-These simulations demonstrate that deleterious effects are around $-0.002$
-and a large fraction of the synonymous mutations needs to be deleterious. This
-picture is also confirmed by the analysis of the marginal distributions of the
-single parameters (see \figurename~S\marginals).}
+These simulations demonstrate that deleterious effects are around $0.002$
+and a large fraction of the synonymous mutations needs to be
+deleterious. The marginal distributions of each parameter are given in
+\figurename~S\marginals.} 
 \label{fig:simheat}

### 2013-08-17 synmut.tex
added data and further corrections

@@
 abram_nature_2010}. In our simulations, we vary the mutations rate
-from $10^{-5} - 3\times 10^{-5}$ per day. Different virions within an
+from $10^{-5} - 4\times 10^{-5}$ per day. Different virions within an
 infected person can recombine their genome by coinfection of the same
@@
 product $N\mu$ is of order one or larger \citep{neher_genetic_2011}. We
-simulate populations between $N=10^4$ and $10^5$. Larger populations
+simulate populations between $N=10^4$ and $5 \times 10^4$. Larger populations
 become prohibitively costly to simulate.
@@
 probability $\alpha$ or a deleterious effect $s_d$ with probability
-$1-\alpha$. In our simulations, $\alpha$ varies from 0 to 1.
+$1-\alpha$. In our simulations, $\alpha$ varies from 0 to 0.75,
+and $s_d$ between $2 \times 10^{-4}$ to $2\times 10^{-2}$ per day.
 Mutations at the first and second positions were assigned deleterious 
@@
 \citep{shankarappa_consistent_1999}, this rate should be on the order of
-$k_A=0.005$ per day. We simulate a range from $10^{-4}$ to $2\times 10^{-2}$ per day.
+$k_A=0.005$ per day. We simulate a range from $10^{-3.8}$ to $2\times 10^{-2}$ per day.
 
@@
 If no fixation happens, this restricted area will be
-$0.125$; if every SNV fixes, the area will be $0.125$. In HIV-1 infected
-patients, we find $P_\text{interm} \approx 0.01$, $A_\text{syn} \approx -0.08$
+$-0.125$; if every SNV fixes, the area will be $0.125$. In HIV-1 infected
+patients, we find $P_\text{interm} \approx 0.01$, $A_\text{syn} \approx -0.09$
 for synonymous changes and $A_\text{nonsyn} \approx 0$ for nonsynonymous
@@
 the distribution of deleterious effects $s_d$ and neutral fraction
-$\alpha$ for which we found $A_\text{syn} < -0.15$ and $0.005 <
+$\alpha$ for which we found $-0.11 < A_\text{syn} < -0.06$ and $0.008 <
 P_\text{interm} < 0.015$. This subset of parameters indicates that in

### 2013-08-17 scripts/plot_colors_sizes_7.py

@@ -1,0 +1,583 @@
+# vim: fdm=marker
+'''
+author:     Fabio Zanini
+date:       14/01/13
+content:    Scatter plot and further visualization of the results of random
+            gradual simulations.
+'''
+# Modules
+import sys
+import os
+import re
+import numpy as np
+
+from matplotlib import rcParams
+rcParams.update({'axes.labelsize': 20, 
+                 'text.fontsize': 20,
+                 'legend.fontsize': 20,
+                 'xtick.labelsize': 20,
+                 'ytick.labelsize': 20,
+                 'text.usetex': True})
+
+from matplotlib.patches import Rectangle
+import matplotlib.cm as cm
+import matplotlib.pyplot as plt
+
+
+# Globals
+datafile = 'parameters_fixation_areas.dat'
+panel_name = 'B'
+fig_name = 'simulations_syn_latin_contour'
+figm_name = 'simulations_marginals_stratified'
+
+# Set limits for good sims
+Asynmin = -0.11
+Asynmax = -0.06
+poly_synmin = 0.008
+poly_synmax = 0.015
+i_final_offset = 2
+
+# Condensation
+recondense = False
+savefig = False
+plot_type = 'contour'
+plot_things = True
+
+
+
+# Functions
+def condense_data(results_dir):
+    '''Condense the results from many simulations into a single file'''
+    results_dir = results_dir.rstrip('/')+'/'
+    table = []
+
+    # List all directories
+    g = os.walk(results_dir)
+    folders = g.next()[1]
+
+    # Collect data from all simulations
+    n_empty = 0
+    for folder in folders:
+        parfile = results_dir+folder+'/parameters.dat'
+        Pfixfile = results_dir+folder+'/Pfix.dat'
+        subfile = results_dir+folder+'/substitutions.dat'
+        polyfile = results_dir+folder+'/poly_25to75.dat'
+
+        if not all(map(os.path.isfile, (parfile, Pfixfile, subfile, polyfile))):
+            n_empty += 1
+            continue
+
+        # Load data on parameters
+        try:
+            pars = np.loadtxt(parfile)
+        except ValueError:
+            continue
+
+        # Load data on Pfix
+        (nu0a, nu0b,
+         Psyn, Pnonsyn,
+         csyn, cnonsyn) = np.genfromtxt(Pfixfile, comments='#', unpack=True,
+                                        missing_values=['--'], usemask=True)
+
+        # Calculate areas
[…]

### 2013-08-18 synmut.tex
little changes

@@
 simulations where each site in the genome is represented by one bit that can be
-in one of two states. Outcrossing rates, crossover rates, mutations rates and
+in one of two states. Outcrossing rates, crossover rates, mutation rates and
 arbitrary fitness functions can be specified. While the best estimates
@@
 exogenous LacZ constructs and the average nucleotide substitution rate is 
-estimated to be $\mu\approx 2 \times 10^{-5}$ per generation \citep{mansky_lower_1995,
+estimated to be $\mu\approx 2 \cdot 10^{-5}$ per generation \citep{mansky_lower_1995,
 abram_nature_2010}. In our simulations, we vary the mutations rate
-from $10^{-5} - 4\times 10^{-5}$ per day. Different virions within an
+from $10^{-5} - 4\cdot 10^{-5}$ per day. Different virions within an
 infected person can recombine their genome by coinfection of the same
@@
 template switching rate of $10^{-3}$ per nucleotide and vary the
-recombination rate between $5\times 10^{-6}- 5\times 10^{-5}$ by adjusting the
+recombination rate between $5\cdot 10^{-6}- 5\cdot 10^{-5}$ by adjusting the
 coinfection rate. The population size relevant to evolution in chronic
@@
 product $N\mu$ is of order one or larger \citep{neher_genetic_2011}. We
-simulate populations between $N=10^4$ and $5 \times 10^4$. Larger populations
+simulate populations between $N=10^4$ and $5 \cdot 10^4$. Larger populations
 become prohibitively costly to simulate.
@@
 $1-\alpha$. In our simulations, $\alpha$ varies from 0 to 0.75,
-and $s_d$ between $2 \times 10^{-4}$ to $2\times 10^{-2}$ per day.
+and $s_d$ between $2 \cdot 10^{-4}$ to $2\cdot 10^{-2}$ per day.
 Mutations at the first and second positions were assigned deleterious 
@@
 10^{-2.5}$ and $2 \cdot 10^{-1.5}$ per day. 
-The rate at which new antibody challenges arise is more difficult to
+The rate $k_A$ at which new antibody challenges arise is more difficult to
 quantify, but a substantial fraction of nonsynonymous substitutions in gp120 are
-probably driven by escape. Since the sequence of the C2-V5 region of
-gp120  changes at a rate of about 1\% per year
-\citep{shankarappa_consistent_1999}, this rate should be on the order of
-$k_A=0.005$ per day. We simulate a range from $10^{-3.8}$ to $2\times 10^{-2}$ per day.
+probably driven by escape. We simulate a range from $10^{-3.8}$ to $2\cdot 10^{-2}$ per day.
 
@@
 fix with a probability of 50\% (see inset in figure
-\ref{fig:aftsyn}). The fixation probability of neutral SNVs is
+\ref{fig:aft}A). The fixation probability of neutral SNVs is
 independent of most model assumptions and is only affected if neutral
@@
 \citet{shankarappa_consistent_1999}). Despite many synonymous SNVs
-reaching high frequency, very few fix (panel~\ref{fig:aftsyn}); in
+reaching high frequency, very few fix (Fig.~\ref{fig:aft}A); in
 contrast, many nonsynonymous mutations fix
-(panel~\ref{fig:aftnonsyn}). This observation seems at odds with
+(Fig.~\ref{fig:aft}B). This observation seems at odds with
 the assumption of neutrality.
@@
 most synonymous SNVs segregate for roughly one year and are lost much more
-frequently than expected under neutrality (panel \ref{fig:fixp1}). The long-time
+frequently than expected under neutrality (Fig.~\ref{fig:fixp}A). The long-time
 probability of fixation, $\pfix$, is shown as a function of the initial
-frequency $\nu_0$ in panel \ref{fig:fixp2}. The neutral expectation is
+frequency $\nu_0$ in Fig.~\ref{fig:fixp}B. The neutral expectation is
 shown as a black dashed line. We found that $\pfix$ of synonymous
@@
 lost (only variants reaching frequencies above 15\%). As shown in
-\FIG{SHAPEA}, the reactivities of fixed SNVs (red histogram) are systematically
+\FIG{SHAPE}A, the reactivities of fixed SNVs (red histogram) are systematically
 larger than those of lost SNVs (blue) (Kolmogorov-Smirnov test on the cumulative
@@
 V loops show little deviation from the neutral signature, see
-\FIG{SHAPEB}. This is suggestive of important RNA helices in conserved
+\FIG{SHAPE}B. This is suggestive of important RNA helices in conserved
 regions between the V loops.
@@
 genome (in \shankaregion{} but not in the rest of \env{}; compare the red
-and green lines in \FIG{fixp2}).
+and green lines in \FIG{fixp}B).
 The region \shankaregion{} undergoes frequent adaptive changes to evade
@@
 The approximate magnitude of the deleterious effects can be estimated from
-\FIG{fixp1}, which shows the distribution of times after which synonymous
+\FIG{fixp}A, which shows the distribution of times after which synonymous
 SNVs at intermediate frequencies become fixed or lost. The typical time to
@@
 
-\FIG{simfixpvar} shows simulation results for the fixation probability and the
+\FIG{simheat}A shows simulation results for the fixation probability and the
 synonymous diversity for different deleterious effects sizes of synonymous mutations.
@@
 probability and the diagonal, which is the neutral expectation
-(\FIG{simfixpvar}, lower inset). We restrict this area up to frequency
+(\FIG{simheat}A, lower inset). We restrict this area up to frequency
 $\nu=0.5$ since the high frequency part is rather noisy.
@@
 for synonymous changes and $A_\text{nonsyn} \approx 0$ for nonsynonymous
-changes. In the three simulations shown in \FIG{simfixpvar}, the fixation
+changes. In the three simulations shown in \FIG{simheat}A, the fixation
 probability of synonymous SNVs decreases from the neutral expectation
@@
 large depression in fixation probability of synonymous SNVs but, simultaneously,
-a moderately high synonymous diversity. Figure \ref{fig:simsfig} shows
+a moderately high synonymous diversity. Figure \ref{fig:simheat}B shows
 the distribution of deleterious effects $s_d$ and neutral fraction
@@
 over the wildtype virus. (iii) The inverse of $s_d$ is of the same order of magnitude as the
-fixation/extinction times estimated above (see \FIG{fixp1}). (iv)
+fixation/extinction times estimated above (see \FIG{fixp}A). (iv)
 The cost $s_d$ tends to be larger than $0.001$, since mutations with smaller
@@
 nonsynonymous SNVs at high frequency often disappear again
-(\FIG{fixp2}). One possible reason for this discrepancy is the fact that
+(\FIG{fixp}B). One possible reason for this discrepancy is the fact that
 the fitness landscape in our model is static: Once a site is designated
@@
 mutations display a fixation pattern as if they were neutral;
-see~\FIG{fixp2}. However, nonsynonymous diversity exceeds
+see~\FIG{fixp}B. However, nonsynonymous diversity exceeds
 synonymous diversity despite the overall much greater constraints on the
@@
 order of $\rho = 10^{-5}$ per base and day. It takes roughly $t_{sw} =
-\epsilon^{-1} \log \nu_0$ generations for an escape SNV with escape rate
+\epsilon^{-1} |\log \nu_0|$ generations for an escape SNV with escape rate
 $\epsilon$ to rise from an initially low frequency $\nu_0$ to frequency
 one. This implies that a region of length $l = (\rho t_{sw})^{-1} = \epsilon /
-\rho \log \nu_0$ remains linked to the adaptive mutation. With
-$\epsilon=0.01$, and $\nu_0\approx 10^{3}$, 
+|\rho \log \nu_0|$ remains linked to the adaptive mutation. With
+$\epsilon=0.01$, and $\nu_0\approx 10^{-3}$, 
 we have $l\approx 100$ bases. Hence we expect strong linkage between the
@@
 \begin{center}
-\subfloat{\includegraphics[width=0.6\linewidth]
-{Shankarappa_allele_freqs_trajectories_syn_p10.pdf}
-\label{fig:aftsyn}}\\
- \subfloat{\includegraphics[width=0.6\linewidth]
-{Shankarappa_allele_freqs_trajectories_nonsyn_p10.pdf}
-\label{fig:aftnonsyn}}
+\includegraphics[width=0.6\linewidth]{fig1}
+%\subfloat{\includegraphics[width=0.6\linewidth]
+%{Shankarappa_allele_freqs_trajectories_syn_p10.pdf}
+%\label{fig:aftsyn}}\\
+% \subfloat{\includegraphics[width=0.6\linewidth]
+%{Shankarappa_allele_freqs_trajectories_nonsyn_p10.pdf}
+%\label{fig:aftnonsyn}}
 \caption{Time series of frequencies
@@
 \begin{center}
-\subfloat{\includegraphics[width=0.6\linewidth]{fixation_times.pdf}
-\label{fig:fixp1}}\\
-\subfloat{\includegraphics[width=0.6\linewidth]{fixation_probabilities.pdf}
-\label{fig:fixp2}}
+\includegraphics[width=0.6\linewidth]{fig2}
+%\subfloat{\includegraphics[width=0.6\linewidth]{fixation_times.pdf}
+%\label{fig:fixp1}}\\
+%\subfloat{\includegraphics[width=0.6\linewidth]{fixation_probabilities.pdf}
+%\label{fig:fixp2}}
 \caption{Fixation and loss of SNVs.
@@
 \begin{center}
-\subfloat{\includegraphics[height=0.3\linewidth]{reactivities_histograms_syn.pdf}\label{fig:SHAPEA}}
-\subfloat{\includegraphics[height=0.3\linewidth]{fixation_probabilities_VnonV.pdf}\label{fig:SHAPEB}}
+\includegraphics[width=0.9\linewidth]{fig3}
+%\subfloat{\includegraphics[height=0.3\linewidth]{reactivities_histograms_syn.pdf}\label{fig:SHAPEA}}
+%\subfloat{\includegraphics[height=0.3\linewidth]{fixation_probabilities_VnonV.pdf}\label{fig:SHAPEB}}
 \caption{Permissible synonymous mutations tend to be unpaired.
@@
 \begin{center}
-\subfloat{\includegraphics[width=0.6\linewidth]{simulations_graduallydel.pdf}
-\label{fig:simfixpvar}}\\
-\subfloat{\includegraphics[width=0.6\linewidth]{simulations_syn_latin_contour.pdf}
-\label{fig:simsfig}}
+\includegraphics[width=0.6\linewidth]{fig4}
+%\subfloat{\includegraphics[width=0.6\linewidth]{simulations_graduallydel.pdf}
+%\label{fig:simfixpvar}}\\
+%\subfloat{\includegraphics[width=0.6\linewidth]{simulations_syn_latin_contour.pdf}
+%\label{fig:simsfig}}
 \caption{Distribution of selection coefficients on synonymous sites. Panel A)

### 2013-08-18 synmut.tex
final edits, simplified ms to make conversion easier

@@
 \usepackage{graphicx}
-\usepackage[caption=false]{subfig}
 \usepackage[numbers]{natbib}
-\usepackage{pslatex}
 \usepackage{lineno}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\linespread{1.5}
+\linespread{1.5}
 \begin{document}
@@
 show substantial population structure and were excluded (see
-\figurename~S\PCApat); a total of 11 patients with 4-23 time points each and
+Fig.~S\PCApat); a total of 11 patients with 4-23 time points each and
 approximately 10 sequences per time point were analyzed. The time intervals
@@
 \citet{bunnik_autologous_2008} (patients with strong viral population structure
-were not considered; see methods and \figurename~S\PCApat). The
+were not considered; see methods and Fig.~S\PCApat). The
 former data set is restricted to the \shankaregion{} region of \env, while the
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-While the observation of some deleterious synonymous variants is not fully
+While the observation that some synonymous variants are deleterious is not
 unexpected, it seems odd that we observe them at high population
@@
 observations and our interpretation, we performed computer simulations of
-evolving viral populations assuming a mix of positive and purifying selection
+a model of evolving viral populations assuming a mix of positive and purifying selection
 and rare recombination.  For this purpose, we use the simulation package
@@
 synonymous diversity for different deleterious effects sizes of synonymous mutations.
-We quantify synonymous diversity via $P_\text{interm}$, the fraction of sites
+We quantify synonymous diversity via $P_\mathrm{interm}$, the fraction of sites
 with an synonymous SNV at frequency $0.25 < \nu < 0.75$. The synonymous diversity
@@
 If no fixation happens, this restricted area will be
-$-0.125$; if every SNV fixes, the area will be $0.125$. In HIV-1 infected
-patients, we find $P_\text{interm} \approx 0.01$, $A_\text{syn} \approx -0.09$
-for synonymous changes and $A_\text{nonsyn} \approx 0$ for nonsynonymous
+$-0.125$; if every SNV fixes, the area will be $0.125$. In data from HIV-1 infected
+patients, we find $P_\mathrm{interm} \approx 0.01$, $A_\mathrm{syn} \approx -0.09$
+for synonymous changes and $A_\mathrm{nonsyn} \approx 0$ for nonsynonymous
 changes. In the three simulations shown in \FIG{simheat}A, the fixation
 probability of synonymous SNVs decreases from the neutral expectation
-($A_\text{syn} \approx 0$) to zero ($A_\text{syn} \approx -0.125$) as the
+($A_\mathrm{syn} \approx 0$) to zero ($A_\mathrm{syn} \approx -0.125$) as the
 fitness cost of the SNVs increases; the synonymous diversity plummets as well, as
@@
 size (see methods).  Among all simulations, we selected the ones
-that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
+that show $A_\mathrm{syn}$ and $P_\mathrm{interm}$ as observed in the data, i.e., a
 large depression in fixation probability of synonymous SNVs but, simultaneously,
@@
 the distribution of deleterious effects $s_d$ and neutral fraction
-$\alpha$ for which we found $-0.11 < A_\text{syn} < -0.06$ and $0.008 <
-P_\text{interm} < 0.015$. This subset of parameters indicates that in
+$\alpha$ for which we found $-0.11 < A_\mathrm{syn} < -0.06$ and $0.008 <
+P_\mathrm{interm} < 0.015$. This subset of parameters indicates that in
 order to be consistent with the data only
@@
 above zero. This is incompatible with the observed fixation probability of
-nonsynonymous SNVs at high frequency often disappear again
+nonsynonymous SNVs that often disappear again
 (\FIG{fixp}B). One possible reason for this discrepancy is the fact that
@@
 the mutant might cease to be beneficial and disappear soon, despite its
-quick initial rise in frequency. In support of this idea, in
-\citet{richman_rapid_2003,wei_antibody_2003,bunnik_autologous_2008},
-antibody responses against escape mutants have been reported. These
+quick initial rise in frequency. In support of this idea,
+\citet{richman_rapid_2003,wei_antibody_2003,bunnik_autologous_2008}
+report antibody responses against escape mutants. These
 responses are delayed by a few months, roughly matching the average time
@@
 hitchhiking, but reduced fixation of nonsynonymous SNVs.
-\figurename~S\timedependence~shows that $\pfix$ of synonymous SNVs is
+Fig.~S\timedependence~shows that $\pfix$ of synonymous SNVs is
 not affected by this change, while $\pfix$ of nonsynonymous SNVs
@@
 all others are lost. The emergence of multiple competing escape variants
-in HIV-1 infections has been shown \citep{moore_limited_2009,
-bar_early_2012,wei_antibody_2003}.  This scenario has been explicitly
+within a single epitope infections has been shown for CTL and anti-body escape 
+\citep{fischer_transmission_2010,moore_limited_2009, bar_early_2012,wei_antibody_2003}.  
+This scenario has also been explicitly
 observed in the evolution of resistance to 3TC, where the mutation M184V
-is often preceded by M184I \citep{hedskog_dynamics_2010} or in the
-context of CTL escape mutations where multiple variants rise to high
-frequencies rapidly but most of them are lost again
-\citep{fischer_transmission_2010}. Consequently, there could be many
-nonsynonymous mutations are only beneficial until the viral population
-has found a ``better solution'' and subsequently disappear.
+is often preceded by M184I \citep{hedskog_dynamics_2010}. Consequently, there could be many
+nonsynonymous mutations that are only beneficial until the viral population
+has found a ``better solution'' and that subsequently disappear.
 
@@
 the simulation data are compatible with observations; see
-\figurename~S\withinepi. The two scenarios, time-dependent selection and
+Fig.~S\withinepi. The two scenarios, time-dependent selection and
 competition between equivalent escape pathways, are not exclusive and
@@
 C2-C4. We suggest through computational modeling that these SNVs have
-deleterious effects on the order of $s_d = 0.002$ and that the majority
-of all synonymous mutations is deleterious. In absence of hitchhiking
+deleterious effects on the order of $s_d = 0.002$ per day and that the majority
+of all synonymous mutations in this region is deleterious. In absence of hitchhiking
 in a large population, deleterious mutations with effect $s_d$ should be
 at a frequency $\mu/s_d\approx 0.01$. The fact that we see many
-synonymous mutations at high frequency that nevertheless disappears over
+synonymous mutations at high frequency and that these SNVs nevertheless disappears over
 a time scale of a year suggests that they are brought to high frequency
@@
 at synonymous sites varies greatly along the HIV-1 genome
-\citep{mayrose_towards_2007} (see also \figurename~S\syndiv). This variation can confound estimates of
+\citep{mayrose_towards_2007} (see also Fig.~S\syndiv). This variation can confound estimates of
 selection on proteins substantially \citep{ngandu_extensive_2008}. 
@@
 The observed hitchhiking highlights the importance of linkage due to
-infrequent recombination for the evolution of HIV-1
+infrequent recombination for the evolution of HIV-1. 
+The recombination rate has been estimated to be on the
+order of $\rho = 10^{-5}$ per base and day
 \citep{neher_recombination_2010, batorsky_estimate_2011,
-josefsson_majority_2011}. The recombination rate has been estimated to be on the
-order of $\rho = 10^{-5}$ per base and day. It takes roughly $t_{sw} =
-\epsilon^{-1} |\log \nu_0|$ generations for an escape SNV with escape rate
-$\epsilon$ to rise from an initially low frequency $\nu_0$ to frequency
+josefsson_majority_2011}. It takes roughly $t_{sw} =
+\epsilon^{-1} |\log \nu_0|$ generations for an escape variant with escape rate
+$\epsilon$ to rise from an initially low frequency $\nu_0$ to a frequency close to
 one. This implies that a region of length $l = (\rho t_{sw})^{-1} = \epsilon /
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliographystyle{unsrtnat}
-\bibliography{bib}
-%\onecolumngrid
+%\bibliographystyle{unsrtnat}
+%\bibliography{bib}
+
+\begin{thebibliography}{48}
+\providecommand{\natexlab}[1]{#1}
+\providecommand{\url}[1]{\texttt{#1}}
+\expandafter\ifx\csname urlstyle\endcsname\relax
+  \providecommand{\doi}[1]{doi: #1}\else
+  \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi
+
+\bibitem[Rambaut et~al.(2004)Rambaut, Posada, Crandall, and
+  Holmes]{rambaut_causes_2004}
+Andrew Rambaut, David Posada, Keith~A. Crandall, and Edward~C. Holmes.
+\newblock The causes and consequences of {HIV} evolution.
+\newblock \emph{Nature Reviews Genetics}, 5\penalty0 (1):\penalty0 52--61,
+  January 2004.
+\newblock ISSN 1471-0056.
+\newblock \doi{10.1038/nrg1246}.
+\newblock URL \url{http://www.nature.com/nrg/journal/v5/n1/full/nrg1246.html}.
+
+\bibitem[Mansky and Temin(1995)]{mansky_lower_1995}
+L~M Mansky and H~M Temin.
+\newblock Lower in vivo mutation rate of human immunodeficiency virus type 1
+  than that predicted from the fidelity of purified reverse transcriptase.
+\newblock \emph{Journal of virology}, 69\penalty0 (8):\penalty0 5087--5094,
+  1995.
+\newblock URL \url{file:///home/fabio/university/evolution/references/J.
+  Virol.-1995-Mansky-5087-94.pdf}.
+
+\bibitem[Abram et~al.(2010)Abram, Ferris, Shao, Alvord, and
+  Hughes]{abram_nature_2010}
+Michael~E. Abram, Andrea~L. Ferris, Wei Shao, W.~Gregory Alvord, and Stephen~H.
+  Hughes.
+\newblock Nature, position, and frequency of mutations made in a single cycle
+  of {HIV-1} replication.
+\newblock \emph{Journal of Virology}, 84\penalty0 (19):\penalty0 9864--9878,
+  October 2010.
+\newblock ISSN 0022-{538X}, 1098-5514.
+\newblock \doi{10.1128/JVI.00915-10}.
+\newblock URL \url{http://jvi.asm.org/content/84/19/9864}.
+
+\bibitem[{McMichael} et~al.(2009){McMichael}, Borrow, Tomaras, Goonetilleke,
+  and Haynes]{mcmichael_immune_2009}
+Andrew~J. {McMichael}, Persephone Borrow, Georgia~D. Tomaras, Nilu
+  Goonetilleke, and Barton~F. Haynes.
+\newblock The immune response during acute {HIV-1} infection: clues for vaccine
+  development.
+\newblock \emph{Nature Reviews Immunology}, 10\penalty0 (1):\penalty0 11--23,
+  December 2009.
+\newblock \doi{10.1038/nri2674}.
+\newblock URL \url{http://www.nature.com/doifinder/10.1038/nri2674}.
+
+\bibitem[Richman et~al.(2003)Richman, Wrin, Little, and
+  Petropoulos]{richman_rapid_2003}
+Douglas~D. Richman, Terri Wrin, Susan~J. Little, and Christos~J. Petropoulos.
+\newblock Rapid evolution of the neutralizing antibody response to {HIV} type 1
+  infection.
+\newblock \emph{Proceedings of the National Academy of Sciences}, 100\penalty0
+  (7):\penalty0 4144--4149, April 2003.
+\newblock ISSN 0027-8424, 1091-6490.
+\newblock \doi{10.1073/pnas.0630530100}.
+\newblock URL \url{http://www.pnas.org/content/100/7/4144}.
+
+\bibitem[Bhatt et~al.(2011)Bhatt, Holmes, and Pybus]{Bhatt:2011p43255}
+S~Bhatt, E.~C Holmes, and O.~G Pybus.
+\newblock The genomic rate of molecular adaptation of the human influenza a
+  virus.
+\newblock \emph{Molecular Biology and Evolution}, 28\penalty0 (9):\penalty0
+  2443--2451, Sep 2011.
+\newblock \doi{10.1093/molbev/msr044}.
+\newblock URL
+  \url{http://mbe.oxfordjournals.org/cgi/content/abstract/28/9/2443?etoc}.
+
+\bibitem[Hurst(2002)]{Hurst:2002p32608}
+Laurence~D Hurst.
+\newblock The {Ka/Ks} ratio: diagnosing the form of sequence evolution.
+\newblock \emph{Trends Genet}, 18\penalty0 (9):\penalty0 486, Sep 2002.
+\newblock URL
+  \url{http://www.sciencedirect.com/science?_ob=GatewayURL&_origin=inwardhub&_urlversion=4&_method=citationSearch&_piikey=S0168952502027221&_version=1&md5=994712d847438f6994ce2a092b0f78a2}.
+
+\bibitem[Chen et~al.(2004)Chen, Perlina, and Lee]{Chen:2004p22606}
+Lamei Chen, Alla Perlina, and Christopher~J Lee.
+\newblock Positive selection detection in 40,000 human immunodeficiency virus
+  ({HIV}) type 1 sequences automatically identifies drug resistance and
+  positive fitness mutations in hiv protease and reverse transcriptase.
+\newblock \emph{J Virol}, 78\penalty0 (7):\penalty0 3722--32, Apr 2004.
+\newblock URL
+  \url{http://jvi.asm.org/cgi/content/full/78/7/3722?view=long&pmid=15016892}.
+
+\bibitem[Fernandes et~al.(2012)Fernandes, Jayaraman, and
+  Frankel]{fernandes_hiv-1_2012}
+Jason Fernandes, Bhargavi Jayaraman, and Alan Frankel.
+\newblock The {HIV-1} rev response element: An {RNA} scaffold that directs the
+  cooperative assembly of a homo-oligomeric ribonucleoprotein complex.
+\newblock \emph{{RNA} Biology}, 9\penalty0 (1):\penalty0 4--9, January 2012.
+\newblock ISSN 1547-6286.
+\newblock \doi{10.4161/rna.9.1.18178}.
+\newblock URL
+  \url{http://www.landesbioscience.com/journals/rnabiology/article/18178/?nocache=886329209}.
+
+\bibitem[Watts et~al.(2009)Watts, Dang, Gorelick, Leonard, Jr, Swanstrom,
+  Burch, and Weeks]{watts_architecture_2009}
+Joseph~M. Watts, Kristen~K. Dang, Robert~J. Gorelick, Christopher~W. Leonard,
+  Julian W.~Bess Jr, Ronald Swanstrom, Christina~L. Burch, and Kevin~M. Weeks.
+\newblock Architecture and secondary structure of an entire {HIV-1} {RNA}
+  genome.
+\newblock \emph{Nature}, 460\penalty0 (7256):\penalty0 711--716, August 2009.
+\newblock ISSN 0028-0836.
+\newblock \doi{10.1038/nature08237}.
+\newblock URL
+  \url{http://www.nature.com/nature/journal/v460/n7256/full/nature08237.html}.
+
+\bibitem[Keating et~al.(2009)Keating, Hill, Hawkes, Smyth, Isel, Le,
+  Palmenberg, Marshall, Marquet, Nabel, and Mak]{keating_rich_2009}
+Cameron~P. Keating, Melissa~K. Hill, David~J. Hawkes, Redmond~P. Smyth,
+  Catherine Isel, Shu-Yun Le, Ann~C. Palmenberg, John~A. Marshall, Roland
+  Marquet, Gary~J. Nabel, and Johnson Mak.
+\newblock The {A-rich} {RNA} sequences of {HIV-1} pol are important for the
+  synthesis of viral {cDNA}.
+\newblock \emph{Nucleic Acids Research}, 37\penalty0 (3):\penalty0 945--956,
+  February 2009.
+\newblock ISSN 0305-1048, 1362-4962.
+\newblock \doi{10.1093/nar/gkn1015}.
+\newblock URL \url{http://nar.oxfordjournals.org/content/37/3/945}.
+
+\bibitem[Forsdyke(1995)]{forsdyke_reciprocal_1995}
+{D.R.} Forsdyke.
+\newblock Reciprocal relationship between stem-loop potential and substitution
+  density in retroviral quasispecies under positive {Darwinian} selection.
+\newblock \emph{Journal of Molecular Evolution}, 41\penalty0 (6), December
+  1995.
+\newblock ISSN 0022-2844, 1432-1432.
+\newblock \doi{10.1007/BF00173184}.
+\newblock URL \url{http://www.springerlink.com/content/n7621n18505m418m/}.
+
+\bibitem[Snoeck et~al.(2011)Snoeck, Fellay, Bartha, Douek, and
+  Telenti]{snoeck_mapping_2011}
+Joke Snoeck, Jacques Fellay, István Bartha, Daniel~C Douek, and Amalio
+  Telenti.
+\newblock Mapping of positive selection sites in the {HIV-1} genome in the
+  context of {RNA} and protein structural constraints.
+\newblock \emph{Retrovirology}, 8\penalty0 (1):\penalty0 87, 2011.
+\newblock ISSN 1742-4690.
+\newblock \doi{10.1186/1742-4690-8-87}.
+\newblock URL \url{http://www.retrovirology.com/content/8/1/87/additional}.
+
+\bibitem[Pollom et~al.(2013)Pollom, Dang, Potter, Gorelick, Burch, Weeks, and
+  Swanstrom]{pollom_comparison_2013}
+Elizabeth Pollom, Kristen~K. Dang, E.~Lake Potter, Robert~J. Gorelick,
+  Christina~L. Burch, Kevin~M. Weeks, and Ronald Swanstrom.
+\newblock Comparison of {SIV} and {HIV-1} genomic {RNA} structures reveals
+  impact of sequence evolution on conserved and non-conserved structural
+  motifs.
+\newblock \emph{{PLoS} Pathog}, 9\penalty0 (4):\penalty0 e1003294, April 2013.
+\newblock \doi{10.1371/journal.ppat.1003294}.
+\newblock URL \url{http://dx.doi.org/10.1371/journal.ppat.1003294}.
+
+\bibitem[Neher and Leitner(2010)]{neher_recombination_2010}
+{R.A.} Neher and Thomas Leitner.
+\newblock Recombination rate and selection strength in {HIV} intra-patient
+  evolution.
+\newblock \emph{{PLoS} Comput Biol}, 6\penalty0 (1):\penalty0 e1000660, January
+  2010.
+\newblock \doi{10.1371/journal.pcbi.1000660}.
+\newblock URL \url{http://dx.plos.org/10.1371/journal.pcbi.1000660}.
+
+\bibitem[Batorsky et~al.(2011)Batorsky, Kearney, Palmer, Maldarelli, Rouzine,
+  and Coffin]{batorsky_estimate_2011}
+Rebecca Batorsky, Mary~F Kearney, Sarah~E Palmer, Frank Maldarelli, Igor~M
+  Rouzine, and John~M Coffin.
+\newblock Estimate of effective recombination rate and average selection
+  coefficient for {HIV} in chronic infection.
+\newblock \emph{Proceedings of the National Academy of Sciences of the United
+  States of America}, 108\penalty0 (14):\penalty0 5661--6, April 2011.
+\newblock \doi{10.1073/pnas.1102036108}.
+\newblock URL
+  \url{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3078368&tool=pmcentrez&rendertype=abstract}.
+
+\bibitem[Shankarappa et~al.(1999)Shankarappa, Margolick, Gange, Rodrigo,
+  Upchurch, Farzadegan, Gupta, Rinaldo, Learn, He,
+  et~al.]{shankarappa_consistent_1999}
+R.~Shankarappa, {J.B.} Margolick, {S.J.} Gange, {A.G.} Rodrigo, David Upchurch,
+  Homayoon Farzadegan, Phalguni Gupta, {C.R.} Rinaldo, {G.H.} Learn, X.~He,
+  et~al.
+\newblock Consistent viral evolutionary changes associated with the progression
+  of human immunodeficiency virus type 1 infection.
+\newblock \emph{Journal of Virology}, 73\penalty0 (12):\penalty0 10489, 1999.
+\newblock URL \url{http://jvi.asm.org/cgi/content/abstract/73/12/10489}.
+
+\bibitem[Liu et~al.(2006)Liu, {McNevin}, Cao, Zhao, Genowati, Wong,
+  {McLaughlin}, {McSweyn}, Diem, Stevens, et~al.]{liu_selection_2006}
+Yi~Liu, J.~{McNevin}, Jianhong Cao, Hong Zhao, Indira Genowati, Kim Wong,
+  S.~{McLaughlin}, {M.D.} {McSweyn}, Kurt Diem, {C.E.} Stevens, et~al.
+\newblock Selection on the human immunodeficiency virus type 1 proteome
+  following primary infection.
+\newblock \emph{Journal of virology}, 80\penalty0 (19):\penalty0 9519, 2006.
+\newblock \doi{10.1128/JVI.00575-06}.
+\newblock URL \url{http://jvi.asm.org/cgi/content/abstract/80/19/9519}.
+
+\bibitem[Bunnik et~al.(2008)Bunnik, Pisas, Van~Nuenen, and
+  Schuitemaker]{bunnik_autologous_2008}
+{E.M.} Bunnik, Linaida Pisas, {A.C.} Van~Nuenen, and Hanneke Schuitemaker.
+\newblock Autologous neutralizing humoral immunity and evolution of the viral
+  envelope in the course of subtype b human immunodeficiency virus type 1
+  infection.
+\newblock \emph{Journal of virology}, 82\penalty0 (16):\penalty0 7932, 2008.
+\newblock ISSN 3120566829.
+\newblock \doi{10.1128/JVI.00757-08}.
+\newblock URL \url{http://jvi.asm.org/cgi/content/abstract/82/16/7932}.
+
+\bibitem[Kuiken et~al.(2012)Kuiken, Leitner, Hahn, Mullins, Wolinsky, Foley,
+  Apetrei, Mizrachi, Rambaut, and Korber]{LANL2012}
+Carla Kuiken, Thomas Leitner, Beatrice Hahn, James Mullins, Steven Wolinsky,
+  Brian Foley, Cristian Apetrei, Ilene Mizrachi, Andrew Rambaut, and Bette
+  Korber.
+\newblock \emph{HIV Sequence Compendium 2012}.
+\newblock Theoretical Biology and Biophysics Group T-6, Mail Stop K710 Los
+  Alamos National Laboratory Los Alamos, New Mexico 87545 U.S.A., 2012.
+
+\bibitem[Edgar(2004)]{edgar_muscle:_2004}
+Robert~C. Edgar.
+\newblock {MUSCLE:} multiple sequence alignment with high accuracy and high
+  throughput.
+\newblock \emph{Nucleic Acids Research}, 32\penalty0 (5):\penalty0 1792--1797,
+  March 2004.
+\newblock ISSN 0305-1048, 1362-4962.
+\newblock \doi{10.1093/nar/gkh340}.
+\newblock URL \url{http://nar.oxfordjournals.org/content/32/5/1792}.
+
+\bibitem[Zanini and Neher(2012)]{zanini_ffpopsim:_2012}
+Fabio Zanini and Richard~A. Neher.
+\newblock {FFPopSim:} an efficient forward simulation package for the evolution
+  of large populations.
+\newblock \emph{Bioinformatics}, October 2012.
+\newblock ISSN 1367-4803, 1460-2059.
+\newblock \doi{10.1093/bioinformatics/bts633}.
+\newblock URL
+  \url{http://bioinformatics.oxfordjournals.org/content/early/2012/10/24/bioinformatics.bts633}.
+
+\bibitem[Perelson et~al.(1996)Perelson, Neumann, Markowitz, Leonard, and
+  Ho]{perelson_hiv-1_1996}
+a~S Perelson, a~U Neumann, M~Markowitz, J~M Leonard, and D~D Ho.
+\newblock Hiv-1 dynamics in vivo: virion clearance rate, infected cell
+  life-span, and viral generation time.
+\newblock \emph{Science {(New York, N.Y.)}}, 271\penalty0 (5255):\penalty0
+  1582--6, March 1996.
+\newblock URL \url{http://www.ncbi.nlm.nih.gov/pubmed/8599114}.
+
+\bibitem[Markowitz et~al.(2003)Markowitz, Louie, Hurley, Sun, Di~Mascio,
+  Perelson, and Ho]{markowitz_novel_2003}
+Martin Markowitz, Michael Louie, Arlene Hurley, Eugene Sun, Michele Di~Mascio,
+  Alan~S Perelson, and David~D Ho.
+\newblock A novel antiviral intervention results in more accurate assessment of
+  human immunodeficiency virus type 1 replication dynamics and {T-cell} decay
+  in vivo.
+\newblock \emph{J Virol}, 77\penalty0 (8):\penalty0 5037--5038, April 2003.
+
+\bibitem[Levy et~al.(2004)Levy, Aldrovandi, Kutsch, and
+  Shaw]{levy_dynamics_2004}
+David~N Levy, Grace~M Aldrovandi, Olaf Kutsch, and George~M Shaw.
+\newblock Dynamics of {HIV}-1 recombination in its natural target cells.
+\newblock \emph{Proc Natl Acad Sci USA}, 101\penalty0 (12):\penalty0 4204--9,
+  Mar 2004.
+\newblock \doi{10.1073/pnas.0306764101}.
+\newblock URL \url{http://www.pnas.org/content/101/12/4204.long}.
+
+\bibitem[Josefsson et~al.(2013)Josefsson, Palmer, Faria, Lemey, Casazza,
+  Ambrozak, Kearney, Shao, Kottilil, Sneller, Mellors, Coffin, and
+  Maldarelli]{josefsson_single_2013}
+Lina Josefsson, Sarah Palmer, Nuno~R. Faria, Philippe Lemey, Joseph Casazza,
+  David Ambrozak, Mary Kearney, Wei Shao, Shyamasundaran Kottilil, Michael
+  Sneller, John Mellors, John~M. Coffin, and Frank Maldarelli.
+\newblock Single cell analysis of lymph node tissue from {HIV-1} infected
+  patients reveals that the majority of {CD4+} t-cells contain one {HIV-1}
+  {DNA} molecule.
+\newblock \emph{{PLoS} Pathog}, 9\penalty0 (6):\penalty0 e1003432, June 2013.
+\newblock \doi{10.1371/journal.ppat.1003432}.
+\newblock URL \url{http://dx.doi.org/10.1371/journal.ppat.1003432}.
+
+\bibitem[Boltz et~al.(2012)Boltz, Ambrose, Kearney, Shao, {KewalRamani},
+  Maldarelli, Mellors, and Coffin]{boltz_ultrasensitive_2012}
+Valerie~F. Boltz, Zandrea Ambrose, Mary~F. Kearney, Wei Shao, Vineet~N.
+  {KewalRamani}, Frank Maldarelli, John~W. Mellors, and John~M. Coffin.
+\newblock Ultrasensitive allele-specific {PCR} reveals rare preexisting
+  drug-resistant variants and a large replicating virus population in macaques
+  infected with a simian immunodeficiency virus containing human
+  immunodeficiency virus reverse transcriptase.
+\newblock \emph{Journal of Virology}, 86\penalty0 (23):\penalty0 12525--12530,
+  December 2012.
+\newblock ISSN 0022-{538X}, 1098-5514.
+\newblock \doi{10.1128/JVI.01963-12}.
+\newblock URL \url{http://jvi.asm.org/content/86/23/12525}.
+
+\bibitem[Neher and Shraiman(2011)]{neher_genetic_2011}
+Richard~A Neher and Boris Shraiman.
+\newblock Genetic draft and quasi-neutrality in large facultatively sexual
+  populations.
+\newblock \emph{Genetics}, 188\penalty0 (4):\penalty0 975--996, 2011.
+\newblock URL \url{http://www.genetics.org/content/188/4/975.short}.
+
+\bibitem[Asquith et~al.(2006)Asquith, Edwards, Lipsitch, and
+  McLean]{Asquith:2006p28003}
+Becca Asquith, Charles T~T Edwards, Marc Lipsitch, and Angela~R McLean.
+\newblock Inefficient cytotoxic t lymphocyte-mediated killing of
+  {HIV}-1-infected cells in vivo.
+\newblock \emph{PLoS Biol}, 4\penalty0 (4):\penalty0 e90, Apr 2006.
+\newblock \doi{10.1371/journal.pbio.0040090}.
+\newblock URL
+  \url{http://www.plosbiology.org/article/info%253Adoi%252F10.1371%252Fjournal.pbio.0040090}.
+
+\bibitem[Moore et~al.(2009)Moore, Ranchobe, Lambson, Gray, Cave, Abrahams,
+  Bandawe, Mlisana, Abdool~Karim, Williamson, Morris, the {CAPRISA} 002~study,
+  and the {NIAID} Center~for {HIV/AIDS} Vaccine
+  Immunology~{(CHAVI)}]{moore_limited_2009}
+Penny~L. Moore, Nthabeleng Ranchobe, Bronwen~E. Lambson, Elin~S. Gray, Eleanor
+  Cave, Melissa-Rose Abrahams, Gama Bandawe, Koleka Mlisana, Salim~S.
+  Abdool~Karim, Carolyn Williamson, Lynn Morris, the {CAPRISA} 002~study, and
+  the {NIAID} Center~for {HIV/AIDS} Vaccine Immunology~{(CHAVI)}.
+\newblock Limited neutralizing antibody specificities drive neutralization
+  escape in early {HIV-1} subtype {C} infection.
+\newblock \emph{PLoS Pathog}, 5\penalty0 (9):\penalty0 e1000598, September
+  2009.
+\newblock \doi{10.1371/journal.ppat.1000598}.
+\newblock URL \url{http://dx.doi.org/10.1371/journal.ppat.1000598}.
+
+\bibitem[{McKay} et~al.(1979){McKay}, Beckman, and
+  Conover]{mckay_comparison_1979}
+M.~D. {McKay}, R.~J. Beckman, and W.~J. Conover.
+\newblock A comparison of three methods for selecting values of input variables
+  in the analysis of output from a computer code.
+\newblock \emph{Technometrics}, 21\penalty0 (2):\penalty0 239--245, May 1979.
+\newblock ISSN 0040-1706.
+\newblock \doi{10.2307/1268522}.
+\newblock URL \url{http://www.jstor.org/stable/1268522}.
+\newblock {ArticleType:} research-article / Full publication date: May, 1979 /
+  Copyright © 1979 American Statistical Association and American Society for
+  Quality.
+
+\bibitem[Coffin(1995)]{coffin_hiv_1995}
+John~M Coffin.
+\newblock {HIV} population dynamics in vivo: implications for genetic
+  variation, pathogenesis, and therapy.
+\newblock \emph{Science}, 267\penalty0 (5197):\penalty0 483--489, January 1995.
+\newblock ISSN 0036-8075.
+
+\bibitem[Sanjuan and Borderia(2011)]{sanjuan_interplay_2011}
+Rafael Sanjuan and Antonio~V. Borderia.
+\newblock Interplay between {RNA} structure and protein evolution in {HIV-1}.
+\newblock \emph{Molecular Biology and Evolution}, 28\penalty0 (4):\penalty0
+  1333--1338, April 2011.
+\newblock ISSN 0737-4038, 1537-1719.
+\newblock \doi{10.1093/molbev/msq329}.
+\newblock URL \url{http://mbe.oxfordjournals.org/content/28/4/1333}.
+
+\bibitem[Jenkins and Holmes(2003)]{jenkins_extent_2003}
+Gareth~M Jenkins and Edward~C Holmes.
+\newblock The extent of codon usage bias in human {RNA} viruses and its
+  evolutionary origin.
+\newblock \emph{Virus Research}, 92\penalty0 (1):\penalty0 1--7, March 2003.
+\newblock ISSN 0168-1702.
+\newblock \doi{10.1016/S0168-1702(02)00309-X}.
+\newblock URL
+  \url{http://www.sciencedirect.com/science/article/pii/S016817020200309X}.
+
+\bibitem[Kuyl and Berkhout(2012)]{kuyl_biased_2012}
+Antoinette C. van~der Kuyl and Ben Berkhout.
+\newblock The biased nucleotide composition of the {HIV} genome: a constant
+  factor in a highly variable virus.
+\newblock \emph{Retrovirology}, 9\penalty0 (1):\penalty0 92, November 2012.
+\newblock ISSN 1742-4690.
+\newblock \doi{10.1186/1742-4690-9-92}.
+\newblock URL \url{http://www.retrovirology.com/content/9/1/92/abstract}.
+
+\bibitem[Williamson(2003)]{williamson_adaptation_2003}
+Scott Williamson.
+\newblock Adaptation in the env gene of {HIV-1} and evolutionary theories of
+  disease progression.
+\newblock \emph{Molecular biology and evolution}, 20\penalty0 (8):\penalty0
+  1318--25, 2003.
+\newblock \doi{10.1093/molbev/msg144}.
+\newblock URL \url{http://www.ncbi.nlm.nih.gov/pubmed/12777505}.
+
+\bibitem[Wei et~al.(2003)Wei, Decker, Wang, Hui, Kappes, Wu, Salazar-Gonzalez,
+  Salazar, Kilby, Saag, Komarova, Nowak, Hahn, Kwong, and
+  Shaw]{wei_antibody_2003}
+Xiping Wei, Julie~M. Decker, Shuyi Wang, Huxiong Hui, John~C. Kappes, Xiaoyun
+  Wu, Jesus~F. Salazar-Gonzalez, Maria~G. Salazar, J.~Michael Kilby, Michael~S.
+  Saag, Natalia~L. Komarova, Martin~A. Nowak, Beatrice~H. Hahn, Peter~D. Kwong,
+  and George~M. Shaw.
+\newblock Antibody neutralization and escape by {HIV-1}.
+\newblock \emph{Nature}, 422\penalty0 (6929):\penalty0 307--312, March 2003.
+\newblock ISSN 0028-0836.
+\newblock \doi{10.1038/nature01470}.
+\newblock URL
+  \url{http://www.nature.com/nature/journal/v422/n6929/full/nature01470.html}.
+
+\bibitem[Smith and Haigh(1974)]{smith_hitch-hiking_1974}
+J~M Smith and J~Haigh.
+\newblock The hitch-hiking effect of a favourable gene.
+\newblock \emph{Genetical research}, 23\penalty0 (1):\penalty0 23--35, February
+  1974.
+\newblock {PMID:} 4407212.
+
+\bibitem[Gillespie(2000)]{gillespie_genetic_2000}
+J~H Gillespie.
+\newblock Genetic drift in an infinite population. the pseudohitchhiking model.
+\newblock \emph{Genetics}, 155\penalty0 (2):\penalty0 909--19, July 2000.
+\newblock URL
+  \url{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=1461093&tool=pmcentrez&rendertype=abstract}.
+
+\bibitem[Bar et~al.(2012)Bar, Tsao, Iyer, Decker, Yang, Bonsignori, Chen,
+  Hwang, Montefiori, Liao, Hraber, Fischer, Li, Wang, Sterrett, Keele, Ganusov,
+  Perelson, Korber, Georgiev, {McLellan}, Pavlicek, Gao, Haynes, Hahn, Kwong,
+  and Shaw]{bar_early_2012}
+Katharine~J. Bar, Chun-yen Tsao, Shilpa~S. Iyer, Julie~M. Decker, Yongping
+  Yang, Mattia Bonsignori, Xi~Chen, Kwan-Ki Hwang, David~C. Montefiori, Hua-Xin
+  Liao, Peter Hraber, William Fischer, Hui Li, Shuyi Wang, Sarah Sterrett,
+  Brandon~F. Keele, Vitaly~V. Ganusov, Alan~S. Perelson, Bette~T. Korber,
+  Ivelin Georgiev, Jason~S. {McLellan}, Jeffrey~W. Pavlicek, Feng Gao,
+  Barton~F. Haynes, Beatrice~H. Hahn, Peter~D. Kwong, and George~M. Shaw.
+\newblock Early low-titer neutralizing antibodies impede {HIV-1} replication
+  and select for virus escape.
+\newblock \emph{{PLoS} Pathog}, 8\penalty0 (5):\penalty0 e1002721, May 2012.
+\newblock \doi{10.1371/journal.ppat.1002721}.
+\newblock URL \url{http://dx.doi.org/10.1371/journal.ppat.1002721}.
+
+\bibitem[Hedskog et~al.(2010)Hedskog, Mild, Jernberg, Sherwood, Bratt, Leitner,
+  Lundeberg, Andersson, and Albert]{hedskog_dynamics_2010}
+Charlotte Hedskog, Mattias Mild, Johanna Jernberg, Ellen Sherwood, Göran
+  Bratt, Thomas Leitner, Joakim Lundeberg, Björn Andersson, and Jan Albert.
+\newblock Dynamics of {HIV-1} quasispecies during antiviral treatment dissected
+  using ultra-deep pyrosequencing.
+\newblock \emph{{PLoS} {ONE}}, 5\penalty0 (7):\penalty0 e11345, July 2010.
+\newblock \doi{10.1371/journal.pone.0011345}.
+\newblock URL \url{http://dx.doi.org/10.1371/journal.pone.0011345}.
+
+\bibitem[Fischer et~al.(2010)Fischer, Ganusov, Giorgi, Hraber, Keele, Leitner,
+  Han, Gleasner, Green, Lo, Nag, Wallstrom, Wang, {McMichael}, Haynes, Hahn,
+  Perelson, Borrow, Shaw, Bhattacharya, and Korber]{fischer_transmission_2010}
+Will Fischer, Vitaly~V. Ganusov, Elena~E. Giorgi, Peter~T. Hraber, Brandon~F.
+  Keele, Thomas Leitner, Cliff~S. Han, Cheryl~D. Gleasner, Lance Green,
+  Chien-Chi Lo, Ambarish Nag, Timothy~C. Wallstrom, Shuyi Wang, Andrew~J.
+  {McMichael}, Barton~F. Haynes, Beatrice~H. Hahn, Alan~S. Perelson, Persephone
+  Borrow, George~M. Shaw, Tanmoy Bhattacharya, and Bette~T. Korber.
+\newblock Transmission of single {HIV-1} genomes and dynamics of early immune
+  escape revealed by ultra-deep sequencing.
+\newblock \emph{{PLoS} {ONE}}, 5\penalty0 (8):\penalty0 e12303, August 2010.
+\newblock \doi{10.1371/journal.pone.0012303}.
+\newblock URL \url{http://dx.doi.org/10.1371/journal.pone.0012303}.
+
+\bibitem[Knoepfel and Berkhout(2013)]{knoepfel_role_2013}
+S.~A. Knoepfel and B.~Berkhout.
+\newblock On the role of four small hairpins in the {HIV-1} {RNA} genome.
+\newblock \emph{{RNA} Biology}, 10\penalty0 (4):\penalty0 512--524, March 2013.
+\newblock URL
+  \url{http://www.landesbioscience.com/journals/rnabiology/article/24133/}.
+
+\bibitem[Mayrose et~al.(2007)Mayrose, Doron-Faigenboim, Bacharach, and
+  Pupko]{mayrose_towards_2007}
+Itay Mayrose, Adi Doron-Faigenboim, Eran Bacharach, and Tal Pupko.
+\newblock Towards realistic codon models: among site variability and dependency
+  of synonymous and non-synonymous rates.
+\newblock \emph{Bioinformatics}, 23\penalty0 (13):\penalty0 i319--i327, July
+  2007.
+\newblock ISSN 1367-4803, 1460-2059.
+\newblock \doi{10.1093/bioinformatics/btm176}.
+\newblock URL
+  \url{http://bioinformatics.oxfordjournals.org/content/23/13/i319}.
+
+\bibitem[Ngandu et~al.(2008)Ngandu, Scheffler, Moore, Woodman, Martin, and
+  Seoighe]{ngandu_extensive_2008}
+Nobubelo~K. Ngandu, Konrad Scheffler, Penny Moore, Zenda Woodman, Darren
+  Martin, and Cathal Seoighe.
+\newblock Extensive purifying selection acting on synonymous sites in {HIV-1}
+  group m sequences.
+\newblock \emph{Virology Journal}, 5\penalty0 (1):\penalty0 160, December 2008.
+\newblock ISSN 1743-{422X}.
+\newblock \doi{10.1186/1743-422X-5-160}.
+\newblock URL \url{http://www.virologyj.com/content/5/1/160/abstract}.
+\newblock {PMID:} 19105834.
+
+\bibitem[Ferguson et~al.(2013)Ferguson, Mann, Omarjee, Ndung'u, Walker, and
+  Chakraborty]{ferguson_translating_2013}
+Andrew L. Ferguson, Jaclyn K. Mann, Saleha Omarjee, Thumbi Ndung'u, Bruce D.
+  Walker, and Arup K. Chakraborty.
+\newblock Translating {HIV} sequences into quantitative fitness landscapes
+  predicts viral vulnerabilities for rational immunogen design.
+\newblock \emph{Immunity}, 38\penalty0 (3):\penalty0 606--617, March 2013.
+\newblock ISSN 1074-7613.
+\newblock \doi{10.1016/j.immuni.2012.11.022}.
+\newblock URL
+  \url{http://www.sciencedirect.com/science/article/pii/S1074761313001076}.
+
+\bibitem[Josefsson et~al.(2011)Josefsson, King, Makitalo, Br\"annstr\"om, Shao,
+  Maldarelli, Kearney, Hu, Chen, Gaines, Mellors, Albert, Coffin, and
+  Palmer]{josefsson_majority_2011}
+Lina Josefsson, Martin~S. King, Barbro Makitalo, Johan Br\"annstr\"om, Wei
+  Shao, Frank Maldarelli, Mary~F. Kearney, Wei-Shau Hu, Jianbo Chen, Hans
+  Gaines, John~W. Mellors, Jan Albert, John~M. Coffin, and Sarah~E. Palmer.
+\newblock Majority of {CD4+} t cells from peripheral blood of {HIV-1} infected
+  individuals contain only one {HIV} {DNA} molecule.
+\newblock \emph{Proceedings of the National Academy of Sciences}, 108\penalty0
+  (27):\penalty0 11199--11204, July 2011.
+\newblock ISSN 0027-8424, 1091-6490.
+\newblock \doi{10.1073/pnas.1107729108}.
+\newblock URL \url{http://www.pnas.org/content/108/27/11199}.
+
+\bibitem[Nijhuis et~al.(1998)Nijhuis, Boucher, Schipper, Leitner, Schuurman,
+  and Albert]{nijhuis_stochastic_1998}
+Monique Nijhuis, Charles A.~B. Boucher, Pauline Schipper, Thomas Leitner, Rob
+  Schuurman, and Jan Albert.
+\newblock Stochastic processes strongly influence {HIV-1} evolution during
+  suboptimal protease-inhibitor therapy.
+\newblock \emph{Proceedings of the National Academy of Sciences}, 95\penalty0
+  (24):\penalty0 14441--14446, November 1998.
+\newblock ISSN 0027-8424, 1091-6490.
+\newblock \doi{10.1073/pnas.95.24.14441}.
+\newblock URL \url{http://www.pnas.org/content/95/24/14441}.
+
+\end{thebibliography}
 
@@
 \includegraphics[width=0.6\linewidth]{fig1}
-%\subfloat{\includegraphics[width=0.6\linewidth]
-%{Shankarappa_allele_freqs_trajectories_syn_p10.pdf}
-%\label{fig:aftsyn}}\\
-% \subfloat{\includegraphics[width=0.6\linewidth]
-%{Shankarappa_allele_freqs_trajectories_nonsyn_p10.pdf}
-%\label{fig:aftnonsyn}}
 \caption{Time series of frequencies
@@
 \includegraphics[width=0.6\linewidth]{fig2}
-%\subfloat{\includegraphics[width=0.6\linewidth]{fixation_times.pdf}
-%\label{fig:fixp1}}\\
-%\subfloat{\includegraphics[width=0.6\linewidth]{fixation_probabilities.pdf}
-%\label{fig:fixp2}}
 \caption{Fixation and loss of SNVs.
@@
 \includegraphics[width=0.9\linewidth]{fig3}
-%\subfloat{\includegraphics[height=0.3\linewidth]{reactivities_histograms_syn.pdf}\label{fig:SHAPEA}}
-%\subfloat{\includegraphics[height=0.3\linewidth]{fixation_probabilities_VnonV.pdf}\label{fig:SHAPEB}}
 \caption{Permissible synonymous mutations tend to be unpaired.
@@
 \includegraphics[width=0.6\linewidth]{fig4}
-%\subfloat{\includegraphics[width=0.6\linewidth]{simulations_graduallydel.pdf}
-%\label{fig:simfixpvar}}\\
-%\subfloat{\includegraphics[width=0.6\linewidth]{simulations_syn_latin_contour.pdf}
-%\label{fig:simsfig}}
 \caption{Distribution of selection coefficients on synonymous sites. Panel A)
@@
 diversity, measured by the probability of a SNV to be found at
-intermediate frequencies $P_\text{interm}$ (first inset). The area observed in
+intermediate frequencies $P_\mathrm{interm}$ (first inset). The area observed in
 the data is also shown (second inset).
@@
 deleterious. The marginal distributions of each parameter are given in
-\figurename~S\marginals.} 
+Fig.~S\marginals.} 
 \label{fig:simheat}
@@
 
-%\setcounter{figure}{0}
-%\include{supplement_body}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2015-07-30 response_to_reviewers.txt

@@
-##############################################################################
-# Reviewer comments:
 ##############################################################################
@@
 readily measured in days and we have chosen to model HIV evolution with
-generation of one day for simplicity. Factor of 2 changes in the
-generation have little influence on the dynamics and only renormalize
-genetic drift. Genetic drift is not a dominant force in evolution of the
-HIV envelope gene. The (effective) population size, the main determinant of drift,
-is not known to this accuracy anyway. We have repeated a subset of the
-simulations with a generation time of one day, finding no
-difference. Our more exhaustive survey of the parameter space includes a
-10-fold variation of the population size. Our results are very
-insensitive and hence also insensitive to the generation time.
+generation of one day for simplicity. We have repeated some of the
+simulations with a generation time of 2 days and find the same results.
+In fact, minor changes in the generation have little influence on the
+dynamics and only renormalize genetic drift. Since the main determinant
+of genetic drift -- the (effective) population size -- is not known
+accurately, we don't think the difference of choosing one or two days as
+generation time is important. Our more exhaustive survey of the
+parameter space includes a 5-fold variation of the population size. Our
+results are very insensitive to the population size and hence also
+insensitive to the generation time.
 
@@
 different values for the mutation rate, the recombination rate, the
-generation time, the fitness effects, and the population size. To
-accomodate this high dimensional data, we have changed figure 4. The
-figure now shows the joint distribution of the deleterious effect and
-the fraction of synonymous mutations that are deleterious that are
-compatible with observations as a heat map. The marginal distribution of
-the individual parameters consistent with the data are shown in the supplement.
+fitness effects of synonymous and non-synonymous mutations, and the
+population size. To accomodate this high dimensional data, we have
+changed figure 4. The figure now shows the joint distribution of the
+deleterious effect and the fraction of synonymous mutations that are
+deleterious that are compatible with observations as a contour plot. The
+marginal distributions of the individual parameters consistent with the
+data are shown in the supplement.
 
-The distribution of allowed deleterious effects is again peaked around
+The distribution of compatible deleterious effects is again peaked around
 0.002 and the fraction of "synonymous" mutations that can be neutral is
@@
 the mutations would fix with a probability equal to their frequency,
-regardless of whether they are subject to drift of draft. We have added
-a sentence to the discussion to make this point clear.
+regardless of whether they are subject to drift of hitchhiking. We have
+added a sentence to the discussion to make this point clear.
 

### 2015-07-30 supplement_body.tex
fixed supplementary marginals

--- 2013-05-31 supplement_body.tex
@@
 The other parameters for the shown simulations are
-the following: deleterious effect $s_d = 10^{-3}$, average escape rate $\epsilon = 0.016$,
-fraction of deleterious synonymous mutations $\alpha = 0.986$, rate of new epitopes
-$k_A=0.0014$ per generation.
+the following: deleterious effect $s_d = 0.002$, average escape rate $\epsilon
+= 0.032$ per day, fraction of deleterious synonymous mutations $\alpha = 0.986$, rate of new epitopes
+$k_A=0.0014$ per day.
 }
@@
 compete and only one of them fixes. The other parameters for the shown simulations are
-the following: deleterious effect $s_d = 10^{-3}$, average escape rate $\epsilon = 0.016$,
-fraction of deleterious synonymous mutations $\alpha = 0.986$, rate of new epitopes
-$k_A=0.0014$ per generation.
+the following: deleterious effect $s_d = 0.002$, average escape rate
+$\epsilon = 0.032$ per day, fraction of deleterious synonymous mutations $\alpha = 0.986$, rate of new epitopes
+$k_A=0.0014$ per day.
 }
@@
 \end{figure}
+
+\newpage
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\section{Simulations with different parameters}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \begin{figure}[h]
+% \begin{center}
+% \includegraphics[width=0.4\linewidth]{simulations_syn_gentime_2days.pdf}\\
+% \includegraphics[width=0.24\linewidth]{simulations_syn_mu_1e-5.pdf}
+% \includegraphics[width=0.24\linewidth]{simulations_syn_latin_suppl.pdf}
+% \includegraphics[width=0.24\linewidth]{simulations_syn_mu_3e-5.pdf}
+% \includegraphics[width=0.24\linewidth]{simulations_syn_mu_4e-5.pdf}\\
+% \includegraphics[width=0.32\linewidth]{simulations_syn_coi_5e-3.pdf}
+% \includegraphics[width=0.32\linewidth]{simulations_syn_latin_suppl2.pdf}
+% \includegraphics[width=0.32\linewidth]{simulations_syn_coi_2e-2.pdf}
+% \caption{
+% Results of simulations with alternative parameters.
+% \textbf{Panel A)} Like \figurename~4B in the main text, but simulations are performed using a
+% generation time of two days. Two selection coefficient of synonymous sites
+% doubles in units of 1/generations, hence keeps constants in units of 1/days.
+% \textbf{Panels B, C, D, and E)} Effect of mutation rate on the simulation
+% results. The mutation rates are respectively $1 \times 10^{-5}$ (B), $2 \times
+% 10^{-5}$ (C), $3 \times 10^{-5}$ (D), and $4 \times
+% 10^{-5}$ (E) per generation per site. Increasing the mutation rate tightens the
+% requirement that almost all synonymous mutations must be deleterious. This can
+% be rationalized as follows: a higher $\mu$ increases the neutral synonymous
+% diversity (we have measured this directly in our simulations), hence more and
+% more synonymous SNVs that reach high frequency are neutral and the
+% effect of deleterious SNVs, although present, becomes too weak to be observed as
+% a depression in $P_\text{fix}$. As a consequence, only simulations with almost
+% no neutral sites still show a reduced $P_\text{fix}$ and are shown in the
+% figures.
+% \textbf{Panels F, G, H)} Effect of recombination rate on the simulation results.
+% The recombination rates are respectively $0.5 \times 10^{-5}$ (F), $10^{-5}$
+% (G), and $2 \times 10^{-5}$ (H) per generation per site. Recombination does not
+% change the results qualitatively in terms of fixation probability. Simulations
+% with high recombination rates, however, have an excessively high synonymous
+% diversity compared to the data (dots at the bottom and right side of panel H).
+% More stringent gating around the data solves this issue.
+% }
+% \label{fig:sims_altparameters}
+% \end{center}
+% \end{figure}
+
+\newpage
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Marginal distributions in simulations}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=0.99\linewidth]{simulations_marginals_stratified.pdf}
+\caption{
+Marginal distributions of simulation parameters that reproduce the
+synonymous fixation probability and diversity seen in the data. Each
+panel contains the prior distribution of parameters (prior), the distribution of
+parameters after imposing the lower or upper bounds on synonymous
+diversity (syn diversity $\gtrless$), the bounds on the fixation
+probability (Area of $P_{fix}$), and the all bounds
+simultaneously (gated). The distribution of population sizes, the recombination rates, the escape
+rate of new epitopes, and the mutation rate remain approximately
+flat. They only two directions parameters whose distribution is strongly
+constrained by the data is the cost of synonymous mutations and the
+fraction of synonymous mutations that are neutral. We note that the
+distribution of compatible parameters is most likely a complicated
+subspace of the high dimensional parameter space and interpretation of
+these marginal projections is difficult. 
+}
+\label{fig:sims_altparameters}
+\end{center}
+\end{figure}
+
+% \newpage
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \section{Divergence and diversity over time in simulations}
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \begin{figure}[h]
+% \begin{center}
+% \includegraphics[width=0.6\linewidth]{simulations_diversity_divergence_example.pdf}
+% \caption{
+% Diversity and divergence over time in an example simulation with realistic
+% parameters. Divergence from the founder genotype accumulates linearly  in time. On the
+% contrary, because of the frequent selective sweeps of escape mutations, diversity saturates
+% relatively quickly. The population dynamics after that point is in a steady
+% state.
+% }
+% \label{fig:sims_altparameters}
+% \end{center}
+% \end{figure}
